\newpage

# Chapter Four 

# Introduction

On October 10, 2020, then Presidential candidate Joe Biden's campaign team aired an advertisement touting the former Vice President's national security credentials. This particular advertisement is narrated by a man identified as Jim, an "Iraq War Veteran," who, with a personal touch, describes the efforts of the former Vice President in delivering Mine Resistant Ambush Protected (MRAP) vehicles to Iraq and Afghanistan. These vehicles were known for their heavy armor, which likely saved the lives and limbs of servicemen and women who were deployed at the time to the Middle East. The advertisement was later pulled by the Biden Campaign after a retired Army Lieutenant General, Sean MacFarland, objected to the advertisement on the grounds that the ad used images of him without his consent [@moore_biden_2020]. The images in question showed then Lieutenant General MacFarland, in uniform, shaking hands with Biden as the former Vice President arrived to Iraq [@moore_biden_2020]. 

In objecting to the ad, MacFarland noted that he had not granted permission to the Biden Campaign to use images which featured the general in such a recognizable manner. Further, MacFarland expressed concern that voters would assume that his appearance in the ad constituted an endorsement of Biden, and noted that, "A number of people have understandably inferred that my appearance [in this Biden Campaign advertisement] constitutes an endorsement of the former Vice President. It does not. To be clear, I have not endorsed President Trump, either. I object to the use of ANY military personnel in uniform in political ads - full stop" [MacFarland quoted in @moore_biden_2020]. 

This incident illustrates one of the many ways in which political candidates use the military, especially during elections.^[To be clear, President Trump's campaign likewise used military symbols and imagery in many advertisements. For a salient example in which the Chairman of the Joint Chiefs of Staff appeared in one of Trump's advertisements, see @seligman_top_2020.] While civilians have long appealed to the public by harnessing the popularity of the military or of individual military figures in a variety of ways, the practice of doing so during a closely contested election season underscores how a non-partisan institution can find itself - willingly, or in the case of MacFarland, unwillingly - thrust into the political spotlight.   

This chapter examines variation in the use of the military during Presidential campaign television advertisements. The content of campaign advertisements is primarily the purview of civilian rather than military actors, and thus, examining this particular form of behavior in depth will facilitate a comparison between political behaviors undertaken by military actors (the focus of last chapter) vice civilian actors. Of interest is whether any discernible variation can be detected in the degree to which civilian actors have used the military in ways that violate the principles of civil-military relations, and to account for this variation. 

This chapter proceeds in four parts. First, this chapter briefly revisits the theory argued previously in this dissertation, which posits that the levels of political polarization and the prestige of the military impact the degree to which military and civilian actors adhere to the central principles of civil-military relations. In the second part of the chapter, I present the methodology and the data used in this chapter, which include roughly 3,600 presidential campaign television advertisements aired on behalf of the five elections occurring inclusively from 2000 - 2016. With the exception of some of the data that is missing, these advertisements comprise all presidential campaign television advertisements used in these elections.^[So far as I know, this dissertation chapter comprises the only scholarly effort undertaken with this particular data for the purposes of coding the use of military symbols, actors, and behaviors.]

The third part of the chapter explores and analyzes the data in detail. The analysis first explores variation in the types of military imagery that have been featured in campaign advertisements, and then explores variation in the ways that various military actors have committed expressly partisan acts during these advertisements, such as endorsing or attacking a political candidate. The results show that there is significant variation in how civilians have used the military, to include its symbols and actors, in presidential campaign television advertisements over the time period examined. Furthermore, the results provide moderately strong support for the notion that the level of military prestige drives the choice of civilians to use the military in ways that challenge one or more of the central principles of civil-military relations, and especially the principle of non-partisanship. In the fourth and final part of the chapter, I conclude by integrating the results of this chapter with the results obtained in the previous chapter. I also lay the groundwork and motivate the qualitative case studies which begin in the subsequent chapter. 

# A Brief Review of the Theory

In this section of the chapter, I briefly review the theory presented earlier in chapter two of this dissertation. I will not review in detail the three central principles of civil-military relations, as these have been covered in depth in the previous two chapters. Instead, the next several paragraphs will focus on a review of the theoretical expectation that the level of military prestige _may_ impact the willingness of _both_ civilian and military leaders to behave in ways that fail to adhere to the central principles of civil-military relations. 

The level of military prestige, I argued in chapter two, is one of two independent variables that shape the constraining influence of the three central principles of civil-military relations, which include the principle of civilian control of the military, the principle of non-partisanship of the military institution, and the principle of non-interference of the military into certain realms of state politics. I argued that high levels of military prestige present military and civilian actors with _opportunities_ to involve the military in politics, and thus, when military prestige is high, civilian and military actors are more likely to engage in behaviors that violate these central principles. 

The other independent variable that impacts the political behavior of civilian or military actors, I argued in chapter two, is the level of political polarization. I argued in chapter two that rising political polarization provides a "motive" for military actors in particular to intervene in politics, and thus, when polarization is high relative to when it is low, military actors are increasingly likely to engage in behaviors that violate the central principles of civil-military relations. 

## The Impact of Military Prestige on Military and Civilian Leaders

Chapter two argued that the level of military prestige is important because higher levels of military prestige increase the _potential_ that the military be used, either by civilian or military actors, to impact public opinion. This central claim arose mainly from the political communication literature about political elites, which argues that the public often looks to elites who are considered or believed by the public to be "knowledgeable" and "trustworthy" [@lupia_democratic_1998, 69-76] when forming decisions or opinions [@zaller_nature_1992]. 

Chapter two noted that although this and other literature on elite communication [@saunders_no_2017; @guisinger_mapping_2017; @baum_relationships_2008] is not specifically written from the perspective of military leaders in particular, applying this literature to military leaders and military elites in particular seems plausible, especially when the variable of military prestige is conceptualized as a relational indicator of the overall level of trust and confidence placed by the public in the military. Chapter two argued for a broad definition of the variable military prestige, and noted that several different pathways might ultimately impact the level of prestige of a military, including defeat or victory in war, and an overall assessment of how central or important the military is thought to be by the nation. The point of doing so was to allow room for the level of military prestige to change as a result of multiple possible factors. Regardless of how it changes, chapter two argued that when the public trusts and respects the military and its leaders at relatively high levels, it is reasonable to conclude that the public simultaneously perceives the military and its leaders as possessing the traits of knowledge and trustworthiness, and thus, as the prestige of the military rises, it is increasingly likely that the public _may_ also listen or become persuaded by the military and its leaders. 

Chapter two also emphasized that the _potential_ to impact public opinion is different from the actual _ability_ or _efficacy_ of the military to impact public opinion, as several experimental studies have shown [@golby_military_2012; @golby_listening_2013]. For the purposes of this dissertation, it is the belief of civilian and military actors in the _potential_ of the military to impact public opinion when the level of military prestige is relatively high that matters, and not whether a more prestigious military actually can and does persuade the public to a greater degree.   

The theory also explained in chapter two that military and civilian leaders both become aware of the military's increasing potential to move public opinion as the prestige of the military rises. In terms of how this impacts civilian and military actor behavior, an important question then becomes what type of outcome civilian and military actors may seek to influence. Civilian leaders, whose goal is to stay in political power or to achieve their desired policy choices, become increasingly willing to violate one or more of the principles of civil-military relations, and therefore to engage in a political behavior or activity that does so. Military leaders, too, may be more willing to behave in ways that violate one or more of the central principles of civil-military relations as the military institution becomes more prestigious. Chapter two also articulated a number of forms or types of behaviors that civilian and military actors engage in that, depending on the context in which these behaviors occur, violate one or more of the central principles of civil-military relations. These included, for example, the civilian actor behaviors of _electoral swaggering_, _partisan stacking_ and _the military as sales reps_, and the military actor behaviors of _explicit insubordination_, _running for office_, and what Risa Brooks has called "politicking" and "shoulder tapping" [@brooks_militaries_2009, 219]. The theory strongly advocated for evaluating each of these and many other types of political behaviors engaged in by civilian and military actors through the lens of their impact to the three central principles of civil-military relations. 

In summary, chapter two argued that rising military prestige presents an _opportunity_ for civilian and military actors to use the military to sway the public. Depending on what policy outcomes or issues civilian and military actors seek to influence - whether it is an election, or a use of force decision - this can result in civilian and military leaders engaging in political behaviors that challenge the principles of civilian control, non-partisanship, and non-interference of the military institution. When civilian and military leaders believe that the military is highly prestigious and therefore capable of swaying the public, both actor types, on the whole, are less likely to adhere to and thus, be constrained by, the central principles of civil-military relations. Combined with the presence (or absence) of a high level of military prestige, which affords military actors with a "motive" to engage in political behaviors that violate the principles of civil-military relations, chapter two argued that in total, four possible domestic structures emerge, each of which corresponds to distinct patterns of political activity that involve the military. 

# Methodology

In order to examine potential variation in the political behaviors undertaken by civilian actors that involve the military, we must examine a behavior that is primarily the purview of civilian actors. Such an approach parallels the approach implemented in the previous chapter, which explored retired military officer opinion commentary, a political behavior that, I argued, is undertaken primarily by military actors. This chapter examines the civilian use of the military in campaign advertisements, a form of behavior that most approximates the civilian actor behavior of _Electoral Swaggering_ that was described previously in chapter two. 

In this chapter, I examine all of the presidential campaign advertisements that aired on television in the five elections that occurred between 2000 and 2016, inclusively, which amount to roughly 3,600 campaign ads in total. During this period, the level of US military prestige, as measured by Gallup polling that tracks the percentage of the public that has a "great deal or quite a lot" of trust in the military, ranges from 64 (in 2000) to 82 (in 2003 and 2009) [@noauthor_confidence_2020].^[In the years that presidential elections occur, the lowest value is 64 (2000) and the highest is 75 (2004 and 2012), however.] 

To be clear, this does not constitute quite the degree of variation that a researcher hopes for, but there are other things that can and should be leveraged. First, from a qualitative perspective, this period captures (albeit barely) the significant change in the international environment marked by the attacks on the United States that occurred on September 11, 2001. These attacks were the proximate cause of two wars that the US military fought beginning that same year. Notwithstanding any debates about its effectiveness during these conflicts, the continual deployment of the US military to Iraq and Afghanistan likely increased the prestige of the military in a sense that may not be entirely captured by the Gallup polling data.

Additionally, when we also factor in the credibility ratings of other predominant political institutions, a greater shift in what we might call "relative" military prestige becomes slightly more evident. For example, the gap between the percentage of Americans surveyed who expressed a "great deal or quite a lot" in the military compared to Congress was 40% (in favor of the military) in 2000 and 64% (also in favor of the military) in 2016. At the very least, the period from 2000 to 2016 generally saw relative increases in the prestige of the military. These increases are small when we consider trust in the military in its raw form, but larger when viewed in relative terms compared to Congress. Both measurements may be useful when exploring the impact of military prestige on the civilian propensity to use the military in presidential campaign advertisements in ways that violate the central principles of civil-military relations.

Polarization over the same period also generally increased, except for measurements of affective polarization that rely on measurements provided by American National Election Survey, which show a dip in affective polarization between 2016 [@jeffrey_b_lewis_voteview_2020; @noauthor_american_nodate; @iyengar_affect_2012]. In short, the time period examined in this chapter, 2000 - 2016, broadly reflects movement between a domestic environment characterized by moderate polarization and moderate military prestige to one marked by higher polarization and higher military prestige. Because both of the independent variables of polarization and military prestige are changing over this time period, the statistical analysis will enable us to better disentangle the relative impacts of these (and other) variables. 

The level of trust Americans placed in Congress and the military as well as the level of congressional polarization between 2000 and 2016 is displayed in Figure \@ref(fig:trust-ch4). This figure also shows the relative prestige of the military. 

```{r trust-ch4-prep, include=FALSE, warning=FALSE, echo=FALSE}

df_trust_polar_ch4 <- read_csv(here("data","polar-inst-trust-time.csv")) %>% view()

df_trust_polar_ch4 <- df_trust_polar_ch4 %>% filter(Year<=2016 & Year >= 2000) %>% 
  mutate(
    Rel_Mil_Pres=Military-Congress,
    house_polar_dim1 = house_polar_dim1 *100,
    sen_polar_dim1 = sen_polar_dim1 *100,
    ) %>% pivot_longer(Congress:Rel_Mil_Pres, names_to = "value", values_to = "Score") %>% distinct() %>% view()
```

```{r trust-ch4, include=T, warning=FALSE, echo=F, message=F, fig.height=4, fig.width=6, fig.cap = "Public Trust (L) and Polarization (R), 1980-2021" }
df_trust_polar_ch4 %>% ggplot(aes(x=Year, y=Score, color=value)) + geom_line() + geom_point(size=0.7) + labs(x="Year", y="Trust in Institutions") + scale_color_discrete(name="Trust;\nPolarization",
                         breaks=c("Congress", "Military",  "Rel_Mil_Pres", "house_polar_dim1", "sen_polar_dim1"),
                         labels=c("Congress", "Military","Relative\nPrestige", "House\nPolarization", "Senate\nPolarization")) + scale_y_continuous(sec.axis = sec_axis(~., name = "Polarization (Scaled)")) + theme(legend.position="bottom") + guides(color = guide_legend(ncol=3,nrow=2,byrow=TRUE))
```

## Civilian Use of Military Imagery and Actors in Campaign Ads

Examining how civilians have featured the military in the content of campaign advertisements provides several advantages from a research perspective that are worth briefly articulating. The first is that, similar to the behavior of retired military officer opinion commentary, there is no _a priori_ expectation that featuring a military image or a military actor in a presidential campaign advertisement must violate, or automatically violates, one or more principles of civil-military relations. In fact, there is no expectation that civilian candidates choose to feature the military in campaign advertisements over and above any other institution, or over any other topic in general. Civilian political candidates and their campaign teams have an almost limitless array of options when producing campaign ads. 

A second significant advantage afforded by examining campaign advertisements is that they comprise an integral part of the democratic process. Like other democracies, the United States holds elections at fixed intervals, and thus, we can expect campaign advertisements to regularly occur in periods leading up to an election. We can and should also expect that the content of campaign ads will likely change over time, and for a variety of factors. The issues that are important to voters are not static, and the broad political atmosphere changes over time. Of course, we would expect civilian political candidates and their campaign teams to construct advertisements around those issues and topics they think will be most effective in persuading the public so as to secure electoral victory. At times, perhaps when wars are fought, for example, we might expect advertisements to feature the military in some noticeable way. While we cannot predict what specific issues will be on the docket four or eight years from now, we can suspect, if the theory offered earlier in this dissertation is accurate, that changes in the relative level of military prestige _could_ impact the degree to which civilian candidates feature or showcase the military in the content of campaign advertisements. 

How might civilian actors fail to adhere to the central principles of civil-military relations through the content of campaign advertisements? It is unlikely that the content of a campaign advertisement would violate the principle of civilian control. This is because, as has been stated earlier in this dissertation, civilians have a responsibility to maintain and enforce the principle of civilian control, but usually do not expressly violate the principle through acts they commit. Rather, civilian behavior "erodes" the principle of civilian control when civilians excessively defer to the military such that civilians no longer adequately influence the policy making process [@beliakova_erosion_2021]. Thus, a campaign advertisement that uses the military in such a way that violates the principle of civilian control would take on a rather extreme form, perhaps if an advertisement featured a retired military actor urging others to mount a coup or stage some other act of obvious insubordination. 

It is far more likely that featuring the military in television campaign advertisements would break the other two principles of civilian control - the principle of non-partisanship and non-interference of the military. A violation of the principle of non-partisanship of the military vis-a-vis campaign advertisements would occur if an advertisement clearly intimated that the military favored or endorsed a particular candidate, or the policies or political stances of a candidate, or a particular political party over another. Such a message, deliberate or not, could and likely would communicate to the American public that the military (or a faction of the military, or particular leaders within the military) favors a particular political entity over another. Examples of such behavior abound in almost every election, as this chapter will show. 

A violation of the principle of "non-interference" of the military into certain realms of state policy might occur if a campaign advertisement featured a military actor weighing in on policy arenas that are largely disconnected from national security, defense, and other essential roles the military is expected to fill in American society [@ben-meir_civil-military_1995]. Hypothetically speaking, this could occur if a retired general appeared in a campaign advertisement and spoke predominantly about a candidate's plan to reform social security or to reduce the national debt, and why particular candidate's plans to tackle these topics are more sound than another candidate's plans on the same topics. In this example, it would be reasonable to conclude that the advertisement violated the principle of non-interference because the retired military officer is likely not an expert on the topics of social security reform and the national debt. I do not question that a large number of reasonably intelligent retired military officers, like many retired individuals at large within American society, are smart and can understand the topics of social security reform and reigning in national debt. At issue, though, in this hypothetical example, is the choice by the civilian candidate or campaign team to use a military figure to convince the American public of the candidate's merits on these particular topics. It is doubtful that a retired general, admiral, colonel, or captain are truly the most qualified experts to weigh the candidates' stances over these issues that are somewhat removed from topics that distinctly involve national security, and thus, such an act violates the principle of non-interference. 

## Hypotheses Explored in this Chapter

The five hypotheses presented in chapter two are listed below. Given that this chapter focuses on a type of political undertaken primarily by civilian actors, this chapter primarily addresses H3. It will not address H1, which was addressed in the previous chapter, nor H2, as both of these hypotheses are addressed towards military rather than civilian actors. This chapter will, however, inform H4 and H5.  

H1: Military actors increasingly violate the central principles of civil-military relations when polarization is high relative to periods of low polarization. 

H2: Military actors increasingly violate the central principles of civil-military relations in public and visible ways when military prestige is high relative to periods of low military prestige. 

H3: Civilian actors increasingly violate the central principles of civil-military relations when military prestige is high relative to periods of low military prestige. 

H4: Military and civilian actors engage in relatively more visible and public forms of political behavior that involve the military institution when military prestige is high relative to periods of low military prestige. 

H5: The most extreme levels of overall political activity occur in environments that are characterized by simultaneously high levels of political polarization and military prestige. 

## The Data

The data consist of all (more than 3,600) presidential campaign television advertisements from the 2000 - 2016 presidential elections. ^[Unfortunately, the raw data used to code the 1996 election is no longer available to review; hence, this study begins with the 2000 Presidential Election. Additionally, raw data for the 2020 Election is not available at the time of this writing, but is projected to become available sometime in late 2022 or 2023.] The data is made available from the Wisconsin Advertising Project (WiscAD) and Wesleyan Media Project (WMP). For the 2000, 2004, and 2008 elections, the data come from the WiscAd, and consist of individual PDF documents that contain a full transcript and several still images of the advertisement [@goldstein_political_2002; @goldstein_presidential_2007; @goldstein_presidential_2011]. For 2012 and 2016 elections, the data comes from the WMP, and consists of full motion video, to include sound [@fowler_presidential_2017; @fowler_presidential_2020].^[See the Appendix for further details regarding how the data was cleaned and merged to enable a comparison over time.]

In the statistical analysis portion of this chapter, I also rely on data covering polarization, military prestige, and military casualties. For polarization data, I primarily rely on DW-Nominate (dynamic-weight, three step estimation) scores, which are constructed using roll-call voting data from the US House of Representatives and the US Senate [@lewis_voteview_nodate]. I also include "affective" polarization measurements, which are generally constructed using a series of "feeling thermometer" scores from the American National Election Study (ANES) [@iyengar_affect_2012; @iyengar_strengthening_2018; @iyengar_origins_2019].  

As a proxy for military prestige, I rely on Gallup public opinion data that captures the level of trust that Americans have placed in the military and a variety of other institutions over the past several decades [@noauthor_confidence_2020]. Finally, I use a variety of open source data sources from the Department of Defense regarding US military casualties to measure levels of hostile casualties sustained by US military forces over the past several decades.^[These sources include @department_of_defense_dcas_2011; @department_of_defense_dcas_2021; @department_of_defense_dcas_2021-1; @department_of_defense_dcas_2021-2; @department_of_defense_dcas_2021-3; @department_of_defense_dcas_2021-4; @department_of_defense_dcas_2021-5; @department_of_defense_dmdc_2021.]  

After gathering the data, I implement an empirical strategy that at its root, looks to answer two major questions. The first question, does this particular advertisement have any type of military image or symbol featured in the advertisement? At a minimum, answering this first question will help paint a picture regarding how frequently military imagery generally appears in campaign advertisements. The second question I seek to answer is whether a particular advertisement features the military in some way that violates the principles of civil-military relations. Answering these two main questions and presenting the results in a meaningful way is the aim of the next section of this chapter. 

# Analysis

```{r data, include=FALSE, echo=FALSE}

# reading in the raw data
df_campaign_ads_read <- read_csv(here("data", "refined_master_imagery_ch4_my_edits.csv"))
summary(df_campaign_ads_read)

# truncate the data after the last row (observation 3666) 
# rename the variable 'missing' to 'missing_ad', and replace all N/A values of 'missing_ad' with #0

df_campaign_ads_raw <- df_campaign_ads_read [1:3666,] %>% select(year:missing) %>% rename(missing_ad=missing) %>% mutate_at(
  .vars = vars(missing_ad), 
  .funs = ~ case_when(is.na(.) ~ 0, TRUE ~1) 
) %>% mutate(
  dummy_04=case_when(
    year == 2004 ~ 1, 
    year != 2004 ~0), 
  party_reg=case_when(
    party=="DEMOCRAT" ~ "DEMOCRAT",
    party=="REPUBLICAN" ~ "REPUBLICAN",
    (party!= "REPUBLICAN") & (party!= "DEMOCRAT") ~ "OTHER"
  ),
  party_reg_numeric=case_when(
    party_reg=="DEMOCRAT" ~ 1,
    party_reg=="REPUBLICAN" ~ 0,
    party_reg=="OTHER" ~2
  ),
  house_polar = case_when(
    year==2000 ~ 77.713,
    year == 2004 ~ 79.302,
    year == 2008 ~ 80.765,
    year == 2012 ~ 86.142,
    year == 2016 ~ 87.639
    ), 
  sen_polar = case_when(
    year==2000 ~ 65.833,
    year == 2004 ~ 64.984,
    year == 2008 ~ 68.736,
    year == 2012 ~ 74.383,
    year == 2016 ~ 81.293
  ),
  aff_polar = case_when(
    year==2000 ~ 61.692,
    year == 2004 ~ 69.27,
    year == 2008 ~ 69.519,
    year == 2012 ~ 84.689,
    year == 2016 ~ 67.754,
  ),
  HostCasRate = case_when(
    year==2000 ~ 1.23874924,
    year == 2004 ~ 52.0801226,
    year == 2008 ~ 25.1029256,
    year == 2012 ~ 29.4917871,
    year == 2016 ~ 1.229532
  ),
  HostCasRateLagged = case_when(
    year==2000 ~ 0,
    year == 2004 ~ 21.92,
    year == 2008 ~ 61.97,
    year == 2012 ~ 31.31,
    year == 2016 ~ 1.75
  ),
  mil_prestige = case_when(
    year== 2000 ~ 64,
    year == 2004 ~ 75,
    year == 2008 ~ 71,
    year == 2012 ~ 75,
    year == 2016 ~ 73
  ), 
  cong_prestige = case_when(
    year== 2000 ~ 24,
    year == 2004 ~ 30,
    year == 2008 ~ 12,
    year == 2012 ~ 13,
    year == 2016 ~ 9
  ),
  prty_ofc = case_when(
    year== 2000 ~ 1,
    year == 2004 ~ 0,
    year == 2008 ~ 0,
    year == 2012 ~ 1,
    year == 2016 ~ 1
  ),
  prestige_gap = mil_prestige-cong_prestige) %>% view()

# create another, identical data frame for conducting MICE later
df_campaign_ads_raw_mice <- df_campaign_ads_raw

```

In this portion of the chapter, I analyze the data and present the results. First, I present and analyze the general frequency and type of military symbols and imagery that appear within and across the data. This first step is meant to serve as a more general overview of the ways in which the military appears - in some fashion - in campaign advertisements over time.  Second, I analyze the frequency with which particular campaign advertisements involve the military in ways that violate one or more of the principles of civil-military relations. This second step is the more substantive of the two in terms of examining which factors impact the degree to which civilian actors violate the principles of civil-military relations.     

In total, the data consist of 3,666 advertisements across the five presidential elections that occurred from 2000 - 2016, inclusively. 89 of these observations (2.4%) are missing.^[After contacting WiscAd and WMP, there is no reason to suspect that the data is not missing at random. Later in the statistical analysis, I conduct multiple imputation to deal with the missing data, but want to first provide the "raw" or "as is" facts given the data that is available.] After cleaning and merging the data across the several data sets, I capture several variables that WiscAd and WMP have already coded for that may be helpful for the statistical analysis conducted later in this chapter. These variables, and a short description of each, are displayed in Table \@ref(tab:existing-variables-ch4). 

```{r existing-variables-ch4, include=TRUE, echo=FALSE}

text_vars <- tribble(~"Variable Name", ~"Question", ~"Possible Values",
  "Creative",  "What is the title of the ad?", "varies by each ad",
  "Count", "How many times did this ad air on a particular media market?", "varies by each ad", 
  "Length", "How long is the ad (in seconds)", "0-120",
  "Party", "What is the Party of the favored candidate?", "Republican; Democrat; Other",
  "Sponsor", "Who sponsored the advertisement?", "1 - Candidate; 2 - Party; 3 - Coordinated Between a Candidate and Party; 4 - Interest Group/Other",
  "f_mention", "Is the favored candidate mentioned in the ad?", "1 - Yes; 0 - No",
  "f_picture", "Is the favored candidate pictured in the ad?", "1 - Yes; 0 - No",
  "f_narrate", "Does the favored candidate narrate the ad?", "1 - Yes; 0- No",
  "o_mention", "Is/are the opposition candidate/s mentioned in the ad?", "1 - Yes; 0 - No",
  "o_picture", "Is/are the opposition candidate/s pictured in the ad?", "1 - Yes; 0 - No",
  "ad_tone", "In your judgement, is the primary purpose of the ad to promote a specific candidate, attack a candidate, contrast the candidate, or other?", "1 - contrast; 2- promote; 3 - attack; 4 - other",
  "prty_mn", "Does the ad mention the party label of the favored candidate or the opponent?", "0 - No, 1 - Yes, favored candidate's party; 2 - Yes, opposition candidate's party; 3 - Yes, both candidate's parties; 4 - yes, favored candidate's party but only in the paid for by line"

  )


text_vars %>% kable(booktabs=T, caption="Existing Variables from WiscAd and WMP") %>%
  kable_styling(full_width = F, font_size=10.5, latex_options="striped" ) %>% 
  column_spec(1, width="8em") %>% 
  column_spec(2, width="12em") %>% 
  column_spec(3, width="15em") 

```

I then develop four dichotomous variables that capture various types of military images and symbols.^[I came up with these four variables myself by first viewing approximately 250 advertisements across the five data sets to determine what categories would be most meaningful in terms of distinguishing the different types of images and symbols.] These four variables denote the use of active duty US soldiers or troops, veterans, combat, or US military hardware. 

The first kind of military imagery, that of US soldiers or troops, include images of US servicemen and women who are in uniform. It could include a lone soldier saluting, or thousands of US servicemen and women training, or a host of other possible scenarios. The key thing I look for are images of US military uniforms, military medals, or insignia that clearly denote US soldiers. The second type of military imagery relates to veterans, which I consider distinct from images of troops. This indicator variable looks for any image of a veteran, which could include gatherings of veterans, such as an American Legion or Veterans of Foreign Wars group, or imagery of a wounded veteran recovering in a hospital, or a veteran, for example, who is in a wheelchair during a later stage of life.

The third indicator accounts for images of combat. This could include things such as an explosion, a soldier driving a tank, or a scene containing fighting that clearly depicts US soldiers involved. This category does not include images of fighting that clearly involve non-US soldiers, such as soldiers of a foreign military or members of a terrorist group. The fourth indicator variable for a type of military imagery accounts for images of US military hardware. This variable captures advertisements that contain images or scenes of any major military hardware item, such as a naval vessel, fighter jet, or military radar system. 

I allow for the possibility that an advertisement portrays or includes more than one type of imagery. Again, the purpose of each of these four variables is to obtain a general idea for the prevalence of military imagery across the data. These four variables do not necessarily indicate that a campaign advertisement has violated any of the central principles of civil-military relations. 

Finally, I develop four additional variables that measure several other items of interest. These include the overall purpose of the advertisement, whether the candidate is a veteran, whether the advertisement explicitly mentions or portrays that the featured candidate is a veteran, and whether the advertisement itself is entirely focused on military, defense, veteran, and/or or national security issues. These variables, the question they intend to answer, and the potential range of values these variables can take, are displayed in Table \@ref(tab:el-ads-additional-ch4).

```{r el-ads-additional-ch4, include=TRUE, echo=FALSE}

text_vars <- tribble(~"Variable Name", ~"Question", ~"Possible Values",
  "Act_Duty_Imagery", "Does the advertisement feature imagery of US troops in uniform?", "1 - Yes; 0 - No",
  "Vet_Imagery", "Does the advertisement feature imagery of Veterans (VFW, veteran gatherings, etc.)?", "1 - Yes; 0 - No",
  "
  US_Mil_Hardware_Imagery", "Does the advertisement feature imagery of US military hardware (planes, tanks, helicopters, ships)?", "1 - Yes; 0 - No",
  "Combat_Imagery", "Does the advertisement feature images of past or present combat involving US military forces (prior wars, scenes of combat)?", "1 - Yes, 0 - No",
  "Def_Vet_Ad_Theme", "Is the ad itself focused entirely on military, defense, veteran, or national security issues?", "1 - Yes, military/defense/veteran issues only; 0 - other topics",
    "Ad_Purpose",  "What is the overall purpose of the ad", "1-promote or disparage candidates for an election; 2 - promote/disparage a specific issue or proposition that is on the ballot, 3 - to discuss a general theme (the economy, immigration, etc.) that does not do 1 or 2;  4 - other",
  "Can_Vet", "Is the candidate a Veteran?", "1 - Yes; 0 - No",
  "Can_Vet_Mention", "Does the ad explicitly state or portray the candidate as a Veteran?", "1 - Yes, 0 - No",
  "Vet_Fam", "Does the ad feature a Retired Military Elite, Veteran, or Veteran's family speaking?", "3 - Veteran and Family; 2 - Veteran's Family; 1 - Veteran; 0 - No",
  "Vet_Fam_Endorse", "Does the Military Figure who appears in a prominent way explicitly endorse the candidate for office?", "1 - Yes; 0 - No; 99-N/A",
  "Vet_Fam_Attack", "Does the Military Figure who appears in a prominent way explicitly attack or disparage other candidates running for office?", "1 - Yes; 0 - No; 99 - N/A",
  "Vet_Fam_Out_Issues", "Does the Military Figure who appears in a prominent way address topics other than the candidate's personal characteristics or defense/veteran (broadly related) issues?", "1 - Yes, the vet or vet's family addresses topics outside of the candidate's characteristics and/or veterans/defense (broadly related) issues; 0 - No, the vet or vet's family only addresses the candidate's characteristics and/or veterans/defense issues; 99 - N/A"
  

  )
 
text_vars %>% kable(booktabs=T, caption="Additional Variables / Anticipated Coding Scheme") %>%
  kable_styling(full_width = F, font_size=10.5, latex_options="striped" ) %>% 
  column_spec(1, width="8em") %>% 
  column_spec(2, width="14em") %>% 
  column_spec(3, width="16em") %>% add_footnote(label = "Vet_Fam_Endorse, Vet_Fam_Attack, and Vet_Fam_Out_Issues can take a value of N/A if the variable Vet_Fam equals 0.")
```

## Summary Statistics and Descriptive Data

Table \@ref(tab:desc-stats-ch4) shows, for each election cycle, the total number of advertisements aired on the nation's top media markets, the number of advertisements missing from the data, the proportion of the data that is missing, the number of advertisements that featured some form of military imagery in each election, and the proportion of each election's ads that contained military imagery. 

```{r initial-stats-ch4, include=F, echo=F, warning=F, message=F}
desc_stats <- tribble(
  ~'Election Cycle', ~'Total Ads (on hand and missing)', ~'# Ads Missing', ~'% Ads Missing', ~'# Ads on hand with Military Imagery', ~'% of Ads on hand with Military Imagery',
  "2000", "347", "39", "11.2", "20", "6.5",
  "2004", "650", "3", ".5", "136", "20.9",
  "2008", "966", "45", "4.7", "138", "15",
  "2012", "789", "1", ".1", "85", "10.8",
  "2016", "914", "1", ".1", "171", "18.7",
  "2000-2016", "3666", "89", "2.4", "550", "15.0")
```

```{r desc-stats-ch4, include=TRUE, echo=FALSE}
 
kable(desc_stats, booktabs=T, caption="Summary Statistics, The Military in Presidential Campaign Advertisements, 2000-2016") %>%
  kable_styling(full_width = F, font_size=10.5, latex_options="striped") %>% 
  column_spec(1:2, width="5em") %>% 
  column_spec(3:4, width="6.5em") %>% 
  column_spec(5:6, width="6.5em") 


```

### The Prominence of Military Imagery in US Presidential Elections

Table \@ref(tab:desc-stats-ch4) shows that 15 percent of the presidential campaign television ads featured military imagery of some kind. The least and most saturated elections in terms of military imagery, respectively, were the 2000 and 2004 elections, when 6.5% and 20.9% of advertisements featured some kind of military imagery.^[The 2000 Election contains the highest incidence of missing data, at 11.2%. Still, it is very unlikely that the 2000 Election is _not_ the least saturated with military imagery. For the 2000 Election to eclipse the second-least saturated election (2012), at least 18 of the 39 missing ads (46%) from the 2000 Election would have to contain military imagery. As the data is missing randomly, this is certainly possible, but very highly unlikely.] Following the 2004 Election, the percentage of campaign ads containing military imagery declined in the 2008 and 2012 Elections before rising sharply again in 2016.

There are a number of possible explanations behind these broad trends. One involves the arc of American foreign policy, and the military's role in shaping it between 2000 and 2016. The September 11, 2001 attacks ushered in a distinct role for the military as the US conducted lengthy military operations, first in Afghanistan in 2001 and later in Iraq in 2003. That military imagery contained high relative proportions of military imagery during the 2004 and 2008 Elections is mostly unsurprising. The war in Iraq was a major theme in the 2004 Election, when the incumbent, George W. Bush, ran against the Democratic challenger, John Kerry, a decorated combat veteran of the Vietnam War. At the time of the 2008 election, the surge of US military forces in Iraq had shown signs that it was likely working to at least alleviate the level of sectarian violence in Iraq. Moreover, US casualties, after peaking in 2007, had also been declining steadily throughout 2008 [@us_department_of_defense_dcas_2021]. This is likely a major reason why the proportion of campaign ads featuring military imagery declined in 2008 relative to the proportion of ads featuring military imagery in the 2004 election. 

The arc of US foreign policy, and the US military's role in it, becomes a bit more complicated from 2008 to 2016. First, from 2008 to 2012, the US' involvement in Iraq ended, but its involvement in Afghanistan increased. Indeed, the last of US forces withdrew from Iraq at the end of 2011 when Operation _New Dawn_ concluded.^[US forces would return to Iraq as part of the counter-ISIS effort several years later, and they remain there, albeit in relatively small numbers, as of writing this dissertation in early 2022.] But, President Barack Obama has also in 2009 decided to increase the number of US forces in Afghanistan as part of a surge of US forces in order to conduct a rigorous counter-insurgency campaign. Though the Afghan surge did not bring the acclaim that the earlier surge to Iraq did, there were nonetheless important victories along the way, perhaps none as important as the killing of Osama bin Laden, the architect of the 9/11 attacks, in 2011. Additionally, the number of US casualties in Afghanistan peaked in the summer of 2011, and declined afterwards, especially in the months leading up to the 2012 election [@us_department_of_defense_dcas_2021-1]. At the time of the election in November of 2012, few observers could point to Afghanistan and say that the surge of forces there had not worked, but at the same time, there was little if any appetite - from politicians and the public, alike - to do anything substantially greater there. Combined with the fact that 2012 was a reelection year for the incumbent, Barack Obama, the relatively low percentage of campaign advertisements featuring the military in 2012 is, in my view, rather unsurprising. 

However, in the 2016 election, the percentage of advertisements featuring military imagery rose sharply compared to the 2012 election. The passage of time facilitated an opportunity for Americans and political candidates to think about the trajectory of US foreign and national security policy in broad and critical terms, and in particular, to evaluate US involvement in a series of messy situations and crises that had unfolded abroad, particularly in the Middle East and Eastern Europe. These included conflict in Libya and an increasingly deteriorating civil war in Syria. In particular, President Obama's handling of events in Syria in 2013, which included making but failing to enforce a now infamous "red line" policy position that alluded to US military action in the event the Assad regime used chemical weapons (which Assad did), troubled many Americans [@rhodes_inside_2018].

In many respects, Hillary Clinton, the Democratic Candidate for President in 2016, was largely seen as an extension of Barack Obama, particularly in the realm of foreign policy. Clinton had served as Obama's Secretary of State, and she was portrayed extremely negatively after a US Ambassador and three other Americans were killed on September 12, 2012, when terrorists attacked the US consulate in Benghazi, Libya [@bixby_parents_2016]. Donald Trump, an outside candidate who ran and ultimately won as a Republican, instead ran on a platform of putting "America First," which included promises to cease fighting endless wars and significantly rebuilding the US military. The combination of perceived foreign policy missteps between 20012 and 2016 and the promise of fortifying the US' strategic position likely led to the relative increase in military imagery in the 2016 presidential election. 

### Specific Types of Military Imagery

Moving beyond the general incidence of military imagery across presidential elections from 2000 - 2016, Figure \@ref(fig:img-type-ch4) displays the number of uses of military imagery by type. The figure shows that imagery of active duty troops and veterans are the most common forms of military imagery. 

```{r data selection results, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE, error=F}
#create a variable called Yrcount that counts the number of ads per election cycle
#create a variable called "mil_img" if there is any type of military imagery category in a 
# given ad
### analysis with original code on the data that we have on hand

df_campaign_ads_raw_on_hand <- df_campaign_ads_raw %>%
  filter(missing_ad != 1) %>% 
  group_by(year) %>% 
  mutate(
    Yrcount=n(),
    mil_img = case_when(
    (vet_fam != 0) | (act_dty_img ==1 | vet_img ==1 | us_mil_hard_img ==1 | us_cmbt_img==1) ~ 1, 
  (vet_fam == 0) & (act_dty_img ==0 & vet_img ==0 & us_mil_hard_img ==0 & us_cmbt_img==0) ~ 0
),
  no_mil_img = case_when(
    (vet_fam != 1) & (act_dty_img !=1 & vet_img !=1 & us_mil_hard_img !=1 & us_cmbt_img!=1) ~ 1, 
  (vet_fam != 0) | (act_dty_img ==1 | vet_img ==1 | us_mil_hard_img ==1 | us_cmbt_img==1) ~ 0
  )
    ) 


# create a new df called 'adj_df_ch4' that 
# takes the original df and adds additional variables 
#(1) sum_mil_img, the total # of ads per election cycle containing military imagery 
#(mil_img ==1); 
#(2) sum_no_mil_img, the total # of ads per cycle containing no military imagery
#(3) mil_fig, if an ad features a prominent veteran or veteran's family in the ad
# (mil_fig==1)
#(4) sum_mil_fig_cycle, the total # of ads per cycle that feature a veteran or family endorser
# sum_mil_fig_cycle
#(5) sum_def_vet_ad_theme, the total # of ads per cycle that address entirely matters
# of national security, defense, or veterans issues
adj_df_ch4 <- df_campaign_ads_raw_on_hand %>% group_by(year) %>% mutate(
    sum_mil_img_cycle=sum(mil_img),
    sum_no_mil_img_cycle=sum(no_mil_img),
    prop_mil_img = sum_mil_img_cycle/Yrcount,
    act_duty_img_total=sum(act_dty_img),
    prop_act_dty_img=(act_duty_img_total/Yrcount),
    vet_img_total=sum(vet_img),
    prop_vet_img=(vet_img_total/Yrcount),
    us_cmbt_img_total=sum(us_cmbt_img),
    prop_us_cmbt_img=(us_cmbt_img_total/Yrcount),
    us_mil_hard_img_total=sum(us_mil_hard_img),
    prop_us_mil_hard_img=(us_mil_hard_img_total/Yrcount),
    mil_fig=case_when(vet_fam != 0 ~1,
                      vet_fam == 0 ~ 0),
    sum_mil_fig_cycle=sum(mil_fig),
    sum_def_vet_ad_theme=sum(def_vet_ad_theme),
    def_theme_and_mil_img=case_when((def_vet_ad_theme==1) & (mil_img==1) ~1,
                                    (def_vet_ad_theme!=1) | (mil_img==0) ~0),
    sum_def_theme_and_mil_img=sum(def_theme_and_mil_img),
    not_def_theme_but_mil_img=case_when((def_vet_ad_theme!=1) & (mil_img==1) ~ 1,
                                         (mil_img==0) | (def_theme_and_mil_img==1) ~0),
    sum_not_def_theme_but_mil_img=sum(not_def_theme_but_mil_img),
    non_def_vet_ad_theme=case_when((def_vet_ad_theme==0) ~1,
                                  (def_vet_ad_theme==1) ~ 0),
    sum_non_def_vet_ad_theme=sum(non_def_vet_ad_theme)
    ) %>% view()

```

```{r img-type-ch4, include=T, echo=FALSE, error=FALSE, warning=FALSE, message=F, fig.height=4, fig.width=6, fig.cap="Number of Presidental Campaign TV Ads Displaying Military Imagery"}

## prepare a graph for raw counts of image type 

# raw counts
graphable_img_types_df <- adj_df_ch4 %>% 
  select(year, act_duty_img_total, vet_img_total, us_cmbt_img_total, us_mil_hard_img_total) %>% pivot_longer(act_duty_img_total:us_mil_hard_img_total, names_to = "statistic", values_to = "Number") %>% distinct () %>% view ()

plot_img_type_count <- ggplot(data=graphable_img_types_df, aes(fill=statistic, x=year, y=Number)) + geom_bar(position="dodge", stat="identity") + labs(x="Election", y="Number of Ads") + scale_fill_discrete(name="Military Imagery Type",
                         breaks=c("act_duty_img_total", "vet_img_total", "us_cmbt_img_total", "us_mil_hard_img_total"),
                         labels=c("Active Military", "Veteran", "Combat", "Mil. Hardware")) + theme(legend.position="bottom") + scale_x_continuous(breaks=seq(2000, 2016, 4)) + guides(fill = guide_legend(ncol=2 ,nrow=2,byrow=TRUE))

plot_img_type_count

```

In 2004 and in 2008, combat imagery is more common than imagery containing military hardware, but in all other elections, more images of military hardware were evident than combat imagery. The most reasonable explanation for this difference is the fact that American combat operations in Iraq and Afghanistan loomed large over the 2004 and 2008 elections, respectively. It is also quite clear from Figure \@ref(fig:img-type-ch4) that the 2000 election contained far less military imagery - across all imagery types - than subsequent elections. This likely reflects the post-9/11 era increase in the general prominence or the centrality of the military as an institution in politics relative to the immediate elections held in the pre-9/11 era. 

Finally, it is also clear that the relative gap between active duty military imagery and veteran imagery has widened, as veteran imagery has emerged as the most common form of imagery beginning in the 2012 election. These trends are also reflected when we plot the type of military imagery featured in election advertisements as proportions, as reflected in Figure \@ref(fig:img-type-prop-ch4). Note that the overall proportions of advertisements in the 2008 and 2012 elections that featured active duty military and veteran imagery were relatively low compared to the 2016 election. This is somewhat interesting in that during the 2012 election, and certainly during the 2016 election, wars in Iraq and Afghanistan were far less active on the whole than they had been in 2008. In other words, we might anticipate a decline in the proportion of advertisements featuring active military troops in 2016 relative to 2012, but instead, the data shows that the trend is increasing. 

```{r img-type-prop-ch4, include=T, echo=FALSE, warning=FALSE, message=FALSE, fig.height=4, fig.width=6, fig.cap="Proportion of Presidental Campaign TV Ads Displaying Military Imagery"}

## prepare a graph for proportion of image type 

graphable_img_prop_df <- adj_df_ch4 %>% 
  select(year, prop_act_dty_img, prop_vet_img, prop_us_cmbt_img, prop_us_mil_hard_img) %>% pivot_longer(prop_act_dty_img:prop_us_mil_hard_img, names_to = "statistic", values_to = "Proportion") %>% distinct () %>% view ()

# proportion graph military imagery type
plot_img_type_prop <- ggplot(data=graphable_img_prop_df, aes(color=statistic, x=year, y=Proportion)) + geom_point() + geom_smooth(se=FALSE)+ labs(x="Election", y="Proportion of Total Ads") + scale_color_discrete(name="Military Imagery Type",
                         breaks=c("prop_act_dty_img", "prop_vet_img", "prop_us_cmbt_img", "prop_us_mil_hard_img"),
                         labels=c("Active Military", "Veteran", "Combat", "Mil.Hardware")) + theme(legend.position="bottom") + scale_x_continuous(breaks=seq(2000, 2016, 4)) + scale_y_continuous(breaks=seq(0, .125, .025)) + guides(color = guide_legend(ncol=2,nrow=2,byrow=TRUE))

plot_img_type_prop

```

### The Use of Military Imagery in non Defense-Themed Ads

We can and should suspect that some form of military imagery would appear in advertisements that are squarely focused on defense, veterans, or other strict national security issues. And it is certainly reasonable to suspect that the prominence of these types of issues might vary over time, as described above, as particular events such as wars or other crises occur. 

Yet there is another way to gauge how civilian candidates choose to leverage the military for their electoral advantages. This involves looking for advertisements that are _not_ centered on defense-related issues, but which feature military imagery anyways. The focus or theme of these advertisements is an issue or set of issues other than defense or national security topics, but by still including some type of military imagery, the candidate is, I argue, still signaling to the public that the military matters to a certain degree.

Figure \@ref(fig:non-def-theme-imagery) shows the proportion of advertisements in each election cycle that do not primarily address defense, veterans, or national security topics but include military imagery anyways. The smoothing line in blue clearly depicts an upward trend over time, suggesting that civilian candidates and their campaign teams have increasingly valued portraying the military in some way to the public, even in ads that do not directly address defense, veterans, or other national security topics. This trend seems to suggest that over time, civilian political candidates increasingly want the viewer to know that the military, in some way, has not been forgotten. 

```{r non-def-theme-imagery, include=T, echo=FALSE, warning=FALSE, message=F, fig.height=4, fig.width=6, fig.cap="Military Imagery in non-Defense or non-Veteran Related Presidential Campaign Advertisements, 2000-2016"}

facet_wrap_nondef <- adj_df_ch4 %>% select(year, Yrcount, sum_def_vet_ad_theme, sum_def_theme_and_mil_img, sum_not_def_theme_but_mil_img) %>% mutate(
  prop_not_def_but_mil_img = sum_not_def_theme_but_mil_img/Yrcount
  ) %>% select(year, prop_not_def_but_mil_img) %>% 
  distinct()

# now make the data into longer format for graphing and prepare to graph

facet_wrap_nondef <- facet_wrap_nondef  %>% pivot_longer(prop_not_def_but_mil_img, names_to = "question", values_to = "value")

plot_nondef_img <- ggplot(data=facet_wrap_nondef, aes(x=year, y=value)) + geom_bar(position="dodge", stat="identity", fill="seagreen") + labs(x="Election", y="Proportion of Total Ads", caption="Presidential Campaign Television Ads that include Military Imagery\nBut do not focus on Defense, Veterans, or National Security Issues") + theme(legend.position="bottom") + scale_x_continuous(breaks=seq(2000, 2016, by = 4)) + geom_smooth()

plot_nondef_img

```

### Prominent Appearances by Military Figures in Campaign Ads

A final orientation measure to the data involves the number and proportion of advertisements that feature a military actor of some sort appearing prominently in an advertisement. Later in this chapter, we will explore what actions, behaviors, and/or statements these military actors engage in or make, but for now, Figure \@ref(fig:mil-endorse) shows the number of times in each election cycle that military figures appear in prominent ways during presidential campaign advertisements. 

```{r mil-endorse, include=T, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, fig.height=4, fig.width=6, fig.cap="Number of Presidential Campaign TV Ads Featuring Military Figures Appearing Prominently, 2000-2016"}

# create a df with the variables I will need.

prom_figs_df <- adj_df_ch4 %>% select (year, count, creative, party, sponsor, ad_purpose, can_vet, can_vet_mention, vet_fam, vet_fam_end, vet_fam_atk, vet_fam_out_issues, prom_mil_fig, Yrcount, dummy_04:prestige_gap) %>% group_by(year, party) %>% view()

# create counts and proportions of advertisements
# denoting pertinent variables such as 
# counts of prominent veteran appearances, prominent vet family appearances, etc.

prom_figs_df <- prom_figs_df %>% group_by(year) %>% mutate(
    veteran_prominent=case_when(prom_mil_fig=="vet" ~ 1, prom_mil_fig!="vet" ~0),
    total_veteran_prominent=sum(veteran_prominent),
    elite_prominent=case_when(prom_mil_fig=="elite" ~1, prom_mil_fig !="elite" ~0),
    total_elite_prominent=sum(elite_prominent),
    family_vet_prominent=case_when(prom_mil_fig=="family" ~1, prom_mil_fig !="family" ~0),
    total_family_vet_prominent=sum(family_vet_prominent),
    elite_prominent_gen_sense=case_when(prom_mil_fig=="elitegeneralsense"~1, prom_mil_fig!="elitegeneralsense" ~0),
    total_elite_mil_fig_gen_sense=sum(elite_prominent_gen_sense),
    elite_pol_mil_fig=case_when(prom_mil_fig=="elitepolfig" ~1, prom_mil_fig!="elitepolfig"~0),
    total_elite_pol_mil_fig=sum(elite_pol_mil_fig),
     total_elite_appearances=total_elite_prominent+total_elite_mil_fig_gen_sense+total_elite_pol_mil_fig,
    total_prominent_appearances=total_veteran_prominent + total_elite_prominent + total_family_vet_prominent + total_elite_mil_fig_gen_sense + total_elite_pol_mil_fig,
    prop_veteran_prominent=(total_veteran_prominent/Yrcount),
     prop_elite_prominent=(total_elite_prominent/Yrcount),
    prop_family_vet_prominent=(total_family_vet_prominent/Yrcount),
    prop_elite_prominent_gen_sense=(total_elite_mil_fig_gen_sense/Yrcount),
     prop_elite_pol_mil_fig=(total_elite_pol_mil_fig/Yrcount),
    prop_elite_appearances=(total_elite_appearances/Yrcount),
    prop_prominent_appearances=(total_prominent_appearances/Yrcount)
    ) %>% view()

# begin making the first facet / panel graph wrap 

facet_wrap1 <- prom_figs_df %>% select(year, veteran_prominent:total_prominent_appearances) %>% distinct() %>% select(-veteran_prominent, -elite_prominent, -family_vet_prominent, -elite_prominent_gen_sense, -elite_pol_mil_fig, -total_elite_appearances) %>% distinct() %>% view()

# now make the data into longer format for graphing and prepare to graph

facet_wrap1 <- facet_wrap1  %>% pivot_longer(total_veteran_prominent:total_prominent_appearances, names_to = "question", values_to = "value") %>% view ()

plot3 <- ggplot(data=facet_wrap1, aes(fill=question, x=year, y=value)) + geom_bar(position="dodge", stat="identity") + labs(x="Election", y="Number of Ads") + scale_fill_discrete(name="Actor",
                         breaks=c("total_veteran_prominent", "total_elite_prominent", "total_family_vet_prominent", "total_elite_mil_fig_gen_sense", "total_elite_pol_mil_fig", "total_prominent_appearances"),
                         labels=c("Vets", "Retired Gen./Adm.","Family Members\nof Vets", "Retired Gen./Adm.\nMentioned Pictured Only", "Pol.Figures w/Mil.\nBackgrounds", "Any Military\nFigure")) + theme(legend.position="bottom") + scale_x_continuous(breaks=seq(2000, 2016, 4)) 

plot3

```

By appearing in a prominent way, I am referring to instances that describe a number of potential circumstances, but the thrust of the idea involves someone who is related to the military in some way - a military figure - that appears in a distinguishable, identifiable, and deliberate manner. This might include a retired military officer appearing by name to endorse a particular candidate for office, or it might include the presentation of a photograph and name of a military actor on the screen, informing the viewer that this particular military actor has endorsed a candidate running for office. It may also include a veteran narrating an advertisement and identifying himself or herself as such. Other scenarios might include the family member(s) of a soldier killed or wounded in action being identified as such and appearing in an advertisement.^[In a few instances where the the viewer is informed that a certain candidate was "endorsed" by several former members of the Joint Chiefs of Staff, whose photo is displayed during the ad, I have labeled this category as "retired generals/admirals pictured only"]. 

These categorical values for military figure type are important because each type is likely used to portray a slightly different type of message. Elite military actors, such as retired general and admirals, convey the rank and authority of high-ranking military officers. Civilian candidates might choose elite military actors to deliver a message in the hopes that the public will listen to them on the basis of their accrued knowledge and expertise. Non-elite veterans, on the other hand, may not be subject matter experts on specific and often complex foreign policy issues, but they might nonetheless be chosen by civilian political candidates to convey a message regarding the broader value of service to one's country. A non-elite veteran is likely in a better position, in other words, to hypothetically state in an advertisement, "I fought for my country; now I hope the country does its part by making prescription drugs more affordable" than a retired multistar general or admiral. Veteran family members, understandably, offer an even more personal and intimate connection to the public. Civilian political candidates may use veterans and/or their family members to craft emotional, heartfelt messages with the public.

Figure \@ref(fig:mil-endorse-prop) expresses these results as a proportion of the total number of advertisements displayed in a given election cycle. 

```{r mil-endorse-prop, include=T, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, fig.height=4, fig.width=6, fig.cap="Proportion of Presidential Campaign TV Ads Featuring Military Figures Appearing Prominently, 2000-2016"}

# begin making the first facet / panel graph wrap 

facet_wrap1_prop <- prom_figs_df %>% select(year, prop_veteran_prominent:prop_prominent_appearances) %>% distinct() %>% select(-prop_elite_appearances) %>% distinct() %>% view()

# now make the data into longer format for graphing and prepare to graph

facet_wrap1_prop <- facet_wrap1_prop  %>% pivot_longer(prop_veteran_prominent:prop_prominent_appearances, names_to = "question", values_to = "value") %>% view ()

plot3_prop <- ggplot(data=facet_wrap1_prop, aes(fill=question, x=year, y=value)) + geom_bar(position="dodge", stat="identity") + labs(x="Election", y="Proportion of All Ads") + scale_fill_discrete(name="Actor",
                         breaks=c("prop_veteran_prominent", "prop_elite_prominent", "prop_family_vet_prominent", "prop_elite_prominent_gen_sense", "prop_elite_pol_mil_fig", "prop_prominent_appearances"),
                         labels=c("Vets", "Retired Gen./Adm.","Family Members\nof Vets", "Retired Gen./Adm.\nPictured Only", "Pol.Figures w/Mil.\nBackgrounds", "Any Military\nFigure")) + theme(legend.position="bottom") + scale_x_continuous(breaks=seq(2000, 2016, 4)) + guides(fill = guide_legend(ncol=3,nrow=2,byrow=TRUE))

plot3_prop
```

Figure \@ref(fig:mil-endorse) and Figure \@ref(fig:mil-endorse-prop) portray an interesting narrative. There were no instances of military figures appearing prominently in campaign advertisements in the 2000 Election. Then, in the 2004 election, there was an explosion of military figure prominent appearances. In 2008 and 2012, military figure prominent appearances gradually declined. Yet in 2016, the number and proportion of military figure prominent appearances increased - not quite to the levels experienced during the 2004 election, but to nonetheless relatively high levels. 

The 2004 presidential election deserves a quick note to orient the reader to some of the underlying reasons for the high number (42 in total) and proportion (nearly 6.5% of all ads) of prominent military figure appearances. In 2004, a number of veterans, family members of veterans, and retired officers coalesced to attack the war record of the Democratic Candidate, Senator John Kerry. The group, which was an official political action committee known as the "Swiftboat Veterans for Truth," claimed that Kerry had exaggerated his war record and was undeserving of many of the heroic medals he had received [@ullman_whats_2008]. The group even published a book that turned out to be a _New York Times_ bestseller, whose title embodied the group's main charge against Kerry: "Unfit for Command" [@buckaloo_swiftboat_nodate]. 

Although it is hard to state conclusively, the attacks levied by the Swiftboat Veterans for Truth against Kerry were almost certainly effective, and likely damaged his chances of winning the Presidency. One scholar argues that the fact that these ads featured real veterans as opposed to paid actors, the timing of their release (the ads were released in August, the "doldrums period between conventions," which facilitated these ads having quite a bit of exposure in the media), and at least a plausible semblance of veracity badly hurt Kerry's chances as a candidate [@devlin_contrasts_2005, 292].^[It is not my central purpose to render a normative judgement on the actions, nor the intent of, the Swiftboat Veterans. One can reasonably conclude, in my view, that their behavior was inherently partisan, for the actions of the group contributed, at least in part, to the prevention of Kerry's ascendance to the Presidency. From this perspective, the actions of the Swiftboat Veterans must be viewed as wrong in that they violated the military's non-partisan ethic. On the other hand, one may take a different view on the topic altogether. The historical record indicates that the Swiftboat Veterans began as an organic veterans group. It was not the brainchild of a political party. The group's founder, retired Admiral Roy Hoffman, insists that his actions were sparked upon reading Douglas Brinkley's biography of Kerry, _Tour of Duty_, and gaining the impression that the book had significantly embellished Kerry's war record. Ultimately, many of Kerry's fellow veterans supported and backed Kerry, while others did not. See @rosin_unfriendly_2004 for a narrative of the beginning of the Swiftboat Veterans organization and @devlin_contrasts_2005 for a description of the group's activities.] 

## Are Campaign Ads Increasingly Challenging Principles of Civil-Military Relations?

I now pivot to answering a second, and for the purposes of the theory argued earlier in this dissertation, more substantive question, which is whether and to what degree the military images, symbols, and military figures who appear prominently in presidential campaign advertisements violate one or more of the principles of civil-military relations. In order to explore and answer this important question, I operationalize potential violations to the principles of civil-military relations. Further, I focus on the actions or behaviors of the military figures who appear prominently in advertisements rather than on the images and symbols that appear in ads, mainly because these actions and behaviors are clear and tangible expressions rather than subjective reactions to how the military is portrayed.^[In other words, a researcher can much more easily determine if a prominent military figure endorses or attacks a political candidate, rather than conclusively determine if the presence of an image of a tank or aircraft carrier violates the principle of non-partisanship. The latter is almost entirely subjective, whereas the former is clearer.] 

I then create three dichotomous indicator variables to capture potential violations to two out of the three central principles of civil-military relations military figures who appear in a prominent way undertake. First, I capture violations of the military's non-partisanship principle by cataloging instances of a military figure explicitly endorsing or attacking a political candidate for office. This type of violation is fairly straightforward. I watch or read the transcript of the ads (in the cases of 2000, 2004, and 2008) and catalog explicit endorsements or attacks by military figures. 

I also capture violations of the principle of non-interference of the military by capturing instances in which a military figure addresses issues that fall outside of the military figure's relevant expertise. This type of violation is not as straightforward, as it requires making a determination of whether the military figure either possesses knowledge or expertise regarding a certain topic, or might reasonably serve as a spokesperson for a particular topic or issue area. For example, if a veteran who appears in an advertisement discusses why a particular candidate's plan for improving social security is better than another candidate's, I count this as a violation of the principle of non-interference.^[Similar to identifying violations to this principle in the previous chapter, I try to be both reasonable and charitable. I code for instances in which there is a gross mismatch between the military figure and whatever topic he or she is addressing.] 

### Violations of the Principles of Non-Partisanship and Non-Interference

Figure \@ref(fig:viol-overview-prop) displays the proportion of advertisements in each election cycle that endorse political candidates, attack political candidates, or address issues that are tangential or altogether outside of military actor expertise. In addition, Figure \@ref(fig:viol-overview-prop) plots both the proportion of advertisements in each election cycle in which one of these violations to the principles of civil-military relations occur, as well as an overall rate of violations to the central principles of civil-military relations.  

```{r viol-overview, include=FALSE, echo=FALSE, error=FALSE, warning=FALSE}

prom_figs_df %>% view()

prom_figs_df %>% 
  select_if(function(x) any(is.na(x))) %>% 
  summarise_each(funs(sum(is.na(.)))) %>% view()

## replace all values for endorse, attack, and issues
## not equal to '1' with '0' to enable summation

prom_figs_df<- prom_figs_df %>% mutate(
  appearance= case_when(
    vet_fam == 1 | vet_fam==2 ~ 1,
    vet_fam != 1 & vet_fam!= 2 ~ 0
            ),
  vet_fam_end = replace(vet_fam_end, vet_fam_end != 1, 0), 
   vet_fam_atk = replace(vet_fam_atk, vet_fam_atk != 1, 0),
  vet_fam_out_issues = replace(vet_fam_out_issues, vet_fam_out_issues != 1, 0)
) %>% view()

# create variables for instances of, the total #, and proportion of 
# all ads on hand by actor type that 
# endorse, attack, or address issues outside of relevant expertise
# or conduct any of these violations. 

# Note that here, I combine as "elites" all retired elites, or political 
# elites with military backgrounds, as well as elites who are pictured in a 
# general sense only. 

violations_wrap <- prom_figs_df %>% group_by(year) %>% mutate(
  endorse_count=sum(vet_fam_end),
  attack_count=sum(vet_fam_atk),        ### instances of violations by type
  non_int_count=sum(vet_fam_out_issues),
  num_violations=sum(vet_fam_end+vet_fam_atk+vet_fam_out_issues),
  endorse_prop=endorse_count/Yrcount,   ### proportion of ads that have a type of 
  attack_prop=attack_count/Yrcount,
  non_int_prop=non_int_count/Yrcount,
  violations_rate=num_violations/Yrcount,       ### rate of violations per election cycle
  ad_violating=case_when(
    vet_fam_end==1 | vet_fam_atk == 1 | vet_fam_out_issues ==1 ~ 1,
    vet_fam_end==0 & vet_fam_atk == 0 & vet_fam_out_issues == 0 ~ 0),
  prop_ads_violating=(sum(ad_violating)/Yrcount)) %>% view()

# now create graphs of proportion of ads violating
prop_viol_ch4 <- violations_wrap %>% 
  select(year, endorse_prop:violations_rate, prop_ads_violating) %>% pivot_longer(endorse_prop:prop_ads_violating, names_to = "statistic", values_to="Proportion") %>% distinct() %>% view()

prop_viol_ch4_graph <- prop_viol_ch4 %>% ggplot(aes(x=year, y=Proportion, color=statistic)) + geom_point() + geom_smooth(se=FALSE) + labs(x="Election", y="Proportion/Rate") + scale_color_discrete(name="Proportion", breaks=c("endorse_prop", "attack_prop", "non_int_prop", "violations_rate", "prop_ads_violating"), labels=c("Endorse Candidate", "Attack Candidate", "Address Topics\nOutside of\nExpertise", "Violation\nRate", "Portion of\nAds Violating")) + theme(legend.position="bottom") + guides(color = guide_legend(ncol=3,nrow=2,byrow=TRUE))

```

```{r viol-overview-prop, include=T, warning=FALSE, error=FALSE, message=FALSE, echo=FALSE, fig.height=4, fig.width=6, fig.cap="Proportion and Rate of All Campaign Advertisements That Violate Central Principles of Civil-Military Relations, 2000-2016" }

prop_viol_ch4_graph

```

Figure \@ref(fig:viol-overview-prop) confirms that, as suspected, the 2004 Presidential Election featured the highest proportion of advertisements that violated one or more of the principles of civil-military relations, followed by 2016, 2008, 2012, and 2000. This is also true of the proportion of advertisements that featured military figures endorsing or attacking political candidates. However, the  2016 Election featured the most instances of military figures addressing topics that are outside or indirectly related to military expertise. However, this involved a small number of instances (n=3 in 2016, and n=1 in 2012 and 2004).

Figure \@ref(fig:viol-overview-prop) also shows, rather interestingly, that when the number of total violations that occur in a given election cycle are expressed as a rate of the number of advertisements that air, the 2016 Election is behind the 2004 Election, but not by much. This measure perhaps better captures the possibility that military actors violate more than one principle of civil-military relations while appearing in a single ad, such as a military figure endorsing and attacking a candidate running for office in the same campaign advertisement. 

Indeed, for every 100 ads run in the 2016 Election, military figures engaged in 4.7 violations to the principles of civil-military relations by attacking or endorsing candidates, or addressing topics that are tangential to military experience. In 2004, this rate was 5.1. The rate of violations is also higher in 2012 than in 2008, whereas the proportion of advertisements containing at least one violation was higher in 2008 than in 2012. This suggests that instances in which military actors who appear in campaign advertisements and both endorse and attack a political candidate running for office is increasing, a trend that also appears to be reflected in that the proportion of advertisements in which military figures attacked candidates was relatively high in 2016. This likely reflects the overall level of vitriol that characterized the election between Donald Trump and Hillary Clinton. 

### Military-Affiliated Interest Groups?

To account for the potential role of interest groups with significant military affiliation in campaign advertisements, I filter the data and drop advertisements that are sponsored by interest groups.^[Unfortunately, the data does not provide listings of which interest groups sponsor particular ads. I am able to hear (or see in the transcripts for older advertisements) most cases of obvious veteran-affiliated interest groups, but I nonetheless ultimately have to rely on the coding for this variable conducted previously by WiscAd and WMP. If anything, filtering the data in this way likely results in undercounting the advertisements that subsequently remain for analysis, but I do not assess the potential impacts from doing so to be severe.] 

Figure \@ref(fig:interest-group-drop-prop-viol) again displays the proportion of advertisements in each election cycle that endorse political candidates, attack political candidates, or address issues that are tangential or altogether outside of military actor expertise, as well as both the proportion of advertisements that violate, and an overall rate of violations to, the central principles of civil-military relations. 

```{r interest-group-drop-prop-viol, include=T, echo=FALSE, warning=FALSE, message=FALSE, error=FALSE, fig.height=4, fig.width=6, fig.cap="Proportion and Rate of Campaign Advertisements Sponsored by Candidates and/or Political Parties That Violate Central Principles of Civil-Military Relations, 2000-2016"}

# drop observations where sponsor==4 using the filter command 

drop_igs_violations_wrap <- prom_figs_df %>% group_by(year) %>% filter(sponsor!= 4) %>% mutate(
  endorse_count=sum(vet_fam_end),
  attack_count=sum(vet_fam_atk),                                ### instances of violations by type
  non_int_count=sum(vet_fam_out_issues),
  num_violations=sum(vet_fam_end+vet_fam_atk+vet_fam_out_issues),
  endorse_prop=endorse_count/Yrcount,                         ### proportion of ads that have a type of violation
  attack_prop=attack_count/Yrcount,
  non_int_prop=non_int_count/Yrcount,
  violations_rate=num_violations/Yrcount,                      ### rate of violations per election cycle
  ad_violating=case_when(
    vet_fam_end==1 | vet_fam_atk == 1 | vet_fam_out_issues ==1 ~ 1,
    vet_fam_end==0 & vet_fam_atk == 0 & vet_fam_out_issues == 0 ~ 0),
  prop_ads_violating=(sum(ad_violating)/Yrcount)) %>% view()

# now create graphs of proportion of ads violating
drop_igs_prop_viol_ch4 <- drop_igs_violations_wrap %>% 
  select(year, endorse_prop:violations_rate, prop_ads_violating) %>% pivot_longer(endorse_prop:prop_ads_violating, names_to = "statistic", values_to="Proportion") %>% distinct() %>% view()

drop_igs_prop_viol_ch4_graph <- drop_igs_prop_viol_ch4 %>% ggplot(aes(x=year, y=Proportion, color=statistic)) + geom_point() + geom_smooth(se=FALSE) + labs(x="Election", y="Proportion/Rate") + scale_color_discrete(name="Proportion", breaks=c("endorse_prop", "attack_prop", "non_int_prop", "violations_rate", "prop_ads_violating"), labels=c("Endorse", "Attack", "Address Outside\nTopics", "Violation Rate", "Portion Ads\nViolating")) + theme(legend.position="bottom") + guides(color = guide_legend(ncol=3,nrow=2,byrow=TRUE)) 

drop_igs_prop_viol_ch4_graph
```

Figure \@ref(fig:interest-group-drop-prop-viol) reveals that even after removing military-affiliated interest groups from consideration, civilian political candidates and political parties show varying degrees of interest in harnessing military figures to endorse candidates, attack candidates, and on occasion, address topics that are indirectly related to areas of military expertise. Somewhat unsurprisingly, there is no question that a sort of "Trump Effect" is on display in Figure \@ref(fig:interest-group-drop-prop-viol); indeed, every possible category sharply increases from 2012 to 2016, thus reflecting the acerbic tone of that election and of the candidates involved. 

To round out the initial analysis of campaign advertisements and the degree to which they have featured violations to the principles of civil-military relations, I then include only those advertisements sponsored by interest groups. These are displayed in Figure \@ref(fig:interest-group-only-viol). 

```{r interest-group-only-viol, include=T, echo=FALSE, warning=FALSE, message=FALSE, error=FALSE, fig.height=4, fig.width=6, fig.cap="Proportion and Rate of Campaign Advertisements Sponsored by Interest Groups That Violate Central Principles of Civil-Military Relations, 2000-2016"}

# drop observations where sponsor==1, 2, or 3 using the filter command 

igs_violations_wrap <- prom_figs_df %>% group_by(year) %>% filter(sponsor== 4) %>% mutate(
  endorse_count=sum(vet_fam_end),
  attack_count=sum(vet_fam_atk),       ### instances of violations by type
  non_int_count=sum(vet_fam_out_issues),
  num_violations=sum(vet_fam_end+vet_fam_atk+vet_fam_out_issues),
  endorse_prop=endorse_count/Yrcount, ### proportion of ads w/any viol
  attack_prop=attack_count/Yrcount,
  non_int_prop=non_int_count/Yrcount,
  violations_rate=num_violations/Yrcount, ### rate of violations per election cycle
  ad_violating=case_when(
    vet_fam_end==1 | vet_fam_atk == 1 | vet_fam_out_issues ==1 ~ 1,
    vet_fam_end==0 & vet_fam_atk == 0 & vet_fam_out_issues == 0 ~ 0),
  prop_ads_violating=(sum(ad_violating)/Yrcount)) %>% view()

# now create graphs of proportion of ads violating
igs_prop_viol_ch4 <- igs_violations_wrap %>% 
  select(year, endorse_prop:violations_rate, prop_ads_violating) %>% pivot_longer(endorse_prop:prop_ads_violating, names_to = "statistic", values_to="Proportion") %>% distinct() %>% view()

igs_prop_viol_ch4_graph <- igs_prop_viol_ch4 %>% ggplot(aes(x=year, y=Proportion, color=statistic)) + geom_point() + geom_smooth(se=FALSE) + labs(x="Election", y="Proportion/Rate") + scale_color_discrete(name="Proportion", breaks=c("endorse_prop", "attack_prop", "non_int_prop", "violations_rate", "prop_ads_violating"), labels=c("Endorse Candidate", "Attack Candidate", "Address Topics\nOutside of Expertise", "Violation Rate", "Portion of\nAds Violating")) + theme(legend.position="bottom") + guides(color = guide_legend(ncol=3,nrow=2,byrow=TRUE))

igs_prop_viol_ch4_graph
```

Our understanding of the picture is now much more clear: military-affiliated interest groups drove the production of advertisements that featured military actors endorsing candidates and attacking candidates in 2004; since that period, however, they have become increasingly involved in elections. Civilian political parties and candidates drove the harnessing of military figures to appear on behalf of and in opposition to the candidates in 2016. Military affiliated interest groups were most involved in the 2004 Election, and have taken on an increasing role and willingness to produce advertisements that violate the principles of civil-military relations, with the exception of the principle of non-interference. All 5 instances of violations to this principle occurred in advertisements produced by civilian candidates or the political parties they represent.  

## Explaining Campaign Ads that Feature Explicitly Partisan Behavior by Military Figures

```{r reg data manipulation, include=FALSE, echo=FALSE, message=FALSE, error=FALSE}

# prep the data 

reg_df_ch4 <- violations_wrap %>%view()

# also, use mutate function to include several control 
# variables to the data frame.  

reg_df_ch4 <- reg_df_ch4 %>% mutate(
  can_vet = replace(can_vet, can_vet != 1, 0),
  partisan_act=case_when(
    vet_fam_end ==1 | vet_fam_atk ==1 | vet_fam_out_issues ==1 ~ 1, 
    vet_fam_end !=1 & vet_fam_atk!=1 & vet_fam_out_issues !=1 ~ 0 )) %>% view()

```

The next several pages reveal the results of a detailed regression analysis. The first set of regressions all use OLS regression, where the dependent variables are the proportion of ads aired in each election cycle that violate at least one of the central principles of civil-military relations and the overall violation rate of advertisements in each election cycle. This first set of regressions should thus be viewed as a macro level view of each election cycle. The units of analysis in this first series of regressions are the five separate election cycles that occur over the date range 2000 - 2016. 

```{r ch4-initreg, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}

### Here I construct my regression models. Begin with OLS and then move towards logit models. 
## create OLS data frames for all advertisements, ads
## produced by political parties, and ads produced by Int Groups. 
## Select only the proportions I am interested in 
## as my DV and drop any other variables I do not need. 

# all ad df
reg_df_ch4_ols <- reg_df_ch4 %>% select(house_polar:prestige_gap, violations_rate, prop_ads_violating) %>% distinct() %>% view()

## OLS Model 1: DV is prop_ads_violating

mod_ols_prop_violating <- lm(prop_ads_violating ~ mil_prestige + house_polar + HostCasRateLagged, data=reg_df_ch4_ols)
summary(mod_ols_prop_violating)

mod_ols_prop_violating_gap <- lm(prop_ads_violating ~ prestige_gap + house_polar + HostCasRateLagged, data=reg_df_ch4_ols)
summary(mod_ols_prop_violating_gap)

## OLS Model 2: DV is violation rate

mod_ols_violations_rate <- lm(violations_rate ~ mil_prestige + house_polar + HostCasRateLagged, data=reg_df_ch4_ols)
summary(mod_ols_violations_rate)

## Now Create a df only for ads sponsored by 
## parties, candidates, or a combination of the two. 
drop_igs_violations_wrap_ols <- drop_igs_violations_wrap %>% select(house_polar:prestige_gap, violations_rate, prop_ads_violating) %>% distinct() %>% view()

## OLS Model 1: DV is prop_ads_violating

no_ig_ols_prop_violating <- lm(prop_ads_violating ~ mil_prestige + sen_polar + HostCasRateLagged, data=drop_igs_violations_wrap_ols)
summary(no_ig_ols_prop_violating)

no_ig_ols_prop_violating_gap <- lm(prop_ads_violating ~ prestige_gap + sen_polar + HostCasRateLagged, data=drop_igs_violations_wrap_ols)
summary(no_ig_ols_prop_violating_gap)

## OLS Model 2: DV is violation rate

no_ig_ols_violations_rate <- lm(violations_rate ~ mil_prestige + sen_polar + HostCasRateLagged, data=drop_igs_violations_wrap_ols)
summary(no_ig_ols_violations_rate)

## No Statistical Significance for these models

## Now Create a df only for ads sponsored by 
## interest groups 

igs_violations_wrap_ols <- igs_violations_wrap %>% select(house_polar:prestige_gap, violations_rate, prop_ads_violating) %>% distinct() %>% view()

## OLS Model 1: DV is prop_ads_violating

ig_ols_prop_violating <- lm(prop_ads_violating ~ mil_prestige + house_polar + HostCasRateLagged, data=igs_violations_wrap_ols)
summary(ig_ols_prop_violating)

ig_ols_prop_violating_gap <- lm(prop_ads_violating ~ prestige_gap + house_polar + HostCasRateLagged, data=igs_violations_wrap_ols)
summary(ig_ols_prop_violating_gap)

## OLS Model 2: DV is violation rate

ig_ols_violations_rate <- lm(violations_rate ~ mil_prestige + house_polar + HostCasRateLagged, data=igs_violations_wrap_ols)
summary(ig_ols_violations_rate_gap)
```

I therefore only control for broad, macro level factors in this first set of regressions. These include, first, the level of military prestige, which is proxied using a measure of public trust in the military that comes from Gallup public opinion polling regarding trust in American institutions [@noauthor_confidence_2020]. This control variable is intended to capture the theoretical expectation that civilians are more likely to use the military for their own partisan gain and purposes as the level of military prestige rises. In some models, I instead use a control for the prestige gap, which uses the same data to create a measure of the gap between the level of public trust in the military versus that of Congress. I include this measure instead of the level of public trust in the military in some models because, as the beginning of this chapter described, the level of trust in the military does not significantly vary over the time horizon of the data explored in this chapter. The prestige gap between the military and Congress varies more, and while the two measures are not the same theoretically, perhaps the prestige gap will intimate what the raw level of military prestige may not.  

I also include levels of congressional polarization in the US House and Senate, which come from a measure of ideology based off of roll-call voting data over time [@jeffrey_b_lewis_voteview_2020].^[In these models, I do not include a measure of affective polarization, which measures how partisans feel about political parties and/or those Americans who hold differing political views, as reported in data from the American National Election Studies (ANES) [@iyengar_affect_2012; @iyengar_strengthening_2018; @iyengar_origins_2019. The reason that I do not include them is because the measure of the level of affective polarization between 2012 and 2016 shows a decline, as I described previously in chapter three of this dissertation. I am not ignoring this reality, but the focus of this particular chapter is on the level of military prestige, and it is more consistent, in my view, to use the two congressional polarization measurements, each of which reflects increases over the date range examined]. Polarization is included as a control because it changes over the time horizon examined, and because one might suspect that civilians likewise react to changing levels in polarization by becoming increasingly willing to engage in overtly partisan behaviors as polarization rises. 

I also include a measure of military casualties, and lag the measure by one year. I lag the level of casualties because election advertisements run over a period of months before an election actually takes place. Although general elections occur near the end of the year in November, the data include advertisements that ran during the primary campaign of each election, which in many cases begin the calendar year prior to an election. The previous year's casualty levels may thus be partly responsible for spurring the decision made by civilian candidates to design advertisements that feature military actors behaving in a particular manner.^[This variable is constructed from a number of sources, and consists of the number of hostile US casualties sustained in a given calendar year per 100,000 Active Duty military forces. I include in this measurement those casualties that are the result of what the Department of Defense classifies as either "hostile action" or a "terrorist attack." Annual data is provided from the Department of Defense from 1980-2010. I then used data from the Department of Defense regarding the wars in Iraq and Afghanistan to construct a measurement for years outside of this time range. See @department_of_defense_dcas_2011; @department_of_defense_dcas_2021; @department_of_defense_dcas_2021-1; @department_of_defense_dcas_2021-2; @department_of_defense_dcas_2021-3; @department_of_defense_dcas_2021-4; @department_of_defense_dcas_2021-5; @department_of_defense_dmdc_2021 for more details.]

In models 1-3, the dependent variable is the proportion of ads (models 1-2) and the violation rate (model 3) of all advertisements aired in an election cycle. In models 4-6, the dependent variables are the same, respectively, but for advertisements sponsored strictly by political parties and/or political candidates (no interest groups). In models 7-9, the dependent variables are the same as well, but only for advertisements produced by interest groups, which would ostensibly include those produced by veteran-affiliated interest groups. 

These results are displayed in Table \@ref(tab:ch4regression1). 
\singlespace
```{r reg results, include=T, echo=FALSE, results="asis", warning=FALSE, message=FALSE}
stargazer(mod_ols_prop_violating, mod_ols_prop_violating_gap, mod_ols_violations_rate, no_ig_ols_prop_violating, no_ig_ols_prop_violating_gap, no_ig_ols_violations_rate, ig_ols_prop_violating, ig_ols_prop_violating_gap, ig_ols_violations_rate,
          type="latex",
          label="tab:ch4regression1",
          title="OLS Regression Results, Proportion/Rate of Presidential Election Ads Violating Principles of Civil-Military Relations", 
          covariate.labels = c('Trust in Military', 'Prestige Gap', 'House Polarization', 'Senate Polarization', 'Casualty Rate (lagged)'),
          dep.var.labels = c('Proportion', 'Rate', 'Proportion', 'Rate', 'Proportion', 'Rate'),
          model.names=F,
          header=FALSE,
          column.sep.width = "-15pt",
          align=TRUE,
          font.size="small",
          no.space=TRUE,
          omit.stat=c("f", "ser"),
          float.env = "sidewaystable"
          )
```
\doublespace

The results shown in Table \@ref(tab:ch4regression1) indicate that none of the variables have much, if any, explanatory power in shaping the proportion of ads or the rate at which ads aired in each election that violate the principles of civil-military relations. Across all model specifications, the coefficient for trust in the military is positive, suggesting that there is a positive relationship, but the coefficient is small in scale and statistically insignificant in each model.

Furthermore, all other variables likewise seem to reflect little importance. The coefficients for the prestige gap, types of polarization and the lagged casualty rate are statistically insignificant in every model. This is true regardless of whether we analyze the levels and rates of all advertisements aired in each election cycle (models 1-3), or strictly those produced by political parties and candidates (models 4-6) or interest groups (models 7-9). In short, the results of Table \@ref(tab:ch4regression1) prevent us from concluding that there is a strong relationship between any of the control variables and either the proportion or the violation rate of presidential campaign ads that violate the principles of civil-military relations.

These results change, however, when we shift the unit of analysis from the election cycle to the individual advertisement. As was shown in the earlier figures of this chapter describing the data, the five election cycles covered in the data are different in many ways. The 2004 Election witnessed a significant level of partisan activity at the hands of military figures, and so did the 2016 Election. This, combined with the fact that the 2000 Election, at least insofar as the data is concerned, showed very little partisan activity by military figures in campaign ads, renders an analysis of the election cycle only somewhat incomplete. The next series of regression models therefore incorporate logistic regression, where the dependent variable is the likelihood that an individual advertisement violates one or more of the central principles of civil-military relations, or features a prominent appearance by a military figure.  

In addition to the controls used in the previous set of models, I control for whether at least one candidate mentioned or pictured in the advertisement is a veteran. The reason for including this control is that perhaps civilian candidates who themselves are veterans may be more inclined to tout their own security credentials, and to use military figures to do so in ways that violate the principles of civil-military relations. Additionally, civilian veteran candidates may likewise have the effect such that their political opponents who are not veterans may seek out military figures to undermine the veteran candidate's service record or views regarding the use of force.  

I also include a control for the party of the political candidate supported in each advertisement. This variable is designed to capture any distinction between the major parties in terms of how each party is inclined to use military figures in campaign ads to begin with. Consistent with the "issue ownership" literature in American politics, it could be that Democratic candidates may be more inclined to employ a military figure in overtly partisan ways on the grounds that doing so may make the Democratic candidate appear to be strong on national security issues. Over the past several decades, Republicans have generally been perceived by the public as stronger than Democrats on defense issues and national security, and thus Democrats be more inclined (and Republicans less so) to engage in efforts to shore up their defense credentials [@petrocik_issue_2003; @sides_consequences_2007].

```{r reg-models2, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}

# Logistic Regression Model 3: DV is Prominent Appearance

mod_appearance <- glm(appearance ~ mil_prestige + house_polar + HostCasRateLagged + can_vet + party_reg, 
    family=binomial(link="logit"),
    data=reg_df_ch4
    )
summary(mod_appearance)


mod_appearance_gap <- glm(appearance ~ prestige_gap + house_polar + HostCasRateLagged + can_vet + party_reg, 
    family=binomial(link="logit"),
    data=reg_df_ch4
    )
summary(mod_appearance_gap)

# Logistic Regression Model 4: DV is Any Violation

mod_anyviol_ch4 <- glm(partisan_act ~ mil_prestige + sen_polar + HostCasRateLagged + can_vet + party_reg, 
    family=binomial(link="logit"),
    data=reg_df_ch4
    )
summary(mod_anyviol_ch4)


mod_anyviol_ch4_gap <- glm(partisan_act ~ prestige_gap + sen_polar + HostCasRateLagged + can_vet + party_reg,
    family=binomial(link="logit"),
    data=reg_df_ch4
    )
summary(mod_anyviol_ch4_gap)

# Logistic Regression Model 5: DV is Endorsement

mod_endorse_ch4 <- glm(vet_fam_end ~ mil_prestige + house_polar + HostCasRateLagged + can_vet + party_reg, 
    family=binomial(link="logit"),
    data=reg_df_ch4
    )
summary(mod_endorse_ch4)

mod_endorse_ch4_gap <- glm(vet_fam_end ~ prestige_gap + sen_polar + HostCasRateLagged + can_vet + party_reg,
    family=binomial(link="logit"),
    data=reg_df_ch4
    )
summary(mod_endorse_ch4_gap)

# Logistic Regression Model 6: DV is attack

mod_atk_ch4 <- glm(vet_fam_atk ~ mil_prestige + sen_polar + HostCasRateLagged + can_vet + party_reg,
    family=binomial(link="logit"),
    data=reg_df_ch4
    )
summary(mod_atk_ch4)

mod_atk_ch4_gap <- glm(vet_fam_atk ~ prestige_gap + sen_polar + HostCasRateLagged + can_vet + party_reg, 
    family=binomial(link="logit"),
    data=reg_df_ch4
    )
summary(mod_atk_ch4_gap)


# Logistic Regression Model 7: DV is non-int

mod_int_ch4 <- glm(vet_fam_out_issues ~ mil_prestige + house_polar + HostCasRateLagged + can_vet + party_reg, 
    family=binomial(link="logit"),
    data=reg_df_ch4
    )
summary(mod_int_ch4)

mod_int_ch4_gap <- glm(vet_fam_out_issues ~ prestige_gap + house_polar + HostCasRateLagged + can_vet + party_reg,
    family=binomial(link="logit"),
    data=reg_df_ch4
    )
summary(mod_int_ch4_gap)

```
The specific dependent variables vary by model specification. In models 1-2, the dependent variable is the log of the odds of a military figure prominently appearing in an advertisement. In models 3-4, the dependent variable is the log of the odds of a military figure engaging in _any_ of the three operationalized behaviors denoting violations to the principles of civil-military relations: endorsing a political candidate, attacking a political candidate, or speaking out on an issue that is tangential to the expertise of the military actor. In models 5-9, the dependent variable is the log of the odds that an advertisement features a military actor who commits each of these three violations separately: endorsing a candidate (models 5-6), attacking a political candidate (models 7-8), or addressing a topic that is tangential to one's military experiences (model 9).^[For readability, I only include one model in this table where the dependent variable denotes a military figure addressing an outside topic because all control variables, regardless of model specification, yield statistically insignificant results.] These results, for all advertisements in the data, are displayed in Table \@ref(tab:ch4regression2).

```{r reg-results-set2, include=T, echo=FALSE, results="asis", warning=FALSE, message=FALSE}
stargazer(mod_appearance, mod_appearance_gap, mod_anyviol_ch4, mod_anyviol_ch4_gap, mod_endorse_ch4, mod_endorse_ch4_gap, mod_atk_ch4, mod_atk_ch4_gap, mod_int_ch4,
          type="latex",
          label="tab:ch4regression2",
          title="Log Odds of a Presidential Campaign TV Ad Featuring Prominent Military Actors Engaging in Political Behaviors", 
          covariate.labels = c('Trust in Military', 'Prestige Gap', 'House Polarization', 'Senate Polarization',  'Casualty Rate (lagged)', 'Candidate-Veteran', 'Party-Other', 'Party-GOP'),
          dep.var.labels = c('Appear', 'Any Viol.', 'Endorse', 'Attack', 'Topic'),
          model.names = FALSE,
          header=FALSE,
          column.sep.width = "-15pt",
          align=TRUE,
          font.size="small",
          no.space=TRUE,
          omit.stat=c("f","ser"),
          float.env="sidewaystable"
)
         
```

In Table \@ref(tab:ch4regression2), the coefficient for trust in the military is positive and statistically significant in three of the five models in which it is used, including when the dependent variable is a military figure appearing prominently (model 1), engaging in any violation (model 3), and attacking a political candidate (model 7). Somewhat interestingly, the prestige gap does not seem to matter, at least not consistently, as the coefficient for the prestige gap changes directions across model specifications and is only statistically significant in one model.  

However, a candidate's veteran status seems to be important. The coefficient for candidate-veteran is positive and statistically significant in all but the two model specifications in which the dependent variable indicates a military figure endorsing a political candidate. The results also show that advertisements that support GOP candidates are consistently less likely to use military figures to appear as well as violate the principles of civil-military relations. The coefficient for Party-GOP is negative and highly statistically significant across all model specifications except in model 9, when the dependent variable indicates a military figure is violating the principle of non-interference by addressing a topic that is tangential or outside of one's military expertise. 

As a whole, the results of Table \@ref(tab:ch4regression2) suggest that the level of public trust in the military moderately shapes the degree to which military figures appear in campaign advertisements as well as how these figures act when they appear. In particular, the results seem to suggest that as the level of military prestige rises, presidential campaign television advertisements are increasingly likely to feature military actors and that these actors are increasingly likely to violate at least one of the principles of civil-military relations when they appear. Furthermore, the principle that these actors are most likely to violate is that of attacking a political candidate running for office. 

### Parties and Candidates Only

In this section, I focus only on those advertisements in the data sponsored by political parties and or political candidates. This will enable perhaps what can be considered a more precise glimpse into how truly civilian actors - candidates and the political parties - are behaving. I include the very same controls as in the previous regression table.   

```{r party-can-data prep, include=FALSE, echo=FALSE, warning=FALSE}

## Advertisement-level data for ads 
## produced by parties/candidates only

no_ig_reg_df_ch4 <- reg_df_ch4 %>% filter(sponsor!=4) %>% view()

# Logistic Regression Model: DV is Prominent Appearance
noig_appearance <- glm(appearance ~ mil_prestige + house_polar + HostCasRateLagged + can_vet + party_reg, 
    family=binomial(link="logit"),
    data=no_ig_reg_df_ch4
    )

noig_appearance_robust<-robustify(noig_appearance, cluster=year)
summary(noig_appearance_robust)

noig_appearance_gap <- glm(appearance ~ prestige_gap + house_polar + HostCasRateLagged + can_vet + party_reg, 
    family=binomial(link="logit"),
    data=no_ig_reg_df_ch4
    )

noig_appearance_gap_robust <-robustify(noig_appearance_gap, cluster=year)
summary(noig_appearance_gap_robust)


# Logistic Regression Model: DV is Any Violation
noig_anyviol_ch4 <- glm(partisan_act ~ mil_prestige + sen_polar + HostCasRateLagged + can_vet + party_reg, 
    family=binomial(link="logit"),
    data=no_ig_reg_df_ch4
    )

noig_anyviol_ch4_robust <- robustify(noig_anyviol_ch4, cluster = year)
summary(noig_anyviol_ch4_robust)

noig_anyviol_ch4_gap <- glm(partisan_act ~ prestige_gap + sen_polar + HostCasRateLagged + can_vet + party_reg,
    family=binomial(link="logit"),
    data=no_ig_reg_df_ch4
    )

noig_anyviol_ch4_gap_robust <- robustify(noig_anyviol_ch4_gap, cluster=year)
summary(noig_anyviol_ch4_gap_robust)


# Logistic Regression Model: DV is Endorsement
noig_endorse_ch4 <- glm(vet_fam_end ~ mil_prestige + house_polar + HostCasRateLagged + can_vet + party_reg, 
    family=binomial(link="logit"),
    data=no_ig_reg_df_ch4
    )

noig_endorse_ch4_robust <- robustify(noig_endorse_ch4, cluster=year)
summary(noig_endorse_ch4_robust)


noig_endorse_ch4_gap <- glm(vet_fam_end ~ prestige_gap + house_polar + HostCasRateLagged + can_vet + party_reg,
    family=binomial(link="logit"),
    data=no_ig_reg_df_ch4
    )

noig_endorse_ch4_gap_robust <- robustify(noig_endorse_ch4_gap, cluster=year)
summary(noig_endorse_ch4_gap_robust)

# Logistic Regression Model: DV is attack
noig_atk_ch4 <- glm(vet_fam_atk ~ mil_prestige + sen_polar + HostCasRateLagged + can_vet + party_reg, 
    family=binomial(link="logit"),
    data=no_ig_reg_df_ch4
    )

noig_atk_ch4_robust <- robustify(noig_atk_ch4, cluster=year)
summary(noig_atk_ch4_robust)

noig_atk_ch4_gap <- glm(vet_fam_atk ~ prestige_gap + sen_polar + HostCasRateLagged + can_vet + party_reg,
    family=binomial(link="logit"),
    data=no_ig_reg_df_ch4
    )
summary(noig_atk_ch4_gap)


# Logistic Regression Model 7: DV is non-int
noig_int_ch4 <- glm(vet_fam_out_issues ~ mil_prestige + house_polar + HostCasRateLagged + can_vet + party_reg, 
    family=binomial(link="logit"),
    data=no_ig_reg_df_ch4
    )
summary(noig_int_ch4)

noig_int_ch4_sen <- glm(vet_fam_out_issues ~ mil_prestige + sen_polar + HostCasRateLagged + can_vet + party_reg,
    family=binomial(link="logit"),
    data=no_ig_reg_df_ch4
    )
summary(noig_int_ch4_sen)

```

The results of this series of regressions is displayed in Table \@ref(tab:ch4regression3).

\singlespace
```{r reg-results-set3, include=T, echo=FALSE, results="asis", warning=FALSE, message=FALSE}
stargazer(noig_appearance, noig_appearance_gap, noig_anyviol_ch4, noig_anyviol_ch4_gap, noig_endorse_ch4, noig_endorse_ch4_gap, noig_atk_ch4, noig_atk_ch4_gap, noig_int_ch4,
          type="latex",
          label="tab:ch4regression3",
          title="Log Odds of a Presidential Campaign TV Ad Sponsored by Parties and/or Candidates Featuring Prominent Military Actors Engaging in Political Behaviors", 
          covariate.labels = c('Trust in Military', 'Prestige Gap', 'House Polarization', 'Senate Polarization',  'Casualty Rate (lagged)', 'Candidate-Veteran', 'Party-Other', 'Party-GOP'),
          dep.var.labels = c('Appear', 'Any Viol', 'Endorse', 'Attack', 'Topic'),
          model.names=FALSE,
          header=FALSE,
          column.sep.width = "-15pt",
          align=TRUE,
          font.size="small",
          no.space=TRUE,
          omit.stat=c("f", "ser"),
          float.env="sidewaystable"
)
         
```
\doublespace

Table \@ref(tab:ch4regression3) contains a couple of interesting findings. Note that the coefficient for trust in the military remains positive and statistically significant for whether a military figure appears in a prominent way (model 1) or commits any violation (model 3), but is now positive and statistically strong when the dependent variable indicates a military figure endorsing a political candidate (model 5). However, the coefficient for trust in the military is not statistically significant when the dependent variable is the log of the odds that a military figure attacks a political candidate (model 7). This is a change from the previous set of regression models, which analyzed all advertisements and not just those produced strictly by political parties and/or candidates.  

In this set of models, the prestige gap seems to become entirely unimportant in that the coefficient for the prestige gap is statistically insignificant in all model specifications and changes sign in two of the four models in which it is used. There is some very mild evidence to suggest that civilian leaders in these models respond to growing polarization by choosing to feature military actors in campaign advertisements and to employ these actors in ways that violate the principles of civil-military relations, but this evidence is inconsistent. Indeed, the coefficient for polarization is statistically significant and positive in three of the nine models used, and in none of the models in which the dependent variable is a specific violation of the principles of civil-military relations. This suggests that the main driver of the choice by parties and candidates to employ military figures in the ads that they produce is not the level of polarization itself. 

However, the a candidate's veteran status again appears to matter significantly. The coefficient for candidate-veteran is positive and statistically significant in seven of nine models, including in models which cover two of the three specific instances of behaviors that violate the principles of civil-military relations (endorse and attack; see models 5, 6, and 8). Additionally, the coefficient for Party-GOP remains negative and statistically significant across eight of the nine model specifications, suggesting that GOP parties and candidates are less likely than Democrats to employ military figures in campaign ads in ways that violate the principles of civil-military relations. 

In short, Table \@ref(tab:ch4regression3) shoes results which are only slightly rather than significantly different from the results displayed in Table \@ref(tab:ch4regression2). 

### Interest Groups (Including Military-Affiliated Interest Groups)

I conduct a final set of logistic regressions on only those advertisements sponsored by interest groups, which would include those ads produced by veteran specific interest groups. I conduct this analysis to determine whether these actors, which could reasonably be considered to more approximate a military rather than civilian actor, respond differently than the truly civilian entities of parties and candidates previously examined.^[Further research is required to determine if this is in fact the case.]I include the very same set of controls as in the preceding set of regression models. The results are displayed in Table \@ref(tab:ch4regression4).

```{r int-groups-only, include=FALSE, echo=FALSE, warning=FALSE}

## No Macro Level Statistical Significance for OLS models

## Now prep advertisement-level data

ig_reg_df_ch4 <- reg_df_ch4 %>% filter(sponsor==4) %>% view()

# Logistic Regression Model: DV is Prominent Appearance
ig_appearance <- glm(appearance ~ mil_prestige + house_polar + HostCasRateLagged + can_vet + party_reg, 
    family=binomial(link="logit"),
    data=ig_reg_df_ch4
    )
summary(ig_appearance)

ig_appearance_gap <- glm(appearance ~ prestige_gap + house_polar + HostCasRateLagged + can_vet + party_reg, 
    family=binomial(link="logit"),
    data=ig_reg_df_ch4
    )
summary(ig_appearance_gap)

# Logistic Regression Model: DV is Any Violation
ig_anyviol_ch4 <- glm(partisan_act ~ mil_prestige + sen_polar + HostCasRateLagged + can_vet + party_reg, 
    family=binomial(link="logit"),
    data=ig_reg_df_ch4
    )
summary(ig_anyviol_ch4)

ig_anyviol_ch4_gap <- glm(partisan_act ~ prestige_gap + sen_polar + HostCasRateLagged + can_vet + party_reg,
    family=binomial(link="logit"),
    data=ig_reg_df_ch4
    )
summary(ig_anyviol_ch4_gap)

# Logistic Regression Model: DV is Endorsement
ig_endorse_ch4 <- glm(vet_fam_end ~ mil_prestige + house_polar + HostCasRateLagged + can_vet + party_reg, 
    family=binomial(link="logit"),
    data=ig_reg_df_ch4
    )
summary(ig_endorse_ch4)


ig_endorse_ch4_gap <- glm(vet_fam_end ~ prestige_gap + house_polar + HostCasRateLagged + can_vet + party_reg,
    family=binomial(link="logit"),
    data=ig_reg_df_ch4
    )
summary(ig_endorse_ch4_gap)

# Logistic Regression Model: DV is attack
ig_atk_ch4 <- glm(vet_fam_atk ~ mil_prestige + sen_polar + HostCasRateLagged + can_vet + party_reg, 
    family=binomial(link="logit"),
    data=ig_reg_df_ch4
    )
summary(ig_atk_ch4)

ig_atk_ch4_gap <- glm(vet_fam_atk ~ prestige_gap + sen_polar + HostCasRateLagged + can_vet + party_reg,
    family=binomial(link="logit"),
    data=ig_reg_df_ch4
    )
summary(ig_atk_ch4_gap)


# Logistic Regression Model 7: DV is non-int
ig_int_ch4 <- glm(vet_fam_out_issues ~ mil_prestige + house_polar + HostCasRateLagged + can_vet + party_reg,
    family=binomial(link="logit"),
    data=ig_reg_df_ch4
    )
summary(ig_int_ch4)

ig_int_ch4_sen <- glm(vet_fam_out_issues ~ mil_prestige + sen_polar + HostCasRateLagged + can_vet + party_reg,
    family=binomial(link="logit"),
    data=ig_reg_df_ch4
    )
summary(ig_int_ch4_sen)

```


\singlespace
```{r reg-results-set4, include=T, echo=FALSE, results="asis", warning=FALSE, message=FALSE}
stargazer(ig_appearance, ig_appearance_gap, ig_anyviol_ch4, ig_anyviol_ch4_gap, ig_endorse_ch4, ig_endorse_ch4_gap, ig_atk_ch4, ig_atk_ch4_gap, ig_int_ch4,
          type="latex",
          label="tab:ch4regression4",
          title="Log Odds of a Presidential Campaign TV Ad Sponsored by Interest Groups Featuring Prominent Military Actors Engaging in Political Behaviors", 
          covariate.labels = c('Trust in Military', 'Prestige Gap', 'House Polarization', 'Senate Polarization',  'Casualty Rate (lagged)', 'Candidate-Veteran', 'Party-Other', 'Party-GOP'),
          dep.var.labels = c('Appear', 'Any Viol', 'Endorse', 'Attack', 'Topic'),
          model.names=FALSE,
          header=FALSE,
          column.sep.width = "-15pt",
          align=TRUE,
          font.size="small",
          no.space=TRUE,
          omit.stat=c("f", "ser"),
          float.env="sidewaystable"
)
         
```
\doublespace

Table \@ref(tab:ch4regression4) shows a couple of brief but somewhat interesting findings. The prestige gap does not appear to impact interest groups in any meaningful way. The level of trust in the military, on the other hand, does appear to be important, but it is not conclusive how or why it is important. The coefficient for trust in the military is positive and statistically significant in models 1 and 7, which denote prominently appearing and attacking candidates, respectively, but negative in model 5 and statistically insignificant in both models 3 and 5. 

Additionally, in this particular set of regression models, the coefficient for candidate-veteran is positive and statistically significant only in models 7 and 8, suggesting that interest groups are particularly prone to feature military figures who attack candidates when the candidate is a veteran. Additionally, the coefficient for Party-GOP is strong and statistically significant only in models 3 and 4 here, whereas the same coefficient was negative and statistically significant across all bot one of the model specifications in the previous two sets of regressions. This difference suggests that the distinction between how interest groups affiliated with the two major parties employ military figures to engage in behaviors that violate the principles of civil-military relations is less pronounced than civilian candidates and political parties, strictly speaking.^[To the extent that the interest group population in the data is dominated by veteran affiliated interest groups (a statement we cannot make conclusively given how the data is coded for interest groups), it appears, then, that veteran affiliated interest groups are less likely to be marked by party distinctions. In other words, veteran affiliated interest groups appear to be less distinct in the party affiliation of whom they endorse or criticize.]

## Predicted Probability

In order to gauge the substantive importance of the level of trust in the military and how it impacts the likelihood that campaign advertisements feature military figures who in some way violate the principles of civil-military relations when they appear in these advertisements, we can transform the logistic regression results from Table \@ref(tab:ch4regression2) into predicted probabilities. Figure \@ref(fig:pred-prob-ch4-sen) plots the predicted probability of a presidential campaign television ad featuring a military figure who engages in any behavior that undermines the principles of civil-military relations. I fix the level of polarization in the Senate and the lagged hostile troop casualty rate at their respective means over the period from 2000 - 2016. I hold the value of candidate-veteran at 0, denoting that the candidates portrayed in a hypothetical ad are not veterans, and hold the value of the party that the advertisement supports as the Republican Party. The vertical colored lines indicate the level of military prestige as measured by public opinion polling in certain points of time in contemporary US history [@noauthor_confidence_2020]. The color of the line indicates the party of the President who was in office at the time. 

```{r pred-prob-ch4-prep, include=FALSE, warning=FALSE, message=FALSE, error=FALSE}

# I create two regression models that mirror 
# mod_anyviol_ch4 and mod_anyviol_ch4_house from 
# Tab:Regression-set 2. 
# Hold hostile casualty rate, sen polarization at means
# set party as GOP in model 1 and Dems in model 2 using 
# the party_reg_numeric variable. 

ch4_pp_sen_anyviol <- glm(partisan_act ~ mil_prestige + sen_polar + HostCasRateLagged + can_vet + party_reg_numeric, 
    family=binomial(link="logit"),
    data=reg_df_ch4
    )
summary(ch4_pp_sen_anyviol)

ch4_pp_house_anyviol <- glm(partisan_act ~ mil_prestige + house_polar + HostCasRateLagged + can_vet + party_reg_numeric,
    family=binomial(link="logit"),
    data=reg_df_ch4
    )
summary(ch4_pp_house_anyviol)

# using senate polar and party = GOP
pp_data_ch4_sen <- tibble(mil_prestige=seq(55, 85, by=1), sen_polar = mean(reg_df_ch4$sen_polar), HostCasRateLagged=mean(reg_df_ch4$HostCasRateLagged), party_reg_numeric=0, can_vet=0) %>% crossing(partisan_act = c(0,1)) %>% view()

augment(ch4_pp_sen_anyviol, newdata = pp_data_ch4_sen, se_fit=TRUE)

# using house polar and party = DEM/Other
pp_data_ch4_house <- tibble(mil_prestige=seq(55, 85, by=1), house_polar = mean(reg_df_ch4$sen_polar), HostCasRateLagged=mean(reg_df_ch4$HostCasRateLagged), party_reg_numeric=1, can_vet=1) %>% crossing(partisan_act = c(0,1)) 

augment(ch4_pp_house_anyviol, newdata = pp_data_ch4_house, se_fit=TRUE)

# prediction data for Senate and party is GOP
predictions_ch4_sen <- augment(ch4_pp_sen_anyviol, newdata=pp_data_ch4_sen, se_fit=TRUE) %>% 
  mutate(
    .prob=plogis(.fitted),
    conf.low = plogis(.fitted - (qnorm(.975) * .se.fit)),
    conf.high = plogis(.fitted + (qnorm(.975) * .se.fit)),
    pred_final_rec = case_when(
      .prob >= .5 ~ "Likely",
      .prob < .5 ~ " Not likely"
    )
  )

predictions_ch4_sen %>% arrange(.prob, .fitted, partisan_act, desc(.fitted)) %>% 
  view()

# now let's try some graphs
# notice that the predictions are not in (0, 1)
ggplot(predictions_ch4_sen) +
  aes(x = .fitted) +
  geom_histogram()

# if we wanted predictions on the probability scale,
#   we predict on the log scale,
#   then apply inverse-logit (logistic CDF)
# use plogis(): cumulative probability of the logistic distribution

# demonstration: scatter log odds (x) and predicted probability (y)
# with 95% CIs
# (CIs don't look smooth in this example because .se.fit depends on X data)
predictions_ch4_sen %>%
  ggplot() +
  aes(x = .fitted, y = .prob) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high)) +
  geom_smooth() 

# prediction data for House Polarization and party is Dem
predictions_ch4_house <- augment(ch4_pp_house_anyviol, newdata=pp_data_ch4_house, se_fit=TRUE) %>% 
  mutate(
    .prob=plogis(.fitted),
    conf.low = plogis(.fitted - (qnorm(.975) * .se.fit)),
    conf.high = plogis(.fitted + (qnorm(.975) * .se.fit)),
    pred_final_rec = case_when(
      .prob >= .5 ~ "Likely",
      .prob < .5 ~ " Not likely"
    )
  )

predictions_ch4_house %>% arrange(.prob, .fitted, partisan_act, desc(.fitted)) %>% 
  view()

# now let's try some graphs
# notice that the predictions are not in (0, 1)
ggplot(predictions_ch4_house) +
  aes(x = .fitted) +
  geom_histogram()

# if we wanted predictions on the probability scale,
#   we predict on the log scale,
#   then apply inverse-logit (logistic CDF)
# use plogis(): cumulative probability of the logistic distribution

# demonstration: scatter log odds (x) and predicted probability (y)
# with 95% CIs
# (CIs don't look smooth in this example because .se.fit depends on X data)
predictions_ch4_house %>%
  ggplot() +
  aes(x = .fitted, y = .prob) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high)) +
  geom_smooth() 

```

```{r pred-prob-ch4-sen, include=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Predicted Probability of Military Military Figures Violating Central Principles of Civil-Military Relations in Presidential Campaign TV Ads", fig.height=4, fig.width=6}
predictions_ch4_sen%>%
  ggplot() +
  aes(x = mil_prestige, y = .prob) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high)) +
  geom_smooth() + labs(
    x="Level of Military Prestige",
    y="Predicted Probability", caption="Senate Polarization Held Constant,\nNon-Veteran Candidates Portrayed\nAd Supports Republican Party"
  ) + scale_y_continuous(breaks=seq(-.1, .9, .1)) + 
  geom_vline(xintercept=85, size=.2, colour="tomato") + 
  geom_vline(xintercept=64, size=.2, colour="dodgerblue") + 
  geom_vline(xintercept=82, size=.2, colour="dodgerblue") +
  geom_vline(xintercept=58, size=.2, colour="tomato") +
  geom_vline(xintercept=69, size=.2, colour="dodgerblue") +
  geom_text(aes(x=85, label="Feb 1991",y=.2), colour="tomato", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=64, label="1994",y=.2), colour="dodgerblue", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=82, label="2003",y=.2), colour="tomato", angle=90, vjust=-.2, text=element_text(size=9)) + 
  geom_text(aes(x=58, label="1975",y=.2), colour="dodgerblue", angle=90, vjust=-.2, text=element_text(size=9))  + 
  geom_text(aes(x=69, label="2021", y=.2), colour="dodgerblue", angle=90, vjust=1, text=element_text(size=9))
```

I then conduct a similar graph, but alter just a few of the variables. In particular, I use the average value of polarization in the house and change the status of the hypothetical candidates appearing in an advertisement to veterans.  I also change the party whom the ad supports to Democrat instead of Republican. This graph is portrayed in Figure \@ref(fig:pred-prob-ch4-house).

```{r pred-prob-ch4-house, include=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Predicted Probability of Military Military Figures Violating Central Principles of Civil-Military Relations in Presidential Campaign TV Ads", fig.height=4, fig.width=6}
predictions_ch4_house%>%
  ggplot() +
  aes(x = mil_prestige, y = .prob) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high)) +
  geom_smooth() + labs(
    x="Level of Military Prestige",
    y="Predicted Probability", caption="House Polarization Held Constant\nVeteran Candidate(s) Portrayed\nAd Supports Democratic Party") + 
  scale_y_continuous(breaks=seq(-.1, .5, .1)) + 
  geom_vline(xintercept=85, size=.3, colour="tomato") + 
  geom_vline(xintercept=64, size=.2, colour="dodgerblue") + 
  geom_vline(xintercept=82, size=.2, colour="dodgerblue") +
  geom_vline(xintercept=58, size=.2, colour="tomato") +
  geom_vline(xintercept=69, size=.2, colour="dodgerblue") +
  geom_text(aes(x=85, label="Feb 1991",y=.3), colour="tomato", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=64, label="1994",y=.2), colour="dodgerblue", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=82, label="2003",y=.2), colour="tomato", angle=90, vjust=-.2, text=element_text(size=9)) + 
  geom_text(aes(x=58, label="1975",y=.2), colour="dodgerblue", angle=90, vjust=-.2, text=element_text(size=9))  + 
  geom_text(aes(x=69, label="2021", y=.2), colour="dodgerblue", angle=90, vjust=1, text=element_text(size=9))
```

Both Figures \@ref(fig:pred-prob-ch4-sen) and \@ref(fig:pred-prob-ch4-house) have similar trajectories and shapes in that they portray a non-linear increase in predicted probabilities, and both charts have fairly substantial standard errors depicted, particularly at higher levels of military prestige. Additionally, truly discerning the impact of the level of military prestige is, as the regression analysis argued, a somewhat difficult endeavor in that other variables also play a role at times in driving the likelihood that a particular advertisement features a military figure who violates the principles of civil-military relations. 

On one hand, both graphs increase in probability as the level of military prestige rises. On the other, we also see a somewhat substantive difference in the predicted probabilities between the graphs, especially at higher levels. For example, the point estimates for each graph when the level of military prestige is fixed at 80 varies: when the candidate portrayed in an ad is not a veteran, the ad supports a Republican, and holding the level of polarization in the Senate fixed, the model predicts roughly a 3.7% chance that such an ad will violate one of the principles of civil-military relations; however, when the candidate portrayed in an ad is a veteran, the ad supports a Democrat, and holding the level of polarization in the House fixed, the model then predicts roughly a 7.2% chance that such an ad will violate one of the principles of civil-military relations. 

Furthermore, the significant size of the standard error intervals, especially at higher values of military prestige, indicate that some caution should be used when using these models to predict how higher levels of military prestige impact the use of military figures to undertake acts that violate the principles of civil-military relations during campaign ads. Both models suggest that at high levels of military prestige, rather extreme levels are possible. Statistically speaking, this is likely the result of several factors, including the fact that the data examined in this chapter include levels of military prestige that range over a somewhat narrow interval (64 - 75 during Election Years only) [@noauthor_confidence_2020].

## Summary of Statistical Results

Overall, this chapter has shown moderate evidence to suggest that the level of military prestige shapes the way in which civilian actors employ military figures in campaign ads. This evidence was strongest based on the regression results of Table \@ref(tab:ch4regression2) and Table \@ref(tab:ch4regression3). Whether we examine all presidential campaign ads or even only those sponsored strictly by political candidates or parties, the statistical analysis showed that as the level of military prestige rises, there is an increased likelihood that campaign ads will feature military actors who appear in prominent ways and engage in behaviors that violate the principles of civil-military relations when they do appear. The statistical results further narrowed the  violations that military actors will engage in presidential advertisements either to attacks (when we consider all advertisements as shown in Table \@ref(tab:ch4regression2)) or endorsements (when we consider only those ads sponsored by candidates and parties as shown in Table \@ref(tab:ch4regression4)). 

However, the statistical results remain nonetheless fuzzy in many places. The OLS regression models, for instance, showed no evidence that broad trends in either the military's raw level of prestige nor the level of relative military prestige shape the overall proportion of ads that violate the principles of civil-military relations,  nor the overall violation rate of advertisements in a given election cycle. This is likely the result of the fact that only five total elections were examined in this analysis. 

The results also showed that military prestige is not the only variable that shapes the ways in which military figures behave while featured in presidential campaign advertisements. The statistical analysis demonstrated that the veteran status of candidates portrayed in advertisements matters a great deal, as does whether the advertisement supports Democratic versus Republican candidates. This is likely, but not definitively, the case insofar as it is consistent with an explanation that is rooted in the "issue ownership" literature [@petrocik_issue_2003]. Over the period examined in this chapter especially, Republican candidates were perceived to be stronger on defense and national security issues than Democrats.^[According to recent Gallup polling from October 2021, 54% of surveyed Americans currently believe that the Republican Party "will do a better job of protecting the country from international terrorism and military threats," compared to 39% who believe the Democratic Party will. Since 2002, Republicans have always led Democrats on this question, with the exception of 2007, when Democrats enjoyed a 5% lead, and 2012, when the parties were tied. See @jones_gop_2021 for more details.]

## Alternate Explanations

Before concluding, it is worth briefly thinking about alternate explanations that could be responsible for driving the variation in political behaviors that challenge civil-military relations principles committed by military figures who appear prominently in campaign advertisements. 

In my view, the most compelling alternate explanation is not necessarily an altogether alternate explanation, but rather a more precise definition of the components that shape military prestige. I noted in chapter two that military prestige is a comprehensive and multi-faceted concept, and likely shaped by several components, to include the degree to which the military is esteemed by the public. But esteem is likewise shaped by multiple subcomponents, including the degree of felt separation between those who serve and those who do not (one of "the gaps" between the military and the public), and the perceived performance of the military in recent events, such as wars [@feaver_soldiers_2001;@schake_warriors_2016]. The point in revisiting the composition of military prestige is simply to argue that it is _possible_ that one or more of these subcomponents - institutional influence, societal esteem, and/or performance in recent wars or conflicts - is more responsible for shaping variation in the level of military prestige than others. 

However, it is difficult, if not impossible, to disentangle some of these subcomponents of military prestige, or at least to do so reliably. For example, did the United States lose in Iraq and in Afghanistan, two conflicts that occurred nearly across the entire time range of the data examined in this chapter? The answer to this question is not clear, and even if it were, the perceived answer to this question might very well have been different if we had asked the same question at the height of these conflicts, say in 2004 or in 2008.^[Future scholarship, with the benefit of hindsight and the passage of time, may render more definitively whether such wars ought to be coded dichotomously, but applying such a measurement scheme to these wars, in this chapter, would be irresponsible, in my view.] 

It could also be that civilian candidates and their campaign teams have simply increasingly adopted the behavior of using the military - images, symbols, and military figures - in campaign ads. But this too is a difficult explanation to prove. Even if norms regarding the use of military symbols and military figures in ads altogether are changing, so is the level of military prestige, and thus, it remains hard to disentangle the impact of simultaneously changing norms and changing levels of military prestige. 

Future research should also examine the ways in which veteran affiliated interest groups influence (or do not) political parties and civilian candidates. This chapter did not find any significant differences between the two entities, but the way that interest groups are coded in the data prevented a precise analysis of the types of interest groups. Furthermore, another important but difficult question then becomes not just whether, but how these entities interact. For example, do parties and candidates feed off of, or "follow" the lead of veteran affiliated groups, as reportedly was the case during the 2004 Election and the Swiftboat Veterans for Truth, or do parties and civilian candidates spur these groups to initiate certain activities in the first place? Obtaining a clear answer may not be possible, but if such an answer exists, such an answer would better inform our understanding of which actor - civilian or military - is responsible for initiating what might be considered the first steps of behaviors that violate the principles of civil-military relations.   

# Conclusion and Way Forward

This chapter found moderate support for H3, which stated that civilian actors increasingly violate the central principles of civil-military relations when military prestige is high relative to periods of low military prestige. The previous chapter, which examined opinion commentary authored by retired military officers, found moderate support for H2, which stated that military actors increasingly violate the central principles of civil-military relations when polarization is high relative to periods of low polarization. 

This chapter, as well as the preceding chapter, arrived at these results by conducting large-N analyses in which the dependent variable in each chapter consists of one type of a political behavior that is primarily engaged in by one type of actor. Furthermore, these results are important because they establish a connection between each type of actor, military and civilian, and a particular form of political behavior that each engages in involving the military: broadly speaking, increasing levels of polarization drives military actors to author opinion commentary in ways that violate the principles of civil-military relations, and increasing levels of military prestige drives civilian actors to use military figures in ways during campaign ads that violate the principles of civil-military relations. 

Still, there is much of the theory offered in chapter two that remains unknown. In particular, the results have not clearly indicated whether and to what degree military actors are also sensitive to changes in the level of military prestige, as the theory predicts they are. Additionally, because the previous chapters explored only one type of political behavior in depth, conclusions could not be reached about the other types of political behavior that each actor can engage in at the same time, or the broad characteristics of the political behaviors that shape the civil-military relationship in a particular era. 

The dissertation now moves to examine two different historical periods. Each of these case studies will focus on the tenor, tone, and characteristics of the political behaviors that occur in a particular time period rather than on one particular form of behavior. This methodological shift into eras other than those which have been examined here will help establish the overall patterns of political behaviors that occur in a given period of time, which then enable the researcher to employ process tracing to determine why these patterns have occurred. This is particularly important because, as the US experience proves and as this dissertation has noted, the levels of polarization and military prestige vary over time. 

\pagebreak