# Chapter Five {-}

# Introduction

On August 19, 1867, _The New York Times_ published a letter sent to Ulysses S. Grant from General John Pope, who at the time commanded one of five military districts that had recently been established by Congress action in spite of a veto by President Johnson. In the letter, Pope respectfully but clearly warned Grant of the significant potential consequences the nation would suffer if the military failed to sufficiently see the task of Reconstruction through. Pope wrote, in part, "Now is the time and this is the opportunity to complete the work so that it will stand. Every dictate of wisdom and patriotism demands this work at our hands. If hastily or partially done, reconstruction will drag with it a train of evils to this country which can never be remedied"[@pope_reconstruction_1867]. 

That Pope's letter was published in the _New York Times_ to begin with, and that its contents expressed an urgent plea, is reflective both of the military's central role in the Reconstruction era and the strong opinions held by senior military leaders during what were undoubtedly some of the nation's darkest and most polarizing political days.^[The historical evidence does not directly indicate that Pope's letter was leaked, either by Johnson or Grant. But it is highly likely that either one of these officers or someone close to them leaked the letter. As I show in this chapter, this type of behavior was fairly common in the period leading up to Johnson's impeachment.] Indeed, the military and its leaders were not just observers, but rather firsthand participants in the events that ultimately led to Johnson's impeachment. Johnson first removed Pope and other military district commanders, and replaced them with commanders who shared his views on Reconstruction policies. Then, after being impeached early in 1868, Andrew Johnson would remain in office, holding on to the Presidency by a single vote. In the fall of 1868, the nation elected the Republican candidate, General Ulysses S. Grant, to the first of two presidential terms.

A little more than a decade later, however, the military no longer occupied nearly as central a role in the life of the nation. The combination of the end of Reconstruction, divisive domestic politics, and uncertainty regarding the proper role of the military at a time when few if any clear international threats existed resulted in a period of frustrating years, especially for the Army, during the late 1870s - early 1880s. So frustrating were these years that in one private letter written in 1877, General William Sherman, the commanding general of the Army from 1869-1883 and himself a major hero of the Civil War, wrote, "Our enemy is not now the Indians but Congress" [Sherman, 1877 as cited in @marszalek_sherman_2007, 432]. 

This chapter identifies, compares, and contrasts the patterns of political behavior engaged in by military and civilian actors from the end of the American Civil War in 1865 to roughly 15 years later, in 1880. This chapter argues that during the post-US Civil War era, the character and severity of the political behaviors involving the military undertaken by civilian and military actors changed significantly. Early in this period, military and civilian actors both undertook highly visible and severe forms of political behavior that sought to leverage the military's political clout. Towards the end of the period, however, civilian and military leaders reacted to the military's diminishing prestige and decreased political influence by engaging in fewer and less severe forms of political behavior involving the military. The impeachment of Andrew Johnson, and the Army's closeness to the events leading up to it, was undoubtedly an atypical event. Yet this period clearly demonstrates, in my view like none other in American history to date, how high levels of polarization and military prestige can combine to generate a toxic political atmosphere with respect to civil-military conduct. 

The chapter proceeds in five parts. First, I briefly describe how we should think of the post-Civil War era in terms of the central principles of civil-military relations. Early in chapter one, I derived the three central principles of civil-military relations. In doing so, however, I noted that these principles have their root in the post-World War Two Era. Huntington and those who criticized him wrote in the 1950s and afterwards, yet the variables of political polarization and military prestige have operated long before the 1950s. This case, perhaps more than any other, shows how each of these variables work in the ways that the theory predicts. The second part of the chapter lays out the predictions regarding the patterns of political behavior that should be evident in this case based on the theory set forth in chapter two of this dissertation. 

The third and fourth parts of this chapter comprise the heart of this case study. The third part explores the political behaviors engaged in by civilian and military actors during the earlier portion of this case study, from approximately 1865 - 1868, while the fourth does the same for a latter portion from approximately 1877 - 1880. These two opposite ends of the case study align with respective periods in which military prestige existed at incredibly high and relatively low levels, respectively. The fifth and final part of this chapter evaluates and compares the differences in the patterns of political behaviors that occurred during these two eras. The main conclusion reached in this chapter is that the decline in the level of military prestige between 1868 and 1877 led to civilian and military actors engaging in less severe forms of political behavior involving the military. 

# Normative Political Principles and the Military in the Late 19th Century

The three central principles of civil-military relations, which were defined in chapter one as the principles of civilian control, non-partisanship of the military, and the non-interference of the military institution, did not exist qua firmly entrenched principles in the post-Civil War era, or for that matter, in any period before the post-World War Two era. These central principles, I pointed out in chapter one, derive mainly from post-World War Two scholars like Huntington and his critics, as well as military officers and statesmen, all of whom theorize about how a state's coercive and politically influential armed forces can secure the United States throughout the Cold War era while simultaneously obeying its civilian masters. During the late 19th Century and for a good portion of the early 20th Century, the United States was not a global superpower. This matters insofar as the military's latent political influence was not always as high as it has been throughout the post-World War Two era. Indeed, as this chapter shows, civilian leaders have not always had to contend with a military that possesses significant political capital.    

However, it would be wrong to state that there were no normative expectations regarding military and civilian leader interaction before World War Two and during the mid-late 19th Century in particular. The three central principles existed, but in nascent form. William Skelton, a prominent US military historian, notes that by the middle decades of the 1800s and well before the outbreak of the Civil War, "professional [military] ideology rejected partisanship and discouraged taking sides on purely civilian matters. The professed ideal was an apolitical officer corps and a rigid separation between the civilian and military spheres" [@skelton_american_1992, 283]. As early as the 1830s, and largely the result of the War of 1812, Skelton notes that, "though seldom expressed as a formal theory, officers developed a conception of the army as an apolitical instrument of public policy. As servants of the nation, they should stand aloof from party and sectional strife and avoid taking sides on civilian political issues" [@skelton_american_1992, 283].

Military leaders also expected a degree of professional autonomy free from civilian meddling, and resented being dragged into partisan politics. This sentiment is clearly visible throughout this case study, and shines through very clearly, for instance, in a private letter written by General Ulysses S. Grant to General William Sherman in which Grant explains his view on having his views publicized. Grant told Sherman that "military men had no objection to the publication of their views as expressed upon official matters properly brought before them, but that they did not like expressions of theirs which are calculated to array them on one or other side of antagonistic political parties to be brought before the public. That such a course would make or was calculated to make a whole party array itself in opposition to the officer and would weaken his influence for good" [Grant, 1866 quoted in @badeau_grant_1887,  40-41]. More than 150 years ago, military leaders strongly resented overt civilian efforts that politicized the military. 

For their part, civilian leaders expected that the military and its officers would obey orders and not obstruct their lawful authority. 19th and early 20th Century Presidents well understood that they were endowed by the Constitution as the Commander in Chief of the Armed Forces, even as the Presidency itself as an institution was not as powerful as it is in the contemporary era [@white_republican_1958, 20-44]. This sentiment also shines clearly throughout this case study. 

For instance, when President Andrew Johnson relieved General Philip Sheridan as the Commander of the 5th Military District in August of 1867, the President grew increasingly frustrated with Grant for what the President deemed to be insubordinate behavior. Grant had written the President a private letter in which he objected to the removal of Sheridan, but Grant's letter also stated that the general "should be consulted" as to the details of future Presidential orders [Grant, 1867 as cited in @sioussat_notes_1913, 113]. The day after receiving the letter, Johnson summoned Grant, and the President told the General that his letter had "amounted to a sort of political essay," and that "if every order he gave was to provoke a political essay from the General, it would be impossible for the Executive and the head of the War Department to work together" [Johnson, 1867 as quoted in @sioussat_notes_1913, 113]. The President then told Grant that in the event the President leaked Grant's letter to the press, the President thought that the public would side against Grant, on the grounds that the general's behavior clearly bucked well understood norms of obedience at the time [@sioussat_notes_1913, 113]. Grant then asked to formally withdraw the letter, and Johnson agreed. 

Tables \@ref(tab:pol-behav) and \@ref(tab:pol-behav-civ) in chapter two identify several potential political behaviors that may be undertaken by military and civilian actors. All of these forms of behaviors occurred in the 19th Century, and we should evaluate the occurrence of these behaviors in much the same way as when they occur in the modern era, with perhaps a few exceptions. Leaking information, giving interviews, writing op-eds, and otherwise making public appeals that openly disparaged either civilians or military actors by the other side was not considered an appropriate or acceptable behavior, as both sides sought to sustain their reputation with the public in an era when printed newspapers and other written outlets carried the headlines to America. Military officers acting explicitly insubordinate, or selectively complying with orders issued by their civilian bosses, was understood to be normatively wrong. Civilians politicizing the military by thrusting the military into the political spotlight for expressly partisan purposes was likewise understood to be normatively inappropriate. 

The one behavior that deserves a bit of further discussion is that of military officers running for office, and in particular, running for President. Norms regarding this type of behavior in the 19th Century were simply not as developed as they are in the contemporary era. For instance, General Winfield Scott, a hero of the Mexican War, actively campaigned for President ahead of the 1852 election while in uniform, an action that modern day Americans would find reprehensible [@skelton_american_1992, 286-287]. The contemporary and late 20th Century American generals and admirals who have run for office - Dwight Eisenhower, Douglas MacArthur, and Wesley Clark - at least waited until their military service was complete before formally announcing their candidacy for office, even if their political ambitions were suspected sooner. 

In summary, the three central principles of civilian control, the non-partisanship of the military institution, and the non-interference of the military were not as firmly developed in the mid-late 19th Century as they would come to be 100 years later and beyond. In this way, we should acknowledge that the individuals examined in this chapter were not guilty of violating principles which had been well worn for decades upon decades. Yet even in nascent form, civilian and military actors both understood and believed in the underlying spirit behind each of the three central principles, with few exceptions. Therefore, in the same way undertaken in the previous two empirical chapters, this case study examines not only what types of political behaviors occurred, but what practical effect they had in terms of challenging civilian supremacy (principle of civilian control), aligning the military with clearly partisan entities (principle of non-partisanship), and using the military to address primarily civilian matters of governance (principle of non-interference). Most importantly, such an analysis clearly demonstrates the role of polarization and military prestige in generating these behaviors. 

# Theoretical Predictions for the Post US Civil War Era, 1865-1881

This section summarizes the predictions that stem from the theory laid out in chapter two of this dissertation. I begin with describing the political behaviors predicted to occur in a domestic environment characterized by high polarization and high prestige, and then describe those predicted to occur in a setting marked by high polarization and lower military prestige. 

## High Polarization and High Military Prestige

According to the theory laid out in this dissertation, significant ideological divisions will exist among military actors when polarization is high. This is because, as the theory argued, military actors become polarized as society becomes polarized. The theory further predicts that high levels of polarization will provide a "motive" for military officers to intervene in politics. This stems from the idea that high levels of polarization point to ideological divisions not just about politics, but about moral issues and values that are at stake in the political sphere. This explains the fact that on the eve of Civil War, the US Army's officer corps splintered, with officers such as Robert E. Lee, who President Lincoln asked to command the Army, resigning his commission and choosing to fight for the Confederacy [@freeman_lee_1993, 110-111]. 

The theory also predicts that as military actors encounter increasingly polarized environments, they increasingly encounter choices and decisions in which they must choose either to adhere to the principles of civil-military relations, or to behave in ways that violate these principles, but are aligned with the moral values and philosophies with which they identify. As polarization worsens, then, a greater number of military actors will choose moral principle over civil-military principle, and thus, increasing polarization leads to military actors increasingly engaging in behaviors that violate one or more of the principles of civil-military relations. This case study will show that even after the Civil War and even among officers who had fought together to preserve the Union, high polarization contributed to these officers adopting very different views on - and thus engaging in different behaviors related to - Reconstruction policies. Simply put, high polarization erodes the political basis of unity because the underlying moral foundation of that unity is no longer sure.   

When the level of military prestige is also high, however, the theory predicts that civilian and military actors both believe that military actors can potentially influence public opinion. Chapter two rooted this expectation in the political communication literature, a portion of which argues that the public looks to voices who are considered to be "knowledgeable" and "trustworthy" before making decisions [@lupia_democratic_1998, 69-76; see also @zaller_nature_1992, 47]. It is reasonable, the theory argued, to assume that as the level of military prestige rises, the public will view the military and its leaders as increasingly "knowledgeable" and "trustworthy." 

The sociological literature of professions and occupations states that two components, centrality and performance, shape the overall level of prestige commanded by the members of a profession [@davis_conceptual_1942; @jaco_prestige_1970]. The two previous empirical chapters examined empirical behaviors in the post-Cold War era, when the centrality of the military to the United States is relatively high and constant. Variation in military prestige in these two previous chapters stemmed from variation primarily in society's evaluation of the military's skill and performance. In this chapter, the source of the variation in military prestige involves the other component involving how central or important the military is viewed by the nation of which it is a part. 

Regardless of which components drive variation in military prestige, when prestige is high, the theory predicts that both civilian and military actors will engage in behaviors that are intended to sway or influence public opinion, even when these behaviors violate accepted norms of principles of civil-military relations. Furthermore, the theory predicts that during periods of high military prestige, the behaviors that civilian and military actors commit will be relatively more visible to the public - in an attempt to influence public opinion - than in periods of low military prestige. 

When we integrate simultaneously high polarization and military prestige, the theory predicts that civilian and military actors will engage in what can be considered a comprehensive form of military politicization. This means that civilian and military actors will undertake a rather extreme level of political activity when polarization and military prestige exist at high levels. 

```{r poss-environment2, echo=FALSE, include=TRUE, error=FALSE, message=FALSE}

text_env_revised <- tribble(~"Environment", ~"Low Polarization", ~"High Polarization",
  "Low Prestige", "No politicization; low levels of overall political activity.", 
  "Military (induced) politicization; moderate levels of overall political activity.",  
  "High Prestige", "Military and Civilian (induced) politicization; moderate levels of overall political activity.", 
  "Comprehensive politicization; extreme levels of overall political activity.  " 
  )
 

  kable(text_env_revised, booktabs=T, caption="Types of Domestic Environments for Civil-Military Relations and their Impact on Political Behaviors") %>%
  kable_styling(full_width = F, font_size=10, latex_options="striped") %>% 
  column_spec(1, width="8em") %>% 
  column_spec(2:3, width="18em") 

```
## High Polarization and Low(er) Military Prestige

If the level of polarization remains high but the level of military prestige declines, the most significant theoretical prediction is that the "opportunity" for civilian and military actors to involve the military in politics diminishes. The theory suggests that declining military prestige is associated with a reduction in the perceived potential of the military to influence public opinion, and therefore, that civilian and military actors will become less likely to undertake political behaviors that involve the military in politics in highly public ways. Nonetheless, because polarization remains high, the theory predicts that military actors maintain a "motive" to intervene in politics. In a domestic environment categorized by "motive" without "opportunity," the theory predicts that military actors will be the primary type of actor inducing political activity, and that they will engage in less severe forms of political behavior than when military prestige is high. Moreover, in an environment marked by high polarization and low military prestige, the theory predicts that civilian actors will engage in far fewer (and far less overt) political behaviors involving the military.

Having laid out these theoretical predictions, this chapter now turns to examining the political behaviors undertaken by civilian and military actors in the years immediately following the end of the Civil War. 

# Johnson, Grant, Congress, and the Army, 1865 - 1868

On numerous occasions from 1865-1868, civilian actors sought to leverage the military's prestige for their own partisan gain. President Andrew Johnson as well as his opponents both attempted to pull Grant and other highly prestigious military actors into their respective political camps, and to portray to the American public that certain military leaders were indeed their political allies. 

At the same time, military leaders, including Grant himself, engaged in several instances of political behavior that, even by historical standards, challenged, if not violated, the normative standards of civil-military conduct that existed at the time. These military leaders did so, I argue, because of deep moral convictions related to significant political issues that were at stake during the period of Reconstruction. Key military leaders who were heroes of the Civil War were also very aware of the enormous political influence they commanded with the public at the time. 

In this front section of the case study, I first set the scene by establishing that the levels of polarization and military prestige were both extremely high at end of the US Civil War in 1865. I then describe the political behaviors undertaken by civilian and military actors in the events leading to Johnson's impeachment and Grant's nomination to the Presidency. In so doing, I describe the roles of polarization and prestige in contributing to these behaviors. 

Before proceeding, it is important to note that in the historiography of this period, clear biases are present. Some historians, such as Milton (1930) are obviously sympathetic towards Andrew Johnson, and thus very critical of Grant. Others, such as Badeau (1887), White (2016) and Chernow (2017) are extremely critical towards Johnson, and thus ebullient in their praise of Grant. Still others, in my view, including Hesseltine (1935) and Simpson (1988) seem to fall somewhere in the middle. Throughout this chapter, I seek to remain as neutral as I possibly can, and to objectively present the events that unfolded and why. 

## High Level of Polarization, 1865-1868

It is difficult to state the precise level of polarization that prevailed at the end of the Civil War, but there is no doubt that the level was significantly high. The fact that a Civil War - the only one in US history - had just occurred, primarily over the issue of slavery, provides an immediate basis for this assertion. Standard measurements of polarization in Congress, however, usually begin with the 45th Congress, which met from 1877-1879 [@jeffrey_b_lewis_voteview_2020]. This is again because of the Civil War and its aftermath. When the first state, South Carolina, seceded from the Union in December 1860, until the Compromise of 1877 effectively ended Reconstruction, several southern states either lacked Congressional representation altogether or had been prevented from electing members of Congress on their own, absent of federal interference.^[The Compromise of 1877 was, in essence, a compromise about the 1876 Presidential Election of 1876. Democrats more or less agreed to a compromise such that the Republican candidate, Rutherford B. Hayes, would become the President, while Republicans in Congress agreed that the last of federal troops would depart Louisiana and South Carolina, effectively ending Reconstruction. See @hoogenboom_rutherford_1995, 274-294 for an account of this compromise by one of Hayes' biographers.] Still, the lack of a precise measure of polarization until the late 1870s need not leave the reader wondering whether polarization was high at the end of the Civil War in 1865.  

Polarization was high at the end of the Civil War as the country sought to "bind up the nation's wounds" [@lincoln_lincolns_1865]. To begin with, the nation had to face and answer questions concerning how best to bestow new political freedoms, to include the right to vote and own land, to more than four million former slaves [@green_reconstructing_2010, 186-193].^[For just a brief sampling of the rich literature that describes the enormous political questions facing the nation at the end of the Civil War, see @mckitrick_andrew_1988; @green_reconstructing_2010; @sheehan-dean_northern_2014; @smith_stormy_2017; @sheehan-dean_northern_2019.] 

The US government also faced other significant legal and moral questions pertaining to how to handle and relate to the states that had seceded. Indeed, McKitrick (1988) argues that Reconstruction, at its heart, boiled down to two central questions: "what had the war done to the rebellious states? And what was the new relation of those states - or districts - to the federal government" [@mckitrick_andrew_1988, 96]? While this chapter cannot unpack all of the differing views on these questions, it is worth summarizing a few of the opinions held by prominent civilian leaders at the time to illustrate how widely opinions varied on these topics. 

Some politicians, such as Thaddeus Stevens, a Pennsylvania Congressman who bitterly opposed President Johnson and who was known as a Republican 'Radical,' believed that the southern states should be administered as "conquered provinces," much in the same way a victor in war might exercise rule over a formerly hostile foreign power and its inhabitants [@mckitrick_andrew_1988, 99-101]. Stevens and others who held to this view believed that the former Confederate states had forfeited any of the legal rights that they had once possessed under the Constitution, that the entire way of life in the south would have to be forcefully remade, and perhaps most controversially, that private property belonging to southerners should be seized, confiscated, and redistributed [@mckitrick_andrew_1988, 99-101]. 

On the other end of the spectrum of views regarding reconstruction is what McKitrick (1988) calls the "southern theory" of Reconstruction [@mckitrick_andrew_1988, 97-98]. This view advocated for a relatively quick restoration of full rights to the former Confederate states on the basis that these states had tried to secede, but had failed, and thus should be welcomed back into the Union with little or no hindrance [@mckitrick_andrew_1988, 97-98]. Now that the war was over, advocate of this view argued that there was little need for punitive action.^[McKitrick (1988) points out that it was not only those from the south who held this view. Among the most ardent supporters of this "southern theory" of reconstruction was none other than General William T. Sherman, who, after Lee's surrender at Appomattox, infamously offered generous peace terms to Confederate General Albert Sidney Johnston. These terms of peace included a general pronouncement of amnesty for former Confederate officers. But Sherman quickly encountered political turmoil after offering these terms. When General Grant, as Sherman's superior, learned what had happened, he intervened and revoked the terms. Grant understood that Sherman, by brokering terms of political restoration, had not only crossed into a realm that only statesmen should occupy, but had offered terms that were politically unpalatable at the time. See @mckitrick_andrew_1988, 98; and @simpson_ulysses_1988, 95-100; for more details.] 

In short, each of the different views regarding how the south should be reconstructed involved a host of significant moral and practical beliefs about what actions the federal government should take with respect to the southern states and its inhabitants, including both former confederates and former slaves. Should former confederate officers and soldiers be granted amnesty for their past actions in taking up arms against the union? Should such persons be allowed to hold political office, and if so, would they have to denounce their past behavior or declare loyalty to the federal government? How much help should the federal government provide, and how quickly should they provide it, to ensure the implementation of new freedoms to former slaves?^[See @mckitrick_andrew_1988, 120-152 and @green_reconstructing_2010, 186-194 for a concise treatment of these issues]. 

Adding to the weight of these significant questions was another crucial issue, which was _who_ was ultimately responsible for overseeing and implementing the nation's reunification. Was the President ultimately responsible, or should Congress set the terms for welcoming the recalcitrant southern states back into the Union after four years of Civil War [@mckitrick_andrew_1988, 103-109]? Although President Lincoln had claimed significant executive powers, some of which rested on perhaps questionable invocations of wartime necessity, Republicans in Congress were not willing to grant Lincoln's successor the same benefit of the doubt, especially in the wake of Lincoln's assassination and because the war, at least in terms of open battles fought, had formally ended [@mckitrick_andrew_1988, 106-107; @hearn_impeachment_2000, 29-32].  

The Army and its senior leaders came to adopt strong opinions about these and related issues as the Army became a main actor involved with Reconstruction. Indeed, the Army came to embark on what should be considered nothing short of an occupation of the American south that would last more than a decade. For much of this unprecedented period, Army forces in the south and their commanders took on incredible responsibilities for overseeing civilian government, administering martial law, and exercising authorities similar to those that exist in wartime [@downs_after_2015; @lang_wake_2017]. 

Like the public, however, military leaders were divided about how Reconstruction should proceed. Some leaders, including Generals Ulysses S. Grant and Phil Sheridan, came to adopt views that were consistent with Republicans in Congress, favoring strong involvement by the Army and the federal government. Other military figures, including Generals William Sherman and Winfield Scott Hancock, held views that were similar to the President, and favored states' rights, and thus a more limited role for the Army in the South [@milton_age_1930; @hesseltine_ulysses_1935; @jordan_winfield_1988;@simpson_let_1991; @marszalek_sherman_2007; @white_american_2016; @chernow_grant_2017]. Many times, the viewpoints that military officers held about Reconstruction served as "motives" for these officers to engage in political behaviors that challenged, and sometimes violated, normative standards of civil-military conduct. 

## Extremely High Level of Military Prestige, 1865-1868

The level of military prestige was also high at the end of the Civil War, and for one central reason: the Army had achieved victory in preventing the dissolution of the nation. This was especially true of many general and admirals who had achieved nationwide fame as a result of their martial exploits.  

While it is not possible to list all of the individuals who achieved heroic stature as a result of their martial exploits during the Civil War, it is worth pointing out several, who, in the years after the war, remained in the Army. Among these included General Philip Sheridan, the famed cavalry officer; General William Sherman, known for his ruthless march through Georgia near the close of the war; General Winfield Scott Hancock, who repelled the Confederate charge at Cemetery Ridge at the battle of Gettysburg; and many, many others.^[These include figures such as John M. Schofield, who would play a significant role in the Army's reform efforts late in the 19th Century.] However, while these officers were well known and popular figures, the prestige that accompanied General Ulysses S. Grant, who became the Army's commanding general at the close of the war, was second to none. 

Grant was the last of Lincoln's commanding generals of the U.S. Army and the military leader who ultimately brought about the defeat of Lee's Army. A West Point graduate and veteran of the Mexican War, Grant had resigned his officer commission in 1854 as a Captain. However, after several years in failed private business ventures, Grant was recommissioned as a Colonel in the Illinois Militia when the Civil War broke out in 1861, and he quickly established himself as an able fighter and commander [@white_american_2016].

So popular and so recognized by the American public was Grant at the end of the Civil War that Hesseltine (1935) argues Grant "was rapidly on his way to becoming a saint" [@hesseltine_ulysses_1935, 47]. Furthermore, multiple biographers of Grant recount that the general spent several weeks during the late summer of 1865 on a traveling tour of the northern states, where he was fervently met at every stop, and often regaled with large and lavish receptions, parties, and other gifts [@hesseltine_ulysses_1935, 57-62; @chernow_grant_2017, 558-560]. A significant portion of the northern press was also infatuated with Grant. For example, a _New York Daily Tribune_ article written on April 20, 1865, said the following of the general: "Three captains created the art of war as it exists today in Europe - Marlborough, Frederick, and Napoleon. Grant's last campaign entitles him to rank with either of the two former, and in several important elements of military greatness places him on a level even with the great French Captain" [@noauthor_grants_1865]. The potential that Grant held to shape public opinion was so great that Adam Badeau, one of Grant's subordinate staff officers who spent many years observing Grant as an eyewitness to the events written about in this chapter, would write in 1887, that "It is impossible to understand either Johnson's forbearance or Grant's authority all through this epoch without bearing constantly in mind that Grant was the most popular man in America" [@badeau_grant_1887, 68-69].

Grant's significant popularity translated into political capital, and many civilians understood this reality. Horace White, the editor for the _Chicago Tribune_, for instance, wrote to Republican Congressman Elihu Washburn at the end of 1865 that, "I am persuaded that Grant is to be the next President if he gives us half a chance to nominate him" [White quoted in @hesseltine_ulysses_1935, 63]. Chernow (2017) goes so far as saying that in the years after the Civil War, Grant "became the supreme prize in American politics, sought by all parties" [@chernow_grant_2017, 597]. With his headquarters in Washington D.C., Ulysses S. Grant found himself squarely in the middle of high stakes political debates about how Reconstruction should proceed, and what policies Reconstruction should entail. Later in this chapter, it will become clear that the same was also true of many of Grant's subordinate officers, many of whom were also highly prestigious and popular figures.  

For the first several weeks, and perhaps months, the relationship between Grant and President Johnson was relatively free of friction, even, as Hesseltine describes, a "political cauldron" was boiling all around [@hesseltine_ulysses_1935, 54]. Over time, however, Grant and several of his subordinate commanders would no longer be able, nor desire, to remain outside of politics. 

In the next section, I narrate and describe the political behaviors engaged in by civilian and military leaders from roughly May 1865 - May 1868 that, even by historical standards, violated the principles of civil-military conduct. The levels of polarization and military prestige were both extremely high, and combined in such a way so as to generate extreme and frequent political behavior by both civilian and military leaders. I describe these behaviors generally in chronological order in the following section.

## How Military Prestige Shaped Political Behaviors Involving the Military, 1865-1868

As divisions over reconstruction became evident in 1865 between President Johnson and his opponents in Congress, both entities sought to bring the military into their respective camps. Johnson and members of his administration, as well as members of Congress, sought General Grant's approval on the specific measures that Reconstruction should entail. Indeed, in the months that followed, as the divisions between Johnson and Congress became increasingly irreconcilable, both Johnson and his opponents in Congress would increasingly wonder, as White (2016) notes, "where did Grant stand" [@white_american_2016, 429]? Badeau observed that as early as 1865, "both parties to the contest wanted to use the prestige of his (Grant's) name; both laid their arguments before him and sought to secure his support" [@badeau_grant_1887, 35, parentheses mine].

That civilian leaders of both political parties maneuvered and jockeyed for Grant fully aligns with the predictions made by the theory advanced in this dissertation. Civilian leaders acted from the belief that General Grant, as the nation's most prestigious figure, commanded not merely the Army, but the potential to shape and influence public opinion in a way that was politically advantageous. Adam Badeau's observations of the President's early overtures towards Grant cleary signify the link between Grant and his utility to civilian leaders to shape public perception. Reflecting back on the first few months of Grant's professional relationship with the President, Badeau notes that "the great popularity of Grant at this period made it important to win him over to the support of the enterprise...It was, however, not so much Grant's real concurrence as the appearance of it before the world that Johnson probably sought" [@badeau_grant_1887, 35, 39]. Grant personally understood that political parties saw him as politically powerful and useful. He wrote his friend, General William Sherman, in 1867 that, "...there is little difference between parties. No matter how close I keep my tongue each tries to interpret from the little I let drop that I am with them" [Ulysses Grant, 1867, cited in @badeau_grant_1887, 73]".

### 1865 - Fall of 1866

One of the first political behaviors the President engaged in was sending Grant on a tour of the South in late 1865. The purpose of sending Grant to the South was to refute a previous report made by Carl Schurz, a journalist and former general whom Johnson had previously sent to assess the general political atmosphere of the southern states. Schurz's report was politically problematic for Johnson, however, because it portrayed southern attitudes as "angry and defiant," and indicated that former slaves who had been freed were "languishing in wretched conditions of poverty, reinforced by Black Codes that trapped them in a new subservience" [@chernow_grant_2017, 563]. In short, Schurz's report did not portray a south that was remorseful and prepared to rapidly reunify with the nation, but rather one that required a lengthier and more robust federal presence [@chernow_grant_2017, 563]. 

Johnson hoped that Grant, after touring the south, would write a report that instead described the political conditions there as conducive to a relatively quick restoration, in line with Johnson's views on Reconstruction [@hesseltine_ulysses_1935, 60; see also @chernow_grant_2017, 563-564]. Such a report made by a highly influential and prestigious military figure like Grant would arm the President with a political weapon to employ against his opponents in Congress. After Grant returned from his visit, Johnson summoned Grant to a cabinet meeting where the general "expressed his conviction that every consideration called for the early reestablishment of the Union [@hesseltine_ulysses_1935, 61]. Grant's comments were immediately sent off to the Senate, and used by Johnson's administration to defend against attacks made by Radical Republicans [@hesseltine_ulysses_1935, 61; @chernow_grant_2017, 565-566].^[Chernow (2017) argues that Grant's visit was superficial in that it was rushed and largely featured sham interactions with southern whites who praised Grant. Chernow also notes that Grant regretted his report soon after furnishing it. See @chernow_grant_2017, 565-566 for more details. Furthermore, both Hesseltine (1935) and Milton (1930) argue that in the wake of Grant furnishing his report, Radical Republicans fervently tried to get Grant to denounce it. See @hesseltine_ulysses_1935, 65; @milton_age_1930, 290 for more details.] 

In sending Grant to the southern states, the President challenged at least two of the normative principles of civil-military conduct that existed at the time. First, Johnson's behavior violated the principle of non-partisanship because there was effectively no possible way that Grant could generate a report that did not anger either Johnson's supporters or his opponents in Congress. Furthermore, Johnson's chief purpose in sending Grant was to provide both the American public and Congress the impression that Grant's views on Reconstruction favored his own. Johnson knew that validation of his views by Grant would disadvantage his opponents in Congress, if even for a short period of time.  

Second, Johnson's dispatching of Grant to the south for the purposes of Grant writing a subsequent public report also violated the principle of non-interference because, in effect, Johnson made a sitting military commander investigate, evaluate, and publicize prevailing political conditions in ways that were more comparable to what a reporter or journalist, rather than a military officer, would do. There is nothing normatively inappropriate for a President to send a military officer to learn more about a situation, or to report back with intelligence on a number of possible scenarios. But Grant did not possess any clearly superior reporting or communication skills than the author of the report for which Grant was sent to refute, Carl Schurz. If anything, it is almost certain that Grant's popularity prevented him from accessing people during the trip who would likely have conveyed to Grant far less rosy of a situation  [@hesseltine_ulysses_1935, 60]. 

Schurz was a career journalist who had volunteered to serve in the Union Army, and would become the editor of the _Detroit Post_ in 1866. This matters because Schurz's initial report was written from the standpoint of a professional reporter or journalist [@simpson_let_1991, 116-188]. Johnson could have dispatched others to refute Schurz's report, including perhaps other press correspondents who likened to his views. Instead, in sending Grant, Johnson sought to harness Grant's enormous prestige, knowing that the public would be interested, and that Congressional Republicans would have to contend with, Grant's evaluation. Still, In asking Grant to fill a role that was better suited for a press correspondent or journalist, President Johnson compelled the nation's top Army officer to assume a role that was tangential, if not altogether unrelated, to his role as the general-in-chief of the Army. 

After Grant's report became public knowledge, the final months of 1865 and early months of 1866 evinced even greater divisions between Johnson and his Republican opponents in Congress. As time passed, it became increasingly clear that Johnson's conception of Reconstruction clearly favored southern whites with ties to the former confederacy, and at the same time, disadvantaged both former slaves and whites who had supported the union. For instance, newly freed blacks frequently faced increasing danger as southern state and local governments enacted a series of "black codes" that authorized a variety of discriminatory measures against blacks and former slaves [see for example, @mckitrick_andrew_1988, 204-205; @chernow_grant_2017, 571]. 

Though the 13th Amendment (which formally abolished slavery in the United States) went into effect in December of 1865, Johnson continually vetoed legislation that sought to expand the rights and political freedoms of the former slaves. These included bills which sought to extend the life of the Freedmen's Bureau, an organization created in 1865 to "provide food, shelter, clothing, medical services, and land to displaced southerners, including newly freed African Americans" [@noauthor_us_nodate], and later, a civil rights bill which, while "silent on voting rights," nonetheless sought to make the federal government the "guarantor of basic liberties" [@chernow_grant_2017, 570]. Both of these pieces of legislation would eventually end up passing as Congress voted to override vetoes by Johnson.

With these legislative battles between Congress and the President unfolding, Johnson engaged in another obvious political behavior involving the military ahead of the upcoming midterm elections of 1866. Despite the already noteworthy shortcomings of Johnson's Reconstruction policies up until that time, Johnson sought to show the nation that his approach towards the south was the right one, and that Grant, his prestigious military commander, thought so too. Though Johnson was not up for reelection in 1866, conservative Republicans and moderate Democrats - who favored Johnson - were. In a very real way, then, the midterm elections were a referendum on Johnson's policies in the south.  

### Need instance of other leaders here----

Johnson and other civilian leaders attempted to harness General Grant's prestige and that of other military leaders for partisan electoral gain in the months leading up to the elections, in ways both large and small. As an example of a relatively small political behavior, on several occasions, civilian actors deliberately orchestrated their public appearances in such a way so as to physically appear alongside popular military figures in order to give the impression that these military leaders approved of their policies. For instance, on August 18, 1866, President Johnson hosted "a delegation from the National Union Convention in Philadelphia" at the White House [@chernow_grant_2017, 576]. In spite of the clearly partisan nature of the gathering, Grant reluctantly accepted Johnson's invitation to attend, as he "felt duty-bound" to do so [@chernow_grant_2017, 576]. Before addressing his supporters in the East Room of the White House, the President summoned Grant, where the two stood next to each other during all of the President's remarks [@chernow_grant_2017, 577]. After the event, Grant supposedly returned to his headquarters, where he was "full of indignation" at having been manipulated by the President [Badeau, 1877, as cited in @chernow_grant_2017, 577]. Ultimately, Grant's physical proximity to the President on that particular August evening insinuated, at least to one reporter present at the event, that Grant fervently endorsed Johnson's policies [McFeely, 2002, as cited in @chernow_grant_2017, 577]. 

A far greater instance of Johnson and other civilian leaders harnessing Grant's prestige and that of other military leaders occurred at the end of the summer of 1866, however, when Johnson insisted that Grant and several other military figures accompany him on a two week long public relations tour of several Northern states and populous cities that would be known as the "Swing Around the Circle." Traveling by train and designed to drum up support ahead of the elections, the President brought with him several of his cabinet members, but also several military figures, including General Grant, Admiral David Farragut, and General George Meade. Much the same way the President had done at the White House earlier in August, appearing alongside popular military figures in overtly partisan settings intended to convey to the public that these military figures approved of Johnson's political policies, which included a strong denunciation of the 14th Amendment, which had been passed by the Senate in June [@hesseltine_ulysses_1935, 74-76; @chernow_grant_2017, 577-581].^[The 14th Amendment granted citizenship to all persons born in the United States, including former slaves. Except for Tennessee, the southern states rejected the 14th Amendment initially, but the Military Restoration Acts of 1867 later required ratification of both the 13th and 14th Amendments for states to be restored to the Union.]

By almost all accounts, the President's trip turned out to be a public relations disaster. Johnson often argued with the crowds at almost every stop, and many of those who turned out asked instead to hear from Grant and Farragut rather than from the President [@hesseltine_ulysses_1935, 75; @chernow_grant_2017, 577]. Interestingly, Milton (1930, pp. 362-369), Hesseltine (1935, pp. 74-75), and Simpson (1991, p. 148) argue that the crowds who cheered for Grant and Farragut and those who heckled Johnson did so as a result of deliberate orchestrations devised by Radical Republicans in Congress and their allies in the press, who, upon learning that several well known military figures would accompany Johnson on his speaking tour, devised a plan to prevent Johnson from capitalizing on the prestige these military figures brought. If these explanations are valid, even in part, then it is clear that not only Johnson, but also Johnson's opponents, acted in ways that sought to appropriate the prestige of military figures for their own partisan advantages. In essence, Johnson and his opponents competed, at times in ways that were not subtle at all, for the prestige of particular military figures, and all did so in order to influence public opinion in ways that would be advantageous ahead of the upcoming elections.^[Some historians point out that the "Swing Around the Circle" likely marked somewhat of an internal turning point for Grant in terms of his convictions regarding his conduct vis-a-vis the central principles of civil-military relations. Hesseltine (1935, p. 75), for instance, argues that in the trip's aftermath, Grant drew conclusions about how unpopular the President and his policies were across the country, and that from that point forward, Grant tried less hard to avoid confrontation with the President.]

### Late 1866 - Johnson's Impeachment in early 1868

Throughout this period, the nation's political climate continued to worsen. Significant race riots in 1866, first in Memphis and then in New Orleans, had killed dozens of freed blacks and sympathetic whites [@mckitrick_andrew_1988, 422-427; @hyman_santon_1960, 90]. Events such as these further cemented partisan battle lines, even among Army leaders: along with the Secretary of War, Edwin Stanton, Generals Grant, Sheridan, Pope, and Meigs favored a theory of Reconstruction that was congruent with most Republicans in Congress, while Generals Sherman and Hancock favored Johnson's [@hyman_santon_1960, 93]. 

Congress and the President continued to battle for control over Reconstruction, which necessarily involved jockeying for control of the Army. Even after the midterm elections, the President continued to engage in several blatant political behaviors that violated the principles of civil-military relations. The most obvious of these involved efforts to either minimize the influence of or outright replace key military commanders on the basis of partisan convictions. 

Johnson first tried to break up the combined team of Grant and Stanton, who by early 1867 had both come to believe that Congressional Republicans, rather than the President, possessed a view of Reconstruction that was best for the nation and for the Army [@hyman_santon_1960]. Too prestigious to dismiss outright, Johnson instead sought to send Grant to Mexico as part of a diplomatic mission, and to simultaneously replace Stanton with General Sherman, who Hyman (1960) argues was the "nation's third most popular man" at the time [@hyman_santon_1960, 94]. Johnson apparently even considered making the move in conjunction with leaking a private letter he had received from Sherman in which the general had expressed support for the President's policies [@hesseltine_ulysses_1935, 78]. 

Johnson's scheme ultimately floundered, however, and neither Grant nor Stanton (this time) were replaced. Though Sherman was sympathetic to Johnson's views, Sherman could not bring himself to be part of an obvious political scheme. Grant, on the other hand, simply refused Johnson's directive, and in a rather dramatic manner. In front his assembled cabinet, the President asked the Attorney General whether his orders to send Grant to Mexico were illegal, ostensibly in order to publicly shame Grant, who was also present for the meeting, into compliance. Hesseltine (1935) depicts Grant as standing up before the Attorney General could respond, stating, "I can answer that question Mr. President, without referring to the Attorney General. I am an officer of the army, and bound to obey your military orders. But this is a civil office, a purely diplomatic duty that you offer me, and I cannot be compelled to undertake itno power on earth can compel me to it" [Badeau, 1877 as cited in @hesseltine_ulysses_1935, 79]. 

Though Johnson decided not to press the issue further, the entire incident reveals how a high level of military prestige shaped the President's and Grant's behavior. To begin with, if replacing Stanton had even a remote chance to work, the President would have to choose someone who was highly regarded by the public. This explains Johnson's selection of Sherman. Though we can never know for sure, had Johnson replaced Stanton with Sherman, Johnson's opponents in Congress would have likely accepted Sherman or at least have been less capable of rendering opposition on the grounds that Sherman was a highly regarded military figure. 

Grant's refusal of Johnson's directive to go to Mexico is also instructive. Grant stated in effect that the order was illegal, and that he therefore did not have to follow it. But is this an accurate statement? Although Johnson's directive to send Grant to Mexico was clearly done in order for the President to obtain a political advantage in that it would remove Grant from interfering with Johnson's political wishes, I am not convinced that the order was itself explicitly illegal. It is more likely the case that in refusing Johnson's order, Grant knew that he could. There is no doubt that Grant wished this incident not to have transpired in the first place, Grant likely knew that the prestige he carried rendered him politically unassailable, as both Republicans in Congress and the public would back him.  

On March 2nd, 1867, Congress passed (again, over a Presidential veto) the first of several Military Reconstruction Acts. This significant legislation split the south into five military districts, each of which would be overseen by an army general. The legislation intended for these military commanders to hold enormous sway over the daily conduct of state and local governments, which were to be regarded by the commanders as "provisional" in nature pending their "full" restoration to the union [Grant, 1867 as cited in @chernow_grant_2017, 587]. Importantly, the legislation also stated that all orders to the Army from the President had to be routed through Grant [@chernow_grant_2017, 587].^[The Military Reconstruction Acts also empowered Grant to appoint these five generals himself, but even Grant had reservations about this, believing that the President, as the Commander in Chief, should have a strong say in who these generals were. See @hesseltine_ulysses_1935, 83 and @chernow_grant_2017, 585 for more details.] 

At the same time, Congress also enacted the Tenure of Office Act, which sought to preserve Stanton's place in Johnson's cabinet, and thus his link to Republicans in Congress [@hearn_impeachment_2000, 110-119; @chernow_grant_2017, 586-587]. This piece of legislation specifically declared that the President could not remove cabinet officials that he had not appointed himself without the Senate's approval [@chernow_grant_2017, 586].^[Chernow (2017, p. 586) notes that Grant personally felt the Tenure of Office Act was "ridiculous" in that it essentially rendered the President powerless, but that he nonetheless sought to obey the law that had been passed.] Indeed, Johnson had inherited Stanton, who had also served as Lincoln's Secretary of War. Together, the Military Reconstruction and the Tenure of Office Acts effectively cut off the President from exercising meaningful control of the Army in the south. Johnson was, in effect, no longer the Commander in Chief.  

The tension between the President and Congress continued to significantly escalate, and the Army remained in the middle of it all. In August 1867, the situation reached a boiling point. The President first informed Stanton that he was suspended (not "removed," so as to avoid being instantly in violation of the Tenure of Office Act), and proceeded to appoint Grant as the Secretary of War _ad interim_ until a permanent replacement might be named [@hyman_santon_1960, 98]. The President then removed General Philip Sheridan, who commanded the 5th Military District (Louisiana and Texas), after Sheridan removed the Governors of both Louisiana and Texas [@white_american_2016, 448-449; @hyman_santon_1960; 98-99]. Sheridan, a favored subordinate of Grant's who favored an aggressive Republican policy towards Reconstruction, had helped register thousands of blacks to vote in the district, earning widespread condemnation from Johnson and southern whites. Johnson also later removed the commanders of the 2nd and 3rd Military Districts, Generals Daniel Sickles and John Pope, for similar reasons.^[Johnson's replacement commanders were all known to be sympathetic to Johnson's views on Reconstruction, especially General Winfield Scott Hancock, who replaced Sheridan in the 5th Military District.]  

In early January of 1868, the situation involving Johnson, Grant, and the Army moved towards a climax. The Senate first released a report refusing to concur with Stanton's suspension, which in turn raised immediate questions as to whether Grant's serving as the _ad interim_ War Secretary was legal or not [@mckitrick_andrew_1988, 501-505]. Grant did not wish to find out the answer, and thus he attempted to vacate the office, which Johnson would not let him easily do [@mckitrick_andrew_1988, 501-502].

Exactly what transpired next is not so easy to discern, even for contemporary historians [@chernow_grant_2017, 602]. The critical question is what sort of mutual understanding, if any, Grant and Johnson had regarding Grant's continued service as the _ad interim_ War Secretary, and when such an understanding was reached. Those who are sympathetic towards Johnson allege that the two men agreed that Grant would either keep the office, even if the Senate or the Supreme Court would hold proceedings to decide the legality of Stanton's removal or that of the Tenure of Office Act altogether, or that Grant would vacate the office sufficiently ahead of time so as to enable Johnson to name a replacement, and thus, to not be found in violation of the law [@milton_age_1930, 481-485]. 

Other historians who are charitable towards Grant take a different view, and argue that Grant made it clear to the President that if Stanton were to be reinstated by the Senate, Grant would comply by returning the office to Stanton immediately [@chernow_grant_2017, 602-607]. Regardless of which view is historically accurate, it is clear that after a tense and heated cabinet meeting in January of 1868, Johnson's administration engaged in multiple efforts to discredit Grant in the press and to portray him as willfully acting in a "duplicitous" manner [@chernow_grant_2017, 605; @milton_age_1930, 483-485; @mckitrick_andrew_1988, 503; @hesseltine_ulysses_1935, 106-107]. For several weeks, the nation hung on press reports fueled by leaked letters, interviews, and other details provided by the Johnson Administration. 

The civilian acts of leaking information and giving interviews to the press in this instance violated the principles of civilian control and non-partisanship. They violated the principle of civilian control because they reflected the fact that by this point, the Administration was no longer attempting to exercise oversight of the military through any established mechanisms. Indeed, to be fair to Johnson and also historically accurate, many of these regular mechanisms had been removed by Congress through the Military Reconstruction Acts, which effectually made the Commander in Chief an observer of, rather than a participant of, the state's civil-military relationship. 

Leaking and giving interviews to the press to discredit Grant also violated the principle of non-partisanship, however. While it was well known that Grant differed with the Commander in Chief over Reconstruction policies by this point, the public nature of the leaks and interviews given by the Johnson Administration only served to publicize these differences. 

In summary, President Johnson and members of Congress undertook several significant behaviors from 1865 - 1868 to court approval with highly influential military actors and to portray the approval of these military figures to the public. These behaviors ranged considerably, from Johnson sending Grant to refute Carl Schurz's initial observations of the south in 1865, to Johnson's harnessing of several military actors on his "Swing Around the Circle Tour" ahead of the 1866 midterm elections. In other instances, military figures undermined the central principles of civil-military relations by asserting the prestige that they carried. This occurred perhaps most recognizably when Grant, arguably the nation's most prestigious figure at the time, refused Johnson's directive to be sent to Mexico in front of an assembled cabinet. The high level of visibility of multiple political behaviors undertaken for the purpose of influencing public opinion by civilian and military actors during the period from 1865-1868 strongly suggests that high military prestige contributed to both sets of actors engaging in these behaviors. 

## The Role of Polarization on Political Behaviors Involving the Military, 1865-1868
			
In this section of the chapter, I briefly show how high levels of polarization impacted the political behaviors engaged in by military actors from 1865-1868. As I have described above, the political questions Army officers faced during Reconstruction involved significant moral components. These included the restoration of one's fellow citizens to the nation, the treatment and awarding of political freedoms to former slaves, and the legacy of a painful and costly Civil War. As the scope of the task of Reconstruction came into focus, the concern of military actors over these and other related topics of a moral nature weakened the constraining influence of the central principles of civil-military relations. Furthermore, military actors engaged in behaviors that simultaneously violated the central principles of civil-military relations but advanced the moral views to which these actors subscribed. 

Before diving into the specific political behaviors that transpired during this period, it is important to make the central point that the fact that the Reconstruction era ever happened in the way that it did speaks to the power of high polarization in shaping a nation's civil-military dynamic. During the Reconstruction era, which lasted more than a decade, there was little if any difference between the nation's political and military policy. In other words, the extremely high level of polarization present, which had contributed to the fighting of the nation's (hopefully) only Civil War over the controversial issue of slavery, had generated a political dynamic such that civilian and military spheres now very closely, if not entirely, overlapped. The fact that during Reconstruction, military commanders in the South at times had power to enact martial law, approve the composition of state and local governments, and influence the courts, attests to this point.  

Multiple scholars shed light on this central point from different angles. Downs (2015) makes the compelling point that in many ways, American Reconstruction was a fight over who - the President or Congress - has the power to end a special type of (Civil) war [@downs_after_2015, 7-8]. Hyman (1960) is perhaps even more stark. He argues that in essence, two armies existed when the Civil War ended; one was controlled by the President, and had duties on the American frontier and along the US border with Mexico. The other army that existed was the one that implemented Reconstruction in the south. This army, Hyman argues, is the one that Congress would never allow President Johnson to control. The stakes in Reconstruction were simply too high [@hyman_santon_1960, 86-87]. The key to thing to note is that this unique and rather extreme dynamic that ran throughout the Reconstruction Era occurred during a period of high polarization, and would have been highly unlikely to occur in an era of low polarization. 

### Behaviors Stemming from Extreme Levels of Polarization

Events surrounding the removal of General Phil Sheridan as military district commander by President Johnson late in 1867 stand as a noteworthy instance of issues pertaining to Reconstruction providing a _motive_ for military actors to violate the principles of civil-military relations. Sheridan was one of several generals whom the President, simply put, did not like. The two men possessed entirely different views on the goals and methods of Reconstruction. When the President began to strongly considering removing Sheridan, he first asked Grant's opinion on the matter. Grant responded to the President in a private letter that it was "unmistakably the expressed wish of the country that General Sheridan not be removedthis is a Republic where the will of the people is the law of the land. I beg that their voice may be heard" [Grant, 1867 as cited in @milton_age_1930, 458].

In June 1867, Sheridan then removed the Governor of Louisiana for impeding Reconstruction, a strong move that provoked considerable reaction among the public, both in the south and other parts of the country [@chernow_grant_2017, 589].^[In July of that year, Sheridan also removed the Governor of Texas. See @dawson_army_1994, 56 for more details.] Johnson had his Attorney General, Henry Stanbery, produce an opinion strongly denouncing Sheridan's actions, and ordered Grant to distribute Stanbery's opinion to his commanders [@hesseltine_ulysses_1935, 85; @chernow_grant_2017, 589]. Grant complied, but not entirely. He distributed Stanbery's opinion as ordered, but in doing so, he also let his commanders know that "they could freely interpret" Stanbery's opinion as they saw fit [@chernow_grant_2017, 589]. Grant's behavior in this instance constitutes a form of selective compliance. Though Grant perhaps thought Sheridan's methods were a bit too strong, he did not fault the underlying reason premise for Sheridan's behavior: the Governor of Louisiana and other civilian officials whom Sheridan had previously removed had been responsible for preventing blacks from registering to vote, facilitating violence during race riots in New Orleans the previous year, and generally thwarting the implementation of key Reconstruction policies [@dawson_army_1994, 46-50]. Those who closely observed this episode noted that, "the President, of course, observed [Grant's] tacit disobedience, but he was powerless to control or punish his subordinates. He had disregarded the will of Congress, and in return the officers of the Army disregarded his. The situation was approaching mutiny on one side, or else treason on the other" [@badeau_grant_1887, 71].

After publication of Stanbery's opinion, Sheridan then wrote Grant a letter in which Sheridan chastised the President [@hesseltine_ulysses_1935, 85-86]. The letter was leaked and published, though it is not clear who was responsible for the leak. The President ultimately decided not to punish Sheridan for the letter on the grounds that it had been written, at least plausibly from Sheridan's perspective, as a private rather than public letter [@hesseltine_ulysses_1935, 85-86]. However, Republicans in Congress were greatly happy with Sheridan's criticism of the President, and used it as grounds to make further overtures towards Grant [@hesseltine_ulysses_1935, 85-86].

After Johnson finally removed Sheridan, an action that Grant greatly protested, Grant told his commanders not to reinstate any of the southern politicians Sheridan had previously removed [@chernow_grant_2017, 596]. Grant also furnished the contents of the previously private letter he had sent to the President about the possibility of removing Sheridan to several press outlets, including the _Army and Navy Journal_, which was popular with military audiences [@chernow_grant_2017, 596]. Grant's behaviors - directing his subordinates not to reinstate any politicians Sheridan had removed, and then leaking a letter to the press - challenged both the principles of civilian control and non-partisanship. 

Shared moral conviction led to the formation of what should be considered an alliance of sorts between Grant, Stanton, many of the 'Radical' Republicans in Congress, and several of Grant's subordinate military commanders. These actors shared the fundamental belief that the Army needed to be more heavily involved in, rather than standing more aloof from, Reconstruction policy in the South. This belief was formed as a result of several developments that had occurred within 18 months of the end of the Civil War, including the institution of Black Codes in southern states, the occurrence of race riots in Memphis and New Orleans and 1866, and the seemingly quick restoration of southern politicians who had advocated rebellion just a few years before. Grant, Sheridan, and many other military actors believed that failing to confront these developments aggressively was tantamount to squandering the purposes for which the Civil War had been fought [@hyman_santon_1960, 87; @benedict_new_1973, 351]. 

Some historians note that Grant in particular became more openly opposed to the President and his policies following the "Swing Around the Circle Tour" in the late Summer of 1866, when Grant clearly perceived how unpopular both Johnson's policies and the President himself were with the American public [for instance, see @hesseltine_ulysses_1935, 75]. This is an important consideration in some respects, especially insofar as the political philosophy that Grant would exercise as President was one that viewed the Executive's office as the executor of the "will of the people" [@hesseltine_ulysses_1935, 93]. I do not doubt that Grant (and perhaps other military commanders) took on a tone that was more willing to challenge the President after learning firsthand how unpopular the President was. But the historical record seems to suggest that Grant's perception of Johnson's lack of popularity with the American public exacerbated, rather than initiated, Grant's clear opposition to the thrust of Johnson's political views regarding Reconstruction. In other words, significant and irreconcilable moral differences existed between the President's views of Reconstruction and those held by Grant and many other military leaders.

## Summary of the 1865-1868 Period

Table \@ref(tab:pol-behav-cs1) captures several substantial political behaviors undertaken by civilian and military actors from 1865-1868. There are at least two important takeaways from this table. The first is that both types of actors, civilian and military, engaged in behaviors that were highly visible to the public. The second important takeaway is that several of these behaviors, especially partisan stacking and explicit insubordination, are rather extreme forms of political behavior in that these forms of behavior have not often occurred in the United States. 

```{r pol-behav-cs1, echo=FALSE, include=TRUE, error=FALSE, message=FALSE}

text_tbl1 <- tribble(~"Actor", ~"Tactic of Political Behavior", ~"Instance", ~"Central Principles Challenged",
  "President Andrew Johnson (Civilian)", "Partisan Stacking", "Andrew Johnson replaces Generals Sheridan, Sickles, and Pope as Military District Commanders", "Non-partisanship",
  "President Andrew Johnson (Civilian)", "The Military as Sales Reps", "Andrew Johnson directs Generals Grant and Meade, along with Admiral Farragut, to accompany him on a tour of the nation ahead of the 1866 midterm elections", "Non-partisanship",
  "President Andrew Johnson (Civilian)", "Brass in the Cabinet", "Johnson seeks to replace Stanton as Secretary of War with General Sherman", "Non-partisanship",
  "President Andrew Johnson (Civilian)", "Public Appeal", "(1) Johnson dispatches Grant to refute a previous report made by Carl Schurz regarding political conditions in the south (2) Johnson leaks the content of a heated cabinet meeting held in early 1868 to the Press for the purposes of discrediting Grant", "Civilian control (2); non-partisanship (1, 2); non-interference (1)",
  "General Grant (military)", "Selective Obedience/Compliance", "Grant publishes Attorney General Stanbery's legal opinion of the Military Reconstruction Acts in conjunction with President Johnson's wishes, but also instructs his commanders to make their own interpretations", "Civilian control",
  "General Sheridan (military)", "Public Appeal (perhaps inadvertent)", "Sheridan writes a letter to Grant in which he upbraids the President's views on Reconstruction", "Civilian control; non-partisanship",
  "General Grant (military)", "Explicit Insubordination", "Grant refuses to be sent to Mexico by Johnson", "Civilian control",
  "General Grant (military)", "Public Appeal", "After Johnson removes Sheridan, Grant leaks a private letter he had sent to Johnson in which he protested the move", "Civilian control; non-partisanship"
  )
 

text_tbl1 %>% kable(booktabs=T, caption="Political Behaviors Involving the Military, 1865-1868") %>%
  kable_styling(full_width = F, font_size=8, latex_options="striped" ) %>% 
  column_spec(1, width="10em") %>% 
  column_spec(2, width="12em") %>% 
  column_spec(3, width="16em") %>% 
  column_spec(4, width="14em") 

```

# Part 2 - The Army, Domestic Disturbance, and Failed Reform Efforts, 1877 - 1880

This chapter now turns to examining the political behaviors undertaken by civilian and military actors beginning roughly a decade later, from approximately 1877 - 1880. Similar to the first portion of this case study, I first set the scene by arguing that while the level of polarization remained high during the period, the relative prestige of the military as an institution declined from 1868-1877. I make the point that this historical period in particular accorded with a decline in the prestige of the military as an institution, even while a number of individual military figures retained an outsized level of personal prestige on the basis of their personal martial exploits during the Civil War. 

I then narrate several significant events that involved the Army from 1877 - 1880, including the use of the Army in several domestic labor disputes and bureaucratic fights. I argue that the general qualities of the political behaviors engaged in by civilian and military actors from 1877 - 1880 were noticeably different than those which both types of actors had engaged in from 1865 - 1868. In particular, civilians did not engage in the same intensity of efforts to harness or leverage the prestige of the military for their own partisan advantage because the military no longer afforded as significant a political advantage as it previously did. Additionally, military leaders likewise were cognizant that they represented an institution which by 1877 carried diminished political clout relative to a decade prior. 

## High Level of Polarization, 1877-1885

The level of polarization remained quite high as Reconstruction ended. When we view Figure \@ref(fig:polar-chamber), we can see that after a couple of very small drops in the level of polarization in the US House, polarization only climbed beginning in the mid 1880s all the way until after the start of the 20th Century. Polarization in the Senate reflects a very similar trajectory. 

Scholars have noted that polarization in the 21st century shares more than just a couple of passing similarities with the polarization that prevailed in the late 19th century. These scholars make the point that in the late 19th Century, issues related to "race, morality, immigration, and themes related to populism" contributed to American polarization [@azari_back_2016, 93]. These themes remained salient as the United States emerged from Reconstruction and embarked on a process of significant industrialization throughout the decades of the late 19th Century [@skowronek_building_1982]. In short, the level of polarization remained relatively high well into the start of the 20th Century.  

## Declining Military Prestige, 1868-1877 

However, the relative level of military prestige declined from 1868 - 1877. This decline was not the result of a single event, but rather the confluence of several trends that converged on the Army during the Reconstruction period and in the years to follow. These interrelated trends included the downsizing of the Army after the Civil War; the perception of many Americans of the Army as a result of its involvement in the political activity of Reconstruction; and a lack of clarity regarding the fundamental purposes to be achieved by the Army in an era in which no clear international threat existed. Together, these factors rendered the Army less central to the life of the nation and less respected relative to what it had been in 1865.^[Johnson's impeachment and Grant's subsequent ascendance to the Presidency ensured that the Army played a pivotal role in the political life of the America throughout the Reconstruction period. This role came to an end, however, when following the contested Presidential Election of 1876, President Grant pulled the last remaining federal troops from the south. This was the result of a deal struck between Republicans and Democrats that ultimately bestowed the White House to the Republican candidate, Rutherford B. Hayes, in exchange for the return of southern local rule free of federal military interference. See @woodward_reunion_1991, 6-8; @marszalek_sherman_2007, 428; and @wooster_united_2021, 242 for more details.]

The first factor that contributed to the Army's diminishing prestige was the significant reduction in manpower the service endured during the decade following the end of the Civil War. For instance, within six months of Lee's surrender at Appomattox, "800,000 of the 1 million men in the Union's Grand Army of the Republic were mustered out of federal service" [@laurie_role_1997, 21]. Four years later, by 1869, the Army's authorized end strength stood at 37,313 [Utley, 1974 and Cooper, 1986 as cited in @clark_preparing_2017, 99]. By the end of 1875, the Army was capped at 25,000 men, including officers and enlisted men [@laurie_role_1997, 21].^[Laurie and Cole (1997, p. 22) also note that only 2,105 officers remained in the Army after 1871 - a figure that did not change by more than 100 men until the close of the century.]

Reductions to the size of the Army touched not only on the issue of numbers, but also on issues related to the composition of the Army itself. The Army that had fought during the Civil War was in actuality a conglomerate of regular, volunteer, and militia forces. Many Americans believed that volunteers or citizen-soldiers had served the nation well during the Civil War. Thus, even as the Army downsized in the years after the Civil War, it also faced significant bureaucratic opposition, especially as Army leaders began to conceptualize a number of long term reforms.^[Clark (2017) tells the full story of the Army's efforts to reform after the Civil War in his book, _Preparing for War: The Emergence of the Modern U.S. Army, 1815-1917_. See especially his fourth chapter on this point, "The Civil War's Legacy."] The 1870s and 1880s would proved to be difficult times for Army leaders to convince Congress and the nation that significant reforms were required, particularly in the absence of a compelling international threat.^[Grandstaff (1998) and Clark (2017) conclude argue that the late 1890s and early 1900s, after the Spanish-American War, is the period in which the Army became more successful at gaining support for their intended reforms. See @grandstaff_preserving_1998, 537 and @clark_preparing_2017 for more details.] In many cases, this opposition was led by politicians who had themselves served as volunteers in the Civil War, many of whom had also achieved high rank and acclaim. This was perhaps most true of Senator John Logan, who in 1870 served as the House Military Affairs Chairman, and who had been passed over for command of the Army of the Tennessee by General William Sherman in 1864 [@clark_preparing_2017, 100]. 

A second factor that contributed to the decline in the Army's prestige in the decade after the Civil War was its involvement in the inherently political tasks of Reconstruction. Though it is difficult to precisely gauge the impact of Reconstruction on the military's relationship with all of American society, we can reasonably ascertain that the regular Army came to be especially disliked and distrusted by Southern White Democrats [@byler_civil-military_2006, 6-11]. For instance, Byler (2006) notes that of particular concern was the way that federal forces had been used at or near voting locations during elections, particularly after 1870, when the military districts created by the Military Reconstruction Acts of 1867 ceased to formally function [@byler_civil-military_2006, 8]. As Reconstruction ended, the Army thus had to navigate entrenched partisan forces that sought to define the terms under which the Army could be and should be used domestically, especially during a number of domestic labor disputes in the late 1870s and early 1880s that involved federal military forces.

Many Army leaders themselves were cognizant that institution of which they were a part occupied a less prominent role than it had previously had when the Civil War ended. For instance, while speaking to a gathering of Civil War Veterans in 1873, Major General John Pope expressed concern that the Army might "properly and naturally become an object of suspicion and dislike" in the years to come [Cozzens, 2000 as cited in @byler_civil-military_2006, 1]. 

A third and final factor contributing to the Army's diminishing prestige after the Civil War involved the lack of clarity regarding the primary purpose for which regular Army forces existed in an era of peace. Until the Spanish-American War and outside of battling Indians in America's Western territories, there were few if any significant international threats that civilian and military leaders agreed warranted extensive preparations for. The Army had also become increasingly scattered in a geographical sense. Laurie and Cole (1997, p.22) note, for instance, that upon the end of Reconstruction, "all of the cavalry forces and three fourths of the infantry" had moved to geographical outposts "west of the Mississippi Rive to battle hostile Indians and otherwise police the frontier." 

In many ways, the heart and soul of Army reform was intended to help answer the question of what the Army should be doing during peacetime to best prepare for the next major conflict. But there was little unity gained in answering this question, at least initially. Major intellectual differences were evident between General Sherman and other military actors advocating reform and their opponents, such as John Logan, on issues such as the degree to which military officers required special training and preparation in order to hone battlefield skills [@clark_preparing_2017]. Logan and others believed, based on their own experiences as volunteers, that "the overwhelming determent of success as a commander was natural 'genius'," and thus they opposed the creation of the special professional military education programs Sherman and others sought [@clark_preparing_2017, 100].^[Key figures who advocated for reforms on behalf of the regular Army included, in addition to General Sherman, Colonel Emory Upton and Major General John M. Schofield. For excellent biographies of these three individuals, respectively, see @marszalek_sherman_2007; @connelly_john_2006; and @fitzpatrick_emory_2017.] Other significant questions about the Army during peacetime, including how the Army should man itself and against whom it should prepare to defend [@grandstaff_preserving_1998, 536-537]. 

## Nationwide Domestic Disturbances and Proposed Army Reforms, 1877-1878

In this section, I narrate two related developments that occurred in the period from 1877 - 1878, and then describe the political behaviors that each actor type, military and civilian, undertook as these developments unfolded. 

The first development was the use of regular Army forces to quell what would end up being several domestic disturbances in the late 1870s - 1890s.^[See @cooper_army_1980 and @laurie_role_1997 for excellent and in-depth accounts of these events.] The first of these disturbances, known as _The Great Railway Strike of 1877_, began in Maryland in July of that year before spreading to West Virginia, Pennsylvania, and later, Chicago. This disturbance, like many subsequent strikes, occurred as workers, particularly in the rail and coal industries, protested against what they perceived to be unfair and inequitable working conditions, to include the enaction of several wage cuts by their employers after the economic downturn of 1873 [@cooper_army_1980, 43-45]. At first, state and local officials tried to respond to these strikes with local law enforcement entities and state National Guard units, but these proved insufficient to handle the strikers, who had not only ceased all train traffic, but in some cases, had engaged in acts of violence [@cooper_army_1980, 43-45; @laurie_role_1997, 30-32]. Despite concerns about involving the Army in what might be perceived as an economic "class war", President Hayes dispatched the Regular Army to maintain law and order [@laurie_role_1997, 32]. 

The timing of this strike is notable in that it intersected with an Army whose political position had been weakened in important ways over the previous several years. Congress had failed to pass an appropriations bill funding the Army past July 1 of 1877, resulting in an Army that was essentially serving without pay. Moreover, the failed appropriations bill was the result of bitter congressional fighting between southern Democrats who were angry with an Army that had taken on significant political roles during Reconstruction, and Republicans who had overseen and directed the use of the Army in precisely these controversial tasks. The use of the Army in southern states to secure polling sites during the contested presidential election of 1876, held just months earlier, further roused these tensions [@cooper_army_1980, 47-48; @laurie_role_1997, 31-32]. In protest of Congressional Republicans failing to commit to any meaningful restrictions on the Army's use domestically after the election, democrats "blocked the 1877 appropriations bill" [@clark_preparing_2017, 115].

The Army was generally successful in quelling the _Great Strike of 1877_, and many citizens, popular press outlets, and Army leaders perceived the Army as having performed admirably [@cooper_army_1980, 60-61; @laurie_role_1997, 52-55; @skowronek_building_1982; 100-101]. General Winfield Scott Hancock, who commanded the Army's response in the East to the civil unrest, wrote to his fellow General John Schofield, stating, "the troops have lost the government no prestige...it was the moral force of the United States government that was displayed - not its physical force" [Hancock, 1877 as cited in @cooper_army_1980, 61]. Yet even after the Army's relative success in responding to this major episode of civil unrest, the general political environment remained apathetic towards the Army.^[Congress would not allocate funds to the Army for the express purposes of dealing with domestic labor disputes until after the _Pullman Strike_ of 1894, which originated near Coeur d'Alene, Idaho. See @cooper_army_1980, 165-196 for more details.] 

A second development involves a congressional effort that tried (and ultimately failed) to reform the Army the following year in 1878. Indeed, the army appropriations bill passed by Congress in 1878 contained two critical provisions. First, at the behest of Congressional Democrats, the bill instituted the _posse comitatus_ law. In response to what they perceived to be the heavy handed and inappropriate use of the Army throughout the Reconstruction period, Democrats "forbade the use of the Army as a posse to aid local officials in dealing with domestic turmoil without the express order of the President, and only then for very specific purposes" [@cooper_army_1980, 83]. Interestingly, historians note that Army leaders remained relatively "quiet" about the passage of this important law [@connelly_john_2006, 238]. Second, the bill established a bipartisan committee known as the _Burnside Commission_ to examine a number of military reorganization and reform topics, including the Army's size as well as systems involving manning, promotion, retirement, and staffing [@clark_preparing_2017, 116].^[Connelly (2006), Fitzpatrick (2017),and Clark (2017) are among the contemporary historians who write about these particular reforms extensively.]   

Chaired by Senator Ambrose Burnside, a former Union General, the _Burnside Commission_ recommended a number of overarching reforms to the Army. But in many ways, the proposals recommended by the Commission were a series of compromises: to placate Republicans, the legislation proposed that the overall size of the Army remain the same; to placate Democrats who largely wished to reduce the Army, the legislation proposed a reduction in the number of Army officers, including the number of generals, as well as a reduction in several types of particular Army units [@fitzpatrick_emory_2017; @clark_preparing_2017, 118-121; @skowronek_building_1982; 103]. 

The Army did not receive and react to the proposed legislation uniformly. Some, including Generals Sherman and Schofield, cautiously and perhaps somewhat reluctantly approved of the legislation on the grounds that it likely served as the Army's best chance at obtaining any meaningful reforms in the foreseeable future, while others, such as Generals Sheridan and Hancock, did not support the legislation [@clark_preparing_2017, 118-121; @jordan_winfield_1988, 254].^[Clark (2017) notes that while Sheridan approved of many aspects of the legislation, including proposed changes to the delegated authorities exercised by the commanding general, Sheridan took issue with the legislation's plans to trim the Army's Cavalry formations.] Several historians explain the legislation's ultimate failure as reflective of the overall national sentiment at the time, which was simply that the nation had moved on to other problems and issues that rendered the military less central to the nation. Nearly four decades later, Francis V. Greene commented on the the Burnside legislation's failure, noting that the prevailing political attitude at the time was that, "this country would never be engaged in war; and therefore, in the midst of such pressing problems as the building of railroads, the resumption of specie payments, the silver question, and the tariff, there was no time to think about the needs of the Army. There was no 'military situation in the United States' worth thinking about" [Greene, 1919 as cited in @fitzpatrick_emory_2017, 223]. 

## Declining Military Prestige and Political Behaviors, 1877-1878

The curtailment of the Army's centrality to the nation impacted how military and civilian leaders advocated in support of or against the proposed legislation. Military actors rarely addressed or appealed to the general public about the proposed reforms and instead, either addressed fellow military actors or select politicians directly. Additionally, most, if not all, of the Army's senior leaders remained remarkably quiet about the legislation, and did not engage in significant efforts either in support of or against the bill's passage. Furthermore, these political activities undertaken by military actors violated the principles of civil-military relations, but only by exception rather than as the rule.  

For starters, a number of military officers wrote letters to the _Burnside Commission_ expressing their opinions on the proposed changes. However, these letters should be seen as appropriate acts of behavior in response to legitimate Congressional inquiry. Some officers went above and beyond the writing of letters in response to Congressmen, however. For instance, Fitzpatrick (2006) notes that many staff bureau officers in Washington D.C., who in particular feared that the legislation would reduce their authority and influence as staff officers with Congress, produced and distributed informational pamphlets to members of congress [@fitzpatrick_emory_2017, 221]. These efforts appear to at least have been partially successful. For example, Senator Burnside himself, upon realizing that the proposed legislation was stalling, noted, "I must say that some of these staff officers have gone beyond the line of duty, particularly...in Washington" [Burnside, 1878 as cited in @fitzpatrick_emory_2017, 221]. 

Military officers handing out informational pamphlets in order to persuade members of congress to support, or in this case, denounce the legislation in question amount to acts of "shoulder tapping" and "alliance building", which were outlined in chapter two of this dissertation [@brooks_militaries_2009]. As Senator Burnside's remarks suggest, the act of furnishing pamphlets - unlike the act of writing in response to an inquiry made by Congress - violated the central principles of civil-military relations. In particular, these acts violated the principle of civilian control in that the _Burnside Commission_, a legitimate Congressional entity attempting to exercise oversight of the military by asking military officers to provide their opinions on a number of reform topics, now had to contend with other military officers functioning as a disparate interest group. Furthermore, these acts also likely violated the principle of non-partisanship because by attempting to persuade members of congress, the officers were essentially taking a position that was consistent with one party, that of the Democrats, over and above the other. 

Military actors also engaged in making a number of what can be described as public appeals regarding the legislation, but these appeals differed in a number of important ways from the leaks and appeals to the press made by Army leaders a decade earlier. For instance, General Sherman passed on letters written by Generals Schofield, Sheridan, and Pope, as well as Lieutenant Colonel Upton, to the _Army and Navy Journal_ to be published [@connelly_john_2006, 243]. These letters addressed the proposed reforms from a number of angles. But it is vital to keep two things in mind when considering this particular action. 

The first is that Sherman's behavior did not constitute a leak per se, but rather an act that was mostly likely undertaken with the author's concurrence _and_ for the purposes of informing rather than influencing. To be clear, I could not find definitive proof that this was the case, but it seems the most likely explanation given Sherman's ambivalent disposition regarding the legislation.^[Marszalek (2007, p. 435) quotes Sherman as having stated, "I shall not advocate the Bill or oppose it, but shall conform my action, official and personal, to the conclusion of the Law Making Powers of Our Government."] Indeed, the historical record indicates that Sherman himself cautiously favored the bill, but had decided nonetheless not to take a highly visible position on the issue [@marszalek_sherman_2007, 435; @connelly_john_2006, 243; @clark_preparing_2017, 118].

The second consideration to note is that the _Army and Navy Journal_ was a periodical primarily read and consumed by military actors, to include soldiers, sailors, and veterans. The fact that Sherman appealed to a publication geared to military actors, rather than to a major popular newspaper with a significant civilian readership, further suggests that Sherman intended to reach military rather than civilian audiences. With these two considerations in mind, Sherman's behavior, then, in this instance, can best be viewed as having not violated the principles of civil-military relations.

Civilian leaders, too, did not engage in as significant of political activities as they had undertaken in the critical period from 1865 - 1868. They lobbied on behalf of (or against) the legislation, but in ways that were consistent with the duties of policymakers and in ways that did not violate the principles of civil-military relations. For instance, Representative James Garfield, who later became President and who had served as a Union General during the Civil War, wrote a two part series in _The North American Review_ in early 1878 in which he advocated that Congress take seriously the topic of Army reform. In these publications, Garfield included a portion of, or in some cases, the entirety of, a number of letters written by senior Army leaders addressing the size of the peacetime Army, the role of the staff vis-a-vis the commanding general and the Secretary of War, and other pertinent items [@garfield_army_1878; @garfield_army_1878-1]. In all likelihood (I do not know for sure; this is speculative), Garfield had the permission of the officers whose letters he reprinted in the publications that he authored, but even if he did not, the tone and tenor with which Garfield writes does not lend the impression that he is politicizing the Army in a significantly nefarious manner. Rather, his articles serve as a sort of alarm bell for the nation, and urge the American people (and Congress in particular) to treat the Army seriously in spite of the very real transition from war to peace that had occurred since the end of the Civil War. Certainly, Garfield's publications take a stake in the debate about Army reform; as a Republican, he was largely in favor of it. Still, However, there is no obvious attempt on Garfield's part to leverage the Army or the topic of Army reform for his own political advantage in the publications. Indeed, there was relatively little - if any - advantage the Army and its leaders could have offered Garfield or other civilian leaders at this particular juncture. 

## Exceptions - Military Actor Presidential Candidates 

Other political behaviors seem not to have been tremendously impacted by the drop in the relative level of military prestige, however, and thus, these behaviors stand out as possible exceptions to the predictions made by the theory earlier in this chapter. Among the clearest of these possible examples involves the political behavior of military figures running for office, and in particular, for the Presidency. 

To see this, consider that during the seven Presidential Elections held from 1868 to 1892, only the election between Grover Cleveland and James Blaine in 1884 featured two non-veteran candidates [@teigen_why_2018, 104-106]. All other elections featured at least one veteran as one of the party's major final candidates. These candidates running for President in elections held between 1868 and 1892 (except for 1884) are displayed in Table \@ref(tab:ch5-candidates). This then raises two important questions: first how could military figures become a political party's final candidate if military prestige truly declined in the decade after the Civil War, and second, what accounts for General Winfield Scott Hancock in particular obtaining the 1880 Democratic nomination at a time when the military as an institution was relatively less prestigious than it used to be?

```{r ch5-candidates, include=T, echo=F, error=F, message=F}

ch5_text_tbl2 <- tribble(~"Election Year", ~"Candidate", ~"Party", ~"Veteran Type",
  "1868", "Ulysses S. Grant", "Republican", "Career General",
  "1872", "Ulysses S. Grant", "Republican", "Career General",
  "1876", "Rutherford Hayes", "Republican", "Battlefield General",
  "1880", "Winfield Scott Hancock (and James Garfield)", "Democrat (Republican)", "Career General (Battlefield General)",
  "1888", "Benjamin Harrison", "Republican", "Battlefield General",
  "1892", "Benjamin Harrison", "Republican", "Battlefield General"
  )
 
ch5_text_tbl2 %>% kable(booktabs=T, caption="Presidential Candidates, Post Civil-War Era") %>%
  kable_styling(full_width = F, font_size=8, latex_options="striped" ) %>% 
  column_spec(1, width="10em") %>% 
  column_spec(2, width="12em") %>% 
  column_spec(3, width="8em") %>% 
  column_spec(4, width="14em") %>% add_footnote(label = "Reproduced from Table 5.1 in Jeremy Teigen's (2018) book, Why Veterans Run (p. 105).")
```

To answer the first question, it is critical to understand that there are distinctions between types of military service. As Teigen (2018) notes and as the far right column of Table \@ref(tab:ch5-candidates) describes, this was especially true during the late 19th Century, though it is also true today. For instance, Grant and Hancock were career generals whose professional roots lie with the regular Army, while Hayes, Garfield, and Harrison had served in the military for considerably shorter periods of time. Harrison in particular served less than four years in uniform, but obtained the rank of Brigadier General before resuming civilian life [@teigen_why_2018, 19]. The point here is that while the military service of Grant and Hancock played perhaps the dominant factor in achieving the political successes that each obtained, it need not be the case that the same be true for the candidacies of Hayes, Garfield, and Harrison. It could certainly be true that the military service of these three men constituted even perhaps significant factor in each man's success in reaching a terminal level of political candidacy, but it is doubtful that these men ran on a platform that showcased their lifetime military service to the nation, as such a platform was simply not the case. 

The second question, however, is also important: if the military was indeed relatively unpopular in 1880, what accounts for General Winfield Scott Hancock obtaining the Democratic nomination for President? A likely answer is that Hancock, like many senior military officers of his time, possessed significant personal prestige that operated at a high level even while the prestige of the Army as an institution was low. If the variable of military prestige, then, as I am suggesting, contains a personal-institutional as well as an institutional-organizational component, is this a problem for the theory?     

While future research is required to fully unpack the personal and institutional domains of military prestige as a variable, it is doubtful that the mere presence of these different components negates the theory previously offered in this dissertation. In broad terms, I maintain that military prestige as a variable still impacts the dependent variable of political activity in the same direction, i.e., greater military prestige increases the likelihood that civilian and actors act in ways to leverage public opinion, even if in the process, these actors violate the principles of civil-military relations. However, and as the case of General Winfield Scott Hancock illustrates, personal and institutional levels of military prestige may in fact operate at different levels, even at the same time.

## Summary of the 1877-1878 Period

In the decade after the Civil War, multiple factors combined to render the Army less prestigious than it had been at the close of the Civil War. The political nature of Reconstruction had been unpopular with many Americans, and particularly with southerners. As the nation reunified, the Army became a perennial target by Democrats who sought to preclude the Army's domestic involvement and trim its size and influence within the government. Indeed, the Army did downsize, and significantly so, as it transitioned to constabulary duties and fighting Indians after the Civil War. In the absence of a clear international enemy to prepare against, the Army was largely at the mercy of a divided Congress that could not be compelled to take up a number of reforms in the late 1870s. 

The Army's reduced standing and political influence in this era greatly impacted the character of the political behaviors Army and civilian leaders alike undertook. Even after performing rather well at quelling a serious labor dispute in 1877, military leaders avoided the public spotlight as Congress passed a bill outlining _posse comitatus_ regulations. When Congress came close to adopting a number of reforms proposed by the _Burnside Commission_ the following year, many military actors engaged with fellow military actors on these topics, but they largely did so in military journals and private correspondence rather than through national newspaper outlets or other civilian publications. Members of the Army's staff bureaus, who stood the most to lose from a number of the legislation's proposed reforms, engaged in more obvious acts such as the construction and distribution of informational pamphlets to members of Congress, but even these acts were less visible than the leaks and interviews with the press conducted by Army leaders at the end of the Johnson Administration. 

A summary of the political behaviors engaged in by both sets of actors from 1877-1880 is captured in Table \@ref(tab:pol-behav-cs1-part2).

```{r pol-behav-cs1-part2, echo=FALSE, include=TRUE, error=FALSE, message=FALSE}

text_tbl3 <- tribble(~"Actor", ~"Tactic Political Behavior", ~"Instance", ~"Central Principles Challenged",
  "Military Staff Officers", "Shoulder Tapping; Alliance Building", "Staff officers distribute informational pamphlets to members of Congress arguing against adopting a number of provisions proposed by the Burnside Commission", "Civilian Control; Non-partisanship",
  "General William Sherman", "Public Appeals", "General Sherman sent letters written by fellow military commanders to be published by The Army and Navy Journal regarding various reforms under consideration by the Burnside Commission", "No principles violated; the public appeal was geared to the military community and conducted for the purposes of informing rather than influencing",
  "Representative James Garfield (Civilian)", "Public Appeal", "Garfield publishes two articles in The North American Review on the topic of Army reform in early 1878", "No principles violated; the articles were written in general terms and did not seek to leverage the Army for personal political or partisan gain",
  "Winfield Scott Hancock", "Running for Office", "General Hancock runs for political candidacy even as he is an active member of the Army in 1880", "Non-partisanship"
  )
 

text_tbl3 %>% kable(booktabs=T, caption="Political Behaviors Involving the Military, 1877-1881") %>%
  kable_styling(full_width = F, font_size=8, latex_options="striped" ) %>% 
  column_spec(1, width="10em") %>% 
  column_spec(2, width="12em") %>% 
  column_spec(3, width="16em") %>% 
  column_spec(4, width="14em") 

```

## Alternate Explanations

Besides the relative decline in the importance of the Army to the nation in the years following the Civil War, a number of other factors might explain the changes in how military and civilian leaders engaged in political activity involving the military. One possible alternate explanation involves the individual personality differences between senior army leaders between 1865 and 1877, and indeed, there may be some merit to this explanation. Grant was no lover of politics as a General, but it is also well known among historians that Sherman and Grant approached politics rather differently. Marszalek (2007) for instance notes that, "...politics was anathema to Sherman. He viewed politicians as venal adversaries he was forced to tolerate but refused to work with" [@marszalek_sherman_2007, 425]. Thus, there seems to be at least some basis for suggesting that the relative decline in the visibility of political behaviors engaged in by military actors that violated the principles of civil-military relations, particularly as the _Burnside Commission_ considered a number of reforms in 1878, was the product of a disgruntled Sherman as the commanding general of the Army, rather than because of the broader political environment. 

Such an explanation is not entirely sufficient, however. I will leave the historians to be the definitive experts on the personalities of Grant, Sherman, and other Army leaders, but even if Sherman was naturally inclined against fully embracing his appropriate political role, the historical literature surveyed for this chapter clearly indicated that Sherman lamented the fact that he was presiding over the Army at a time that coincided with the loss of the Army's popularity or standing in the country. There is also little to suggest that any major professional military norms concerning behavior changed in the 15 years between 1865 and 1880. 

The conflict or threat environment, however, did change between 1865 and 1868. The Civil War was a major conflict that devastated roughly half of the United States. Nonetheless, it seems mostly incorrect to state that Lee's surrender in 1865 ushered in an era of peace. The fact is that the Reconstruction era spurred the continuation of significant wartime authorities and the undertaking of what should be considered a major military effort, an occupation, of the southern states [@downs_after_2015]. On this basis, it appears, then, that a change in the conflict environment did not in itself cause a change in the degree to which civilian and military actors engaged in political behaviors that violated the principles of civil-military relations.  

A more accurate rendering, based on the evidence described in this chapter, suggests that a gradual change in the conflict environment, as well as evolving political circumstances that occurred during the Reconstruction era, rendered the Army both less popular as well as less central to the life of the nation. In 1865, there was little if any difference between the nation's political and military policies, as both centered on the important work of undertaking Reconstruction. As Johnson and Congress wrestled for control of the military, Grant and other Army leaders eventually could not remain neutral due to the significant moral undertones of Reconstruction itself. In perhaps what was America's gravest civil-military relations crisis in its history, civilian and military leaders engaged in a number of behaviors that challenged the principles of civil-military relations. 

By the late 1870s, even as polarization remained high, the Army became increasingly distanced from major civilian political priorities. During this period, civilian leaders in particular undertook fewer instances of harnessing the prestige of the Army for the purposes of shifting public opinion. Military leaders, for their part and with few exceptions, engaged in relatively less overt forms of political behavior challenging the principles of civil-military relations.   

# Conclusion

This chapter ultimately finds moderate support for H4, which stated that military and civilian actors engage in relatively more visible and public forms of political behavior that involve the military institution when military prestige is high relative to periods of low military prestige. It also finds moderate support for H5, which stated that the most extreme levels of overall political activity occur in environments that are characterized by simultaneously high levels of political polarization and military prestige. Furthermore, this support for H5 follows previous support already obtained for H1 and H3 in chapters three and four. 

It is not entirely clear, however, that the civil-military environment that produced the political activities examined in the second half of this case study could only arise as the result of a combination of high polarization and low military prestige. The second half of the case study demonstrated military actors initiating political activity more so than civilian actors, but not on the basis of the high level of polarization or the polarizing topics that existed at the time. Thus, while this chapter has helped to confirm several hypotheses, it has been less helpful clarifying the presence of four unique environments, as shown in Table \@ref(tab:poss-environment) of chapter two. 

The next chapter now examines an era in which the level of polarization predominantly changes. 
\pagebreak