
\newpage

# Appendix to Chapter Three 

This chapter Appendix roughly aligns with the chronological order of the chapter.   

## Word and Data-Base Searches to Obtain Data

I conducted two different sets of searches to obtain the data used in this study using data bases made available through the University of Wisconsin online library. Data collection occurred over the period from April 2020 - March 2021. This first set of search terms was applied to the Factiva and Nexus-Uni data bases. In these databases, I had to select the "type of article" that I was searching for, and I included options such as "commentary/opinion" and "letters to the editor".

The specific terms of search included were as follows:

"retired Army General" or "former Army General" or "retired Navy Admiral" or "former Navy Admiral" or "retired Marine Corps General" or "former Marine Corps General" or "former Air Force General" or "retired Air Force General" or "retired Coast Guard Admiral" or "former Coast Guard Admiral" or "retired from the Marine Corps" or "retired from the Army" or "retired from the Navy" or "retired from the Air Force" or "retired from the Coast Guard" or "supreme commander of" or "commander of" or "Chairman of the Joint Chiefs" or "chairman of the joint chiefs" or "Supreme Commander of" or "Chief of Staff" or "Chief of Naval Operations" or "retired Army" or "retired Navy" or "retired Air Force" or "retired Marine Corps" or "retired Coast Guard" or "retired military" or "commandant of" or "director of" or "retired from the military"

The second set of search terms was applied to a data base known as Pro Quest Historical Newspapers, which the University of Wisconsin also had access to for both the _Wall Street Journal_ and the _New York Times_. Factiva included articles from the Wall Street Journal only back to the middle of 1994, so I relied heavily on Pro Quest to search through the Wall Street Journal from 1979-1994. Unlike the search terms used for Factiva and Nexus-Uni, however, I came to quickly realize that the opinion pieces I sought were not categorized as opinion-editorials or letters to the editor, but rather as "articles."  This required significantly more time searching for the desired data, as more articles had to be searched through and sifted. The specific terms of search included:

(retired OR former OR “chief of staff”) AND (“Army General” OR Gen. OR general OR "Navy Admiral" OR Adm. OR admiral OR "Marine Corps General" OR "Coast Guard Admiral" OR "Air Force General" OR "Chairman of the Joint Chiefs" OR "colonel" OR "captain" OR "commander" OR "chief of naval operations" OR "commandant" OR "director of")

## Main Regression Table Observations 

Table \@ref(tab:appx-ch3-1) contains those observations that correspond to the dependent variable equaling '1' in models 3-6 of Table \@ref(tab:regression1). In other words, these are the retired officer commentary pieces whose content either blatantly criticized civilian leaders, adopted explicitly partisan stances, or addressed issues that can be considered tangential to military expertise. 

\singlespace

```{r appx-ch3-1, include=TRUE, echo=FALSE, message=FALSE, error=FALSE}
#create a table for the Appendix
TA1 <- quest_prof %>% rename(CivViol=CivConViol, PartViol=NonPartViol, IntViol=NonPartViol) %>% 
  kable(format="latex", caption="Op-Eds authored by Retired Military Officers that Strongly Criticized Civilian Leaders, Expressed Partisan Positions, or Addressed Topics Outside of Military Expertise, 1979-2020", longtable=T) %>%
  kable_styling(font_size=9, latex_options="striped", repeat_header_continued=T) %>% 
  column_spec(1, width="8em") %>% 
  column_spec(2, width="12em") %>% 
  column_spec(3:4, width="2.5em") %>%
  column_spec(5, width="6em") %>%
  column_spec(6:8, width="4.5em") %>%
  row_spec(0, bold=T)

TA1
```

## Robustness Tests

\doublespace

In this section of the Appendix, I demonstrate a series of robustness checks.

### Robustness to the Trump Presidency Years

I first include a series of robustness tests to the years of the Trump Presidency to determine whether the Trump years are in fact vital to obtaining the result that polarization drives the degree to which retired military officer commentary violates one or more of the principles of civil-military relations. To investigate this possibility, I truncate the data at certain years and then proceed with regression analysis that mirrors the models included in the chapter's main regression table. The results are displayed in Table \@ref(tab:regression4).

```{r rob-test-trump, include=FALSE, warning=FALSE, echo=FALSE, message=FALSE}
## Robustness Tests
# Is the data robust to periods NOT including Trump?
# Answer: Data is generally NOT robust to data thru 2016, but is generally robust to data thru 2017/2018

# first make a series of dfs that truncate the data
# at or near the Trump Presidency

violationregdata_thru2015 <- reg_df_macro %>% filter(PubYr <=2015)
regdata_thru2015 <- reg_df %>% filter(PubYr<= 2015)

violationregdata_thru2016 <- reg_df_macro %>% filter(PubYr <=2016)
regdata_thru2016 <- reg_df%>% filter(PubYr<= 2016)

violationregdata_thru2017 <- reg_df_macro %>% filter(PubYr <=2017)
regdata_thru2017 <- reg_df %>% filter(PubYr<= 2017)

violationregdata_thru2018 <- reg_df_macro %>% filter(PubYr <=2018)
regdata_thru2018 <- reg_df %>% filter(PubYr<= 2018)

# DV - total proportion of annual publications that violate standards, with house and senate polarization
mod5_prop_ann_violate_house_robtest <- lm(Prop_Pubs_Viol ~ Inst_Cred_Gallup + elecyear + house_polar + HostCasRateLagged, data=violationregdata_thru2016)
summary(mod5_prop_ann_violate_house_robtest)

mod6_prop_ann_violate_house_robtest <- lm(Prop_Pubs_Viol ~ Inst_Cred_Gallup + elecyear + house_polar + HostCasRateLagged, data=violationregdata_thru2017)
summary(mod6_prop_ann_violate_house_robtest)

mod7_prop_ann_violate_house_robtest <- lm(Prop_Pubs_Viol ~ Inst_Cred_Gallup + elecyear + house_polar + HostCasRateLagged, data=violationregdata_thru2018)
summary(mod7_prop_ann_violate_house_robtest)

mod5_prop_ann_violate_sen_robtest <- lm(Prop_Pubs_Viol ~ Inst_Cred_Gallup + elecyear + senate_polar + HostCasRateLagged, data=violationregdata_thru2016)
summary(mod5_prop_ann_violate_sen_robtest)

mod6_prop_ann_violate_sen_robtest <- lm(Prop_Pubs_Viol ~ Inst_Cred_Gallup + elecyear + senate_polar + HostCasRateLagged, data=violationregdata_thru2017)
summary(mod6_prop_ann_violate_sen_robtest)

mod7_prop_ann_violate_sen_robtest <- lm(Prop_Pubs_Viol ~ Inst_Cred_Gallup + elecyear + senate_polar + HostCasRateLagged, data=violationregdata_thru2018)
summary(mod7_prop_ann_violate_sen_robtest)


# DV - total # of violations by year, with house and senate polarization
mod5_num_ann_violate_house_robtest_2015 <- lm(Total_Viol_Count ~ Inst_Cred_Gallup + elecyear + house_polar + HostCasRateLagged, data=violationregdata_thru2015)
summary(mod5_num_ann_violate_house_robtest_2015)

mod5_num_ann_violate_house_robtest <- lm(Total_Viol_Count ~ Inst_Cred_Gallup + elecyear + house_polar + HostCasRateLagged, data=violationregdata_thru2016)
summary(mod5_num_ann_violate_house_robtest)

mod6_num_ann_violate_house_robtest <- lm(Total_Viol_Count ~ Inst_Cred_Gallup + elecyear + house_polar + HostCasRateLagged, data=violationregdata_thru2017)
summary(mod6_num_ann_violate_house_robtest)

mod7_num_ann_violate_house_robtest <- lm(Total_Viol_Count ~ Inst_Cred_Gallup + elecyear + house_polar + HostCasRateLagged, data=violationregdata_thru2018)
summary(mod7_num_ann_violate_house_robtest)

mod5_num_ann_violate_sen_robtest_2015 <- lm(Total_Viol_Count ~ Inst_Cred_Gallup + elecyear + senate_polar + HostCasRateLagged, data=violationregdata_thru2015)
summary(mod5_num_ann_violate_sen_robtest_2015)

mod5_num_ann_violate_sen_robtest <- lm(Total_Viol_Count ~ Inst_Cred_Gallup + elecyear + senate_polar + HostCasRateLagged, data=violationregdata_thru2016)
summary(mod5_num_ann_violate_sen_robtest)

mod6_num_ann_violate_sen_robtest <- lm(Total_Viol_Count ~ Inst_Cred_Gallup + elecyear + senate_polar + HostCasRateLagged, data=violationregdata_thru2017)
summary(mod6_num_ann_violate_sen_robtest)

mod7_num_ann_violate_sen_robtest <- lm(Total_Viol_Count ~ Inst_Cred_Gallup + elecyear + senate_polar + HostCasRateLagged, data=violationregdata_thru2018)
summary(mod7_num_ann_violate_sen_robtest)

# DV - that the op-ed violates any of the central norms of civ-mil relations
mod5_anyviol_house_robtest <- glm(any_CMR_viol ~ Inst_Cred_Gallup + elecyear + house_polar + HostCasRateLagged + AuthRank,
    family=binomial(link="logit"),
    data=regdata_thru2016
    )
summary(mod5_anyviol_house_robtest)

mod6_anyviol_house_robtest <- glm(any_CMR_viol ~ Inst_Cred_Gallup + elecyear + house_polar + HostCasRateLagged + AuthRank,
    family=binomial(link="logit"),
    data=regdata_thru2017
    )
summary(mod6_anyviol_house_robtest)

mod7_anyviol_house_robtest <- glm(any_CMR_viol ~ Inst_Cred_Gallup + elecyear + house_polar + HostCasRateLagged + AuthRank,
    family=binomial(link="logit"),
    data=regdata_thru2018
    )
summary(mod7_anyviol_house_robtest)

mod5_anyviol_house_robtest_2015 <- glm(any_CMR_viol ~ Inst_Cred_Gallup + elecyear + house_polar + HostCasRateLagged + AuthRank,
    family=binomial(link="logit"),
    data=regdata_thru2015
    )
summary(mod5_anyviol_house_robtest_2015)

mod5_anyviol_sen_robtest <- glm(any_CMR_viol ~ Inst_Cred_Gallup + elecyear + senate_polar + HostCasRateLagged + AuthRank,
    family=binomial(link="logit"),
    data=regdata_thru2016
    )
summary(mod5_anyviol_sen_robtest)

mod6_anyviol_sen_robtest <- glm(any_CMR_viol ~ Inst_Cred_Gallup + elecyear + senate_polar + HostCasRateLagged + AuthRank,
    family=binomial(link="logit"),
    data=regdata_thru2017
    )
summary(mod6_anyviol_sen_robtest)

mod7_anyviol_sen_robtest <- glm(any_CMR_viol ~ Inst_Cred_Gallup + elecyear + senate_polar + HostCasRateLagged + AuthRank,
    family=binomial(link="logit"),
    data=regdata_thru2018
    )
summary(mod7_anyviol_sen_robtest)
```

The results seem to indicate that including the Trump years matters, but only somewhat. For example, when the dependent variable is the proportion of annual commentary that violates one or more of the central principles of civil-military relations (models 1-4), the coefficient for polarization is not statistically significant in models 1 and 2, when the data is truncated to exclude commentary authored in the years 2018, 2019, and 20. However, when I include the year 2018 (models 3 and 4), polarization becomes statistically significant and positive. 

\singlespace
```{r Regression-Robustness-Tests, include=TRUE, echo=FALSE, results="asis", warning=FALSE, message=FALSE}
stargazer(mod6_prop_ann_violate_house_robtest, mod6_prop_ann_violate_sen_robtest, mod7_prop_ann_violate_house_robtest, mod7_prop_ann_violate_sen_robtest, mod5_anyviol_house_robtest_2015, mod5_anyviol_sen_robtest, mod6_anyviol_house_robtest, mod7_anyviol_sen_robtest,
          type="latex",
          label="tab:regression4",
          title="Regression Results: Robustness Checks (Data Cutoff Thru Years Near and During Trump Administration)", 
          covariate.labels = c('Trust in Military', 'Election Year', 'House Polarization', 'Senate Polarization', 'Hostile Casualty Rate (lagged)', 'Rank'),
          dep.var.labels = c('Proportion Violating', 'Log Odds of Breaking a Central Principle'),
          column.labels=c('thru 2017','thru 2017', 'thru 2018', 'thru 2018', 'thru 2015','thru 2016', 'thru 2017', 'thru 2018'),
          model.names=TRUE,
          header=FALSE,
          column.sep.width = "1pt",
          align=TRUE,
          font.size="small",
          no.space=TRUE,
          omit.stat=c("f", "ser"),
          float.env="sidewaystable"
          )
```
\doublespace

In models 5-8, where the dependent variable is the log of the odds that a particular piece of commentary violates the central principles of civil-military relations, a statistically significant result for polarization is obtained beginning in model 7, and the result holds for model 8. 

Thus, the Trump Presidency years matter in terms of obtaining the positive and statistically significant results for polarization, but only a portion of the Presidency matters (2017 and 2017 matter when the dependent variable is the proportion that violates a central principle of civil-military relations, and only 2017 matters when the dependent variable is the log of the odds that a particular piece violates the principles).

## Robustness Checks to Interaction of Polarization and Military Prestige

I now run a series of models that check to see if any interaction between polarization and military prestige matters. The results, which are displayed in Table \@ref(tab:regression5), show that there is no meaningful impact of interacting polarization and military prestige, either when the dependent variable is the total number of pieces authored (model 1), the proportion of annual pieces that violate the principles of civil-military relations (model 2), or the log of the odds that a single piece violates one or more of the central principles (models 3 - 6). 

```{r appendix-interaction-terms, include=FALSE, warning=FALSE, echo=FALSE, message=FALSE}

# dv - num annual pubs
mod1_num_ann_pubs_house_lagged_int <- lm(Yrcount ~ Inst_Cred_Gallup + elecyear + house_polar + HostCasRateLagged + Inst_Cred_Gallup * house_polar, data=reg_df_macro)
summary(mod1_num_ann_pubs_house_lagged_int)

# dv - proportion of pubs violating

mod1_prop_ann_violate_sen_lagged_int <- lm(Prop_Pubs_Viol ~ Inst_Cred_Gallup + elecyear +   senate_polar + HostCasRateLagged + Inst_Cred_Gallup*senate_polar, data=reg_df_macro)
summary(mod1_prop_ann_violate_sen_lagged_int)

# dv - log of the odds a piece violates
mod2_anyviol_house_lagged_int <- glm(any_CMR_viol ~ Inst_Cred_Gallup + elecyear + house_polar + HostCasRateLagged + Inst_Cred_Gallup*house_polar + AuthRank,
    family=binomial(link="logit"),
    data=reg_df
    )
summary(mod2_anyviol_house_lagged_int)

mod2_anyviol_sen_lagged_int <- glm(any_CMR_viol ~ Inst_Cred_Gallup + elecyear + senate_polar + HostCasRateLagged + Inst_Cred_Gallup * senate_polar  + AuthRank,
    family=binomial(link="logit"),
    data=reg_df
    )
summary(mod2_anyviol_sen_lagged_int)

mod3_anyviol_house_mixed_lagged_int <- glmer(any_CMR_viol ~ Inst_Cred_Gallup + elecyear + house_polar + HostCasRateLagged + Inst_Cred_Gallup*house_polar + AuthRank + Source + (1 | Author), family=binomial, data=reg_df_slim)
summary(mod3_anyviol_house_mixed_lagged_int)


mod3_anyviol_sen_lagged_int <- glm(any_CMR_viol ~ Inst_Cred_Gallup + elecyear + senate_polar + HostCasRateLagged + Inst_Cred_Gallup*senate_polar + AuthRank +Source ,
    family=binomial(link="logit"),
    data=reg_df_slim
    )
summary(mod3_anyviol_sen_lagged_int)

mod3_anyviol_sen_robust_lagged_int <- robustify(mod3_anyviol_sen_lagged_int, cluster=Author)
summary(mod3_anyviol_sen_robust_lagged_int)


# Marginal Effects Prep and Check
mod3_anyviol_house_int_lagged <- glm(any_CMR_viol ~ Inst_Cred_Gallup + elecyear + house_polar + HostCasRateLagged + AuthRank + Source + Inst_Cred_Gallup*house_polar,
    family=binomial(link="logit"),
    data=reg_df_slim
    )
summary(mod3_anyviol_house_int_lagged)
  
margins(mod3_anyviol_house_int_lagged, at = list(house_polar = seq(30,110, by=5)))
summary(margins(mod3_anyviol_house_int_lagged, variables = "Inst_Cred_Gallup", at = list(house_polar = seq(30,110, by=5))))

margins(mod3_anyviol_house_int_lagged, at = list(Inst_Cred_Gallup = seq(20,95, by=5)))
summary(margins(mod3_anyviol_house_int_lagged, variables = "house_polar", at = list(Inst_Cred_Gallup = seq(20,95, by=5))))

```

\singlespace
```{r int-terms-ch3, include=TRUE, echo=FALSE, results="asis", warning=FALSE, message=FALSE}
stargazer(mod1_num_ann_pubs_house_lagged_int, mod1_prop_ann_violate_sen_lagged_int, mod2_anyviol_house_lagged_int, mod2_anyviol_sen_lagged_int, mod3_anyviol_house_mixed_lagged_int, mod3_anyviol_sen_robust_lagged_int,
          type="latex",
          label="tab:regression5",
          title="Checking for Impacts of Interacting Prestige with Polarization", 
          covariate.labels = c('Trust in Military', 'Election Year', 'House Polarization',  'Senate Polarization', 'Hostile Casualty Rate (lagged)', 'Author Rank', 'Washington Post', 'Wall Street Journal', 'Prestige*Polarization (House)', 'Prestige*Polarization(Senate)'),
          dep.var.labels = c("Num Annual Pubs", 'Proportion Violating', 'Log Odds of Violating Any Civ-Mil Principles'),
          model.names=TRUE,
          header=FALSE,
          column.sep.width = "1pt",
          align=TRUE,
          font.size="small",
          no.space=TRUE,
          omit.stat=c("f", "ser"),
          float.env="sidewaystable"
          )
```

\doublespace

Table \@ref(tab:regression5) shows that when we interact trust in the military with the respective measurements of congressional polarization, the interaction terms are inconsistent in terms of direction and only statistically significant in three of the six models (models 4, 5, and 6). In model 4, the interaction term is statistically significant, but so are the coefficients for both trust in the military and polarization. Both of these are positive, while the interaction term is negative. So it becomes somewhat difficult to really isolate the potential impact of any sort of interaction between the two main independent variables of interest in model 4. Model 5 yields similar difficulties, but in the opposite manner: both of the independent variables by themselves are negative and statistically significant, while the interaction term is minimally positive but statistically significant. The results of model 6 are similar to model 4 in that the interaction term is negative and statistically significant, while the separate coefficients for the variables of military prestige (trust in the military) and polarization are each positive and statistically significant. 

In short, there are really no consistent patterns that emerge when an interaction term between polarization and military prestige are included that merit noteworthy attention. 

## Endogeneity Checks

The last major item of interest in this appendix include a series of checks for endogeneity. In line with the theory developed in chapter two, the most potentially problematic relationship involves instances of political behavior that violate the principles of civil-military relations and the level of trust in the military (military prestige). We might be particular concerned that acts of political behavior, in this case, the writing of opinion commentary by retired military officers, somehow drives the level of trust the public places in the military in a meaningful way. 

Table \@ref(tab:regression6) presents the results of several OLS regressions in which the dependent variable is the level of trust the public places in the military. I include several control variables, including dichotomous indicators of instances of commentary that violate one or more of the central principles of civil-military relations (models 1-2), commentary that violates the principle of non-partisanship in particular (model 3), as well as an interaction term that captures the combined effect of polarization and an instance of retired officer commentary violating any of the central principles of civil-military relations or the principle of non-partisanship in particular (models 2 and 3, respectively).  

```{r ch3-endog-test, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
## Endogeneity Checks
# Regress military prestige on crit_part_interfere and polarization

mod4_endog_1 <- lm(Inst_Cred_Gallup ~ any_CMR_viol + senate_polar + elecyear + HostCasRateLagged, data=reg_df)
summary(mod4_endog_1)

# Now include an interaction term for violation * polarization

mod4_endog_int <- lm(Inst_Cred_Gallup ~ any_CMR_viol + senate_polar + elecyear + HostCasRateLagged + any_CMR_viol*senate_polar, data=reg_df)
summary(mod4_endog_int)

mod4_endog_int_nonpart <- lm(Inst_Cred_Gallup ~ NonPartViol + house_polar + elecyear + HostCasRateLagged + NonPartViol*house_polar, data=reg_df)
summary(mod4_endog_int_nonpart)
```

\singlespace
```{r Regression Endogeneity Checks, include=TRUE, echo=FALSE, results="asis", warning=FALSE, message=FALSE}
stargazer(mod4_endog_1, mod4_endog_int, mod4_endog_int_nonpart,
          type="latex",
          label="tab:regression6",
          title="Regression Results: Endogeneity Checks (Regress Military Prestige on Polarization and an Instance of Retired Military Actor Opinion Commentary Breaking Any Central Principle / Non-Partisanship in Particular)", 
          covariate.labels = c('Instance of Violation of Any Central Principle', 'Senate Polarization', 'Violation of Non-Partisanship Principle', 'House Polarization', 'Election Year', 'Hostile Casualty Rate (lagged)', 'Violation of Central Principle * Polarization (Senate)', 'Violation of Non-Partisanship Principle * Polarization (House)'),
          dep.var.labels = c('Military Prestige'),
          model.names=TRUE,
          header=FALSE,
          column.sep.width = "1pt",
          align=TRUE,
          font.size="small",
          no.space=TRUE,
          omit.stat=c("f", "ser"),
          float.env="sidewaystable"
          )
```
\doublespace

The results of Table \@ref(tab:regression6) indicate that there is not a strong and statistically significant relationship between instances of violations to the principles of civil-military relations and the level of trust the public places in the military. The coefficients for instance of violation of any central principle (models 1 and 2) and violation of the principle of non-partisanship (model 3) are all statistically insignificant. The coefficients on the interaction terms in models 2 and 3 are likewise statistically insignificant. Therefore, at least with respect to the behavior of authoring opinion commentary, endogeneity does not appear to pose a significant problem. The level of trust the public places in the military is not the result of instances of opinion commentary violating the central principles of civil-military relations. 

## Acknowlegements

The author would like to thank participants from the UW-Madison International Relations Colloquium, held on 2 March 2021, for feedback on an early version of this project, which later formed the basis for this chapter of the dissertation. The author would also like to especially thank fellow US-Madison graduate student and cohort member Laura Felone, who agreed to do a "second reading" of all of the op-eds in this data set in order to verify the coding scheme used. 

All regression tables used in chapter three were produced with the help of the Stargazer package, designed by Marek Hlavac [@hlavac_stargazer_2018].

\pagebreak





