---
title: "Chapter 1 (Introduction): The Governing Standards or Central Principles of US Civil-Military Relations"
author: "Peter M. Erickson"
date: |
  `r format(Sys.time(), '%B %d, %Y')`
abstract: |
  
  _This chapter introduces the research question to be explored throughout this dissertation. Using Samuel Huntington's 1957 book, The Soldier and the State, as well as critiques of his work, this chapter argues that US civil-military relations have been bounded and influenced by the adherence by both civilian and military actors to three central principles or governing standards, including that of civilian control of the military, the non-partisanship of the military institution, and a third principle which I coin the "non-interference" of the military into certain realms of state policy, which describes the degree to which the military weighs in on issues that are indirectly or altogether unrelated to national security policy. After describing each of these three principles in detail and the gap in the scholarly literature this dissertation seeks to fill, this chapter then presents the broad plan of the dissertation. Ultimately, this dissertation seeks to provide a unique perspective regarding the influence of a state's domestic characteristics - namely, the level of political polarization and military prestige present in society - on instances of political behaviors committed by the military institution and military actors._  
  
geometry: margin = 1.15in              # margin widths
fontsize: 12pt                         # text size
bibliography: dissertation.bib         # bibliography file
biblio-style: authoryear               # citation style
urlcolor: blue                         # links to external URLs
citecolor: black                       # links to bibliography
linkcolor: magenta                     # internal figures, footnotes, etc.
indent: true                          # paragraph format
                                       # true for indented paragraphs
                                       
header-includes:
    - \usepackage{setspace}
    - \usepackage{multirow}
    - \usepackage{indentfirst}
   
    
output: 
  bookdown::pdf_document2:
    keep_tex: true
    latex_engine: pdflatex
    number_sections: false
    toc: false
    fig_caption: true
    citation_package: biblatex
    includes: 
      in_header: 
        - latex-preamble.tex
  bookdown::htmldocument2: default
  bookdown::word_document2:
      toc: false
---

\doublespacing 

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo=FALSE,
  include=FALSE,
  message=FALSE,
  error=FALSE,
  fig.height=3,
  fig.width=4,
  out.width="90%")
```

```{r, include = FALSE, echo = FALSE}
library("here")       # file paths
library("tidyverse")  # workhorse package
library("tidylog")
library("kableExtra")
library("knitr")
library("ggdag")
library("dagitty")
library("gridExtra")
library("broom")
library("dplyr")
library("reshape2")
library("pacman")
library("ggrepel")
```

\newpage

# Introduction and Research Question

Scholars, civilians, diplomats, and military leaders regularly discuss why and how militaries intervene in politics. These discussions, moreover, often include the theoretical question of whether military actors _should_ intervene in politics, to what degree, and for what purpose. Especially in the United States context, the political activity of the military is a subject that has undergone intense scrutiny in recent years. 

In the United States, rising political polarization over the past several years has thrust the military institution into the political spotlight - for better or for worse - on a number of occasions. Though it would be impossible to list all of the contemporary instances in which the military has become entangled in politics, several stand out. From a now-retired Army Lieutenant Colonel, Alexander Vindman, testifying before Congress during impeachment proceedings against former President Trump [@stolberg_meet_2019]; to the removal of Navy Captain Brent Crozier for leaking a letter to the press in which he castigated the chain of command's response to the onset of the COVID-19 pandemic [@associated_press_teddy_2020]; to a remarkable panoply of senior retired military officers excoriating President Trump's threat to use active duty forces to quiet protests and riots in cities across America during the summer of 2020 [@brooks_let_2020-1], instances in recent years abound which illustrate the variety of domestic challenges that the US military and its members face during periods of heightened political polarization. 

A significant amount of previous scholarship within civil-military relations has explored important facets of the ways in which militaries can and do engage in political behaviors that frustrate their civilian bosses [@brooks_militaries_2009; @snider_strategic_2014; @brooks_subverting_2021]. Other scholars have explored the various ways in which polarized politics make particular features of civil-military relations more difficult, or at least, potentially problematic [@robinson_danger_2018; @robinson_dissents_2020]. In this dissertation, I seek to advance both sets of scholarship by systematically exploring the impacts of two important structural domestic factors on civil-military relations in general, and more specifically, their impacts on several distinct forms of political behavior that civilian and military actors commit which involve the military institution. These two structural factors are the degree of _political polarization_ prevalent in society, and the degree to which the military institution is _popular_ or _prestigious_. In the United States, both of these factors are at relatively high levels, and will likely remain high for the foreseeable future. Therefore, it is vital that scholars, civilian leaders, and military officers alike understand how and why these structural factors shape civil-military relations in general, and in particular, the forms and types of behaviors that both sets of actors, civilian and military, are likely to commit regarding the military, when these structural factors exist at various levels. 

The central argument made throughout this dissertation is that the levels of polarization and military prestige in American society shape the ability of fundamental principles or standards of civil-military relations to constrain the behavior of both civilian and military actors regarding the military institution. Said differently, both the degree of polarization within a society and the prestige of the military alter the willingness of civilian and military actors to abide by fundamental principles and standards which, from a normative perspective, guide civil-military relations in the US context. These principles and standards are (1) civilian control of the military; (2) non-partisanship of the military, and (3) the non-interference of the military into certain realms of state policy or policy making. ^[These principles or standards are defined in great detail in Chapter 2 of this dissertation.]

In making this argument, furthermore, I build upon a general "motives" and "opportunity" framework first advanced by Finer (1962) and later by Taylor (2003) to describe military intervention in politics, and claim that in the American context, the relative levels of polarization and prestige serve the same functions [@finer_man_1962; @taylor_politics_2003]. By conducting a systematic exploration of the impacts of polarization and military prestige on civil-military relations, across time and several domestic contexts, this dissertation sheds light on what is, I argue, an under-appreciated task: the changing difficulty that civilian and military leaders face in harnessing, managing, and orchestrating the domestic behavior of the military as an institution. Far from being the hierarchical and monolithic actor most people think that it is, even the US military, especially in periods of high polarization, can be remarkably divided regarding its normative role in politics and therefore, how it should act.

Examining the impacts of polarization and military prestige on political behaviors that involve the military is an important undertaking from both normative and empirical perspectives. From a normative perspective, such a study helps observers and participants better understand why the conduct of civil-military relations can seem relatively smooth or, on the contrary, relatively turbulent, during different periods in history. Scholars often examine the particular leaders who are involved in particular moments or crises, such as wars, as well as the specific context surrounding the conduct of wars, to understand why civil-military relations were or were not troubled as the war proceeded [for just two of the countless possible examples, see @mcmaster_dereliction_1997 regarding the US during the Vietnam War; @kaiser_no_2014 regarding the US during World War Two]. And scholars are right to focus on these things. But I also contend that the story is incomplete if we do not give adequate and equal attention to the broader, long-running, structural and political factors that are examined in this dissertation, and their impact on civil-military relations. In this dissertation, I argue that these factors, especially political polarization and military prestige, shape the environment in profound ways in which civil-military relations occur.

There is much at stake empirically as well. By analyzing certain forms of behaviors over time, particularly as the conditions of polarization and military prestige change, scholars are better positioned to gain an accurate awareness for just how significant the roles of political polarization and military prestige are in shaping civil-military relations, and in particular, the behaviors of both military and civilian leaders. It is one thing to acknowledge that high polarization and high military prestige make the conduct of civil-military relations more challenging, but another to demonstrate just how much more challenging, and with respect to what behaviors. These normative and empirical perspectives compliment each other.

This introductory chapter of the dissertation proceeds in three main parts. First, it introduces the general field of civil-military relations. In doing so, emphasis is placed on identifying and expositing three central and fundamental standards or principles of American civil-military relations: civilian control of the military, the non-partisanship of the military institution, and the non-interference of the military into certain realms of state policy making. The second part of the chapter describes how the intended dissertation topic fits within the scholarly literature and the gaps it intends to fill. The third and final part of this introductory chapter describes the general methodology and the plan of the dissertation.   

# The Field of Civil-Military Relations

Truly one of the interdisciplinary topics of scholarly research, civil-military relations spans the fields of political science, history, and sociology. The topic of civil-military relations, moreover, involve a set of dynamic questions and issue areas, including the relationship between militaries and governments, the relationships between militaries and societies, strategic and military effectiveness, the occurrence of coups, and a host of other pivotal topics. Within the field of political science in particular, one of the areas of civil-military relations that is most studied is the relationship between prominent state officials and military leaders. Indeed, the relationship between these actors is what one scholar has termed, "leadership at the state's apex" [@brooks_risa_integrating_2019, 380], and it is here that numerous weighty decisions that impact national security are made, including whether to go to war, how to form and implement effective strategy, how to prepare for the next conflict, and how large the military and its budget should be. 

In the most basic sense, the study of civil-military relations explores aspects of the question of how to ensure that a military is strong enough to defeat threats to the state without threatening the state itself. One scholar of civil-military relations, Peter Feaver, refers to this central question of civil-military relations as the "problematique," and notes that this dilemma is both "vexing" and a great "paradox" in that the military must be "strong enough to prevail in war," yet submissive to authority such that it cannot and will not "destroy or prey on the society it is intended to protect" [@feaver_civil-military_1996; @feaver_civil-military_1999, 214]. As a result of this "paradox" [@feaver_civil-military_1996] - a state needs a military, and yet, one that is powerful and dangerous - that civil-military relations are, at least to some degree, fraught with tension that is, while perhaps capable of attenuation, never entirely absent.

A number of scholars have offered various solutions and frameworks for solving the "problematique" Feaver describes, but none has been more lasting or consequential than the late Samuel Huntington, whose 1957 work, _The Soldier and the State: The Theory and Politics of Civil-Military Relations_, often serves as the point of departure for a number of debates within the field of civil-military relations. Although Huntington's theory has endured important and substantive critiques over the years [for example, see @janowitz_professional_1960; @finer_man_1962; @cohen_supreme_2003; @feaver_armed_2003; @brooks_paradoxes_2020], it is essential to briefly exposit Huntington's conception for achieving effective civil-military relations, and what lies at the heart of it - a theory of civilian control of the military that maximizes what Huntington called "military professionalism" [@huntington_soldier_1957]. 

In the section to follow, I briefly summarize Huntington's argument and logic, as well as several prominent critiques of his work. The purpose of summarizing both Huntington and his critics is to argue that in spite of the important differences between them, Huntington and his critics share much common ground regarding the aspirational prominence of three central principles, or standards, that help foster healthy and effective civil-military relations. These principles, which I articulate and describe in the section to follow, are the principles of civilian control of the military, the "non-partisanship" of the military institution, and what I will term the "non-interference" of the military into realms of politics that are the _prerogative_ of duly-elected and appointed civilian leaders. These related principles, each of which is the subject of a fair bit of scholarship, are distinct and hold different normative implications for behaviors committed by both civilian and military actors with respect to the military institution.  

## Huntington and The Soldier and the State

Though it is not always framed as such, _The Soldier and the State_ is a book that lies at the very intersection of domestic politics and international relations. The fundamental problem that Huntington sets out to solve is how liberal democracies such as the United States can balance the simultaneous demands of military security and the ideology of political liberalism. Writing after the Korean War, Huntington argues that American security requirements constitute a "functional imperative" that require the US to maintain a military that is sufficiently strong and capable of defeating the external threats of the Cold War [@huntington_soldier_1957, 1-3]. At the same time, Huntington argues that the liberal character of American society generate a "societal imperative" such that the pursuit of American security requirements could not cause the US to deviate from its liberal political ideology in fundamental ways, including abandonment of the principle of civilian control of the military [@huntington_soldier_1957, 1-3]. Like the "problematique" pointed out by Feaver, Huntington aims to solve how to render the military strong enough to protect the United States, but sufficiently subservient and obedient to civilian power [@feaver_civil-military_1996]. 

Huntington argues, importantly, that striking this balance is made more difficult by the fact that the military officer corps is different from the rest of American society in several important ways, a claim that is not without controversy. In particular, Huntington claims that those who serve in the military hold to a distinct "mind" and "ethic" that views conflict as part of a "universal pattern throughout nature," and thus, that "the military view of man is...decidedly pessimistic" [@huntington_soldier_1957, 62-63]. Huntington then juxtaposes the mindset and ethic of the US military officer corps with tenets of American liberal ideology, noting that American liberalism, at least how he conceives of it, stresses "individualism" and that liberalism "either denies the existence of power, minimizes its importance, or castigates it as inherently evil" [@huntington_soldier_1957, 90]. It is, Huntington argues, this contrast between the unique character and ethic of militaries and the attributes of American political liberalism that explains why the US has been suspicious of standing military forces since its founding [@huntington_soldier_1957, 143-151]. 

Although Huntington remains agnostic while tracing the general tension between a unique military mindset and the facets of American liberal ideology, he then takes a clear stand as he presents his main argument. Huntington argues that the Cold War poses a unique problem in terms of the security requirements facing the US. To overcome these security requirements, sizable standing forces are a must. However, the US public tends to view such arrangements unfavorably. In order to overcome the suspicion of the US public in building and maintaining a large military machine, Huntington offers a solution whereby the United States adopts and implements a system of civil-military relations which Huntington terms "objective control" of the military [@huntington_soldier_1957, 83]. Under such an arrangement, civilian leaders grant the military a high degree of professional autonomy, which Huntington claims will enable the military to develop and hone its professional expertise and skills, while also discouraging military leaders from entering partisan politics. 

"Objective civilian control" of the military, in other words, involves civilian and military actors both recognizing that each actor owns a distinct sphere of figurative territory that is to remain separate from that owned by the other [@huntington_soldier_1957, 83-85]. Civilian leaders are not to encroach into areas that are the military's purview, which would damage military autonomy, and conversely, military leaders should not enter the world of partisan politics, as doing so subverts civilian control and the apolitical nature of the military. 

Huntington further claims that the reason his idea can and will work is the rise of the military as a profession. This is, in fact, a vital link in Huntington's logic. Distinct from other vocations or jobs, professions possess several unique attributes, including "expertise, responsibility, and corporateness" [@huntington_soldier_1957, 8-10]. Huntington is essentially making the case that the military officer corps, like those who practice medicine, lawyers, and clergy members, a professional body.

With respect to the attribute of expertise, Huntington has in mind the idea that military professionals possess unique "knowledge and skills" in the arena of warfare [@huntington_soldier_1957, 8-10] and, using Harold Lasswell's terms, are experts in the "management of violence" [Lasswell quoted in @huntington_soldier_1957, 11]. Regarding the attribute of responsibility, Huntington has in mind the idea that the military officer corps is to employ such expertise only in the service and "at the direction" of the state [@nielsen_american_2012, 370]. And with respect to the attribute of corporateness, Huntington sees the military officer corps as a group that certifies its own leaders and polices itself, or in the words of Nielsen, "as a distinct, bureaucratized body, with a common identity fostered through shared educational, training, and service experiences" [@nielsen_american_2012, 370].

The professional nature of the military, so claims Huntington, would guide the military's relationship to politics. In essence, by instituting Huntington's conception of _objective control_, the military would be better postured, enabled, and equipped to further its professional autonomy, which, as a profession, it would greatly desire to do. Similarly, such an arrangement would please politicians, as it would lead to a military that does not interfere in politics yet one that is also exceptionally skilled in completing its tasks, namely, defeating external threats on the battlefield. Ultimately, the posture or arrangement of _objective control_, so claimed Huntington, would lead to a civil-military relationship that simultaneously strengthens and preserves the principle of civilian control of the military and defends the nation [@huntington_soldier_1957, 83-85].

Finally, it is important to emphasize that Huntington's argument works to the extent that one agrees and accepts that the US military, and its officer corps in particular, constitutes a profession. Among contemporary scholars of civil-military relations, this point is not significantly disputed, even as they warn, rightly in my view, that to remain a relevant profession is never an automatic guarantee, but rather a perpetual challenge that requires deliberate and perpetual effort on the part of both soldiers and civilians [for one scholar's warning about this dynamic, see @snider_once_2012]. 

Scholars posit slightly different timelines regarding the US military becoming a profession. Weigley (1977), for instance, asserts that the first signs of the US military becoming a profession were evident prior to the Civil War [@weigley_american_1977], while Heiss (2012) pinpoints The War of 1812 as the catalyst of the US military starting to professionalize [@heiss_professionalization_2012].^[For an excellent book that explores the process of the US Navy becoming a profession, see @mobley_progressives_2018.] Huntington, on the other hand, takes a slightly different view, and generally argues that the military became a profession in earnest over the course of a somewhat lengthy process that began during the post-US Civil War years [@huntington_soldier_1957, 222-269]. For the purposes of this dissertation and the analysis contained herein, the exact year in which the US military became a profession is not as important as recognizing, as Huntington argues at length for, that the professional nature of the US military is both what makes it and its members distinct from the rest of society.

## Critics of Huntington

Huntington's work, and specifically, his concept of "objective civilian control" of the military has generated a number of important and meaningful critiques over the years, and it would be impossible to list all of these critiques here. However, a major theme of these critiques is that the type of stark separation between military and political spheres, which Huntington argued for, is unwise, impractical, and potentially damaging to US security. In levying such a critique, Huntington's critics often point to and assert Clausewitz's famous dictum that war is a "mere continuation of policy by other means" [@clausewitz_war_1976, 69, 605]. Thus, argue the critics of Huntington, the military is, at least to some extent, an inherently political institution that should not, indeed cannot, conduct military operations entirely absent of political influence. 

In making this critique, the critics of Huntington are also leveraging broader democratic theory. Critics of Huntington rightly assert that an extreme divorcing of the duties and responsibilities of statesmen and soldiers runs counter to the notion that in democracies, civilians are superior to military leaders, and thus, hold ultimate responsibility for what militaries do and fail to do. According to such logic, critics of Huntington, such as Eliot Cohen, claim that civilian leaders should not abdicate, under any circumstances, the ultimate responsibilities of statesmanship, which includes at times, decisions regarding whether to go to, and the conduct of, war [@cohen_supreme_2003]. Peter Feaver has parsimoniously captured the essence of this critique of Huntington by noting that the principle of civilian supremacy over the military means that "civilians have the right to be wrong" [@feaver_right_2011, 117; @feaver_armed_2003]. 

Other critiques of Huntington have further developed the possible domestic and international consequences that arise if military officers strictly adopt a Huntingtonian concept of professionalism that starkly separates civilian and military spheres of influence. For instance, critics warn that doing so can lead to military officers failing to recognize the inherent political nature of their roles, which can subsequently lead to military officers unknowingly engaging in political behaviors under the guise of acting in an apolitical manner [@brooks_paradoxes_2020; see also @janowitz_professional_1960; @feaver_civil-military_1996]. Others warn that a serious consequence of military officers adhering to a strict separation between military and civilian spheres in wartime is the failure to develop, implement, and achieve sound objectives and strategy, which leads to disastrous effects on overall US national security policy [@rapp_civil-military_2015].   

In spite of these insightful and important concerns and critiques of Huntington, Huntington retains an out-sized influence in the field of civil-military relations, even today. Most critiques of Huntington simultaneously evince a respect for both the author and his ideas. As Nielsen has argued, any reputable course taught on civil-military relations likely includes Huntington's _The Soldier and the State_, not only because of the "boldness and ambition" of Huntington's work, but also because Huntington was truly the first serious scholar who sought to develop a comprehensive theory to help steer the conduct of civil-military relations [@nielsen_american_2012, 369]. 

Moreover, it is my contention that the critics of Huntington share much in common with Huntington, particularly concerning the broader aims of Huntingtonian thought. Indeed, I contend that Huntington and his critics agree that at least three central principles or standards should govern the relationship between the military and civilian leaders, even if these critics may disagree (and they do) on the specific ways in which these principles or standards should be applied, achieved, developed, and maintained. These three central principles include that of civilian control of the military, the non-partisanship of the military institution, and a principle of "non-interference" of the military into certain areas or realms of state policy making. 

The most obvious principle upon which Huntington and his critics both agree is that of civilian control of the military. Indeed, this is one of the two chief purposes of Huntington's formulation of the concept of _objective control_ (the other being the ability of the military to defeat external threats). Yet maintaining civilian control of the military is likewise the chief concern of Huntington's critics. For example, the scholar Eliot Cohen, himself a critic of much of Huntingtonian thought [see @cohen_supreme_2003, 242-250, 254-255], proffered a descriptive theory of civil-military relations that he has termed "the unequal dialogue" [@cohen_supreme_2003, 208-224]. 

In developing his theory, however, Cohen argues strongly that the most successful wartime civil-military relations have been those that have involved "an unequal dialogue - a dialogue, in that both sides expressed their views bluntly, indeed, sometimes offensively, and not once but repeatedly - and unequal, in that the final authority of the civilian leader was unambiguous and unquestioned..."[@cohen_supreme_2003, 209]. Cohen, as is the case with other scholars of civil-military relations who have critiqued Huntington, is not concerned that Huntington's concept of objective control awards civilians too much power, but rather the opposite - that it prevents civilians from appropriately exercising appropriate oversight of the military. The key point here is that both Huntington and his critics agree that maintenance of the principle of civilian control is a paramount concern.   

Huntington and his critics likewise share a concern that the military stay out of, or least minimize its impact on, the partisan world of politics. Nielsen argues, for instance, that "Huntington's principle of objective control has both merits and shortcomings. On the positive side, it preserves democratic control, speaks to the importance of an apolitical military and protects military professionalism" [@nielsen_american_2012, 375]. And Cohen warns of an outright dismissal of Huntingtonian thought, warning that "to reject Huntington's ideas of sequestering issues of policy from those of military administration or operations is to open the way to a military that is politicized and, by virtue of its size and discipline, a potentially dominant actor in the conduct of foreign and internal affairs" [@cohen_supreme_2003, 264]. Huntington and his critics disagree on exactly how stark the separation between military and political policy should be, but indeed, both share a concern that the military avoid inappropriate partisan political involvement. 

Finally, I argue that both Huntington and his critics agree that the military should not interfere or influence all realms of state policy making. To be clear, Huntington and his critics _do not_ agree that the stark separation between the world of military policy and politics proposed by Huntington's concept of civilian control should exist. Nonetheless, it is vital to understand that critiques of Huntington on this concept are one directional in that they are concerned that civilians do not wield sufficient influence on the military. I am arguing here that critiques of Huntington on this point are never made from the opposite direction; that is, critics of Huntington do not argue in favor of the military holding greater sway and influence into the realm of domestic politics, and particularly, those areas of domestic politics that are indirectly related or entirely unrelated to military affairs. Here again, Cohen is helpful when he warns of dismissing outright Huntington's ideas concerning objective control of the military. Cohen warns that a full dismissal of Huntingtonian theory on this point, especially "in states with less-established democratic traditions _(than the United States)_...would open the path to direct military intervention in politics" [@cohen_supreme_2003, 264, emphasis mine].

In summary, there is much that Samuel Huntington and his critics disagree about concerning civil-military relations. Huntington advocated for separating the spheres of military operations and politics as much as possible on the basis that the military profession would be interested in pursuing its own autonomy, which included further developing its unique and special military expertise. His critics, on the other hand, have pointed out that there are limits to how separate military affairs and politics can reside, both in theory and in practice, and that therefore, the notion of military expertise and the concept of a military as a profession, while they exist, never exist divorced from political context. In spite of these critiques, Huntington's theory of civil-military holds tremendous staying power in both academic circles and within the military, even today. Moreover, I have demonstrated that at least in very broad terms, Huntington and his critics share a concern for at least three overarching principles of civil-military relations. The first is a principle of civilian control of the military. The second is a principle that the military, even if it is a political actor, should avoid partisan influence and entanglement as much as possible. And the third principle is that of the military not interfering in realms of state policy making that are indirectly related, or entirely unrelated, to the conduct of military affairs.  

# Three Central Principles of Civil-Military Relations

In the section below, I define and exposit in greater detail each of the three central standards or principles of civil-military relations. As is often the case in academic scholarship and particularly within the study of civil-military relations, these standards and principles are generally conceptualized and defined not in a dichotomous, black or white fashion, but rather along a gradual continuum, or scale. Thus, ample room exists for scholars of civil-military relations to debate the precise meaning of these principles, or similarly, the extent to which the absence or presence of certain indicators constitute an affront to one or more of these principles. Such nuance is all part of what makes the study of civil-military relations so vibrant.  

For the purposes of this dissertation, I claim that these three principles - that of civilian control of the military, the non-partisanship of the military institution, and the "non-interference" of the military into certain areas of state policy making - constitute a _baseline_ set of principles that can inform, guide, and facilitate healthy and harmonious civil-military relations. I further claim that when these three principles are upheld and respected by both civilian and military sets of actors, civilian and military leaders are less likely to engage in certain forms of behavior that run afoul of these principles. Ultimately, this matters and should concern scholars and practitioners of civil-military relations because when when one or more of these central principles is shunned, bucked, disregarded or blatantly violated by one or both actors, then a state's civil-military relationship is likely to be discordant in some way.  

## Central Principle 1: Civilian Control of the Military

The first central principle of civil-military relationships is that of civilian control of the military. It is listed first deliberately, as it is the central component by which the health of civil-military relationships, particularly in democracies, is maintained. Without strong adherence to the principle of civilian control of the military, it is doubtful that a democratic state's civil-military relationship can be healthy nor successful in the long run. So, what is civilian control of the military, and how does one know if military and civilian actors within a state are adhering to it as a principle of civil-military relations? 

Scholars have not agreed upon a precise definition of the concept of civilian control of the military, as it is a massive and encompassing concept. Scholars share the idea that civilian control generally refers to the "relative political power" that exists between a nation's armed forces and its civilian leaders [@bruneau_civil-military_2019, 7; see also @brooks_shaping_2008]. As such, modern conceptions of civilian control are not dichotomous, but rather exist along a scale that, at least theoretically, can range from instances of "full civilian control to complete military dominance over the system" [@bruneau_civil-military_2019, 7; @fitch_armed_1998]. 

Contemporary scholars agree that serious inquiries of civil-military relations must look beyond the presence of coups to determine the health of the principle of civilian control [@croissant_beyond_2010; see also @cohen_supreme_2003, 242; @beliakova_erosion_2021]. Feaver suggests that observers should expect such nuance and thus examine the "patterns" of civilian control that exist within a state, and not merely look for the presence of it [@feaver_civil-military_1996, 167]. Brooks, Golby, and Urben distill all of these ideas further, noting that in the US context, the concept of civilian control is "about the extent to which political leaders can realize the goals the American people elected them to accomplish" [@brooks_crisis_2021, 65]. They note, correctly in my view, that senior military leaders can exercise degrees of discretion with regard to the information that they share with civilians, the speed with which they implement civilian directives, and the options that they provide to civilians in the first place [@brooks_crisis_2021]. 

The principle of civilian control of the military, therefore, is about much more than simply "overt insubordination" of the military [@brooks_crisis_2021, 64]. It rather has much more to deal with the quality and the extent of the processes in which the military and a nation's political leadership participate together to both develop and implement military plans and strategy that furthers the policy goals of a nation's elected civilian officials. Coups are certainly the most glaringly visible manifestation of the absence of the principle of civilian control, but in democracies such as the United States, where coups are (thankfully) an uncommon occurrence, we need to look carefully for other indicators of the principle of civilian control failing to be fully implemented. 

In one such study that examines the many ways and degrees in which militaries violate civilian control, Beliakova (2021) notes that there are many "forms of erosion" of the principle of civilian control [@beliakova_erosion_2021]. She identifies these forms as "insubordination," in which the military can flat out refuse to obey the government; "deference," in which the government increasingly delegates powers to the military; and "competition," in which the military directly challenges the government for power [@beliakova_erosion_2021, 1400]. Beliakova and other scholars have, in other words, confirmed the warning made by Finer nearly 60 years ago regarding the large aperture that manifestations of violation of the principle of civilian control can take. Indeed, in 1962, Finer simply warned that militaries can and will violate the principle of civilian control through "acts of commission, but also by acts of omission" [@finer_man_1962, 20].

In the United States, the principle of civilian control of the military receives some assistance in the fact that the US Constitution prescribes that civilian leaders control the military. Far from being a guarantee that the principle of civil-military relations is adhered to in healthy ways, these statutory guidelines nonetheless serve as a normative and sometimes a legal marker for civilian and military actors to aspire to. For instance, the US Constitution grants the President the power to serve as the Commander in Chief of the Armed Forces [@united_states_us_1787, Article 2, Section 2], yet also grants Congress unique and specific powers of oversight of the military [@united_states_us_1787-1, Article 1, Section 8]. 

Before moving on to address the second central principle of civil-military relations, it should briefly be mentioned that of the two major sets of actors civil-military relations, only the military can _violate_ the principle of civilian control of the military. To be clear, civilian actors certainly play a part and have a major responsibility to ensure that the principle of civilian control of the military is healthy; as such, civilians can fail in this regard, either through being derelict in their duties to enforce civilian control, or by failing to demand that the military behave in a manner that is consistent with the principle of civilian control. Nonetheless, speaking in the strictest possible sense, only military actors can expressly _violate_ the principle of civilian control. 

## Central Principle 2: Non-Partisanship of the Military Institution

The second central principle of democratic civil-military relations, particularly in the US context, is the principle of non-partisanship of the military institution. In general terms, the principle of non-partisanship of the military stipulates that the military not align itself - nor that it be made to align by civilians - with a particular political party or its platform. The principle of non-partisanship means in part that the military fully obeys the lawful orders of whichever political party is in office. Yet the principle of non-partisanship also means the military cannot and will not identify itself, nor be made to identify by civilian leaders, as a partisan actor. Thus, both sets of actors, civilian and military, have a responsibility to maintain this second standard of non-partisanship of the military. 

However, determining the degree to which civilian or military actors uphold the principle of non-partisanship in practice is a difficult undertaking, largely because it is difficult to both identify and measure what forms of behavior constitute a violation of this principle. After all, if militaries are, in fact, as Clausewitz and the critics of Huntington have contended, inherently political creatures who "serve at the pleasure" of their civilian bosses, then it stands to reason that the military will, at some point, enact the partisan policies, wishes, desires, and goals of their elected civilian leaders, who belong to a political party [@mullen_chairman_2011]. 

For this reason, scholars of civil-military relations have often separated the various meanings and implications of certain forms of behavior and their impact on politics. While there is no clear delineation of which forms of behavior and which contexts violate the principle of non-partisanship of the military, recent scholarship has attempted to create figurative space to allow greater acceptance - even encouragement - of military officers committing some forms of political behavior. For instance, scholars have argued that military officers and civilians should expect military leaders to increasingly contribute to debates about security policy on the grounds that doing so will lead to the better formulation of strategy and thus, to better national security outcomes [@brooks_perils_2013; @owens_military_2015; @rapp_civil-military_2015]. 

Simultaneously, scholars and military officers alike recognize that other forms of behavior, such as retired military officers speaking at political conventions on behalf of political candidates, comprise blatant violations of the military's non-partisan ethic [@dempsey_military_2016]. In other words, there is agreement that while there may be room for political behavior of the military, at some point such behavior can and does cross a figurative line such that the behavior is not merely political in nature, but partisan in nature. It is at this juncture that such behavior becomes problematic.  

Similar logic holds for civilian leaders with respect to the military as an institution. In short, civilian leaders, according to the principle of non-partisanship, should not use the military in strictly partisan ways. Admittedly, however, there is far less clarity for civilian actors concerning when such usage of the military becomes inappropriate, precisely because of the fact that civilian leaders are elected to fulfill partisan campaign promises! Several examples, nonetheless, are instructive. Many scholars and former military officers, former instances, blasted the decision of President Trump to deploy US Troops to the US southern border in 2018, citing the move as a political stunt ahead of the 2018 midterm elections [@adams_trumps_2018]. Similarly, other critics alleged that President Trump's threat to use active duty forces to dispel protesters and rioters in the summer of 2020 likewise violated the military's non-partisan ethic [@brooks_dismay_2020]. Whether these instances, and to what degree, actually posed violations of the military's non-partisan standard is answerable in terms of gradations and degrees, rather than stark dichotomous terms, and often serves as the subject of civil-military relations commentary.^[I do not provide an all-encompassing rubric regarding the extent to which certain behaviors are and are not more partisan than other behaviors in this dissertation. That said, I do defend in great detail, both in Chapter 2 and Chapter 4 of this dissertation, the decision to examine the use of military symbols and actors during Presidential campaign advertisements as a clear indication of how partisan the military, or certain military actors, appear in particular political races.]

In summary, the non-partisanship of the military is an aspirational standard of democratic civil-military relations, particularly in the United States. Furthermore, both sets of actors - military and civilian - have a responsibility to uphold the principle of non-partisanship, each by not aligning the military with partisan goals and aspirations. However, where the line between appropriate political involvement and inappropriate partisan involvement of the military is drawn is not always clear. This is especially true when examining the behavior of civilian leaders because they, by their very nature, are partisan entities. Still, the principle of non-partisanship, at least in general terms, applies: citizens should not be able to look at the military and to easily detect what their partisan preferences are. 

## Central Principle 3: The "Non-Interference" of the Military 

A third central principle or governing standard of American civil-military relations that both Huntington and his critics affirm is what I am terming the _"non-interference"_ of the military into certain realms, or areas, of state politics. The key aspects of the principle of non-interference involve the areas of state policy making that the military seeks to influence, and thus, by extension, the types of missions that the military conducts. In essence, the principle of non-interference states that there are policy arenas of the state which are and should remain "off limits" for US military actors. According to the critics of Huntington, civilian leaders may weigh in, influence, and ultimately oversee all things that involve military affairs. But the reverse notion, that military leaders should influence and have a say on all areas of civilian policy making, is an idea that both Huntington as well as his critics would find fundamentally flawed. 

It is important to point out that unlike the previous principles of civilian control and non-partisanship, the principle of "non-interference" of the military is not explicitly articulated in the American civil-military relations literature. This is unfortunate, and stems, in my view, from the fact that the principle of non-interference is largely subsumed and thus, strongly implied, by the principle of civilian control. Still, the origins of the principle of non-interference can be directly traced to Huntington's conception of military expertise. 

Recall that for Huntington, the concept of military expertise and military professionalism are inextricably linked, and involve the unique knowledge, skills, and behaviors that are relevant to the military officer corps [@huntington_soldier_1957, 7-18]. Huntington later devotes an entire chapter to the development of the notion of a unique and distinct "military mind," which to him consists of the "values, attitudes, and perspectives which inhere in the performance of the professional military function and which are deducible from the nature of that function" [@huntington_soldier_1957, 61]. In short, Huntington is arguing that the uniqueness and distinction of the military - in terms of its mindset, corporate ethic, and values - stems from the fact that the military's fundamental purpose is to defend the nation from external threats. The implication here is simple but profound: a professional military is one that is used only to fill roles that are directly related to national defense and/or the security of the nation. 

And US history, certainly in the 20th Century at least, proves that this has indeed been the case. The US military has been primarily used to fight external rather than internal threats. Thus, there has been little if any need for the US military officer corps to assume any type of political role in managing the state's domestic problems. To be clear, the American military has certainly been used for a variety of non-traditional types of operations in the nation's history, and actually, quite often, from desegregating schools, responding to natural disasters, and most recently, assisting Federal authorities in responding to the Coronavirus Pandemic. But these types of missions, while frequent, have not come anywhere near to comprising the bulk of the US military's efforts.  

It is also important to note that the literature within comparative civil-military relations very much describes the principle of non-interference, even if in a descriptive way to denote that in many countries, such a principle does not exist! Indeed, militaries in many parts of the world influence and sometimes direct policy areas of the state to such an extent that doing so in the US context would be unthinkable. This scholarship includes explorations of the varying mission sets that militaries around the world conduct [for instance, see @vennesson_is_2009; @pion-berlin_democratization_2005; @bruneau_civil-military_2019], and the degree to which certain militaries are used to conduct internal state security missions such as domestic counter-terrorism vice external warfare [@stepan_new_1973; @pion-berlin_civil-military_2020]. For instance, Stepan's concept of "the new professionalism of internal security and national development" traces the development of the Brazilian military's role expansion into domestic affairs as a result of having to primarily confront internal rather than external threats [@stepan_new_1973].

But especially in democratic countries, a principle of "non-interference" of the military seems to firmly established. This is true even as the exact boundaries of separation regarding which policy areas are off limits for military influence are not the same across multiple democratic countries. In his excellent work on Israeli civil-military relations, Yehuda Ben-Meir (1995) separates the affairs of the state into four broad areas, including political affairs, domestic affairs, national security, and the armed forces [@ben-meir_civil-military_1995]. He argues that civilians should and do influence all four of these areas of politics, whereas the military should influence the three areas of domestic affairs, national security, and the armed forces [@ben-meir_civil-military_1995].^[In terms of the activities which constitute political affairs, Ben-Meir includes items such as "taking control of the government (coups), influencing political appointments, or interfering in the decision making process. See @ben-meir_civil-military_1995, 4-5 for an excellent description and diagram.] Ben-Meir is addressing the Israeli context, and thus, his paradigm may not map squarely onto the US context. This is not problematic, however. What is important for the purposes of my argument here is to highlight and to acknowledge, as Ben-Meir has, that the relative scope and specific areas of state policy that are influenced by military leaders is not the same as that of civilians.

Finally, it is important to briefly address what the principle of non-interference of the military implies for the conduct of the US military. The principle of non-interference implies that the military will not seek to perform, and that civilians will not assign the military to perform, roles for which military forces are not suitably designed, nor those which are tangentially related to national defense, except in cases of great crisis or need. Of course, there is a greater debate concerning what policy areas do in fact relate to national defense, and to what degree. Still, the principle of non-interference remains. When the Chairman of the Joint Chiefs of Staff, General Mark Milley, foresaw no role for the military in the 2020 Presidential Election, he was implicitly invoking the principle of "non-interference" of the military [@silva_gen_2020]. Other instances also abound. For example, the US military's role in the COVID-19 crisis, and in particular, in distributing vaccines, has not been without controversy.^[At one point in late 2020, an active duty four star general apologized to the nation for a mix-up in information regarding the distribution of COVID vaccine, leading some critics to express concern that the military was making inherently political decisions that, from a normative perspective, posed some problems. See @passy_gen_2020 for more details.]

## Summary of Three Central Principles of US Civil-Military Relations

In summary, Huntington's _The Soldier and the State_ has served for decades as a starting point for numerous debates involving civil-military relations. Despite the important critiques of Huntington's work, scholars of civil-military relations have generally agreed upon the importance of three foundational principles to guide the behavior of both military and civilian leaders. These three standards are that of civilian control of the military, the non-partisanship of the military institution, and the "non-interference" of the military into certain arenas of state policy. 

When adhered to, followed, and respected, these principles generally constrain the behavior of both civilian and military actors in important ways, and have helped to facilitate relatively harmonious civil-military relations. Therefore, these three principles constitute a _baseline_ set of principles which can be examined and analyzed in order to evaluate the health of American civil-military relations. Note here that the argument that is beginning to take shape is not that abiding by these principles will ensure that civil-military relations will result in the implementation of wise decisions. As Feaver has argued, harmonious civil-military relations "is not synonymous with good policy" [Peter Feaver in @schake_masters_2021]. At the same time, it is not an exaggeration to claim that the health of the environment in which civil-military relations occur and good policy are entirely unrelated. 

Table \@ref(tab:principles) captures these three principles, a concise definition, and the actor that can expressly violate each principle.  

```{r principles, include=TRUE, echo=FALSE}

text_tbl1 <- tribble(~"Governing Principle or Standard", ~"Definition/Description", ~"Potential Violating Actor",
  "Civilian Control of the Military", "Military and Civilian Leaders Engage in a Robust Dialogue and a Process Such That Civilian Political Goals are Fully Actualized and Implemented; Mechanisms of Civilian Oversight Function; No Overt Military Insubordination", "Military (civilian actor involvement is nonetheless vital to make this principle work)",
  "Non-Partisanship of the Military Institution",  "The Military Institution Exists and Operates outside of Partisan Politics; Military Actors Fully Obey Political Leaders, and Do Not Advocate for Partisan Policies, People, or Platforms", "Civilian or Military",
  "Non-Interference of the Military into Certain Realms of State Policy", "There are Areas or Realms of State Policy making into which the Military does not Enter or Seek to Influence", "Civilian or Military")
 

text_tbl1 %>% kable(booktabs=T, caption="Central Principles of US Civil-Military Relations") %>%
  kable_styling(full_width = F, font_size=10, latex_options="striped" ) %>% 
  column_spec(1, width="10em") %>% 
  column_spec(2, width="15em") %>% 
  column_spec(3, width="10em") 
  

```

# The Gap in the Scholarly Literature

In this section of this Chapter, I briefly review the literature that seeks to explain variation in the military's role in politics in general terms. I then describe how current scholarship has explored the consequences of political polarization and military prestige on US civil-military relations. Ultimately, I seek to merge these two strands of literature, and argue that doing so will help fill a needed gap in the scholarly US civil-military relations literature.  

Previous scholarship that explains variation in the military's role in politics is found in works of both civil-military relations theory as well as comparative civil-military relations. While there are dozens of relevant works that explain military intervention in politics in both general and specific terms [@finer_man_1962; @stepan_new_1973; @fitch_armed_1998; @taylor_politics_2003; @croissant_beyond_2010; @bove_beyond_2020; @beliakova_erosion_2021], I focus here on three works of scholars that primarily address _variation_ in how and why militaries intervene politically. These three include Samuel Finer's _The Man on Horseback_ (1957), J. Samuel Fitch's _The Armed Forces and Democracy in Latin America_ (1989), and Brian Taylor's _Politics and the Russian Army_ (2003). In the following paragraphs, I briefly describe these works and point out where, in my opinion, they fall short. 

Finer (1962) argues that militaries intervene in politics as a broad result of having the "motives and opportunities" to do so [@finer_man_1962, 20-76]. With respect to the motives that might spur a military to inappropriately intervene in politics, Finer argues that these vary, but could include a military's own bureaucratic interests as well as the belief or worldview that the military should serve as the "custodian" of the "national interest," particularly in times of national crisis [@finer_man_1962, 30-31]. Additionally, in what is an insightful critique of the Huntingtonian concept of military professionalism, Finer also warns that militaries might actually be compelled to intervene in politics precisely as a result of adopting a very strict conception of professionalism whereby it strictly adopts the idea of serving the state rather than the state's duly-elected civilian leaders [@finer_man_1962, 22-23]. In other words, Finer does not buy the Huntingtonian idea that military professionalism is sufficient to keep the military out of politics; instead, he warns that it might actually induce the military to become involved in politics! Finer's theory is comprehensive, but it falls short in that it does not identify with precision what the motives and opportunities that may exist in the modern United States. Still, Finer's analysis provides many positive clues, and in fact, I build off of his motives and opportunities framework as I develop my comprehensive theory in Chapter 2.  

Fitch explores why military officers of certain countries in Latin America in the late 1980s adopt one of several types of roles with respect to the civilian government. Based off of interview with a rich array of military officers from several Latin American countries, Fitch develops and explains the concept of "military role beliefs," which refers, in general terms, to how military officers conceptualize and view "their role in politics" [@fitch_armed_1998, 61]. Fitch identifies a total of five distinct "role beliefs" that range from conceptions of a military that is truly subordinate to democratic civilian leaders, to one that exercises a voice in "policy matters related to national security policy broadly defined" [@fitch_armed_1998, 69]. These different types of role beliefs therefore run the spectrum from one that fully adheres to the principle of civilian control, to one that intervenes politically at will. Importantly, Fitch argues that the adoption of one type of role belief over another by the military officer in Latin America is a function of the international and domestic environments, as well as previous lessons learned by the militaries of a given country regarding that military's previous experiences of intervening in politics [@fitch_armed_1998, 61-105]. Fitch's work is valuable in that it argues that a wide range of factors drive the military's conception of its normative role in the political space. Still, the detailed reasons that Fitch explores are, as the book's title demonstrates, specific to the Latin American context, and likely offer some, but limited, transport to the modern American context. 

Finally, Taylor (2003) also builds off of Finer's motives and opportunities framework, and applies it with respect to the Russian military from the period 1689-2000 [@taylor_politics_2003, 2]. His impressive book, _Politics and the Russian Army_, explores 19 incidents in which the Russian military intervenes (or might have intervened) in state "sovereign power" issues [@taylor_politics_2003, 3]. Taylor's work is valuable insofar as it serves as a  model for how to to examine the various motives and opportunities that face a state's military when considering to intervene in politics. However, it's direct utility for answering the questions that I seek to answer in the US context falls short, mainly because the political dynamics, political history, and military history are vastly different in the United States than in Russia.

In addition to these scholarly works of comparative civil-military relations, there has been an enormous rise among scholars concerning the potential pernicious impacts of political polarization on the conduct of US civil-military relations [for example, see @robinson_michael_danger_2018; @burbach_partisan_2019; @feaver_military_2020; @feaver_we_2016; @barno_how_2016; @golby_jim_americas_2020; @reid_retired_2020]. To a significant degree, scholarly interest in this topic rose, and for good reason, during the years of the Trump Administration, when American political life proved, in many ways, to be more divisive than in previous political eras. 

There have also been other scholarly works which have examined specific acts of political behavior by the US military over time, to include the endorsement of political candidates by retired military officers, and social media behavior by members of the military [@griffiths_not_2019; @dempsey_our_2010; @urben_party_2013; @urben_wearing_2014]. These have also been extremely valuable in improving our understanding of the subtle ways that the US military collectively and individually engage in acts of political activities [@brooks_militaries_2009]. Additionally, for many years, there has long been a steady stream of valuable and insightful works that examine specific historical contexts and the behavior of individual political and military figures during various era in United States history [for instance, see @mcmaster_dereliction_1997; @kaiser_no_2014; @feaver_soldiers_2001; @schake_warriors_2016]. 

While both the recent and traditional scholarship of US civil-military relations is incredibly insightful and valuable, it does not, in my view, fully explore if, how, and the extent to which underlying domestic political factors - such as political polarization and the prestige of the military - help set conditions that facilitate (or stymie!) adherence to the central principles of civil-military relations by those who practice them. In my view, what is largely missing from much of the contemporary US civil-military relations literature is a full exploration of the possibility that much of what occurred during the Trump Administration with respect to civil-military relations was not directly a result of the actions of Trump the person, but rather, the political dynamics that lay underneath the rise of Donald Trump, the divisive person who was nonetheless elected as the President in 2016.  

To be clear, my aim is not to argue that the decisions and choices made by individual civilian or military figures do not matter (they certainly do), nor that the immediate historical contexts surrounding events such as wars or crises are unimportant (these contexts are certainly very important!). Furthermore, my goal is not to absolve the actual or perceived mistakes and errors of military or civilian leaders committed during the former President's tenure, or any President, for that matter. However, my goal _is_ to extend the US civil-military relations literature by arguing that certain underlying domestic factors - political polarization and the prestige of the military in particular - play much greater roles in shaping the conduct of civil-military relations than has previously been appreciated. In this vein, this dissertation constitutes a study of how particular political dynamics influence the ways and the extent to which civilian and military actors behave with respect to the military in the United States.  

## How this Dissertation Fills This Gap

This dissertation fills a gap in the American civil-military relations literature by exploring the underlying factors that shape why and how civilian and military leaders adhere to the three central principles of civil-military relations previously described and articulated. In doing so, this dissertation sheds light on the extent to which broader, structural aspects of domestic society influence the conduct of civil-military relations, particularly in the United States. There have been similar studies undertaken in the comparative civil-military relations literature that examine the underlying factors that contribute to manifestations of various forms of civil-military conflict [@taylor_politics_2003; @bove_beyond_2020; @beliakova_erosion_2021], but few, if any, have been done that examine the US context in particular. 
I also attempt to explore the military's intervention in politics from both the military and the civilian perspectives. In other words, I ask how polarization and military prestige might induce the military to intervene in politics, but also, how these same factors might also compel civilian political leaders to use the military to do the same. The key difference in these perspectives is precisely how it sounds: is the military seeking to involve itself in politics, or, are civilian leaders seeking to involve the military in politics? By examining both perspectives, I again hope to shed new light on the impacts of the factors explored here - polarization and military prestige - and how the central actors of civil-military relations respond to and leverage each. 

Perhaps the central implication of this dissertation, particularly with respect to the practice of US civil-military relations, is that adherence to the three central principles of civilian control, non-partisanship, and "non-interference" may have in fact been easier to obtain in previous decades, and thus, achieving a posture of adherence to the three central principles, until such dynamics change, is an aspirational goal that, if it is to be achieved, will require deliberate and perpetual hard work by committed civilian and military leaders. The flip side of this implication, if true, is that when the central actors involved in civil-military relations fail to adhere to the principles of civilian control, non-partisanship, and non-interference, the broad levels of polarization and military prestige that exist in society should help better inform our understanding.  

Finally, it must be said that exploring (and, hopefully, filling) this gap in the literature is appropriate at this point in time in the United States. The nation is divided economically in the sense that the share of wealth resides in fewer hands than seen in recent decades, if not longer [@horowitz_most_2020]. The nation is divided politically, as measurements of polarization are at or exceed measures of polarization that existed more than 140 years ago [@lewis_voteview_nodate]. And yet, in the midst of all of these divisions, the American public _still_ places a relatively high degree of public trust in the military as an institution [@noauthor_confidence_2020].

As I write this dissertation in the fall of 2021 and into the Spring of 2022, it is rather remarkable to consider the relative tumult that the United States has experienced over the past 24 months. The death of George Floyd and the tensions that followed, disagreement about the response to the Coronavirus Pandemic, the turmoil that followed the 2020 Presidential Election, and most recently, the raw national emotion following the ignominious US withdrawal from Afghanistan, have combined to cause many to examine what it truly means to be an American, our relative standing in the world, and much more. It is equally remarkable to consider that the US military, for better or for worse, has featured as an actor somewhere near these scenes - sometimes at the periphery, and at other times, closer to the center. 

Because these are likely not the final events of consequence in which the US military will play some role, it is of vital importance that scholars and soldiers alike better understand how broad domestic factors, such as polarization and military prestige, impact the behavior of civilian and military actors in the United States. This dissertation is one small but important step in this endeavor. 

As such, this dissertation seeks to answer several important questions, including:

* How do varying levels of political polarization and military prestige shape the constraining influence of fundamental principles of civil-military relations?
* What types of behaviors might each actor - civilian and military - commit when the fundamental principles of civil-military relations no longer have a strong constraining influence, and why?
* Do the empirical and historical records validate the notion that the levels in political polarization and military prestige shape the ways in which military and civilian actors behave in relation to one another?
* Does a theory that emphasizes the influence of political polarization and the prestige of the military on political behaviors only apply when the military institution is professionally developed?
* What are possible counter-arguments? If political polarization and military prestige are not responsible for variation in the political behaviors that involve the military, what else might be responsible?
* If the theory is correct, how does the presence or absence of high levels of polarization and/or military prestige impact the role of senior military leaders in orchestrating, leveraging, and harnessing the military institution as a political actor? What are the implications for civilian actors?
* How does a military's ability to remain professional hinge on the levels of underlying polarization and military prestige that exist within a society? 

# Plan and Methodology of the Dissertation

The rest of this dissertation proceeds in three main steps. First, the dissertation develops an overarching theory (Chapter 2) for how the levels of polarization and military prestige shape the willingness of military and civilian actors to adhere to the fundamental principles of civil-military relations described above. Second, the dissertation tests this theory by examining several empirical behaviors committed by both military and civilian actors over time, as well as examining several historical periods to see if the predictions made by the theory are borne out in the historical and empirical record. Third, the dissertation then concludes by offering several implications of the dissertation's main findings. 

The main, central argument made in this dissertation is that the levels of polarization and prestige alter the constraining influence of the fundamental principles of civil-military relations, which then drives changes in the behaviors that both military and civilian actors commit involving the military as an institution. Building on previous "motives" and opportunities" frameworks previously used to describe military intervention in politics, my argument is that in the contemporary American context, political polarization provides the "motive" and military prestige the "opportunity" for the military's intervention into politics" [@finer_man_1962; @taylor_politics_2003].

In terms of methodology, I implement a mixed-methods approach in this dissertation that incorporates both quantitative and qualitative analysis. In the empirical chapters (3 and 4), I examine in each chapter a specific type of behavior that is committed by military actors (retired military officer opinion commentary, or writing "op-eds") and civilian actors (incorporating the military into Presidential campaign television advertisements). Each of these forms of behavior make a good form of behavior to study in depth, as each behavior regularly occurs. Chapters 3 and 4 in particular constitute truly original scholarly research. The data collected and examined are, to my knowledge, the first systematic scholarly efforts to do so. In the case study chapters (5 and 6), I then examine whether, in broad and general terms, the predictions of the general theory "fit the case." I then conclude the dissertation by drawing out important implications in Chapter 7, the final chapter. 

## Chapter 2 - Theory

In the second chapter, the dissertation develops and derives an overarching theory that explains variation in political behaviors that involve the military as a function of changing levels of political polarization and military prestige. After defining the variables of polarization and prestige in depth, attention is placed on explaining how the levels of polarization and military prestige fundamentally change the overall constraining influence of the three central principles of civil-military relations previously described. Chapter two concludes with the formulation of distinct hypotheses which are then tested throughout the remainder of the dissertation. 

## Chapter 3 - Retired Military Officer Public Commentary

In the third chapter, the dissertation conducts a quantitative analysis of one distinct form of political behavior that military actors commit - that of making "public appeals," such as giving interviews and writing opinion-editorial publications [@brooks_militaries_2009]. The data examined and analyzed in this chapter include an original data set assembled by the author of nearly 400 opinion-editorials and letters to the editor authored by retired military officers and published in _The New York Times, The Washington Post, The Wall Street Journal, The Los Angeles Times, and The USA Today_ since roughly 1979. Because the data span more than four decades, it is possible to control for changes in military prestige (which has increased since 1979) and polarization (which has also increased since 1979). This chapter shows that retired military officers are criticizing civilian officials at a greater frequency, adopting expressly partisan positions, and weighing in on topics that fall outside of traditional military expertise more frequently than in years past. These results, furthermore, suggest that retired military actors, who represent the military institution to the American public, are increasingly influenced by the nation's polarized politics and the military's relative popularity. 

## Chapter 4 - The Military and Presidential Campaign Advertisements

In Chapter four, this dissertation pivots from examining a behavior conducted by military actors (retired military officer commentary) to one that is expressly the purview of civilian actors: Presidential Campaign advertisements. Using data assembled from the Wisconsin Advertising Project and the Wesleyan Media Project, I conduct an original quantitative analysis of campaign advertisements used in the five Presidential Elections from 2000 - 2016. During this time period, the popularity or prestige of the military has remained relatively constant and at relatively high levels, while the level of political polarization rises drastically. As such, changes over this time period allows the researcher to examine if and how rising political polarization alters how the military institution is featured in campaign advertisements. To measure how the military is featured in campaign advertisements, the chapter examines various patterns in the prominence of military actors and military imagery, such as the use of veterans who speak and endorse candidates during these advertisements, and the frequency with which ads incorporate various military symbols and images. The results suggest that over the past several decades, civilian actors of both parties have harnessed the prestige of the military to advance their political futures, and that polarization has led both parties to use the military in ways that increasingly challenge the principle of non-partisanship. 

## Chapter 5 - Case Study 1: After the US Civil War (1865-1890)

Chapter five of the dissertation serves as the first of two case studies. In the first case study, the dissertation analyzes the the behavior of military and civilian actors between the end of the American Civil War and roughly 1890, or the height of the "Gilded Age" of American 19th Century politics. Over this roughly 25 year time period, the level of political polarization in the United States remains relatively high, but the level of overall military prestige decreases. Examining this time period in depth, and in particular, the role of political polarization and military prestige on various political behaviors that involve the military institution, serves as an excellent opportunity to test the predictions laid out by the theory in this dissertation.  

## Chapter 6 - Case Study 2: From Vietnam to 9/11 (1973-2001)

In the sixth chapter, the dissertation then focuses on the period after the Vietnam War until the events of September 11, 2001. During this time period, political polarization rises slightly, and the prestige of the military drastically increases, especially after the US victory during the Gulf War. Thus, this time period serves as an excellent opportunity to further examine, in detail, the predictions made by the theory about the role of changing levels of political polarization and military prestige on various types of political behaviors involving the military institution. 

It is important to briefly note the importance of examining two case studies that historically speaking, are roughly 100 years apart. The military profession exists at different levels of maturity during each of the case studies. In particular, the professional characteristics of the US military after the US Civil War are in their infancy at best [@huntington_soldier_1957]. On the other hand, the military profession by the end of Vietnam and in the decades to follow is significantly more mature and developed, even as the _health_ of the profession itself is not exceptional, particularly after Vietnam. Yet the differences in the maturity of the military profession between the case studies are important, as they could to some degree serve as a confounding variable. In other words, critics of this dissertation might assert that it is not political polarization and military prestige that are altering the constraining influence of central principles of civil-military relations, but rather the degree to which the military profession itself is mature and developed. This is an important potential critique, and not one to bypass or overlook. The timelines of the two case studies examined in this dissertation help answer this important question, which is addressed at the end of Chapter six.  

## Chapter 7 - Implications and Conclusion

In the seventh and final chapter, the dissertation's main findings are summarized and critical implications are discussed. The chapter also contains recommendations for future research.

\pagebreak
\singlespace
\newpage


