---
title: "Chapter 1 (Introduction): The Governing Standards or Central Principles of US Civil-Military Relations"
author: "Peter M. Erickson"
date: |
  `r format(Sys.time(), '%B %d, %Y')`
abstract: |
  
  _This chapter derives and explains the field of civil-military relations and introduces the research question to be explored throughout the dissertation. Using Samuel Huntington's 1957 book, The Soldier and the State, as well as critiques of his work, this chapter argues that US civil-military relations have been bounded and influenced by the adherence by both civilian and military actors to three central principles or governing standards, including the principle of civilian control of the military, that of non-partisanship of the military, and a third principle which I coin the "non-interference" of the military into certain realms of state policy, which describes the degree to which the military weighs in on issues that are indirectly or altogether unrelated to national security policy. After describing each of these three principles in detail, this chapter then presents the broad plan of the dissertation. Ultimately, this dissertation seeks to provide a unique perspective regarding the influence of a state's domestic characteristics - namely, the level of political polarization and military prestige present in society - on instances of political behaviors committed by the military institution and military actors._  
  
geometry: margin = 1.15in              # margin widths
fontsize: 12pt                         # text size
bibliography: dissertation.bib         # bibliography file
biblio-style: authoryear               # citation style
urlcolor: blue                         # links to external URLs
citecolor: black                       # links to bibliography
linkcolor: magenta                     # internal figures, footnotes, etc.
indent: true                          # paragraph format
                                       # true for indented paragraphs
                                       
header-includes:
    - \usepackage{setspace}
    - \usepackage{multirow}
    - \usepackage{indentfirst}
   
    
output: 
  bookdown::pdf_document2:
    keep_tex: true
    latex_engine: pdflatex
    number_sections: false
    toc: false
    fig_caption: true
    citation_package: biblatex
    includes: 
      in_header: 
        - latex-preamble.tex
  bookdown::htmldocument2: default
  bookdown::word_document2:
      toc: false
---

\doublespacing 

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo=FALSE,
  include=FALSE,
  message=FALSE,
  error=FALSE,
  fig.height=3,
  fig.width=4,
  out.width="90%")
```

```{r, include = FALSE, echo = FALSE}
library("here")       # file paths
library("tidyverse")  # workhorse package
library("tidylog")
library("kableExtra")
library("knitr")
library("ggdag")
library("dagitty")
library("gridExtra")
library("broom")
library("dplyr")
library("reshape2")
library("pacman")
library("ggrepel")
```

\newpage

# Introduction and Research Question

Scholars, civilians, diplomats, and military leaders regularly discuss why and how militaries intervene in politics. These discussions, moreover, often include the theoretical question of whether military actors _should_ intervene in politics, to what degree, and for what purpose. Especially in the United States context, the political activity of the military is a subject that has undergone intense scrutiny in recent years. 

In the United States, rising political polarization over the past several years has thrust the military institution into the political spotlight on a number of occasions. Though it would be impossible to list all of the contemporary instances in which the military has become entangled in politics, several stand out. From a now-retired Army Lieutenant Colonel, Alexander Vindman, testifying before Congress during impeachment proceedings against former President Trump [@stolberg_meet_2019]; to the removal of Navy Captain Brent Crozier for leaking a letter to the press in which he castigated the chain of command's response to the onset of the COVID-19 pandemic [@associated_press_teddy_2020]; to a remarkable panoply of senior retired military officers excoriating President Trump's threat to use active duty forces to quiet protests and riots in cities across America during the summer of 2020 [@brooks_let_2020-1], instances in recent years abound which illustrate the variety of domestic challenges that the US military and its members face during periods of heightened political polarization. 

A significant amount of previous scholarship within civil-military relations has explored important facets of the ways in which militaries can and do engage in political behaviors that frustrate their civilian bosses [@brooks_militaries_2009; @snider_strategic_2014; @brooks_subverting_2021]. Other, more recent scholarship, has explored various impacts of polarization on particular features of civil-military relations [@robinson_danger_2018; @robinson_dissents_2020]. In this dissertation, I seek to advance both sets of scholarship by systematically exploring the impacts of two important structural domestic factors on civil-military relations in general, and more specifically, their impacts on distinct forms of political behavior that civilian and military actors commit which involve the military institution. These two structural factors are _political polarization_, and a relatively _popular_ or _prestigious_ military institution. As both factors are likely to endure for the foreseeable future in the United States, it is important that scholars, civilian leaders, and military officers alike understand how and why these factors shape civil-military relations, and in particular, the forms and types of behaviors that both sets of actors, civilian and military, are likely to commit when these structural factors are present. 

The central argument made throughout this dissertation is that the levels of polarization and military prestige in American society shape the ability of fundamental principles or standards of civil-military relations to effectively constrain the behavior of both civilian and military actors regarding the military institution. Said differently, both the degree of polarization within a society and the prestige of the military alter the willingness of civilian and military actors to abide by fundamental principles, standards, and norms that, from a normative perspective, guide civil-military relations in the US context. In making this argument, furthermore, I build upon a general "motives" and "opportunity" framework first advanced by Finer (1962) and later by Taylor (2003) to describe military intervention in politics, and claim that in the American context, the relative levels of polarization and prestige serve the same functions [@finer_man_1962; @taylor_politics_2003]. By conducting a systematic exploration of the impacts of polarization and military prestige on civil-military relations, across time and several domestic contexts, this dissertation sheds light on what is, I argue, an under-appreciated task: the changing difficulty that military leaders in particular face in harnessing, managing, and orchestrating the domestic behavior of the military as an institution. Far from being the hierarchical and monolithic actor most people think that it is, even the US military, especially in periods of high polarization, can be remarkably divided regarding its normative role in politics and therefore, how it should act.

Examining the impacts of polarization and military prestige on political behaviors that involve the military is an important undertaking, in both normative and empirical senses. From a normative perspective, analyzing the effects of polarization and the prestige of the military reveals increasing awareness and appreciation for the domestic factors that have enabled the United States to foster the fundamental principles of civil-military relations for so long. Despite episodic civil-military troubles in US history, some of which have been severe, the United States has a relatively solid track record in terms of developing, fostering, and abiding by civil-military principles. In an empirical sense, analyzing certain forms of behaviors over time as the conditions of polarization and military prestige change likewise increases scholarly understanding of the factors that drive the actors in civil-military relations to behave a certain way. In addition to providing several valuable empirical contributions to the fields of civil-military relations, American politics, and international relations, this dissertation analyzes new data in ways that future scholars can use to further explore various aspects of civil-military behavior. 

This introductory chapter of the dissertation proceeds in three main parts. First, this introduction introduces the general field of civil-military relations. In doing so, emphasis is placed on identifying and expositing what have already been referred to in these introductory paragraphs: three central and fundamental standards or principles of American civil-military relations. The second part of the chapter describes how the intended dissertation topic fits within the scholarly literature and the gaps it intends to fill. The third and final part of this introductory chapter describes the general methodology and the plan of the dissertation.   

# The Field of Civil-Military Relations

Truly one of the interdisciplinary topics of scholarly research, civil-military relations spans the fields of political science, history, and sociology. The topic of civil-military relations, moreover, involve a set of dynamic questions and issue areas, including the relationship between militaries and governments, the relationships between militaries and societies, strategic and military effectiveness, the occurrence of coups, and a host of other pivotal topics. Within the field of political science in particular, one of the areas of civil-military relations that is most studied is the relationship between prominent state officials and military leaders. Indeed, the relationship between these actors is what one scholar has termed, "leadership at the state's apex" [@brooks_risa_integrating_2019, 380], and it is here that numerous weighty decisions that impact national security are made, including whether to go to war, how to form and implement effective strategy, how to prepare for the next conflict, and how large the military and its budget should be. 

In the most basic sense, the study of civil-military relations explores aspects of the question of how to ensure that a military is strong enough to defeat threats to the state without threatening the state itself. One scholar of civil-military relations, Peter Feaver, refers to this central question of civil-military relations as the "problematique," and notes that this dilemma is both "vexing" and a great "paradox" in that the military must be "strong enough to prevail in war," yet submissive to authority such that it cannot and will not "destroy or prey on the society it is intended to protect" [@feaver_civil-military_1996; @feaver_civil-military_1999, 214]. Furthermore, it for this reason - a state needs a military, and yet, a military is powerful and dangerous - that civil-military relations are, at least to some degree, fraught with tension that is, while perhaps capable of attenuation, never entirely absent.

A number of scholars have offered various solutions and frameworks for solving the "problematique" Feaver describes, but none has been more lasting or consequential than the late Samuel Huntington, whose 1957 work, _The Soldier and the State: The Theory and Politics of Civil-Military Relations_, often serves as the point of departure for a number of debates within the field of civil-military relations. Although Huntington's theory has endured important and substantive critiques over the years [for example, see @janowitz_professional_1960; @finer_man_1962; @cohen_supreme_2003; @feaver_armed_2003; @brooks_paradoxes_2020], it is essential to briefly exposit Huntington's conception for achieving effective civil-military relations, and what lies at the heart of it - a theory of civilian control of the military that maximizes what Huntington called "military professionalism" [@huntington_soldier_1957]. 

In the section to follow, I briefly summarize Huntington's argument and logic, as well as several prominent critiques of his work. The purpose of summarizing both Huntington and his critics is to argue that in spite of the important differences between them, Huntington and his critics share much common ground regarding the aspirational prominence of three central principles, or standards, that help foster healthy and effective civil-military relations. These principles, which I articulate and describe in the section to follow, are the principles of civilian control of the military, the "non-partisanship" of the military institution, and what I will term the "non-interference" of the military into realms of politics that are the _prerogative_ of duly-elected and appointed civilian leaders. These related principles, each of which is the subject of a fair bit of scholarship, are distinct and hold different normative implications for behavior committed by civilian and military actors.  

## Huntington, The Soldier and the State, and His Critics

Though it is not always framed as such, _The Soldier and the State_ is a book that lies at the very intersection of domestic politics and international relations. The fundamental problem that Huntington sets out to solve is how liberal democracies such as the United States can balance the simultaneous demands of military security and the ideology of political liberalism. Writing after the Korean War, Huntington argues that American security requirements constitute a "functional imperative" that require the US to maintain a military that is sufficiently strong and capable of defeating the external threats of the Cold War [@huntington_soldier_1957, 1-3]. At the same time, Huntington argues that the liberal character of American society generated a "societal imperative" such that the pursuit of American security requirements could not cause the US to deviate from its liberal political ideology in fundamental ways, including abandonment of the principle of civilian control of the military [@huntington_soldier_1957, 1-3]. Like the "problematique" pointed out by Feaver, Huntington aims to solve how to render the military strong enough to protect the United States, but sufficiently subservient and obedient to civilian power [@feaver_civil-military_1996]. 

Huntington argues, importantly, that striking this balance is made more difficult by the fact that the military officer corps is different from the rest of American society in several important ways, a claim that is not without controversy. In particular, Huntington claims that those who serve in the military hold to a distinct "mind" and "ethic" that views conflict as part of a "universal pattern throughout nature," and thus, that "the military view of man is...decidedly pessimistic" [@huntington_soldier_1957, 62-63]. Huntington then juxtaposes the mindset and ethic of the US military officer corps with tenets of American liberal ideology, noting that American liberalism, at least how he conceives of it, stresses "individualism" and that liberalism "either denies the existence of power, minimizes its importance, or castigates it as inherently evil" [@huntington_soldier_1957, 90]. It is, Huntington argues, this contrast between the unique character and ethic of militaries and the attributes of American political liberalism that explains why the US has been suspicious of standing military forces since its founding [@huntington_soldier_1957, 143-151]. 

Although Huntington remains agnostic while tracing the general tension between a unique military mindset and the facets of American liberal ideology, he then takes a clear stand as he presents his main argument. Huntington argues that the Cold War poses a unique problem in terms of the security requirements facing the US. To overcome these security requirements, sizable standing forces are a must. However, the US public tends to view such arrangements unfavorably. In order to overcome the suspicion of the US public in building and maintaining a large military machine, Huntington offers a solution whereby the United States adopts and implements a system of civil-military relations which Huntington terms "objective control" of the military [@huntington_soldier_1957, 83]. Under such an arrangement, civilian leaders grant the military a high degree of professional autonomy, which Huntington claims will enable the military to develop and hone its professional expertise and skills, while also discouraging military leaders from entering partisan politics. 

"Objective civilian control" of the military, in other words, involves civilian and military actors both recognizing that each actor owns a distinct sphere of figurative territory that is to remain separate from that owned by the other [@huntington_soldier_1957, 83-85]. Civilian leaders are not to encroach into areas that are the military's purview, which would damage military autonomy, and conversely, military leaders should not enter the world of partisan politics, as doing so subverts civilian control and the apolitical nature of the military. Figure \@ref(fig:huntington) graphically represents the overall problem and solution Huntington attempts to solve.^[As Feaver points out in a 1996 article, Huntington did not use modern political science language, such as "dependent variable" and "independent variable" when developing this concept. I constructed this puzzle using Feaver's 1996 recapitulation of Huntington. See @feaver_civil-military_1996 for more.] 

```{r figure one, include=FALSE, echo=FALSE}
huntington_dag <- dagify(
    MSCS ~ CC,
    CC ~ FI + SI,
    SI~I + S,
    coords = tribble(
      ~ name,      ~ x,    ~ y,
      "MSCS", 1, .5,
      "CC", .7, .5,
      "FI", .5, .6,
      "SI", .5, .4,
      "I", .3, .3,
      "S",.7, .3,
    ))
    
huntington_dag %>% 
  ggplot(aes(x=x, y=y, xend=xend, yend=yend))+
  geom_dag_edges()+
  geom_dag_point()+
  geom_dag_text_repel(aes(label=name), show.legend=FALSE)+
  theme_dag()
```

```{r huntington, include=TRUE, echo=FALSE, fig.height=5, fig.width=7, fig.cap="Samuel Huntington's Theory Of Civilian Control of the Military"}

huntington_dag %>% tidy_dagitty() %>% 
  dag_label(labels=c("MSCS"="Military Security and\n Civilian Supremacy", 
                     "CC"="Patterns of\n Civilian Control\n 'Objective Control'", 
                     "FI"="Functional\n Imperative", 
                     "SI"="Societal\n Imperative",
                     "I"="Liberal Ideology",
                     "S"="Constitutional Structure")) %>% 
  ggplot(aes(x=x, y=y, xend=xend, yend=yend))+
  geom_dag_edges()+
  geom_dag_point()+
  geom_dag_text()+ 
  geom_dag_label_repel(aes(label=label, fill=label), col="white", show.legend=FALSE)+ theme_dag(plot.title=element_text(hjust=.5), plot.subtitle=element_text(hjust=.5)) + labs(subtitle="As Explained in The Soldier and The State, 1957")
  
```

Huntington further claims that a key ingredient in making such an arrangement work is the rise of the military as a profession - a development that generally occurred during the 19th and 20th centuries. ^[It should be noted, however, that determining a precise date for when the US military became a "profession" is a matter of some debate. Huntington makes the point in his book that he sees the rise of the military profession as dating back to as early as 1808, when the Prussian government issued a decree pointing to the content of material that officers in its Army were expected to study. But he does not provide a definitive date for the beginning of the US military as a profession. My reading of Huntington is that he infers the "professionalization" of the US military to have begun in the post-US Civil War period. He dedicates several chapters of his book to describe the post-US Civil War growth of the US military.] Unlike the military that had existed at the founding of the United States, which was comprised of citizen-soldiers armed with very basic weapons and materials, the military by the end of World War Two employed sophisticated weapons systems such as battleships, fighter jets, and nuclear weapons throughout the course of complex military operations, such that the officer corps after World War Two comprised a body truly skilled in, using Harold Lasswell's terms, the "management of violence" [Lasswell quoted in @huntington_soldier_1957, 11]. Such significant advancements in warfare, and the accompanying increase in the skills of a professional military officer corps, were developments that US founding fathers simply could not foresee. 

The professional nature of the military, so claims Huntington, would also guide the military's relationship to politics. Huntington notes that one of the chief aims of the military officer corp is to develop and foster its own professional attributes and characteristics [@huntington_soldier_1957]. These include the development of a special body of knowledge and expertise, as well as the establishment of a degree of corporateness, which refers to the way that other professions, such as law and medicine, formalize and govern the standards by which members of the profession are certified [@huntington_soldier_1957, 11-18]. In short, by instituting Huntington's conception of "objective control" of the military, the military is enabled to further its professional autonomy, leading to a military with greater skill on the battlefield and less interest to interfere in politics, leading to a civil-military relationship that ultimately strengthens and preserves the principle of civilian control of the military [@huntington_soldier_1957, 83-85].

Huntington's work, and specifically, his concept of "objective civilian control" of the military has generated a number of important and meaningful critiques over the years, and it would be impossible to list all of these critiques here. However, a major theme of these critiques is that the stark separation between military and political spheres, which Huntington certainly advocated for, is unwise, impractical, and potentially damaging to US security. Citing Clausewitz's famous dictum that war is a "mere continuation of policy by other means", a theme of Huntington's critics is that the military is an inherently political institution that cannot conduct military operations in a vacuum, so to speak, absent from political influence [@clausewitz_war_1976, 69, 605]. 

In addition to the inherent political nature of militaries in general, which renders Huntington's conception of civil-military relations impractical, his critics also point to broader democratic theory,  which they argue does not square with the idea of separating the duties and responsibilities of statesmen and soldiers, which they claim Huntington argues for. This general critique argues that by virtue of their superior position within democratic governments, elected civilian officials are in fact those who ultimately must shoulder the responsibility for what militaries do and fail to do, and thus should not abdicate, under any circumstances, these crucial responsibilities to the military [@cohen_supreme_2003]. Peter Feaver has parsimoniously captured the essence of this critique of Huntington by noting that the principle of civilian supremacy over the military means that "civilians have the right to be wrong" [@feaver_right_2011, 117; @feaver_armed_2003]. 

Other recent critiques of Huntington have further developed the possible domestic and international consequences that arise if military officers strictly adopt a Huntingtonian concept of professionalism that starkly separates civilian and military spheres. For instance, critics warn that doing so can lead to military officers failing to recognize the inherent political nature of their roles, which can subsequently lead to military officers unknowingly engaging in political behaviors under the guise of acting in an apolitical manner [@brooks_paradoxes_2020; see also @janowitz_professional_1960; @feaver_civil-military_1996]. Others warn that a serious consequence of military officers adhering to a strict separation between military and civilian spheres in wartime is the failure to develop, implement, and achieve sound objectives and strategy, which leads to disastrous effects on overall US national security policy [@rapp_civil-military_2015].   

In spite of these insightful and important concerns and critiques of Huntington, Huntington and his critics nonetheless share important common ground regarding several fundamental aspects of civil-military relations. Indeed, Huntington and his critics agree that three central principles or standards should govern the relationship between the military and civilian leaders, even if they may disagree on the specific ways in which these principles or standards should be applied. The central principles or governing standards of civil-military relations, shared by Huntington as well as his critics, constitute a _baseline_ set of principles that can inform and guide civil-military relationships, to include the behaviors that each set of actors undertakes. These three principles or standards are that of civilian control of the military, non-partisanship of the military institution, and what I am terming the "non-interference" of the military into certain areas or realms of state affairs. 

In the section below, I define and exposit each of these standards or principles. As is often the case in academic scholarship and particularly within the study of civil-military relations, these standards and principles are generally conceptualized not in dichotomous fashion, but on a gradual continuum, or scale. Additionally, I claim that when these three principles are upheld and respected by both sets of actors, the civil-military relationship between civilian leaders and the military is likely to be relatively harmonious. Alternatively, when one or more of these central principles is shunned, bucked, disregarded or blatantly violated by one or both parties, then a state's civil-military relationship is likely to be discordant.  

## Central Principle 1: Civilian Control of the Military

The first central principle of civil-military relationships is that of civilian control of the military. It is listed first deliberately, as it is the central component or standard by which the health of civil-military relationships in democratic countries is maintained. Without strong adherence to the principle of civilian control of the military, it is doubtful that a state's civil-military relationship can be healthy nor successful.  

What is civilian control of the military, and how does one know if military and civilian actors within a state are adhering to it as a principle of civil-military relations? Scholars have not agreed upon a precise definition of the concept of civilian control of the military, as it is a giant concept in and of itself. Nonetheless, scholars share the idea that civilian control generally refers to the "relative political power" that exists between a nation's armed forces and its civilian leaders [@bruneau_civil-military_2019, 7; see also @brooks_shaping_2008]. Furthermore, as was mentioned previously, modern conceptions of civilian control are not dichotomous, but rather exist along a scale that, at least theoretically, can range from instances of "full civilian control to complete military dominance over the system" [@bruneau_civil-military_2019, 7; @fitch_armed_1998]. Contemporary scholars, moreover, have argued that scholars and practitioners of civil-military relations must look beyond coups to determine the health of the principle of civilian control [@croissant_beyond_2010]. Other scholars who recognize the nuance involved in the principle of civilian control further recommend that scholars should examine the "patterns" of civilian control that exist within a state, and not merely look for the presence of it [@feaver_civil-military_1996, 167]. 

In the United States, the principle of civilian control of the military receives some assistance in the fact that the US Constitution prescribes that civilian leaders control the military. Far from being a guarantee that the principle of civil-military relations is adhered to in healthy ways, these statutory guidelines nonetheless serve as a normative and sometimes a legal marker for civilian and military actors to aspire to. For instance, the US Constitution grants the President the power to serve as the Commander in Chief of the Armed Forces [@united_states_us_1787], yet also grants Congress unique and specific powers of oversight of the military [@united_states_us_1787-1]. 

Finally, it is reasonably clear that both Huntington and his critics share a similar concern for the supremacy of the principle of civilian control of the military. Although critiques of Huntington have been plentiful across the years, these critiques are aimed at Huntington's conception of _how_ to best achieve civilian control. Indeed, Huntington's theory seeks to achieve two major goals: the security of the United States _and_ the maintenance of civilian supremacy [@huntington_soldier_1957]. Before moving on to address the second central principle of civil-military relations, it should briefly be mentioned that of the two major sets of actors civil-military relations, only the military can _violate_ the principle of civilian control of the military. To be clear, civilian actors certainly play a part and have a responsibility to ensure that the principle of civilian control of the military is healthy; as such, civilians can fail in this regard, either through being derelict in their duties to enforce civilian control, or by failing to demand that the military behave in a manner that is consistent with the principle of civilian control. Nonetheless, speaking in a strict sense, only military actors can _violate_ the principle of civilian control. 

## Central Principle 2: Non-Partisanship of the Military Institution

A second central principle or standard of civil-military relations, particularly in the US context, is the principle of non-partisanship of the military institution. In general terms, the principle of non-partisanship of the military stipulates that the military not align itself - nor that it be made to align by civilians - with a particular political party or its platform. Thus, both sets of actors, civilian and military, have a responsibility to maintain the standards of non-partisanship of the military. 

However, determining the degree to which civilian or military actors have upheld the principle of non-partisanship in practice is a difficult undertaking, largely because it is difficult to measure what forms of behavior constitute a violation of this principle. After all, if militaries are, in fact, as Clausewitz and the critics of Huntington have contended, inherently political creatures who "serve at the pleasure" of their civilian bosses, then it stands to reason that military will, at some point, enact the partisan policies, wishes, desires, and goals of their elected civilian leaders, who belong to a political party [@mullen_chairman_2011]. 

For this reason, scholars of civil-military relations have often separated the various meanings and implications of certain forms of behavior and their impact on politics. While there is no clear delineation of which forms of behavior and which contexts violate the principle of non-partisanship of the military, recent scholarship has attempted to create figurative space to allow greater acceptance - even encouragement - of military officers committing some forms of political behavior. For instance, scholars have argued that military officers and civilians should expect military leaders to increasingly contribute to debates about security policy on the grounds that doing so will lead to the better formulation of strategy and thus, to better national security outcomes [@brooks_perils_2013; @owens_military_2015; @rapp_civil-military_2015]. Simultaneously, scholars and military officers alike recognize that other forms of behavior, such as retired military officers speaking at political conventions on behalf of political candidates, comprise blatant violations of the military's non-partisan ethic [@dempsey_military_2016]. 

Similar sentiments hold for the behavior of civilian actors, but with far less clarity in rendering a judgement regarding whether a particular behavior truly violated the principle of non-partisanship. For example, many scholars and former military officers blasted the decision of President Trump to deploy US Troops to the US southern border in 2018, citing the move as a political stunt ahead of the 2018 midterm elections [@adams_trumps_2018]. Other critics alleged that President Trump's threat to use active duty forces to dispel protesters and rioters in the summer of 2020 likewise violated the military's non-partisan ethic [@brooks_dismay_2020]. Whether these instances, and to what degree, actually posed violations of the military's non-partisan standard, is answerable in terms of gradations and degrees, rather than stark dichotomous terms. 

For purposes of this dissertation, it is important to point out that both Huntington and his critics share the concern that the military operate as an actor outside of the world of partisan politics, even if they disagree on how best to structure civil-military relations to achieve this principle. Even today, while scholars may not agree on precisely which instances of behavior conducted in specific political contexts constitute challenges or violations to the military's non-partisan aspirations, these scholars would nonetheless agree that as a principle, the non-partisanship of the military is a vital component to maintaining healthy civil-military relations [@owens_military_2015; @brooks_paradoxes_2020]. Although Huntington offers a structure of civil-military relations such that the military operating more distinctly from the political sphere than his critics deem wise, both Huntington and his critics share the concern that the military be as free as possible from the partisan aspects of politics. 

## Central Principle 3: The "Non-Interference" of the Military 

A third central principle or governing standard that Huntington and his critics affirm is what I am terming the _"non-interference"_ of the military into certain realms, or areas, of state politics. The principle of "non-interference" of the military is not, unlike the two previous principles, one which is explicitly mentioned in the American civil-military relations literature, but it is nonetheless a standard that is heavily implied throughout Huntington's _The Soldier and the State_, and one which is generally accepted by his critics. In general terms, the principle of "non-interference" generally describes the policy arenas or issues areas upon which the military institution and its leaders influence and weigh in, and by extension, the resulting types of missions and activities that the military conducts. 

In further defining the standard of _non-interference_ by the US military, such a principle conveys the idea that there are policy arenas which are and should remain "off limits" for US military actors in the sense that military actors are not and should not be responsible for solving certain types of state problems or issues. To be clear, a standard or principle of _non-interference_ accepts that the specific policy areas or issues upon which military actors should not solve may change over time, but nonetheless accepts that even if civilian leaders have, by virtue of their superiority in democracies, "a right" to encroach into all areas of state politics, including military affairs, the converse is not true for the military [@cohen_supreme_2003; @feaver_armed_2003]. In other words, the principle of "non-interference" simply means that military leaders should not called to solve any issue that they desire to, and thus, they do not weigh in on certain subsets of state policy.

Huntington devotes an entire chapter in which he exposits the notion of a "military mind" as being distinct and unique from that of civilians [@huntington_soldier_1957, 59-79]. Furthermore, Huntington asserts that a unique "military mind" consists of the "values, attitudes, and perspectives which inhere in the performance of the professional military function and which are deducible from the nature of that function" [@huntington_soldier_1957, 61]. In other words, Huntington believed that the uniqueness of the military - in terms of its mindset, corporate ethic, and values - stemmed entirely from the fact that the military's purpose is to provide national defense for the nation. In drawing this conclusion, Huntington implies that the military should be used to fill roles that are related to national defense or the security of the nation. Failing to do so will render the "unique" nature of the "military mind" less distinct [@huntington_soldier_1957, 59-79].

There is vast scholarship within comparative civil-military relations that relates to and describes the principle of "non-interference" of the military into certain realms of state politics, even as militaries across adhere to such a principle to widely varying degrees. This scholarship includes explorations of the varying mission sets that militaries around the world conduct [for instance, see @vennesson_is_2009; @pion-berlin_democratization_2005; @bruneau_civil-military_2019], and the degree to which certain militaries are used to conduct internal state security missions such as domestic counter-terrorism vice external warfare [@stepan_new_1973; @pion-berlin_civil-military_2020]. In his excellent work on Israeli civil-military relations, Yehuda Ben-Meir (1995) separates the affairs of the state into four broad areas, including political affairs, domestic affairs, national security, and the armed forces [@ben-meir_civil-military_1995]. He argues that civilians should and do influence all four of these areas of politics, whereas the military should influence the three areas of domestic affairs, national security, and the armed forces [@ben-meir_civil-military_1995].^[In terms of the activities which constitute political affairs, Ben-Meir includes items such as "taking control of the government (coups), influencing political appointments, or interfering in the decision making process. See @ben-meir_civil-military_1995, 4-5 for an excellent description and diagram.] While he address the Israeli context specifically, Ben-Meir's argument is similar to that which I am making here in calling for a distinct norm of "non-interference" of the military into state affairs.

In practice within the US context, the principle of "non-interference" of the military into certain realms of state politics features prominently anytime one hears critiques that levy something to the extent of, "the military should not be used to do that," or, "doing that is not the military's job." When the Chairman of the Joint Chiefs of Staff, General Mark Milley, foresaw no role for the military in the 2020 election, he was implicitly invoking the principle of "non-interference" of the military [@silva_gen_2020] in at least, what was, a fairly clear example in terms of US history. Other instances of the military's use to solve other types of problems are not as clear. For example, the US military's role in the COVID-19 crisis, and in particular, in distributing vaccines, has not been without controversy.^[At one point in late 2020, an active duty four star general apologized to the nation for a mix-up in information regarding the distribution of COVID vaccine, leading some critics to express concern that the military was making inherently political decisions that, from a normative perspective, posed some problems.  See @passy_gen_2020 for more details.]

## Summary of 3 Central Principles / Standards of US Civil-Military Relations

In summary, Huntington's _The Soldier and the State_ has served for decades as a starting point for numerous debates involving civil-military relations. Despite the important critiques of Huntington's work, scholars of civil-military relations have generally agreed upon the importance of three foundational principles to guide the behavior of both military and civilian leaders. These three standards are that of civilian control of the military, non-partisanship of the military institution, and "non-interference" of the military into certain arenas of state policy. 

While the precise meaning and implications of each of these principles varies significantly across time and country context, these principles, when adhered to, have generally constrained the behavior of both civilian and military actors in important ways. In the US in particular, which has certainly witnessed its own share of civil-military trials and travails throughout its history, civil-military relations have generally been stable and harmonious when these three principles are adhered to by civilian and military actors. Therefore, these three principles constitute a _baseline_ set of principles which can be examined and analyzed in order to evaluate the health of civil-military relations. Table \@ref(tab:principles) captures these three principles, their definitions, and the actor that can expressly violate each principle.  

```{r principles, include=TRUE, echo=FALSE}

text_tbl1 <- tribble(~"Governing Principle or Standard", ~"Definition/Description", ~"Potential Violating Actor",
  "Civilian Control of the Military", "The Military Institution is Subordinate to its Duly Elected Civilian Leaders; the Degree of Relative Power Between Civilian Leaders and the Military ", "Military",
  "Non-Partisanship of the Military Institution",  "The Military Institution Exists and Operates outside of Partisan Politics; the degree to which Civilian or Military Actors Attempt to Harness the Military for Distinctly Partisan Ends", "Civilian or Military",
  "Non-Interference of the Military into Certain Realms of State Policy", "There are Areas of State Policy into which the Military does not Enter; the degree to which the Military Weighs in on/Influences Policies of the State that have Little to Deal with the Military", "Civilian or Military")
 

text_tbl1 %>% kable(booktabs=T, caption="Central Principles of US Civil-Military Relations") %>%
  kable_styling(full_width = F, font_size=10, latex_options="striped" ) %>% 
  column_spec(1, width="10em") %>% 
  column_spec(2, width="15em") %>% 
  column_spec(3, width="10em") 
  

```

Before describing the gap in the current literature that this dissertation seeks to fill, it is important to draw the reader's attention to the fact that in terms of structure, thinking about the military's relationship to guiding principles of civil-military relations is related to other topics in the field of civil-military relations. In particular, the study of military officer "role beliefs" examines a similar topic, though in ways that slightly differ from those used in this dissertation. Nonetheless, thinking about how militaries behave, or are made to behave, in relation to guiding principles has some precedent. 

# Military Role Beliefs

One helpful concept that demonstrates similarity between the relationship of military actors to fundamental principles of civil-military relations in the United States and the relationship between military actors and their role in politics in other parts of the world is the concept of military "role beliefs," which was first developed and used by J. Samuel Fitch in his book, _The Armed Forces and Democracy in Latin America._ In this section, I briefly describe and review the concept of military "role beliefs" in order to demonstrate that while the fundamental principles of civil-military relations may vary significantly across countries, scholarly frameworks for thinking about the role of the military in politics, broadly speaking, is a rich and developed undertaking. 

Through rich interview data conducted in the 1990s, Fitch traces how military leaders in Argentina and Ecuador adopted one of five distinct "role beliefs," referring to the perceptions and attitudes held by these officers regarding their normative role in politics, and importantly, their roles as military officers with respect to democracy as a political system [@fitch_armed_1998, 61]. These "role beliefs" vary in several dimensions, most notable of which are the degree of subordination and assertiveness militaries should espouse relative to civilian authority, especially in times of crises, and the duties and responsibilities militaries should fulfill within the government. 

The first type of role belief Fitch identifies is "democratic professionalism," and involves both "a subjective commitment to democracy as the preferred political system," and a "belief in a professional ethic that differentiates and separates military and political affairs" [@fitch_armed_1998, 66]. This type of "role belief", I would point out, is closely aligned with the normative standards of _military professionalism_ raised by Huntington. A second type of "role belief" Fitch identifies is "classical professionalism," which differs from "democratic professionalism" in that while it "argues for the subordination of the military...to the constitutionally designated civilian authorities", it does not espouse a "normative commitment to a democratic regime" [@fitch_armed_1998, 66-67]. Where a clear departure from American Huntingtonian norms emerges is beginning with a third type of "role belief" that Fitch identifies. When militaries conceive of themselves as a "power factor," they adopt a belief that their defense role within society entitles them to actively participate in "the national project" [@fitch_armed_1998, 67]. Though the boundaries of participation are ambiguous, this third belief suggests that military officers should and do seek to fulfill their "rightful place" within the regime [@fitch_armed_1998, 68]. 

Continuing a trend of increasing assertiveness of the military institution, Fitch identifies that a fourth type of "role belief" is one of "conditional separation" in which the military stays out of politics "in normal circumstances" but not necessarily if and when crises exist [@fitch_armed_1998, 68]. This type of role belief warrants the military to "safeguard the highest interests of the nation" during times of crisis, "even if that entails overthrowing the incumbent government and assuming power temporarily" [@fitch_armed_1998, 68]. A final possible role belief militaries can adopt is a "tutelary" model in which the civilian government grants the military both a "high degree of autonomy" and the ability to become a regular and "institutional voice in policy matters" [@fitch_armed_1998, 68-69]. Military actors who espouse this role belief deem the military the final "guardian of national security," and thus, that the military has "the right" to participate in a wide variety of policy debates that impact the state [@fitch_armed_1998, 69]. Fitch argues that in the Latin American context, the particular role beliefs militaries adopt are a function of three primary factors: the lessons militaries draw from previous domestic intervention in politics, a wide-ranging assessment of current domestic indicators (the economy, the effectiveness of institutions, etc.), and the prevailing international context [@fitch_armed_1998].  

The specific "role beliefs" that Fitch develops and describes for existing in Latin America do not map directly well onto the United States. As Fitch has described, the variation in military "role beliefs" in Latin America is far wider, thankfully, than that which the American experience has demonstrated (at least up until this point in time). Nonetheless, an underlying similarity between Fitch's study and the research project I undertake in this dissertation concerns the centrality of principles that guide and shape the various actors in civil-military relations, and their willingness to follow them.  

# The Gap in the Current Literature

While there have been many historical episodes in the United States in which civilian or military actors failed to abide by the central principles of civilian control, non-partisanship, and the "non-interference" of the military, most explanations have focused on specific historical contextual factors in order to explain these events. While these factors are important and accurate, scholarly attention has mostly neglected the possibility that broad structural aspects of domestic society, in addition to the historical and contextual circumstances surrounding specific events or instances, have also influenced the willingness of civilian and military actors to abide by and follow the principles of civil-military relations.  

It is a central premise of this dissertation that deviation from one or more of the three central principles or standards of civil-military relations - principles which I have argued constitute a baseline of normative behavioral principles for both military and civilian leaders - will result in military and civilian committing specific political behaviors that seem to run against these principles. If this is true, a key starting question to consider is, "what factors or variables might influence or shape the degree to which military and civilian actors follow, abide, and adhere to the principles of civil-military relations described above?" This dissertation considers that question as its starting point, and posits that polarization and military prestige alter the ability of the central principles of civil-military relations to constrain the behavior of civilian and military actors. 

Exploring the drivers of political behaviors that involve the US military fills an important and timely gap in the scholarly literature. Previous works have explored the topic of military intervention and involvement in politics, both in a general sense [@finer_man_1962] and in several comparative contexts, such as Russia [@taylor_politics_2003] and Latin America [@fitch_armed_1998]. As of yet, however, such a work does not exist within the context of the United States. This is likely because, thankfully, the United States does not have a tradition of prolonged and extensive inappropriate intervention by the military into politics. While scholars have developed helpful typologies for understanding the types of political behaviors that the US military engages in, these scholars have not attempted to explain what causes variation in these types of behaviors, and why [@brooks_militaries_2009]. Other contemporary civil-military scholars have made tremendous contributions exploring the various impacts of polarization on the military, to include its impact on the public credibility of the military [@robinson_michael_danger_2018; @burbach_partisan_2019]. Still, these do not holistically consider the influence of polarization and other factors on civil-military relations, and on particular behaviors that civilian and military actors commit. 

Furthermore, while these are all important and scholarly contributions, none have yet attempted to explain variation in types of political behaviors that involve the American military institution. Though the rise of political polarization in America has occurred for several decades, civil-military scholars have only somewhat recently given serious attention to the topic, particularly in the age of former President Donald Trump [@feaver_we_2016; @barno_how_2016; @golby_jim_americas_2020; @reid_retired_2020]. These works are important, but mainly impressionistic rather than analytic. Those that have analyzed military behaviors over time (such as military endorsements of political candidates) [@griffiths_not_2019] or military attitudes about certain behaviors (such as voting and social media use) [@dempsey_our_2010; @urben_party_2013; @urben_wearing_2014] have stopped short of developing and testing a holistic theoretical argument against multiple types of behaviors across periods of time. 

Moreover, it is appropriate to explore, at this point in time in the United States, the impact of political polarization and the prestige of the military on civil-military relations. As the year 2020 and all of its tumult - from the death of George Floyd and the aftermath, to the nation's response to COVID-19, to a very spirited Presidential Election whose outcome was in question for at least several weeks afterward - the United States is divided to a level that has not been seen at least for a very long time. The nation is divided economically in the sense that the share of wealth resides in fewer hands than seen in recent decades, if not longer [@horowitz_most_2020]. The nation is divided politically, as measurements of polarization are at or exceed measures of polarization that existed at the end of the American Civil-War [@lewis_voteview_nodate]. America is also divided racially. Over the past few years, the "Black Lives Matter" movement, combined with questions about the meaning of concepts such as equality, diversity, and even the fundamental purposes upon which America was founded, have ignited a civil-rights movement that is not only moving at incredible speed, but is altogether different than that which began some 60 years ago by Dr. Martin Luther King Jr. And yet, in the midst of all of this division, the American public _still_ places a relatively high degree of public trust in the military as an institution [@noauthor_confidence_2020], even after two decades of involvement in inconclusive wars in Iraq and Afghanistan. 

Therefore, it is of vital importance that scholars understand if and how broad domestic factors impact the behavior of civilian and military actors. This dissertation is one small step in this important endeavor. As such, this dissertation seeks to answer several important questions, including:

* How do varying levels of political polarization and military prestige shape the constraining influence of fundamental principles of civil-military relations?
* What types of behaviors might each actor - civilian and military - commit when the fundamental principles of civil-military relations no longer have a strong constraining influence, and why?
* Do the empirical and historical records validate the notion that the levels in political polarization and military prestige shape the ways in which military and civilian actors behave in relation to one another?
* Does a theory that emphasizes the influence of political polarization and the prestige of the military on political behaviors only apply when the military institution is professionally developed?
* What are possible counter-arguments? If political polarization and military prestige are not responsible for variation in the political behaviors that involve the military, what else might be responsible?
* If the theory is correct, how does the presence or absence of high levels of polarization and/or military prestige impact the role of senior military leaders in orchestrating, leveraging, and harnessing the military institution as a political actor? What are the implications for civilian actors?
* How does a military's ability to remain professional hinge on the levels of underlying polarization and military prestige that exist within a society? 

# Plan of the Dissertation

The rest of this dissertation proceeds in three main steps. First, the dissertation develops an overarching theory (Chapter 2) for how the levels of polarization and military prestige shape the willingness of military and civilian actors to adhere to the fundamental principles of civil-military relations described above. Second, the dissertation tests this theory by examining several empirical behaviors committed by both military and civilian actors across time, as well as examining several historical periods to see if the predictions made by the theory are borne out in the historical and empirical record. Third, the dissertation then concludes by offering several implications of the dissertation's main findings. 

In this section of the prospectus, I will describe the overall plan for my dissertation. In short, I plan to implement a mixed-methods approach that incorporates quantitative and qualitative analysis over the course of what I have now planned to take seven chapters. The main argument made in this dissertation is that polarization and prestige alter the constraining influence of fundamental principles of civil-military relations, which further drive changes in the behaviors that both military and civilian actors commit involving the military as an institution. Building on previous "motives" and opportunities" frameworks previously used to describe military intervention in politics, my argument is that in the contemporary American context, political polarization provides the "motive" and military prestige the "opportunity" for the military's intervention into politics" [@finer_man_1962; @taylor_politics_2003].

## Chapter 2 - Theory

In the second chapter, the dissertation develops and derives an overarching theory that explains variation in political behaviors that involve the military as a function of changing levels of political polarization and military prestige. After defining the variables of polarization and prestige in depth, attention is placed on explaining how the levels of polarization and military prestige fundamentally change the overall constraining influence of the three central principles of civil-military relations previously described. Chapter two concludes with the formulation of distinct hypotheses which are then tested throughout the remainder of the dissertation. 

## Chapter 3 - Retired Military Officer Public Commentary

In the third chapter, the dissertation conducts a quantitative analysis of one distinct form of political behavior that military actors commit - that of making "public appeals," such as giving interviews and writing opinion-editorial publications [@brooks_militaries_2009]. The data examined and analyzed in this chapter include an original data set assembled by the author of nearly 400 opinion-editorials and letters to the editor authored by retired military officers and published in _The New York Times, The Washington Post, The Wall Street Journal, The Los Angeles Times, and The USA Today_ since roughly 1979. Because the data span more than four decades, it is possible to control for changes in military prestige (which has increased since 1979) and polarization (which has also increased since 1979). This chapter shows that retired military officers are criticizing civilian officials at a greater frequency, adopting expressly partisan positions, and weighing in on topics that fall outside of traditional military expertise more frequently than in years past. These results, furthermore, suggest that retired military actors, who represent the military institution to the American public, are increasingly influenced by the nation's polarized politics and the military's relative popularity. 

## Chapter 4 - The Military and Presidential Campaign Advertisements

In Chapter four, this dissertation pivots from examining a behavior conducted by military actors (retired military officer commentary) to one that is expressly the purview of civilian actors: Presidential Campaign advertisements. Using data assembled from the Wisconsin Advertising Project and the Wesleyan Media Project, I conduct an original quantitative analysis of campaign advertisements used in the five Presidential Elections from 2000 - 2016. During this time period, the popularity or prestige of the military has remained relatively constant and at relatively high levels, while the level of political polarization rises drastically. As such, changes over this time period allows the researcher to examine if and how rising political polarization alters how the military institution is featured in campaign advertisements. To measure how the military is featured in campaign advertisements, the chapter examines various patterns in the prominence of military actors and military imagery, such as the use of veterans who speak and endorse candidates during these advertisements, and the frequency with which ads incorporate various military symbols and images. The results suggest that over the past several decades, civilian actors of both parties have harnessed the prestige of the military to advance their political futures, and that polarization has led both parties to use the military in ways that increasingly challenge the principle of non-partisanship. 

## Chapter 5 - Case Study 1: After the US Civil War (1865-1890)

Chapter five of the dissertation serves as the first of two case studies. In the first case study, the dissertation analyzes the the behavior of military and civilian actors between the end of the American Civil War and roughly 1890, or the height of the "Gilded Age" of American 19th Century politics. Over this roughly 25 year time period, the level of political polarization in the United States remains relatively high, but the level of overall military prestige decreases. Examining this time period in depth, and in particular, the role of political polarization and military prestige on various political behaviors that involve the military institution, serves as an excellent opportunity to test the predictions laid out by the theory in this dissertation.  

## Chapter 6 - Case Study 2: From Vietnam to 9/11 (1973-2001)

In the sixth chapter, the dissertation then focuses on the period after the Vietnam War until the events of September 11, 2001. During this time period, political polarization rises slightly, and the prestige of the military drastically increases, especially after the US victory during the Gulf War. Thus, this time period serves as an excellent opportunity to further examine, in detail, the predictions made by the theory about the role of changing levels of political polarization and military prestige on various types of political behaviors involving the military institution. 

It is important to briefly note the importance of examining two case studies that historically speaking, are roughly 100 years apart. The military profession exists at different levels of maturity during each of the case studies. In particular, the professional characteristics of the US military after the US Civil War are in their infancy at best [@huntington_soldier_1957]. On the other hand, the military profession by the end of Vietnam and in the decades to follow is significantly more mature and developed, even as the _health_ of the profession itself is not exceptional, particularly after Vietnam. Yet the differences in the maturity of the military profession between the case studies are important, as they could to some degree serve as a confounding variable. In other words, critics of this dissertation might assert that it is not political polarization and military prestige that are altering the constraining influence of central principles of civil-military relations, but rather the degree to which the military profession itself is mature and developed. This is an important potential critique, and not one to bypass or overlook. The timelines of the two case studies examined in this dissertation help answer this important question, which is addressed at the end of Chapter six.  

## Chapter 7 - Implications and Conclusion

In the seventh and final chapter, the dissertation's main findings are summarized and critical implications are discussed. The chapter also contains recommendations for future research.

# Summary

A high degree of political polarization and a popular military institution are features of the contemporary US political landscape that are likely to remain for some time. As we have seen in this introductory chapter, civil-military relations in general are marked by a certain degree of irreducible tension based on the fact that while states desire their militaries to provide security for the state, enabling them to do so provides militaries with coercive power that can, if left unchecked, pose a threat to the state itself [@feaver_civil-military_1996]. 

This chapter also made the case that in spite of the inherent tension that exists in the practice of civil-military relations, adherence by both civilian and military actors to three central principles or governing standards of civil-military relations drives the health of the civil-military relationship within a state. These principles include that of civilian control of the military, the non-partisanship of the military institution, and what I have termed the "non-interference" of the military into certain areas of state policy. Additionally, although the precise meaning and implications for application of each principle varies significantly across time and across country context, both Samuel Huntington, whose model of civil-military relations outlined in his 1957 book, _The Soldier and the State_, still serves as a scholarly reference point for numerous debates within civil-military relations, and many of his critics, support and advocate for adherence to these three central principles. 

In the pages to follow, I develop and test a theory that argues that the levels of polarization and military prestige influence the degree to which the various actors in civil-military relations adhere to these central principles, and why. Of course, the ultimate evidence for whether military and civilian actors adhere to these principles is particular instances of behavior that challenge or violate these principles. The goal of this dissertation is to convince you, the reader, that there is in fact empirical and historical proof that the level of political polarization and military prestige combine to influence how military and actors adhere to the fundamental principles of civil-military relations, and thus, the types of behaviors that each commits. While contextual circumstances may also influence the willingness of civilian and military actors to adhere to central principles of civil-military relations, as evidenced by numerous civil-military relations imbroglios or incidents throughout US history, this dissertation undertakes a systemic evaluation of the broader political and societal factors and their influence on US civil-military relations.

\pagebreak
\singlespace
\newpage


