---
title: "Chapter 1 (Introduction): The Governing Standards or Central Principles of US Civil-Military Relations"
author: "Peter M. Erickson"
date: |
  `r format(Sys.time(), '%B %d, %Y')`
abstract: |
  
  _This chapter introduces the research question to be explored throughout this dissertation. Using Samuel Huntington's 1957 book, The Soldier and the State, as well as critiques of his work, this chapter argues that the US military profession has been bounded and influenced by the adherence by both civilian and military actors to three central principles or governing standards, which include the principles of civilian control of the military, non-partisanship of the military institution, and a third principle which I coin the "non-interference" of the military into certain realms of state policy, which describes the degree to which the military weighs in on issues that are indirectly or altogether unrelated to national security policy. After describing each of these three principles in detail and the gap in the scholarly literature this dissertation seeks to fill, this chapter then presents the broad plan of the dissertation. Ultimately, this dissertation argues that particular domestic characteristics - in the US, the level of political polarization and military prestige present in society - profoundly impact the quality and severity of political behaviors engaged in by members of the military institution and civilian actors, and thus, the conduct of civil-military relations in general._  
  
geometry: margin = 1.15in              # margin widths
fontsize: 12pt                         # text size
bibliography: dissertation.bib         # bibliography file
biblio-style: authoryear               # citation style
urlcolor: blue                         # links to external URLs
citecolor: black                       # links to bibliography
linkcolor: magenta                     # internal figures, footnotes, etc.
indent: true                          # paragraph format
                                       # true for indented paragraphs
                                       
header-includes:
    - \usepackage{setspace}
    - \usepackage{multirow}
    - \usepackage{indentfirst}
   
    
output: 
  bookdown::pdf_document2:
    keep_tex: true
    latex_engine: pdflatex
    number_sections: false
    toc: false
    fig_caption: true
    citation_package: biblatex
    includes: 
      in_header: 
        - latex-preamble.tex
  bookdown::htmldocument2: default
  bookdown::word_document2:
      toc: false
---

\doublespacing 

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo=FALSE,
  include=FALSE,
  message=FALSE,
  error=FALSE,
  fig.height=3,
  fig.width=4,
  out.width="90%")
```

```{r, include = FALSE, echo = FALSE}
library("here")       # file paths
library("tidyverse")  # workhorse package
library("tidylog")
library("kableExtra")
library("knitr")
library("ggdag")
library("dagitty")
library("gridExtra")
library("broom")
library("dplyr")
library("reshape2")
library("pacman")
library("ggrepel")
```

\newpage

# Introduction and Research Question

Scholars, civilians, diplomats, and military leaders regularly discuss why and how militaries intervene in politics. These discussions, moreover, often include the theoretical question of whether military actors _should_ intervene in politics, to what degree, and for what purpose. Especially in the United States context, political activity by members of the military and the politicization of the military at the hands of civilian leaders are subjects that have undergone intense scrutiny in recent years. 

In the United States, rising political polarization over the past several years has thrust the military institution into the political spotlight - for better or for worse - on a number of occasions. Though it would be impossible to list all of the contemporary instances in which the military has become entangled in politics, several stand out. From former President Trump's appointment of several retired military officers to key postings within his administration [@brooks_dont_2016], the removal of Navy Captain Brent Crozier for leaking a letter to the press in which he castigated the chain of command's response to the onset of the COVID-19 pandemic [@associated_press_teddy_2020], and the remarkable panoply of senior retired military officers excoriating former President Trump's threat to use active duty forces to quiet protests and riots in cities across America during the summer of 2020 [@brooks_let_2020-1], instances abound in recent years which illustrate some of the challenges facing both the military and civilian leaders during periods of heightened political polarization. 

A significant amount of previous scholarship within civil-military relations has explored important facets of the ways in which militaries can and do engage in various political behaviors that frustrate their civilian bosses [@brooks_militaries_2009; @snider_strategic_2014; @brooks_subverting_2021]. Other scholarship has explored this problem from the opposite vantage point; that is, how civilian leaders often bring the military into politics, or politicize it [@golby_jim_americas_2020; @golby_uncivil-military_2021]. Other scholars have explored the various ways in which polarized politics make particular features of civil-military relations more difficult, or at least, potentially problematic [@robinson_danger_2018; @robinson_dissents_2020]. 

In this dissertation, I seek to advance this related scholarship by theorizing and measuring the likelihood that military and civilian actors engage in distinct forms and patterns of political behaviors which involve the military. The central argument emphasized throughout this dissertation is that the patterns of political behavior and the actors who engage in them are largely influenced by key features of the domestic environment in which these actors operate. In particular, the degree of _political polarization_ prevalent in society, and the degree to which the military institution is _prestigious_, combine in such a way so as to shape the ways in which military and civilian actors behave with respect to the military.   

In making this argument, I build upon a general "motives" and "opportunity" framework first advanced by Finer (1962) and later by Taylor (2003) to describe military intervention in politics [@finer_man_1962; @taylor_politics_2003]. I claim that the relative levels of polarization and prestige serve similar functions in that each influences the degree to which both sets of actors adhere to what I define as the the central principles of civil-military relations, which are derived and explained later in this chapter. By conducting a systematic exploration of the impacts of polarization and military prestige on civil-military relations, across time and several domestic contexts, this dissertation sheds light on what is, I argue, an under-appreciated task: the changing difficulty that both civilian and military leaders face in harnessing, managing, and orchestrating their own behavior as it pertains to the military, as well as possessing awareness of how the other actor is inclined to behave. 

Examining the impacts of polarization and military prestige on political behaviors that involve the military is an important undertaking from both normative and empirical perspectives. From a normative perspective, this study helps observers and participants better understand why the conduct of civil-military relations can seem relatively smooth or, on the contrary, relatively turbulent, during different periods in history. Scholars often examine the particular leaders who are involved in particular moments or crises, such as wars, as well as the specific context surrounding the conduct of wars, to understand why civil-military relations were or were not troubled as the war proceeded [for just two of the countless possible examples, see @mcmaster_dereliction_1997 regarding the US during the Vietnam War; @kaiser_no_2014 regarding the US during World War Two]. Certainly, scholars are right to place significant emphasis on these things. 

But I also contend that the story is incomplete if we do not give adequate attention to the broader, long-running, political factors that are examined in this dissertation, and their impact on civil-military relations. If for a moment the conduct of civil-military relations can be viewed analogously to a sporting event, then the players who take the field are the civilian and military actors who engage in the conduct of civil-military relations. This dissertation, however, attempts to do something a little different in that a significant part of it examines the stadium or the field - _the context_ - in which the players compete. At the risk of sticking with the sporting analogy for just a bit more, this dissertation sheds light on the fact that just as different weather conditions (for example, accumulating snow at a football game, or whether a pitcher in baseball has to throw in the rain or wind) may influence how athletes perform during a game, so too do different domestic conditions influence how civilian and military actors behave.

Yet there is much at stake empirically as well. By analyzing certain forms of behaviors over time, particularly as the conditions of polarization and military prestige change, scholars are better positioned to gain an accurate awareness for just how significant the roles of political polarization and military prestige are in shaping civil-military relations, and in particular, the behaviors of both military and civilian leaders. It is one thing to acknowledge that high polarization and high military prestige make the conduct of civil-military relations more challenging, but another to demonstrate just how much more challenging, and with respect to what specific, or at least types of, behaviors. It is my hope is that these normative and empirical perspectives compliment each other.

This introductory chapter of the dissertation proceeds in three main parts. First, it introduces the general field of civil-military relations. In doing so, emphasis is placed on identifying and expositing three central principles of American civil-military relations: that of civilian control of the military, that of non-partisanship of the military institution, and that of non-interference of the military into certain realms of state policy making. The second part of the chapter describes how the intended dissertation topic fits within the scholarly literature and the gaps it intends to fill. The third and final part of this introductory chapter describes the general plan of and the methodology employed throughout the dissertation   
# The Field of Civil-Military Relations

Truly one of the interdisciplinary topics of scholarly research, civil-military relations spans the fields of political science, history, sociology, and ethics. The study of civil-military relations also covers an incredibly rich and immense set of questions involving enormous topics such as the relationship between militaries and governments, the relationships between militaries and societies, strategic and military effectiveness, and the occurrence of coups. Within the field of political science in particular, one of the areas of civil-military relations that is most studied is the relationship between prominent state officials and military leaders. Indeed, the relationship between these actors is what one scholar has termed, "leadership at the state's apex" [@brooks_risa_integrating_2019, 380], as it is here that numerous weighty decisions that impact national security are made, including whether to go to war, how to form and implement effective strategy, how to prepare for the next conflict, and how large the military and its budget should be. 

In the most basic sense, the study of civil-military relations explores aspects of the question of how to ensure that a military is strong enough to defeat threats to the state without threatening the state itself. One scholar of civil-military relations, Peter Feaver, refers to this central question of civil-military relations as the "problematique," and notes that this dilemma is "vexing," and a great "paradox" in that the military must be "strong enough to prevail in war," yet submissive to authority such that it cannot and will not "destroy or prey on the society it is intended to protect" [@feaver_civil-military_1996; @feaver_civil-military_1999, 214]. As a result of the twin facts that a state needs a military to defeat external threats and that militaries capable of doing so are at least to some degree powerful and dangerous, civil-military relations are always filled with a degree of tension.  

A number of scholars have offered various solutions for reducing the tensions that are inherent within the "problematique" Feaver describes, but none has been more lasting or consequential than the late Samuel Huntington, whose 1957 work, _The Soldier and the State: The Theory and Politics of Civil-Military Relations_, often serves as the point of departure for a number of debates within the field of civil-military relations. Although Huntington's theory has endured important and substantive critiques over the years [for example, see @janowitz_professional_1960; @finer_man_1962; @cohen_supreme_2003; @feaver_armed_2003; @brooks_paradoxes_2020], it is essential to briefly exposit Huntington's conception for achieving effective civil-military relations, and what lies at the heart of it - a theory of civilian control of the military that maximizes what Huntington called "military professionalism" [@huntington_soldier_1957]. 

In the section to follow, I briefly summarize Huntington's argument and logic, as well as several prominent critiques of his work. The purpose of summarizing both Huntington _and_ his critics is to argue that in spite of the important differences between them, Huntington and his critics share much common ground regarding the aspirational prominence of three central principles, or standards, that help foster healthy and effective civil-military relations. These principles, which I articulate and describe in the section to follow, are the principles of civilian control of the military, the "non-partisanship" of the military institution, and what I will term the "non-interference" of the military into realms of politics that are the _prerogative_ of duly-elected and appointed civilian leaders. These related principles, each of which is the subject of a fair bit of scholarship, are distinct and hold different normative implications for behaviors committed by both civilian and military actors with respect to the military institution.  

## Huntington and The Soldier and the State

Though it is not always framed as such, _The Soldier and the State_ is a book that lies at the very intersection of domestic politics and international relations. The fundamental problem that Huntington sets out to solve is how liberal democracies such as the United States can balance the simultaneous demands of maintaining military security and its liberal political character. Writing after the Korean War, when the danger confronting the United States as a result of the Cold War were quite tangible, Huntington argues that American security requirements constitute a "functional imperative" in that they require the US to maintain a military that is sufficiently strong and capable of defeating the external threats posed by the Cold War [@huntington_soldier_1957, 1-3]. At the same time, Huntington argues that the liberal character of American society generate a "societal imperative" such that the pursuit of American security requirements could never cause the US to deviate from its liberal political ideology in fundamental ways, including abandonment of the principle of civilian control of the military [@huntington_soldier_1957, 1-3]. In this way, Huntington's puzzle is essentially an earlier framing of the "problematique" pointed out by Feaver decades later: Huntington aims to solve how to render the military strong enough to protect the United States, but sufficiently subservient and obedient to civilian power [@feaver_civil-military_1996]. 

Huntington forcefully argues that achieving the simultaneous demands of maintaining the state's security and it's liberal political character is made more difficult by the fact that the American military officer corps is different from the rest of American society in several important ways -- a claim that is not without controversy. In particular, Huntington asserts that the military officer corps holds to a distinct "mind" and "ethic" that views conflict as part of a "universal pattern throughout nature," and thus, that "the military view of man is...decidedly pessimistic" [@huntington_soldier_1957, 62-63]. These views, argues Huntington, clash with the generally liberal views cherished by the majority in American society, views which emphasize that the "natural relation" of mankind is one of peace rather than war [@huntington_soldier_1957, 90]. In short, the gulf between America's liberal political character and the ethic of professional militaries is significant, and accounts for the strong suspicion and skepticism Americans have had for standing military forces since its founding [@huntington_soldier_1957, 143-151]. 

Huntington argues that these two requirements - securing the United States during the Cold War and modern warfare in general as well as maintaining America's liberal political character - pose a fundamentally unique problem for the conduct of American civil-military relations. From the perspective of national security, sizable standing military forces are a requirement that is likely to endure. At the same time, the United States public would never agree to discard or abandon its liberal political essence. 

Huntington's solution for solving these twin dilemmas was adoption of a scheme of civil-military relations characterized by what Huntington called "objective civilian control" of the military [@huntington_soldier_1957, 83]. Such a scheme, in essence, requires that civilian political leaders grant the military a high degree of professional autonomy, which Huntington claims will facilitate both the military's development of its professional expertise and skills, and its eschewal of partisan politics - a claim that is not without controversy, and which is discussed further in this chapter. In practice, achieving "objective civilian control" of the military, according to Huntington, requires recognition by both military and civilian actors that each operates within a distinct sphere of territory [@huntington_soldier_1957, 83-85]. Civilian leaders are not to encroach into areas that are the military's purview, which would damage the military's professional autonomy, and conversely, military leaders should not enter the world of partisan politics, as doing so subverts civilian control and the apolitical nature of the military.

A key aspect of Huntington's logic that undergirds the concept of objective civilian control centers on the US military officer corps as a professional body. According to Huntington, professions are distinct from other vocations or jobs in that they possess several unique attributes, including that of "expertise, responsibility, and corporateness" [@huntington_soldier_1957, 8-10]. With respect to the attribute of expertise, Huntington has in mind the idea that military professionals possess unique "knowledge and skills" in the arena of warfare [@huntington_soldier_1957, 8-10] and, using Harold Lasswell's terms, are experts in the "management of violence" [Lasswell quoted in @huntington_soldier_1957, 11]. Regarding the attribute of responsibility, Huntington has in mind the idea that the military officer corps is to employ such expertise only in the service and "at the direction" of the state [@nielsen_american_2012, 370]. And with respect to the attribute of corporateness, Huntington sees the military officer corps as a group that certifies its own leaders and polices itself, or in the words of Nielsen, "as a distinct, bureaucratized body, with a common identity fostered through shared educational, training, and service experiences" [@nielsen_american_2012, 370].^[Huntington's argument is dependent on the notion that the US military, and its officer corps in particular, constitutes a profession. Among contemporary scholars of civil-military relations, this point is not significantly disputed, even as they warn, rightly in my view, that to remain a relevant profession is never an automatic guarantee, but rather a perpetual challenge that requires deliberate and perpetual effort on the part of both soldiers and civilians. For one scholar's warning regarding this dynamic, see @snider_once_2012.]

As a professional body seeking to further its autonomy, the military officer corps would welcome the concept of objective civilian control, Huntington argued. But he also believes that civilians would welcome the concept, as it would render a professional military that is exceedingly skilled and capable.^[Huntington asserts that the development of the US military as a profession largely occurred after the US Civil War, and was the function of the isolation experienced by the military officer corps. The timeline of this claim, as well as the supposed isolation of the US military officer corps that spurred professionalization in the late 19th Century, has been rejected by many military historians [see also @coffman_long_1991]. Those who reject the timeline offered by Huntington assert that the US military exhibited real signs of professionalization well before the post-US Civil War Era. For instance, Skelton (1992) argues that the US military made substantive strides towards professionalization beginning after the War of 1812 [@skelton_american_1992; see also @heiss_professionalization_2012]. Grandstaff (1998) as well as Connelly (2005) take a more nuanced view, arguing that the process of military professionalization occurred in two distinct waves during the 19th Century, one before and one after the Civil War [@grandstaff_preserving_1998; @connelly_american_2005]. In chapter five of this dissertation, which involves the post-US Civil War military and its leaders, I substantiate that even in the process of professionalizing, military leaders and officers in particular both understood and assented to the importance of basic principles of civil-military relations, including that of civilian control and an avoidance of engaging in overtly partisan politics.] Ultimately, the posture or arrangement of objective control, so claimed Huntington, would lead to a civil-military relationship that simultaneously does not significantly taint the liberal political character of the country, defends the nation from external threats, and most importantly, maintains the principle of civilian control of the military [@huntington_soldier_1957, 83-85].

## Critiques of Huntington

Critics have assailed Huntington's concept of _objective civilian control_ for years, and it would be impossible to list all of those critiques here. However, perhaps they key theme of these critiques centers around the stark separation between figurative military and political spheres - autonomy - favored by Huntington. In levying such a critique, Huntington's critics often point to and assert Clausewitz's famous dictum that war is a "mere continuation of policy by other means" [@clausewitz_war_1976, 69, 605]. Because the military is an inherently political institution that cannot exist and operate absent from political influence, so contend the critics of Huntington, the concept of objective civilian control, which relies on the separation between civilian and military spheres, should not be pursued. 

Some critiques of Huntington contend that the design of objective civilian control is failed from the beginning. Critics who hold this view, such as Cohen and Feaver, argue that a stark divorce between the duties and responsibilities of statesmen and soldiers can never achieve maintenance of the principle of civilian control. The central idea of this critique is that, especially in democracies, civilian leaders are legally and authoritatively superior to military leaders, and thus, must hold ultimate responsibility for what militaries do and fail to do. Eliot Cohen argues forcefully that the world's greatest heads of state have never abdicated responsibilities that concern the conduct of war, and at times [@cohen_supreme_2003]. Peter Feaver has parsimoniously captured the essence of this critique of Huntington by noting that the principle of civilian supremacy over the military means that "civilians have the right to be wrong" [@feaver_right_2011, 117; @feaver_armed_2003]. 

Other critiques of Huntington have arrived at a similar conclusion, but by focusing on the practical shortfalls of Huntington's theory. For instance, critics such as Brooks (2020) warn that the pursuit of Huntington's theory of objective control leads to some military officers developing "blind spots," such that these officers are led to engage in detrimental political actions under the rationalization that because they are "professional" officers, their actions are apolitical by default [@brooks_paradoxes_2020; 17]. Others warn that facilitation of a strict separation between military and civilian spheres in practice, and especially during wartime, fails to recognize the degree to which military and civilian spheres must overlap in order to develop, implement, and achieve the goals of national security policy [@rapp_civil-military_2015].   

In spite of these insightful and important concerns and critiques of Huntington, Huntington retains an out-sized influence in the field of civil-military relations, even today. As Nielsen has argued, any reputable course taught on civil-military relations likely includes Huntington's _The Soldier and the State_, not only because of the "boldness and ambition" of Huntington's work, but also because Huntington was truly the first serious scholar who sought to develop a comprehensive theory to help steer the conduct of civil-military relations [@nielsen_american_2012, 369]. Moreover, in my view, many of the issues and discussions within US civil-military relations today - how to maintain civilian control of the military, how separate should the military be from society, how much overlap between civilian and military spheres should there be - are all topics that Huntington, albeit perhaps with error, addressed in some way. 

Furthermore, there is much that Huntington and his critics agree upon, in my view. In particular, I contend that Huntington and his critics agree on the centrality of three central principles that should govern civil-military relations, even if these critics disagree on the ways in which these principles should be achieved and maintained. These three central principles include that of civilian control of the military, the non-partisanship of the military institution, and a principle of "non-interference" of the military into certain areas or realms of state policy making.

The most obvious principle upon which Huntington and his critics both agree is that of ensuring civilian control of the military. Indeed, one of the two chief purposes of Huntington's formulation of the concept of _objective control_ (the other being the ability of the military to defeat external threats) in the first place is to ensure civilian control of the military. In the same way, the critics of Huntington assail his argument on the grounds that such a concept fails to ensure civilian control. For example, Eliot Cohen's descriptive theory of civil-military relations, which he has termed "the unequal dialogue" [@cohen_supreme_2003, 208-224], and Feaver's contention that "civilians have a right to be wrong" [@feaver_right_2011, 117; @feaver_armed_2003], both establish the principle of civilian control as the most essential characteristic of healthy civil-military relations.^[Cohen argues strongly that the most successful wartime civil-military relations have been those that have involved "an unequal dialogue - a dialogue, in that both sides expressed their views bluntly, indeed, sometimes offensively, and not once but repeatedly - and unequal, in that the final authority of the civilian leader was unambiguous and unquestioned..."[@cohen_supreme_2003, 209]. Cohen, as is the case with other scholars of civil-military relations who have critiqued Huntington, is concerned that Huntington's concept of objective control prevents civilian leaders from exercising appropriate oversight of the military.]   

A second principle that Huntington and his critics share is that the military stay out of, or at least minimize its impact within, the world of partisan politics. Nielsen, for instance, acknowledges that "Huntington's principle of objective control has both merits and shortcomings. On the positive side, it preserves democratic control, speaks to the importance of an apolitical military and protects military professionalism" [@nielsen_american_2012, 375]. And Cohen cautions against an outright dismissal of Huntingtonian thought, warning that "to reject Huntington's ideas of sequestering issues of policy from those of military administration or operations is to open the way to a military that is politicized and, by virtue of its size and discipline, a potentially dominant actor in the conduct of foreign and internal affairs" [@cohen_supreme_2003, 264]. And in article appropriately titled, "Military Officers: Political without Partisanship," Mackubin Thomas Owens echoes a similar point made by both Brooks (2020) and Rapp (2015), namely, that military officers need to engage with and understand the political process, but "without becoming swept up in partisan politics" [@owens_military_2015, 97]. Indeed, while Huntington and his critics disagree on exactly how stark the separation between military and political spheres should be, both share a concern that the military avoid inappropriate partisan political entanglement. 

Finally, there is a third principle upon which Huntington and his critics share common ground. However, this principle is one that is implied by Huntington and his critics, rather than one explicitly articulated. This principle involves the _non-interference_ of the military, and centers on the notion that the military should not interfere or seek to influence all realms of state policy making. In introducing this principle of _non-interference_, I am underscoring the fact that while Huntington and his critics _do not_ agree that a stark separation should exist between political and military realms, the critiques levied against Huntington are one directional in that they allege that under Huntington's concept of objective civilian control, civilians will not wield sufficient and appropriate influence over the military. However, the critiques of Huntington on this point are never made from the opposite direction; that is, critics of Huntington do not argue in favor of the military holding greater sway and influence into the realm of domestic politics, and particularly, those areas of domestic politics that are indirectly related or entirely unrelated to military affairs. Graphically, Huntington imagined two largely separate circles, one political and one military, while his critics retort that civilian control can only be properly achieved if the military circle rests entirely or at least largely within a larger political circle. My point is that both Huntington and his critics would be against a smaller political circle residing within a larger military circle, which perhaps exists in states that have military dictatorships. In fact, Cohen warns that an outright dismissal of Huntingtonian theory, especially "in states with less-established democratic traditions _(than the United States)_...would open the path to direct military intervention in politics" [@cohen_supreme_2003, 264, emphasis mine].

In summary, there is much that Samuel Huntington and his critics disagree about concerning civil-military relations. Huntington advocated for separating the spheres of military operations and politics as much as possible on the basis that the military profession would be interested in pursuing its own autonomy, which included further developing its unique military expertise. His critics, on the other hand, have pointed out that there are limits to how separate military affairs and politics can reside, both in theory and in practice, and that therefore, the notion of military expertise and the concept of a military as a profession should never exist entirely separate from political context. In spite of these critiques, I have demonstrated that at least in very broad terms, Huntington and his critics share a concern for at least three overarching principles of civil-military relations. The first is a principle of civilian control of the military. The second is a principle that the military, even if it is a political actor, should avoid partisan influence and entanglement as much as possible. And the third principle speaks to the idea of the military not interfering in realms of state policy making that are entirely unrelated or indirectly related to the conduct of military affairs.  

# Three Central Principles of Civil-Military Relations

In the section below, I define and exposit in greater detail each of these three central principles of civil-military relations. Together, I argue that these three principles - civilian control of the military, the non-partisanship of the military institution, and the "non-interference" of the military into certain areas of state policy making - constitute a _baseline_ set of principles that inform, guide, and facilitate the conduct of healthy and harmonious civil-military relations. Both actor types, military and civilian, have a role in adhering to these principles. Ultimately, these principles are important for the study and practice of civil-military relations because when when one or more of these central principles is disregarded or violated by one or both actors, then a state's civil-military relationship is likely to be discordant in some way.  

## Central Principle 1: Civilian Control of the Military

The first central principle of civil-military relationships is that of civilian control of the military. It is listed first deliberately, as it is the central component by which the health of civil-military relationships, particularly in democracies, is maintained. Without strong adherence to the principle of civilian control of the military, it is doubtful that a democratic state's civil-military relationship can be healthy nor successful in the long run. So, what is civilian control of the military, and how does one know if military and civilian actors within a state are adhering to it as a principle of civil-military relations? 

On the surface, the principle of civilian control is fairly straightforward: civilians should be in charge of the military. Yet a significant scholarship has pointed out that focusing on coups or other extreme forms of insubordinate military behavior fails to recognize the myriad other ways in which militaries often challenge the principle of civilian control [@croissant_beyond_2010; see also @cohen_supreme_2003, 242; @beliakova_erosion_2021]. Finer, for instance, warns that militaries can and will violate the principle of civilian control through "acts of commission, but also by acts of omission" [@finer_man_1962, 20].  For this reason, Feaver argues that observers interested in the health of the principle of civilian control should examine the "patterns" of civilian control, rather than merely looking for whether the principle exists within a state [@feaver_civil-military_1996, 167]. 

Thus, while civilian control may at first glance seem is thought it is a straightforward concept, it is in actuality a massive and all-encompassing one. Scholars share the idea that civilian control more accurately refers to the "relative political power" that exists between a nation's armed forces and its civilian leaders [@bruneau_civil-military_2019, 7; see also @brooks_shaping_2008]. Brooks, Golby, and Urben argue more precisely that civilian control refers to "the extent to which political leaders can realize the goals the American people elected them to accomplish" [@brooks_crisis_2021, 65]. 

Thus, in my view,  the degree to which the military adheres to the principle of civilian control involves both outcomes (does the military do what it is told) _and_ process (what is the spirit of the military in doing what it is told). This focus on process is important. Military leaders may ultimately obey the orders of their bosses, but in the process engage in a range of behaviors that thwart, stymie, and/or frustrate the will and policy objectives of elected civilian leaders, all of which harm the principle of civilian control. 

The behaviors or actions that challenge the principle of civilian control range from the very subtle to the very obvious. For example, as Brooks, Golby, and Urben argue, military officers may choose to share little information with civilians about an issue, or comply with a civilian directive at a leisurely pace, rather than with spirited initiative [@brooks_crisis_2021]. In these types of cases, it is possible that civilian leaders and the public will never know that the military is willfully challenging the principle of civilian control! Other types of behaviors, which are discussed more in chapter two of this dissertation, are far more obvious. For example, an Army general who writes an opinion piece strongly criticizing the President's foreign policy views challenges the principle of civilian control in that it likely undermines popular support for the President, and because it imposes a cost that the President now has to deal with, particularly if the military officer who authored the piece is a popular figure.  

Can civilian leaders also violate the principle of civilian control? I contend that they can, but that when they do, it is not typically the result of a single action or behavior that carries immediate, tangible consequences. Rather, it is more typical that civilians violate the principle of civilian control by failing to ensure that an overall climate of civilian control is established and maintained. This is consistent with recent scholarship by Beliakova (2021), who argued that one pathway through which the "erosion" of civilian control occurs is "deference," that is, by civilians delegating too much power to the military [@beliakova_erosion_2021]. This can be the result of civilian failure to assert themselves sufficiently in an overall civil-military dialogue, perhaps during the course of a major military operation such as a war, or over the course of a lengthier time horizon. In the contemporary United States, for example, several critics have expressed concern that the principle of civilian control has been threatened for the past several years as a result of an extremely slow and politically charged process of the confirmation of senior civilian Department of Defense appointees, resulting in a shift in the overall balance of power within the Pentagon towards the uniformed military [@seligman_civilian_2020]. 

Finally, it should be pointed out that in the United States, the principle of civilian control is embedded within the US Constitution, which effectively establishes that the principle of civilian control involves the military obeying two civilian bosses. The first is the President, who is endowed as the Commander in Chief of the Armed Forces [@united_states_us_1787, Article 2, Section 2], and the second is the Congress, which the Constitution grants powers of oversight [@united_states_us_1787-1, Article 1, Section 8]. Thus, from the US military's perspective, adhering to the principle of civilian control involves appeasing two different bosses - a prospect that is undoubtedly difficult at times. 

## Central Principle 2: Non-Partisanship of the Military Institution

The second central principle of democratic civil-military relations, particularly in the US context, is the principle of non-partisanship of the military institution. In general terms, the principle of non-partisanship of the military stipulates that the military not align itself - nor that it be made to align by civilians - with a particular political party or its platform. The principle of non-partisanship means in part that the military fully obeys the lawful orders of whichever political party is in office, yet it also means that the military cannot and will not identify itself, nor be made to identify by civilian leaders, as a partisan actor. Thus, both sets of actors, civilian and military, have a responsibility to maintain and adhere to the principle of non-partisanship of the military. 

However, determining the degree to which civilian or military actors uphold the principle of non-partisanship in practice is a difficult undertaking, largely because it is difficult to both identify and measure what forms of behavior constitute a violation of this principle. After all, if militaries are, in fact, as Clausewitz and the critics of Huntington have contended, inherently political creatures who "serve at the pleasure" of their civilian bosses, then it stands to reason that the military will, at some point, enact the partisan policies, wishes, desires, and goals of their elected civilian leaders, who belong to a political party [@mullen_chairman_2011]. 

For this reason, scholars of civil-military relations have often separated the various meanings and implications of certain forms of behavior and their impact on politics. While there is no clear delineation of which forms of behavior and which contexts violate the principle of non-partisanship of the military, recent scholarship has attempted to create figurative space to allow greater acceptance - even encouragement - of military officers engaging in _some_ forms of political behavior. For instance, scholars have argued that military officers and civilians should expect military leaders to increasingly contribute to debates about security policy on the grounds that doing so will lead to the better formulation of strategy and thus, to better national security outcomes [@brooks_perils_2013; @owens_military_2015; @rapp_civil-military_2015]. In other words, scholars and senior military officers recognize that the military needs to be part of and embrace its role within the national security policy-making process. 

Simultaneously, scholars and military officers alike recognize that other forms of behavior, such as retired military officers speaking at political conventions on behalf of political candidates, comprise blatant violations of the military's non-partisan ethic [@dempsey_military_2016]. Thus, there is agreement that while there may be room for the military to engage in some political behaviors, there is a point at which other behaviors cross a figurative line such that they are not merely political in nature, but rather partisan in nature. It is at this juncture that such behavior becomes problematic.  

Similar logic holds for civilian leaders with respect to the principle of non-partisanship of the military. In short, civilian leaders, according to the principle of non-partisanship, should not use the military in overtly partisan ways. Admittedly, there is far less clarity for civilian actors concerning when such usage of the military becomes inappropriate, precisely because of the fact that civilian leaders are elected to fulfill partisan campaign promises! 

Several examples, nonetheless, are instructive. Many scholars and former military officers blasted the decision of President Trump to deploy US Troops to the US southern border in 2018, citing the move primarily as a political stunt ahead of the 2018 midterm elections [@adams_trumps_2018]. Other critics alleged that President Trump's threat to use active duty forces to dispel protesters and rioters in the summer of 2020 likewise violated the military's non-partisan ethic [@brooks_dismay_2020]. Finally, consider the case of President George W. Bush and his administration employing a number of senior retired military officers to boost popular support during a period of the Iraq War when both his policies and he personally were relatively unpopular. A fascinating yet somewhat disturbing picture emerges from Barstow (2008) as he describes efforts by the Bush Administration to court retired military officers for a major public opinion campaign. In particular, Barstow alleges that the Bush Administration politicized the latent popularity of dozens of retired military officials by first, warming up to them, and second, by prodding them to speak favorably to the media over several controversial issues, including the detention facility at Guantanamo Bay and the broader "surge" strategy to Iraq [@barstow_behind_2008]. 

The degree to which each of these examples constitute mainly civilian violations of the military's non-partisan ethic, or were simply indicative of politics as usual, is the subject of considerable civil-military relations debate.^[I do not provide an all-encompassing rubric regarding the extent to which certain behaviors are and are not more partisan than other behaviors in this dissertation. That said, I do defend in great detail, both in Chapter 2 and Chapter 4 of this dissertation, the decision to examine the use of military symbols and actors during Presidential campaign advertisements as a clear indication of how partisan the military, or certain military actors, appear in particular political races.] What is clear, however, is that both sets of actors - military and civilian - have a responsibility to uphold the principle of non-partisanship, each by not aligning the military with overtly partisan goals and aspirations. However, where the line between appropriate political involvement and inappropriate partisan involvement of the military should be drawn is not always clear. This is especially true when examining the behavior of civilian leaders because they, by their very nature, are partisan entities. Still, the principle of non-partisanship, at least in general terms, applies: citizens should not be able to look at the military and to easily detect what their partisan preferences are. 

## Central Principle 3: The "Non-Interference" of the Military 

Finally, a third central principle of American civil-military relations which I argue both Huntington and his critics affirm is what I am terming the _"non-interference"_ of the military into certain realms or areas of policy. This principle, I argue, is strongly implied in both Huntington's argument and critiques of his work. In short, the principle of non-interference simply means that there are policy arenas which are and should remain "off limits" for US military actors. By off limits, I mean that the military should not seek to influence some areas of government policy. 

The origins of the principle of non-interference can be directly traced to Huntington's conception of military expertise. Recall that for Huntington, the concept of military expertise and military professionalism are inextricably linked, and involve the unique knowledge, skills, and behaviors that are relevant to the military officer corps [@huntington_soldier_1957, 7-18]. Huntington later devotes an entire chapter to the development of the notion of a unique and distinct "military mind," which to him consists of the "values, attitudes, and perspectives which inhere in the performance of the professional military function and which are deducible from the nature of that function" [@huntington_soldier_1957, 61]. In short, Huntington is arguing that the uniqueness and distinction of the military - in terms of its mindset, corporate ethic, and values - stems from the fact that the military's fundamental purpose is to defend the nation from external threats. The implication here is simple but profound: a professional military is one that is used only to fill roles that are directly related to national defense and/or the security of the nation. 

And US history, certainly in the 20th Century at least, proves that this has indeed been the case. The US military has been primarily used to fight external rather than internal threats. Thus, there has been little if any need for the US military officer corps to assume any type of political role in managing the state's domestic problems. To be clear, the American military has certainly been used for a variety of non-traditional types of operations in the nation's history, and actually, quite often, from desegregating schools, responding to natural disasters, and most recently, assisting Federal authorities in responding to the Coronavirus Pandemic. But these types of missions, while frequent, have not come anywhere near to comprising the bulk of the US military's efforts.

Critics of Huntington likewise support a principle of non-interference. According to these critics, who criticize Huntington for separating military and civilian spheres too starkly, civilian leaders may weigh in, influence, and ultimately oversee all things that involve military affairs. But these critics never argue the reverse!  That is, Huntington and his critics never argue in favor of the military influencing all areas of civilian policy making. Indeed, both Huntington and his critics would find such an idea reprehensible and flawed.

It is also important to note that the literature within comparative civil-military relations very much describes the principle of non-interference, even if in a descriptive way to denote that in many countries, such a principle does not exist! Indeed, militaries in many parts of the world influence and sometimes direct policy areas of the state to such an extent that doing so in the US context would be unthinkable. This scholarship includes explorations of the varying mission sets that militaries around the world conduct [for instance, see @vennesson_is_2009; @pion-berlin_democratization_2005; @bruneau_civil-military_2019], and the degree to which certain militaries are used to conduct internal state security missions such as domestic counter-terrorism vice external warfare [@stepan_new_1973; @pion-berlin_civil-military_2020]. For instance, Stepan's concept of "the new professionalism of internal security and national development" traces the development of the Brazilian military's role expansion into domestic affairs as a result of having to primarily confront internal rather than external threats [@stepan_new_1973].

But especially in democratic countries, a principle of "non-interference" of the military seems to be firmly established. This is true even as the exact boundaries of separation regarding which policy areas are off limits for military influence are not the same across multiple democratic countries. In his excellent work on Israeli civil-military relations, Yehuda Ben-Meir (1995) separates the affairs of the state into four broad areas, including political affairs, domestic affairs, national security, and the armed forces [@ben-meir_civil-military_1995]. He argues that civilians should and do influence all four of these areas of politics, whereas the military should influence the three areas of domestic affairs, national security, and the armed forces [@ben-meir_civil-military_1995].^[In terms of the activities which constitute political affairs, Ben-Meir includes items such as "taking control of the government (coups), influencing political appointments, or interfering in the decision making process. See @ben-meir_civil-military_1995, 4-5 for an excellent description and diagram.] Ben-Meir is addressing the Israeli context, and thus, his paradigm may not map squarely onto the US context. This is not problematic, however. What is important for the purposes of my argument here is to highlight and to acknowledge, as Ben-Meir has, that the relative scope and specific areas of state policy that are influenced by military leaders is not the same as that of civilians.

Finally, it is important to briefly address what the principle of non-interference of the military implies for the conduct of the US military. The principle of non-interference implies that the military will not seek to perform, and that civilians will not assign the military to perform, roles or missions for which military forces are not suitably designed, nor those which are tangentially or indirectly related to national defense, except in cases of great crisis or need. Thus, both civilian and military actors have a responsibility to adhere to the principle of non-interference. 

A few examples are illustrative of this dynamic. Hypothetically, most Americans, as well as the military, would probably not want a President to place the Pentagon in charge of designing a plan to overhaul social security, nor would they want the Chief of Naval Operations advocating that the Navy solve this problem for the nation. Non-hypothetical examples also abound. When the Chairman of the Joint Chiefs of Staff, General Mark Milley, foresaw no role for the military in the 2020 Presidential Election, he was implicitly invoking the principle of "non-interference" of the military, implying that it was not appropriate for the military to become involved in settling electoral disputes [@silva_gen_2020]. Furthermore, across the world, the role of militaries in responding to the COVID-19 pandemic has raised numerous normative concerns over whether the military should occupy the roles that they have, including in the provision of healthcare and logistics, to include the distribution and contracting of vaccines.^[At one point in late 2020, an active duty four star general apologized to the nation for a mix-up in information regarding the distribution of COVID vaccine, leading some critics to express concern that the military was making inherently political decisions that, from a normative perspective, posed some problems. See @passy_gen_2020 for more details.] 

## Summary of Three Central Principles of US Civil-Military Relations

In summary, Samuel Huntington and his critics both agree on the primacy and the importance of three central principles of civil-military relations. These principles are that of civilian control of the military, the non-partisanship of the military institution, and the "non-interference" of the military into certain arenas of state policy. When adhered to, followed, and respected, these principles generally constrain the behavior of both civilian and military actors in important ways, and have helped to facilitate relatively harmonious civil-military relations. 

Table \@ref(tab:principles) captures these three principles, a concise definition of each principle, the actor that can violate each principle, and a few pertinent examples of behaviors that violate each principle.  

```{r principles, include=TRUE, echo=FALSE}

text_tbl1 <- tribble(~"Central Principle", ~"Description", ~"Military Example of Violation", ~"Civilian Example of Violation",
  "Civilian Control", "Civilian Political Goals are Actualized and Implemented; Mechanisms of Civilian Oversight Function; No Overt Military Insubordination", "Resigning in protest of policy; slow-rolling policy implementation; authoring an op-ed that criticizes a President's policy preferences", "Failing to establish mechanisms and processes of oversight; delegating too much power to the military",
  "Non-Partisanship",  "The Military Institution Exists and Operates outside of Partisan Politics; Military Actors Fully Obey Political Leaders, and Do Not Advocate for Partisan Policies, People, or Platforms", "Advocating for the platform of a political party or denouncing that of another; Declaring candidacy for partisan political office while in uniform", "Urging several military generals and admirals to speak at a party political convention",
  "Non-Interference", "There are Areas or Realms of State Policy making into which the Military does not Enter or Seek to Influence", "Advocating that the President place the military in charge of overhauling social security", "Appointing a serving unformed military officer as the Secretary of Labor or Education")
 

text_tbl1 %>% kable(booktabs=T, caption="Central Principles of US Civil-Military Relations") %>%
  kable_styling(full_width = F, font_size=10, latex_options="striped" ) %>% 
  column_spec(1, width="6em") %>% 
  column_spec(2, width="12em") %>% 
  column_spec(3, width="12em") %>% 
  column_spec(4, width="12em") 
  

```

# The Gap in the Scholarly Literature

Having established the the baseline principles of civil-military relations as that of civilian control, non-partisanship, and non-interference, as well as the central role these principles play with respect to the facilitation of healthy democratic civil-military relations, I will briefly trace arguments made by scholars who seek to explain the military's role in politics. The goal of doing so is two-fold: first, to orient the reader to the value of this scholarship, and second, to explain the gap that currently exists within the scholarship. 

Explaining why and how militaries intervene in politics is a massive topic within civil-military relations [for a sampling of this scholarship, see @finer_man_1962; @taylor_politics_2003; @croissant_beyond_2010; @bove_beyond_2020; @beliakova_erosion_2021]. Some works focus on specific types or forms of military intervention in politics, such as coups [for instance, see @horowitz_coup_1980;], while others focus on explaining a range of intervention outcomes that can occur within a particular country or region of the world [for instance, see @stepan_new_1973;  @fitch_armed_1998;]. I first summarize the contributions of five particular scholars whose works, in my view, are highly relevant to the topic this dissertation addresses. These are Samuel Finer's _The Man on Horseback_ (1957); Donald Horowitz's _Coup Theories and Officer Motives: Sri Lanka in Comparative Perspective_ (1980); J. Samuel Fitch's _The Armed Forces and Democracy in Latin America_ (1989); Bove, Rivera, and Ruffa's _Beyond Coups: Terrorism and Military Involvement in Politics_ (2020); and Brian Taylor's _Politics and the Russian Army_ (2003). Then, I briefly describe several works that address political activities in mature democracies.  

Finer (1962) is among the first of scholars to provide a comprehensive theory for why militaries intervene in politics. His central argument is that militaries intervene in politics as a broad result of having the "motives and opportunities" to do so [@finer_man_1962, 20-76], a framework that is adopted in this dissertation and further explained in the next chapter. Finer delineates several classes of motives, arguing that these motives may include a desire for the military to protect or enhance its own bureaucratic interests, as well as the belief or worldview that the military should serve as the "custodian" of the "national interest," particularly in times of national crisis [@finer_man_1962, 30-31]. Importantly, Finer also is among the first of scholars to point out, in what is an insightful critique of Huntington that others have advanced in similar ways [for instance, see @brooks_paradoxes_2020], that a strictly Huntingtonian concept of military professionalism might actually compel, rather than prevent, the military from intervening in politics [@finer_man_1962, 22-23].^[The mechanism through which this may occur, argues Finer, is different from that posited by Brooks, however. Finer argues that the desire to remain professional may lead military officers to refuse to obey a particular civilian leader because he or she can claim that in doing so, he or she is actually serving the broader concept of the state's interest, rather than obeying a particular civilian regime. In the US context, however, this point of Finer's is perhaps not as pertinent because indeed, the US military officer's oath is to support and defend _the Constitution of the United States_, rather than to support or advance the agenda of a particular President. Brooks, in here 2020 article, "Paradoxes of Professionalism," describes more of a general mindset that is unwittingly adopted by the "professional" military officer in which he or she believes that on the basis of serving as a military officer, his or her conduct, regardless of its form, carries little or no political ramifications. The point here that both scholars are emphasizing is that the concept of professionalism can be abused by military officers in such a way that elicits the very conduct that Huntington seeks to circumscribe.] Finer's major contribution, in my view, is theoretical, as he is the first to posit the significance of an underlying "motives and opportunities" framework to explain military intervention in politics. The main shortcoming of Finer's work, however, is that while it is theoretically rich, his dependent variable of military intervention is somewhat imprecise. Indeed, the dependent variables Finer uses are the "levels of intervention" undertaken by the military, which range from asserting "influence" to engaging in "supplantment" [@finer_man_1962, 77-116]. These outcomes, and the range between them, include a significant amount of variation, however. While this range of outcomes is insightful, it does not facilitate granularity in examining the dependent variable. 

Horowitz (1980) advances an understanding of Finer's "motives and opportunities" framework by examining the factors that produced an attempted military coup in Sri Lanka in 1962. Using qualitative interview data from nearly two dozen military officers who participated in the coup attempt, Horowitz analyzes the motives of the participating officers. He concludes, convincingly in my view, that identifying a specific motive, in the specific instance examined but also likely in any other attempted coup, is a complex undertaking, and that multiple different officer motives, to include organizational interests, societal cleavages, or bureaucratic politics, likely exist simultaneously [see @horowitz_coup_1980; 3-30, 179-221]. His main contribution is theoretical and empirical in that he highlights the dynamic and complex nature of officer motives, and in a way, warns against adopting a uniform explanation for the occurrence of coups. The main drawback to Horowitz's work again involves the dependent variable, as he is focused squarely on the explanation of coups.   

A third scholar whose work touches on this dissertation is Fitch. Fitch seeks to explain how and why officers adopt unique beliefs regarding their normative roles in politics and within the government. Using rich interview data from the officers of several Latin American militaries in the 1980s, Fitch's develops the dependent variable of "military role beliefs," and argues that officers adopt a particular belief as the result of their military's organizational culture and lessons learned from previous political interventions by the military [@fitch_armed_1998, 61-105]. Fitch's dependent variable is different from that explored by either Finer or Horowitz, mainly in that it emphasizes the linkages between the beliefs adopted by an officer, and the general behaviors these officers engage in [see especially @fitch_armed_1998, 65-100]. This is an important contribution, in my view, because it explains how and why officers arrive at conclusions regarding how they should behave relative to civilians within government, and with respect to the principle of civilian control more broadly.  

A fourth scholarly work is Bove, Rivera, and Ruffa's (2020) _Beyond Coups: Terrorism and Military Involvement in Politics_. This work is important because it is among the first that explores the role of each type of actor, civilian and military, in spurring military involvement in politics. Using instances of terrorism as a theoretical starting point, the authors assert that military actors, by virtue of the information advantages they hold over civilians, "push" themselves into politics out of a desire to secure the state, while civilian leaders, who demand that the military employ its expertise by responding to terrorism quickly and successfully, "pull" the military into politics [@bove_beyond_2020, 268]. However, the dependent variable used in this study is a composite measure of the military's involvement in politics based on the International Country Risk Guide (IRCG). This measurement is fine for the intended use that the authors have in mind, but it is not a measurement of specific forms or types of political behaviors that occur.  

Finally, Taylor (2003) provides another important contribution by applying Finer's motives and opportunities framework to the case of Russia [@taylor_politics_2003, 2]. His impressive book, _Politics and the Russian Army_, explores 19 incidents in which the Russian military intervenes (or might have intervened) in state "sovereign power" issues [@taylor_politics_2003, 3]. Finer's contribution is important because his application of a motives and opportunities framework to a particular context is precise and sound. Moreover, his analysis covers an impressive period of time - more than three centuries!  

While the works of these five scholars are incredibly rich and valuable, there are two broad shortcomings that this dissertation seeks to address. The first is an issue related to the dependent variable examined in each study. In each of the studies described above, the dependent variables and concepts of military intervention (in the case of Fitch, role beliefs, which is different yet similar) are generally wider and more extreme than the instances of political behavior that occur within mature democracies. Therefore, I can (and do) seek to employ many of these scholars' theoretical staring points, including a general motives and opportunities framework, but in a way that seeks to explain meaningful variation in specific forms of political behaviors that occur within mature democracies. 

The second issue with the works described above, in my view, is that they tend to overlook the important role of civilian leaders in eliciting military intervention in politics. This is not the case for Bove, Rivera, and Ruffa (2020), as they accurately identify, in my view, the notion that civilian leaders seek to "pull" the military into politics at the same time that military leaders "push" themselves to do the same [@bove_beyond_2020]. The other works described above, however, tend to focus primarily if not exclusively on the behavior of military actors. This dissertation follows Bove and Ruffa (2020) in examining the simultaneous interplay that occurs between both sets of actors, civilian and military, but builds on their analysis by examining the role of domestic variables in explaining military intervention in politics rather than a type of threat (terrorism).

## How This Dissertation Fills This Scholarly Gap 

Thus, there is a gap in the civil-military relations literature regarding the causes and measurements of specific political activities involving the military within mature political democracies. To be clear, a number of other scholars have raised alarms about the impacts of rising political polarization on the conduct of US civil-military relations [for example, see @robinson_michael_danger_2018; @burbach_partisan_2019; @feaver_military_2020; @feaver_we_2016; @barno_how_2016; @golby_jim_americas_2020; @reid_retired_2020], and others have examined several political behaviors by the US military over time, to include the endorsement of political candidates by retired military officers and social media habits by members of the military [@griffiths_not_2019; @dempsey_our_2010; @urben_party_2013; @urben_wearing_2014]. However, these works, while incredibly valuable, have nonetheless not attempted to develop an encompassing theoretical justification, nor a detailed measurement scheme, to explain variation in particular behaviors over time. 

This dissertation differs from previous works and advances civil-military relations scholarship in several respects. First, this work seeks to develop a theory regarding the causes as well as a measurement strategy of political behaviors involving the military in democracies. In doing so, I conceptualize the three central principles of civil-military relations, which were described earlier in this chapter, as a necessary point of departure on the grounds that political behaviors are problematic insofar as the degree to which they violate one or more of these central principles. Second, this work harmonizes considerations that involve both types of actors, civilian and military, who can politicize the military. A perspective that considers simultaneous military and civilian efforts to politically engage with the military in various ways is consistent with previous research [@bove_beyond_2020], yet it also builds on recent efforts which have focused on one type of actor [for recent scholarship describing civilian politicization of the military, see @karlin_case_2020; @golby_uncivil-military_2021]. 

The main argument of this dissertation is that the confluence of two factors, political polarization and military prestige, greatly impacts the willingness of civilian and military actors to adhere to the central principles of civil-military relations in important ways. Furthermore, when civilian and military actors are less constrained to adhere to the principles of civil-military relations, each actor will behave in ways that violate these principles. 

If this argument is true, there are several immediate repercussions that are worth briefly touching upon. The first is that to a significant extent, the political behaviors that occur involving the military are the product not primarily of mere circumstances and personalities, but rather circumstances and personalities that operate within a context marked by the broad political factors of polarization and military prestige. A second and related repercussion is that those who practice and observe civil-military relations might better understand the occurrence of, even if they do not (and I believe that they should not!) excuse the practice of, political behaviors that involve the military. An Army Chief of Staff or a Marine Corps Commandant have a difficult and demanding job during any point in time, but could it be more difficult when polarization is high, at least in terms of managing the behavior of those who serve in their respective organizations? Similarly, a President or a Secretary of State's duties and responsibilities are always difficult, but how are they made different and perhaps more difficult when the military is exceedingly prestigious? Unpacking and exploring these dynamics is the goal of the remaining chapters.

# Methodology and Plan of the Dissertation

I first develop a theory regarding the causes of political behavior in democracies, and then implement a mixed-methods empirical approach to empirically test this theory. Chapters three and four constitute mainly quantitative, large-N studies of specific political behaviors undertaken by military and civilian leaders, while chapters five and six rely mainly on process tracing by examining two historical, qualitative case studies. The final chapter of the dissertation summarizes the results of this investigation, discusses its implications, and recommends future avenues of research. 

## Chapter 2 - Theory

This chapter develops an overarching theory that explains variation in political behaviors involving the military in mature democracies. The theory posits that two variables in particular, the levels of political polarization and military prestige, shape the degree to which military and civilian actors are constrained by the central principles of civil-military relations, and by extension, the quality of the behavior each actor engages in that involve the military. This chapter also formulates distinct hypotheses, which are then tested throughout the remainder of the dissertation.  

## Chapter 3 - Retired Military Officer Opinion Commentary

This chapter investigates retired military officer opinion commentary in depth by conducting a quantitative analysis of retired US military officer authored opinion pieces over the past four decades. Opinion commentary authored by retired military officers are one form of a political behavior undertaken mainly at the hands of military actors. The analysis reveals that retired military officers are criticizing civilian officials at a greater frequency, adopting expressly partisan positions, and weighing in on topics that fall outside of traditional military expertise more frequently than in past years. Furthermore, the analysis reveals that these results are driven largely by changes in the level of political polarization that have occurred over the past several decades. 

## Chapter 4 - The Civilian Use of the Military in Presidential Campaign Advertisements

In Chapter four, this dissertation pivots from examining a behavior conducted by military actors (retired military officer commentary) to one that is primarily the purview of civilian actors - that of the content of Presidential campaign advertisements. Using data assembled from the Wisconsin Advertising Project and the Wesleyan Media Project, this chapter performs a quantitative analysis of all television campaign advertisements aired during the five Presidential Elections held from 2000 - 2016. The analysis explores the degree to which military symbols and images appear in the advertisements, as well as the instances in which various military figures endorse or attack political candidates, or engage in other behaviors that challenge the central principles of civil-military relations. The statistical analysis reveals that the level of military prestige strongly drives changes in the way that civilian candidates harness military figures to engage in behaviors during campaign advertisements that violate the principle of non-partisanship. 

## Chapter 5 - Civilian and Military Leaders after the US Civil War (1865-1890)

Chapter five of the dissertation serves as the first of two case studies. In the first case study, the dissertation analyzes the the behavior of military and civilian actors between the end of the American Civil War and roughly 1890, or the height of the "Gilded Age" of American 19th Century politics. Over this roughly 25 year time period, the level of political polarization in the United States remains relatively high, but the level of overall military prestige decreases dramatically. This case study thus involves a change in domestic environments marked by high polarization and high military prestige to one characterized by high polarization and low military prestige.    

## Chapter 6 -Civilian and Military Leaders From Vietnam to the end of the Gulf War (1973-1992)

In the sixth chapter, the dissertation then focuses on the period after the Vietnam War until the end of the Gulf War in the early 1990s. During this time period, political polarization rises slightly, but the prestige of the military drastically increases, especially after the US victory in the Gulf War against Iraq. This case study involves an examination of a period that changes from low polarization and low military prestige to one marked by low polarization and high military prestige. 

Together, the two case studies contribute to the understanding of the the theory offered in the next chapter, and enable a meaningful comparison across the cases. The cases are roughly 100 years apart from each other, which allows the researcher to determine what, if any, historical differences result in how polarization and military prestige impact military and civilian actors. More importantly, the variable of military prestige varies significantly in each case study, but in different directions (from high to low in the first case study, and from low to high in the second), while the level of polarization remains constant but at different relative levels (high in the first case study, and low in the second). These attributes and characteristics will help isolate the role that the level of military prestige plays, under various levels of polarization, in impacting the political behaviors that military and civilian actors commit that violate the central principles of civil-military relations.    

## Chapter 7 - Implications and Conclusion

In the seventh and final chapter, the dissertation's main findings are summarized and critical implications are discussed. This chapter also contains recommendations for future research. 

\pagebreak
\singlespace
\newpage


