---
title: "Chapter 2: A Theory of How Polarization and Prestige Impact the Constraining Influence of the Principles of US Civil-Military Relations"
author: "Peter M. Erickson"
date: |
  `r format(Sys.time(), '%B %d, %Y')`
abstract: |
  
  _This chapter develops a theory that argues that two domestic political factors, the level of domestic political polarization and the degree of prestige of the military, greatly shape the degree to which the actors involved in civil-military relations adhere to the central principles of civilian control, non-partisanship, and non-interference of the military into certain sectors of state policy. The theory advanced here further argues that as these principles lose their constraining influence, civilian and military actors are more likely to commit political behaviors that violate these principles. The chapter develops and explains a typology in which four possible domestic environments are possible. Each of these environments correspond to relative levels of polarization and military prestige, and is associated with predictable patterns of political behavior conducted by civilian and military actors involving the military._  
  
geometry: margin = 1.15in              # margin widths
fontsize: 12pt                         # text size
bibliography: dissertation.bib         # bibliography file
biblio-style: authoryear               # citation style
urlcolor: blue                         # links to external URLs
citecolor: black                       # links to bibliography
linkcolor: magenta                     # internal figures, footnotes, etc.
indent: true                          # paragraph format
                                       # true for indented paragraphs
                                       
header-includes:
    - \usepackage{setspace}
    - \usepackage{multirow}
    - \usepackage{indentfirst}
   
    
output: 
  bookdown::pdf_document2:
    keep_tex: true
    latex_engine: pdflatex
    number_sections: false
    toc: false
    fig_caption: true
    citation_package: biblatex
    includes: 
      in_header: 
        - latex-preamble.tex
  bookdown::htmldocument2: default
  bookdown::word_document2:
      toc: false
---

\doublespacing 

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo=FALSE,
  include=FALSE,
  message=FALSE,
  error=FALSE,
  fig.height=3,
  fig.width=4,
  out.width="90%")
```

```{r, include = FALSE, echo = FALSE}
library("here")       # file paths
library("tidyverse")  # workhorse package
library("tidylog")
library("kableExtra")
library("knitr")
library("ggdag")
library("dagitty")
library("gridExtra")
library("broom")
library("dplyr")
library("reshape2")
library("pacman")
library("ggrepel")
library("writexl")

```

\newpage

# Introduction

This chapter develops a comprehensive theory that explains how and why civilian and military actors might be less constrained by the central principles of civil-military relations, which include the principle of civilian control, non-partisanship, and non-interference of the military into certain sectors of state policy. These principles were identified and explained in the previous chapter. 

The core argument developed in this chapter and tested throughout the remainder of this dissertation is that the levels of domestic polarization and the prestige of the military impact the _constraining influence_ of the principles of civilian control of the military, non-partisanship of the military institution, and "non-interference" of the military into certain realms of state politics. When the constraining influence of these principles is high, civilian and military actors are less likely to engage in behaviors that violate these principles, and conversely, when their constraining influence is low, civilian and military actors are more likely to commit political behaviors that violate these principles. 

This chapter proceeds in four parts. First, I define the relevant terms that are used throughout this study. I begin by defining and tracing the developments of the dependent variable, the _political behaviors_ which involve the military institution, as well as the independent variables examined in this study, _political polarization_ and _military prestige_. 

In the second and main part of this chapter, I explain how the levels of polarization and military prestige impact the overall constraining influence that the governing principles of civil-military relations have on the propensity of military and civilian actors to engage in various political behaviors that involve the military. I argue that the levels of polarization and military prestige serve as the respective "motives and opportunities" that induce the military's intervention in politics [@finer_man_1962]. I also argue, importantly, that the levels of polarization and military prestige induce the different major actors involved in civil-military relations (civilian and military leaders) to engage in different types of political behaviors, and in different ways. I ultimately argue that the levels of polarization and military prestige can combine to form a total of four possible domestic political environments, each of which is likely to produce certain patterns of political behavior committed by civilian and military actors. 

In the third section, I discuss concerns regarding endogeneity, and examine several potential alternative explanations. In the fourth and final section of this chapter, I present a total of six hypotheses to be tested, analyzed, and explored throughout the remainder of the dissertation.  
# Components of the Theory

A visual representation of the overall theory explained in this chapter and throughout the dissertation is shown below in Figure \@ref(fig:overview) . From left to right, the core argument is that the two independent variables of political polarization and military prestige influence an intervening variable, the constraining influence of the fundamental principles or standards of civil-military relations. This core argument further suggests that the overall constraining influence of the fundamental principles of civil-military relations on civilian and military actors impacts the dependent variable, patterns of political behaviors. 

```{r theory-shell, include=FALSE, echo=FALSE}
problem_dag <- dagify(
    PB ~ CI,
    CI ~ MP + PP,
    coords = tribble(
      ~ name,      ~ x,    ~ y,
      "PB", 1,.5,
      "CI", .7, .5,
      "PP", .3, .7,
      "MP", .3, .3,
    ))
    

problem_dag %>% 
  ggplot(aes(x=x, y=y, xend=xend, yend=yend))+
  geom_dag_edges()+
  geom_dag_point()+
  geom_dag_text_repel(aes(label=name), show.legend=FALSE)+
  theme_dag()
```

```{r overview, include=TRUE, echo=FALSE, fig.height=4, fig.width=5, fig.cap="The Impact of Polarization and Military Prestige on Principles of Civil-Military Relations and Political Behaviors"}

problem_dag %>% tidy_dagitty() %>% 
  dag_label(labels=c("PB"="DV\n Political\n Behaviors",
                     "CI"="Int Var\n Constraining Influence\n of Foundational\n Civil-Military Principles", 
                     "MP"="IV\n Military\ Prestige", 
                     "PP"="IV\n Political\n Polarization"
                      )) %>% 
  ggplot(aes(x=x, y=y, xend=xend, yend=yend))+
  geom_dag_edges()+
  geom_dag_point()+
  geom_dag_text()+ 
  geom_dag_label_repel(aes(label=label, fill=label), col="white", show.legend=FALSE)+ theme_dag(plot.title=element_text(hjust=.5))
  
```

## The DV: Political Behaviors

The dependent variable, the types and patterns of political behaviors involving the military, must be carefully defined and examined. This is especially true because military and civilian actors can each commit certain forms, types, and patterns of behaviors that are political in nature and that adversely impact civil-military relations. 

In this section, I describe what kinds of political behaviors are able to adversely impact civil-military relations, and how, and why. I first address the behaviors that can be committed by military actors, and then address the behaviors that can be committed by civilian actors. I describe these behaviors in general terms and then operationalize them with an eye towards the US context in particular. 

There are three critical takeaways from this section of the chapter. The first is simply an acknowledgement that there are, in fact, a wide range of behaviors that both sets of actors can commit that can threaten civil-military relations. A second takeaway is that each of these behaviors threatens civil-military relations in different ways, and to different extents. As I will argue, the political behaviors themselves that are committed by military and civilian actors that involve the military institution cannot be entirely separated from the broader political contexts in which these actions are committed. Thus, when scholars and observers of civil-military relations render normative judgments regarding the political behaviors committed by civil and military actors, and the extent to which such behaviors challenge or threaten civil-military relations, they must always, in my view, examine the full political context surrounding these behaviors. 

The third takeaway is recognition that because civil-military relations are a dynamic interaction between civilian and military actors, these types of behaviors often occur simultaneously. At the same time that military actors engage in various forms of political behaviors that may harm civil-military relations, so may civilian leaders. Too often, in my opinion, scholars focus on what the military is doing or has done that harms civil-military relations. This is certainly an important task. However, scholars need to also examine the same question from the perspective of civilian actors. There has, thankfully, been some growing emphasis on this question in the literature [@karlin_case_2020; @golby_uncivil-military_2021]. 

Combined, the insights that civil and military actors can engage in political actions at the same time, to varying degrees, and with varying impacts on civil-military relations, is intended to foreshadow what the rest of the theory laid out in this chapter argues: that certain domestic conditions facilitate both sides, civilian and military, to behave in ways that challenge the principles of civil-military relations at the same time. When domestic political conditions do not induce both sides to engage in certain forms of behavior that challenge civil-military relations, it is reasonable to expect that a state's civil-military relations will be relatively harmonious. On the other hand, if it is true that domestic conditions might actually compel or induce both sides to engage in behaviors that blatantly challenge one or more of the central principles of civil-military relations, we can and should expect a state's civil-military relationship to be more difficult. 

### Political Behaviors by Military Actors

Scholars offer a number of considerations when determining whether and the degree to which certain forms of behavior conducted by militaries and military actors are inappropriately political vice part and parcel of their regular duties, which, as other scholars have argued, are inherently political to begin with [@brooks_paradoxes_2020; @owens_us_2011; see also @clausewitz_war_1976]. These considerations include whether the behavior of a military actor was committed for the purpose of "influencing" policy outcomes, and the degree to which that behavior generates political costs for civilians [@brooks_militaries_2009, 213]. 

What should immediately jump out to the reader about these twp general considerations is that both the military actor's intent as well as the impact or effect of the behavior are important. As one might suspect, this complicates efforts by researchers to capture just how inappropriate a certain act or behavior committed by a military officer may in fact be. The intent of actors is not always visible, and at the very minimum, can be misunderstood. Furthermore, the impact of a certain behavior committed by a military officer is not always immediately evident, either. This is further complicated by the fact that some behaviors can have negative, but also very positive, repercussions. 

These multifaceted considerations are evident in the following, hypothetical example. An observer may conclude that a military officer telling the President the truth is not only good, but perfectly aligned with the duties and expectations of a military officer. This may in fact be true, but if we were to further learn that the President had already decided on a course of action that was contrary to the opinion of the officer, and that the officer informed the President while seated at a conference table along with other cabinet secretaries and members of the press, we might instead conclude that in being truthful, the military officer's behavior at a minimum made the President's job more difficult, if not sparked a genuine crisis in civil-military relations! If we later learned, however, that in speaking up, the military officer altered the President's decision by abandoning a course of action that would have resulted in the unnecessary deaths of more than 50,000 Troops and countless civilians, we might approve of the officer's conduct, regardless of when and how such advice had been offered! 

This hypothetical example illustrates that the political context and circumstances surrounding a particular act of behavior committed by a military actor are at least as important, and sometimes perhaps more important, than the act itself, especially with respect to determining whether a certain instance of behavior challenged the conduct of civil-military relations. Furthermore, as Brooks (2009) rightly notes, certain aspects of engaging in political behaviors and activities can also yield positive repercussions [@brooks_militaries_2009]. At the end of the hypothetical example above, the military officer's behavior in speaking out may have in fact challenged civil-military relations, but it also had the positive effect of saving lives.

It would be impossible to render an exhaustive list of the political behaviors or actions that military officers have at their disposal to influence policy outcomes and/or generate political costs for civilians, and more difficult to attempt to parse out the possible types of contexts in which these behaviors are appropriate vice harmful for the conduct of civil-military relations. In general, along the range of a spectrum of political behaviors committed by the military that are immediately dynamic and visible to observers, one would include coups or other forms of the military taking over government. On the other end of this spectrum would be behaviors that are less visible and thus more subtle, such as the military selectively giving advice, or conversely, selectively complying with civilian orders.^[Selectively giving advice and selectively complying with civilian directives constitutes an important literature in contemporary US civil-military relations. See @brooks_paradoxes_2020 for a recent example, particularly with respect to how military leaders shaped advice regarding the surge to Afghanistan in the early stages of the Obama Administration.] Somewhere between the ends of this spectrum fall a range of many other behaviors, including retired military officers running for elected office, and the use of the military to appear in the media. To sometimes greater and other times lesser extent, these (and likely many other) types of behaviors may be committed by military officers to shape policy outcomes. In doing so, these behaviors and the officers who commit them can and often do generate political costs for civilian leaders. 

In the US context, some behaviors are more relevant and more frequent than others. Risa Brooks (2009) has developed one helpful framework for better understanding these behaviors, particularly in the US context. These types of behaviors are displayed below in Table \@ref(tab:pol-behav).  

\singlespace

```{r pol-behav, echo=FALSE, include=TRUE, error=FALSE, message=FALSE}

text_tbl1 <- tribble(~"Tactic", ~"Nature of Activity", ~"Audience", ~"Examples",
  "Public Appeal", "Public, value-laden commentary", "Mass public",  "Colin Powell's 1992 statements on intervention in Bosnia", "Grandstanding", "Threat or actual resignation in protest of policy", "Mass public, Congress, executive branch", "Hypothetical; if Eric Shinseki had resigned in 2003",
  "Politicking",  "Retired officer endorsements; organizing vote drives", "Mass public", "Crowe's endorsement of Clinton in 1992; Frank's endorsement of Bush in 2004", 
  "Alliance Building",  "Military leaders form ties with civilian interest groups", "Congress", "Air Force and Lockheed Martin over procurement of F-22 raptor",
  "Shoulder Tapping", "Military leaders setting an agenda by bringing issues to attention of politicians and engaging in lobbying-like activities on behalf of those issues", "Congress", "Military Mobilization of key members of Congress over gays in the military in early 1990s")
 

text_tbl1 %>% kable(booktabs=T, caption="Political Activities of the Military") %>%
  kable_styling(full_width = F, font_size=8, latex_options="striped" ) %>% 
  column_spec(1, width="10em") %>% 
  column_spec(2, width="15em") %>% 
  column_spec(3, width="10em") %>% 
  column_spec(4, width="15em") %>% 
  add_footnote(label = "The above table is reproduced from Brooks, Risa (2009), 'Political Activities of the Military' in\n American Civil-Military Relations: The Soldier and the State in a New Era, Eds. Suzanne Nielsen and Don Snider.")

```
\doublespace

Two implications are worth emphasizing from these types of behaviors. Underscoring the importance of considering the context and specific circumstances surrounding these behaviors, the first implication is that not _all_ of these behaviors are of the qualitative nature such that they seek to influence policy outcomes and/or generate significant political costs in every instance. Military officers making "public appeals" illustrates this point. For example, when General Colin Powell wrote a series of op-eds in the early 1990s while serving as the Chairman of the Joint Chiefs of Staff in which he articulated his views regarding the use of force, his behavior undoubtedly generated political costs for the candidate, and later, President, Bill Clinton [@powell_opinion_1992; @powell_us_1992]. Moreover, engagement in this behavior by a highly respected and prestigious figure likely influenced public opinion and thus, constrained the options available to Clinton as a number of crises erupted across the world in the Balkans, Somalia, Haiti, and Rwanda. Similar levels of generating political costs and impacting policy outcomes are not likely generated every time a retired General or Admiral pens an op-ed, or gives an interview to the press. 

Conversely, every instance of a military officer publicly endorsing a political candidate is an inherently political act that seeks to influence a policy outcome (the result of an election) and which generates political costs (for the candidate not endorsed by the military actor). Regardless of whether the military actor's endorsement of a political candidate is effective and actually persuades the public to change its vote choice [see @golby_military_2012], it is impossible for a military officer to make a nonpartisan endorsement of a political candidate. Other behaviors, such as resignation, have sparked considerable debate among scholars of civil-military relations. Some have argued that the public nature of resignation constitutes an inherently political act that should not be practically considered [@golby_beyond_2015], while others have left room for the possibility that in extreme circumstances, resignation is appropriate, regardless of the potential public fallout that may be generated [@dubik_principled_2014; @snider_strategic_2014].

The second implication of the behaviors listed in Table \@ref(tab:pol-behav) is that military actors have a wide variety of activities at their disposal from which to choose to generate political costs for civilian leaders. The well known collisions that have occurred between the President and various wartime Commanders throughout American history - Lincoln and McClellan, Truman and MacArthur, Obama and McChrystal - perhaps threaten civil-military relations in the most recognizable and public way, but there are other subtle and nuanced means available to the military which also generate political costs that civilian leaders must contend with.  

For instance, when military actors offer commentary to the public, the information they provide can help citizens better understand and process information that is often complex and disconnected from everyday life, particularly if the issue is somewhat technical or complex [@brooks_militaries_2009, 225-231; see also @schultz_democratic_1999]. Furthermore, militaries can alert Congress to potentially destructive decisions an executive may be considering, thus helping the nation avoid undertaking a course of action that will lead to disastrous consequences [@brooks_militaries_2009, 225-231]. While the final decisions ultimately belong to civilian leaders to make, there are nonetheless real and potentially positive benefits of military actors engaging in the political process in overt ways. 

Brooks (2009) ultimately notes that military actors who deliberately commit these types of political activities do so while facing four unique risks [@brooks_militaries_2009, 234]. A first risk involves society's esteem of the military [@brooks_militaries_2009, 234]. Part of the reason the military has maintained its popularity with American society is precisely because it has been seen and viewed as a professional institution that does not routinely engage in inappropriate political activities. A second risk of engaging in these acts is that doing so provides civilian leaders with an estimate of a particular military leader's political views, which could jeopardize how civilian leaders consult with that particular but also other military actors [@brooks_militaries_2009, 234]. A third risk facing military actors when engaging in political activities involves the tenets of "military professionalism" laid out by Huntington and previously discussed in the previous chapter: by engaging in political activities, military officers become less focused on their professional duties of securing the Republic [@brooks_militaries_2009, 235; @huntington_soldier_1957]. A fourth and final risk, one which Brooks argues is the most important to consider, is that doing so may "undermine" civilian leaders and ultimately, the governing principle of civilian control [@brooks_militaries_2009, 235]. Ultimately, Brooks (2009) concludes that the potential benefits of military actors engaging in political activities must be weighed against the potential consequences, and recommends that for the sake of long-term, stable civil-military relations, military actors rarely engage in these and other political activities [@brooks_militaries_2009].

In summary, military actors have an arsenal of behaviors they can use to shape the choices of civilian leaders, raise the political costs borne by civilian leaders, or both. Some of these behaviors, such as the military endorsing political candidates, are almost always problematic with respect to upholding one or more of the principles of civil-military relations, while other behaviors, such as making "public appeals" [@brooks_militaries_2009], require observers of civil-military relations to fully evaluate the context in which they are committed before assessing the appropriateness and consequences of such behavior. In the United States in particular, where coups and other stark forms of military political behavior are (thankfully) not a regular occurrence, scholars and those who practice civil-military relations should stay highly attuned to how certain political behaviors committed by the military impact the principles of civilian control, non-partisanship, and non-interference, and thus, the conduct of civil-military relations. 

## Political Behaviors Committed by Civilians

Civilian leaders can also engage in behaviors that challenge the conduct of civil-military relations in general, and, in the United States in particular, the principles of civilian control, non-partisanship, and non-interference of the military. In the broadest possible sense, these behaviors are those undertaken by civilian leaders that inappropriately "politicize" the military. However, such a broad definition is, in my view, rather insufficient. A far better definition to describe those behaviors committed by civilian leaders that potentially harm civil-military relations has been offered recently by Jim Golby (2021) as the concept of "civilian activation," which involves those behaviors "by civilian political elites to court or co-opt the military for personal, partisan, or electoral gain" [@golby_uncivil-military_2021, 153]. 

Characteristics such as regime type influence the type of behaviors civilians will commit. For instance, in autocratic or emerging-democratic contexts, civilians might undertake relatively drastic actions to prevent military forces from undertaking a coup, such as establishing "parallel" militaries who hold a "special loyalty" to the regime in power [@quinlivan_coup-proofing_1999] and other forms of "counterbalancing" [@de_bruin_will_2019]. Similarly, Beliakova (2021) notes that in democracies, one of the ways in which the principle of civilian control can erode is through the behavior of civilian "deference," which describes the act of "civilians delegating policy making prerogatives to the members of the military profession," a behavior that could manifest itself in a number of ways [@beliakova_erosion_2021, 1400].

In established democracies such as the United States, the types of behaviors that civilian leaders commit involving the military institution varies considerably. On the more visible end of the spectrum of such behaviors would be the appointment of one or more active or recently retired officers to high visibility political postings, or the use of recently retired military officers to serve as keynote speakers at major political conventions. Such behaviors are, unfortunately, relevant in the modern US context. For instance, in the aftermath of the Presidential Campaign of 2016, Retired Army General Martin Dempsey, who previously served as the Chairman of the Joint Chiefs of Staff, published an opinion-editorial piece in _The Washington Post_ in which he criticized two former Generals (retired Army Lieutenant General Mike Flynn and retired Marine Corps General John Allen) for speaking on behalf of then-candidate Donald Trump and the Democratic challenger, Hillary Clinton, at the respective conventions of each political party. Dempsey stated in the op-ed, "As generals, they have an obligation to uphold our apolitical traditions. They have just made the task of their successors – who continue to serve in uniform and are accountable for our security - more complicated. It was a mistake for them to participate as they did. _It was a mistake for our presidential candidates to ask them to do so_" [@dempsey_military_2016, emphasis mine; see also @exum_dangerous_2017]. 

On the less-visible end of this spectrum of behaviors are those that can take on a far more subtle form. For example, in election years, high-ranking retired military officers often sign a form letter that endorses a particular candidate for office, and this letter is sometimes released by itself or published in a newspaper [for an analysis of this behavior, see @griffiths_not_2019]. To be sure, letters of this kind constitute a partisan act, but in a way that is ostensibly less-visible than a military officer speaking at a party convention. 

For their part, civilian actors also use the military in subtle ways that may - but do not necessarily have to - challenge one or more of the principles of civil-military relations. For instance, during political campaigns, civilian candidates and their campaign staffs may plan visits that occur at or near a military base in order to signify to the public that candidate's security credentials or ties to the military. Such a behavior may not be intended as overtly partisan, but it nonetheless pushes the military closer to a partisan line that at some point becomes normatively unwise, if not appropriate, depending on the particular context in which these types of behaviors occur, and how a candidate or civilian leader specifically behaves when such behavior occurs. Presidents regularly visit military bases, just as President Obama did in 2011 to mark the formal end of the war in Iraq in 2011 [@obama_remarks_2011]. However, observers were concerned when President Trump visited Iraq in 2018 when some US Troops reportedly adorned hats and other "Make America Great Again" (MAGA) gear, which President Trump in some cases happily signed [@fritze_trump_2018]. The President appearing with Troops was not the problem; what gave observers concern, and rightfully so, was the appearance of the President, combined with what seemed to be his support for, Troops who were outwardly endorsing the partisan nature of his office. Finally, it is important to point out that in general, rendering definitive normative evaluations of particular behaviors involving the military are _more_ difficult with respect to civilian leaders than with military actors. This is because, as Golby notes, "elected political leaders often provide alternative explanations to legitimize or excuse their behavior," and observers cannot readily or easily observe the "intent" of those civilian leaders who engage in such behaviors [@golby_uncivil-military_2021, 152]. 

Furthermore, as was the case with respect to evaluating political behaviors committed by military actors, the broad domestic political context matters tremendously when assessing whether and the extent to which civilian leader behavior involving the military threatens or challenges the conduct of civil-military relations. For instance, observers of US civil-military relations did not, to my knowledge, sound a major alarm when an incumbent President, Ronald Reagan, named Lieutenant General Colin Powell, then on active duty, as his National Security Advisor in 1987. However, 30 years later, scholars and observers of civil-military relations alike were concerned when President Trump named several retired and active generals to key postings across his cabinet and within his administration.^[These included Jim Mattis as Secretary of Defense; Mike Flynn, and later, H.R. McMaster, as National Security Advisor; John Kelly, first as head of the Department of Homeland Security and, later, White House Chief of Staff; and Mark Inch as head of the Bureau of Prisons]. In briefly comparing these two cases, in my view, it is both the number of relative military figures appointed to an administration _and_ the fact that unlike President Reagan, who had been in office for several years and, by that time, previously established his national security bona fides, President Trump was not only new, but a contentious figure. 

In the US context, civilian leaders have faced consequences for "politicizing" the military previously. Several isolated instances reveal that this has been the case. For example, some observers alleged that President George W. Bush's use of General David Petraeus in garnering Congressional support for the "surge" strategy in Iraq politicized the military in unhealthy ways [@coll_generals_2008]. Under President Donald Trump's tenure in office, charges of politicizing the military were more common, including after the former President's 2018 decision to deploy several thousand troops to the US border with Mexico in response to an impending caravan of migrants from South America. The headline of an op-ed published by several retired Army Colonels in response to this incident aptly summarizes the feelings held by many Americans at the time: "Trump's Border Stunt is a Profound Betrayal of Our Military" [@adams_trumps_2018]. Examining these singular instances of inappropriate political behavior is insightful, as are those that look to the behaviors of an entire administration at large [@golby_uncivil-military_2021]. 

At the same time, such analyses fall short, in my opinion, of thoroughly investigating how, why, and the extent to which particular shifts in a society's domestic political factors impact the behavior of civilian and military actors with respect to the military, and by extension, whether and how these changes impact civil-military relations. In order to this, systematic empirical work that examines behavior over time is required. ^[For this reason, I undertake systematic explorations of one type of political behavior committed by each actor over time (opinion commentary authored by retired military officers in chapter 3, and the use of military figures and imagery in civilian political candidate advertisements in chapter 4). These particular behaviors are especially worthy of examination, as I argue later, for two main reasons. The first is because these behaviors are of the quality such that they do not necessarily have to violate one or more of the principles of civil-military relations, although they can, and sometimes do. In other words, each behavior can be operationalized in such a way as to capture instances of challenging one or more of the principles of civilian control, non-partisanship, and non-interference. The second reason these particular behaviors are important is because they occur regularly, over time, as important features of American democracy (e.g. a free press and regular elections). Even in these democratic institutions, there is military involvement! The important questions, which will become clear in these later chapters, is what kind of involvement, and whether such involvement adversely impacts civil-military relations.]

In summary, civilian leaders can engage in certain behaviors which challenge civil-military relations, much in the way that military actors can and do. These behaviors undertaken by civilians take on a variety of forms, both in general and in the United States specifically. Some behaviors are visible, such as a civilian candidate selecting a military officer to speak at a major political convention, or appointing one or more military officers to high-level political posts. Other forms of behavior are far more subtle, such as the decision by civilian candidate campaign teams to feature military imagery and symbols during campaign advertisements, or to schedule a campaign stop at or near a major military installation. What is common to these (and other) forms of political behaviors regarding the military is that they are conducted by civilian leaders or candidates "for personal, partisan, or electoral gain" [@golby_uncivil-military_2021, 153]. To the extent that they are intended to have such an effect, and that they do, matters on the particular context in which these behaviors are committed.  

Throughout the rest of this chapter, I argue that changes in two domestic factors stand out in terms of altering the extent to which military and civilian actors commit political behaviors in ways that challenge the principles of civilian control, non-partisanship, and non-interference. These two domestic factors are the levels of political polarization and military's prestige in a society. In the next section of the paper, I define each of these variables and trace their respective levels in contemporary America.

# IV1 - Political Polarization

The first independent variable of study in this dissertation is political polarization. Polarization is a concept that is difficult to define precisely. In American politics, polarization generally refers to the movement of political thoughts and attitudes within a political community away from the political center and towards more ideological extremes [@hare_polarization_2014; @abramowitz_great_2018]. Within international relations, polarization is often defined the same way, but scholars of international relations are interested in different applications of polarization, such as how polarization contributes to a state's ability to recognize common threats, learn and derive lessons from foreign policy failure, and agree on what role a nation should play on the world stage [@schultz_perils_2017; @myrick_external_2020]. From the perspective of political theory, political polarization describes the degree to which a society shares fundamental values, is capable of incorporating divergent worldviews, and the overall degree to which these values unite versus divide a nation [@marsden_twilight_2014; @deneen_why_2018]. In this dissertation, I rely on both the relatively narrower conception of polarization as the divergence of political attitudes away from an ideological center and towards the extreme, as well as a conception of the term that recognizes the difficulties which arise when a society increasingly disagrees about the fundamental values and worldviews which should govern it. 

## Perspectives on Polarization

Perspectives regarding the _causes_ of political polarization in the United States vary. Scholars have pointed to a variety of causes of polarization, including differences of opinion over issues pertaining to race and religion [@abramowitz_great_2018], the size and role of government [@sides_two_2015], normative conceptions of personal and national identity [@sides_identity_2018], and other specific policy issues [@mann_its_2012]. Other scholars, mainly political theorists and cultural historians, have argued that the current polarized environment in the United States is the result of a slow yet steady political shift that began in the 1950s, a shift that reflects both the incompatibility of American pluralism to reconcile and "embrace a wide range of both religious and nonreligious viewpoints" [@marsden_twilight_2014, 171] and the inability of liberalism as a political philosophy to deliver on its intended promises of securing individual rights in a manner that is equitable and fair [@deneen_why_2018]. What is common to most explanations of polarization in the United States is the sense that American society is increasingly divided because of the loss of support for the collective values and purposes that should guide and unite society, and that the various perspectives about which values and purposes should perform this task today are inherently at odds with each other.   

Scholars note the presence of three characteristics or features that further describe the current polarized political landscape in the US. The first feature of polarization in the contemporary United States is that it goes far beyond political differences between political elites [@mann_its_2012]. In other words, today's polarized landscape extends beyond the halls of Congress in Washington D.C. While one can argue that everyday citizens may not feel as strongly as elected politicians about every single political issue at stake today, the fact is that the level of disdain held by regular Americans for both political parties as well as those who identify as being on the opposite side of the political spectrum has risen sharply over the past several decades. This phenomenon is known in the American politics literature "affective polarization" [@iyengar_affect_2012; @iyengar_origins_2019] or "negative partisanship" [@abramowitz_great_2018]. 

A second feature of contemporary polarized politics in the United States is the increasing scope and number of issues that have become included as part of the partisan battleground. Politicians and citizens do not disagree and wage political war over a handful of important issues, but rather, at least seemingly, over everything! Scholars of American politics describe this dynamic of increasing political warfare as "conflict extension" [@layman_party_2002]. As one might expect, the dynamic of "conflict extension," in which every political issue seems to constitute a hefty partisan political battle, has dampened the prospect for meaningful cooperation across partisan aisles and other displays of bipartisanship [@layman_activists_2010]. 

A third and final feature of contemporary political polarization in the United States is that issues of personal and collective identity are increasingly associated with partisanship - to the point that partisan identity itself is an important component of how people view themselves and others [@sides_identity_2018; @robinson_michael_danger_2018]. This feature of polarized politics is perhaps most clearly illustrated by the rise and influence of American "identity politics" in which the normative roles of race, gender, religion, and immigration invoke very strong commitments from individuals on both sides of the partisan aisle [@white_values_2002; @mann_its_2012; @sides_identity_2018]. The prominence of these identity-related issues in politics affirms that these issues are not merely peripheral interests, but rather issues that individual citizens care deeply about and consider to be of fundamental concern. 

## Polarization in the US Over Time

```{r polar-measurement, include=FALSE, echo=FALSE}
nom_dat <- read_csv("https://voteview.com/static/data/out/members/HSall_members.csv")

south <- c(40:49,51,53)
polar_dat <- nom_dat %>% 
    filter(congress>45 & 
           chamber != "President") %>%
    mutate( 
      year = 2*(congress-1) + 1789,
    ) %>%
    group_by(chamber,congress,year) %>% 
    summarize(
      party.mean.diff.d1 = mean(nominate_dim1[party_code==200],na.rm=T) - 
                           mean(nominate_dim1[party_code==100],na.rm=T),
      prop.moderate.d1 = mean(abs(nominate_dim1)<0.25,na.rm=T),
      prop.moderate.dem.d1 = mean(abs(nominate_dim1[party_code==100])<0.25,na.rm=T),
      prop.moderate.rep.d1 = mean(abs(nominate_dim1[party_code==200])<0.25,na.rm=T),
      overlap = (sum(nominate_dim1[party_code==200] <
                       max(nominate_dim1[party_code==100],na.rm=T),na.rm=T)  +
                 sum(nominate_dim1[party_code==100] >
                       min(nominate_dim1[party_code==200],na.rm=T),na.rm=T))/
                 (sum(!is.na(nominate_dim1[party_code==100]))+
                  sum(!is.na(nominate_dim1[party_code==200]))),
      chamber.mean.d1 = mean(nominate_dim1,na.rm=T),
      chamber.mean.d2 = mean(nominate_dim2,na.rm=T),
      dem.mean.d1 = mean(nominate_dim1[party_code==100],na.rm=T),
      dem.mean.d2 = mean(nominate_dim2[party_code==100],na.rm=T),
      rep.mean.d1 = mean(nominate_dim1[party_code==200],na.rm=T),
      rep.mean.d2 = mean(nominate_dim2[party_code==200],na.rm=T),
      north.rep.mean.d1 = mean(nominate_dim1[party_code==200 & 
                                             !(state_icpsr %in% south)],na.rm=T),    
      north.rep.mean.d2 = mean(nominate_dim2[party_code==200 & 
                                             !(state_icpsr %in% south)],na.rm=T),    
      south.rep.mean.d1 = mean(nominate_dim1[party_code==200 & 
                                              (state_icpsr %in% south)],na.rm=T),    
      south.rep.mean.d2 = mean(nominate_dim2[party_code==200 & 
                                             (state_icpsr %in% south)],na.rm=T),    
      north.dem.mean.d1 = mean(nominate_dim1[party_code==100 & 
                                              !(state_icpsr %in% south)],na.rm=T),    
      north.dem.mean.d2 = mean(nominate_dim2[party_code==100 & 
                                              !(state_icpsr %in% south)],na.rm=T),    
      south.dem.mean.d1 = mean(nominate_dim1[party_code==100 & 
                                              (state_icpsr %in% south)],na.rm=T),    
      south.dem.mean.d2 = mean(nominate_dim2[party_code==100 & 
                                              (state_icpsr %in% south)],na.rm=T),    
    ) 

head(polar_dat)
write_csv(polar_dat,path="voteview_polarization_data.csv")

polar_dat %>% write_xlsx(here("data", "cong_polarization.xlsx"))

```


```{r include=FALSE, echo=FALSE}
# Here are up to date versions of the famous party means over time plots for each chamber:
polar_dat_long <- polar_dat %>% gather(score,value,-chamber,-year,-congress)
labels <- c("dem.mean.d1"="DEM",
            "rep.mean.d1"="REP",
            "north.dem.mean.d1"="N. DEM",
            "south.dem.mean.d1"="S. DEM")

polarized_plot <- function(chamb) {
  pdatl <- polar_dat_long %>% 
                filter(chamber==chamb,
                       score %in% c("dem.mean.d1","rep.mean.d1",
                                    "north.dem.mean.d1","south.dem.mean.d1")) %>%
                mutate(party=labels[score]) %>%
                ungroup()
  
  gg <- ggplot(data=pdatl,
               aes(x=year,y=value,group=party,col=party)) +
               scale_x_continuous(expand = c(0.15, 0), 
                                 breaks=seq(1880, max(pdatl$year), by=8)) +
               geom_line() + geom_point(size=0.7) + 
               xlab("Year") + ylab("Liberal-Conservative") + 
               geom_text_repel(data=pdatl %>%
                    filter(year == min(year)),
                  aes(label = party, color = party),
                  size = 3,
                  nudge_x = -8,
                  point.padding = 0.1,
                  segment.color = NA,
                  show.legend = FALSE) +
              theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
              scale_color_manual(values=c("REP"="#c70828","DEM"="#1372ad",
                                         "N. DEM"="#6194F4",
                                         "S. DEM"="#81c4e4"),guide="none") +
              theme_bw()
  gg
}
```

```{r include= FALSE, echo=FALSE}
# House Polarization Plot
gg1 <- polarized_plot("House")
ggsave("voteview_house_party_means.png", plot=gg1, device="png", 
       width=7, height=4, dpi=300)
ggsave("voteview_house_party_means.pdf", plot=gg1, device="pdf", 
       width=7, height=4)
gg1
```


```{r include = FALSE, echo=FALSE}
# Senate Plot
gg2 <- polarized_plot("Senate")
ggsave("voteview_senate_party_means.png", plot=gg2, device="png", 
       width=7, height=4, dpi=300)
ggsave("voteview_senate_party_means.pdf", plot=gg2, device="pdf", 
       width=7, height=4)
gg2

```

```{r polar-chamber, include=TRUE, echo=FALSE, fig.align='center', fig.cap="Liberal-Conservative Partisan\nPolarization by Chamber, 1880-Present"}
# liberal-conserative polarization by chamber
gg3 <- ggplot(polar_dat_long %>% 
              filter(score == "party.mean.diff.d1"),
              aes(x=year,y=value,group=chamber,col=chamber)) +
              scale_x_continuous(expand = c(0.10, 0), breaks=seq(1880,max(polar_dat_long$year),by=8)) +
              geom_line() + geom_point(size=0.5) + 
              xlab("Year") + ylab("Distance between party means") + 
              geom_text_repel(data=polar_dat_long %>%
                    ungroup() %>%
                    filter(year == min(year) & score=="party.mean.diff.d1"),
                  aes(label = chamber, color = chamber),
                  size = 3,
                  nudge_x = -10,
                  point.padding = 0.1,
                  segment.color = NA,
                  show.legend = FALSE) +
              theme(axis.text.x = element_text(angle = 60, hjust = 1), plot.title=element_text(hjust=.5), plot.subtitle=element_text(hjust=.5)) + scale_color_manual(values=c("House"="#c70828","Senate"="#1372ad"), guide="none") +
               labs(subtitle="Data Available from Lewis et. al. (2020) at Voteview.com")


gg3

```

The most widespread and historical (though certainly, not perfect) measurement of polarization has been that of dynamically-weighted, three step estimation scores (DW-NOMINATE), which are derived from examining roll call voting patterns among members of Congress [@jeffrey_b_lewis_voteview_2020]. Figure \@ref(fig:polar-chamber) shows polarization levels present in the US House and Senate since the latter portions of the 19th Century [@hare_polarization_2014; @jeffrey_b_lewis_voteview_2020]. As Figure \@ref(fig:polar-chamber) demonstrates, polarization levels in the United States were high around 1880, and remained high before gradually decreasing early in the 20th century. The figure further shows that polarization reached it's nadir in the years before World War Two, and remained at relatively low levels until approximately the mid-1970s. Since that time, however, polarization in both chambers of the US Congress has increased dramatically, to the point that contemporary levels now exceed those previously recorded.  

Polarization has been measured in a second way in the American Politics literature. First derived by Iyengar et al. (2012), measurements of "affective polarization" or "negative partisanship" have been derived by using data from the American National Elections Studies (ANES) [@iyengar_affect_2012]. Beginning in the late 1970s, the ANES has asked a series of "feeling thermometer" questions, in which respondents are asked to provide, on a scale ranging from 0-100, how they feel about a range of topics, including their feelings about political parties. Figure \@ref(fig:affect-polar) graphs the results of how partisans feel about members of their own party, members who affiliate with the opposite party, the difference between these two answers, and a total level of affective polarization, which is the sum of the difference for each of the two main parties in the United States.

As Figure \@ref(fig:affect-polar) shows, measurements for how favorable each party is have remained more or less constant over the past several decades. However, partisans have increasingly rated those on the opposite side of the political spectrum as increasingly less favorable over the same time period. In other words, data shows that partisans do not necessarily like the party with whom they affiliate more, but rather that they disdain those on the other side of the aisle at greater levels. 

```{r affect-polar, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.height=5.2, fig.width=7.2, fig.cap="Affective Polarization in the US, 1978-2020"}
df <- read_csv(here("data", "Aff_Polarization.csv"))

slim_df <- df %>% select(Year:AffPol) %>% na.omit %>% view()
melted_df <- melt(slim_df, id="Year") %>% view()

melted_df %>% ggplot(aes(x=Year, y=value, color=variable)) + geom_point() +
                       geom_smooth(se=FALSE) + labs(color="ANES\nFeeling Thermometer\nValue") +   scale_color_manual(breaks=c("DemIn", "DemOut", "DemDiff","RepIn","RepOut","RepDiff","AffPol"), labels=c("Democrat-In","Democrat-Out","Democrat-Diff","Republican-In","Republican-Out","Republican-Diff","Affective Polarization"), values=c("red", "blue", "green","dodgerblue","darkorange","cyan1","black")) + theme(legend.position="bottom", plot.title=element_text(hjust=.5))

```

## Why Does Polarization Matter?

One does not have to look far today in the American political culture to grasp the fact that current levels of polarization are high. Indeed, it is little wonder that both candidates running for President of the United States in 2020 claimed that the very "soul of the nation" was at stake [@dias_biden_2020]. The questions with which the United States is wrestling about itself reflect not only political differences among America's elected leaders, but significant gaps between the worldviews of the nation's citizens and the fundamental core values that should be adopted, implemented, and practiced. 

A cursory examination of just a handful of these polarizing questions confirms that they are not easy questions, nor likely to be solved without eliciting a strong response from individuals, depending on which political party, or political ideology, an individual supports. To be clear, my aim in listing these questions is to demonstrate that there are a host of consequential and important political questions in the US today that, depending on one's ideology, political party affiliation, or worldview, are likely to be answered in a particular way. And, what is critical for understanding the degree of polarization that is prevalent in the contemporary US, those who answer one or more of these questions in a certain way will likely view those who might answer a different way, and the worldviews from which these answers arise, as disdainful.     


* What does it mean to be an American? Do we have a slate of common "American Values"? What are they?
* Should America be proud of its history? 
* Is the United States an "inherently racist" or "systematically racist" country?
* To what degree should the United States be proud of its Constitution, and those American "founding fathers" who wrote it?
* To what degree should the United States have and enforce borders? Do all people have a "human right" to come to the United States?
* To what degree are the constructs of sex and gender important anymore? Are men and women different in any substantive way (physically, physiologically, biologically)? Or are all people the same, in every way? 

\doublespace
These questions, and many more, illustrate how different worldviews reflect drastically different ideas of how the United States should look moving forward. These questions are qualitatively different from policy questions such as what percentage of national GDP should be spent on the military, whether charter schools should be supported or opposed, and whether the debt ceiling should be raised in times of fiscal crisis and uncertainty. While there are undoubtedly significant policy ramifications that stem from these questions, these listed above (and others) bear significantly on an individual's deeply held beliefs, values, and overall worldview [@white_values_2002; @mann_its_2012; @sides_identity_2018]. How the nation answers them will likely set a tone that will have lasting impacts on the nation's values, now and into the future. 

In this dissertation, I argue that members of the US military are not immune from holding strong views and opinions on the questions listed above and similar issues that are considered to be central. I argue that the views held on these questions by various military actors stem from their personal worldviews and values that they hold dear and cherish. Therefore, I argue that when these types of central and consequential issues and concerns are at stake, members of the military will sometimes behave in a way that advances the values and worldview that they believe in, even if doing so runs counter to the foundational principles of civil-military relations: maintaining civilian control, the military's non-partisan ethic, and the military not interfering in certain areas of state policy. 

# IV2 - The Military's Prestige 

The second independent variable of focus in this study is the military's overall level of prestige. Military prestige does not refer to one single indicator or measurement, but rather a conglomerate of factors, including the military's relative level of influence within the government and its standing among and within society. In this vein, the concept of military prestige is similar, though not identical, to previous studies within civil-military relations that have examined similar variables, such as the "balance of power" which between political and military leaders [@brooks_shaping_2008]. In short, military prestige refers to the degree of power, respect, and influence the military institution (and specific military actors) has and can wield - both within the government and across society. 

Measuring military prestige is a bit more difficult than measuring polarization. On one hand, scholars have routinely used proxies such as public opinion of the military, including measures of public trust and confidence in the military [@noauthor_confidence_2020]. On the other hand, measuring a concept such as influence, and in particular, how much influence the military truly has, is more difficult. Indeed, as Richard Betts has stated, "influence...is one of the most important and most elusive concepts in the study of politics. In decisions on use of force it is impossible to detect objectively whose influence is greatest; it can only be inferred because only the President - and maybe not even he - really knows what convinces him or changes his mind" [@betts_soldiers_1977, 5]. 

In the paragraphs to follow, I summarize two different but related contributors of the contemporary US military's high level of prestige. The first is the US military's growing institutional scope and influence within the government. In spite of the fact that influence is difficult to measure, it is nonetheless clear that since World War Two, the US military's relative scope of responsibility within government, and within the realm of foreign policy making in particular, has grown. The result is that the contemporary military is a behemoth of an organization _relative_ to other government agencies. Moreover, the US military possesses an outsized relative budget to prove it. The second contributor to the contemporary military's level of prestige deals with how it is viewed by American society more broadly. Today, the modern military is somewhat exclusive or distant from many segments of American society, as fewer and fewer Americans can relate to the military writ large. This has rendered the military institution increasingly popular, but also somewhat mysterious. Each of these factors, the military's relative scope and influence within government, and its societal exclusivity, have contributed to the military's high level of prestige in the modern United States.^[It is important to point out that as I write this dissertation, it is possible that several events which have unfolded over the past few months may in fact reduce the military's level of prestige. For instance, the full impacts of the chaotic withdrawal of the US from Afghanistan in August 2021 on the military's level of prestige may not be tangible or fully show up in American public opinion for some time.]    

## The Military's Growing Relative Institutional Scope and Influence

Over the past seventy-five years, the US military's relative scope and influence within the government has grown considerably. Today, the military is a massive organization that influences a wide range of national security policy arenas. The military's potential to influence policy - for better, or for worse - is, in many ways, the result of notable institutional reforms to the Department of Defense and its precursor, as well as to the respective individual services, over the past several decades.  

One of the most prominent reforms to the military came in the wake of World War Two. The National Security Act of 1947, which was signed by President Harry Truman, sought to unify the defense and intelligence establishments. Despite achieving victory in World War Two, the military services, and the leaders in charge of them, had nonetheless squabbled a fair bit throughout the war [@kaiser_no_2014]. The National Security Act created new institutions, including the National Security Council (NSC), the Central Intelligence Agency (CIA), and the US Air Force [@noauthor_national_1947], and more formally, merged the Departments of War and Navy into the "National Military Establishment," which was essentially the precursor to what became the Department of Defense in 1949 [@locher_has_2001]. 

In many ways, the reforms wrought by the National Security Act proved to be quite successful. So influential was the National Security Act that fifty years after its signing, President Bill Clinton stated that as a result of the institutions created by the National Security Act and the people who had served in them, a generation had "helped to secure the peace and prosperity that America enjoys...the success of their efforts and of the historic legislation enacted half a century ago is reflected in an outstanding record of achievement: nuclear war averted, the Cold War won, and the nations of the world turning to democracy and free markets" [@clinton_weekly_1997].

Like most institutional reforms, however, the National Security Act of 1947 could not nor foresee all of the difficulties that the future could bring. A second set of major institutional defense reforms was undertaken in the mid-1980s. Known as the Goldwater-Nichols Act after Senator Barry Goldwater and Congressman William "Bill" Nichols, several incidents had exposed the fact that the US military lacked the requisite capabilities to operate in _joint_ operational environments, or those which required military personnel and equipment from across the individual military services [@kitfield_prodigal_1997].^[This included the the failed Iranian hostage rescue mission in 1980. For an excellent account of how this tragic event spurred the evolution of joint military organizations and doctrine, see @kitfield_prodigal_1997.] The ultimate goal of the Goldwater-Nichols Act was to ensure that in an age in which military technologies and equipment were becoming increasingly sophisticated, American national defense interests would be identified and prioritized over and above the parochial organizational interests of the separate military services [@locher_has_2001]. 

Therefore, among the most significant of the changes wrought by the Goldwater-Nichols Act was codification of the Chairman of the Joint Chiefs of Staff as _the_ President's principal military adviser, a change which, in essence, meant that all advice from the respective individual service chiefs of the Army, Navy, Air Force, and Marine Corps would flow through the Chairman of the Joint Chiefs as the single point of contact [@locher_has_2001; @wills_effect_2016]. Furthermore, the Goldwater-Nichols Act created several regional combatant commanders who would report directly to the Secretary of Defense, also bypassing the military service chiefs [@locher_has_2001; @wills_effect_2016]. In essence, the Goldwater-Nichols act institutionalized who from the military would, by law, advise the President and Secretary of Defense on how to employ the military in the world. The Chairman of the Joint Chiefs of Staff and the Combatant Commanders were institutionalized to be potentially powerful voices.  

Finally, in addition to these important institutional changes that cemented the roles of particular military commanders as potentially influential figures within the government, it is also clear that the scope of responsibility of the military has grown, at least in some respects, especially during the post 9-11 era. Although the wars in Afghanistan and Iraq each involved a sizable civilian component for at least a portion of the war, they witnessed the military assuming control for a significant portion of the day to day decision making of the conflict, even over issues that once were, or ostensibly should be, civilian endeavors.^[To be clear, the military and other governmental organizations may not, and in fact, likely do not, advocate that the American military's efforts in waging counterinsurgency wars in Iraq and Afghanistan were examples of how to prosecute a low intensity conflict the "right way." Hopefully, the next several years will provide military and importantly, also other governmental organizations, the opportunity to continue reflecting on what went well and poorly from the wars in Iraq and Afghanistan. The topic of the interaction (or lack thereof) of military and civilian efforts during the Iraq and Afghanistan conflicts is a burgeoning literature that includes a mixture of official government reports, first-hand memoirs, and works that offer theoretical/historical insights. For a sample of this literature, see @iraq_study_group_us_iraq_2006; @ricks_fiasco_2006; @chandrasekaran_imperial_2007; @hodge_armed_2011; @chandrasekaran_little_2013; @gallagher_day_2019]

In summary, today's military has become incredibly prestigious in part because of the institutional reforms enacted over the past several decades. In particular, the National Security Act of 1947 and the Goldwater-Nichols Act of 1986 empowered certain military positions, to include the Chairman of the Joint Chiefs of Staff and the Combatant Commanders, to have the potential to be tremendously influential voices, particularly with respect to advising the Secretary of Defense and the President. Additionally, the military's experiences of prolonged and repeated deployments in the post-9/11 era have also served to strengthen the military relative to other US institutions. The legislative changes wrought over the previous decades, combined with the past two decades of overseas experience in Iraq and Afghanistan, one scholar has claimed, have had the practical effect of the Pentagon replacing the State Department as the government's "organizational center of gravity for U.S. national security and foreign policy" [@schmidt_silent_2019]. So great has the US military's institutional influence become, claims another scholar, that "Pentagon 'help' is like being hugged by an eight-hundred-pound gorilla: it can squeeze the life right out of civilian efforts" [@brooks_how_2016, 91]. 

As an indicator of the relative disparity between the Department of Defense and perhaps its closest institutional partner, the Department of State, Figure \@ref(fig:dodstate-budgets) displays each department's total budget in 2016 real dollars over the period from 1962-2016. One can clearly detect from the Figure that though the size of the Defense budget has dwarfed that of State for a number of decades, the relative degree to which the Defense budget is larger than that of State's has grown sharply in the post-9/11 era.^[In this dissertation, the predominant measure used to measure military prestige is public trust and confidence in the military, as reported through various public opinion polling. The relative size of the DoD versus State Department Budget could also be used, but the budgetary differences may also reflect different developments, such as the institution of the All-Volunteer Force, which is very expensive to maintain. Therefore, in my view, public opinion measurements of trust and confidence in the military are better (though still imperfect) measures of the military's level of prestige.]

```{r dodstate-budgets, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.height=4.5, fig.width=6, fig.cap="Budgets for US Departments of Defense and State, 1962-2016"}

budgetDF <- read_csv(here("data", "DoDStateBudgets.csv")) %>% view()

#scale the data from millions to billions of annual dollars
budgetDF_adj <- budgetDF %>% mutate(
  Defense=Defense/1000, 
  State=State/1000
) %>% view()

melted_budgetDF <- melt(budgetDF_adj, id="Year")

melted_budgetDF %>% ggplot(aes(x=Year, y=value, color=variable)) + geom_point() + geom_smooth(se=FALSE) + labs(color="Department", subtitle="Data Available from Table 4.1, Outlays by Agency,\n https://obamawhitehouse.archives.gov/omb/budget/Historicals") +   scale_color_manual(values=c("black", "blue")) + theme(legend.position="right", plot.title=element_text(hjust=.5)) +  scale_y_continuous(name="Billions of Dollars (2016 Real Terms)", limits=c(0, 700)) 


```

## The Military's Growing Exclusivity Within American Society

A second contributor of the US military's growing prestige is the degree to which the military institution is esteemed by society. In the post-Vietnam War era, even as the US military has accrued significant influence relative to other government institutions, the Armed Forces have also emerged, at least by many measurements, as the public's most trusted institution [@noauthor_confidence_2020]. While there is nothing inherently wrong with a military that is highly regarded by the public, contemporary scholars have expressed concern about the military being regarded _too highly_. Within the civil-military relations scholarly literature, the topic of the military's relational distance between the military and society is often referred to as the civil-military "gap" [@feaver_soldiers_2001]. For the past twenty years or so, numerous scholars have addressed the gaps's origins, scope, implications, and have raised numerous solutions for shrinking the gap [@feaver_soldiers_2001; @schake_warriors_2016]. Scholars further assert that in the years since the Vietnam War, at least three separate gaps have formed between the military and American society: a geographical gap [@liebert_midlife_2017], a partisan gap [@golby_duty_2011; @liebert_midlife_2017], and a generational gap [@colford_young_2016]. 

Before summarizing each of these three separate gaps, it is important to assert a couple of observations about the concept of a gap in general between society and the military. At first glance, one might think that a gap between society and the military might diminish, rather than enhance, the military's overall level of prestige. After all, a nation that is entirely separated from its military, in the sense that it is unable to observe, relate to, and know anything about the military or those that serve in it, is unlikely to foster appreciation and respect for the military as an institution. However, as I summarize in the paragraphs below, most scholars have expressed the opposite concern: too big of a gap between society and the military leads to society placing too much trust in - or at least, too much blind trust in - the military. 

In the United States, the gap between society and the military has changed over time. In the American context, Americans have generally been suspicious of large numbers of standing military forces since the founding of the United States. In fact, the American founders held significant differences of opinion over how large the nation's military forces should be, and whether these forces should be comprised of volunteers (favored by Thomas Jefferson) or a standing military (favored by Alexander Hamilton) [@weigley_american_2001]. In the end, the founders established a hybrid concept, forming a moderately-sized standing Army that could be augmented, as required, by state militias comprised of "citizen soldiers" [@weigley_american_2001]. 

Furthermore, for most of its history, the United States has relied on the "citizen soldier" concept when fielding military forces, especially for war. It was largely not until the 20th century when the government faced wartime demands that required instituting mandatory military service, particularly during World War Two and later, Vietnam [@rostker_i_2006]. The Vietnam War proved to be tremendously unpopular with domestic American audiences as the length of the war increased, especially after the North Vietnamese initiated their "Tet Offensive" in 1968 [@crane_avoiding_2002; @summers_strategy_1982; @petraeus_american_1987]. Moreover, in the latter years of the war, Americans came to view the existing draft system as inequitable from racial and economic perspectives [@rostker_i_2006]. The growing population of the United States, which from a "supply" perspective, translated into more available manpower, combined with severe disciplinary issues that plagued many of the military services at the end of the war [@romjue_active_1984], led many Americans to believe that instituting an All-Volunteer Force (AVF) was at least a viable alternative to maintaining the draft [@rostker_i_2006]. The watershed moment arrived in 1973, when the United States formally abandoned the draft and created the All-Volunteer Force (AVF), the result of recommendations made by the Gates Commission, established previously by President Richard Nixon [@rostker_i_2006]. Since that time, only volunteers have served in the military, for better or for worse. 
The first gap that scholars write about is a geographical gap between parts of the United States and the military. Golby and Liebert, for instance, note that a disproportionate number of those serving in the military hail from rural and southern states, and that a large proportion of military recruitment stations, Reserve Officer Training Corps (ROTC) programs, and military installations are located in the southern United States [@liebert_midlife_2017]. The presence of this particular gap, moreover, raises concerns that the military is disproportionately attached to particular regions of the country. 

A second gap that scholars have identified is a partisan gap between the military and portions of American society. This gap is a little more complicated to fully understand. Although the military officer corps does in fact disproportionately identify with the Republican Party versus the Democratic Party [@golby_duty_2011; @liebert_midlife_2017], studies have also shown that differences between the military and civilians become smaller if not altogether diminished after conditioning on partisan affiliation [@golby_duty_2011]. In other words, the data shows that the military officer corps does identify with the Republican Party more so than with the Democratic Party, but not that military and civilian values or goals are altogether drastically different from each other. 

Finally, and perhaps most problematically in terms of looking ahead into the future, studies have shown that a third gap exists between the military and younger Americans, a gap that has grown as fewer Americans have a firsthand connection to someone either serving or who has served in the military [@colford_young_2016]. Such a gap is even reflect in recent surveys, which actually show the lowest rates of trust in the military are among those under "30 years of age" [@mcintyre_people_2021].

To the extent that these particular three gaps should, and actually do, matter is the subject of considerable debate. Several scholars argue that as the gap between society and the military has grown, it has simply become too easy to send the military to war, as there is little if any corresponding burden society must bear in doing so [@bacevich_breach_2014; @fallows_tragedy_2014]. Furthermore, these critics argue that as society's corresponding burden in going to war is reduced, the public decreasingly holds elected leaders accountable for going to war [@bacevich_breach_2014; @fallows_tragedy_2014]. The result is that the nation ultimately fights irresponsible wars of choice, which yield indecisive outcomes and an altogether feckless foreign policy [@bacevich_breach_2014]. 

As proof of their argument, these critics point to the military's post-9/11 involvement in Iraq and Afghanistan [@bacevich_breach_2014]. For instance, the historian Andrew Bacevich, himself a retired Army Colonel, points out that as the American military was invading and stabilizing Afghanistan and Iraq in 2001 and 2003 respectively, President George W. Bush lowered taxes and encouraged Americans to travel [@bacevich_breach_2014]. Bacevich further asserts that the net result of the military enduring repeated deployments to Iraq and Afghanistan, and achieving outcomes in these places that fall far short of strategic victory, is that "the state now owns war, with the country consigned to observer status" [@bacevich_breach_2014, 44]. Furthermore, this sentiment has likewise been captured by the journalist James Fallows, who, in an influential article, alleges that Americans in the post-9/11 era have developed a "reverent but disengaged attitude toward the military - we love the Troops, but we'd rather not think about them" [@fallows_tragedy_2014].

In addition to its impacts on American foreign policy, other scholars worry that an excessive gap between society and the military harms the quality of civil-military dialogue by constraining civilian input into the policy process, which in turn can likewise threaten a range of foreign policy outcomes. When interacting with civilian leaders in the context of a sizable gap between the military and society, certain military actors may adopt an attitude such that they believe their military service entitles them to be heard in a deferential manner [@beehner_matter_2020]. A similar dynamic can occur if civilian leaders fail to sufficiently assert themselves into the policy process when interacting with military actors who are incredibly popular. Karlin and Friend, for instance, note the dynamic of "military worship" in which military service is somehow viewed as better, more relevant, or more appropriate than other types of civilian service which ultimately harms civilian control of the military and other democratic norms [@friend_military_2018]. As a result, military actors may insist on solving problems in the manner that they want the job done, and not in a way that aligns with the desire of civilian leaders. 

```{r code, warning=FALSE, echo=FALSE, include=FALSE}
df_trust <- read_csv(here("data","inst_trust.csv")) %>% print()
df_trust <- rename(df_trust, SupremeCourt="Supreme Court")

df_trust_consolidated <- df_trust %>% 
  select(Year, Military, Congress, SupremeCourt, Church, Newspapers) %>% 
  gather(key="Institution", value="value", -Year) 
```

```{r insttrust, warning=FALSE, echo=FALSE, include=TRUE, fig.height=4, fig.width=6, fig.cap = "Public Trust in American Institutions Over Time"}

df_trust_consolidated %>% ggplot(aes(x=Year, y=value)) + geom_line(aes(color=Institution, color = Institution)) + scale_color_manual(values=c("dodgerblue", "darkgreen", "black", "seagreen2", "salmon3")) + labs(x="Year", y="Percent Surveyed Who Place\n A 'Great Deal' or 'Quite a Lot' of Trust In...", subtitle="Data Available from 'Confidence in Institutions,'\n https://news.gallup.com/poll/1597/Confidence-Institutions.aspx")

```

While recent polls show that trust in the military may in fact be slipping [@mcintyre_people_2021], the fact is that relative to other institutions, Americans greatly respect their military. According to recent Gallup public opinion polls, 72 percent of Americans indicate that they trust the military a "great deal" or "quite a lot" - the most of any institution, to include Congress (13%), the Presidency (39%), the police (48%), the Supreme Court (40%), and the church/organized religion (42%) [@noauthor_confidence_2020]. Moreover, trends in trust of various American institutions over time reveal that whereas trust in the military has generally increased over the past 45 years, trust in the other institutions have generally declined or remained the same. These trends are displayed in Figure \@ref(fig:insttrust).

In this dissertation, I argue that a high level of military prestige provides civilian and military actors with an opportunity to leverage public opinion, and thus, with an opportunity to behave in such a way that flaunts the principles of civil-military relations. The next section of this chapter describes how the respective levels of polarization and military prestige combine to form unique domestic political environments that impact the constraining influence of the central principles of civil-military relations on the civilian and military actors who practice them. 

# How Polarization and Prestige Impact the Central Principles of Civil-Military Relations

Having now defined and provided an overview of both the dependent and independent variables, this section of the chapter connects these variables into a cohesive theory. In broad terms, I argue that in the modern American context, high levels of political polarization provide a "motive" for military and civilian leaders to commit political behaviors involving the military that challenge the central standards of civil-military relations, while a high level of military prestige provides these leaders with the "opportunity" to do so [@finer_man_1962]. The application of a "motive" and "opportunities" framework to the contemporary American context follows other works which have done the same to explain military intervention in politics broadly speaking [@finer_man_1962] and in the comparative setting of the Russian military [@taylor_politics_2003]. 

In short, my argument is that high levels of polarization and military prestige alter the willingness of civilian and military actors to abide by the central principles of civil-military relations, which were identified earlier in this dissertation as the principles of civilian control of the military, non-partisanship of the military institution, and the non-interference of the military into certain areas of state policy. In other words, the levels of polarization and military prestige change the degree to which civilian and military leaders are bound in practice by the principles of civil-military relations. Therefore, when both polarization and military prestige are high, both sets of actors commit political behaviors that challenge these principles, creating a rather combustible environment in which civil-military relations occur. When either polarization or military prestige is high, but the other variable is low, we can expect one or both actors to violate these principles, but at a level that is less than when both polarization and military prestige are high. The political environment in which civil-military relations occur may certainly have problems and be filled with tension, but less so on the whole than when both polarization and military prestige are high. When both polarization and military prestige exist at relatively low levels, we can expect fewer instances of political behaviors that violate the central principles of civil-military relations still. Of course, this also does not mean that civil-military relations occur in these types of environments without trouble. 

This section proceeds in three parts. First, I argue how high levels of polarization can provide a motive for civilian and military actors to violate the principles of civil-military relations. I then argue how high levels of military prestige provide an opportunity for civilian and military actors to do the same. Finally, I argue that the combination of relative levels of polarization and military prestige lead to the formulation of four distinct political environments in which civil-military relations occur, and describe how each environment corresponds to unique degrees to which civilian and military actors adhere to the principles of civilian control of the military, non-partisanship, and non-interference. From a broad perspective, this theory therefore attempts to increase the awareness held by scholars, civilian leaders, and military officers of the political environment in which civil-military relations are practiced in the US. The individuals who practice them, and the immediate contexts in which they occur _still_ matter, in my view, but so too do the respective levels of the political factors of political polarization and military prestige.  

## Polarization - A "Motive" to Challenge the Principles of Civil-Military Relations

I argue that as polarization rises, the degree of _ideological unity_ among military and civilian actors decreases. This occurs because highly polarized climates are marked not merely by strong political disagreements (which are nonetheless present when polarization is high), but also by strong disagreement and division over which moral values and purposes should govern or influence society, and to what extent. As polarization rises, therefore, military and civilian actors increasingly encounter situations in which they must make judgments about whether to abide by the fundamental standards of civil-military relations, or to engage in acts of behavior that, while challenge or violate the central governing principles of civil-military relations, nonetheless advance a particular belief system, set of values, or ideology that the officer or civilian leader supports.

Especially in the case of military officers, the tension that arises when faced with a choice to obey either one's professional responsibilities or the dictates of one's personal conscience is neither a new nor rarely discussed phenomenon. Indeed, recent scholarship has recognized the difficulty military officers face when encountering situations in which they must choose between following one's "professional versus private morality" [@cohn_what_2019; @robinson_dissents_2020].^[For a context of morality in war, see also @dubik_just_2016] As scholars have noted, military officers in particular routinely face competing demands of loyalty. That is, the duties required of professional military officership requires that military officers remain loyal to multiple entities: to their civilian leaders, whose orders they follow; to the public, for whom the military institution exists to ultimately defend; to their fellow troops, with whom they serve; and to themselves [@cohn_what_2019; @robinson_dissents_2020]. 

Because these entities can all desire different things from time to time, there are occasions when choosing to be loyal to one entity comes at the direct exclusion of another. Furthermore, even among scholars of civil-military relations, there is relatively little agreement regarding how officers should choose when encountering situations in which their loyalties are divided, with some asserting that civilians should and must always be followed [@kohn_richard_2010], and others arguing that because officers are moral agents, they should never follow an order which they believe is inherently wrong [@milburn_breaking_2010; @milburn_when_2019]. Moreover, making decisions in these types of situations is not easy under any circumstances, nor are these circumstances confined to eras in which the level of polarization in society is high.

What is new from a theoretical perspective is acknowledgement that relative to less polarized settings, highly polarized environments increase the frequency with which military officers encounter these types of situations, as well as the magnitude of the disparity _between_ the outcomes of adhering to the central principles of civil-military relations versus following one's conscience. While it is not possible to predict all of the possible issue areas or situations which could provide a "motive" for civilian and military actors to challenge the governing principles of civil-military relations, the characteristics of today's highly polarized environment in the United States adequately capture why this is so.  

Some military officers will be willing to violate the principle of civilian control when they deem that the preferred policies held and promulgated by a particular civilian leader are not simply politically unwise, but wrong in a moral sense. In a similar vein, some military officers may be willing to violate the principle of non-partisanship if they believe that one of the major political parties possesses a political agenda that is simply wiser from a political perspective than the other party, but rather one which aligns to the moral views of the officer on a particular issue or sets of issues. Finally, some military officers will be more inclined to violate a standard of "non-interference" of the military into certain areas of state policy if they believe that the military as an institution _should_ play a greater role in advancing the underlying ideology or values associated with a particular issue or set of issues.  

Similar dynamics also impact certain civilian leaders in highly polarized environments. While civilian leaders cannot expressly violate the standard of civilian control, certain civilian leaders may be inclined to challenge the principle of non-partisanship of the military if they believe that the advantages to doing so outweigh the potential risks. Likewise, certain civilian leaders may challenge the standard of non-interference by directing the military to fill a role that falls outside of its traditional military expertise if doing so advances the leader's underlying preferred values, ideology, or policies. Moreover, these expectations of behavior follow recent experimental research that shows that a significant majority of Americans believe it is more important to advance candidates for office with whom they share a partisan or ideological affiliation than it is candidates who abide by democratic norms [@graham_democracy_2020], and that citizens find certain military elites credible conditional on these elites sharing similar ideological and partisan convictions [@robinson_michael_danger_2018]. A common theme from this scholarly research is that support is granted on based on what ideological positions one follows, and not whether actors (civilian or military) follow normative guidelines. 

Among the clearest contemporary examples of the dynamics in which polarization can provide a "motive" for civilian and military actors to engage in political behaviors by altering the constraining influence of the standards of civil-military relations is the issue of race in America. For any observer or participant of American politics in the 2020 or 2021 time frame, it comes as no surprise that race has become a if not _the_ polarizing issue in the United States. 

Some Americans, mainly those who identify or associate with the progressive wing of the Democratic Party, hold fast to a set of ideas and principles that are part of a worldview known as _Critical Race Theory_, a concept borne out of academia in the late 1970s and 1980s that views "race [as] a political construction that was invented by white people to give themselves power while excluding all other races from it..." [@lindsay_what_2021]. Furthermore, proponents of Critical Race Theory see the problem of racism as inherent in the very power structures and societal fabric of the West, and thus, view the key to overcoming racism as reshaping or dismantling the "very foundations of the liberal order, including equality theory, legal reasoning, Enlightenment rationalism, and neutral principles of constitutional law" [@lindsay_what_2021]. Finally, proponents of this worldview, such as the historian Ibram Kendi, who authored the wildly popular 2019 book, _How To Be An Antiracist_, claim that it is not enough to be against racism, but that racism - which is inherently in and part of everything - must be aggressively searched for, found, and eradicated [@lindsay_what_2021]. 

Other Americans, mainly those who identify with the political right, reject many of these tenets. While these Americans may disagree among themselves about the extent to which racism in America is inherently "structural" or "systemic," the majority of these Americans acknowledge and regret American's ugly past with racism, even as they affirm that the United States is and has nonetheless been a relatively noble country that has pursued, however imperfectly, positive ideals and goals that have helped millions of people, both within the US and throughout the world. ^[To be clear, there are undoubtedly staunch racists in the United States who truly believe that certain classes of people are inferior to others on the basis of race. I do not have this group of staunch racists in mind when describing the opponents of Critical Race Theory. I have in mind, rather, everyday Americans, including a large number of African Americans, who hold to the view that Critical Race Theory is a somewhat dangerous ideology and worldview.] Furthermore, these Americans view Critical Race Theory as a totalitarian effort that attempts to solve the problem of racism by instituting a form of racism of its own by elevating the experiences of groups who identify as oppressed at the expense of other groups [@lindsay_eight_2020]. Unlike many proponents of Critical Race Theory, these Americans believe that the names of George Washington, Abraham Lincoln, and the like should remain as examples for all Americans to aspire to, despite any connections the historical individuals may have had with racism [@ho_san_2021].

How might the issue of race in America serve as a motive for military and civilian actors to engage in behaviors that challenge governing standards of civil-military relations? Much of the answer depends on where one stands with respect to the issue of Critical Race Theory in the United States. For example, proponents of Critical Race Theory are likely to point to the lack of proportional minority representation among senior general and flag officers and among key occupations such as fighter pilots as proof of "systemic" or "structural" racism in the US, while opponents of Critical Race Theory are unlikely to conceptualize these realities, despite being genuine challenges that require diligence and commitment to rectify, as eternal weights that prevent the US from reaching its full potential as a nation. 

Furthermore, reflecting the dynamics of "conflict extension" previously discussed, in which polarization tends to increase the scope of issues that comprise the political battlefield [@layman_party_2002], Critical Race Theory and the "anti-racist" ideology that stems from it tends to view the world in very sharp contrast in that a person is either an "anti-racist" or a racist - there is no room for middle or neutral ground. From this perspective, the polarizing issue of race in America is one of many that can provide a motive, even for a normatively non-partisan actor such as the military, to commit behaviors that challenge or violate central principles of civil-military relations. Not only is the ideology of Critical Race Theory incompatible with remaining politically neutral, but military actors are likely among the ideology's supporters and opponents. Actors from both camps may believe that the morally and ethically right thing to do is to support/oppose the ideology, even if doing so threatens principles of civilian control, non-partisanship, and the "non-interference" of the military into certain state affairs. 

## Prestige - An "Opportunity" to Challenge the Principles of Civil-Military Relations

Whereas rising political polarization lessens the degree of ideological unity among military and civilian actors, thus providing a "motive" for the military's intervention into politics, rising military prestige advances the military's ability to constitute a  "latent political influence" [@brooks_militaries_2009, 231], and thus provides an "opportunity" for the military to challenge standards of American civil-military relations [@finer_man_1962; @taylor_politics_2003]. In short, the more popular and prestigious a military becomes, the greater capacity the military possesses to engage in political behaviors that defy normative principles of civil-military relations. It is important to underscore that, for purposes of this theory, increasing military prestige - and holding all else equal - increases only the _potential_ for civilian and military actors to engage in political behaviors that defy principles of civil-military relations - and not necessarily that these types of behaviors will in fact increase. The main reason that rising military prestige leads to the military having a greater _potential_ to challenge the governing standards of civil-military relations lies in the relationship between prestige and the potential to shape public opinion. 

A significant amount of scholarship in the subject area of political communication describes how, why, and the extent to which elites can and do influence public opinion [for a sampling of this scholarship, see @saunders_no_2017; @guisinger_mapping_2017; @baum_relationships_2008; and @kertzer_bottom-up_2017]. While scholars differ on the degree to which elites actually sway public opinion, scholars who claim that elites do in fact possess an outsized ability to impact public opinion note that the public looks for two central attributes among elite actors - "knowledge" and "trustworthiness" - when determining which elites to listen to as reputable sources of information [@lupia_democratic_1998, 69-76]. Other scholars claim that the more elites are perceived by the public to hold these attributes in abundance, the greater their ability to "cue" or shape public opinion [@zaller_nature_1992]. 

Furthermore, multiple episodes in American history show that popular military figures have influenced public opinion in significant ways. Ulysses S. Grant and Dwight D. Eisenhower, for instance, were both wartime commanders who, largely as a result of the popularity they commanded with the public, ascended to the Presidency [@teigen_why_2018]. In short, a large body of scholarship shows that the more respected an elite is, especially military elites, the greater _capacity_ or _potential_ he or she poses to move public opinion. 

Even currently-serving high-ranking military officers, who are one class of military "elites", acknowledge the relationship between prestige and public opinion, although they do so, understandably, by eschewing the term "military prestige," and instead use the term "trust" [@odierno_repairing_2014]. These senior military officers understand that without a certain level of prestige (positive reputation among the public), the military's ability to effectively accomplish its mission drops. For instance, when speaking to a group of Army officers, presumably a gathering of Colonels who were scheduled to soon be promoted to the rank of Brigadier General, the former Chief of Staff of the Army, Ray Odierno, made several comments that tied the Army's ability to accomplish its mission with maintaining the public's trust. He stated, for instance, that, "Of great importance within the Army Profession is the duty of each of us to be a steward of our profession and to maintain its essential and hard won trust with society"[@odierno_repairing_2014]. Odierno continued, _"our ability to fulfill our mission depends on trust"_ [@odierno_repairing_2014]. 

The overall level of prestige held by a military dictates the degree to which it is both willing and able to challenge the fundamental principles of civil-military relations. For example, when Colin Powell wrote opinion commentary that challenged Candidate and later, President, Clinton's views on using military force in the Balkans in 1992 and 1993 [@powell_opinion_1992; @powell_us_1992], actions which challenged if not violated the fundamental principle of civilian control of the military, one cannot ignore that Powell's ability to do so stemmed in part from his immense public popularity, or prestige. Having been perceived by the American public as playing a critical role in the US victory during the Gulf War of 1991, Colin Powell indeed posed a considerable "latent political influence"[@brooks_militaries_2009], which he used repeatedly as Bill Clinton took office. Madeleine Albright, Clinton's Ambassador to the United Nations and later, Secretary of State, noted Powell's remarkably high potential to influence policy during Clinton's early for Powell to influence policy. Albright, in a 2006 interview, for instance, noted that, "On a regular basis Colin would come in and do a presentation...he’d go through this and say, 'We can take that hill and we can do that and we can do this. You know we have the best military in the world, but it’s going to take 500,000 men and $500 billion and 50 years.' So he’d lead you up the hill of possibilities and then drop you off the other side, and you’d end up with no options" [@albright_william_2006]. Albright, later in the interview, remarked that although she and Powell were and are close friends, during their time as colleagues under Clinton, there is one question that she wishes that Powell would have answered: "what are you saving your political capital for?" [@albright_william_2006].

The example of Colin Powell constraining options for the Balkans in the early 1990s illustrates how a prestigious military can, and did, challenge the fundamental principle of civilian control of the military. However, there are also examples when civilian actors can harness the military's prestige and similarly challenge the standards or principles of civil-military relations. George W. Bush's use of General David Petraeus to "sell" the Iraq Surge to Congress and the American public [@coll_generals_2008], Donald Trump's deployment of the military to the US southern border ahead of the 2018 midterm elections [@adams_trumps_2018], and a growing number of political candidates who have used the military in campaign advertisements [@moore_biden_2020] all reflect instances in which civilians deliberately harnessed the popularity of the military to sway public opinion, in one form or another. Whether these instances, and to what degree, constitute normatively inappropriate actions that challenge or violate fundamental principles of civil-military relations, and especially the military's non-partisan ethic, remain open debates. In any event, it is clear that a popular and prestigious military that is well liked by the public is often sought to shape opinion, either by civilian actors seeking to win office or implement a desired policy, and/or by military actors who think that they have a certain clout with the public.  

Additionally, _whether_ a prestigious military or military figures are effective and actually do influence public opinion is not a cut and dry issue. Recent civil-military relations studies, such as one that uses survey experiments to examine whether endorsements of presidential candidates by military leaders actually changes public opinion, indicates that they are not [@golby_listening_2013]. In this dissertation, I do not primarily consider the efficacy of these efforts. Rather, for purposes of the theory used throughout this dissertation, it is clear that rising military prestige generally leads to an increase in the _potential_ for the military to sway the public. Thus, rising military prestige provides an "opportunity" for civilian and military actors to engage in behavior that challenge fundamental principles of civil-military relations [@finer_man_1962].  

## Polarization and Prestige Lead to Four Types of Domestic Structures

When we consider the impact of polarization and prestige in tandem, four possible domestic structures or environments emerge, each of which corresponds to a relatively high or low degree of polarization and military prestige. These four structures are displayed in Table \@ref(tab:poss-environment). 

```{r poss-environment, echo=FALSE, include=TRUE, error=FALSE, message=FALSE}

text_env <- tribble(~"Environment", ~"Low Polarization", ~"High Polarization",
  "Low Prestige", "Little Motive or Opportunity to Challenge Principles of Civil-Military Relations. Civilian and Military Actors are Relatively United Ideologically; Political Behaviors Involving the Military Institution are Relatively Private", 
  "There exists Motive, but little Opportunity to Challenge Principles of Civil-Military Relations. Civilian and Military Actors are Relatively Divided Ideologically; Behaviors Involving the Military Institution are Relatively Private",  
  "High Prestige", "Little Motive, but Opportunity to Challenge Principles of Civil-Military Relations. Civilian and Military Actors are Relatively United Ideologically; Behaviors Involving the Military Institution are Public and Visible", 
  "Motive and Opportunity Exists to Challenge Principles/Standards of Civil-Military Relations. Civilian and Military Actors are Relatively Divided Ideologically; Behaviors Involving the Military Institution are Public and Visible" 
  )
 

  kable(text_env, booktabs=T, caption="Types of Domestic Environments for Civil-Military Relations and their Impact on Political Behaviors") %>%
  kable_styling(full_width = F, font_size=10, latex_options="striped") %>% 
  column_spec(1, width="10em") %>% 
  column_spec(2:3, width="15em") 
```

## Low Polarization, Low Military Prestige
Under conditions when polarization and military prestige are both relatively low, military actors possess neither a "motive" nor "opportunity" to intervene in politics [@finer_man_1962]. Low polarization levels indicate there is little if any ideological substance driving military actors to challenge foundational principles of civil-military relations, and low prestige suggests that  the military lacks the ability to leverage public opinion. This leads to the theoretical expectation that while there might certainly be conflict between military and civilian leaders over isolated, context-driven events, there will likely not be widespread and substantive instances of political behavior that challenge or violate the standards or principles of civil-military relations. Historical periods which pose as examples of this domestic structure include the post-Vietnam and pre-World War two eras.

## High Polarization, Low Military Prestige
When polarization is high but military prestige is low, military and civilian actors may in fact be divided ideologically. Some of these actors likely possess a true "motive" to advance particular ideologies, even if doing so involves conducting behaviors that challenge, threaten or violate the governing standards of civil-military relations [@finer_man_1962]. However, as a result of the low level of military prestige that exists in this particular domestic structure, civilian and/or military actors lack the "opportunity" to significantly involve the military institution in ways that challenge the central principles of civil-military relations. This leads to the theoretical expectation that while there might be some instances of civilian or military actors engaging in less-visible forms of behavior that challenge the foundational principles of civil-military relations, these efforts will stop short of attempting to harness the military to sway public opinion. Historical examples of this era include the Gilded Age through the beginning of American involvement in World War One, or approximately 1875-1915.

## Low Polarization, High Military Prestige
Under conditions of low polarization but relatively high military prestige, military and civilians actors are relatively united ideologically, and therefore do not possess a strong ideological "motive" to engage in political behaviors that challenge or threaten the principles of civil-military relations [@finer_man_1962]. However, the "opportunity" to use the military institution to leverage public opinion exists [@finer_man_1962]. Under these conditions, it is possible that military and/or civilian actors will engage in forms of political behaviors that harness or leverage the military to move public opinion. This leads to the theoretical expectation that under these domestic conditions, military actors in particular pose a considerable "latent political influence" because they are popular with the public [@brooks_militaries_2009, 231]. The military under these conditions may pose a considerable force to block, obstruct, or influence debates about the use of military force and national security more broadly.  Examples of this period include the post-World War Two until the Korean War, and the post-Gulf War period (1991) until approximately 2010.

## High Polarization, High Military Prestige
When both polarization and military prestige exist at relatively high levels, military and civilian actors possess both a "motive" and an "opportunity" to use the military in ways that challenge, threaten, or violate the fundamental principles of civil-military relations [@finer_man_1962]. Under these conditions, there is a greater likelihood that certain civilian and/or military actors believe so strongly in particular political issues or sets of political issues that they are willing to push the boundaries of the principles of civilian control, non-partisanship, and "non-interference" of the military by engaging in behaviors activities that do so than when polarization is low. Moreover, as a result of the military's relatively high degree of prestige, military and civilian actors will seek to use the military in public and visible ways, as they deem the military institution capable of impacting public opinion. Examples of this period include the post-US Civil War period during American Reconstruction, 1865-1870, and from approximately 2010 - present day. 

# Hypotheses to be Tested

Based on the theoretical predictions that follow from Table \@ref(tab:poss-environment), it is possible to derive the following six hypotheses: 

H1: Military actors are increasingly active in a partisan manner when polarization is high relative to periods of low polarization.

H2: Military elites increasingly challenge civilian authorities in a visible, public manner during periods of high military prestige relative to periods of low military prestige.

H3: When military prestige is high, the level of political polarization moderates the issues over which military actors challenge civilian actors. When military prestige is high and political polarization is low, military actors will likely challenge civilians on use of force initiatives, military policy, and other issues that are directly related to national security. However, when military prestige is high and polarization is high, military actors are more likely to challenge civilians over political issues that do not directly involve the military, or are indirectly related to national security policy. 

H4: When military prestige is high, civilian leaders will attempt to leverage or harness the military (either as an institution, or individual military figures who possess outsized levels of prestige/popularity) for their own political purposes and advantages relative to when military prestige is low.

H5: When polarization is high, the level of military prestige moderates the propensity of civilian leaders to employ the military (as an institution, or individual military actors) in public ways to support (or denounce) policy initiatives or political candidates. When polarization is high and military prestige is high, civilian actors will use the military in public ways to advocate publicly for policies and/or partisan political candidates. When polarization is high and military prestige is low, however, civilians actors are less inclined to use the military to publicly advocate for policies and/or partisan political candidates.  

H6:  When polarization and prestige are high, there is increased public competition between the civilian political parties for the military institution relative to when polarization and military prestige are low. 

## Summary

This chapter argued that the two independent variables of political polarization and military prestige combine to influence the degree to which civilian and military actors are willing to adhere to the fundamental principles or standards of civil-military relations. In doing so, this chapter argued that a high level of polarization can provide a "motive" for civilian and military actors to behave in ways that challenge or violate standards of civil-military behavior [@finer_man_1962]. This is because, this chapter argued, increasing political polarization reflects disagreement in the populace not only about specific policy matters, but about the values and worldviews that should govern society, to what extent, and why. When polarization is high, civilian and military actors are increasingly willing to challenge civilian leaders, adopt positions that align with partisan entities, and appeal to the military to weigh in on or solve issue areas that are indirectly related to topics that are traditionally associated with the military. 

This chapter also argued that high levels of military prestige can serve as an "opportunity" for the military to engage in politics in ways that threaten or violate principles of civil-military relations [@finer_man_1962]. High levels of military prestige suggest that the military can impact public opinion, and that both civilian and military actors are cognizant of this reality. When the prestige of the military is high, the military as an institution is at a greater risk of featuring in more visible and public ways than when the level of military prestige is low. 

Finally, this chapter argued that both variables need to be considered simultaneously in order to determine if and how military and civilian actors will challenge the fundamental principles of civil-military relations. Because each variable can theoretically take on a relatively high or low value, a total of four structures or environments are possible. The types, forms, and overall pattern of political behaviors that each type of actor, civilian and military, are likely to commit involving the military institution, correspond to one or more of these environments.  

The next four chapters of the dissertation test the predictive power of the theory laid out thus far. The first two chapters each explore one "side" of the civil-military relationship - not in isolation from the other, but in terms of a distinct behavior that each actor can commit. The next chapter focuses on retired military officer and the behavior of public opinion commentary, while the subsequent chapter focuses on civilian candidates, interest groups, and political parties and the behavior of campaign advertisements.

Following these two "large-N" chapters, each of which incorporates mainly a mainly quantitative analysis, the dissertation then conducts two separate case studies. In each of these studies, I trace the impacts of polarization and military prestige on the form and pattern of the distinct behaviors committed by civilian and military actors that involve the military institution.

\pagebreak
\singlespace
\newpage


