---
title: "Chapter 2: How Polarization and Prestige Impact the Constraining Influence of Principles of US Civil-Military Relations"
author: "Peter M. Erickson"
date: |
  `r format(Sys.time(), '%B %d, %Y')`
abstract: |
  
  _This chapter advances a theory that describes how the political behaviors that involve the US military are impacted by two distinct factors: the level of domestic political polarization within a society, and the degree of prestige the military institution commands. I argue that these two factors greatly shape the constraining influence of three central principles of civil-military relations: civilian control of the military, the non-partisanship of the military institution, and the "non-interference" of the military into certain realms of state politics. The chapter further argues that a total of four possible domestic environments are possible, each of which is marked by relative levels of polarization and military prestige, and that each environment leads to unique patterns of political behavior conducted by civilian and military actors which involve the military._  
  
geometry: margin = 1.15in              # margin widths
fontsize: 12pt                         # text size
bibliography: dissertation.bib         # bibliography file
biblio-style: authoryear               # citation style
urlcolor: blue                         # links to external URLs
citecolor: black                       # links to bibliography
linkcolor: magenta                     # internal figures, footnotes, etc.
indent: true                          # paragraph format
                                       # true for indented paragraphs
                                       
header-includes:
    - \usepackage{setspace}
    - \usepackage{multirow}
    - \usepackage{indentfirst}
   
    
output: 
  bookdown::pdf_document2:
    keep_tex: true
    latex_engine: pdflatex
    number_sections: false
    toc: false
    fig_caption: true
    citation_package: biblatex
    includes: 
      in_header: 
        - latex-preamble.tex
  bookdown::htmldocument2: default
  bookdown::word_document2:
      toc: false
---

\doublespacing 

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo=FALSE,
  include=FALSE,
  message=FALSE,
  error=FALSE,
  fig.height=3,
  fig.width=4,
  out.width="90%")
```

```{r, include = FALSE, echo = FALSE}
library("here")       # file paths
library("tidyverse")  # workhorse package
library("tidylog")
library("kableExtra")
library("knitr")
library("ggdag")
library("dagitty")
library("gridExtra")
library("broom")
library("dplyr")
library("reshape2")
library("pacman")
library("ggrepel")
```

\newpage

# Introduction

The goal of this chapter is to develop a comprehensive yet parsimonious theory that explains how and why civilian and military actors might be less constrained by the central principles of civil-military relations, which were explained in detail in the previous chapter. The core argument developed in this chapter and tested throughout the remainder of this dissertation, is that changes in the levels of domestic polarization and the prestige of the military ultimately impact the _constraining influence_ of the principles of civilian control of the military, non-partisanship of the military institution, and "non-interference" of the military into certain realms of state politics. Furthermore, this argument claims that when the principles of civil-military relations no longer effectively constrain military and civilian actors, these actors will commit political behaviors that align with recognizable and, to some degree, predictable patterns. In making this argument, this chapter emphasizes _why_ levels in polarization and military prestige alter the constraining influence of civil-military principles and describes the patterns of political behavior to expect when these variables exist at relatively high versus low levels.  

This chapter proceeds in three parts. First, I define the relevant terms that are used throughout this study. Having already established a basic definition and understanding of the origin and meaning of the three foundational principles of American civil-military relations in the previous chapter, I define and trace the developments of the dependent variable, the _political behaviors_ which involve the military institution, and the independent variables, _political polarization_ and _military prestige_. In the second part of this chapter, I explain how the levels of polarization and military prestige impact the constraining influence that the governing principles of civil-military relations have on the propensity of military and civilian actors to engage in various political behaviors that involve the military institution. This second part of the chapter concludes by arguing that a total of four domestic political structures, or environments, are possible, each of which is marked by the presence of relatively low versus high levels of polarization and military prestige, and that each environment is likely to produce certain patterns of political behavior committed by civilian and military actors involving the military institution. Before concluding, the third and final part of the chapter then presents a total of six hypotheses to be tested, analyzed, and explored throughout the remainder of the dissertation.  

# Components of the Theory

A visual representation of the overall theory explained in this chapter and throughout the dissertation is shown below in Figure \@ref(fig:overview) . From left to right, the core argument is that the two independent variables of political polarization and military prestige influence an intervening variable, which is the constraining influence of the fundamental principles or standards of civil-military relations. This core argument further suggests that the overall constraining influence of the fundamental principles of civil-military relations on civilian and military actors impacts the dependent variable, which is certain types and patterns of political behaviors. 

```{r theory-shell, include=FALSE, echo=FALSE}
problem_dag <- dagify(
    PB ~ CI,
    CI ~ MP + PP,
    coords = tribble(
      ~ name,      ~ x,    ~ y,
      "PB", 1,.5,
      "CI", .7, .5,
      "PP", .3, .7,
      "MP", .3, .3,
    ))
    

problem_dag %>% 
  ggplot(aes(x=x, y=y, xend=xend, yend=yend))+
  geom_dag_edges()+
  geom_dag_point()+
  geom_dag_text_repel(aes(label=name), show.legend=FALSE)+
  theme_dag()
```

```{r overview, include=TRUE, echo=FALSE, fig.height=4, fig.width=5, fig.cap="The Impact of Polarization and Military Prestige on Principles of Civil-Military Relations and Political Behaviors"}

problem_dag %>% tidy_dagitty() %>% 
  dag_label(labels=c("PB"="DV\n Political\n Behaviors",
                     "CI"="Int Var\n Constraining Influence\n of Foundational\n Civil-Military Principles", 
                     "MP"="IV\n Military\ Prestige", 
                     "PP"="IV\n Political\n Polarization"
                      )) %>% 
  ggplot(aes(x=x, y=y, xend=xend, yend=yend))+
  geom_dag_edges()+
  geom_dag_point()+
  geom_dag_text()+ 
  geom_dag_label_repel(aes(label=label, fill=label), col="white", show.legend=FALSE)+ theme_dag(plot.title=element_text(hjust=.5))
  
```

## The DV: Political Behaviors

### Military Actors

The dependent variable, the types and patterns of political behaviors which involve the military institution, must be carefully defined. This is especially true because _both_ military and civilian actors can each commit behaviors that challenge or threaten the principles of civil-military relations particularly when examining both military _and_ civilian actors. Yet at the same time, these actors and the particular behaviors they commit are not the same. Elected civilian leaders are inherently political to begin with, and thus, everything they say and do should, at least to some extent, be seen as "political."

This is not the case for military actors, even as scholars have argued that the military will always retain an inherently political flavor that can never be entirely discarded [@brooks_paradoxes_2020; @owens_us_2011; see also @clausewitz_war_1976]. In the scholarly literature, perhaps the key consideration when separating the "political" behaviors of military actors vice "non-political" behaviors that military actors commit is whether the behavior is conducted for the purpose of "influencing" policy outcomes [@brooks_militaries_2009, 213]. While there an almost limitless number of forms such behaviors can take, one helpful framework of these types of behaviors, developed by Risa Brooks (2009), is displayed below in Table \@ref(tab:pol-behav).  

\singlespace

```{r pol-behav, echo=FALSE, include=TRUE, error=FALSE, message=FALSE}

text_tbl1 <- tribble(~"Tactic", ~"Nature of Activity", ~"Audience", ~"Examples",
  "Public Appeal", "Public, value-laden commentary", "Mass public",  "Colin Powell's 1992 statements on intervention in Bosnia", "Grandstanding", "Threat or actual resignation in protest of policy", "Mass public, Congress, executive branch", "Hypothetical; if Eric Shinseki had resigned in 2003",
  "Politicking",  "Retired officer endorsements; organizing vote drives", "Mass public", "Crowe's endorsement of Clinton in 1992; Frank's endorsement of Bush in 2004", 
  "Alliance Building",  "Military leaders form ties with civilian interest groups", "Congress", "Air Force and Lockheed Martin over procurement of F-22 raptor",
  "Shoulder Tapping", "Military leaders setting an agenda by bringing issues to attention of politicians and engaging in lobbying-like activities on behalf of those issues", "Congress", "Military Mobilization of key members of Congress over gays in the military in early 1990s")
 

text_tbl1 %>% kable(booktabs=T, caption="Political Activities of the Military") %>%
  kable_styling(full_width = F, font_size=8, latex_options="striped" ) %>% 
  column_spec(1, width="10em") %>% 
  column_spec(2, width="15em") %>% 
  column_spec(3, width="10em") %>% 
  column_spec(4, width="15em") %>% 
  add_footnote(label = "The above table is reproduced from Brooks, Risa (2009), 'Political Activities of the Military' in\n American Civil-Military Relations: The Soldier and the State in a New Era, Eds. Suzanne Nielsen and Don Snider.")

```
\doublespace

Determining which behaviors listed in Table \@ref(tab:pol-behav) are normatively problematic - and when - is a difficult if not impossible endeavor. As Brooks (2009) notes, rendering a normative assessment regarding political activities committed by military officers depends on specific contexts and circumstances, as certain aspects of engaging in these behaviors and activities can also yield positive repercussions [@brooks_militaries_2009]. For instance, when military actors offer commentary to the public, the information they provide can help citizens better understand and process information that is often complex and disconnected from everyday life, particularly if the issue is somewhat technical or complex [@brooks_militaries_2009, 225-231; see also @schultz_democratic_1999]. Furthermore, militaries can alert Congress to potentially destructive decisions an executive may be considering, thus helping the nation avoid undertaking a course of action that will lead to disastrous consequences [@brooks_militaries_2009, 225-231]. While the final decisions ultimately belong to civilian leaders to make, there are nonetheless real and positive benefits of military actors engaging in the political process in overt ways. 

As Brooks notes, however, any benefits of military actors committing political behaviors in the short-term may be outweighed by more deleterious long-term impacts [@brooks_militaries_2009]. When military actors engaging in political activities, they generate  _political costs_ for civilian leaders. These costs vary, but include constraining the options available to civilian leaders, sparking public backlash, and otherwise influencing debate about an impending policy choice away from the true preferences of civilian leaders [@brooks_militaries_2009]. Therefore, and as Brooks (2009) notes, military actors who engage in political activities do so while facing four unique risks. 

A first risk of military actors engaging in political activities is to society's esteem of the military [@brooks_militaries_2009, 234]. Part of the reason the military has maintained its popularity with American society is precisely because it has been seen and viewed as a professional institution that does not routinely engage in inappropriate political activities. A second risk of military leaders committing politically inappropriate acts is that doing so provides civilian leaders with an estimate of a particular military leader's political views, which could jeopardize how civilian leaders consult with that particular but also other military actors [@brooks_militaries_2009, 234]. A third risk which military actors face when engaging in political activities is to the tenets of "military professionalism" laid out by Huntington directly: military actors engaging in political activities renders them less focused on their professional duties of securing the Republic [@brooks_militaries_2009, 235; @huntington_soldier_1957]. A fourth and final risk, one which Brooks argues is the most important to consider, of military actors engaging in appropriate political expression is that doing so could "undermine" civilian leaders and ultimately, the governing principle of civilian control [@brooks_militaries_2009, 235]. Ultimately, Brooks (2009) concludes that the potential benefits of military actors engaging in political activities must be weighed against the potential consequences, and recommends that for the sake of long-term, stable civil-military relations, military actors rarely engage in these and other political activities [@brooks_militaries_2009].

Two other implications are worth noting regarding the activities listed in Table \@ref(tab:pol-behav). First, not _all_ of these listed behaviors are of the qualitative nature such that they generate significant political costs for civilian leaders in every instance. For example, military officers who engage in making "public appeals" can do so, theoretically, without inducing significant political costs. For example, when General Colin Powell wrote a series of op-eds in the early 1990s while serving as the Chairman of the Joint Chiefs of Staff in which he articulated his views regarding the use of force, his behavior undoubtedly generated political costs for the candidate, and later, President, Bill Clinton [@powell_opinion_1992; @powell_us_1992]. Engagement in this behavior by a highly respected and prestigious figure likely influenced public opinion and constrained the options available to Clinton as a number of crises erupted across the world in the Balkans, Somalia, Haiti, and Rwanda. Similar levels of costs are not likely generated every time a retired General or Admiral pens an op-ed, or gives an interview to the press. 

Conversely, every instance of a military officer publicly endorsing a political candidate is an inherently political act that by definition, violates the governing principle of non-partisanship, and creates at least some political costs for civilian leaders (i.e., the candidate not endorsed). Regardless of whether the endorsement of a political candidate made by a military actor is effective in persuading the public to change its vote choice [see @golby_military_2012], the endorsement it is impossible for a military officer to make a nonpartisan endorsement of a political candidate. Other behaviors, such as resignation, have sparked considerable debate among scholars of civil-military relations. Some have argued that the public nature of resignation constitutes an inherently political act that should not be practically considered [@golby_beyond_2015], while others have left room for the possibility that in extreme circumstances, resignation is appropriate, regardless of the potential public fallout that may be generated [@dubik_principled_2014; @snider_strategic_2014].

The second implication of Table \@ref(tab:pol-behav) is simply that military actors have a variety of activities at their disposal from which to choose to generate political costs for civilian leaders. The well known collisions that have occurred between the President and various wartime Commanders throughout American history - Lincoln and McClellan, Truman and MacArthur, Obama and McChrystal - perhaps threaten civil-military relations in the most recognizable and public way, but there are other subtle and nuanced means available to the military which also generate political costs that civilian leaders must contend with. 

## Behaviors Committed by Civilians

Civilian leaders can also engage in behaviors that challenge the governing principles previously exposited, though in ways that are different from military actors. To begin with, several of the activities listed in Table \@ref(tab:pol-behav) involve varying degrees of civilian actor acquiescence or involvement. Take, for example, the behavior of military actors endorsing political candidates for office. Most instances of this behavior constitute a figurative two way street in which a candidate asks a military officer to make an endorsement, and a military officer agrees to do so. It was for this reason that, in the aftermath of the Presidential Campaign of 2016, Retired Army General Martin Dempsey, who previously served as the Chairman of the Joint Chiefs of Staff, published an opinion-editorial piece in _The Washington Post_ in which he criticized both the military officers (retired Army Lieutenant General Mike Flynn, who spoke on behalf of Donald Trump at the Republican convention, and retired Marine Corps General John Allen, who spoke on behalf of Hillary Clinton at the Democratic Convention) who spoke at the respective conventions as well as the civilian leaders who sought their participation. Dempsey stated in the op-ed, "As generals, they have an obligation to uphold our apolitical traditions. They have just made the task of their successors – who continue to serve in uniform and are accountable for our security - more complicated. It was a mistake for them to participate as they did. It was a mistake for our presidential candidates _to ask them to do so_" [@dempsey_military_2016]. 

Whereas the political activities committed by military actors can begin to be considered problematic insofar as they attempt to influence policy outcomes and/or generate political costs for civilian leaders [@brooks_militaries_2009], inappropriate activities at the hands of civilian actors which involve the military institution are those that "politicize" the military. More precisely, the behaviors committed by civilian leaders which "politicize" the military are those that cause the military to either breach or be perceived by the public as breaching the principles of non-partisanship or non-interference (by definition, as was argued in the previous chapter, it is not possible for civilian actors to violate the principle of civilian control). 

Civilian leaders have faced consequences for "politicizing" the military previously. For example, some observers alleged that President George W. Bush's use of General David Petraeus in garnering Congressional support for the "surge" strategy in Iraq politicized the military in unhealthy ways [@coll_generals_2008], while other observers note the increase in retired military generals and admirals who endorse political candidates at conventions [@exum_dangerous_2017] as troubling indicators of the military's politicization. Moreover, under President Donald Trump's tenure in office, charges of politicizing the military were common, perhaps none as vociferous as the reaction to the former President's decision 2018 decision to deploy several thousand troops to the US border with Mexico in response to an impending caravan of migrants from South America. The headline of an op-ed published by several retired Army Colonels in response aptly summarizes the feelings held by many Americans at the time: "Trump's Border Stunt is a Profound Betrayal of Our Military" [@adams_trumps_2018]. 

In summary, military and civilian leaders can engage in certain behaviors which violate the fundamental governing principles of civil-military relations. Military actors have a variety of behaviors at their disposal which, if committed, can generate substantial political costs for civilian leaders. These costs include constraining the policy choices that are available to civilian leaders, generating negative public opinion against civilian leadership, and otherwise presenting options that fail to satisfy the true preferences of civilians [@brooks_militaries_2009]. Alternately, civilians too can engage in political behaviors which politicize the military, leading to an actual or perceived breach of the military institution's non-partisan ethic or a violation of a norm of non-interference by asking or directing the military to do a task that is at best indirectly or, at worst, entirely unrelated to the defense of the nation. 

Having described the dependent variable of study, the political behaviors that can be committed by military and civilian actors which jeopardize the governing principles of civil-military relations, the next task is to describe each of the independent variables: political polarization and military prestige. After defining each of the variables, I trace the respective levels of each variable in contemporary America. 

# IV1 - Political Polarization

The first independent variable of study in this dissertation is political polarization. Polarization is a difficult concept to precisely define. From the perspective of the American politics sub-field, polarization refers to the movement of political thoughts and attitudes away from the political center and towards more ideological extremes [@hare_polarization_2014; @abramowitz_great_2018]. In recent years, the concept of polarization has also risen within the sub field of international relations, particularly as a domestic political feature that impacts various international outcomes, including whether a polarized country can unite to recognize common threats, learn and derive lessons from foreign policy failure, and agree on what role a nation should play on the world stage [@schultz_perils_2017; @myrick_external_2020]. From a broad perspective, one that is more often explored and discussed within the sub-field of political theory, political polarization describes the degree to which a society shares fundamental values, is capable of incorporating divergent worldviews, and the overall degree to which these values unite versus divide a nation [@marsden_twilight_2014; @deneen_why_2018]. In this dissertation, I rely on both the relatively narrower conception of polarization as the divergence of political attitudes away from an ideological center and towards the extreme, as well as a broader conception of the term that recognizes the difficulties which arise when a society can no longer agree upon the fundamental values and worldviews which should govern it. 

## Perspectives on Polarization

Perspectives regarding the _causes_ of political polarization vary. Americanists tend to explain polarization as stemming from differences of opinion over issues pertaining to race and religion [@abramowitz_great_2018], the size and role of government [@sides_two_2015], normative conceptions of personal and national identity [@sides_identity_2018], and other policy issues [@mann_its_2012]. Political theorists and cultural historians, on the other hand, recognize the current polarizing environment as a symptom of a slow yet steady political shift that began in the 1950s, a shift that reflects both the incompatibility of American pluralism to reconcile and "embrace a wide range of both religious and nonreligious viewpoints" [@marsden_twilight_2014, 171] and the inability of liberalism as a political philosophy to deliver on its intended promises of securing individual rights in a manner that is equitable and fair [@deneen_why_2018]. Both explanations of polarization acknowledge that American society is strained because there is no longer a common vision for the values and purposes that should guide and unite society, and that the various perspectives regarding the values that should perform this immensely important task are inherently at odds with each other.   

Scholars note the presence of three characteristics or features that describe the current polarized political landscape. The first feature is that polarization in the US goes far beyond mere political differences between political elites [@mann_its_2012]. In other words, today's polarized landscape extends far beyond the halls of Congress in Washington D.C. While one can argue that everyday citizens may not feel as strongly as elected politicians about every single political issue at stake today, the fact is that the level of disdain held by regular Americans for those who sit on the opposite side of the political spectrum has risen sharply over the past several decades. This phenomenon is known in the American politics literature in particular as "affective polarization" [@iyengar_affect_2012; @iyengar_origins_2019] or "negative partisanship" [@abramowitz_great_2018]. 

A second feature of contemporary polarized politics in the United States is the increasing scope and number of issues that have become included as part of the partisan battleground. Politicians and citizens do not disagree and wage political war over a handful of important issues, but rather, at least seemingly, over everything! Scholars of American politics describe this dynamic of increasing political warfare as "conflict extension" [@layman_party_2002]. As one might expect, the dynamic of "conflict extension," in which every political issue seems to be fought over bitterly, dampens the prospects for bipartisanship and meaningful cooperation across partisan aisles [@layman_activists_2010]. 

A third and final feature of contemporary political polarization in the United States is that issues of identity are increasingly associated with partisanship - to the point that partisan identity itself is an important component of how people view themselves and others [@sides_identity_2018; @robinson_michael_danger_2018]. This feature of polarized politics is perhaps most clearly illustrated by the rise and influence of American "identity politics" in which the normative roles of race, gender, religion, and immigration invoke very strong commitments from individuals on both sides of the partisan aisle [@white_values_2002; @mann_its_2012; @sides_identity_2018]. The prominence of these identity-related issues in politics affirms that these issues are not merely peripheral interests, but rather issues that individual citizens care deeply about and consider to be of fundamental concern. 

## How Bad is Contemporary Polarization?

```{r polar-measurement, include=FALSE, echo=FALSE}
nom_dat <- read_csv("https://voteview.com/static/data/out/members/HSall_members.csv")

south <- c(40:49,51,53)
polar_dat <- nom_dat %>% 
    filter(congress>45 & 
           chamber != "President") %>%
    mutate( 
      year = 2*(congress-1) + 1789,
    ) %>%
    group_by(chamber,congress,year) %>% 
    summarize(
      party.mean.diff.d1 = mean(nominate_dim1[party_code==200],na.rm=T) - 
                           mean(nominate_dim1[party_code==100],na.rm=T),
      prop.moderate.d1 = mean(abs(nominate_dim1)<0.25,na.rm=T),
      prop.moderate.dem.d1 = mean(abs(nominate_dim1[party_code==100])<0.25,na.rm=T),
      prop.moderate.rep.d1 = mean(abs(nominate_dim1[party_code==200])<0.25,na.rm=T),
      overlap = (sum(nominate_dim1[party_code==200] <
                       max(nominate_dim1[party_code==100],na.rm=T),na.rm=T)  +
                 sum(nominate_dim1[party_code==100] >
                       min(nominate_dim1[party_code==200],na.rm=T),na.rm=T))/
                 (sum(!is.na(nominate_dim1[party_code==100]))+
                  sum(!is.na(nominate_dim1[party_code==200]))),
      chamber.mean.d1 = mean(nominate_dim1,na.rm=T),
      chamber.mean.d2 = mean(nominate_dim2,na.rm=T),
      dem.mean.d1 = mean(nominate_dim1[party_code==100],na.rm=T),
      dem.mean.d2 = mean(nominate_dim2[party_code==100],na.rm=T),
      rep.mean.d1 = mean(nominate_dim1[party_code==200],na.rm=T),
      rep.mean.d2 = mean(nominate_dim2[party_code==200],na.rm=T),
      north.rep.mean.d1 = mean(nominate_dim1[party_code==200 & 
                                             !(state_icpsr %in% south)],na.rm=T),    
      north.rep.mean.d2 = mean(nominate_dim2[party_code==200 & 
                                             !(state_icpsr %in% south)],na.rm=T),    
      south.rep.mean.d1 = mean(nominate_dim1[party_code==200 & 
                                              (state_icpsr %in% south)],na.rm=T),    
      south.rep.mean.d2 = mean(nominate_dim2[party_code==200 & 
                                             (state_icpsr %in% south)],na.rm=T),    
      north.dem.mean.d1 = mean(nominate_dim1[party_code==100 & 
                                              !(state_icpsr %in% south)],na.rm=T),    
      north.dem.mean.d2 = mean(nominate_dim2[party_code==100 & 
                                              !(state_icpsr %in% south)],na.rm=T),    
      south.dem.mean.d1 = mean(nominate_dim1[party_code==100 & 
                                              (state_icpsr %in% south)],na.rm=T),    
      south.dem.mean.d2 = mean(nominate_dim2[party_code==100 & 
                                              (state_icpsr %in% south)],na.rm=T),    
    ) 

head(polar_dat)
write_csv(polar_dat,path="voteview_polarization_data.csv")

```


```{r include=FALSE, echo=FALSE}
# Here are up to date versions of the famous party means over time plots for each chamber:
polar_dat_long <- polar_dat %>% gather(score,value,-chamber,-year,-congress)
labels <- c("dem.mean.d1"="DEM",
            "rep.mean.d1"="REP",
            "north.dem.mean.d1"="N. DEM",
            "south.dem.mean.d1"="S. DEM")

polarized_plot <- function(chamb) {
  pdatl <- polar_dat_long %>% 
                filter(chamber==chamb,
                       score %in% c("dem.mean.d1","rep.mean.d1",
                                    "north.dem.mean.d1","south.dem.mean.d1")) %>%
                mutate(party=labels[score]) %>%
                ungroup()
  
  gg <- ggplot(data=pdatl,
               aes(x=year,y=value,group=party,col=party)) +
               scale_x_continuous(expand = c(0.15, 0), 
                                 breaks=seq(1880, max(pdatl$year), by=8)) +
               geom_line() + geom_point(size=0.7) + 
               xlab("Year") + ylab("Liberal-Conservative") + 
               geom_text_repel(data=pdatl %>%
                    filter(year == min(year)),
                  aes(label = party, color = party),
                  size = 3,
                  nudge_x = -8,
                  point.padding = 0.1,
                  segment.color = NA,
                  show.legend = FALSE) +
              theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
              scale_color_manual(values=c("REP"="#c70828","DEM"="#1372ad",
                                         "N. DEM"="#6194F4",
                                         "S. DEM"="#81c4e4"),guide="none") +
              theme_bw()
  gg
}
```

```{r include= FALSE, echo=FALSE}
# House Polarization Plot
gg1 <- polarized_plot("House")
ggsave("voteview_house_party_means.png", plot=gg1, device="png", 
       width=7, height=4, dpi=300)
ggsave("voteview_house_party_means.pdf", plot=gg1, device="pdf", 
       width=7, height=4)
gg1
```


```{r include = FALSE, echo=FALSE}
# Senate Plot
gg2 <- polarized_plot("Senate")
ggsave("voteview_senate_party_means.png", plot=gg2, device="png", 
       width=7, height=4, dpi=300)
ggsave("voteview_senate_party_means.pdf", plot=gg2, device="pdf", 
       width=7, height=4)
gg2

```

```{r polar-chamber, include=TRUE, echo=FALSE, fig.align='center', fig.cap="Liberal-Conservative Partisan\nPolarization by Chamber, 1880-Present"}
# liberal-conserative polarization by chamber
gg3 <- ggplot(polar_dat_long %>% 
              filter(score == "party.mean.diff.d1"),
              aes(x=year,y=value,group=chamber,col=chamber)) +
              scale_x_continuous(expand = c(0.10, 0), breaks=seq(1880,max(polar_dat_long$year),by=8)) +
              geom_line() + geom_point(size=0.5) + 
              xlab("Year") + ylab("Distance between party means") + 
              geom_text_repel(data=polar_dat_long %>%
                    ungroup() %>%
                    filter(year == min(year) & score=="party.mean.diff.d1"),
                  aes(label = chamber, color = chamber),
                  size = 3,
                  nudge_x = -10,
                  point.padding = 0.1,
                  segment.color = NA,
                  show.legend = FALSE) +
              theme(axis.text.x = element_text(angle = 60, hjust = 1), plot.title=element_text(hjust=.5), plot.subtitle=element_text(hjust=.5)) + scale_color_manual(values=c("House"="#c70828","Senate"="#1372ad"), guide="none") +
               labs(subtitle="Data Available from Lewis et. al. (2020) at Voteview.com")


gg3

```

The most widespread and historical (though certainly, not perfect) measurement of polarization has been that of dynamically-weighted, three step estimation scores (DW-NOMINATE), which are derived from examining roll call voting patterns among members of Congress [@jeffrey_b_lewis_voteview_2020]. Figure \@ref(fig:polar-chamber) shows polarization levels present in the US House and Senate since the latter portions of the 19th Century [@hare_polarization_2014; @jeffrey_b_lewis_voteview_2020]. As Figure \@ref(fig:polar-chamber) demonstrates, polarization levels were high at the end of the American Civil War, and remained high before gradually decreasing early in the 20th century. The figure further shows that polarization reached it's nadir in the years before World War Two, and remained at relatively low levels until approximately the mid-1970s. Since that time, however, polarization in both chambers of the US Congress has increased dramatically, to the point that contemporary levels exceed those previously recorded.  

Polarization can and has been measured in a second way. Formally derived by Iyengar et al (2012) measurements of "affective polarization" have been derived by using data from the American National Elections Studies (ANES). Beginning in the late 1970s, the ANES has asked a series of "feeling thermometer" questions, in which respondents are asked to provide, on a scale ranging from 0-100, how they feel about a range of topics, including their feelings about political parties. Figure \@ref(fig:affect-polar) graphs the results of how partisans feel about members of their own party, members who affiliate with the opposite party, the difference between these two answers, and a total level of affective polarization, which is the sum of the difference for each of the two main parties in the United States.

As Figure \@ref(fig:affect-polar) shows, measurements for how favorable each party is have remained more or less constant over the past several decades. However,partisans have increasingly rated those on the opposite side of the political spectrum as increasingly less favorable over the same time period. In other words, data shows that partisans do not necessarily like the party with whom they affiliate more, but rather that they disdain those on the other side of the aisle at greater levels. 

```{r affect-polar, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.height=5.2, fig.width=7.2, fig.cap="Affective Polarization in the US, 1978-2020"}
df <- read_csv(here("data", "Aff_Polarization.csv"))

slim_df <- df %>% select(Year:AffPol) %>% na.omit %>% view()
melted_df <- melt(slim_df, id="Year") %>% view()

melted_df %>% ggplot(aes(x=Year, y=value, color=variable)) + geom_point() +
                       geom_smooth(se=FALSE) + labs(color="ANES\nFeeling Thermometer\nValue") +   scale_color_manual(breaks=c("DemIn", "DemOut", "DemDiff","RepIn","RepOut","RepDiff","AffPol"), labels=c("Democrat-In","Democrat-Out","Democrat-Diff","Republican-In","Republican-Out","Republican-Diff","Affective Polarization"), values=c("red", "blue", "green","dodgerblue","darkorange","cyan1","black")) + theme(legend.position="bottom", plot.title=element_text(hjust=.5))

```

## Why Does Polarization Matter?

One does not have to look far today in the American political culture to grasp the fact that current levels of polarization are high. Indeed, it is little wonder that both candidates running for President of the United States in 2020 claimed that the very "soul of the nation" was at stake [@dias_biden_2020]. The questions with which the United States is wrestling about itself reflect not only political differences among America's elected leaders, but significant gaps between the worldviews of the nation's citizens and the fundamental core values that should be adopted, implemented, and practiced. A cursory examination of just a handful of these polarizing questions confirms that they are not easy questions, nor likely to be solved without eliciting a strong response from the public. 


* What does it mean to be an American? What are our "American Values"? Are there any we can agree on?
* Should America be proud of its history? 
* Is the United States an inherently racist country?
    + Should we enact a "cancel culture", and if so, who should be cancelled? What should the nation do with its founding fathers, many of whom owned slaves? 
* Should the United States have and enforce borders? Do all people have a "human right" to come to the United States?
    + Should taxpayers fund medical care for illegal immigrants? 
* Are men and women different, and to what degrees? Or are all people the same, in every way? Are the concepts of gender and sex even relevant anymore?

\doublespace
These questions, and many more, illustrate how different worldviews reflect drastically different ideas of how the United States should look moving forward. These questions are qualitatively different from policy questions such as what percentage of national GDP should be spent on the military, whether charter schools should be supported or opposed, and whether the debt ceiling should be raised in times of fiscal crisis and uncertainty. While there are undoubtedly significant policy ramifications that stem from these questions, these listed above (and others) bear significantly on an individual's deeply held beliefs, values, and overall worldview [@white_values_2002; @mann_its_2012; @sides_identity_2018]. How the nation answers them will likely set a tone that will have lasting impacts on the nation's values, now and into the future. 

# IV2 - The Military's Prestige 

The second independent variable of focus in this study is the military's overall level of prestige. Military prestige does not refer to a single indicator but rather a conglomerate of factors, including the military's relative level of influence within the government and its standing among and within society. In this vein, the concept of military prestige is similar, though not identical, to previous studies within civil-military relations that have examined similar variables, such as that of the "balance of power" which exists between political and military leaders [@brooks_shaping_2008]. In short, military prestige refers to the degree of power, respect, and influence the military institution (and specific military actors) has and can wield - both within the government and across society. 

While there are several ways to measure political polarization, measuring the military's prestige is a bit more difficult. On one hand there are clear measurements of several proxies of military prestige, such as public opinion measures of public trust and confidence in the military [@noauthor_confidence_2020]. On the other hand, measuring a concept such as influence - which I argue should be considered when esteeming the overall level of prestige of the military - is problematic. Indeed, as Richard Betts has stated, "influence...is one of the most important and most elusive concepts in the study of politics. In decisions on use of force it is impossible to detect objectively whose influence is greatest; it can only be inferred because only the President - and maybe not even he - really knows what convinces him or changes his mind" [@betts_soldiers_1977, 5]. Nonetheless, with a general understanding of what military prestige is and what it refers to, the next task is to briefly summarize the growth in the US military's prestige in the post World War Two era. Two important drivers of the military's prestige after World War Two have been the growing institutional influence of the military within the national security apparatus and the military's growing exclusivity within American society.  

## The Military's Growing Institutional Influence

One of the main drivers of the US military's prestige over the past seventy-five years has been its increasing institutional power and influence in relation to other government departments and institutions. By virtually every objective measurement, to include budgets/spending and manpower, the US military apparatus is a behemoth of an organization, and one that influences a wide range of policy areas. This has not always been the case, however, in American history. The contemporary influence of the US military as an institution is the result of decades worth of organizational and institutional reform and evolution to the Department of Defense and the respective individual military services. 

One of the most prominent reforms to the military came in the wake of World War Two. The National Security Act of 1947, which was signed by President Harry Truman, sought to unify the defense and intelligence establishments, which despite achieving victory in World War Two, had nonetheless squabbled a fair bit as well [@kaiser_no_2014]. The National Security Act created new institutions, including the National Security Council (NSC), the Central Intelligence Agency (CIA), and the US Air Force [@noauthor_national_1947], and more formally, merged the Departments of War and Navy into what would be first called the "National Military Establishment"- essentially a precursor to what became in 1949 the Department of Defense [@locher_has_2001]. 

For the next several decades, the reforms wrought by the National Security Act helped the defense and intelligence communities prevail in a number of conflicts, including the Cold War, and to ultimately ensure the security of the United States. So influential was the National Security Act that fifty years after its signing, President Bill Clinton stated that as a result of the institutions created by the National Security Act and the people who had served in them, a generation had "helped to secure the peace and prosperity that America enjoys...the success of their efforts and of the historic legislation enacted half a century ago is reflected in an outstanding record of achievement: nuclear war averted, the Cold War won, and the nations of the world turning to democracy and free markets" [@clinton_weekly_1997].

Like most institutional reforms, however, the National Security Act of 1947 could not nor foresee all of the difficulties that the future would bring. A second set of major institutional defense reforms was undertaken in the mid-1980s. Known as the Goldwater-Nichols Act after Senator Barry Goldwater and Congressman William "Bill" Nichols, several incidents made clear that the military lacked the requisite capabilities to operate in "joint" environments, or those which required military personnel and equipment from across the military services [@kitfield_prodigal_1997].^[This included the the failed Iranian hostage rescue mission in 1980. For an excellent account of this event and the demand for "joint" services afterwards, see @kitfield_prodigal_1997.] Especially as military technologies and equipment grew increasingly complex and sophisticated, the Goldwater-Nichols legislation sought to ensure that national defense interests were identified and prioritized over and above the parochial interests of the individual military services [@locher_has_2001]. 

Among the most significant of the changes wrought by the Goldwater-Nichols Act was codification of the Chairman of the Joint Chiefs of Staff as the President's principal military adviser, a change which in essence meant that all advice from the respective individual service chiefs of the Army, Navy, Air Force, and Marine Corps would flow through one single point of contact: the Chairman of the Joint Chiefs [@locher_has_2001; @wills_effect_2016]. Additionally, the Goldwater-Nichols Act created several regional combatant commanders who would report directly to the Secretary of Defense, thus bypassing the military service chiefs [@locher_has_2001; @wills_effect_2016]. 

As important as these and other institutional changes have been in cementing the immense roles of particular military commanders, the scope of responsibilities undertaken by those serving in these roles has also grown considerably, especially during the post-9/11 era. Although the wars in Afghanistan and Iraq each involved a sizable civilian component for at least a portion of the war, they witnessed the military assuming control for a significant portion of the day to day decision making of the conflict, even over issues that once were, or ostensibly should be, civilian endeavors. ^[The topic of the military and civilian experience and expertise, particularly during the Iraq and Afghanistan conflicts, is a burgeoning literature that includes a mixture of official government reports, first-hand memoirs, and works that offer theoretical/historical insights. For a sample of this literature, see @iraq_study_group_us_iraq_2006; @ricks_fiasco_2006; @chandrasekaran_imperial_2007; @hodge_armed_2011; @chandrasekaran_little_2013; @gallagher_day_2019]

In summary, the National Security Act of 1947 and the Goldwater-Nichols Act of 1986 generated significant institutional changes for key positions and organizations within the US Military. Additionally, the military's experiences of prolonged and repeated deployments in the post-9/11 era have also served to strengthen the military relative to other US institutions. The legislative changes wrought over the previous decades, combined with the past two decades of overseas experience in Iraq and Afghanistan, one scholar has claimed, have had the practical effect of the Pentagon replacing the State Department as the government's "organizational center of gravity for U.S. national security and foreign policy" [@schmidt_silent_2019]. So great has the US military's institutional influence become, claims another scholar, that "Pentagon 'help' is like being hugged by an eight-hundred-pound gorilla: it can squeeze the life right out of civilian efforts" [@brooks_how_2016, 91]. 

One graphical depiction of the current relative disparities between the Department of Defense and perhaps its closest institutional partner, the Department of State, is portrayed in Figure \@ref(fig:dodstate-budgets), which shows each department's total budget in 2016 real dollars over the period from 1962-2016. One can clearly detect from the Figure that though the size of the Defense budget has dwarfed that of State for a number of decades, the relative degree to which the Defense budget is larger than that of State's has grown sharply in the post-9/11 era.

```{r dodstate-budgets, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.height=4.5, fig.width=6, fig.cap="Budgets for US Departments of Defense and State, 1962-2016"}

budgetDF <- read_csv(here("data", "DoDStateBudgets.csv")) %>% view()

#scale the data from millions to billions of annual dollars
budgetDF_adj <- budgetDF %>% mutate(
  Defense=Defense/1000, 
  State=State/1000
) %>% view()

melted_budgetDF <- melt(budgetDF_adj, id="Year")

melted_budgetDF %>% ggplot(aes(x=Year, y=value, color=variable)) + geom_point() + geom_smooth(se=FALSE) + labs(color="Department", subtitle="Data Available from Table 4.1, Outlays by Agency,\n https://obamawhitehouse.archives.gov/omb/budget/Historicals") +   scale_color_manual(values=c("black", "blue")) + theme(legend.position="right", plot.title=element_text(hjust=.5)) +  scale_y_continuous(name="Billions of Dollars (2016 Real Terms)", limits=c(0, 700)) 


```

## The Military's Growing Exclusivity Within American Society

A second driver of the US military's growing prestige is the degree to which the military institution is esteemed by society. In the post-Vietnam War era, even as the US military has accrued significant influence relative to other government institutions, the Armed Forces have also emerged, at least by many measurements, as the public's most trusted institution [@noauthor_confidence_2020]. While there is nothing wrong at the surface level with a military that is highly regarded by the public, contemporary scholars have expressed concerns about the military being regarded _too highly_ by society. Within the civil-military relations scholarly literature, the topic of the military's relational distance between the military and society is often referred to as the civil-military "gap" [@feaver_soldiers_2001]. For the past twenty years or so, numerous scholars have addressed the gaps's origins, scope, implications, and have raised numerous solutions for shrinking the gap [@feaver_soldiers_2001; @schake_warriors_2016].

At first glance, one might think that a gap between society and the military might diminish, rather than enhance, the military's overall level of prestige. After all, a nation that is entirely separated from its military, in the sense that it is unable to observe, relate to, and know anything about the military or those that serve in it, is unlikely to foster appreciation and respect for the military institution. While a national military that is fully separated from society is certainly a plausible theoretical outcome, most scholars have expressed the opposite concern for the contemporary United States: the distance and quality of the separation between the modern US military and American society has resulted in the military becoming so prominently respected and trusted that the potential harm caused by this development far outweighs any potential benefits. 

In the United States, and in most countries, the gap between society and the military has changed over time. In the American context, Americans have generally been suspicious of large numbers of standing military forces since the founding of the United States. In fact, the American founders held significant differences of opinion over how large the nation's military forces should be, and whether these forces should be comprised of volunteers (favored by Thomas Jefferson) or a standing military (favored by Alexander Hamilton) [@weigley_american_2001]. In the end, the founders established a hybrid concept, forming a moderately-sized standing Army that could be augmented, as required, by state militias comprised of "citizen soldiers" [@weigley_american_2001]. 

Furthermore, for most of its history, the United States has relied on the "citizen soldier" concept when fielding military forces, especially for war. It was predominantly not until the the 20th century when the government faced wartime demands that required instituting mandatory military service, particularly during World War Two and later, Vietnam [@rostker_i_2006]. The Vietnam War proved to be tremendously unpopular with domestic American audiences as the length of the war increased, especially after the North Vietnamese initiated their "Tet Offensive" in 1968 [@crane_avoiding_2002; @summers_strategy_1982; @petraeus_american_1987]. Moreover, in the latter years of the war, Americans came to view the existing draft system as inequitable from racial and economic perspectives [@rostker_i_2006]. The growing population of the United States, which from a "supply" perspective, translated into more available manpower, combined with severe disciplinary issues that plagued many of the military services at the end of the war [@romjue_active_1984], led many Americans to believe that instituting an All-Volunteer Force (AVF) was at least a viable alternative to maintaining the draft [@rostker_i_2006]. The watershed moment arrived in 1973, when the United States formally abandoned the draft and created the All-Volunteer Force (AVF), the result of recommendations made by the Gates Commission, established previously by President Richard Nixon [@rostker_i_2006]. 

While it is not possible to fully exposit here the relative benefits and disadvantages of the All-Volunteer Force, a task which would require a full complement of empirical and normative insights, a couple of broad summary statements speak to the overall impact of the All-Volunteer Force. On one hand, it is difficult to find a scholar, pundit, or senior military officer who does not laud the tactical proficiency and overall skill of the AVF. Indeed, there is near-universal acceptance of the tactical acumen developed and honed by the AVF since its inception. On the other hand, scholars assert that in the years since the military draft ended, there have been at least three noteworthy gaps that have formed between the military and American society: a geographical gap [@liebert_midlife_2017], a partisan gap [@golby_duty_2011; @liebert_midlife_2017], and a generational gap [@colford_young_2016]. 

Regarding the geographical gap, Golby and Liebert note that a disproportionate number of those serving in the military hail from rural and southern states, and that a large proportion of military recruitment stations, Reserve Officer Training Corps (ROTC) programs, and military installations are located in the southern United States [@liebert_midlife_2017]. The presence of this particular gap, moreover, raises concerns that the military is disproportionately attached to particular regions of the country. The partisan gap is a little more complicated. Although the military officer corps does in fact disproportionately identify with the Republican Party versus the Democratic Party [@golby_duty_2011; @liebert_midlife_2017], studies have also shown that differences between the military and civilians become smaller if not altogether diminished after conditioning on partisan affiliation [@golby_duty_2011]. In other words, the data shows that the military officer corps does identify with the Republican Party more so than with the Democratic Party, but not that military and civilian values or goals are altogether drastically different. Finally, and perhaps most problematically, studies have shown that a third gap exists between the military and younger Americans, a gap that has grown as fewer Americans have a firsthand connection to someone either serving or who has served in the military [@colford_young_2016]. Furthermore, in recent surveys which actually display that trust in the military is beginning to decrease, the data shows that the lowest rates of trust in the military are among those under "30 years of age" [@mcintyre_people_2021].

To the extent that these particular three gaps should, and actually do, matter is the subject of considerable debate. In more general terms, critics of the AVF allege that as the gap between society and the military has grown, regardless of any increase in the military's tactical abilities, it has simply become too easy to send the military to war, as there is little if any corresponding burden society must bear in doing so [@bacevich_breach_2014; @fallows_tragedy_2014]. These critics further allege that as society's burden in going to war shrinks, the public does not hold elected leaders as accountable for going to war, and that the nation ultimately will fight irresponsible wars of choice, which ultimately result in both indecisive outcomes and an altogether feckless foreign policy [@bacevich_breach_2014]. 

As proof of their argument, these critics point to the military's post-9/11 involvement in Iraq and Afghanistan [@bacevich_breach_2014]. For instance, the historian Andrew Bacevich, himself a retired Army Colonel, points out that as the American military was invading and stabilizing Afghanistan and Iraq in 2001 and 2003 respectively, President George W. Bush lowered taxes and encouraged Americans to travel vacation [@bacevich_breach_2014]. Bacevich further asserts that the net result of the military enduring repeated deployments to Iraq and Afghanistan, and achieving outcomes in these places that fall far short of strategic victory, is that "the state now owns war, with the country consigned to observer status" [@bacevich_breach_2014, 44]. Furthermore, this sentiment has likewise been captured by the journalist James Fallows, who, in an influential article, alleges that Americans in the post-9/11 era have developed a "reverent but disengaged attitude toward the military - we love the Troops, but we'd rather not think about them." [@fallows_tragedy_2014].

In addition to its impacts on American foreign policy, other scholars worry that an excessive gap between society and the military harms the quality of civil-military dialogue by constraining civilian input into the policy process, which in turn can likewise threaten a range of foreign policy outcomes. When interacting with civilian leaders in the context of a sizable gap between the military and society, certain military actors may adopt an attitude such that they believe their military service entitles them to be heard in a deferential manner [@beehner_matter_2020]. A similar dynamic can occur if civilian leaders fail to sufficiently assert themselves into the policy process when interacting with military actors who are incredibly popular. Karlin and Friend, for instance, note the dynamic of "military worship" in which military service is somehow viewed as better, more relevant, or more appropriate than other types of civilian service which ultimately harms civilian control of the military and other democratic norms [@friend_military_2018]. As a result, military actors may insist on solving problems in the manner that they want the job done, and not in a way that aligns with the desire of civilian leaders. 

Regardless of the mechanism through which it works, there is a general consensus among scholars that the contemporary United States military is either very close to or at a point in which it is esteemed too highly. While recent polls show that trust in the military may in fact be slipping [@mcintyre_people_2021], the fact is that relative to other institutions, Americans greatly respect their military. According to recent Gallup public opinion polls, 72 percent of Americans indicate that they trust the military a "great deal" or "quite a lot" - the most of any institution, to include Congress (13%), the Presidency (39%), the police (48%), the Supreme Court (40%), and the church/organized religion (42%) [@noauthor_confidence_2020]. Moreover, trends in trust of various American institutions over time reveal that whereas trust in the military has generally increased over the past 45 years, trust in the other institutions have generally declined or remained the same. These trends are displayed in Figure \@ref(fig:insttrust).

```{r code, warning=FALSE, echo=FALSE, include=FALSE}
df_trust <- read_csv(here("data","inst_trust.csv")) %>% print()
df_trust <- rename(df_trust, SupremeCourt="Supreme Court")

df_trust_consolidated <- df_trust %>% 
  select(Year, Military, Congress, SupremeCourt, Church, Newspapers) %>% 
  gather(key="Institution", value="value", -Year) 
```

```{r insttrust, warning=FALSE, echo=FALSE, include=TRUE, fig.height=4, fig.width=6, fig.cap = "Public Trust in American Institutions Over Time"}

df_trust_consolidated %>% ggplot(aes(x=Year, y=value)) + geom_line(aes(color=Institution, color = Institution)) + scale_color_manual(values=c("dodgerblue", "darkgreen", "black", "seagreen2", "salmon3")) + labs(x="Year", y="Percent Surveyed Who Place\n A 'Great Deal' or 'Quite a Lot' of Trust In...", subtitle="Data Available from 'Confidence in Institutions,'\n https://news.gallup.com/poll/1597/Confidence-Institutions.aspx")

```

# How Polarization and Prestige Impact Central Principles of Civil-Military Relations

Having now defined and provided an overview of both the dependent and independent variables, this theory chapter will now connect these variables into a cohesive theory. As Figure \@ref(fig:overview) at the beginning of this chapter shows, the main argument of this dissertation is that polarization and military prestige impact the constraining influence of the three governing standards of American civil-military relations (civilian control of the military, non-partisanship of the military institution, and the non-interference of the military into civilian political affairs). Furthermore, as Figure \@ref(fig:overview) shows, when these three central principles of civil-military relations lose their power to constrain the political behavior of civilian and military actors involving the military institution, we can expect to see increasingly blatant behaviors committed by both civilian and military actors that challenge, if not violate, these three central principles of civil-military relations. In other words, my argument is that the level of polarization within a society and the degree of prestige of the military impact the degree to which the principles of civilian control, non-partisanship, and the non-interference of the military into civilian affairs are considered sacred or inviolable by both military and civilian actors. When these principles are considered sacred, there are less instances of political behavior which challenge or violate these principles; on the other hand, when these principles are thought of as idealistic but altogether not binding, or generally less important than something else, there will be more instances of civilian and military actors behaving politically in ways that challenge or violate these standards. 

Furthermore, in the American context, I argue that the level of political polarization provides a "motive" for military and civilian leaders to commit acts of political behavior that challenge the central standards of civil-military relations, while the level of military prestige provides the "opportunity" to do so [@finer_man_1962]. The application of a "motive" and "opportunities" framework to the contemporary American context follows other works which have done the same to explain military intervention in politics broadly speaking [@finer_man_1962] and in the comparative setting of the Russian military [@taylor_politics_2003]. 

## Polarization - A "Motive" to Challenge the Principles of Civil-Military Relations

I argue that as polarization rises, the degree of _ideological unity_ among military and civilian actors decreases. This occurs because highly polarized climates are marked not merely by strong political disagreements (which are nonetheless present when polarization is high), but also by strong disagreement and division over which moral values and purposes should govern or influence society, and to what extent. As polarization rises, therefore, military and civilian actors increasingly encounter situations in which they must make judgments about whether to abide by the fundamental standards of civil-military relations, or to engage in acts of behavior that, while challenge or violate the central governing principles of civil-military relations, nonetheless advance a particular belief system, set of values, or ideology that the officer or civilian leader supports.

Especially in the case of military officers, the tension that arises when faced with a choice to obey either one's professional responsibilities or the dictates of one's personal conscience is neither a new nor rarely discussed phenomenon. Indeed, recent scholarship has recognized the difficulty military officers face when encountering situations in which they must choose between following one's "professional versus private morality" [@cohn_what_2019; @robinson_dissents_2020].^[For a context of morality in war, see also @dubik_just_2016] As scholars have noted, military officers in particular routinely face competing demands of loyalty. That is, the duties required of professional military officership requires that military officers remain loyal to multiple entities: to their civilian leaders, whose orders they follow; to the public, for whom the military institution exists to ultimately defend; to their fellow troops, with whom they serve; and to themselves [@cohn_what_2019; @robinson_dissents_2020]. 

Because these entities can all desire different things from time to time, there are occasions when choosing to be loyal to one entity comes at the direct exclusion of another. Furthermore, even among scholars of civil-military relations, there is relatively little agreement regarding how officers should choose when encountering situations in which their loyalties are divided, with some asserting that civilians should and must always be followed [@kohn_richard_2010], and others arguing that because officers are moral agents, they should never follow an order which they believe is inherently wrong [@milburn_breaking_2010; @milburn_when_2019]. Moreover, making decisions in these types of situations is not easy under any circumstances, nor are these circumstances confined to eras in which the level of polarization in society is high.

What is new from a theoretical perspective is acknowledgement that relative to less polarized settings, highly polarized environments increase the frequency with which military officers encounter these types of situations, as well as the magnitude of the disparity _between_ the outcomes of adhering to the central principles of civil-military relations versus following one's conscience. While it is not possible to predict all of the possible issue areas or situations which could provide a "motive" for civilian and military actors to challenge the governing principles of civil-military relations, the characteristics of today's highly polarized environment in the United States adequately capture why this is so.  

Some military officers will be willing to violate the principle of civilian control when they deem that the preferred policies held and promulgated by a particular civilian leader are not simply politically unwise, but wrong in a moral sense. In a similar vein, some military officers may be willing to violate the principle of non-partisanship if they believe that one of the major political parties possesses a political agenda that is simply wiser from a political perspective than the other party, but rather one which aligns to the moral views of the officer on a particular issue or sets of issues. Finally, some military officers will be more inclined to violate a standard of "non-interference" of the military into certain areas of state policy if they believe that the military as an institution _should_ play a greater role in advancing the underlying ideology or values associated with a particular issue or set of issues.  

Similar dynamics also impact certain civilian leaders in highly polarized environments. While civilian leaders cannot expressly violate the standard of civilian control, certain civilian leaders may be inclined to challenge the principle of non-partisanship of the military if they believe that the advantages to doing so outweigh the potential risks. Likewise, certain civilian leaders may challenge the standard of non-interference by directing the military to fill a role that falls outside of its traditional military expertise if doing so advances the leader's underlying preferred values, ideology, or policies. Moreover, these expectations of behavior follow recent experimental research that shows that a significant majority of Americans believe it is more important to advance candidates for office with whom they share a partisan or ideological affiliation than it is candidates who abide by democratic norms [@graham_democracy_2020], and that citizens find certain military elites credible conditional on these elites sharing similar ideological and partisan convictions [@robinson_michael_danger_2018]. A common theme from this scholarly research is that support is granted on based on what ideological positions one follows, and not whether actors (civilian or military) follow normative guidelines. 

Among the clearest contemporary examples of the dynamics in which polarization can provide a "motive" for civilian and military actors to engage in political behaviors by altering the constraining influence of the standards of civil-military relations is the issue of race in America. For any observer or participant of American politics in the 2020 or 2021 time frame, it comes as no surprise that race has become a if not _the_ polarizing issue in the United States. 

Some Americans, mainly those who identify or associate with the progressive wing of the Democratic Party, hold fast to a set of ideas and principles that are part of a worldview known as _Critical Race Theory_, a concept borne out of academia in the late 1970s and 1980s that views "race [as] a political construction that was invented by white people to give themselves power while excluding all other races from it..." [@lindsay_what_2021]. Furthermore, proponents of Critical Race Theory see the problem of racism as inherent in the very power structures and societal fabric of the West, and thus, view the key to overcoming racism as reshaping or dismantling the "very foundations of the liberal order, including equality theory, legal reasoning, Enlightenment rationalism, and neutral principles of constitutional law" [@lindsay_what_2021]. Finally, proponents of this worldview, such as the historian Ibram Kendi, who authored the wildly popular 2019 book, _How To Be An Antiracist_, claim that it is not enough to be against racism, but that racism - which is inherently in and part of everything - must be aggressively searched for, found, and eradicated [@lindsay_what_2021]. 

Other Americans, mainly those who identify with the political right, reject many of these tenets. While these Americans may disagree among themselves about the extent to which racism in America is inherently "structural" or "systemic," the majority of these Americans acknowledge and regret American's ugly past with racism, even as they affirm that the United States is and has nonetheless been a relatively noble country that has pursued, however imperfectly, positive ideals and goals that have helped millions of people, both within the US and throughout the world. ^[To be clear, there are undoubtedly staunch racists in the United States who truly believe that certain classes of people are inferior to others on the basis of race. I do not have this group of staunch racists in mind when describing the opponents of Critical Race Theory. I have in mind, rather, everyday Americans, including a large number of African Americans, who hold to the view that Critical Race Theory is a somewhat dangerous ideology and worldview.] Furthermore, these Americans view Critical Race Theory as a totalitarian effort that attempts to solve the problem of racism by instituting a form of racism of its own by elevating the experiences of groups who identify as oppressed at the expense of other groups [@lindsay_eight_2020]. Unlike many proponents of Critical Race Theory, these Americans believe that the names of George Washington, Abraham Lincoln, and the like should remain as examples for all Americans to aspire to, despite any connections the historical individuals may have had with racism [@ho_san_2021].

How might the issue of race in America serve as a motive for military and civilian actors to engage in behaviors that challenge governing standards of civil-military relations? Much of the answer depends on where one stands with respect to the issue of Critical Race Theory in the United States. For example, proponents of Critical Race Theory are likely to point to the lack of proportional minority representation among senior general and flag officers and among key occupations such as fighter pilots as proof of "systemic" or "structural" racism in the US, while opponents of Critical Race Theory are unlikely to conceptualize these realities, despite being genuine challenges that require diligence and commitment to rectify, as eternal weights that prevent the US from reaching its full potential as a nation. 

Furthermore, reflecting the dynamics of "conflict extension" previously discussed, in which polarization tends to increase the scope of issues that comprise the political battlefield [@layman_party_2002], Critical Race Theory and the "anti-racist" ideology that stems from it tends to view the world in very sharp contrast in that a person is either an "anti-racist" or a racist - there is no room for middle or neutral ground. From this perspective, the polarizing issue of race in America is one of many that can provide a motive, even for a normatively non-partisan actor such as the military, to commit behaviors that challenge or violate central principles of civil-military relations. Not only is the ideology of Critical Race Theory incompatible with remaining politically neutral, but military actors are likely among the ideology's supporters and opponents. Actors from both camps may believe that the morally and ethically right thing to do is to support/oppose the ideology, even if doing so threatens principles of civilian control, non-partisanship, and the "non-interference" of the military into certain state affairs. 

## Prestige - An "Opportunity" to Challenge the Principles of Civil-Military Relations

Whereas rising political polarization lessens the degree of ideological unity among military and civilian actors, thus providing a "motive" for the military's intervention into politics, rising military prestige advances the military's ability to constitute a  "latent political influence" [@brooks_militaries_2009, 231], and thus provides an "opportunity" for the military to challenge standards of American civil-military relations [@finer_man_1962; @taylor_politics_2003]. In short, the more popular and prestigious a military becomes, the greater capacity the military possesses to engage in political behaviors that defy normative principles of civil-military relations. It is important to underscore that, for purposes of this theory, increasing military prestige - and holding all else equal - increases only the _potential_ for civilian and military actors to engage in political behaviors that defy principles of civil-military relations - and not necessarily that these types of behaviors will in fact increase. The main reason that rising military prestige leads to the military having a greater _potential_ to challenge the governing standards of civil-military relations lies in the relationship between prestige and the potential to shape public opinion. 

A significant amount of scholarship in the subject area of political communication describes how, why, and the extent to which elites can and do influence public opinion [for a sampling of this scholarship, see @saunders_no_2017; @guisinger_mapping_2017; @baum_relationships_2008; and @kertzer_bottom-up_2017]. While scholars differ on the degree to which elites actually sway public opinion, scholars who claim that elites do in fact possess an outsized ability to impact public opinion note that the public looks for two central attributes among elite actors - "knowledge" and "trustworthiness" - when determining which elites to listen to as reputable sources of information [@lupia_democratic_1998, 69-76]. Other scholars claim that the more elites are perceived by the public to hold these attributes in abundance, the greater their ability to "cue" or shape public opinion [@zaller_nature_1992]. 

Furthermore, multiple episodes in American history show that popular military figures have influenced public opinion in significant ways. Ulysses S. Grant and Dwight D. Eisenhower, for instance, were both wartime commanders who, largely as a result of the popularity they commanded with the public, ascended to the Presidency [@teigen_why_2018]. In short, a large body of scholarship shows that the more respected an elite is, especially military elites, the greater _capacity_ or _potential_ he or she poses to move public opinion. 

Even currently-serving high-ranking military officers, who are one class of military "elites", acknowledge the relationship between prestige and public opinion, although they do so, understandably, by eschewing the term "military prestige," and instead use the term "trust" [@odierno_repairing_2014]. These senior military officers understand that without a certain level of prestige (positive reputation among the public), the military's ability to effectively accomplish its mission drops. For instance, when speaking to a group of Army officers, presumably a gathering of Colonels who were scheduled to soon be promoted to the rank of Brigadier General, the former Chief of Staff of the Army, Ray Odierno, made several comments that tied the Army's ability to accomplish its mission with maintaining the public's trust. He stated, for instance, that, "Of great importance within the Army Profession is the duty of each of us to be a steward of our profession and to maintain its essential and hard won trust with society"[@odierno_repairing_2014]. Odierno continued, _"our ability to fulfill our mission depends on trust"_ [@odierno_repairing_2014]. 

The overall level of prestige held by a military dictates the degree to which it is both willing and able to challenge the fundamental principles of civil-military relations. For example, when Colin Powell wrote opinion commentary that challenged Candidate and later, President, Clinton's views on using military force in the Balkans in 1992 and 1993 [@powell_opinion_1992; @powell_us_1992], actions which challenged if not violated the fundamental principle of civilian control of the military, one cannot ignore that Powell's ability to do so stemmed in part from his immense public popularity, or prestige. Having been perceived by the American public as playing a critical role in the US victory during the Gulf War of 1991, Colin Powell indeed posed a considerable "latent political influence"[@brooks_militaries_2009], which he used repeatedly as Bill Clinton took office. Madeleine Albright, Clinton's Ambassador to the United Nations and later, Secretary of State, noted Powell's remarkably high potential to influence policy during Clinton's early for Powell to influence policy. Albright, in a 2006 interview, for instance, noted that, "On a regular basis Colin would come in and do a presentation...he’d go through this and say, 'We can take that hill and we can do that and we can do this. You know we have the best military in the world, but it’s going to take 500,000 men and $500 billion and 50 years.' So he’d lead you up the hill of possibilities and then drop you off the other side, and you’d end up with no options" [@albright_william_2006]. Albright, later in the interview, remarked that although she and Powell were and are close friends, during their time as colleagues under Clinton, there is one question that she wishes that Powell would have answered: "what are you saving your political capital for?" [@albright_william_2006].

The example of Colin Powell constraining options for the Balkans in the early 1990s illustrates how a prestigious military can, and did, challenge the fundamental principle of civilian control of the military. However, there are also examples when civilian actors can harness the military's prestige and similarly challenge the standards or principles of civil-military relations. George W. Bush's use of General David Petraeus to "sell" the Iraq Surge to Congress and the American public [@coll_generals_2008], Donald Trump's deployment of the military to the US southern border ahead of the 2018 midterm elections [@adams_trumps_2018], and a growing number of political candidates who have used the military in campaign advertisements [@moore_biden_2020] all reflect instances in which civilians deliberately harnessed the popularity of the military to sway public opinion, in one form or another. Whether these instances, and to what degree, constitute normatively inappropriate actions that challenge or violate fundamental principles of civil-military relations, and especially the military's non-partisan ethic, remain open debates. In any event, it is clear that a popular and prestigious military that is well liked by the public is often sought to shape opinion, either by civilian actors seeking to win office or implement a desired policy, and/or by military actors who think that they have a certain clout with the public.  

Additionally, _whether_ a prestigious military or military figures are effective and actually do influence public opinion is not a cut and dry issue. Recent civil-military relations studies, such as one that uses survey experiments to examine whether endorsements of presidential candidates by military leaders actually changes public opinion, indicates that they are not [@golby_listening_2013]. In this dissertation, I do not primarily consider the efficacy of these efforts. Rather, for purposes of the theory used throughout this dissertation, it is clear that rising military prestige generally leads to an increase in the _potential_ for the military to sway the public. Thus, rising military prestige provides an "opportunity" for civilian and military actors to engage in behavior that challenge fundamental principles of civil-military relations [@finer_man_1962].  

## Polarization and Prestige Lead to Four Types of Domestic Structures

When we consider the impact of polarization and prestige in tandem, four possible domestic structures or environments emerge, each of which corresponds to a relatively high or low degree of polarization and military prestige. These four structures are displayed in Table \@ref(tab:poss-environment). 

```{r poss-environment, echo=FALSE, include=TRUE, error=FALSE, message=FALSE}

text_env <- tribble(~"Environment", ~"Low Polarization", ~"High Polarization",
  "Low Prestige", "Little Motive or Opportunity to Challenge Principles of Civil-Military Relations. Civilian and Military Actors are Relatively United Ideologically; Political Behaviors Involving the Military Institution are Relatively Private", 
  "There exists Motive, but little Opportunity to Challenge Principles of Civil-Military Relations. Civilian and Military Actors are Relatively Divided Ideologically; Behaviors Involving the Military Institution are Relatively Private",  
  "High Prestige", "Little Motive, but Opportunity to Challenge Principles of Civil-Military Relations. Civilian and Military Actors are Relatively United Ideologically; Behaviors Involving the Military Institution are Public and Visible", 
  "Motive and Opportunity Exists to Challenge Principles/Standards of Civil-Military Relations. Civilian and Military Actors are Relatively Divided Ideologically; Behaviors Involving the Military Institution are Public and Visible" 
  )
 

  kable(text_env, booktabs=T, caption="Types of Domestic Environments for Civil-Military Relations and their Impact on Political Behaviors") %>%
  kable_styling(full_width = F, font_size=10, latex_options="striped") %>% 
  column_spec(1, width="10em") %>% 
  column_spec(2:3, width="20em") 
```

## Low Polarization, Low Military Prestige
Under conditions when polarization and military prestige are both relatively low, military actors possess neither a "motive" nor "opportunity" to intervene in politics [@finer_man_1962]. Low polarization levels indicate there is little if any ideological substance driving military actors to challenge foundational principles of civil-military relations, and low prestige suggests that  the military lacks the ability to leverage public opinion. This leads to the theoretical expectation that while there might certainly be conflict between military and civilian leaders over isolated, context-driven events, there will likely not be widespread and substantive instances of political behavior that challenge or violate the standards or principles of civil-military relations. Historical periods which pose as examples of this domestic structure include the post-Vietnam and pre-World War two eras.

## High Polarization, Low Military Prestige
When polarization is high but military prestige is low, military and civilian actors may in fact be divided ideologically. Some of these actors likely possess a true "motive" to advance particular ideologies, even if doing so involves conducting behaviors that challenge, threaten or violate the governing standards of civil-military relations [@finer_man_1962]. However, as a result of the low level of military prestige that exists in this particular domestic structure, civilian and/or military actors lack the "opportunity" to significantly involve the military institution in ways that challenge the central principles of civil-military relations. This leads to the theoretical expectation that while there might be some instances of civilian or military actors engaging in less-visible forms of behavior that challenge the foundational principles of civil-military relations, these efforts will stop short of attempting to harness the military to sway public opinion. Historical examples of this era include the Gilded Age through the beginning of American involvement in World War One, or approximately 1875-1915.

## Low Polarization, High Military Prestige
Under conditions of low polarization but relatively high military prestige, military and civilians actors are relatively united ideologically, and therefore do not possess a strong ideological "motive" to engage in political behaviors that challenge or threaten the principles of civil-military relations [@finer_man_1962]. However, the "opportunity" to use the military institution to leverage public opinion exists [@finer_man_1962]. Under these conditions, it is possible that military and/or civilian actors will engage in forms of political behaviors that harness or leverage the military to move public opinion. This leads to the theoretical expectation that under these domestic conditions, military actors in particular pose a considerable "latent political influence" because they are popular with the public [@brooks_militaries_2009, 231]. The military under these conditions may pose a considerable force to block, obstruct, or influence debates about the use of military force and national security more broadly.  Examples of this period include the post-World War Two until the Korean War, and the post-Gulf War period (1991) until approximately 2010.

## High Polarization, High Military Prestige
When both polarization and military prestige exist at relatively high levels, military and civilian actors possess both a "motive" and an "opportunity" to use the military in ways that challenge, threaten, or violate the fundamental principles of civil-military relations [@finer_man_1962]. Under these conditions, there is a greater likelihood that certain civilian and/or military actors believe so strongly in particular political issues or sets of political issues that they are willing to push the boundaries of the principles of civilian control, non-partisanship, and "non-interference" of the military by engaging in behaviors activities that do so than when polarization is low. Moreover, as a result of the military's relatively high degree of prestige, military and civilian actors will seek to use the military in public and visible ways, as they deem the military institution capable of impacting public opinion. Examples of this period include the post-US Civil War period during American Reconstruction, 1865-1870, and from approximately 2010 - present day. 

# Hypotheses to be Tested

Based on the theoretical predictions that follow from Table \@ref(tab:poss-environment), it is possible to derive the following six hypotheses: 

H1: Military actors are increasingly active in a partisan manner when polarization is high relative to periods of low polarization. 

H2: Military elites increasingly challenge civilian authorities in a visible, public manner during periods of high military prestige relative to periods of low military prestige. 

H3: When military prestige is high but political polarization is low, military actors are increasingly likely to challenge civilians on use of force initiatives, military policy, and other national security issues rather than political issues that are unrelated to these issue areas.

H4: When military prestige is high, civilian leaders will increasingly attempt to leverage or harness the military (either as an institution, or individual military figures who possess outsized levels of prestige/popularity) for their own political purposes and advantages relative to when military prestige is low. 

H5: When polarization is high, civilian leaders will employ the military (either as an institution, or individual military figures who possess outsized levels of prestige/popularity) to support (or denounce) policy initiatives that align with (or oppose) underlying ideologies associated with a polarizing issue or cause relative to when polarization is low. 

H6: When polarization and military prestige are high, civilian political parties will increasingly compete for the military institution and in increasingly public ways relative to when polarization and military prestige are low. 

## Summary

This chapter argued that the two independent variables of political polarization and military prestige combine to influence the degree to which civilian and military actors are willing to adhere to the fundamental principles or standards of civil-military relations. In doing so, this chapter argued that a high level of polarization can provide a "motive" for civilian and military actors to behave in ways that challenge or violate standards of civil-military behavior [@finer_man_1962]. This is because, this chapter argued, increasing political polarization reflects disagreement in the populace not only about specific policy matters, but about the values and worldviews that should govern society, to what extent, and why. When polarization is high, civilian and military actors are increasingly willing to challenge civilian leaders, adopt positions that align with partisan entities, and appeal to the military to weigh in on or solve issue areas that are indirectly related to topics that are traditionally associated with the military. 

This chapter also argued that high levels of military prestige can serve as an "opportunity" for the military to engage in politics in ways that threaten or violate principles of civil-military relations [@finer_man_1962]. High levels of military prestige suggest that the military can impact public opinion, and that both civilian and military actors are cognizant of this reality. When the prestige of the military is high, the military as an institution is at a greater risk of featuring in more visible and public ways than when the level of military prestige is low. 

Finally, this chapter argued that both variables need to be considered simultaneously in order to determine if and how military and civilian actors will challenge the fundamental principles of civil-military relations. Because each variable can theoretically take on a relatively high or low value, a total of four structures or environments are possible. The types, forms, and overall pattern of political behaviors that each type of actor, civilian and military, are likely to commit involving the military institution, correspond to one or more of these environments.  

The next four chapters of the dissertation test the predictive power of the theory laid out thus far. The first two chapters each explore one "side" of the civil-military relationship - not in isolation from the other, but in terms of a distinct behavior that each actor can commit. The next chapter focuses on retired military officer and the behavior of public opinion commentary, while the subsequent chapter focuses on civilian candidates, interest groups, and political parties and the behavior of campaign advertisements.

Following these two "large-N" chapters, each of which incorporates mainly a mainly quantitative analysis, the dissertation then conducts two separate case studies. In each of these studies, I trace the impacts of polarization and military prestige on the form and pattern of the distinct behaviors committed by civilian and military actors that involve the military institution.

\pagebreak
\singlespace
\newpage


