---
title: "Servants, Syndicates, and Saviors: How Polarization and Prestige Impact the Political Behavior of the Military (Prospectus)"
author: "Peter M. Erickson"
date: |
  `r format(Sys.time(), '%B %d, %Y')`
abstract: |
  
  _This dissertation explores variation in the political behavior of the US military. I advance a theory in which the political behavior of the US military is impacted by two distinct factors: the level of domestic political polarization within a society, and the degree of prestige the military institution commands. I argue that these two factors greatly shape the constraining influence of three central norms of civil-military relations: civilian control of the military, the non-partisanship of the military institution, and the degree to which the military fills an active role in domestic politics. As the constraining influence of these norms varies, distinct political behaviors conducted by or involving the military also change. To support my theoretical argument, I conduct a mixed-methods empirical strategy focusing on military behavior within the United States across time. This dissertation advances the scholarly understanding of US civil-military relations, and provides a unique perspective about how domestic structure influences the ability of the military to act professionally._  
  
geometry: margin = 1.15in              # margin widths
fontsize: 12pt                         # text size
bibliography: dissertation.bib         # bibliography file
biblio-style: authoryear               # citation style
urlcolor: blue                         # links to external URLs
citecolor: black                       # links to bibliography
linkcolor: magenta                     # internal figures, footnotes, etc.
indent: true                          # paragraph format
                                       # true for indented paragraphs
                                       
header-includes:
    - \usepackage{setspace}
    - \usepackage{multirow}
    - \usepackage{indentfirst}
   
    
output: 
  bookdown::pdf_document2:
    keep_tex: true
    latex_engine: pdflatex
    number_sections: false
    toc: false
    fig_caption: true
    citation_package: biblatex
    includes: 
      in_header: 
        - latex-preamble.tex
  bookdown::htmldocument2: default
  bookdown::word_document2:
      toc: false
---

\doublespacing 

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo=FALSE,
  include=FALSE,
  message=FALSE,
  error=FALSE,
  fig.height=3,
  fig.width=4,
  out.width="90%")
```

```{r, include = FALSE, echo = FALSE}
library("here")       # file paths
library("tidyverse")  # workhorse package
library("tidylog")
library("kableExtra")
library("knitr")
library("ggdag")
library("dagitty")
library("gridExtra")
library("broom")
library("dplyr")
library("reshape2")
library("pacman")
library("ggrepel")
```

\newpage

# Introduction

In this chapter, attention is devoted to developing a comprehensive theory that explains how political behaviors involving the military institution are impacted by changes in the degrees of domestic polarization and the prestige of the military within a society. The previous chapter asserted that the vibrant and relatively healthy civil-military relations that the United States has enjoyed throughout most of the nation's history is the result of civilian and military actors remaining willing to abide by three central norms: civilian control of the military, non-partisanship of the military institution, and the "non-interference" of the military into certain arenas and policy realms of state policy. The core argument of the theory developed in this chapter is that changes in the levels of domestic polarization and the prestige of the military ultimately impact the _constraining influence_ of these norms on military and civilian actors. That is, levels of polarization and military prestige influence the willingness of military and civilian actors to adhere to norms and standards of civil-military relations, which influences the political behaviors that civilian and military actors commit involving the military as an institution. 

This chapter proceeds in three parts. First, I define the relevant terms that are used throughout this study. Having already established a basic definition and understanding of the origin and meaning of the three foundational principles of American civil-military relations in the previous chapter, I define and trace the developments of the dependent variable, _political behaviors_ which involve the military institution, and independent variables, _political polarization_ and _military prestige_, used in this study. In the second part of this chapter, I explain how the levels of polarization and military prestige impact the constraining influence that the governing principles of civil-military relations have on the propensity of military and civilian actors to engage in various political behaviors that involve the military institution. This second part of the chapter concludes by arguing that a total of four domestic political structures, or environments, are possible, each of which is marked by the presence of relatively low versus high levels of polarization and military prestige, and that each environment is likely to produce certain patterns of political behavior commited by civilian and military actors involving the military institution. The third and final part of the chapter then presents a series of hypotheses to be tested, analyzed, and explored throughout the remainder of the dissertation.  

# Components of the Theory

A visual representation of the overall theory explained in this chapter and throughout the dissertation is shown below in Figure \@ref(fig:theory-overview-fig) . From left to right, the two independent variables of political polarization and military prestige combine to influence an intervening variable, the ability of fundamental norms of civil-military relations to constrain civilian and military actors from the dependent variable, engaging in certain types and forms of political behaviors. 

```{r theory-shell, include=FALSE, echo=FALSE}
problem_dag <- dagify(
    PB ~ CI,
    CI ~ MP + PP,
    coords = tribble(
      ~ name,      ~ x,    ~ y,
      "PB", 1,.5,
      "CI", .7, .5,
      "PP", .3, .7,
      "MP", .3, .3,
    ))
    

problem_dag %>% 
  ggplot(aes(x=x, y=y, xend=xend, yend=yend))+
  geom_dag_edges()+
  geom_dag_point()+
  geom_dag_text_repel(aes(label=name), show.legend=FALSE)+
  theme_dag()
```

```{r theory-overview-fig, include=TRUE, echo=FALSE, fig.height=4, fig.width=5 }

problem_dag %>% tidy_dagitty() %>% 
  dag_label(labels=c("PB"="DV\n Political\n Behaviors",
                     "CI"="Int Var\n Constraining Influence\n of Foundational\n Civil-Military Norms", 
                     "MP"="IV\n Military\ Prestige", 
                     "PP"="IV\n Political\n Polarization"
                      )) %>% 
  ggplot(aes(x=x, y=y, xend=xend, yend=yend))+
  geom_dag_edges()+
  geom_dag_point()+
  geom_dag_text()+ 
  geom_dag_label_repel(aes(label=label, fill=label), col="white", show.legend=FALSE)+ theme_dag(plot.title=element_text(hjust=.5)) + labs(title="Determinants of Political Behavior by the Military")
  
```

## The DV: Political Behaviors

### Military Actors

The concept of political behaviors which involve the military institution must be carefully defined, particularly when examining both military _and_ civilian actors. After all, elected civilian leaders are inherently political to begin with, and thus, everything they say and do should, at least to some extent, be seen as "political." Therefore, the normative standards exposited in the previous chapter will serve as an appropriate point of departure or reference point when discussing political behaviors which involve the military institution, that both sides, military and civilian, commit. From the perspective of military actors, political behaviors are those behaviors that military leaders engage in for the purposes of "influencing" policy outcomes [@brooks_militaries_2009, 213]. There are a number of specific behaviors that military leaders can engage in which might in fact influence policy outcomes. One helpful framework of these behaviors, developed by Risa Brooks (2009), is displayed below in Table \@ref(tab:pol-behav-tab) .  

\singlespace

```{r pol-behav-tab, echo=FALSE, include=TRUE, error=FALSE, message=FALSE}

text_tbl1 <- tribble(~"Tactic", ~"Nature of Activity", ~"Audience", ~"Examples",
  "Public Appeal", "Public, value-laden commentary", "Mass public",  "Colin Powell's 1992 statements on intervention in Bosnia", "Grandstanding", "Threat or actual resignation in protest of policy", "Mass public, Congress, executive branch", "Hypothetical; if Eric Shinseki had resigned in 2003",
  "Politicking",  "Retired officer endorsements; organizing vote drives", "Mass public", "Crowe's endorsement of Clinton in 1992; Frank's endorsement of Bush in 2004", 
  "Alliance Building",  "Military leaders form ties with civilian interest groups", "Congress", "Air Force and Lockheed Martin over procurement of F-22 raptor",
  "Shoulder Tapping", "Military leaders setting an agenda by bringing issues to attention of politicians and engaging in lobbying-like activities on behalf of those issues", "Congress", "Military Mobilization of key members of Congress over gays in the military in early 1990s")
 

text_tbl1 %>% kable(booktabs=T, caption="Political Activities of the Military") %>%
  kable_styling(full_width = F, font_size=8, latex_options="striped" ) %>% 
  column_spec(1, width="10em") %>% 
  column_spec(2, width="15em") %>% 
  column_spec(3, width="10em") %>% 
  column_spec(4, width="15em") %>% 
  add_footnote(label = "The above table is reproduced from Brooks, Risa (2009), 'Political Activities of the Military' in\n American Civil-Military Relations: The Soldier and the State in a New Era, Eds. Suzanne Nielsen and Don Snider.")

```
\doublespace

Determining which behaviors listed in Table @\ref(tab:pol-behav) are normatively problematic - and when - is a difficult if not impossible endeavor. As Brooks (2009) notes, rendering a normative assessment regarding political activities committed by military officers depends on specific contexts and circumstances, as certain aspects of engaging in these behaviors and activities can also yield positive repercussions [@brooks_militaries_2009]. For instance, when military actors offer commentary to the public, the information they provide can help citizens better understand and process information that is often complex and disconnected from everyday life, particularly if the issue is somewhat technical or complex [@brooks_militaries_2009, 225-231; see also @schultz_democratic_1999]. Furthermore, militaries can alert Congress to potentially destructive decisions an executive may be considering, thus helping the nation avoid undertaking a course of action that will lead to disastrous consequences [@brooks_militaries_2009, 225-231]. While the final decisions ultimately belong to civilian leaders to make, there are nonetheless real and positive benefits of military actors engaging in the political process in overt ways. 

As Brooks notes, however, any benefits of military actors commiting political behaviors in the short-term may be outweighed by more deleterious long-term impacts [@brooks_militaries_2009]. When military actors engaging in political activities, they generate  _political costs_ for civilian leaders. These costs vary, but include constraining the options available to civilian leaders, sparking public backlash, and otherwise influencing debate about an impending policy choice away from the true preferences of civilian leaders [@brooks_militaries_2009]. Therefore, and as Brooks (2009) notes, military actors who engage in political activities do so while facing four unique risks. 

A first risk of military actors engaging in political activities is to society's esteem of the military [@brooks_militaries_2009, 234]. Part of the reason the military has maintained its popularity with American society is precisely because it has been seen and viewed as a professional institution that does not routinely engage in inappropriate political activities. A second risk of military leaders committing politically inappropriate acts is that doing so provides civilian leaders with an estimate of a particular military leader's political views, which could jeopardize how civilian leaders consult with that particular but also other military actors [@brooks_militaries_2009, 234]. A third risk which military actors face when engaging in political activities is to the tenets of "military professionalism" laid out by Huntington directly: military actors engaging in political activities renders them less focused on their professional duties of securing the Republic [@brooks_militaries_2009, 235; @huntington_soldier_1957]. A fourth and final risk, one which Brooks argues is the most important to consider, of military actors engaging in appropriate political expression is that doing so could "undermine" civilian leaders and ultimately, the governing principle of civilian control [@brooks_militaries_2009, 235]. Ultimately, Brooks (2009) concludes that the potential benefits of military actors engaging in political activities must be weighed against the potential consequences, and recommends that for the sake of long-term, stable civil-military relations, military actors rarely engage in these and other political activities [@brooks_militaries_2009].

Two other implications are worth noting regarding the activities listed in Table \@ref(tab:pol-behav). First, not _all_ of these listed behaviors are of the qualitative nature such that they generate significant political costs for civilian leaders in every instance. For example, military officers who engage in making "public appeals" can do so, theoretically, without inducing significant political costs. For example, when General Colin Powell wrote a series of op-eds in the early 1990s while serving as the Chairman of the Joint Chiefs of Staff in which he articulated his views regarding the use of force, his behavior undoubtedly generated political costs for the candidate, and later, President, Bill Clinton [@powell_opinion_1992; @powell_us_1992]. Engagement in this behavior by a highly respected and prestigious figure likely influenced public opinion and constrained the options available to Clinton as a number of crises erupted across the world in the Balkans, Somalia, Haiti, and Rwanda. Similar levels of costs are not likely generated every time a retired General or Admiral pens an op-ed, or gives an interview to the press. 

Conversely, every instance of a military officer publicly endorsing a political candidate is an inherently political act that by definition, violates the governing principle of non-partisanship, and creates at least some political costs for civilian leaders (i.e., the candidate not endorsed). Regardless of whether the endorsement of a political candidate made by a military actor is effective in persuading the public to change its vote choice [see @golby_military_2012], the endorsement it is impossible for a military officer to make a nonpartisan endorsement of a political candidate. Other behaviors, such as resignation, have sparked considerable debate among scholars of civil-military relations. Some have argued that the public nature of resignation constitutes an inherently political act that should not be practically considered [@golby_beyond_2015], while others have left room for the possibility that in extreme circumstances, resignation is appropriate, regardless of the potential public fallout that may be generated [@dubik_principled_2014; @snider_strategic_2014].

The second implication of Table \@ref(tab:pol-behav) is simply that military actors have a variety of activities at their disposal from which to choose to generate political costs for civilian leaders. The well known collisions that have occurred between the President and various wartime Commanders throughout American history - Lincoln and McClellan, Truman and MacArthur, Obama and McChrystal - perhaps threaten civil-military relations in the most recognizable and public way, but there are other subtle and nuanced means available to the military for creating costs intended for civilian leaders. 

## Behaviors Committed by Civilians

Civilian leaders can also engage in behaviors that challenge the governing principles previously exposited, though in ways that are different from military actors. Indeed, several of the activities listed in Table \@ref(tab:pol-behav) involve varying degrees of civilian actor acquiescence. Take, for example, the behavior of military actors endorsing political candidates for office. Most instances of this behavior constitute a two way street in which a candidate asks a military officer to make an endorsement, and a military officer agrees to do so. For this reason, in the aftermath of the Presidential Campaign of 2016, for instance, Retired Army General Martin Dempsey, who previously served as the Chairman of the Joint Chiefs of Staff, published an opinion-editorial piece in _The Washington Post_ which he criticized retired Army Lieutenant General Mike Flynn and Marine Corps General John Allen for participating in the respective Republican and Democratic conventions. In the op-ed, Dempsey wrote in part, "As generals, they have an obligation to uphold our apolitical traditions. They have just made the task of their successors – who continue to serve in uniform and are accountable for our security - more complicated. It was a mistake for them to participate as they did. It was a mistake for our presidential candidates _to ask them to do so_" [@dempsey_military_2016; italics mine]. 

Whereas the political activities committed by military actors are considered problematic in that they generate political costs for civilian leaders, inappropriate activities at the hands of civilian actors which involve the military institution are those that "politicize" the military. More precisely, the behaviors commited by civilian leaders which "politicize" the military are those that cause the military to either breach or be perceived by the public as breaching the principles of non-partisanship or non-interference (by definition, it is not possible, I argue, for civilian actors to violate the principle of civilian control). 

Civilian leaders have faced consequences for "politicizing" the military previously. For example, some observers alleged that President George W. Bush's use of General David Petraeus in garnering Congressional support for the "surge" strategy in Iraq politicized the military in unhealthy ways [@coll_generals_2008], while other observers note the increasing of retired military generals and admirals who endorse political candidates at conventions [@exum_dangerous_2017] as troubling indicators of the military's politicization. Moreover, under President Donald Trump's tenure in office, charges of politicizing the military were common, perhaps none as vociferous as the reaction to the former President's decision 2018 decision to deploy several thousand troops to the US border with Mexico in response to an impending caravan of migrants from South America. The headline of an op-ed published by several retired Army Colonels in response aptly summarizes the feelings held by many Americans at the time: "Trump's Border Stunt is a Profound Betrayal of Our Military" [@adams_trumps_2018]. 

In summary, military and civilian leaders can engage in certain behaviors which violate the fundamental governing principles of civil-military relations. Military actors have a variety of behaviors at their disposal which, if committed, can generate substantial political costs for civilian leaders. These costs include constraining the policy choices that are available to civilian leaders, generating negative public opinion against civilian leadership, and otherwise presenting options that fail to satisfy the true preferences of civilians [@brooks_militaries_2009]. Alternately, civilians too can engage in political behaviors which politicize the military, leading to an actual or perceived breach of the military institution's non-partisan ethic or a violation of a norm of non-interference by asking or directing the military to do a task that is indirectly or altogether unrelated to the defense of the nation. 

Having described the dependent variable of study, the political behaviors that can be committed by military and civilian actors which jeopardize the governing principles of civil-military relations, the next task is to describe each of the independent variables: political polarization and military prestige. After defining each of the variables, I frame the respective levels of each in contemporary America. 

# IV1 - Political Polarization

## What is Polarization?

The first independent variable of study in this dissertation is political polarization. Polarization is a difficult concept to precisely define. From the perspective of the American politics sub-field, polarization refers to the movement of political thoughts and attitudes away from the political center and towards more ideological extremes [@hare_polarization_2014; @abramowitz_great_2018]. In recent years, the concept of polarization has also risen within the subfield of international relations, particulary as a domestic political feature that impacts various international outcomes, including whether a polarized country can unite to recognize common threats, learn and derive lessons from foreign policy failure, and agree on what role a nation should play on the world stage [@schultz_perils_2017; @myrick_external_2020]. From a broad perspective, one that is more often explored and discussed within the sub-field of political theory, political polarization describes the degree to which a society shares fundamental values, is capable of incorporating divergent worldviews, and the overall degree to which these values unite versus divide a nation [@marsden_twilight_2014; @deneen_why_2018]. In this dissertation, I rely on both the relatively narrower conception of polarization as the divergence of political attitudes away from an ideological center and towards the extreme, as well as a broader conception of the term that recognizes the difficulties which arise when a society can no longer agree upon the fundamental values and worldviews which should govern it. 

Perspectives regarding the _causes_ of political polarization vary. Americanists tend to explain polarization as stemming from differences of opinion over issues pertaining to race and religion [@abramowitz_great_2018], the size and role of government [@sides_two_2015], normative conceptions of personal and national identity [@sides_identity_2018], and other policy issues [@mann_its_2012]. Political theorists and cultural historians, on the other hand, recognize the current polarizing environment as a symptom of a slow yet steady political shift that began in the 1950s, a shift that reflects both the incompatability of American pluralism to reconcile and "embrace a wide range of both religious and nonreligious viewpoints" [@marsden_twilight_2014, 171] and the inability of liberalism as a political philosophy to deliver on its intended promises of securing individual rights in a manner that is equitable and fair [@deneen_why_2018]. Both explanations of polarization acknowledge that American society is strained because there is no longer a common vision for the values and purposes that should guide and unite society, and that the various perspectives regarding the values that should perform this immensely important task are inherently at odds with each other.   

Scholars note the presence of three characteristics or features that desribe the current polarized political landscape. The first feature is that polarization in the US goes far beyond mere political differences between political elites [@mann_its_2012]. In other words, today's polarized landscape extends far beyond the halls of Congress in Washington D.C. While one can argue that everyday citizens may not feel as strongly as elected politicians about every single political issue at stake today, the fact is that the level of disdain held by regular Americans for those who sit on the opposite side of the political spectrum has risen sharply over the past several decades. This phenomenon is known in the American politics literature in particular as "affective polarization" [@iyengar_affect_2012; @iyengar_origins_2019] or "negative partisanship" [@abramowitz_great_2018]. 

A second feature of contemporary polarized politics in the United States is the increasing scope and number of issues that have become included as part of the partisan battleground. Politicians and citizens do not disagree and wage political war over a handful of important issues, but rather, at least seemingly, over everything! Scholars of American politics describe this dynamic of increasing political warfare as "conflict extension" [@layman_party_2002]. As one might expect, the dynamic of "conflict extension," in which every political issue seems to be fought over bitterly, dampens the prospects for bipartisanship and meaningful cooperation across partisan aisles [@layman_activists_2010]. 

A third and final feature of contemporary political polarization in the United States is that issues of identity are increasingly associated with partisanship - to the point that partisan identity itself is an important component of how people view themselves and others [@sides_identity_2018; @robinson_michael_danger_2018]. This feature of polarized politics is perhaps most clearly illustrated by the rise and influence of American "identity politics" in which the normative roles of race, gender, religion, and immigration invoke very strong commitments from individuals on both sides of the partisan aisle [@white_values_2002; @mann_its_2012; @sides_identity_2018]. The prominence of these identity-related issues in politics affirms that these issues are not merely peripheral interests, but rather issues that individual citizens care deeply about and consider to be of fundamental concern. 

## How Bad is Contemporary Polarization?

The most widespread and historical (though certainly, not perfect) measurement of polarization has been that of dynamically-weighted, three step estimation scores (DW-NOMINATE), which are derived from examining roll call voting patterns among members of Congress [@jeffrey_b_lewis_voteview_2020]. Figure \@ref(fig:polar-measurement) shows polarization levels present in the US House and Senate since the latter portions of the 19th Century [@hare_polarization_2014; @jeffrey_b_lewis_voteview_2020]. As Figure \@ref(fig:polar-measurement) demonstrates, polarization levels were high at the end of the American Civil War, and remained high before gradually decreasing early in the 20th century. The figure further shows that polarization reached it's nadir in the years before World War Two, and remained at relatively low levels until approximately the mid-1970s. Since that time, however, polarization in both chambers of the US Congress has increased dramatically, to the point that contemporary levels exceed those previously recorded.  

```{r polar-measurement, include=FALSE, echo=FALSE}
nom_dat <- read_csv("https://voteview.com/static/data/out/members/HSall_members.csv")

south <- c(40:49,51,53)
polar_dat <- nom_dat %>% 
    filter(congress>45 & 
           chamber != "President") %>%
    mutate( 
      year = 2*(congress-1) + 1789,
    ) %>%
    group_by(chamber,congress,year) %>% 
    summarize(
      party.mean.diff.d1 = mean(nominate_dim1[party_code==200],na.rm=T) - 
                           mean(nominate_dim1[party_code==100],na.rm=T),
      prop.moderate.d1 = mean(abs(nominate_dim1)<0.25,na.rm=T),
      prop.moderate.dem.d1 = mean(abs(nominate_dim1[party_code==100])<0.25,na.rm=T),
      prop.moderate.rep.d1 = mean(abs(nominate_dim1[party_code==200])<0.25,na.rm=T),
      overlap = (sum(nominate_dim1[party_code==200] <
                       max(nominate_dim1[party_code==100],na.rm=T),na.rm=T)  +
                 sum(nominate_dim1[party_code==100] >
                       min(nominate_dim1[party_code==200],na.rm=T),na.rm=T))/
                 (sum(!is.na(nominate_dim1[party_code==100]))+
                  sum(!is.na(nominate_dim1[party_code==200]))),
      chamber.mean.d1 = mean(nominate_dim1,na.rm=T),
      chamber.mean.d2 = mean(nominate_dim2,na.rm=T),
      dem.mean.d1 = mean(nominate_dim1[party_code==100],na.rm=T),
      dem.mean.d2 = mean(nominate_dim2[party_code==100],na.rm=T),
      rep.mean.d1 = mean(nominate_dim1[party_code==200],na.rm=T),
      rep.mean.d2 = mean(nominate_dim2[party_code==200],na.rm=T),
      north.rep.mean.d1 = mean(nominate_dim1[party_code==200 & 
                                             !(state_icpsr %in% south)],na.rm=T),    
      north.rep.mean.d2 = mean(nominate_dim2[party_code==200 & 
                                             !(state_icpsr %in% south)],na.rm=T),    
      south.rep.mean.d1 = mean(nominate_dim1[party_code==200 & 
                                              (state_icpsr %in% south)],na.rm=T),    
      south.rep.mean.d2 = mean(nominate_dim2[party_code==200 & 
                                             (state_icpsr %in% south)],na.rm=T),    
      north.dem.mean.d1 = mean(nominate_dim1[party_code==100 & 
                                              !(state_icpsr %in% south)],na.rm=T),    
      north.dem.mean.d2 = mean(nominate_dim2[party_code==100 & 
                                              !(state_icpsr %in% south)],na.rm=T),    
      south.dem.mean.d1 = mean(nominate_dim1[party_code==100 & 
                                              (state_icpsr %in% south)],na.rm=T),    
      south.dem.mean.d2 = mean(nominate_dim2[party_code==100 & 
                                              (state_icpsr %in% south)],na.rm=T),    
    ) 

head(polar_dat)
write_csv(polar_dat,path="voteview_polarization_data.csv")

```


```{r include=FALSE, echo=FALSE}
# Here are up to date versions of the famous party means over time plots for each chamber:
polar_dat_long <- polar_dat %>% gather(score,value,-chamber,-year,-congress)
labels <- c("dem.mean.d1"="DEM",
            "rep.mean.d1"="REP",
            "north.dem.mean.d1"="N. DEM",
            "south.dem.mean.d1"="S. DEM")

polarized_plot <- function(chamb) {
  pdatl <- polar_dat_long %>% 
                filter(chamber==chamb,
                       score %in% c("dem.mean.d1","rep.mean.d1",
                                    "north.dem.mean.d1","south.dem.mean.d1")) %>%
                mutate(party=labels[score]) %>%
                ungroup()
  
  gg <- ggplot(data=pdatl,
               aes(x=year,y=value,group=party,col=party)) +
               scale_x_continuous(expand = c(0.15, 0), 
                                 breaks=seq(1880, max(pdatl$year), by=8)) +
               geom_line() + geom_point(size=0.7) + 
               xlab("Year") + ylab("Liberal-Conservative") + 
               geom_text_repel(data=pdatl %>%
                    filter(year == min(year)),
                  aes(label = party, color = party),
                  size = 3,
                  nudge_x = -8,
                  point.padding = 0.1,
                  segment.color = NA,
                  show.legend = FALSE) +
              theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
              scale_color_manual(values=c("REP"="#c70828","DEM"="#1372ad",
                                         "N. DEM"="#6194F4",
                                         "S. DEM"="#81c4e4"),guide="none") +
              theme_bw()
  gg
}
```

```{r include= FALSE, echo=FALSE}
# House Polarization Plot
gg1 <- polarized_plot("House")
ggsave("voteview_house_party_means.png", plot=gg1, device="png", 
       width=7, height=4, dpi=300)
ggsave("voteview_house_party_means.pdf", plot=gg1, device="pdf", 
       width=7, height=4)
gg1
```


```{r include = FALSE, echo=FALSE}
# Senate Plot
gg2 <- polarized_plot("Senate")
ggsave("voteview_senate_party_means.png", plot=gg2, device="png", 
       width=7, height=4, dpi=300)
ggsave("voteview_senate_party_means.pdf", plot=gg2, device="pdf", 
       width=7, height=4)
gg2

```

```{r include=TRUE, echo=FALSE, fig.align='center'}
# liberal-conserative polarization by chamber
gg3 <- ggplot(polar_dat_long %>% 
              filter(score == "party.mean.diff.d1"),
              aes(x=year,y=value,group=chamber,col=chamber)) +
              scale_x_continuous(expand = c(0.10, 0), breaks=seq(1880,max(polar_dat_long$year),by=8)) +
              geom_line() + geom_point(size=0.5) + 
              xlab("Year") + ylab("Distance between party means") + 
              geom_text_repel(data=polar_dat_long %>%
                    ungroup() %>%
                    filter(year == min(year) & score=="party.mean.diff.d1"),
                  aes(label = chamber, color = chamber),
                  size = 3,
                  nudge_x = -10,
                  point.padding = 0.1,
                  segment.color = NA,
                  show.legend = FALSE) +
              theme(axis.text.x = element_text(angle = 60, hjust = 1), plot.title=element_text(hjust=.5), plot.subtitle=element_text(hjust=.5)) + scale_color_manual(values=c("House"="#c70828","Senate"="#1372ad"), guide="none") +
               labs(title="Figure 3: Liberal-Conservative Partisan\nPolarization by Chamber", subtitle="1880-present", caption="Data Available from Lewis et. al. (2020) at Voteview.com")


gg3

```

Polarization can and has been measured in a second way. Formally derived by Iyengar et al (2012) measurements of "affective polarization" have been derived by using data from the American National Elections Studies (ANES). Beginning in the late 1970s, the ANES has asked a series of "feeling thermometer" questions, in which respondents are asked to provide, on a scale ranging from 0-100, how they feel about a range of topics, including their feelings about political parties. Figure \@ref(fig:affect-polar) graphs the results of how partisans feel about members of their own party, members who affiliate with the opposite party, the difference between these two answers, and a total level of affective polarization, which is the sum of the difference for each of the two main parties in the United States.

As Figure \@ref(fig:affect-polar) shows, respective levels of favorability for each party have remained more or less constant over the past several decades. However, the favorability levels among partisans for those on the opposite side of the political spectrum have dropped significantly over the same time period. In short, partisans do not necessarily like the party with whom they affiliate more; they disdain members of the other side more. Regardless of how one measures or conceives of polarization, indications are strong that polarization in the contemporary United States is higher than it has been in any recent period since the end of the US Civil War. 

```{r affect-polar, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.height=5.2, fig.width=7}
df <- read_csv(here("data", "Aff_Polarization.csv"))

slim_df <- df %>% select(Year:AffPol) %>% na.omit %>% view()
melted_df <- melt(slim_df, id="Year") %>% view()

melted_df %>% ggplot(aes(x=Year, y=value, color=variable)) + geom_point() +
                       geom_smooth(se=FALSE) + labs(title="Affective Polarization in the US, 1978-2020", color="ANES\nFeeling Thermometer\nValue") +   scale_color_manual(breaks=c("DemIn", "DemOut", "DemDiff","RepIn","RepOut","RepDiff","AffPol"), labels=c("Democrat-In","Democrat-Out","Democrat-Diff","Republican-In","Republican-Out","Republican-Diff","Affective Polarization"), values=c("red", "blue", "green","dodgerblue","darkorange","cyan1","black")) + theme(legend.position="bottom", plot.title=element_text(hjust=.5))

```

## Why Does Polarization Matter?

One does not have to look far today in the American political culture to grasp the fact that current levels of polarization are high. Indeed, it is little wonder that both candidates running for President of the United States in 2020 claimed that the very "soul of the nation" was at stake [@dias_biden_2020]. The questions with which the United States is wrestling about itselfreflect not only political differences among America's elected leaders, but significant gaps between the worldviews of the nation's citizens and the fundamental core values that should be adopted, implemented, and practiced. A cursory examination of just a handful of these polarizing questions confirms that they are not easy questions, nor likely to be solved without eliciting a strong response from the public. 
\singlespace

* What does it mean to be an American? What are our "American Values"? Are there any we can agree on?
* Should America be proud of its history? 
* Is the United States an inherently racist country?
    + Should we enact a "cancel culture", and if so, who should be cancelled? What should the nation do with its founding fathers, many of whom were slaveowners? 
* Should the United States have and enforce borders? Do all people have a "human right" to come to the United States?
    + Should taxpayers fund medical care for illegal immigrants? 
* Are men and women different, and to what degrees? Or are all people the same, in every way? Are the concepts of gender and sex even relevant anymore? 

\doublespace
These questions, and many more, illustrate how different worldviews reflect drastically different ideas of how the United States should look moving forward. These questions are qualitatively different from policy questions such as what percentage of national GDP should be spent on the military, whether charter schools should be supported or opposed, and whether the debt ceiling should be raised in times of fiscal crisis and uncertainty. While there are undoubtedly significant policy ramifications that stem from these questions, these listed above (and others) bear significantly on an individual's deeply held beliefs, values, and overall worldview [@white_values_2002; @mann_its_2012; @sides_identity_2018]. How the nation answers them will likely set a tone that will have lasting impacts on the nation's values, now and into the future. 

# IV2 - The Military's Prestige 

The second independent variable of focus in this study is the military's overall level of prestige. Military prestige does not refer to a single indicator but rather a conglomerate of factors, including the military's relative level of influence within the government and its standing among and within society. In this vein, the concept of military prestige is similar, though not identical, to previous studies within civil-military relations that have examined similar variables, such as that of the "balance of power" which exists between political and military leaders [@brooks_shaping_2008]. In short, military prestige refers to the degree of power, respect, and influence the military institution (and specific military actors) has and can wield - both within the government and across society. 

While there are several ways to measure political polarization, measuring the military's prestige is a bit more difficult. On one hand there are clear measurements of several proxies of military prestige, such as public opinion measures of public trust and confidence in the military [@noauthor_confidence_2020]. On the other hand, measuring a concept such as influence - which I argue should be considered when esteeming the overall level of prestige of the military - is problematic. Indeed, as Richard Betts has stated, "influence...is one of the most important and most elusive concepts in the study of politics. In decisions on use of force it is impossible to detect objectively whose influence is greatest; it can only be inferred because only the President - and maybe not even he - really knows what convinces him or changes his mind" [@betts_soldiers_1977, 5]. Nonetheless, with a general understanding of what military prestige is and what it refers to, the next task is to briefly summarize the growth in the US military's prestige in the post World War Two era. Two important drivers of the military's prestige after World War Two have been the growing institutional influence of the military within the national security apparatus and the military's growing exclusivity within American society.  

## The Military's Growing Institutional Influence

One of the main drivers of the US military's prestige over the past seventy-five years has been its increasing institutional power and influence in relation to other government departments and institutions. By virtually every objective measurement, to include budgets/spending and manpower, the US military apparatus is a behemoth of an organization, and one that influences a wide range of policy areas. This has not always been the case, however, in American history. The contemporary influence of the US military as an institution is the result of decades worth of organizational and institutional reform and evolution to the Department of Defense and the respective individual military services. 

One of the most prominent reforms to the military came in the wake of World War Two. The National Security Act of 1947, which was signed by President Harry Truman, sought to unify the defense and intelligence establishments, which despite achieving victory in World War Two, had nonetheless squabbled a fair bit as well []. The National Security Act created new institutions, including the National Security Council (NSC), the Central Intelligence Agency (CIA), and the US Air Force [@noauthor_national_1947], and more formally, merged the Departments of War and Navy into what would be first called the "National Military Establishment"- essentially a precursor to what became in 1949 the Department of Defense [@locher_has_2001]. 

For the next several decades, the reforms wrought by the National Security Act helped the defense and intelligence communities prevail in a number of conflicts, including the Cold War, and to ultimately ensure the security of the United States. So influential was the National Security Act that fifty years after its signing, President Bill Clinton stated that as a result of the institutions created by the National Security Act and the people who had served in them, a generation had "helped to secure the peace and prosperity that America enjoys...the success of their efforts and of the historic legislation enacted half a century ago is reflected in an outstanding record of achievement: nuclear war averted, the Cold War won, and the nations of the world turning to democracy and free markets" [@clinton_weekly_1997].

Like most institutional reforms, however, the National Security Act of 1947 could not nor foresee all of the difficulties that the future would bring. A second set of major institutional defense reforms was undertaken in the mid-1980s. Known as the Goldwater-Nichols Act after Senator Barry Goldwater and Congressman William "Bill" Nichols, several incidents made clear that the military lacked the requisite capabilities to operate in "joint" environments, or those which required military personnel and equipment from across the military services [@kitfield_prodigal_1997].^[This included the the failed Iranian hostage rescue mission in 1980. For an excellent account of this event and the demand for "joint" services afterwards, see @kitfield_prodigal_1997.] Especially as military technologies and equipment grew increasingly complex and sophisticated, the Goldwater-Nichols legislation sought to ensure that national defense interests were identified and prioritized over and above the parochial interests of the individual military services [@locher_has_2001]. 

Among the most significant of the changes wrought by the Goldwater-Nichols Act was codification of the Chairman of the Joint Chiefs of Staff as the President's principal military adviser, a change which in essence meant that all advice from the respective individual service chiefs of the Army, Navy, Air Force, and Marine Corps would flow through one single point of contact: the Chairman of the Joint Chiefs [@locher_has_2001; @wills_effect_2016]. Additionally, the Goldwater-Nichols Act created several regional combatant commanders who would report directly to the Secretary of Defense, thus bypassing the military service chiefs [@locher_has_2001; @wills_effect_2016]. 

As important as these and other institutional changes have been in cementing the immense roles of particular military commanders, the scope of responsibilities undertaken by those serving in these roles has also grown considerably, especially during the post-9/11 era. Although the wars in Afghanistan and Iraq each involved a sizeable civilian component for at least a portion of the war, they witnessed the military assuming control for a significant portion of the day to day decision making of the conflict, even over issues that once were, or ostensibly should be, civilian endeavors. ^[The topic of the military and civilian experience and expertise, particularly during the Iraq and Afghanistan conflicts, is a burgeoning literature that includes a mixture of official government reports, first-hand memoirs, and works that offer theoretical/historical insights. For a sample of this literature, see @iraq_study_group_us_iraq_2006; @ricks_fiasco_2006; @chandrasekaran_imperial_2007; @hodge_armed_2011; @chandrasekaran_little_2013; @gallagher_day_2019]

In summary, the National Security Act of 1947 and the Goldwater-Nichols Act of 1986 generated significant institutional changes for key positions and organizations within the US Military. Additionally, the military's experiences of prolonged and repeated deployments in the post-9/11 era have also served to strengthen the military relative to other US instititions. The legislative changes wrought over the previous decades, combined with the past two decades of overseas experience in Iraq and Afghanistan, one scholar has claimed, have had the practical effect of the Pentagon replacing the State Department as the government's "organizational center of gravity for U.S. national security and foreign policy" [@schmidt_silent_2019]. So great has the US military's institutional influence become, claims another scholar, that "Pentagon 'help' is like being hugged by an eight-hundred-pound gorilla: it can squeeze the life right out of civilian efforts" [@brooks_how_2016, 91]. 

One graphical depiction of the current relative disparities between the Department of Defense and perhaps its closest institutional partner, the Department of State, is portrayed in Figure \@ref(fig:dodstate-budgets), which shows each department's total budget in 2016 real dollars over the period from 1962-2016. One can clearly detect from the Figure that though the size of the Defense budget has dwarfed that of State for a number of decades, the relative degree to which the Defense budget is larger than that of State's has grown sharply in the post-9/11 era.

```{r dodstate-budgets, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.height=4.5, fig.width=6}

budgetDF <- read_csv(here("data", "DoDStateBudgets.csv")) %>% view()

#scale the data from millions to billions of annual dollars
budgetDF_adj <- budgetDF %>% mutate(
  Defense=Defense/1000, 
  State=State/1000
) %>% view()

melted_budgetDF <- melt(budgetDF_adj, id="Year")

melted_budgetDF %>% ggplot(aes(x=Year, y=value, color=variable)) + geom_point() + geom_smooth(se=FALSE) + labs(title="Budgets for US Departments of Defense and State, 1962-2016", color="Department", caption="Data Available\nfrom Table 4.1, Outlays by Agency, https://obamawhitehouse.archives.gov/omb/budget/Historicals") +   scale_color_manual(values=c("black", "blue")) + theme(legend.position="right", plot.title=element_text(hjust=.5)) +  scale_y_continuous(name="Billions of Dollars (2016 Real Terms)", limits=c(0, 700)) 


```

## The Military's Growing Exclusivity Within American Society

A second driver of the US military's growing prestige is the degree to which the military institution is esteemed by society. In the post-Vietnam War era, even as the US military has accrued significant influence relative to other government institutions, the Armed Forces have also emerged, at least by many measurements, as the public's most trusted institution [@noauthor_confidence_2020]. While there is nothing wrong at the surface level with a military that is highly regarded by the public, contemporary scholars have expressed concerns about the military being regarded _too highly_ by society. Within the civil-military relations scholarly literature, the topic of the military's relational distance between the military and society is often referred to as the civil-military "gap" [@feaver_soldiers_2001]. For the past twenty years or so, numerous scholars have addressed the gaps's origins, scope, implications, and have raised numerous solutions for shrinking the gap [@feaver_soldiers_2001; @schake_warriors_2016].

At first glance, one might think that a gap between society and the military might diminish, rather than enhance, the military's overall level of prestige. After all, a nation that is entirely separated from its military, in the sense that it is unable to observe, relate to, and know anything about the military or those that serve in it, is unlikely to foster appreciation and respect for the military institution. While a national military that is fully separated from society is certainly a plausible theoretical outcome, most scholars have expressed the opposite concern for the contemporary United States: the distance and quality of the separation between the modern US military and American society has resulted in the military becoming so prominently respected and trusted that the potential harm caused by this development far outweighs any potential benefits. 

In the United States, and in most countries, the gap between society and the military has changed over time. In the American context, Americans have generally been suspicious of large numbers of standing military forces since the founding of the United States. In fact, the American founders held significant differences of opinion over how large the nation's military forces should be, and whether these forces should be comprised of volunteers (favored by Thomas Jefferson) or a standing military (favored by Alexander Hamilton) [@weigley_american_2001]. In the end, the founders established a hybrid concept, forming a moderately-sized standing Army that could be augmented, as required, by state militias comprised of "citizen soldiers" [@weigley_american_2001]. 

Furthermore, for most of its history, the United States has relied on the "citizen soldier" concept when fielding military forces, especially for war. It was predominantly not until the the 20th century when the government faced wartime demands that required instituting mandatory military service, particularly during World War Two and later, Vietnam [@rostker_i_2006]. The Vietnam War proved to be tremendously unpopular with domestic American audiences as the length of the war increased, especially after the North Vietnamese initiated their "Tet Offensive" in 1968 [@crane_avoiding_2002; @summers_strategy_1982; @petraeus_american_1987]. Moreover, in the latter years of the war, Americans came to view the existing draft system as inequitable from racial and economic perspectives [@rostker_i_2006]. The growing population of the United States, which from a "supply" perspective, translated into more available manpower, combined with severe disciplinary issues that plagued many of the military services at the end of the war [@romjue_active_1984], led many Americans to believe that instituting an All-Volunteer Force (AVF) was at least a viable alternative to maintaining the draft [@rostker_i_2006]. The watershed moment arrived in 1973, when the United States formally abandoned the draft and created the All-Volunteer Force (AVF), the result of recommendations made by the Gates Commission, established previously by President Richard Nixon [@rostker_i_2006]. 

While it is not possible to fully exposit here the relative benefits and disadvantages of the All-Volunteer Force, a task which would require a full complement of empirical and normative insights, a couple of broad summary statements speak to the overall impact of the All-Volunteer Force. On one hand, it is difficult to find a scholar, pundit, or senior military officer who does not laud the tactical proficiency and overall skill of the AVF. Indeed, there is near-universal acceptance of the tactical acumen developed and honed by the AVF since its inception. On the other hand, scholars assert that in the years since the military draft ended, there have been at least three noteworthy gaps that have formed between the military and American society: a geographical gap [@liebert_midlife_2017], a partisan gap [@golby_duty_2011; @liebert_midlife_2017], and a generational gap [@colford_young_2016]. 

Regarding the geographical gap, Golby and Liebert note that a disproportionate number of those serving in the military hail from rural and southern states, and that a large proportion of military recruitment stations, Reserve Officer Training Corps (ROTC) programs, and military installations are located in the southern United States [@liebert_midlife_2017]. The presence of this particular gap, moreover, raises concerns that the military is disproportionally attached to particular regions of the country. The partisan gap is a little more complicated. Although the military officer corps does in fact disproportionately identify with the Republican Party versus the Democratic Party [@golby_duty_2011; @liebert_midlife_2017], studies have also shown that differences between the military and civilians become smaller if not altogether diminished after conditioning on partisan affiliation [@golby_duty_2011]. In other words, the data shows that the military officer corps does identify with the Republican Party more so than with the Democratic Party, but not that military and civilian values or goals are altogether drastically different. Finally, and perhaps most problematically, studies have shown that a third gap exists between the military and younger Americans, a gap that has grown as fewer Americans have a firsthand connection to someone either serving or who has served in the military [@colford_young_2016]. Furthermore, in recent surveys which actually display that trust in the military is beginning to decrease, the data shows that the lowest rates of trust in the military are among those under "30 years of age" [@mcintyre_people_2021].

To the extent that these particular three gaps should, and actually do, matter is the subject of considerable debate. In more general terms, critics of the AVF allege that as the gap between society and the military has grown, regardless of any increase in the military's tactical abilities, it has simply become too easy to send the military to war, as there is little if any corresponding burden society must bear in doing so [@bacevich_breach_2014; @fallows_tragedy_2014]. These critics further allege that as society's burden in going to war shrinks, the public does not hold elected leaders as accountable for going to war, and that the nation ultimately will fight irresponsible wars of choice, which ultimately result in both indecisive outcomes and an altogether feckless foreign policy [@bacevich_breach_2014]. 

As proof of their argument, these critics point to the military's post-9/11 involvement in Iraq and Afghanistan [@bacevich_breach_2014]. For instance, the historian Andrew Bacevich, himself a retired Army Colonel, points out that as the American military was invading and stabilizing Afghanistan and Iraq in 2001 and 2003 respectively, President George W. Bush lowered taxes and encouraged Americans to travel vacation [@bacevich_breach_2014]. Bacevich further asserts that the net result of the military enduring repeated deployments to Iraq and Afghanistan, and achieving outcomes in these places that fall far short of strategic victory, is that "the state now owns war, with the country consigned to observer status" [@bacevich_breach_2014, 44]. Furthermore, this sentiment has likewise been captured by the journalist James Fallows, who, in an influential article, alleges that Americans in the post-9/11 era have developed a "reverent but disengaged attitude toward the military - we love the Troops, but we'd rather not think about them." [@fallows_tragedy_2014].

In addition to its impacts on American foreign policy, other scholars worry that an excessive gap between society and the military harms the quality of civil-military dialogue by mconstraining civilian input into the policy process, which in turn can likewise threaten a range of foreign policy outcomes. When interacting with civilian leaders in the context of a sizeable gap between the military and society, certain military actors may adopt an attitude such that they believe their military service entitles them to be heard in a deferential manner [@beehner_matter_2020]. A similar dynamic can occur if civilian leaders fail to sufficiently assert themselves into the policy process when interacting with military actors who are incredibly popular. Karlin and Friend, for instance, note the dynamic of "military worship" in which military service is somehow viewed as better, more relevant, or more appropriate than other types of civilian service which ultimately harms civilian control of the military and other democratic norms [@friend_military_2018]. As a result, military actors may insist on solving problems in the manner that they want the job done, and not in a way that aligns with the desire of civilian leaders. 

Regardless of the mechanism through which it works, there is a general consensus among scholars that the contemporary United States military is either very close to or at a point in which it is esteemed too highly. While recent polls show that trust in the military may in fact be slipping [@mcintyre_people_2021], the fact is that relative to other institutions, Americans greatly respect their military. According to recent Gallup public opinion polls, 72 percent of Americans indicate that they trust the military a "great deal" or "quite a lot" - the most of any institution, to include Congress (13%), the Presidency (39%), the police (48%), the Supreme Court (40%), and the church/organized religion (42%) [@noauthor_confidence_2020]. Moreover, trends in trust of various American institutions over time reveal that whereas trust in the military has generally increased over the past 45 years, trust in the other institutions have generally declined or remained the same. These trends are displayed in Figure \@ref(fig:insttrust). 
```{r code, warning=FALSE, echo=FALSE, include=FALSE}
df_trust <- read_csv(here("data","inst_trust.csv")) %>% print()
df_trust <- rename(df_trust, SupremeCourt="Supreme Court")

df_trust_consolidated <- df_trust %>% 
  select(Year, Military, Congress, SupremeCourt, Church, Newspapers) %>% 
  gather(key="Institution", value="value", -Year) 
```

```{r insttrust, warning=FALSE, echo=FALSE, include=TRUE, fig.height=4.5, fig.width=6, fig.cap = "Public Trust in American Institutions Over Time"}

df_trust_consolidated %>% ggplot(aes(x=Year, y=value)) + geom_line(aes(color=Institution, color = Institution)) + scale_color_manual(values=c("dodgerblue", "darkgreen", "black", "seagreen2", "salmon3")) + labs(x="Year", y="Percent Surveyed Who Place\n A 'Great Deal' or 'Quite a Lot' of Trust In...", caption="Data Available from 'Confidence in Institutions,'\n https://news.gallup.com/poll/1597/Confidence-Institutions.aspx")

```

# How Polarization and Prestige Impact Central Principles of Civil-Military Relations

Having now defined and provided an overview of both the dependent and independent variables, this theory chapter will now connect these variables into a cohesive theory. As Figure \@ref(fig:theory-overview) at the beginning of this chapter shows, the main argument of this dissertation is that polarization and military prestige impact the constraining influence of the three governing standards of American civil-military relations (civilian control of the military, non-partisanship of the military institution, and the non-interference of the military into civilian political affairs). Furthermore, as Figure \@ref(fig:theory-overview) shows, when these three central principles of civil-military relations lose their power to constrain the political behavior of civilian and military actors involving the military institution, we can expect to see increasingly blatant behaviors committed by both civilian and military actors that challenge, if not violate, these three central principles of civil-military relations. In other words, my argument is that the level of polarization within a society and the degree of prestige of the military impact the degree to which the principles of civilian control, non-partisanship, and the non-interference of the military into civilian affairs are considered sacred or inviolable by both military and civilian actors. When these principles are considered sacred, there are less instances of political behavior which challenge or violate these principles; on the other hand, when these principles are thought of as idealistic but altogether not binding, or generally less important than something else, there will be more instances of civilian and military actors behaving politically in ways that challenge or violate these standards. 

Furthermore, in the American context, I argue that the level of political polarization provides a "motive" for military and civilian leaders to commit acts of political behavior that challenge the central standards of civil-military relations, while the level of military prestige provides the "opportunity" to do so [@finer_man_1962]. The application of a "motive" and "opportunities" framework to the contemporary American context follows other works which have done the same to explain military intervention in politics broadly speaking [@finer_man_1962] and in the comparative setting of the Russian military [@taylor_politics_2003]. 

## Polarization - A "Motive" to Challenge the Principles of Civil-Military Relations

I argue that as polarization rises, the degree of _ideological unity_ among military and civilian actors decreases. This occurs because highly polarized climates are marked not merely by strong political disagreements (which are nonetheless present when polarization is high), but also by strong disagreement and division over which moral values and purposes should govern or influence society, and to what extent. As polarization rises, therefore, military and civilian actors increasingly encounter situations in which they must make judgements about whether to abide by the fundamental standards of civil-military relations, or to engage in acts of behavior that, while challenge or violate the central governing principles of civil-military relations, nonetheless advance a particular belief system, set of values, or ideology that the officer or civilian leader supports.

Especially in the case of military officers, the tension that arises when faced with a choice to obey either one's professional responsibilities or the dictates of one's personal conscience is neither a new nor rarely discussed phenomenon. Indeed, recent scholarship has recognized the difficulty military officers face when encountering situations in which they must choose between following one's "professional versus private morality" [@cohn_what_2019; @robinson_dissents_2020].^[For a context of morality in war, see also @dubik_just_2016] As scholars have noted, military officers in particular routinely face competing demands of loyalty. That is, the duties required of professional military officership requires that military officers remain loyal to multiple entities: to their civilian leaders, whose orders they follow; to the public, for whom the military institution exists to ultimately defend; to their fellow troops, with whom they serve; and to themselves [@cohn_what_2019; @robinson_dissents_2020]. 

Because these entities can all desire different things from time to time, there are occasions when choosing to be loyal to one entity comes at the direct exclusion of another. Furthermore, even among scholars of civil-military relations, there is relatively little agreement regarding how officers should choose when encountering situations in which their loyalties are divided, with some asserting that civilians should and must always be followed [@kohn_richard_2010], and others arguing that because officers are moral agents, they should never follow an order which they believe is inherently wrong [@milburn_breaking_2010; @milburn_when_2019]. Moreover, making decisions in these types of situations is not easy under any circumstances, nor are these circumstances confined to eras in which the level of polarization in society is high.

What is new from a theoretical perspective is acknowledgement that relative to less polarized settings, highly polarized environments increase the frequency with which military officers encounter these types of situations, as well as the magnitude of the disparity _between_ the outcomes of adhering to the central principles of civil-military relations versus following one's conscience. While it is not possible to predict all of the possible issue areas or situations which could provide a "motive" for civilian and military actors to challenge the governing principles of civil-military relations, the characteristics of today's highly polarized environment in the United States adequately capture why this is so.  

Some military officers will be willing to violate the principle of civilian control when they deem that the preferred policies held and promulgated by a particular civilian leader are not simply politically unwise, but wrong in a moral sense. In a similar vein, some military officers may be willing to violate the principle of non-partisanship if they believe that one of the major political parties possesses a political agenda that is simply wiser from a political perspective than the other party, but rather one which aligns to the moral views of the officer on a particular issue or sets of issues. Finally, some military officers will be more inclined to violate a standard of "non-interference" of the military into certain areas of state policy if they believe that the military as an institution _should_ play a greater role in advancing the underlying ideology or values associated with a particular issue or set of issues.  

Similar dynamics also impact certain civilian leaders in highly polarized environments. While civilian leaders cannot expressly violate the standard of civilian control, certain civilian leaders may be inclined to challenge the principle of non-partisanship of the military if they believe that the advantages to doing so outweigh the potential risks. Likewise, certain civilian leaders may challenge the standard of non-interference by directing the military to fill a role that falls outside of its traditional military expertise if doing so advances the leader's underlying preferred values, ideology, or policies. Moreover, these expectations of behavior follow recent experimental research that shows that a significant majority of Americans believe it is more important to advance candidates for office with whom they share a partisan or ideological affiliation than it is candidates who abide by democratic norms [@graham_democracy_2020], and that citizens find certain military elites credible conditional on these elites sharing similar ideological and partisan convictions [@robinson_michael_danger_2018]. A common theme from this scholarly research is that support is granted on based on what ideological positions one follows, and not whether actors (civilian or military) follow normative guidelines. 

Among the clearest contemporary examples of the dynamics in which polarization can provide a "motive" for civilian and military actors to engage in political behaviors by altering the constraining influence of the standards of civil-military relations is the issue of race in America. For any observer or participant of American politics in the 2020 or 2021 timeframe, it comes as no surprise that race has become a if not _the_ polarizing issue in the United States. 

Some Americans, mainly those who identify or associate with the progressive wing of the Democratic Party, hold fast to a set of ideas and principles that are part of a worldview known as _Critical Race Theory_, a concept borne out of academia in the late 1970s and 1980s that views "race [as] a political construction that was invented by white people to give themselves power while excluding all other races from it..." [@lindsay_what_2021]. Furthermore, proponents of Critical Race Theory see the problem of racism as inherent in the very power structures and societal fabric of the West, and thus, view the key to overcoming racism as reshaping or dismantling the "very foundations of the liberal order, including equality theory, legal reasoning, Enlightenment rationalism, and neutral principles of constitutional law" [@lindsay_what_2021]. Finally, proponents of this worldview, such as the historian Ibram Kendi, who authored the wildly popular 2019 book, _How To Be An Antiracist_, claim that it is not enough to be against racism, but that racism - which is inherently in and part of everything - must be aggressively searched for, found, and eradicated [@lindsay_what_2021]. 

Other Americans, mainly those who identify with the political right, reject many of these tenets. While these Americans may disagree among themselved about the extent to which racism in America is inherently "structural" or "systemic," the majority of these Americans acknowledge and regret American's ugly past with racism, even as they affirm that the United States is and has nonetheless been a relatively noble country that has pursued, however imperfectly, positive ideals and goals that have helped millions of people, both within the US and throughout the world.^[To be clear, there are staunch racists in the United States, as well as within other nations, who truly believe that certain classes of people are inferior to others on the basis of skin color. I do not have this group of staunch racists in mind when describing the opponents of Critical Race Theory. I have in mind, rather, everyday Americans, including a large number of African Americans, who hold to the view that Critical Race Theory is a dangerous ideology and worldview.] Furthermore, these Americans view Critical Race Theory as a totalitarian effort that attempts to solve the problem of racism by instituting a form of racism of its own by elevating the experiences of groups who identify as oppressed at the expense of other groups [@lindsay_eight_2020]. Unlike many proponents of Critical Race Theory, these Americans believe that the names of George Washington, Abraham Lincoln, and the like should remain as examples for all Americans to aspire to, despite any connections the historical indviduals may have had with racism [@ho_san_2021].

How might the issue of race in America serve as a motive for military and civilian actors to engage in behaviors that challenge governing standards of civil-military relations? Much of the answer depends on where one stands with respect to the issue of Critical Race Theory in the United States. For example, proponents of Critical Race Theory are likely to point to the lack of proportional minority representation among senior general and flag officers and among key occupations such as fighter pilots as proof of "systemic" or "structural" racism in the US, while opponents of Critical Race Theory are unlikely to conceptualize these realities, despite being genuine challenges that require diligence and commitment to rectify, as eternal weights that prevent the US from reaching its full potential as a nation. 

Furthermore, reflecting the dynamics of "conflict extension" previously discussed, in which polarization tends to increase the scope of issues that comprise the political battlefield [@layman_party_2002], Critical Race Theory and the "anti-racist" ideology that stems from it tends to view the world in very sharp contrast in that a person is either an "anti-racist" or a racist - there is no room for middle or neutral ground. From this perspective, the polarizing issue of race in America is one of many that can provide a motive, even for a normatively non-partisan actor such as the military, to commit behaviors that challenge or violate central principles of civil-military relations. Not only is the ideology of Critical Race Theory incompatible with remaining politically neutral, but military actors are likely among the ideology's supporters and opponents. Actors from both camps may believe that the morally and ethically right thing to do is to support/oppose the ideology, even if doing so threatens principles of civilian control, non-partisanship, and the "non-interference" of the military into certain state affairs. 

## Prestige - An "Opportunity" to Challenge the Principles of Civil-Military Relations

Whereas rising political polarization lessens the degree of ideological unity among military and civilian actors, thus providing a "motive" for the military's intervention into politics, rising military prestige advances the military's ability to pose a "latent political influence" [@brooks_militaries_2009, 231], thus offering an "opportunity" for the military to challenge standards of American civil-military relations [@finer_man_1962; @taylor_politics_2003]. The more popular and prestigious the military becomes, either as a corporate institution or as individual military figures, the greater capability the military possesses to engage in political behaviors that defy normative principles of civil-military relations. 

It is important to note that I am not predicting that rising military prestige _will_, holding all else equal, inevitably lead to a military that increasingly challenges normative expectations of civil-military relations. Rather, I am predicting, for the purposes of theory, that higher levels of military prestige will lead to a greater _capacity_ for the military to challenge these expectations. Whether the military will choose to defy or challenge principles of civilian control, non-partisanship, and non-interference of the military, or whether civilian leaders will do the same with respect to the military institution, also depends on other factors, namely, as I argue, the degree of polarization that exists within society. 

To understand why and how rising prestige enables the military to potentially challenge normative expectations of civil-military relations, one can look to both theoretical scholarship of political communication as well as American history. Indeed, a large body of research on political communication describes how and why elites are able to influence public opinion, and multiple historical episodes in the American context provide examples of military elites who, harnessing the high degree of prestige and respect afforded to them by the public, engaged in political behaviors that both challenged principles of civil-military relations, and which were likely not possible without the enormous degree of credibility, trust, and overall prestige these military elites commanded at the time. 

A vast literature describes the role of elites in influencing public opinion [@saunders_no_2017; @guisinger_mapping_2017], as well as the role of the media in this important process [@baum_relationships_2008; @kertzer_bottom-up_2017]. Among scholars who conceptualize elites as possessing an outsized ability to impact public opinion, Lupia and McCubbins (1998) note that the public looks for two central attributes among elite actors - "knowledge" and "trustworthiness" - when determining which elites to listen to as reputable sources of information [@lupia_democratic_1998, 69-76]. Further, scholars note that the more these elites are perceived by the public to hold these attributes in abundance, the greater their ability to "cue" public opinion [@zaller_nature_1992].^[Note that not all scholars believe that elites have the capacity to shape public opinion as extensively as other scholars have theorized. In particular, see @kertzer_bottom-up_2017 for a view that, using survey experiment data, argues that the mass public does not rely on elites to the extent that other scholars have claimed.] 

Currently-serving military officers - who approximate the closest thing to military "elites" referenced in the scholarly literature - likewise acknowledge the importance of maintaining a positive relationship with the American public, although they do so in language that, understandably, eschews discussion of the military's "prestige" and instead adopts the concept of "trust." Without the trust and support of the American public, the military loses its abil These scholars also emphasize the ability of elites Moreover, the public will likely ascribe these attributes to the military when the overall reputation of the military is high rather than low. Furthermore, as General Odierno's comments to senior Army leaders previously discussed reveal, military leaders generally want to maintain a high level of public trust (or to acquire one if the current level of trust enjoyed by the military is low). Therefore, when the prestige of the military is high, military actors possess the ability to leverage _public opinion_ if they choose to. This follows other research that notes the "latent political influence" that military actors can and have wielded at various points in time [@brooks_militaries_2009]. 

In addition, a vast body of research has explored the impact of military prestige on the ultimate efficacy of certain forms of political behavior, to include whether behaviors such as endorsements of presidential candidates by military leaders actually changes public opinion (most indicate they do not) [@golby_listening_2013], and the role of prestige in enabling military veteran candidates, such as Ulysses S. Grant and Dwight Eisenhower, to run for - and successfully win - public office [@teigen_why_2018]. While the contexts of military prestige explored in these specific studies is specific, and thus different from that which is intended here, this research is nonetheless consistent with the notion that in general terms, high military prestige provides military officers and the military as an institution with an "opportunity" to leverage public opinion [@finer_man_1962].  

\singlespace

```{r table2, echo=FALSE, include=TRUE, error=FALSE, message=FALSE}

text_tbl2 <- tribble(~"Environment", ~"Low Polarization", ~"High Polarization",
  "Low Prestige", "Low Overall Risk. Military actors are united ideologically. Military actors are unable to harness/mobilize public opinion. The military acts as a unified institution and adopts a belief that its members are servants of the Republic. The norm at greatest risk is civilian control.", 
  "Moderate Overall Risk. Military actors may be divided ideologically, but are unable to harness public opinion or bureaucratic power in pursuit of ideological causes. The military acts mostly as a unified institution and adopts a belief that its members are servants of the Republic. The norms at greatest risk are civilian control and non-partisanship.",  
  "High Prestige", "Moderate Overall Risk. Military actors are united ideologically. The military can harness their influence in government and public opinion, but are likely to do so only when their prestige is threatened. The military acts as a unified institution and adopts a belief that while they are servants of the Republic, they also can and should advocate for specific positions related to military prestige. The norm at greatest risk is civilian control.", 
  "High Overall Risk. Military actors are divided ideologically and can harness both public opinion and bureaucratic power. Some military actors believe that the pursuit of ideological causes outweighs adherence to foundational civil-military norms. The military is less likely to act as a unified institution, but certain individual military actors emerge as saviors or guardians of the national interest. All foundational civil-military norms are at risk." 
  )
 

  kable(text_tbl2, booktabs=T, caption="Overall Risk Level to Norms of Civil-Military Relations and Role Beliefs of Military Actors As Political Polarization and Military Prestige Vary") %>%
  kable_styling(full_width = F, font_size=8, latex_options="striped") %>% 
  column_spec(1, width="10em") %>% 
  column_spec(2:3, width="20em") 
```
\doublespace
The overall risk level posed by military actors to foundational norms of civil-military relations is low when polarization and prestige are both at relatively low levels. Under these conditions, military actors possess neither a motive nor opportunity to intervene in politics [@finer_man_1962]. When polarization is low, there is little if any ideological substance for military actors to engage in political behaviors that threaten norms of civil-military relations. Moreover, because the level of military prestige is also low under these conditions, the military has little institutional influence to leverage, and it lacks the ability to harness widespread public opinion. Therefore, the military is able to act as a unified institution. This is not to suggest that there is in fact no risk at all to civil-military norms under these conditions. Rather, conflict between civilians and military actors is likely to occur in a manner that is less visible to the public, and over substance areas that are directly related to the military's traditional duties and responsibilities, such as maintaining military readiness, and fighting and winning during combat. The norm of civil-military relations at greatest risk under these conditions is civilian control of the military. 

Moving to the top right of the table, when polarization is high but military is prestige is low, the overall risk level stemming from military actors to foundational civil-military norms is moderate. Stemming from conditions of high polarization, military actors may in fact possess a "motive" to intervene in politics, but lack the "opportunity" to do so as a result of low institutional prestige [@finer_man_1962]. Under these conditions, military actors may be divided ideologically. Some military actors may desire to advance ideological causes while others believe it is more important that the military remain apolitical and non-partisan at all costs. As a result of low levels in military's prestige, however, military actors who wish to advance the causes with which they strongly identify are unable to harness public opinion or institutional influence in support of these causes. For these reasons, the military will be mostly capable of acting as a unified institution. The norms at greatest risk under these conditions are civilian control and non-partisanship of the military.  

The bottom left of the table depicts a world in which domestic polarization levels are low, but military prestige is high. Under these conditions, the military possesses an opportunity to intervene in politics, but lacks the motive to do so outside of protecting its high degree of prestige. Thus, the overall risk level to civil-military norms is moderate [@finer_man_1962]. When these conditions exist, there are few partisan or ideological issues that serve to divide military actors, and thus, military actors are relatively unified. Because military actors enjoy a high level of prestige, they are willing to challenge norms of civilian control and are more assertive in their overall behavior. However, the substance of these challenges to civilian control are less likely to involve ideological concerns than they are debates about traditional military topics or matters, such as military readiness and the use of force abroad. The civil-military norm at greatest risk in this environment is civilian control of the military. 

The bottom right of the table reflects the conditions of high polarization and high military prestige. Under these conditions, the overall risk to norms of civil-military relations posed by military actors is high. Military actors have both a motive and an opportunity to intervene in politics [@finer_man_1962]. Under these conditions, the military institution faces a collective action problem and therefore, cannot act as a unified institution. Some military actors may deem it more important to support ideological causes than it is to uphold the foundational norms of civil-military relations. Due to the high level of polarization that exists in society, other military actors who wish to uphold norms of civilian control and non-partisanship at all costs will find it increasingly difficult to do so. Furthermore, because the military's prestige is also high under this set of conditions, some military actors are likely to behave in a way that harnesses the military's high level of influence in order to support or advance the causes which they support. In some cases, the actors who desire to advance the ideological causes with which they identify may seek an expanded role for the military, or themselves fill a role that is traditionally the purview of civilian leaders. Some actors believe that the military's role, above all else, is to remain subservient to civilian control and outside the realm of partisan politics. Other military actors, however, believe that their normative role as members of a prestigious institution compels them to support specific causes, even if the causes with which these actors identify are heavily conditioned by their partisan affiliation.  

## How Prestige and Polarization Impact Political Behaviors by Military Actors

Having now identified an overall level of risk to the norms of civilian control of the military, non-partisanship of the military institution, and military expertise under varying conditions of domestic political polarization and military prestige, it is now possible to develop a set of predictions and observable indicators about specific political behaviors that military actors may commit when these conditions prevail. Based on the theory laid out in this prospectus thus far, the goal is to make predictions that are as granular as possible without extending past what the theoretical groundwork allows. I build off of the specific political activities highlighted by Brooks but now project how the combined influence of polarization and prestige impact several of these activities [@brooks_militaries_2009]. 

The first type of political behavior, or "tactic," that military actors can commit is a "public appeal" [@brooks_militaries_2009]. In this category, I include writing op-eds, giving interviews, and even leaking information. In environments where polarization and prestige are both low, we would expect military actors to make few public appeals. If military actors do make public appeals under these conditions, the topical substance of these appeals will likely center less on ideological issues, but rather directly on topics that are related to traditional matters of military expertise, such as military deployments, readiness, weapons systems, etc. Under conditions of low prestige and high polarization, military actors would still likely conduct a limited number of public appeals, but the few they do make might occasionally address a topic outside of the traditional realm of military expertise, or overtly criticize a civilian leader for ideological differences. The overall number and frequency of public appeals under conditions of low military prestige, however, is likely to be low relative to conditions of high military prestige. 

Under conditions of high prestige and low polarization, the military will appeal to the public more often than they would under conditions of low military prestige. However, in terms of content, we expect that unified military actors will limit the topical content of their public appeals to issues or concerns that are directly related to traditional matters of military expertise. Under conditions of high polarization and high military prestige, however, military actors (who are no longer unified ideologically) might frequently make public appeals. The content of these public appeals, will include a higher frequency and overall level of appeals that stray outside of topics that are directly related to traditional matters of military expertise. Some of the public appeals made by military actors will adopt an increasingly partisan tone, and some will criticize civilian leaders in topical areas that stray further from the topical expertise of military actors. 

The second type of political behavior is "grandstanding" [@brooks_militaries_2009], which refers to the resignation (or threat of resignation) of a military actor. In political environments of both low polarization and military prestige, there would exist little if any ideological basis for military actors to resign, and so, any instances of resignation would likely be for disagreeing with civilians on distinctly military issues or topics. But owing to low levels of military prestige, there would be relatively little public impact if a military actor were to resign. Thus, we would expect few instances of resignation, and the few instances that may occur to be based not on partisan ideological differences, but on concerns regarding traditional military topics. When polarization is high but military prestige is still low, there is still little if any public impact to be gained by a military actor for resigning. However, the likelihood increases that the basis for a military actor's resignation includes ideological differences with civilian leaders.  

Under conditions of high prestige and low polarization, military actors now possess the ability to leverage public opinion to impose political costs on civilian leaders through resignation. Thus, there is a reasonable expectation to see more resignation or threats of resignation under these conditions. However, there is little if any reason to expect an ideological basis for the resignation of a military actor. Rather, instances of resignation are likely to involve disagreement over traditional military topics. These expectations change in environments marked by both high polarization and high military prestige. Under these conditions, there is an expectation that an increased number of officers may threaten to or actually resign relative to conditions of low military prestige, and also an expectation that the basis for the resignation of these military actors will include ideological issues that are beyond the traditional realm of military expertise. 

The third type of political behavior Brooks identifies is "politicking," which refers to the act of retired military officers making official endorsements of partisan political candidates [@brooks_militaries_2009]. I include in this category behaviors such as military officers speaking at political conventions, or campaigning on behalf of a political candidate. Under conditions of low military prestige and low polarization, the expectation is that few officers participate in this type of behavior. There exists little if any demand from civilian political candidates for the endorsement of active or retired military officers because military actors do not possess a high level of prestige. There is also under these conditions little if any ideological motivation to propel an officer to support a partisan candidate for office over another candidate. Under conditions of high polarization and low military prestige, some officers might actually want to put their support behind a candidate with whom they share similar ideological or partisan convictions, but a candidate for office still has little incentive to gain from the public in obtaining a military actor's endorsement. Under both of these conditions, the expectation is that there are few instances of military actor endorsement of political candidates. 

Under conditions of low polarization and high military prestige, expectations are different. Because of the high prestige of the military institution, political candidates will seek the endorsement of military actors. Although some (perhaps even most) military actors will avoid making an endorsement of a political candidate so as to not jeopardize the norm of non-partisanship of the military institution, other actors will agree to endorse. However, because the level of political polarization is low, visible endorsements may be lopsided. There is a greater likelihood under conditions of low polarization relative to high polarization that military actors (who are relatively ideologically united) will coalesce around a candidate whom they view as supportive of the military, i.e. a candidate who will increase military spending, take care of the troops, etc. Endorsements made by military actors will focus on what a particular candidate will do for the military. In environments marked by high military prestige and high political polarization, several differences are likely to emerge. I expect the endorsement of political candidates by military actors not to be lopsided, but more evenly spread. Moreover, under these conditions, military actors are more likely to base their endorsements not on narrow conceptions of what a political candidate will do for the military, but rather on shared ideological convictions between the military actor and the political candidate. Under these conditions, the question of which candidate is better for the military thus takes on an entirely different meaning than it does under conditions of high military prestige and low polarization. Endorsements by military actors under these conditions are likely to involve content that focuses less on why the candidate whom a military actor is endorsing is better for the military institution, and more on why the candidate is better for America. In making endorsements under these conditions, military actors will thus likely discuss topics that fall outside of the actor's traditional military expertise, and are likely to attack or criticize the other political candidate on partisan ideological grounds. 

In her typology, Brooks identifies two additional types of political behaviors, each of which has to deal with how the military engages Congress.  The first is that of "alliance building," in which "military leaders form ties with interest groups" in an effort to lobby Congress, and the second is "shoulder tapping," which refers to "military leaders setting an agenda by bringing issues to the attention of politicians and engaging in lobbying-like activities on behalf of those issues" [@brooks_militaries_2009, 219]. I combine these behaviors under a more general description of the military engaging Congress as a third party. When political polarization and military prestige are low, the military is likely to engage in this sort of behavior to obtain funding or weapons systems that will enable the military to win on the battlefield in order to improve its overall level of prestige. Because military prestige is low, there is less reason to expect that members of Congress will listen to the military than if the level of military prestige is high. When political polarization is high and prestige is low, some military actors may engage in lobbying efforts to advance causes other than the military's ability to fight and win on the battlefield, but these efforts will likely be limited and occur out of the public eye.

When military prestige is high but political polarization is low, military actors are likely to lobby Congress for matters that the military insists are directly related to their combat mission, such as readiness. Moreover, such lobbying efforts are likely to be successful because members of Congress do not want to be viewed by their constituents as failing to support the military. Additionally, military actors will likely focus their lobbying efforts on pro-military, influential members of Congress who can subsequently put pressure on other members of Congress to support the items for which the military is lobbying. When polarization and military prestige are high, some military actors may engage in efforts to advance ideological causes not directly related to the military's combat mission. These efforts may result in visible connections between members of the military and members of Congress that that fall outside of the political center.  

These political behaviors, key indicators, and observable descriptions are displayed in Table 3, below, along with corresponding historical periods of the United States in which general conditions of polarization and military prestige prevailed. 

```{r table3, echo=FALSE, include=TRUE, error=FALSE, message=FALSE}

text_tbl3 <- tribble(~" Political Behavior", ~"Key Indicators", ~"Low Polarization, Low Prestige", ~"High Polarization, Low Prestige", ~"Low Polarization, High Prestige", ~"High Polarization, High Prestige",
  "Public Appeals and Leaks", "What topics do the public appeals address? Do they address matters of military expertise? Do they criticize civilian leaders or take partisan sides? What is the frequency/number of public appeals made?", "Topics directly relate to military expertise. Criticism of civilians or taking partisan sides is rare. The level and frequency of public appeals is low. Leaks are uncommon.", "Topics mostly address matters of military expertise. Public criticism of civilians or taking partisan sides is uncommon. The level and frequency of public appeals is low. Leaks are possible but infrequent.", "Topics directly relate to matters of military expertise. Criticism of civilians is possible, but related to military affairs. Taking ideological partisan sides unlikely. Level and frequency of public appeals is high. Leaks possible.", "Topics include matters of military expertise and ideological concerns that are tangential to military affairs. Criticism of civilians and taking partisan sides more frequent. Level and frequency of public appeals is high. Leaks likely.",
  "Resignation", "What is the reason or basis of resignation (threat of resignation)? How often do they occur, and at what level of military leadership?", "Reasons for resignation are directly related to traditional military expertise. Resignations are rare and occur at the pinnacle of military leadership.", "Some resignations or threats to resign may occur for ideological differences tangentially related to military expertise. Resignations are rare, but may occur at subordinate levels of military leadership.", "Reasons for resignation are directly related to matters of military expertise. Resignations or threats to resign are possible and impose real costs to civilian leaders, and occur at the pinnacle of military leadership.","Reasons for resignation may include ideological differences between military and civilian leaders or matters directly related to military expertise. Resignation or threats of resignation may occur at subordinate levels of military command.",
  "Political Candidate Endorsement", "Do endorsements occur at all? Are endorsements one-sided, or evenly distributed? Is the basis of the military actor's endorsement that the candidate is 'pro-military,' or is the basis ideological?", "Few endorsements made, as little demand among civilian actors exists for military actor endorsement.", "Few endorsements made, as little demand among civilians exists for military actor endorsement.", "Endorsements exist, but are lopsided. Military actors endorse the candidate whom they think is better for the military.", "Endorsements exist, and are evenly spread. Military actors endorse the candidate whom they align with ideologically and whom they believe will be better for America.",
  "Militaries Use Congress as a 3rd Party", "For what purpose does the military lobby Congress (related directly to military expertise)? Do members of Congress face pressure from constituents to support the military? Is the lobbying effort of the military effective?", "Lobbying efforts by the military are directly related to traditional military expertise. Members of Congress are not significantly pressured to support. Efforts are somewhat effective.", "Lobby efforts are mostly related to traditional military expertise, with some deviation. Members of Congress are not significantly pressured to support. Efforts are somewhat effective.", "Lobby efforts are directly related to traditional military expertise. Members of Congress feel pressure to support the military. Efforts are very effective.", "Lobby efforts are mostly related to traditional military expertise, with some deviation. Efforts are very effective, and members of Congress feel pressure to support the military.",
  "Historical Eras", "","Post Vietnam; Pre World War Two", "Gilded Age (1875) Through Pre-World War One", "Post World War Two; Post Gulf War", "Reconstruction after the Civil War; 2010-Present Day"
  )


  kable(text_tbl3, booktabs=T, caption="Expectation of Political Behaviors by Military Actors as Political Polarization and Military Prestige Vary") %>%
  kable_styling(full_width = F, font_size=8, latex_options="striped") %>% 
  column_spec(1, width="6em") %>% 
  column_spec(2:4, width="13em") %>% 
  column_spec(5, width="15em") %>%
  column_spec(6, width="17em") %>%
    landscape()
  
```

At a minimum, eleven hypotheses can be uncovered from the predictions posited in Table 3. These are: 

_H1: When political polarization is high, the military (or certain military figures) will increasingly criticize civilian leadership (indicating a drop in the principle of civilian control)._

_H2: When political polarization is low, the military (or certain military figures) will publicly weigh in on traditional national security debates (and not on domestic political concerns or issues outside of relevant military expertise)._

_H3: During periods of high prestige and credibility, the frequency of military involvement in quasi-political activities (such as writing op-eds) will increase relative to periods of low prestige and credibility._

_H4: When both political polarization and high prestige are present, the military will be increasingly vocal and at higher frequency about certain domestic ideological or partisan causes than when these factors are not present._

_H5: During periods of active ground combat, military elites will likely comment on the war, but with less frequency on other domestic issues._

_H6: Resignations (or threats of) by military actors increase as the prestige of the military rises_. 

_H7: The basis of resignation (or threat of) becomes increasingly ideological as polarization increases_.  
_H8: Endorsements by military actors become more evenly distributed as polarization increases_. 

_H9: The number and level of endorsements made by military actors increases as military prestige increases_.  

_H10: The effectiveness of military lobbying efforts with Congress increases as military prestige rises._

_H11: The ideological purpose of military lobbying efforts with Congress will widen as polarization increases. When polarization is low, the purpose of military lobbying efforts will center on matters of traditional military expertise._

Before laying out the plan of the dissertation, a few scope conditions are required to help focus the dissertation. The first involves which military actors this dissertation plans to focus on. I will focus my attention in the dissertation on _military elites_, whom I take to mean those active _and_ retired career military officers who serve in influential and visible command and staff positions. The term _military elite_ has a strong scholarly basis, though previous scholars have defined it differently and in rather loose terms. Huntington and Janowitz saw military elites as the entire officer active duty officer corps, a group that for the purposes of this study, is both too broad and yet also not broad enough [@huntington_soldier_1957; @janowitz_professional_1960]. 

The entire active duty officer corps includes junior officers who are, I argue, still in the process of adopting and learning professional norms, and importantly, what adherence to these norms looks like in practice. The political behavior of junior lieutenants and captains, while potentially interesting and certainly worth scholarly exploration, is not likely to impose costs on political leaders that are high and which result in political leaders adjusting course. However, active duty officers in the rank of roughly Lieutenant Colonel and higher, who have served close to a full-length career of twenty years or more, have learned and know the normative standards of civilian control, non-partisanship, and military expertise such that deviation from these standards can and should be considered as qualitatively different from junior officers who do the same. 

In the same vein, I argue that _military elites_ must include retired military officers, and particularly those who have achieved the rank of general or admiral. Commonly referred to as flag officers (referring to the literal flag accompanying these officers while on active duty, which bears the number of stars of their rank), these officers play a substantial role in representing the military institution to America. Today, they increasingly dialogue with the media, and as other scholars have noted, many if not most Americans cannot distinguish the behavior of retired military officers from those on active duty, which has generated concern that the American public perceives retired military officers as speaking on behalf of the military [@golby_military_2012; @golby_jim_americas_2020]. Therefore, in line with dozens of works of previous scholarship, the term _military elite_ refers to both active _and_ retired career military officers, and especially to those who have achieved the rank of general or admiral. 

The second and final scope condition involves the time horizon of the study. Here, I think there are two potential options, both of which have relative strengths and weaknesses. The primary option is to focus entirely on the post-World War Two period. Limiting the study to this time still allows the researcher to observe variation in both of the key independent variables. In addition, the military institution has more or less existed during this time period as a profession. While the threat environment has changed drastically since World War Two, the United States has more or less remained the world's prominent power in this time period. 
 
\pagebreak
\singlespace
\newpage

-------

## Military Actors

First, I will explain how the independent variables of polarization and prestige impact and influence the beliefs and attitudes of military actors regarding the inviolability of three foundational norms described in the previous section: civilian control of the military, non-partisanship of the military institution, and military expertise. In this first section, building off of the American politics literature and military history, I trace the rise of political polarization and military prestige in the contemporary US context, as well as describe the presence of war as a scope condition. This step concludes with the development of a basic "2 x 2" table in which I assign an overall level of risk that military actors pose to the foundational norms of civil-military relations as political polarization and military prestige vary. In determining these overall levels of risk, I focus on four factors: whether military actors are likely to be unified or divided in their ideological beliefs; whether the military can leverage either public opinion or their positional influence in government in challenging norms of civil-military relations; what is the main source of corporate motivation among military actors under these conditions, and which of the three foundational civil-military norms are most likely to be challenged as a result of the varying levels of polarization and prestige.  

Second, I then project how these overall levels of risk to the foundational norms of civil-military relations under varying conditions of polarization and prestige translate into observable characteristics or attributes of military actors engaging in forms of political behavior. These observable characteristics build off of the typologies Finer and Brooks previously develop, but the goal is to determine how the presence (or absence) of the independent variables of political polarization and military prestige ultimately influence which military actors (who), commit which forms of political behavior (what), for what purpose (why), with what frequency (when), and with what other actors from across government or society ("where" are these behaviors targeted, i.e. Congress, the Media, interest groups, etc.). At the conclusion of this step, I present a total of eleven hypotheses to be tested, analyzed, and explored throughout the course of the dissertation. 


# Figure Out Where


-----
A similar concept to that of prestige in the international relations literature is the concept of reputation. In this literature, the concept of reputation almost always refers to the reputation of either a state or head of state [@lupton_reexamining_2018; @renshon_leader_2018]. Additionally, the substance of one's reputation in the international relations literature generally refers to their level of resolve, a concept that is highly connected to theories of deterrence [@powell_nuclear_2003].Two points are worth briefly highlighting in carrying over portions of the concept of reputation into a working definition of military prestige. The first is that similar to the international relations literature, the prestige of the military has both an institutional or corporate component as well as an individual component. Not only is it possible for the military writ large to possess a level of prestige; it is also possible for individual military elites to also have a level of prestige attached to them. The most glaring instance of this is when individual military elites, such as high ranking generals and admirals, become familiar household names: Grant, Pershing, Eisenhower, MacArthur, and Petraeus are all examples who, in each's respective times, commanded incredible prestige as a result of the public's perception of each's contributions to ongoing wartime efforts. The second point to make concerns the substance of the military's prestige. I argue that unlike the literature on reputation which generally focuses on the single factor of an individual's or state's level of resolve, there is no single substantive factor that captures the prestige of the military. How the military is perceived to perform during war, to what extent the military is a powerful actor within the governmental bureaucracy, to what extent the public perceives the military behaving and acting ethically, and to what extent the military upholds foundational norms of civil-military relations are all important aspects of the military's prestige. In very general terms, we would expect the military's prestige to be high during or after achieving victory in times of war (post World War Two, post Gulf War), or when the military is perceived by the public to have filled a significant role in responding to a devastating natural disaster (after Hurricane Katrina), and to be low when the military is perceived to have either lost a war (post Vietnam) or to have been plagued by troubling scandals (the Navy and the Marine Corps, specifically, after the 1991 "Tailhook" Scandal). 

## Ways in Which Prestige Can Influence Civ-Mil Norms

Even if "military worship," does not occur, the fact that the military commands more resources may make it more difficult for civilians, in some cases, to assert themselves. Over the past several decades, for instance, the budget of the State Department has remained relatively constant when measured in real dollars, whereas the Defense Department's Budget has exploded [@matisek_crisis_2018]. Over the same time period, the military has taken on an expansive set of missions and responsibilities that are far outside the traditional scope of conventional combat operations [@brooks_how_2016]. Thus, even if military leaders do not willingly seek to assert themselves over civilians because they believe that their opinions are somehow "better" than that of civilians, their position of influence makes them, by default, very powerful actors capable of subverting civilian control with relative ease. Additionally, trends in the military accumulating responsibility for tasks that are outside the scope of traditional combat operations might lead to the military advocating for assuming responsibility for additional roles and tasks, which likely bring with it increased budgetary power and institutional influence, developments which could further threaten the norm of military expertise. 

These concerns are amplified, moreover, when civilian leaders are literally not present in the room and at the table when civil-military dialogue occurs. Indeed, during the Trump administration, numerous civilian positions (more than 1/3rd) in the Pentagon went unfilled, prompting concern from Congress and other lawmakers that the first half of the civilian-military dialogue process was notably absent [@seligman_civilian_2020]. While the vacancy of these positions was not a result of high levels of military prestige, a major concern was the insufficient civilian oversight of a very powerful military establishment. 

A second, and equally powerful way in which high levels of military prestige can threaten the norm of civilian control of the military is if military actors hijack the public credibility they possess in order to subvert civilian leaders. There are a number ways in which this can occur, and to varying degrees. Leaking information or making public statements on purpose, in one way or another, brings the American public squarely into the fold of any actual or perceived disagreement between civilian officials and military leaders. Engaging in this type of behavior is likely a very deliberate decision by military leaders, because one of the reasons that the military likely enjoys a high degree of public prestige is precisely because they do not routinely engage in this sort of behavior. However, depending on the circumstances, the military may conclude that the particular situation warrants making a public appeal that relies on the high credibility the military enjoys.

In summary, when militaries possess a high degree of prestige, military actors are increasingly willing to challenge norms of civilian control. Military leaders may believe that their experience entitles them to fill a certain role in the policy process, or to provide advice that deserves to be listened to over other types of advice. Similarly, civilians may fail to push back or challenge military actors when the military enjoys a powerful degree of prestige and influence. Moreover, as the government continues to rely on the military to solve problems at the exclusion of other governmental agencies and institutions, the scope of the military's duties and responsibilities grow. Thus, when military prestige is high, the military may increasingly believe that it has the solution to how to solve a particular problem, and may seek a further expansion of its roles and responsibilities. This could further threaten the norm of military expertise. 


------ more on Finer

he first of these frameworks, that espoused by the British political scientist S.E. Finer in 1962, argues that military intervention in politics is ultimately a function of militaries possessing both the "opportunity" and "motive" to do so [@finer_man_1962]. Regarding possible motives that could spur political behavior by the military, Finer argues that soldiers might adopt a specific form of "manifest destiny," or a worldview in which the military acts as "saviors" of the country [@finer_man_1962, 29]. Finer adds that such a motive can arise as a result of the confluence of the military's unique mission, sense of power, and claim of being "outside party politics" [@finer_man_1962, 29]. 

A second motive that could cause the military to intervene in politics is what Finer calls the military's "unique identification with the national interest" [@finer_man_1962, 30]. When this motive is present, Finer argues that militaries see themselves as the "custodians" of the "national interest" [@finer_man_1962, 30]. Finer adds that militaries assuming the role of custodian can take different forms, including a "duty to arbitrate or veto...to exercise it when some convulsion or decision of the civil authorities seems to them [the military] to threaten what they think are the permanent interests of the nation" [@finer_man_1962, 31]. Finally, Finer argues that militaries intervene in politics because of their own bureaucratic interests. These can include, Finer says, any of a wide-ranging set of "sectional interests," including those based on social class, regional affiliation, personal career motivations, or organizational interests [@finer_man_1962, 34-53]. 

Like Fitch, Finer introduces a typology of political interventionism by the military in which he describes four possible levels ranging from the least harmful (and therefore, appropriate) to the most damaging (and therefore, inappropriate). Militaries intervene politically at the lowest level when they "influence," which involves  making appeals "to reason or emotion" in an attempt to persuade the civilian government to adopt a certain course of action [@finer_man_1962, 77]. Finer adds that political intervention by the military at this level is not problematic and entirely "consistent with the supremacy of the civil power" [@finer_man_1962, 77]. Militaries can also intervene politically by "pressuring" or "blackmailing" civilian leaders in an effort to "convince the civil power by threat of some sanction" [@finer_man_1962, 77]. Finer adds that militaries intervening in this way are not normatively wrong in every circumstance, but certain forms of behavior, such as threatening to stage a coup, are clearly wrong from a normative perspective [@finer_man_1962, 77]. A third level of intervention is that of "displacement," and refers to the "removal" of government officials by the military as a result of "violence or the threat of violence" [@finer_man_1962, 77]. The fourth level, that of "supplantment," occurs when the military "sweeps away the civilian regime and establishes the military in its place" [@finer_man_1962, 78]. 

In describing the ways in which militaries can intervene politically, Finer is providing one of the early critiques of Huntington. Finer specifically takes issue with Huntington's concept of _military professionalism_, or the idea that the military's professional nature ensures that military actors will act in a way that does not threaten norms of civil-military relations. Finer actually points out that while the concept of military professionalism should focus the military on itself by developing the skills and expertise it need to fight and win wars, and thus, away from politics, the danger of the concept of military professionalism is that it can also be abused, particularly by military leaders who adopt the view that they serve the larger concept of "the state" rather than a particular civilian government [@finer_man_1962, 22-23]. Pointing to numerous military leaders who intervened inappropriately in politics, including von Seeckt and Gröner under the German Weimar Republic and MacArthur under President Truman in the United States, Finer argues that the reason these leaders intervened was precisely because of how they conceptualized their duties and responsibilities as professional officers [@finer_man_1962]. Finer's point is that the principle of civilian control, and not military professionalism, ought to serve as the only bulwark to keep the military out of politics. 
