---
title: "When the Pen Becomes a Sword: Retired Military Officer Commentary in a Polarized Era"
author: "Peter M. Erickson"
date: |
  `r format(Sys.time(), '%B %d, %Y')`
abstract: |
  
  _This chapter explores variation in the public commentary of retired US military officers. Drawing on an original data set of 390 opinion-editorial pieces authored by retired military officers and published in major US newspapers since 1979, the paper argues that there has been an increase in published opinion commentary which challenge - if not violate - fundamental normative standards and traditions of American civil-military relations. The chapter explains this variation as the result of increasing political polarization and military prestige, and argues that the concept of "politicizing the military," actions which are typically ascribed to those undertaken by civilian officials, must be widened to account for retired military officers as well._  
geometry: margin = 1.15in              # margin widths
fontsize: 12pt                         # text size
bibliography: dissertation.bib         # bibliography file
biblio-style: authoryear               # citation style
urlcolor: blue                         # links to external URLs
citecolor: black                       # links to bibliography
linkcolor: magenta                     # internal figures, footnotes, etc.
indent: true                          # paragraph format
                                       # true for indented paragraphs
                                       
header-includes:
    - \usepackage{setspace}
    - \usepackage{multirow}
    - \usepackage{indentfirst}
    - \usepackage{caption}
    
    
output: 
  bookdown::pdf_document2:
    keep_tex: true
    latex_engine: pdflatex
    number_sections: false
    toc: false
    fig_caption: true
    citation_package: biblatex
    includes: 
      in_header: 
        - latex-preamble.tex
  bookdown::htmldocument2: default
  bookdown::word_document2:
      toc: false
---

\doublespacing 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, include = FALSE, echo = FALSE}
library("here")       # file paths
library("tidyverse")  # workhorse package
library("tidylog")
library("kableExtra")
library("knitr")
library("ggdag")
library("dagitty")
library("gridExtra")
library("broom")
library("dplyr")
library("reshape2")
library("stargazer")
library("rstatix")
library("lme4")
library("brglm")
library("bucky")
library("scales")
```

\newpage

# Introduction

In May of 1959, retired General of the Army Omar Bradley gave an address on the occasion of Armed Forces Day in New Canaan, Connecticut, in which he stated, "I am convinced that the best service a retired general can perform is to turn in his tongue along with his suit and mothball his opinions" [Omar Bradley in @times_bradley_1959]. Bradley's comments came just days after an old colleague from across the Atlantic, retired British Field Marshall Viscount Bernard Montgomery, publicly criticized the leadership of President Dwight Eisenhower, yet another former colleague of both military heroes [@times_bradley_1959]. In admonishing Montgomery, Bradley understood the potential consequences of a popular military figure weighing in on contemporary issues. Later in his address, Bradley offered a bit of guidance to his fellow retired military officers when he stated, "It is not my purpose to contest the right of anyone - even a retired officer - to speak what he chooses. But when he presumes to speak with an authority which derives from his retired rank, he should exercise a sensible degree of circumspection and be discreet in the choice of causes to which he lends his name" [Bradley in @times_bradley_1959].

Sixty-one years later, in the Summer of 2020, a remarkable turn of domestic events in the United States sparked a flurry of retired military officer commentary that encourage fresh examination of Bradley's advice. Following the death of George Floyd in Minneapolis, and after several days of protests, which in some cases devolved into instances of rioting and looting, the military found itself at the center of attention that it likely wished to avoid altogether [@feaver_military_2020]. For starters, both the Secretary of Defense, Mark Esper, and the Chairman of the Joint Chiefs of Staff, General Mark Milley, had accompanied President Donald Trump to a controversial photo shoot at St. John's Church in Washington after security forces had displaced throngs of protesters in neighboring Lafayette Square. Within days, Esper publicly disagreed with President Trump's threat of invoking the Insurrection Act of 1807 to use active-duty military forces to quell subsequent domestic riots [@esper_secretary_2020], and Milley also apologized, telling senior military officers graduating from the National Defense University that he "should not have been there," as accompanying the President to St. John's had "created a perception of the military involved in domestic politics" [@milley_official_2020].

Moreover, several senior retired military officers addressed these events in a series of public opinion pieces and interviews, leading some civil-military relations scholars to conclude that the events of the Summer of 2020 marked "the most intense division between a president and retired officers in a generation" [@brooks_let_2020]. ^[These officers included several retired four star generals and admirals.] However, there was also discernible variation in the tone and tenor of the remarks made by these officers: although each expressed deep concern about the prospect of active-duty forces confronting peaceful protesters, and thus, violating protections held by the First Amendment of the US Constitution [@brooks_dismay_2020; @dempsey_former_2020], some officers went much further in their criticisms, leveling direct insults of the President, including his fitness to lead [@goldberg_james_2020], his administration's broader policies on race in America [@allen_moment_2020], and his general leadership abilities [@mullen_i_2020]. 

In the weeks and months to follow, several other retired and even active-duty military officers publicly wrote commentaries, manifestos, open letters, and other opinion pieces on a host of issues that challenged, if not directly violated, the military's non-partisan and politically neutral ethic. For instance, a group of junior Army officers, all of whom were recent graduates of West Point who finished at or near the top of their respective classes, wrote a lengthy treatise calling for West Point to acknowledge its culture of "whiteness" and to adopt a host of "anti-racist" measures, including that the leaders of the military academy explicitly state that "Black Lives Matter" [@askew_anti-racist_2020]. Later, a pair of retired Army Lieutenant Colonels known in defense circles for playing a prominent role in devising the military's revised counter-insurgency doctrine ahead of the respective troop surges in Iraq and Afghanistan more than a decade ago penned an open letter to the Chairman of the Joint Chiefs of Staff in which they called on the Chairman to prepare to remove the President from office after the election "by force" if necessary [@nagl__2020]. Moreover, the authors of this open letter framed the choice for the Chairman in rather stark terms: "In the Constitutional crisis described above [the President not leaving office at the end of his term], your duty is to give unambiguous orders directing US military forces to support the Constitutional transfer of power. Should you remain silent, you will be complicit in a coup d'etat" [@nagl__2020]. Throughout the Summer of 2020 and well into the fall ahead of the Presidential election, the public commentary authored by members affiliated with the military institution, active and retired, pushed if not broke the normative boundaries that have constrained members of the military from regularly engaging in this type of behavior previously. Indeed, 2020 has revealed, among many other things, that like the nation itself, members of the military feel very strongly about certain political issues of the day.

This chapter systematically examines how public commentary authored by retired military officers has changed over time, and proceeds in four parts. First, this chapter derives three fundamental norms of American civil-military relations, and further argues that these three norms constitute a "baseline" set of normative principles that shape, govern, and ultimately constrain military actors from engaging in certain types of political behaviors. Second, this chapter summarizes a theory in which two variables, political polarization and military prestige, weaken the constraining influence of these fundamental norms of civil-military relations. The third section of this chapter introduces, explores, and analyzes nearly 400 hundred opinion-editorial pieces authored by retired military officers since 1979. The data shows that although the vast majority of op-eds published by retired military officers adhere to normative standards of professional conduct, retired military officers are increasingly pushing against these norms by criticizing civilian leaders, taking expressly partisan positions, and addressing topics that are, at best, tangentially related to national defense and military security. The chapter ultimately argues that the primary driver of this variation is the changing level of political polarization within American society. The fourth and final part of the chapter briefly discusses the implications of these findings for the military, the American public, and civilian leaders. 

# Normative Principles of Military Behavior

There are at least three normative and aspirational principles that inform, shape, and guide the behavior of the American military as an institution. These normative standards have multiple origins, to include democratic tradition, scholarly literature, and several prominent values, creeds, and oaths that have been promulgated within the military institution. These fundamental principles are that of civilian control of the military, the non-partisanship of the military institution, and a norm of _non-interference_ by military actors into certain state policy areas, mainly those that are not related or at best tangentially related to national defense and military security.^[I fully develop these norms in the theory chapter of my dissertation, and provide only a summary here in this paper.] 

The first norm, that of civilian control of the military, is the most basic and fundamental question within all of civil-military relations, and describes the principle that in democracies, elected civilian leaders control the military instrument. As scholars have noted, an inherent tension exists for democratic states in that they require a military strong enough to defeat external threats, yet sufficiently subservient to its leaders so that the military cannot and will not threaten the "society it is intended to protect" [@feaver_civil-military_1996; @feaver_civil-military_1999, 214]. In the United States, the principle of civilian control is expressed prominently in the Constitution, which grants the President the power to serve as the Commander in Chief, while granting Congress unique and specific powers relate to oversight of the military.    

A second central norm of American civil-military relations is a norm of non-partisanship of the military institution. This norm of non-partisanship stipulates that the military not align itself, nor be made to align by civilian leaders, with any political party or its platform. It is the idea that the military exist as an actor outside of partisan politics [@huntington_soldier_1957; @owens_military_2015; @brooks_militaries_2009], even if, as an actor, the military must understand, acknowledge, and exercise awareness of the political ramifications of its actions [@brooks_paradoxes_2020].

A third and final norm of American civil-military relations is somewhat less emphasized than the previous two, yet one that has significant and important relevance for contemporary civil-military relations in America. This is a norm of _non-interference_ of the military into certain realms of politics, which simply means that there are policy arenas which are and should be "off limits" for military actors. Importantly, a norm of _non-interference_ does not mean that that the military should not understand the world of politics nor the political implications of its own actions, but rather that military actors are not and should not be responsible for solving certain types of state problems. 

One the prominent scholars of civil-military relations, Samuel Huntington, implied the existence of such a norm (though he did not explicitly define it) when he referred to the uniqueness of a "military mind," devoting an entire chapter of his seminal book, _The Soldier and the State_, to the topic [@huntington_soldier_1957, 59-79]. Huntington asserted that military officers had a unique "military mind" which consisted of the "values, attitudes, and perspectives which inhere in the performance of the professional military function and which are deducible from the nature of that function" [@huntington_soldier_1957, 61]. In other words, Huntington believed that the uniqueness of the military - it's mindset, it's ethic, and it's values - was the result of the functional nature of defense provided by the military for society. An implied conclusion, though one not explicitly stated by Huntington, was that if the military were to fill roles that are not related to national defense or the security of the state, the uniqueness of the "military mind" would fade, which could result in the state's security being threatened. 

Prominent American military leaders and figures have also played a role furthering these fundamental norms of civil-military relations, even as American history is rife with instances in which senior military leaders have disregarded, challenged, or violated these norms. For instance, in a well known address given to the cadets at West Point in May of 1962, General Douglas MacArthur, who certainly pushed the boundaries of the norm of civilian control of the military when he and President Truman clashed regarding American policy on the Korean Peninsula, encouraged the Corps of Cadets to focus on winning in combat and leaving other types of problems to politicians: 

\singlespacing

> **Yours is the profession of arms, the will to win, the sure knowledge that in war there is no substitute for victory, that if you lose, the Nation will be destroyed, that the very obsession of your public service must be Duty, Honor, Country. Others will debate the controversial issues, national and international, which divide men’s minds. But serene, calm, aloof, you stand as the Nation’s war guardians, as its lifeguards from the raging tides of international conflict, as its gladiators in the arena of battle. For a century and a half you have defended, guarded and protected its hallowed traditions of liberty and freedom, of right and justice. Let civilian voices argue the merits or demerits of our processes of government. Whether our strength is being sapped by deficit financing indulged in too long, by federal paternalism grown too mighty, by power groups grown too arrogant, by politics grown too corrupt, by crime grown too rampant, by morals grown too low, by taxes grown too high, by extremists grown too violent; whether our personal liberties are as firm and complete as they should be. These great national problems are not for your professional participation or military solution. Your guidepost stands out like a tenfold beacon in the night: Duty, Honor, Country** [@macarthur_duty_1962].

\doublespacing

The overall collective influence of the three fundamental norms of civil-military relations has been one of general constraint. To be clear, instances abound throughout American history in which military actors have expressly violated, or at a minimum, pushed against the boundaries of these norms. ^[The well known collisions that have occurred between Presidents and wartime Commanders - Lincoln and McClellan during the Civil War, Truman and MacArthur during the Korean War, and Obama and McChrystal during the war in Afghanistan - bear witness to this point. It would be nearly impossible to recount all of the instances in which military actors have come close to or actually violated these norms throughout American history.] Still, generally speaking, most military actors think or have thought twice, especially in the modern era, before engaging in behaviors that challenge one or more of these fundamental norms of civil-military relations. The military cares deeply about sustaining the level of trust that it has with American society, trust that has historically been based on the military's ability to not only fight and win in combat, but also to do what its civilian leaders tell it to [@odierno_repairing_2014]. As former Chief of Staff of the Army General Ray Odierno told the service's several hundred general officers during a conference in 2014, as "stewards of the Army Profession...we play a critical role in ensuring our nation’s security, both on and off the battlefield. Of great importance...is the duty of each of us to be a steward of our profession and to maintain its essential and hard won trust with society" [@odierno_repairing_2014]. 

Yet an important question remains: are there limits to the efficacy of civil-military norms to constrain particular types of behaviors committed by military actors? If there are, are these limits a function merely of contexts and personalities, or are there structural variables that also impact the ability of the normative principles of civil-military relations to constrain military behaviors? In the paragraphs to follow, I briefly review the literature on political behaviors committed by military actors before introducing a theory that suggests two domestic factors, political polarization and the prestige of the military, strongly shape the constraining influence that the norms of civil-military relations have on military actors to engage in certain types of political behaviors.  

## Political Behaviors of the Military

A basic definition of political behaviors that military actors can commit are those behaviors which generate any number of potential _political costs_ for civilian leaders. These costs vary, but include costs such as constraining the options which are available to civilian leaders, sparking public backlash against civilian leaders, and influencing or steering debate about an impending policy choice away from the actual preferences of civilian leaders [@brooks_militaries_2009]. While there are a number of specific behaviors that military leaders can engage in which can generate these types of costs, one helpful framework of these behaviors was developed by the Risa Brooks (2009), which is reproduced in Table \@ref(tab:pol-behav-tab) below. 

\singlespace

```{r pol-behav-tab, echo=FALSE, include=TRUE, error=FALSE, message=FALSE}

text_tbl1 <- tribble(~"Tactic", ~"Nature of Activity", ~"Audience", ~"Examples",
  "Public Appeal", "Public, value-laden commentary", "Mass public",  "Colin Powell's 1992 statements on intervention in Bosnia", "Grandstanding", "Threat or actual resignation in protest of policy", "Mass public, Congress, executive branch", "Hypothetical; if Eric Shinseki had resigned in 2003",
  "Politicking",  "Retired officer endorsements; organizing vote drives", "Mass public", "Crowe's endorsement of Clinton in 1992; Frank's endorsement of Bush in 2004", 
  "Alliance Building",  "Military leaders form ties with civilian interest groups", "Congress", "Air Force and Lockheed Martin over procurement of F-22 raptor",
  "Shoulder Tapping", "Military leaders setting an agenda by bringing issues to attention of politicians and engaging in lobbying-like activities on behalf of those issues", "Congress", "Military Mobilization of key members of Congress over gays in the military in early 1990s")
 

text_tbl1 %>% kable(booktabs=T, caption="Political Activities of the Military") %>%
  kable_styling(full_width = F, font_size=8, latex_options="striped" ) %>% 
  column_spec(1, width="10em") %>% 
  column_spec(2, width="15em") %>% 
  column_spec(3, width="10em") %>% 
  column_spec(4, width="15em") %>% 
  add_footnote(label = "The above table is reproduced from Brooks, Risa (2009), 'Political Activities of the Military' in\n American Civil-Military Relations: The Soldier and the State in a New Era, Eds. Suzanne Nielsen and Don Snider.")

```
\doublespace

At least two implications are worth noting from Table \@ref(tab:pol-behav-tab). The first is that not _all_ of the listed behaviors generate political costs for civilian leaders in every instance, whereas some do. For example, military officers who engage in making "public appeals" can do so, theoretically, without inducing significant political costs for civilian leaders [@brooks_militaries_2009]. When retired Army General Colin Powell wrote a series of op-eds in the early 1990s while serving as the Chairman of the Joint Chiefs of Staff in which he articulated his views regarding the use of force in the post-Cold War era, his behavior likely influenced public opinion, which likely constrained the choices available to President Bill Clinton as a number of crises erupted across the world [@powell_opinion_1992]. Because Powell was a tremendously popular figure after the Gulf War, publicly airing his views likely made President Clinton's job of determining how to respond to unfolding situations in the Balkans, Somalia, and other places more difficult. Similar political costs to those generated by Powell are not generated every time a retired General or Admiral pens an op-ed or gives an interview to the press. On the other hand, every instance of a military officer publicly endorsing a political candidate for office, or speaking at a political convention, generates costs for civilian leaders because it clearly gives the impression that the candidate who is not endorsed is officially less approved of by the military, which normatively exists as a non-partisan institution.  

A second implication from Table \@ref(tab:pol-behav-tab) is that military actors have a wide variety of tools, or behaviors, at their disposal from which to generate political costs for civilian leaders. Some of these behaviors are more visible to the public than other types of activities. In fact, two of the five activities listed in Table \@ref(tab:pol-behav-tab)) deal with military interactions with Congress, which are on the whole not as visible to the American public except in significant instances. In short, military actors have a wide range of possible options with which to thwart, frustrate, or otherwise complicate the preferences of civilian leaders, and these options range along a spectrum of visibility to the public. 

Public commentary in particular is a tricky behavior to analyze. Retired military officers (especially in the case of those who have achieved high rank) have accumulated unique knowledge and experiences over the course of lengthy careers (at least 20, and in some cases, upwards of 40 or even more years of service). Moreover, retired military actors who have achieved the rank of multistar general or admiral have interacted with Presidents, members of the cabinet, Congress, and foreign dignitaries throughout the course of their careers, and thus, their perspectives on a number of issues can be helpful in informing the public [@brooks_militaries_2009]. In a democracy especially, the perspectives of the military can help improve the free flow of information, leading to a better informed citizenry. 

In addition to Colin Powell's public commentary in the 1990s, which raised questions regarding civilian control of the military, another episode resulting in the same occurred in 2006 when, during the so-called "Revolt of the Generals," several retired military generals and admirals publicly criticized and called for the removal of Defense Secretary Donald Rumsfeld in response to what they perceived to be his poor leadership and its impacts on the war effort in Iraq. Scholars noted that the event was unprecedented in the way that multiple officers "devoid of partisan political motive, felt the need or obligation to speak out publicly regarding policy and leadership during an ongoing conflict" [@cook_revolt_2008, 7]. While the "Revolt of the Generals" generated concern that the military's professional ethic had been strained, others applauded the generals' honesty and forthrightness and claimed that concerns about a dip in the professionalism among military officers was overblown. Frederick Kagan, for instance, claimed, "there is no danger to the republic in a handful of retired generals speaking their minds. There is great danger in making vital decisions about an on-going armed struggle without hearing the views of all available experts" [@kagan_let_2006]. 

Whether one agrees with Kagan's general assessment that the positive aspects of public commentary by retired military officers outweighs the potentially negative impacts on normative standards of civil-military relations is a matter that is open for (in some cases, considerable) debate. Differences abound on this question, among both scholars and among military actors, except perhaps in the most egregious of potential cases in which a military actor blatantly assails a sitting US official. Fortunately, these cases are relatively rare, but as the data will later show, they are becoming increasingly common. In the next section of this chapter, I explain how two variables, polarization and prestige, impact the ability of the norms of civil-military relations to constrain the political behavior of military actors. 

# The Impact of Polarization and Prestige

In the American context, I argue that the level of political polarization provides the "motive" for military and civilian leaders to commit acts of political behavior (and to challenge the norms of civil-military relations in doing so), while the level of military prestige provides military actors with an "opportunity" to do so [@finer_man_1962]. Application of a "motive" and "opportunities" framework to the contemporary American context follows other works which have done the same to explain military intervention in politics broadly speaking [@finer_man_1962] and in the comparative setting of the Russian military [@taylor_politics_2003]. There are several theoretical reasons to suspect that polarization provides a "motive" for military actors to challenge the norms of civil-military relations and military prestige to provide an "opportunity" for the same [@finer_man_1962].

Polarization is a difficult concept to precisely define. From the perspective of the American politics sub-field, polarization refers to the movement of political thoughts and attitudes away from the political center and towards more ideological extremes [@hare_polarization_2014; @abramowitz_great_2018]. In a broader sense, one that is more often explored and discussed within the sub-field of political theory, political polarization describes the degree to which a society shares fundamental values, is capable of incorporating divergent worldviews, and the overall degree to which these values unite versus divide a nation [@marsden_twilight_2014; @deneen_why_2018]. Moreover, the sets of issues which characterize highly polarized environments are those which citizens care deeply about [@iyengar_affect_2012; @iyengar_origins_2019] and for which there is little room for ideological compromise [@layman_party_2002]. Although previous research has found that senior military officers identify politically with the Republican Party at greater levels than the Democratic Party, this research focused primarily on opinions regarding the use of force and was conducted when levels of political polarization were relatively low [@golby_duty_2011]. There is little if any reason to suspect that military actors would be less impacted by polarization than American society at large, and thus, I argue that polarization ultimately impacts the _ideological unity_ of military actors. To be specific, as polarization rises, the _ideological unity_ among military actors decreases. 

Military actors often encounter situations in which an officer must make judgments about whether to follow the dictates of personal conscience or adhere to their responsibilities qua professional officer. These decisions are difficult in any situation, and a fair amount of literature describes the inherent tension facing military officers in their dual capacities as moral agents and professional officers of the state [@cohn_what_2019; @dubik_just_2016]. A polarized environment complicates these challenges, however, in at least two important ways. First, polarization likely increases the frequency with which military actors encounter these situations in which they face a choice between following one's conscience or fulfilling one's professional duties, and second, polarization likely increases the magnitude of the disparity in the resulting consequences of these choices [@cohn_what_2019; @dubik_just_2016]. The underlying polarizing environment leads some military actors to choose to remain professional while others follow the dictates of their personal conscience, which in some cases will include committing one or more of the political behaviors listed in Table \@ref(tab:pol-behav-tab). 

The net result of military officers increasingly encountering situations in which they have to starkly choose between following their personal consciences versus adhering to normative standards required as professional military officers is the erosion of the ability of the norms of civil-military relations to constrain the officer's behavior, which results in an increase in the overall volume and frequency of political behaviors committed by military actors. While it is not possible to predict all of the possible issues or sets of issues that can lead to such a result, one can reasonably predict that during highly polarized political times, which are marked by contentious political differences over what citizens believe to be the fundamental values and purposes that should govern society, some military officers will be willing to violate a norm of civilian control when they deem that the preferred policies held and promulgated by a particular civilian leader are not simply politically unwise, but wrong in a moral sense. In a similar vein, some military officers may be willing to violate a norm of non-partisanship if they believe that one of the major political parties possesses not merely a more politically-savvy agenda than the other party, but rather one which aligns to the moral views of the officer on a particular issue or set of issues. Finally, some military officers will be more inclined to violate a norm of "non-interference" of the military into certain areas of state policy if they believe that the military as an institution should play a greater role in advancing the underlying ideology associated with a particular political issue or set of issues. 

If increasing political polarization provides a "motive" for challenging fundamental norms of civil-military relations, high levels of military prestige provide the corresponding "opportunity" for military intervention into politics [@finer_man_1962]. Like the concept of polarization, the concept of military prestige is difficult to precisely define, but it generally refers to the degree to which the military institution is venerated within society and across the government. Unlike polarization, which can be measured in several different ways, the concept of military prestige is slightly trickier to measure^[For example, there is public opinion time-series data on various public opinion surveys for institutions in America, but these date back to the middle of the 20th Century. It is more difficult to compare across wide time periods. For example, comparing the prestige of the US Army after the Civil War to that of the US military after World War Two is more difficult, lending itself to more qualitative comparisons].

Scholars of public opinion have noted that two of the attributes that the public looks for when determining reputable sources of information are knowledge and trustworthiness [@lupia_democratic_1998, 69]. Moreover, the public will likely ascribe these attributes to the military when the overall reputation of the military is high rather than low. Furthermore, as General Odierno's comments to senior Army leaders previously discussed reveal, military leaders generally want to maintain a high level of public trust (or to acquire one if the current level of trust enjoyed by the military is low). Therefore, when the prestige of the military is high, military actors possess the ability to leverage _public opinion_ if they choose to. This follows other research that notes the "latent political influence" that military actors can and have wielded at various points in time [@brooks_militaries_2009]. 

In addition, a vast body of research has explored the impact of military prestige on the ultimate efficacy of certain forms of political behavior, to include whether behaviors such as endorsements of presidential candidates by military leaders actually changes public opinion (most indicate they do not) [@golby_listening_2013], and the role of prestige in enabling military veteran candidates, such as Ulysses S. Grant and Dwight Eisenhower, to run for - and successfully win - public office [@teigen_why_2018]. While the contexts of military prestige explored in these specific studies is specific, and thus different from that which is intended here, this research is nonetheless consistent with the notion that in general terms, high military prestige provides military officers and the military as an institution with an "opportunity" to leverage public opinion [@finer_man_1962]. 

## The Collision of Polarization and Prestige

The theoretical expectations of polarization and military prestige on the constraining influence of fundamental principles that broadly shape civil-military relations and the political behavior of military actors specifically ultimately drive two questions: are military actors united or divided ideologically?, and can military actors leverage public opinion? These two questions, I argue, need to be asked and answered _simultaneously_ in order to determine whether military actors will engage in political behaviors that reveal a weakening in the ability of the normative principles of civil-military relations to constrain certain types of political acts at the hands of military actors. Consistent with the idea that polarization provides a "motive" for military officers to commit political behaviors by weakening the ideological unity among military actors and that military prestige provides the "opportunity" for military officers to leverage public opinion if they so choose [@finer_man_1962], four distinct political arenas in which military actors must operate are possible. These four environments are displayed in Table \@ref(tab:pol-pre-envir-tab) , below.

\singlespace

```{r pol-pre-envir-tab, echo=FALSE, include=TRUE, error=FALSE, message=FALSE}

text_tbl2 <- tribble(~"Environment", ~"Low Polarization", ~"High Polarization",
  "Low Prestige", "Ideologically United; Private Behaviors Likely", 
  "Ideologically Divided; Private Behaviors Likely",  
  "High Prestige", "Ideologically United; Public Behaviors Likely", 
  "Ideologically Divided; Public Behaviors Likely" 
  )
 

  kable(text_tbl2, booktabs=T, caption="The Impact of Polarization and Military Prestige on the Ideological Unity of Military Actors and the Military Institution's Ability to Leverage Public Opinion") %>%
  kable_styling(full_width = F, font_size=8, latex_options="striped") %>% 
  column_spec(1, width="10em") %>% 
  column_spec(2:3, width="20em") 
```
\doublespace

## Low Polarization, Low Military Prestige
Under conditions when polarization and military prestige are both relatively low, military actors possess neither a "motive" nor "opportunity" to intervene in politics [@finer_man_1962]. Low polarization levels indicate there is little if any ideological substance driving military actors to challenge foundational principles of civil-military relations, and low prestige suggests that  the military lacks the ability to leverage public opinion. This leads to a theoretical expectation that while there might certainly be conflict between military and civilian leaders over isolated, context-driven events, there will not be systematic, widespread engagement in any substantive political behavior at the hands of military actors. Examples of this era, historically, include the post-Vietnam and pre-World War two periods.

## High Polarization, Low Military Prestige
When polarization is high but military prestige is low, military actors may in fact be divided ideologically. Some military actors might possess a true "motive" to engage in acts that threaten or violate normative standards of military actor behavior [@finer_man_1962]. However, these actors lack the "opportunity" to do so because of the relatively low level of military prestige present in the domestic environment. As a result, there might be some instances of military actors engaging in less visible forms of behavior that challenge foundational norms of civil-military relations, but these efforts will be out of the public spotlight. Examples of this era include the Gilded Age through the beginning of American involvement in World War One, or approximately 1875-1915.

## Low Polarization, High Military Prestige
Under conditions of low polarization but relatively high military prestige, military actors are united ideologically, and therefore do not have possess an ideological "motive" to engage in political behaviors [@finer_man_1962]. However, the military possesses the "opportunity" to leverage public opinion if they choose to do so [@finer_man_1962]. Under these conditions, it is possible that military actors will engage in some forms of political behaviors. When they do, however, they will do so in ways that exhibit a unified front as a military institution. Moreover, under these conditions, military actors recognize that engaging in public behaviors is a double-edged sword. That is, military actors are cognizant that one of the reasons for the high degree of prestige which they hold under these conditions is the trust that society places in them, which is driven partially as a result by American society's belief that the military is not acting inappropriately, i.e. that the military is subservient to civilian control, acting in a non-partisan manner, and not interfering into certain areas of state policy that have little to do with national defense or military policy. Thus, under these domestic conditions, the military may engage in political behaviors that invoke the public, but they will do so selectively and with caution. Examples of this period include the post-World War Two until the Korean War, and the post-Gulf War period (1991) until approximately 2010.

## High Polarization, High Military Prestige
When both polarization and military prestige exist at relatively high levels, military actors possess both a "motive" and an "opportunity" to engage in politics [@finer_man_1962]. High levels of polarization suggest that military actors, like society itself, are not united ideologically. Some actors may believe so strongly in particular issues or sets of issues that are part of the polarized political landscape such that they are willing to push the boundaries of the principles of civilian control, non-partisanship, and "non-interference" of the military by engaging in activities that do so. Moreover, these military actors are able to commit these behaviors because they are capable of leveraging public opinion. Examples of this period include the post-US Civil War period during American Reconstruction, 1865-1870, and from approximately 2010 - present day. 

## Hypotheses

The theoretical expectations concerning the impacts of polarization and military prestige enable the crafting of a total of three hypotheses, which are listed below. 

H1: Military actors are increasingly active in a partisan manner when polarization is high relative to periods of low polarization.

H2: Military elites increasingly challenge civilian authorities in a visible, public manner during periods of high military prestige relative to periods of low military prestige.

H3: When military prestige is high, the level of political polarization moderates the issues over which military actors challenge civilian actors. When military prestige is high and political polarization is low, military actors will likely challenge civilians on use of force initiatives, military policy, and other issues that are directly related to national security. However, when military prestige is high and polarization is high, military actors are more likely to challenge civilians over political issues that do not directly involve the military, or are indirectly related to national security policy. 

# Methodology and Analysis

```{r introduction, include = FALSE}

# Read in the Data

dataset <- read_csv(here("data", "MilEditorials.csv"))
dataset %>% print()
# Only read in the first 390 rows - is there a better way to get rid of N/A data?

# Replace NA with 0 in the data for variables letter_ed to online_only, and leg_act #to role
adjdataset <- dataset[1:390,] %>%
  mutate_at(
    .vars = vars(letter_ed:online_only, Leg_Act:Role),
    .funs = ~ case_when(is.na(.) ~0, TRUE ~ 1)
) %>% mutate(
    PubYr = lubridate::year(PubDate)
) %>% view()

summary(adjdataset)
names(adjdataset) 

# count the number of observations by type of binary variable and store as new dataframe
countbytype  <- adjdataset %>% 
  count(Topic) %>% view()


# count the number of observations by source
countbysource <- adjdataset %>% 
  count(Source) %>% view()

#count the number of times each source occurs in the data
WSJcount <- adjdataset %>% count(Source=="WSJ")
WSJcount

NYTcount <- adjdataset %>% count(Source=="NYT")
NYTcount

WaPOcount <- adjdataset %>% count(Source=="WaPO")
WaPOcount

USATodaycount <- adjdataset %>% count(Source=="USA Today")
USATodaycount

LATimescount <- adjdataset %>% count(Source=="LA Times") %>% print()
LATimescount %>% print()

# Create a new variable called Yrcount, which is a count of the yearly number of published op-eds

adjdataset <- adjdataset %>% 
  group_by(PubYr) %>% 
  mutate(Yrcount=n()) %>% 
  view()

# number of authors and author teams
adjdataset %>% 
  group_by(Author) %>% 
  count() %>% 
  view()
```

To test these hypotheses, I examine approximately four decades worth of letters to the editor and op-eds published by retired military officers in major US newspapers. I focus on retired military officers instead of active-duty military officers mainly because  Examining written commentary, especially written opinion-editorial pieces, affords the researcher several advantages over other types of public commentary, such as television or radio interviews or other social media postings. The first advantage is that written opinion pieces which are published in major newspapers are intended for public consumption. Social media posts may certainly end up public and shared, perhaps even reaching a larger audience, but there is a qualitative difference in the views one posts on a personal social media site versus deliberately authoring for a wider media outlet. 

A second advantage of examining opinion-editorials is that the author's intent is likely more discernible than other forms of public commentary, such as television interviews, where there is a greater chance of someone misspeaking or later being misquoted. Publishing an opinion piece or letter in a major newspaper is often a competitive process, as it is not uncommon to see opinion pieces in prominent newspapers published by heads of state, members of congress, governmental figures, and leaders of industry and academia. The reader of an op-ed, regardless of whether he or she agrees with the author, is likely to conclude that the author intended what the author wrote.

A third advantage of examining op-eds and letters to the editor, one which stems from the second advantage above, is that opinion pieces generally make an argument for something, and thus, the researcher can discern (assuming the op-ed is clearly written) the object the author is attempting to pursue. The content of an op-ed can take any number of forms, but generally, op-eds are intended to address a policy issue, or a newsworthy topic such as current events, or to argue that the government or institution take a stand for something. In short, examining opinion commentary that is written in form enables the researcher to discern what message(s) the author intends to send.^[Because I am interested in messages that required the officer to craft a deliberate message, I do not include published interviews with retired military officers, nor any reprinted speeches or remarks that a retired military officer gave at another venue. I also do not include short forms of analysis, such as that provided by the "Monkey Cage" from _The Washington Post_.]

It should be noted that the overall influence and impact of newspapers is likely less than it was several decades ago - not more. After all, social media has taken over as the primary means in which a majority of Americans receive their news [@martin_how_2018]. Furthermore, even after accounting for the advent of online editions of newspapers in the late 1990s and early 2000s, the overall circulation of newspapers (print and online editions combined) is still down today compared to earlier years - as a means of rough comparison, the estimated total daily newspaper circulation in the US in 2000 was 62 million for weekday and 54 million for weekend editions, respectively, while in 2018, the same figures were 29 million and 31 million [@barthel_leading_2019]. Nonetheless, I do not view these trends in newspaper circulation as significant for the purposes of this study, as I am interested in more in the question of what military officers are stating in the newspapers that are published over time.

## The Data

The data consists of 390 observations authored from 1979-2020 and published in a total of five major newspapers: _The Wall Street Journal_ (WSJ), _The New York Times_ (NYT), _The Washington Post_ (WaPo), _The Los Angeles Times_, and _The USA Today_. With the exception of _The USA Today_, all of the newspapers have long and established histories (the _USA Today_ began in 1982 [@noauthor_about_nodate]). These five newspapers, and in particular, the _The Wall Street Journal_ (WSJ), _The New York Times_ (NYT), and _The Washington Post_ (WaPo) are likely the most circulated among the defense and national security communities, as they are located in New York City and Washington D.C., where a significant number of prominent think tanks and international organizations, as well as the Pentagon, are based.

To obtain the data, I use a series of key-word searches in several different databases, including Factiva, Lexus-Uni, and ProQuest Historical Newspapers, that generally look for words such as "retired military officer," "former General," "former Admiral," etc. ^[For the exact search parameters used to collect the data, please see the Appendix. Overall, there were two main challenges in collecting the data. The first was that not all retired military officers clearly identify themselves as such. Some used different titles, such as "Former Commander of US Central Command." This required the author to further investigate and to conclusively determine whether the author was in fact a retired military officer. The second challenge was that each of the databases used to collect the data organizes and classifies historical newspaper articles somewhat differently. Some databases, for instance, include editorials as a source type to search for, making the acquisition of data easier, while others do not, which requires conducting a more-exhaustive search of a much larger set of articles that contain the search parameters.] I include in the data articles or letters to the editor regardless of whether they were published in print or online, but count identical articles published in both mediums only one once, even if the online and print versions contain relatively different titles (sometimes they do). 

In some instances, especially where the same retired military officer authored or co-authored multiple publications, the officer did not always refer to himself or herself as a retired military officer. This occurred in several instances in which the officer was, ostensibly, writing from the perspective of someone other than a retired military officer, such as that of a political appointee to a high level governmental position (after retirement from the military, Army General Barry McCaffrey, for instance, served as President Clinton's Director of the Office of National Drug Control Policy). Nonetheless, because these officers had previously identified themselves as retired military officers in earlier publications, and thus, because members of the public might reasonably recognize the author as a prominent retired military figure, I include the subsequent publications in the data, and simply control for whether the author self-identifies as a retired military officer in a specific publication. 

The methodological approach used in the analysis to follow consists of two broad steps. The first is to analyze the general _topics_ which retired military officers are addressing, and the second is to determine whether a particular opinion piece or letter to the editor has challenged any of the normative principles of civilian control, non-partisanship, or "non-interference" of the military into certain areas of state policy. After presenting summary statistics, I describe how I implement this two-step approach as well as the subsequent results. 

## Summary Statistics

The data set includes a total of 390 observations, penned by 216 distinct different authors or author teams. Eight of the observations are written by anonymous authors. Included in the data are cases in which at least one retired military officer co-authored a piece with one or more civilian authors. ^[There are several co-authored pieces. For example, where General (Retired) David Petraeus wrote one publication and another one was written by General (Retired) David Petraeus and Michael O'Hanlon, I counted these as two different authors for coding purposes.] There are also several retired military officers who penned several observations. The most prolific of these officers includes Retired Army Lieutenant General William Odom (22 publications in total), Retired Air Force General Michael Hayden (20 publications in total), and Retired Marine Lieutenant Colonel Robert McFarlane, who later served as President Reagan's National Security Adviser (18 publications). 

```{r, include=FALSE, echo=FALSE}
# show summary stats for yearly count, rank, and then a breakdown according to source
# yearly count
yearstats <- adjdataset %>% 
  select(PubYr, Yrcount) %>% 
  distinct(PubYr, Yrcount) %>% 
  print()

summary(yearstats)
sd(yearstats$Yrcount)
var(yearstats$Yrcount)

#rank
rankstats <- adjdataset %>% 
  filter(AuthRank <=10) %>% 
  select(Author, AuthRank) %>% print()

summary(rankstats)
sd(rankstats$AuthRank)
var(rankstats$AuthRank)

#sex
adjdataset %>% group_by(AuthSex) %>% count()

#service
adjdataset %>% group_by(Service) %>% count()

Descriptive_Stats <- tribble(
  ~Variable, ~Min, ~Max, ~Mean, ~SD, ~N, 
  'Author Rank', 4, 10, 8.34, 1.99, 390,
  'Number of Yearly Publications', 1, 27, 9.51, 6.84, 390)

Sex_Stats <- tribble(~Breakdown, ~Observations, ~Percentage,
   'Male Authors', 374, 95.9,
  'Female Authors', 3, .8,
  'Mixed  Authorship', 5, 1.3,
  'Sex Unknown', 8, 2,)
  
Service_Stats <- tribble(~Breakdown, ~Observations, ~Percentage,
  'Army', 205, 52.5,
  'Navy',  56, 14.3,
  'Marine Corps', 47, 12.1,
  'Air Force', 63, 16.2,
  'Coast Guard',3,.8,
  'Mixed Service Authorship', 16, 4.1)
  

 rankdistrib <- adjdataset %>% group_by(AuthRank) %>% count()
 rankdistrib_scaled <- rankdistrib %>% 
   mutate(
     Rank=case_when(
       AuthRank== 4 ~ "4: Major/Lieutenant Commander",
       AuthRank== 5 ~ "5: Lieutenant Colonel/Commander",
       AuthRank ==6 ~ "6: Colonel/Captain",
       AuthRank ==7 ~ "7: Brigadier General/Rear Admiral (Lower Half)",
       AuthRank == 8 ~ "8: Major General/Rear Admiral (Upper Half)",
       AuthRank ==9 ~ "9: Lieutenant General/Vice Admiral",
       AuthRank == 10 ~ "10: General/Admiral"
     )
   )

```

In addition to coding each publication for the source and a count of publications in a given year, several bits of biographical information about each author is collected, including the service in which the author served, the rank achieved by the author during his or her time in service, and the sex of the author. Summary statistics for the variables of author rank and the annual count of publications are displayed in Table \@ref(tab:desc-stats), below. Of note, the variable for author rank is coded equivalently to the officer's military pay grade. For example, a rank of "10" means the officer achieved the pay grade of O-10, which is a four star general or admiral. A rank value of 7 refers to a one star general or admiral. A rank of 4 corresponds to a pay grade of O-4, which in the Army, Air Force, and Marine Corps, refers to the rank of Major. In the Navy and Coast Guard, a pay grade of O-4 refers to the rank of Lieutenant Commander. A rank of 5 in the Army, Air Force, and Marine Corps refers to a Lieutenant Colonel, whereas in the Navy and Coast Guard, a rank of 5 denotes a Commander.  Finally, a rank of 6 in the Army, Air Force, and Marine Corps refers to a Colonel, whereas in the Navy and Coast Guard, a rank of 6 corresponds to the rank of Captain. 


```{r desc-stats, echo=FALSE, include=TRUE, warning=FALSE, error=FALSE}
T1 <- kable(Descriptive_Stats, 
            format = 'simple', 
            booktabs = TRUE, 
            caption = "Descriptive Statistics for Author Rank and Yearly Number of Publications Breakdown of Observations by Author Sex and Service Affiliation",
            align = c('l', 'c', 'c', 'c'),
            digits = 2)

T1
```

Table \@ref(tab:sex-stats) and Table \@ref(tab:service-stats) below, show the breakdown of observations by author sex and service affiliation, respectively. Mixed Authorship in Table \@ref(tab:sex-stats) refers to publications that were authored by a mixture of male and female authors, whereas Mixed Service authorship in Table \@ref(tab:service-stats)) denotes publications that were authored by more than one individual and together, as an author team, had served in multiple service branches - for example, a retired Colonel who had served in the Air Force and a retired Admiral who had served in the Navy write an op-ed together. 

```{r sex-stats, echo=FALSE, include=TRUE, warning=FALSE, error=FALSE}
T2 <- kable(Sex_Stats, 
            format = 'simple', 
            booktabs = TRUE, 
            caption = "Breakdown of Observations by Author Sex",
            align = c('l', 'c', 'c', 'c'),
            digits = 2)

T2
```


```{r service-stats, echo=FALSE, include=TRUE, warning=FALSE, error=FALSE}
T3 <- kable(Service_Stats, 
            format = 'simple', 
            booktabs = TRUE, 
            caption = "Breakdown of Observations by Author Service Affiliation",
            align = c('l', 'c', 'c', 'c'),
            digits = 2)

T3
```

In terms of the sex of the author, male authors comprise the bulk of the sample (95.9%). Because the sample includes publications dating back to 1979-2020, a probable reason for this is the difference in sex and gender at the highest ranks of service - a feature that will likely change in the coming years as the restrictions on women serving in combat occupational specialties have been removed. Unsurprisingly, as the military's largest service, retired military elites who served in the Army comprise the largest segment (52.5%) of authors in the sample, as Table \@ref(tab:service-stats) shows. Authors who served in the Navy and Air Force, whose active duty populations in 2020 were of comparable size, featured similarly in the sample (14.3% and 16.2%, respectively). Representation from authors who served in the Marine Corps (12.1%) and Coast Guard (.8%) round out the sample. 

Figure \@ref(fig:rank-pubs), below, displays the distribution of publications by author rank. As the distribution plot according to author rank shows, the publications tend to be authored by those of a higher rank. 185 of the 390 publications (47.4%) were authored by retired 4 star general or flag officers (Generals or Admirals). Generals or Admirals account for 276 of the total 390 publications (70.8%). The fact that high ranking officers tend to publish in the newspapers examine is unsurprising, and confirms the competitive nature of publishing an opinion piece in one of the premier newspapers in the nation. At the same time, there are enough authors who are not retired generals or admirals (29.2%) to show that those who do not attain the rank of general or admiral still have a respectable opportunity to publish their views in the examined sources. 

```{r rank-pubs, include=TRUE, echo=FALSE, warning=FALSE, fig.cap="Opinion Pieces authored by Retired Military Officers Broken Down by Author Rank, 1979-2020", fig.height=4.5, fig.width=6}
colplot1 <- ggplot(data=rankdistrib_scaled, aes(x=Rank, y=n)) + labs(x="Author Pay Grade / Rank", y="Number of Publications") + geom_col() + geom_text(aes(label=n), position=position_dodge(width=.9), vjust=-.25)+ theme(axis.text.x = element_text(angle=90)) + scale_x_discrete(labels=wrap_format(18))

colplot1

```


The distribution of the data according to newspaper source is further graphed in Figure \@ref(fig:source-pubs), below. The data shows that the overwhelming number (379, or 97.2%) of opinion publications were authored by retired military officers in the _New York Times_, _Washington Post_, or _Wall Street Journal._ Only 11 of the 390 publications (2.8%) were authored in either the _USA Today_ or the _Los Angeles Times_.  This is not surprising. Not only are these three newspapers among the most prestigious and most circulated across the world, but they are based in the nation's capital and New York City, where think tanks, international organizations, and government offices are highly concentrated.

```{r source-pubs, include=TRUE, echo=FALSE, warning=FALSE, fig.cap= "Opinion Pieces Authored by Retired Military Officers Broken Down by Newspaper Source, 1979-2020", fig.width=4, fig.height=3}

colplot2 <- ggplot(data=countbysource, aes(x=Source, y=n)) + labs(x="Source", y="Number of Publications") + geom_col() + geom_text(aes(label=n), position=position_dodge(width=.9), vjust=-.25) 

colplot2

```


Finally, Figure \@ref(fig:year-count), below, displays the annual count of publications authored by retired military officers across the dataset. The smoothed trend line indicates that the average yearly count of publications rose steadily from 1979 until the early 2000s, at which time, a decline occurred until approximately 2008. From 2008 until the present day, the average yearly count of opinion publications rose heavily during the Obama administration years before briefly falling, but then resuming the high count during the Trump administration. The data also shows a marked increase occurring in the latter half of the 1990s under President Bill Clinton's tenure in office. 

```{r year-count, fig.align="center", warning=FALSE, error=FALSE, message=FALSE, include=TRUE, echo=FALSE, fig.cap="Annual Number of Op-Eds and/or Letters to the Editor Authored by Retired Military Officers and Published in Major US Newspapers (WSJ, NYT, WaPo, USA Today, LA Times), 1979-2020", fig.width=6, fig.height=4.5}
  
# construct a scatter plot with line and smoothing

ggplot(data=adjdataset, aes(x=PubYr, y=Yrcount)) + geom_point() + geom_line() + geom_smooth(method="loess") + labs(
  x="Year", 
  y="yearly count", 
  caption="Note: Data as of 1 December 2020"
  ) + scale_x_continuous(limits=c(1979, 2020)) + scale_y_continuous(limits=c(-7, 28)) + 
geom_vline(xintercept = 1981, size=.2, colour="tomato") + 
  geom_vline(xintercept=1989, size=.2, colour="tomato") + 
  geom_vline(xintercept=1993, size=.2, colour="dodgerblue") + geom_vline(xintercept=2001, size=.2, colour="tomato") + 
  geom_vline(xintercept=2009, size=.2, colour="dodgerblue") +
  geom_vline(xintercept=2017, size=.2, colour="tomato") +
  geom_text(aes(x=1981, label="Reagan",y=10), colour="tomato", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=1989, label="Bush 41",y=15), colour="tomato", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=1993, label="Clinton",y=20), colour="dodgerblue", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=2001, label="Bush 43",y=-2), colour="tomato", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=2009, label="Obama",y=-2), colour="dodgerblue", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=2017, label="Trump",y=-2), colour="tomato", angle=90, vjust=1, text=element_text(size=9)) 
  
```

From a macro perspective, it is interesting to consider the breadth of the events that have occurred during the past four decades. In terms of external threats and wars, this period includes the final years of the Cold War against the Soviet Union (through 1991), the Gulf War in Iraq (1990-1991), operations in the Balkans (primarily 1993-1995), and the post-9/11 wars in Iraq (2003-present day) and Afghanistan (2001-present day), as well as a host of smaller but important military operations in Grenada (1983), Panama (1989), Somalia (1992-1993), and Haiti (1994-1995). The military has also undergone significant social and institutional changes in the time period examined. For example, the late 1970s came just years after the US Government ended the draft (1973). The debates over the "Don't Ask, Don't Tell" policies, instituted in 1994 before being repealed in 2011), are captured in the data set, along with numerous other debates, including those regarding women serving in combat arms branches of the military. 

## What Are Retired Military Officers Writing About?

The first critical step in the analysis is to determine what topics retired military officers have been addressing through their published commentary. This requires classifying each of the observations into one of several possible topic types. Because historical and contextual circumstances change over time, however, the possible range of topics should be broad enough to classify every observation yet narrow enough to generate meaningful comparison across topics and across time. Therefore, I develop a categorical variable for topic that can take one of eight possible values according to the scheme shown in Table \@ref(tab:topic-scheme), below. 

```{r topic-scheme, include=TRUE, echo=FALSE}
topic_categories <- tribble(~Topic, ~Description, ~Examples,
   'Foreign Policy','If the publication primarily addresses American Foreign Policy or overarching National Security Policy', 'A Disaster Puts Putin in a Bind, by William Odom; The Test Ban Solution, by John Shalikashvili',
  'War-Fighting and Operations', 'If the publication primarily addresses an ongoing conflict or a central aspect related to the ability of the military to fight and win on the battlefield', 'A Month in Macedonia Will Not Be Enough, by Wesley Clark; A Plan To Save Iraq from ISIS and Iran, by Jack Keane',
  'Social Policy', 'If the publication primarily addresses questions or issues related to who should serve, and/or why', 'Banning Transgender Troops Only Hurts Us, by Mike Mullen; US Military Readiness Requires the Draft, by William Westmoreland',
  'Domestic Policy', 'If the publication primarily addresses issues that do not have an immediate and direct impact on the military, or that describe the general operating atmosphere and political conditions in Washington D.C.', 'The Travel Ban Hurts American Spies - and America, by Michael Hayden; Our Republic is Under Attack from the President, by William McRaven',
  'Budget/Weapons/Troops', 'If the publication primarily addresses issues such as defense budgets, a particular weapons system, or troop benefits/pay', 'Washington Tightwads are Creating a Hollow Military, by Gordon Sullivan; We Need More Troops, by Barry McCaffrey',
  'Support', 'If the publication primarily makes general or specific calls of support for the military or for particular military or defense figures', 'In Defense of Donald Rumsfeld, by John Crosby and Thomas McInerney; From Forward Operating Base to Boardroom, by Stanley McChrystal',
  'Civil-Military Relations', 'If the publication primarily addresses how military and civilians can better interact and relate to each other, or discusses how organizations and departments can be best structured to offer advice to civilian leaders, issues of strategy, etc.', 'Who Decides When We Go To War, by David Barno; Military Leaders Do Not Belong at Political Conventions, by Martin Dempsey',
  'Service Culture', 'If the publication primarily addresses an ongoing issue - positive or negative - that is occurring within the military itself and that bears on the internal culture of the military', 'Yet Another Insult to Women, by Wayne Johnson; The Navys Blues, by David Evans',)


  kable(topic_categories, booktabs=T, caption="Categorical Values of the Topic of Retired Military Officer Public Commentary") %>%
  kable_styling(full_width = F, font_size=11, latex_options="striped") %>% 
  column_spec(1, width="10em") %>% 
  column_spec(2:3, width="16em") 
```

Classifying each observation into one of the eight possible topics requires some judgement on the part of the researcher. After all, some observations address multiple topics. Ultimately, reading the content of each publication enables the researcher to identify a publication's thesis statement, which helps answer the fundamental question, "what is the central argument made in this letter or editorial?" 

Table \@ref(tab:pub-subject), below, displays the breakdown of observations by topic category. The topics of _War-fighting and Operations_ and _Foreign policy_ account for a total of 229 of the 390 observations (59% of the total sample). The categories of _Support_, _Domestic Policy_, and _Budget/Weapons/Troops_ account for a total of 120 observations, or 30.5% of the sample. Finally, the topics of _Social Policy_, _Civil-military Relations_, and _Service Culture_ combined account for the remaining 41 observations, or 10.5% of the sample. 

```{r pub-subject, fig.align="center", include=TRUE, echo=FALSE, warning=FALSE}
countbytype %>%
  kable(format="latex", caption="Publication Breakdown by Subject") %>%
  kable_styling(latex_options = "striped", font_size=10) %>% 
  row_spec(0, bold=T)
```

The breakdown of observations by topic type is not necessarily surprising. We expect retired military officers to address matters that relate to _War-fighting and Operations_ and _Foreign Policy_, as these topics directly relate to the military expertise and experiences of retired military leaders. However, if we isolate the instances in which retired military officers primarily address topics related to _Domestic Politics_, a comparison over time becomes somewhat noteworthy. 

Figure \@ref(fig:dompol-pubs), below, plots the proportion of annual publications authored by retired military leaders that address topics related to _Domestic Politics_ since 1979. Three interesting trends stand out. The first is that with the exception of an outlier year (1983), no publications authored by retired military officers addressed primarily topics related to _Domestic Politics_ until 1999. Nearly half of the time duration of the sample are years in which retired military officers nearly shun topics of _Domestic Politics_ altogether. The second interesting trend, as shown in the trend line in blue, is that retired military officers appear to have become increasingly active in addressing topics of _Domestic Politics_ during the second half of President Obama's term in office. The third interesting trend is that retired military officers increasingly addressed topics related to domestic politics under President Trump's term in office.  

```{r, include=FALSE, echo=FALSE}
## Create several new variables.  

# The first is a variable called "case-count" that counts the number of times by publishing year an #article published that calls for legislative action, criticizes or is #insubordinate, or is blatantly #partisan. 

#The second variable, prop_yr, is simply a 
# proportion of annual publications where these conditions hold true.

# The third variable is called dom_pol_count which counts the number of publications
# by publishing year that are written on topics of domestic policy, and the fourth 
# variable is a proportion of this type of article by year. 

adjdataset <- adjdataset %>% group_by(PubYr) %>% 
  mutate(
    case_count=sum(Leg_Act ==1 | Crit_Insub==1 | Endorse_Partisan==1), 
    prop_yr=(case_count/Yrcount),
    dom_pol_count=sum(Topic=="DomesticPolicy"),
    dom_pol_prop=(dom_pol_count/Yrcount)
  ) %>% view()
```

```{r dompol-pubs, include=TRUE, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, fig.cap="Proportion of Opinion Publications authored by Retired Military Officers Which Primarily Address Domestic Politics, 1979-2020", fig.width=6, fig.height=4.5}
ggplot(data=adjdataset, mapping=aes(x=PubYr, y=dom_pol_prop)) + geom_point() + geom_smooth() + labs(x="Year", y="Proportion", color=NULL)+ geom_vline(xintercept = 1981, size=.2, colour="tomato")+
  geom_vline(xintercept=1989, size=.2, colour="tomato") + 
  geom_vline(xintercept=1993, size=.2, colour="dodgerblue") + geom_vline(xintercept=2001, size=.2, colour="tomato") + 
  geom_vline(xintercept=2009, size=.2, colour="dodgerblue") +
  geom_vline(xintercept=2017, size=.2, colour="tomato") +
  geom_text(aes(x=1981, label="Reagan",y=.25), colour="tomato", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=1989, label="Bush 41",y=.25), colour="tomato", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=1993, label="Clinton",y=.25), colour="dodgerblue", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=2001, label="Bush 43",y=.3), colour="tomato", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=2009, label="Obama",y=.3), colour="dodgerblue", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=2017, label="Trump",y=.4), colour="tomato", angle=90, vjust=1, text=element_text(size=9))  
  
```

In summary, the variation in retired military officers addressing matters of _Domestic Politics_ in public commentary over time leads to the conclusion that although retired military officers accelerated this behavior under President Trump, the behavior itself started much earlier. Retired military actors began addressing domestic political matters through published opinion commentary late in the Clinton Administration, and increased this behavior under the administrations of Presidents George W. Bush and Barack Obama. Having presented the data, to include a topical analysis of retired military officer commentary, I now proceed to the more substantive step of determining whether the published opinion commentary pushes against any of the three fundamental norms of civil-military relations, and if so, why. 

## Second Layer Analysis - Are Retired Military Officers Challenging Fundamental Norms?

To measure adherence to the norms of civilian control, I develop several dichotomous indicator variables that correspond to each norm. The first, which is intended to capture instances of retired military officers pushing against or perhaps violating the principle of civilian control, I classify those observations in which the author heavily and blatantly criticizes a civilian leader. Because the observations are all opinion pieces, and thus somewhat argumentative in nature, the goal is not to simply identify publications which point out shortcomings of policy, or make recommendations, as we would expect this type of content in any published commentary. However, blatant and express criticism of a civilian official, and particularly, the Commander in Chief, is not a behavior that military officers, active or retired, have traditionally taken lightly. To be clear, public criticism of a civilian leader at the hands of a retired military officer, even harsh criticism, is not synonymous with the act of plotting and undertaking a coup, but the frequency and content of the criticism offered by retired military officers may nonetheless reveals clues about the primacy of the principle of civilian control in governing their behavior.   

A second indicator is whether a publication explicitly endorses or rebukes a party platform explicitly, to include the endorsement or rebuke of candidates running for office. This indicator measures instances in which retired military actors challenge the principle of non-partisanship of the military. I therefore classify any publications authored by retired military officers which explicitly endorse or rebuke a political party, portions of the party platform, or candidates running for office.

The third indicator corresponds to instances in which retired military actors address topics that are indirectly, tangentially, or altogether unrelated to the military or defense policy. The intent of this indicator is to capture instances in which retired military officers push the boundaries of the norm of "non-interference" into realms of state policy by weighing in on policy areas for which military expertise may not be relevant, appropriate, or applicable. Two points of clarification about this particular indicator are worth underscoring, however. 

The first is that the roles, responsibilities, and missions undertaken by the military have significantly expanded over the past several decades, both overseas and domestically [@brooks_how_2016]. The military, for better or for worse, now plays a prominent role not only in fighting wars, but also implementing other aspects of America's expansive foreign policy, while also providing support to civilian authorities domestically in responding to various types of natural disasters and civil unrest.^[The military's role during the COVID-19 Pandemic has only magnified these tendencies. For example, Operation Warp Speed, the logistics operation designed by the Trump Administration to help manufacture, produce, and deliver a vaccine for COVID-19, has heavily thrust the military into conversation and relationships with the healthcare community, private corporations, and transportation companies. Operation Warp Speed is run, moreover, by a four-star Army General, Gustave Perna. Some civil-military relations scholars have expressed concerns, shared by this author, that placing the military in charge of this operation has had, at times, the unintended consequences of the military being placed in a position to make decisions (related to the distribution of the vaccine) that from a normative perspective should have been made by civilian leaders, not military officers.] 

A second point of clarification, one which relates to the point just made, is that it is theoretically possible to connect many issues - perhaps almost any "hot-button" political issue - to the nation's "security." Problems such as racism, extremism, climate change, polarization, immigration, the lack of civil discourse in America, and many more, are problems that to one degree or another in fact do pose a potential danger to the world in which we live, broadly defined. A critical question then in coding for an indicator of addressing topics which may fall outside of the realm of military expertise is, "should the military be actors who address this issue?" Ultimately, this is a normative question open to a high degree of personal opinion. Therefore, I only code issues or topics that strike me as considerably outside or tangential to relevant military expertise. It is also important to note that some retired military officers do in fact go on to serve in political administrations, or run for office and serve as politicians themselves. For this reason, it is important to control whether a retired military officer wrote a publication while serving or having previously served in a role that would legitimately correspond to the particular officer advocating on behalf of a certain topic or issue area, in which case, dialogue that challenges or violates a normative standard of behavior by military officers would not apply in the same sense. For example, we would expect that Senator John McCain, who retired from the Navy after a career as an aviator, would make partisan statements in his role as an elected official, and thus, observations that challenge normative standards made by Senator McCain are removed from the analysis to follow. 

Unlike the categorical topic variable which can only take on a unique value, I allow for the possibility that a single observation takes on a positive value for more than one of these dichotomous variables, to reflect an instance in which a publication challenges or violates more than one of the norms. Figure \@ref(fig:norms-challenged), below, plots the number of challenges to each of the three fundamental norms, as well as the total count of annual challenges to the foundational norms of civil-military relations, made by retired military officers in published opinion commentary from 1979-2020. 

```{r norms-challenged, include=FALSE, include=TRUE, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="Instances of Retired Military Officers Challenging Norms through Public Commentary, 1979-2020"}

# create a count, by year, of the observations which broke one of the three norms:
# criticized politicians, strayed into strange topic areas, or a norm of non-partisanship
violationdf <- adjdataset %>% group_by(PubYr) %>% filter(Role != 1) %>% select(Author, PubYr, Title, AuthRank, Leg_Act, Crit_Insub, Issue_Area, Endorse_Partisan) %>% mutate(
  Crit_Insub_Count = sum(Crit_Insub),
  Issue_Area_Count = sum(Issue_Area),
  Endorse_Partisan_Count=sum(Endorse_Partisan),
  Total_Count = Issue_Area_Count+Crit_Insub_Count+Endorse_Partisan_Count) %>% view()

# create a data frame but summarize the counts of violations by type per publishing year
violationdf_by_year <- violationdf %>% select(PubYr, Crit_Insub_Count, Endorse_Partisan_Count, Issue_Area_Count, Total_Count) %>% group_by(PubYr, Crit_Insub_Count, Endorse_Partisan_Count, Issue_Area_Count, Total_Count) %>% summarise() %>% view()

#using the pivot_longer function, turn this DF into a graphaple data frame
violationdf_by_year_graph <- violationdf_by_year %>% pivot_longer(Crit_Insub_Count:Total_Count, names_to = "Violation_Type", values_to = "Violation_Count") %>% view()

## now graph this puppy using ggplot!

violationdf_by_year_graph %>% ggplot(aes(x=PubYr, y=Violation_Count)) + geom_point(aes(shape=Violation_Type, color=Violation_Type), position=position_jitter(width=.1, height=.1), alpha=.6, size=5) + scale_color_manual(values=c("tomato","dodgerblue2", "seagreen", "black"))+ labs(x="Publishing Year", y="Number of Violations By Type") + theme(legend.position="bottom", axis.text.x=element_text(angle=90)) + scale_x_continuous(breaks=seq(1979, 2020, 1))

```

Figure \@ref(fig:norms-challenged) reveals similar trends as the analysis of topics addressed by retired military officers examined in the previous section. For the first ten years of the data, from 1979-1988, there were no instances of retired military officer commentary expressly challenging the norms of civil-military relations. From 1989 throughout the 1990s, the frequency and relative number of instances in which retired military officer commentary challenged normative standards of civil-military relations slowly increased. All of these challenges or violations were to the norm of civilian control, and manifested themselves as retired military officers criticized former President Clinton for his leadership during a number of crises overseas, including in the Balkans and Somalia. In these crises, the US military was engaged in combat operations. Even if one does not condone the criticism given by these officers, one can reasonably understand that their experiences as retired military officers directly translated into many of the events addressed in their commentary.

From 2000 until 2010, Figure \@ref(fig:norms-challenged) shows that the total number of challenges to normative standards of retired military officer behavior exhibited in published commentary increased, particularly in the period from 2004 - 2006. During this period, several newly-retired military officers endorsed the two Presidential Candidates for office, the incumbent, George W. Bush, and Senator John Kerry, Bush's opponent, violating the normative standard of non-partisanship of the military institution. Additionally, the "Revolt of the Generals," previously discussed, occurred during this period - an event reflected in the challenges denoted to the norm of civilian control during this period. 

From 2007 through 2012, as Figure \@ref(fig:norms-challenged) shows, retired military officers did not obviously challenge normative standards through the opinion commentary they published. From 2013 - 2020, however, the collective level of retired military officer commentary pushing against normative standards has risen dramatically, and in ways which differ from the preceding period. Retired military officers began to address topics that can reasonably be considered outside of the relevant military expertise of military actors, thus challenging the norm of "non-interference" of the military into certain policy areas. The volume of criticism for civilian officials also increased significantly, reaching a crescendo under the administration of Donald Trump, but as the plot also shows, this behavior had been well established before Trump's Presidency. Finally, challenges to the norm of non-partisanship have also been evident during the past decade through published retired officer opinion commentary. 

For a visual comparison that shows the proportion of the annual number of published opinion commentaries which challenged one of the three norms, please see Figure \@ref(fig:prop-violate-norms), below. Viewing the proportion instead of the total number of annual opinion commentary published by retired military officers informs a comparison across time, as the overall amount of opinion commentary has increased over the past several decades.

```{r prop-violate-norms, include = TRUE, echo=FALSE, warning = FALSE, error=FALSE, message=FALSE, fig.cap="Proportion of Op-Eds Which Either Criticizes Civilian Leader, Adopts Partisan Stances, or Advocates for Topics Not Relevant to Military Expertise, 1979-2020"}

ggplot(data=adjdataset, mapping=aes(x=PubYr, y=prop_yr)) + geom_point() + geom_smooth() + labs(x="Year", y="Annual Proportion", color=NULL) + geom_vline(xintercept = 1981, size=.2, colour="tomato") + 
  geom_vline(xintercept=1989, size=.2, colour="tomato") + 
  geom_vline(xintercept=1993, size=.2, colour="dodgerblue") + geom_vline(xintercept=2001, size=.2, colour="tomato") + 
  geom_vline(xintercept=2009, size=.2, colour="dodgerblue") +
  geom_vline(xintercept=2017, size=.2, colour="tomato") +
  geom_text(aes(x=1981, label="Reagan",y=.2), colour="tomato", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=1989, label="Bush 41",y=.2), colour="tomato", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=1993, label="Clinton",y=.4), colour="dodgerblue", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=2001, label="Bush 43",y=.3), colour="tomato", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=2009, label="Obama",y=.4), colour="dodgerblue", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=2017, label="Trump",y=.05), colour="tomato", angle=90, vjust=1, text=element_text(size=9))  
```

Collectively, Figure \@ref(fig:dompol-pubs), Figure \@ref(fig:norms-challenged), and Figure \@ref(fig:prop-violate-norms) reveal that retired military officers are addressing domestic political topics relative to other topics and challenging each of the three normative standards in doing so at a greater rate than they did in previous decades. Having established that these trends exist, the next task is to determine what roles, if any, polarization and military prestige are playing in driving this variation.

## Explaining A More Assertive Retired Military Officer Corps 

Fortunately, measurements for both of the suspected independent variables, polarization and prestige, are available over the time frame of the data, albeit with important caveats. I include two measures of polarization, following a majority of the American Politics literature. The first is a measure of DW-NOMINATE (dynamic-weight, nominal three step estimation) scores, which use roll-call voting data to assign members of Congress a score on the liberal-conservative dimension, ranging from -1 (extremely liberal) to 1 (extremely conservative) [@jeffrey_b_lewis_voteview_2020]. A second measurement of polarization measures affective polarization, which relies on "feeling-thermometer" data from the American National Election Study (ANES) to gauge how partisans who identify with one political party generally feel about the other political party [@iyengar_affect_2012; @iyengar_strengthening_2018; @iyengar_origins_2019].^[Data on Congressional Polarization extends further back in time, while measurements of "Affective Polarization" go back only for the past several decades. Both measures indicate that polarization has worsened over the past several recent decades.] 

```{r, include = FALSE, warning=FALSE, echo=FALSE}

# Create a variable, quest_prof, that shows all commentary in which the variables 
# of Issue Area, Crit_Insub, or Endorse_Partisan was equal to 1
## AND in which the author was NOT serving in a role that may have made norm 
## violation understandable, i.e. Role !=1

quest_prof <- adjdataset %>% arrange(PubYr) %>% filter(Role != 1) %>% 
  filter(Crit_Insub== 1 | Endorse_Partisan ==1 | Issue_Area== 1)  %>% 
  select (Author, Title, Source, PubYr, Topic, Crit_Insub, Endorse_Partisan, Issue_Area) %>% 
  view()

# The same as above, but shows all authors, to include those serving in Roles where criticism
# was perhaps warranted. 

adjdataset %>% arrange(PubYr) %>% 
  filter(Issue_Area ==1 | Crit_Insub== 1 | Endorse_Partisan ==1)  %>% 
  rename(Crit=Crit_Insub, Endorse=Endorse_Partisan, Interference=Issue_Area
  ) %>% 
  select (Author, Title, Source, PubYr, Topic, Crit, Endorse, Interference, Role) %>% 
  view ()

# Create a df called regdata in which I 
# Create several new variables, including crit_part_interfere and crit_part that take 
# the value of '1' if certain 
# conditions are met, including if the variable "Role" ==0.  
# Also scale the data for congressional polarization by multiplying by 100
# This will come in handy for the glm regressions that will occur later. 

regdata <- adjdataset %>% mutate(
    crit_part_interfere=case_when(
      (Role==0) & (Crit_Insub ==1 | Endorse_Partisan==1 | Issue_Area ==1) ~ 1,
      (Role==1) | (Crit_Insub != 1 & Endorse_Partisan !=1 & Issue_Area != 1) ~ 0
    ),
    crit_part=case_when(
      (Role==0) & (Crit_Insub ==1 | Endorse_Partisan==1) ~ 1,
      (Role==1) | (Crit_Insub != 1 & Endorse_Partisan != 1) ~ 0
    ),
  House_distance_dim1=abs(House_Dem_mean_nom1-House_Rep_mean_nom1)*100,
  Sen_distance_dim1=abs(Sen_Dem_mean_nom1-Sen_Rep_mean_nom1)*100,
  House_distance_dim2=abs(House_Dem_mean_nom2-House_Rep_mean_nom2)*100,
  Sen_distance_dim2=abs(Sen_Dem_mean_nom2-Sen_Rep_mean_nom2)*100,
  aff_polar=dem_diff_aff+`rep-diff_aff`
  ) %>% view()

# create a variable decade
authorcalcs <- regdata %>% 
  mutate(
  decade =case_when(
    (PubYr < 1980) ~ "1970s",
    (PubYr >= 1981 & PubYr < 1991) ~ "1980s",
    (PubYr >= 1991 & PubYr < 2001) ~ "1990s",
    (PubYr >= 2001 & PubYr < 2011) ~ "2000s",
    (PubYr >= 2011 & PubYr < 2021) ~ "2010s"
  )
) %>% select(Author:Service, Title:Topic, crit_part_interfere, crit_part, decade) %>% print()

#count the number of publications by GO/FO author in each decade
authorcalcs %>% group_by(decade) %>% filter(AuthRank >=7) %>% select(Author) %>% count(Author) %>% view()
```

I include a proxy for military prestige using Gallup public opinion data which measures the level of trust that the public places in various institutions in America, such as the Presidency, Congress, Church, and the Police. Since the late 1970s, the credibility of the American military as an institution has gradually increased [@noauthor_confidence_2020]. For instance, in 1979, 54% of those surveyed expressed either a "great deal" or "quite a lot" of confidence in the military, while in 2020, 72% expressed the same [@noauthor_confidence_2020]. Trends in both types of polarization, along with military prestige, are plotted in Figure \@ref(fig:trends) below.

```{r trends, include=TRUE, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, fig.cap="Polarization Levels in the US House of Representatives, Affective Polarization Levels Among the US Electorate, and American Confidence in the Military, 1979-2020", fig.height=4.5, fig.width=6}

polar_and_cred <- melt(subset(regdata, select=c(PubYr, House_distance_dim1, aff_polar_alt, Inst_Cred_Gallup)), id.var="PubYr")

ggplot(polar_and_cred, aes(x=PubYr, y=value)) + geom_line(aes(color=variable)) + facet_grid(variable~., scales="free_y") + theme(legend.position = "none")
```

```{r logit analysis, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
# check for separation (complete or quasi-complete)
regdata %>% 
  filter(Issue_Area == 1 | Crit_Insub ==1 | Endorse_Partisan == 1) %>% group_by(Source) %>% 
  view()

# logit regression where the DV is first, values of crit_part_interfere and
# then, where the DV is crit_part
# model 1 

mod1 <- glm(crit_part_interfere ~ Inst_Cred_Gallup + House_distance_dim1 + AuthRank,
    family=binomial(link="logit"),
    data=regdata
    )
summary(mod1)
tidy(mod1)
glance(mod1)

# model 2 - logit with different DV

mod2 <- glm(crit_part ~ Inst_Cred_Gallup + House_distance_dim1 + AuthRank,
    family=binomial(link="logit"),
    data=regdata
    )

summary(mod2)
tidy(mod2)
glance(mod2)

#model 3 - logit using aff polar 
mod3 <- glm(crit_part_interfere ~ Inst_Cred_Gallup + aff_polar_alt + AuthRank,
    family=binomial(link="logit"),
    data=regdata
    )

summary(mod3)

#model 4 - logit using aff polar with different DV

mod4 <- glm(crit_part ~ Inst_Cred_Gallup + aff_polar_alt + AuthRank,
    family=binomial(link="logit"),
    data=regdata
    )

summary(mod4)

# models 5-8 include Source as an IV, but the standard error estimates are too large
# likely because there is not enough variation within Sources (LA Times is the problem)

mod5 <- glm(crit_part_interfere ~ Inst_Cred_Gallup + House_distance_dim1 + AuthRank + Source,
    family=binomial(link="logit"),
    data=regdata
    )
summary (mod5)


mod6 <- glm(crit_part ~ Inst_Cred_Gallup + House_distance_dim1 + AuthRank + Source,
    family=binomial(link="logit"),
    data=regdata
    )
summary (mod6)

mod7 <- glm(crit_part_interfere ~ Inst_Cred_Gallup + aff_polar_alt + AuthRank + Source,
    family=binomial(link="logit"),
    data=regdata
    )
summary(mod7)

mod8 <- glm(crit_part ~ Inst_Cred_Gallup + aff_polar_alt + AuthRank + Source,
    family=binomial(link="logit"),
    data=regdata
    )
summary(mod8)

# I am now going to drop observations from the LA Times and USA Today  which will enable me to obtain some #better, more #precise estimates of the data. 

xtabs(crit_part_interfere ~ Source, regdata)
xtabs(crit_part ~ Source, regdata)
slimdata <- regdata %>% filter(Source != "LA Times" & Source !="USA Today") %>% print()
xtabs(crit_part_interfere ~ Source, slimdata)
xtabs(crit_part ~ Source, slimdata)

slimdata %>% view()

slimdata %>% filter(crit_part_interfere ==1) %>% group_by(Author) %>% count(Author) %>% arrange(n) %>% view() 

slimdata %>% filter(crit_part ==1) %>% group_by(Author) %>% count(Author) %>%  arrange(n) %>% view ()

# Now run a straght logit GLM including Source as an IV
# Them conduct model with robust standard errors at the author level
# Then conduct a mixed model where we include author random effects

# DV is Crit_part_interfere, house_distance_dim1 is IV
mod9 <- glm(crit_part_interfere ~ Inst_Cred_Gallup + House_distance_dim1 + AuthRank + Source, family=binomial(link="logit"), data=slimdata)
summary(mod9)

mod9robust <- robustify(mod9, cluster=Author)
summary(mod9robust)

mod9mixed <- glmer(crit_part_interfere ~ Inst_Cred_Gallup + House_distance_dim1 + AuthRank + Source + (1 | Author), family=binomial, data=slimdata)
summary(mod9mixed)

# DV is crit_part_interfere, aff_polar_alt is IV

mod11 <- glm(crit_part_interfere ~ Inst_Cred_Gallup + aff_polar_alt + AuthRank + Source, family=binomial(link="logit"), data=slimdata)
summary(mod11)

mod11robust <- robustify(mod11, cluster=Author)
summary(mod11robust)

mod11mixed <- glmer(crit_part_interfere ~ Inst_Cred_Gallup + aff_polar_alt + AuthRank + Source + (1 | Author), family=binomial, data=slimdata)
summary(mod11mixed)

# DV is crit_part, IV is house_distance_dim1

mod10 <- glm(crit_part ~ Inst_Cred_Gallup + House_distance_dim1 + AuthRank + Source, family=binomial(link="logit"), data=slimdata)
summary(mod10)

mod10robust <- robustify(mod10, cluster=Author)
summary(mod10robust)

mod10mixed <- glmer(crit_part ~ Inst_Cred_Gallup + House_distance_dim1 + AuthRank + Source + (1 | Author), family=binomial, data=slimdata)
summary(mod10mixed)

# DV is crit_part, IV is aff_polar_alt

mod12 <- glm(crit_part ~ Inst_Cred_Gallup + aff_polar_alt + AuthRank + Source, family=binomial(link="logit"), data=slimdata)
summary(mod12)

mod12robust <- robustify(mod12, cluster=Author)
summary(mod12robust)

mod12mixed <- glmer(crit_part ~ Inst_Cred_Gallup + aff_polar_alt + AuthRank + Source + (1 | Author), family=binomial, data=slimdata)
summary(mod12mixed)

```

The dependent variable is a binary indicator of whether the opinion piece challenged one of the three norms of civil-military relations by strongly criticizing a civilian leader, adopting an expressly partisan position, or addressing a topic that falls outside of policy areas that are directly relevant to military expertise. I control for the rank of the author as well as the source in which an opinion piece was published to control for newspaper sources which may have adopted a more-pronounced ideological bent along the liberal - conservative spectrum as polarization has worsened [for an informative view of where media outlets are along such a spectrum, see @noauthor_allsides_nodate]. Due to complete separation between the dependent variable involving observations published in _The Los Angeles Times_ and _USA Today_, these 11 observations are dropped from the regression analysis.^[I include several types of models, including a regular logit model, a logit model that clusters standard errors at the author level, and a mixed model that introduces random effects at the author level. The intent of the models which cluster standard errors or introduce random effects at the author level is to determine the degree to which the data can distinguish if the individual authors themselves are an important determinant of whether an opinion piece challenges one of the norms of civil-military relations. The logit model without clustered errors at the author level assumes that authors have no distinguishing characteristics that would incline them towards behavior that would result in the dependent variable taking on a value of 1, whereas both the mixed model with author random effects and the logit model that clusters errors at the author level allows for the possibility that individual author characteristics may explain the dependent variable in important ways that the model does not otherwise account for.] The results are displayed in Table \@ref(tab:regression1) below. 

\singlespace
```{r Regression Results, include=TRUE, echo=FALSE, results="asis", warning=FALSE, message=FALSE}
stargazer(mod9, mod9robust, mod9mixed, mod11, mod11robust, mod11mixed,
          type="latex",
          label="tab:regression1",
          title="Regression Results", 
          covariate.labels = c('Mil Cred', 'Cong. Polar', 'Aff. Polar', 'Rank', 'SourceWaPo'),
          dep.var.caption = "Odds (log) of Opinion Commentary Authored by Retired Military Officers",
          dep.var.labels = "Challenging Normative Standards of Military Behavior",
          model.names=TRUE,
          header=FALSE,
          column.sep.width = "-15pt",
          align=TRUE)
         
```
\doublespace
The results shown in Table \@ref(tab:regression1) indicate that across all model specifications, the coefficient for polarization, regardless of how it is measured, is statistically significant.  Substantively, a coefficient of .118 for Congressional Polarization in Models 1 and 2 indicate that a one unit increase in the mean ideological distance between political parties in Congress is associated with a 12.5 percent increase in the odds that a commentary publication authored by a retired military officer challenges or violates one of the normative standards of military behavior - the principle of civilian control, non-partisanship, or "non-interference" of the military into policy realms that deviate from military relevant military expertise. Similarly, the coefficient for affective polarization in Models 4 and 5 of .045 indicate that for every one unit increase the level of disdain for partisans to feel for those members who affiliate with the other side, there is a 4.6 percent increase in the odds that a retired military officer violates or pushes up against a normative standard in the opinion commentary he or she publishes.  

While the coefficients for the different measurements of polarization are of different size, scaling these measures to understand how each measure shifts over time is important. For instance, within the House of Representatives in the 112th Congress, which began in January 2011, polarization levels were nearly six points higher than the level that existed in the 111th Congress. Such a drastic change in the polarization level within Congress would predict a roughly 75 percent increase in the odds of an opinion publication authored by a retired military officer strongly criticizing a civilian leader, taking a clear partisan stance, or addressing a topic that falls outside of relevant military experience. Regarding changes in the levels of affective polarization, data from the 2012 ANES indicate that the aggregate level of affective polarization increased by just over 16 points from the previous Presidential Election year (2008), which would correspond to a 73.6 percent increase in the odds of a publication authored by a retired military officer violating one of the three foundational norms of civil-military relations. Thus, the coefficients for the respective levels of polarization turn out to be substantively very similar, which increases the confidence that the theory underlying these predictions works in the ways in which it is anticipated to. 

In terms of the other coefficients besides measurements of polarization, two points are worth underscoring. The first is that the coefficients for the estimated impacts of military prestige are negative and not statistically significant across all six models. Relative to the statistical significance and positive direction of the coefficients for polarization, we conclude that changes in military credibility have less explanatory power, if any, in whether a publication authored by retired military officers fails to adhere to norms of civilian control, non-partisanship, and non-interference. The second interesting point is that the coefficients for the author's rank and the source in which a commentary piece is published are both statistically insignificant across all of the models. These results refute alternative explanations which suggest a retired military officer's rank is a driving factor in leading an author to challenge a normative standard in the opinion material he or she publishes, or that a particular newspaper source is significantly more likely than the others examined to publish an opinion piece which challenges or violates normative standards. 

Finally, it is important to note that in Models 3 and 6, which are both mixed models that include random effects at the author level in order to allow for the possibility that individual authors are driving results in the dependent variable rather than other variables, the coefficients for the respective measurements of polarization increase relative to other model types, but also drop in statistical significance. These models, therefore, lead to the conclusion that as polarization increases, it is having a greater impact on certain individual authors, who in turn challenging normative standards through the opinion publications which they author. 

## Predicted Probability

Using the regression results from Table \@ref(tab:regression1) to transform the results of our logistic regressions into predicted probabilities, we can plot the predicted probability of retired military opinion commentary violating normative standards as the level of polarization changes. Using the results of the logistic regression from Model 4, which includes a measure of affective polarization, the source is set for _The New York Times_, while values of the institutional credibility of the military and author rank are held fixed at their respective means (71.1 and 8.39). These predicted probabilities are graphed in Figure \@ref(fig:pred-prob-ANES) below. 

```{r predicted probabilities, include=FALSE, warning=FALSE, message=FALSE, error=FALSE}

# Use Model 11 (Logit) from Above, where IV is aff_polar, and there are no author random Effects
# Hold values of Author Rank and Institutional Credibility at their mean
# Vary the Level of Affective Polarization from its minimum value to an arbitrary 
# level of 120, which is higher than it currently is but not theoretically impossible.  

predictiondata <- tibble(aff_polar_alt=seq(min(slimdata$aff_polar_alt), 120, by=1), Inst_Cred_Gallup = mean(slimdata$Inst_Cred_Gallup), Source="NYT", AuthRank=mean(slimdata$AuthRank)) %>% crossing(crit_part = c(0,1))

predictiondata
augment(mod11, newdata = predictiondata, se_fit=TRUE)

# prediction data
predictions <- augment(mod12, newdata=predictiondata, se_fit=TRUE) %>% 
  mutate(
    .prob=plogis(.fitted),
    conf.low = plogis(.fitted - (qnorm(.975) * .se.fit)),
    conf.high = plogis(.fitted + (qnorm(.975) * .se.fit)),
    pred_final_rec = case_when(
      .prob >= .5 ~ "Likely",
      .prob < .5 ~ " Not likely"
    )
  )

predictions %>% arrange(.prob, .fitted, crit_part, desc(.fitted)) %>% 
  view()

# now let's try some graphs
# notice that the predictions are not in (0, 1)
ggplot(predictions) +
  aes(x = .fitted) +
  geom_histogram()

# if we wanted predictions on the probability scale,
#   we predict on the log scale,
#   then apply inverse-logit (logistic CDF)
# use plogis(): cumulative probability of the logistic distribution

# demonstration: scatter log odds (x) and predicted probability (y)
# with 95% CIs
# (CIs don't look smooth in this example because .se.fit depends on X data)
predictions %>%
  ggplot() +
  aes(x = .fitted, y = .prob) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high)) +
  geom_smooth() 

```

```{r pred-prob-ANES, include=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Predicted Probability of Polarization\n on Retired Military Elite Opinion Publications", fig.height=4.5, fig.width=6}
predictions%>%
  ggplot() +
  aes(x = aff_polar_alt, y = .prob) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high)) +
  geom_smooth() + labs(
    x="Level of Affective Polarization\n Among the US Electorate",
    y="Predicted Probability of an Op-Ed Published in NYT\n by Retired Military Officers which Violate\n Normative Standards"
  ) + scale_y_continuous(breaks=seq(-.1, .8, .1)) + 
  geom_vline(xintercept = 48.81, size=.2, colour="tomato") + 
  geom_vline(xintercept=59.62, size=.2, colour="tomato") + 
  geom_vline(xintercept=68.001, size=.2, colour="dodgerblue") + 
  geom_vline(xintercept=84.4, size=.2, colour="dodgerblue") +
  geom_vline(xintercept=79.55, size=.2, colour="tomato") +
  geom_vline(xintercept=100.35, size=.2, colour="dodgerblue") +
  geom_text(aes(x=48.81, label="Reagan, 1980 ANES",y=.4), colour="tomato", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=59.62, label="Bush 43, 2000 ANES",y=.4), colour="tomato", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=68.001, label="Obama, 2008 ANES",y=.4), colour="dodgerblue", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=84.4, label="Obama, 2012 ANES",y=.62), colour="dodgerblue", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=79.55, label="Trump, 2016 ANES",y=.55), colour="tomato", angle=90, vjust=1, text=element_text(size=9))  + 
  geom_text(aes(x=100.35, label="Biden, 2020 ANES*", y=.75), colour="dodgerblue", angle=45, vjust=1, text=element_text(size=9))
```

Figure \@ref(fig:pred-prob-ANES) clearly reveals that the probability of retired military officer commentary violating normative standards rises as the level of affective polarization among the US electorate increases. Regardless of the specific point estimates of probability revealed in Figure \@ref(fig:pred-prob-ANES), the more consequential and important takeaway is that if polarization remains - and all signs are that it will, at least in the short term - the commentary and dialogue offered by retired military officers to the American public will consistently challenge normative standards.

Collectively, the results offer mixed support for the four hypotheses this chapter set out to explore. The chapter provides strong support for H1, which theorized that military elites are less united ideologically during periods of high polarization relative to periods of low polarization. The instances with which retired military officers challenged or expressly violated one or more of the three normative standards of behavior greatly increased as polarization rose over time, indicating that like the public, military elites feel very strongly about certain ideological issues or sets of issues, and are willing to address these issues, and even strongly criticize civilian leadership or express partisan positions in doing so. 

The chapter found mixed support for H2, which hypothesized that military elites are increasingly active in a partisan manner when polarization was high relative to periods of low polarization. The data indicate that while elites certainly were increasingly active in a partisan sense as polarization increased, there were also isolated periods when military elites were active in a partisan sense when polarization was relatively low. In the data, one of these periods was the 2004 time frame, when several retired military officers endorsed candidates for President in the Press, or took to denouncing or supporting former Secretary of Defense Donald Rumsfeld. 

I find mixed support for H3, which hypothesized that military elites are more assertive and thus, are willing to challenge civilian authority during periods of high military prestige relative to periods when military prestige is low. The data clearly support this general trend, but the quantitative analysis revealed that the independent variable of polarization holds more explanatory power than that of military prestige in driving retired military officers to challenge civilian authority by offering blunt and pointed criticism in published opinion commentary. 

Finally, I find mixed support for H4, which hypothesized that military elites increasingly challenge civilian authorities in a visible, public manner during periods of high military prestige relative to periods of low military prestige. Here again, the data seem to generally support such a trend, but more analysis is required to determine if military actors are deliberately choosing to engage in behaviors that are increasingly visible rather than other, less-visible forms of behavior during the same time period. Ideally, qualitative case studies which isolate the varying levels of polarization and military prestige will further bear this out. 

# Can A Military Be Professional In an Era of Polarization?

This study demonstrated that although the vast number of opinion publications authored by retired military officers uphold the foundational principles and norms that have been the hallmark of American civil-military relations for decades, retired military elites are increasingly failing to uphold these norms in a significant portion of the messaging that they conduct with the American public. Specifically, the written opinion commentary which retired military officers publish in major US newspapers is increasingly challenging, if not expressly violating, one ore more of the principles of civilian control of the military, non-partisanship of the military institution, and a norm of "non-interference" of the military into certain realms of state policy. 

This study supports the general theory that among the pernicious consequences of polarization is a weakening in the ability of foundational normative standards of civil-military relations to constrain members of the military institution from engaging in certain forms of behavior, likely for similar reasons that polarization is dividing the American public. Not only does polarization reflect different policy preferences for a number of important political topics, which is routine in everyday domestic politics, but high levels of polarization also indicate that strong divisions exist regarding the worldviews, values, and overall purposes that should govern society. Military actors may genuinely desire to stay as politically neutral as possible, but as they encounter challenges with increasing frequency and in which a stark difference in consequences between two competing courses of action are present, they are increasingly likely to choose the dictates of personal conscience, even if doing so violates or challenges their professional duties [@cohn_what_2019].

Does it really matter if retired military officers write opinion commentary that pushes the boundaries of normative standards of behavior? Does it matter if today's retired military officers simply see the issue of expressing their opinions publicly differently than the Omar Bradley's of decades ago? Research has already confirmed that some citizens, especially partisans who strongly identify with a political affiliation, are happy when a retired military officer publicly quarrels with an elected official of the opposite political party, a trend that we know is becoming more problematic in an age of polarization [@robinson_michael_danger_2018]. A less explored impact, and one worthy of future research, is whether and to what degree increased retired office dialogue has on those who still serve in the active duty military. For many officers, the retired generals and admirals who are pushing the boundaries of normative standards are the former bosses and leaders with whom leaders in the active duty force once served. There is a risk that those still serving will view the behavior of retired military officers as increasingly normatively appropriate, which could risk these officers on active duty engaging in the same. 

Even if this were not to occur, the fact is that retired military officers have a lot of experience, much of which can be helpful in illuminating the knowledge of the public, especially when issues that involve operations are concerned. Instances in which a retired military officer criticizes a civilian official indiscriminately, adopts a partisan stance, or addresses a topic that is tangential to his or her military experience is a lost opportunity to engage the public on a topic that may be more beneficial in the short term. More seriously, it may have long-term consequences not just on the level of trust that Americans place in the military, but on the propensity of those serving in, and who will join, the active duty military, to disregard normative standards of behavior altogether. 
\singlespace
\pagebreak

# Appendix 1

\doublespace
## Word and Data-Base Searches to Obtain Data

I did two different sets of searches.  This first set of search terms was applied to the Factiva and Nexus-Uni data bases. In each of these databases, I had to select the "type of article" and included options such as commentary/opinion and letters to the editor.
The specific terms of search included:

"retired Army General" or "former Army General" or "retired Navy Admiral" or "former Navy Admiral" or "retired Marine Corps General" or "former Marine Corps General" or "former Air Force General" or "retired Air Force General" or "retired Coast Guard Admiral" or "former Coast Guard Admiral" or "retired from the Marine Corps" or "retired from the Army" or "retired from the Navy" or "retired from the Air Force" or "retired from the Coast Guard" or "supreme commander of" or "commander of" or "Chairman of the Joint Chiefs" or "chairman of the joint chiefs" or "Supreme Commander of" or "Chief of Staff" or "Chief of Naval Operations" or "retired Army" or "retired Navy" or "retired Air Force" or "retired Marine Corps" or "retired Coast Guard" or "retired military" or "commandant of" or "director of" or "retired from the military"

The second set of search terms was applied to Pro Quest Historical Newspapers, which the University of Wisconsin had access to for both the Wall Street Journal and the New York Times. Factiva included article from the Wall Street Journal only back to the middle of 1994, so I relied heavily on Pro Quest to search through the Wall Street Journal from 1979-1994. Unlike the search terms used for Factiva and Nexus-Uni, however, I came to quickly realize that the opinion pieces I sought were not categorized as opinion-editorials or letters to the editor, but rather as "articles."  This required significantly more time searching for the desired data, as more articles had to be searched through and sifted. The specific terms of search included:

(retired OR former OR “chief of staff”) AND (“Army General” OR Gen. OR general OR "Navy Admiral" OR Adm. OR admiral OR "Marine Corps General" OR "Coast Guard Admiral" OR "Air Force General" OR "Chairman of the Joint Chiefs" OR "colonel" OR "captain" OR "commander" OR "chief of naval operations" OR "commandant" OR "director of")

## Observations used for Regression Analysis

The following table contains those observations whose content either called criticized civilian leaders, adopted partisan stances, or addressed issues which are reasonably outside of military expertise.

\singlespace

```{r tab-appx, include=TRUE, echo=FALSE, message=FALSE, error=FALSE}
#create a table for the Appendix
TA1 <- quest_prof %>%rename(Crit=Crit_Insub, Part=Endorse_Partisan, Issue=Issue_Area) %>% 
  kable(format="latex", caption="Op-Eds authored by Retired Military Officers that Strongly Criticized Civilian Leaders, Expressed Partisan Positions, or Addressed Topics Outside of Military Expertise, 1979-2020", longtable=T) %>%
  kable_styling(font_size=9, latex_options="striped", repeat_header_continued=T) %>% 
  column_spec(1, width="9em") %>% 
  column_spec(2, width="14em") %>% 
  column_spec(3:4, width="2.5em") %>%
  column_spec(5, width="5.6em") %>%
  column_spec(6:7, width="1.4em") %>%
  column_spec(8:9, width="3.1em") %>% 
  row_spec(0, bold=T)

TA1
```

\pagebreak









