---
title: "When the Pen Becomes a Sword: Retired US Military Officers and Op-Ed Use in Contemporary America"
author: "Peter M. Erickson"
date: |
  `r format(Sys.time(), '%B %d, %Y')`
abstract: |
  
  _This chapter explores variation in the professionalism of opinion pieces authored by retired US military officers and published in major US newspapers since 1979. Drawing on an original data set of 390 opinion-editorial pieces published by retired military officers in The Wall Street Journal, The Washington Post, The New York Times, The Los Angeles Times, and the USA Today since 1979, the paper argues there has been an increase in published opinion pieces which fail to uphold the baseline norms of civil-military relations. This paper argues that this variation is the result of changes in two domestic political factors: political polarization and the overall prestige of the military institution. The paper concludes by arguing that the concept of "politicization of the military" --- something that previous civil-military relations scholars have identified and described --- must be widened to account not only for efforts by politicians to use the military for political advantages, but also for the phenomenon of retired military officers engaging the public to achieve political ends. The author argues that such a development, if uncorrected, will cause significant damage to the military as a profession._  
geometry: margin = 1.15in              # margin widths
fontsize: 12pt                         # text size
bibliography: dissertation.bib # bibliography file
biblio-style: authoryear               # citation style
urlcolor: blue                         # links to external URLs
citecolor: black                       # links to bibliography
linkcolor: magenta                     # internal figures, footnotes, etc.
indent: true                          # paragraph format
                                       # true for indented paragraphs
                                       
header-includes:
    - \usepackage{setspace}
    - \usepackage{multirow}
    - \usepackage{indentfirst}
    - \usepackage{caption}
    
    
output: 
  bookdown::pdf_document2:
    keep_tex: true
    latex_engine: pdflatex
    number_sections: false
    toc: false
    fig_caption: true
    citation_package: biblatex
    includes: 
      in_header: 
        - latex-preamble.tex
  bookdown::htmldocument2: default
  bookdown::word_document2:
      toc: false
---

\doublespacing 




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, include = FALSE, echo = FALSE}
library("here")       # file paths
library("tidyverse")  # workhorse package
library("tidylog")
library("kableExtra")
library("knitr")
library("ggdag")
library("dagitty")
library("gridExtra")
library("broom")
library("dplyr")
library("reshape2")
library("stargazer")
library("rstatix")
library("lme4")
library("brglm")
library("bucky")
```

\newpage

# Introduction

In late May and early June 2020, racial tension and concerns about police brutality gripped the American public following the death of George Floyd, an unarmed black man, by police in Minneapolis. In some cities and communities across the nation, peaceful protests were followed by periods of rioting and looting. On June 1, after giving remarks at the Rose Garden, President Donald Trump walked to St. John's Church after security forces visibly displaced throngs of protesters in neighboring Lafayette Square. Flanked by members of his cabinet, to include the Secretary of Defense, and with the Chairman of the Joint Chiefs of Staff not far behind, the President held up a Bible at the church, stared into the cameras of the media, and, several minutes later, returned to the White House. In the days and weeks that followed, a civil-military relations crisis ensued  [@feaver_military_2020]. The Secretary of Defense, Mark Esper, publicly disagreed with the President's threat to invoke the Insurrection Act of 1807 to use active-duty forces to quell domestic riots [@esper_secretary_2020], and General Mark Milley, the Chairman of the Joint Chiefs of Staff, later told senior military officers graduating from the National Defense University that he "should not have been there" (accompanying the President on the walk to St. John's Church), as his presence "created a perception of the military involved in domestic politics" [@milley_official_2020]. 

Furthermore, several retired military officers weighed in not only on the Trump administration's handling of specific events near the White House, but also on the broader issues of race in America.^[These officers included several retired four star generals and admirals]. However, there was marked variation in the tone and tenor of the remarks made by these officers: although each expressed deep concern about the prospect of active-duty forces confronting peaceful protesters, and thus, violating protections held by the First Amendment of the US Constitution [@brooks_dismay_2020; @dempsey_former_2020], some officers went much further in their criticisms, leveling direct insults of the President, including his fitness to lead [@goldberg_james_2020], his administration's broader policies on race in America [@allen_moment_2020], and his general leadership abilities [@mullen_i_2020]. Moreover, the comments made by these retired senior leaders renewed interest, at least for a while, in the topic of retired military officer speaking out on current events - a topic that civil-military relations scholars have debated [@kagan_let_2006; @brooks_let_2020], but one that has yet, in the author's opinion, to be studied empirically and systematically. In a domestic political age in which polarization is high, the topic of retired military officer speech is important for a variety of reasons. Not only does the speech of retired military officers carry the potential to impact public opinion, but the very act of retired senior leaders speaking out may also compel those who are still in the military to follow suit. More importantly, the topics that military leaders choose to address in the public square, and the tone of their remarks, is reflective of the conceptualizations these officers have regarding their role in society and ultimately, their purpose and professional identity as military officers. Indeed, in 2020, retired as well as active duty military officers grabbed the pen and argued in support of a variety of topics that clearly pushed the traditional boundaries of professional conduct by military officers by openly arguing for a variety of topics related to President Donald Trump, including including race in America [@askew_anti-racist_2020] and even removing the President from the White House [@nagl__2020].  

This chapter examines variation in the publication of op-ed pieces by retired military officers since 1979. There are two main arguments of this chapter. The first is that the data clearly demonstrates a significant change in the number, topic, and tone of opinion-editorial publications published by retired military officers. The second argument holds that the primary driver of this change is the increase in political polarization that has occurred in America over the same time period. While other factors certainly play a role in influencing retired military officers to pick up the pen and write pieces that challenge various norms of officership, such as non-partisanship and civilian control of the military, the data clearly reveals that political polarization holds more explanatory power than other factors in explaining the variation of retired military officer speech in America over the past four decades. Together, this analysis highlights the extent to which retired military officer speech has strayed from professional normative standards.  

This paper proceeds in three parts. First, the paper connects the behavior of retired military officer speech to the larger concept of military "role beliefs," a concept that is, I argue, worthy of reflection today, when both political polarization and the degree of societal prestige afforded to the military are at relatively high levels. Second, the paper explores and analyzes an original data set of nearly 390 opinion-editorial pieces published in major US newspapers by retired military officers since the late 1970s, and concludes that although the vast majority of op-eds published by retired military officers maintain professional standards, there is clear evidence that retired military officers are using the pen both to argue about topics that fall outside of the realm of traditional national security topics, and to violate norms of civilian control of the military and non-partisanship when doing so. Third, this paper discusses the implications of these findings on the challenges that face the military in what is perhaps the most difficult challenge it has faced in a long time: remaining professional in a highly polarized and divided society.  

\singlespace
# How the Military Views Itself
\doublespace
## Military Role Beliefs 
The concept of military "role beliefs" was first introduced by J. Samuel Fitch in his book, _The Armed Forces and Democracy in Latin America._ In the book, Fitch describes military "role beliefs" as the perceptions held by military officers regarding their normative role in politics [@fitch_armed_1998, 61]. At the center of the concept of role beliefs are the ideas that military officers have about the extent to which they should participate in various aspects of the political process and engage in various forms of political behavior. While Fitch's focus is on militaries in Latin America, it is possible to use the concept of "role beliefs" to develop a normative starting point applicable to military officers in the United States. 

Theoretically, there are at least two major determinants of military role beliefs, or the conceptions held by members of the military regarding their normative role in politics [@fitch_armed_1998]. In the most general terms, the military's involvement in politics requires both the _motive_ and the _opportunity_ to do so [@finer_man_1962]. I argue that in today's domestic political environment,  rising political polarization serves as the _motive_ for the military's intervention in politics, while high levels of prestige, or credibility of the military institution, provides the _opportunity_ for doing so. It is important to briefly elaborate how each of these offers the military in particular the ability to intervene in politics. 

With respect to the first determinant, political polarization, the literature in American politics enables the development of two plausible relationships and corresponding mechanisms that connect polarization to  military role beliefs. These relationships and mechanisms further stem from a particular conception or definition of what polarization is and how it works. One possible conception of polarization is that it marks the ideological widening between political elites, to include members of Congress or the leaders of political parties [@hare_polarization_2014]. According to this view of polarization, one possible mechanism by which the military intervenes in politics is by adopting a belief that elites from both dominant political parties have failed to advance and implement an effective governing agenda, and thus, the military steps in to do what political elites are not doing. This particular dynamic has been noted, in fact, by scholars in the comparative politics sub-field describing the actions of militaries in other parts of the world [@fitch_armed_1998]. 

Yet a second, and I believe the more likely, relationship between political polarization and military role beliefs, particularly in the contemporary United States, stems from a conception of polarization as the strength of feelings that one has concerning a number of ideological issues. This view of polarization is more consistent with a concept scholars have labeled "affective polarization", or "negative partisanship" [@abramowitz_great_2018]. If the variables of polarization and military role beliefs are linked in this way, the military does not intervene as a monolithic block due to a frustration with the political class's inability writ large to effectively govern. Rather, certain military elites, who share beliefs with a polarized electorate on any number of ideological issues, such as those regarding race, gender, and immigration [@abramowitz_great_2018], will selectively intervene in order to advance their personal beliefs regarding these topics. In other words, this mechanism posits that certain military elites feel as strongly as average citizens about certain polarizing issues, to the point that they will break from historical apolitical or non-partisan military traditions and norms in order to advance their beliefs. Thus, when polarization is high, there is a _motive_ among certain members of the the military, to include retired military elites, to deviate from their historical non-partisan and apolitical traditions. In short, increasing political polarization may make it increasingly difficult for members of the military to remain politically neutral as partisan conflicts subsume a greater number of issues, and as strong feelings of dislike or disdain surface against those who hold on opposing view.

The second determinant of military role beliefs is the level of prestige, or credibility, that the public places in the military institution. When these levels are high, I contend that members of the military possess the _opportunity_ to engage with the public. This includes engaging in behaviors such as making public arguments in the press, publishing opinion pieces, or more generally, sharing one's views. When society overall esteems the military at low levels, the expectation is that there would be fewer instances of the military publicly sharing its views, because there is less desire from the public (and thus, less desire from press outlets) to hear from members of the military community. There may also be, during periods when the military commands low levels of prestige and credibility, less inclination among members of the military to make public pronouncements. When the opportunities to engage with the public are more prevalent as a result of high levels of prestige, certain members of the military, I contend, are more likely to feel empowered to take on increasingly active roles, particularly in helping to weigh in or solve problems. Thus, we would expect to see a greater level of overall engagement with the press, and therefore, a greater number of overall opinion pieces authored by retired military officers, when military credibility is high. 

## Civilian Control, and Non-Partisanship

There are at two seminal principles that have helped guide and influence the military's "role beliefs" for most of US history: the tandem norms of civilian control of the military, and the non-partisanship of the military as an institution. Though these norms have been episodically challenged in a variety of specific instances and events, both norms have prominently featured within the military's organizational culture, and increasingly so as the military has modeled the behaviors of a profession, particularly in the post World War Two period [@weigley_american_2001]. The first norm of civilian control of the military stems directly from several provisions outlined in the US Constitution, which expressly grants control of the military (as well as a division of specific roles and responsibilities associated with maintaining control) to the President, Congress, and the individual states [@noauthor_us_1787-2]. The second norm of non-partisanship, while not expressly spelled out in the Constitution, has been cultivated throughout the military's collective experience of fighting and prevailing in the nation's wars and conflicts. 

A number of scholars have likewise underscored the importance of and relationship between the norms of civilian control and non-partisanship of the military. Perhaps the most influential of these scholars has been Samuel Huntington, whose book, _The Soldier and the State: The Theory and Politics of Civil-Military Relations_, argued that maintaining the norms of civilian control of the military and non-partisanship would also ensure that the military prevailed against the dangerous threat environment of the Cold War. Huntington specifically argues in favor of a model of civilian control of the military known as "objective control" in which the military and civilian occupy largely separate spheres of operation such that civilian leaders do not interfere with the details of military operations, and likewise, the military does not intervene in politics [@huntington_soldier_1957]. Only by such an arrangement, Huntington argues, would the military exist as a "politically sterile and neutral," and thus, non-partisan, institution, while simultaneously ensuring civilian control of the military [@huntington_soldier_1957, 83-84]. 

## Opinion-Editorial Publication as a Political Activity?

Scholars have long commented on where the line should be drawn that separates appropriate vs. inappropriate political behavior by military officers, to include military retirees [@brooks_militaries_2009; @owens_military_2015; @golby_beyond_2015]. Most scholars note that this is in many ways a balancing act: while the norms of civilian control of the military and non-partisanship are as close to what can reasonably be considered "sacred" within and among the military, military officers are (especially in the case of those who achieve high rank) have accumulated unique knowledge and experiences over the course of lengthy careers (in the case of retired military officers, at least 20 and in some cases, upwards of 40 or even more years of service). In many cases, these military elites have interacted with Presidents, members of the cabinet, members of Congress, and foreign dignitaries. Though they possess an abundance of expertise in many tactical matters that pertain to war-fighting, they are also familiar with issues ranging from how wars should be fought, to future security threats looming on the horizon, to fixing the institutional culture of large organizations such as the military services, and much more. Moreover, especially in a democracy such as the United States, there is a reasonable expectation that these leaders use their military experience to contribute to the free flow of information that helps inform the citizens of a Republic and enables elected leaders to weigh the decisions before them. Nonetheless, military elites are not experts in everything, nor should they try to be. Even General of the Army Douglas MacArthur, one of several men who ever wore the five star rank, and a hero of World War Two who later clashed with President Truman regarding American strategic goals in Asia, recognized that the military should have focus areas. Addressing the Corps of Cadets at West Point in 1962, the then-82 year old retired General told the Corps:

\singlespacing

> **Yours is the profession of arms, the will to win, the sure knowledge that in war there is no substitute for victory, that if you lose, the Nation will be destroyed, that the very obsession of your public service must be Duty, Honor, Country. Others will debate the controversial issues, national and international, which divide men’s minds. But serene, calm, aloof, you stand as the Nation’s war guardians, as its lifeguards from the raging tides of international conflict, as its gladiators in the arena of battle. For a century and a half you have defended, guarded and protected its hallowed traditions of liberty and freedom, of right and justice. Let civilian voices argue the merits or demerits of our processes of government. Whether our strength is being sapped by deficit financing indulged in too long, by federal paternalism grown too mighty, by power groups grown too arrogant, by politics grown too corrupt, by crime grown too rampant, by morals grown too low, by taxes grown too high, by extremists grown too violent; whether our personal liberties are as firm and complete as they should be. These great national problems are not for your professional participation or military solution. Your guidepost stands out like a tenfold beacon in the night: Duty, Honor, Country** [@macarthur_duty_1962].

\doublespacing

There have likewise been both formal and informal boundaries placed on various forms of political activities by military officers over the years. Formally, Department of Defense (DoD) Directive 1344.10 contains a "standing list of political behavior do's and don'ts for members of the active duty military" [@urben_wearing_2014, 1]. This list includes authorizing members of the military to engage in activities such as voting and joining a partisan or non-partisan club and to attend this club's meetings when not in uniform, and prohibits members of the military from activities such as participating in partisan fundraising, or advocating for or against a partisan cause or figure on radio or television [@department_of_defense_political_2008]. Informally, a number of recent Chairmen of the Joint Chiefs of Staff have likewise urged members of the military to refrain from engaging in overly political activities, particularly in election years, when such activity is more likely to negatively reflect on the military. For example, prior to the 2016 Presidential Election, then-Chairman of the Joint Chiefs of Staff General Joseph Dunford reminded the military that the basis upon which it is able to maintain the public's trust is it's ability to remain "apolitical and neutral" [@dunford_upholding_2016]. Clearly present in both the formal regulations and informal admonition regarding the political behavior of military officers is concern for the military's non-partisan ethos. 

The specific act of authoring opinion commentary has occasionally garnered increased attention in recent years, perhaps starting in 1992 when then-Chairman of the Joint Chiefs of Staff General Colin Powell authored a piece that presented his views regarding the potential use of American military force in several developing "hot spots" across the world (including the Balkans, Somalia, and other places) entitled, "Why Generals Get Nervous." Published in the _New York Times_ less than a month away from the 1992 Presidential Election, Powell wrote in part, "decisive means and results are always to be preferred, even if they are not always possible. So you bet I get nervous when so-called experts suggest that all we need is a little surgical bombing or a limited attack. When the desired result isn't obtained, a new set of experts then comes forward with talk of a little escalation. History has not been kind to this approach" [@powell_opinion_1992]. More than a decade later, in 2006, the so-called "Revolt of the Generals" occurred in which several retired military officers publicly criticized and called for the removal of Defense Secretary Donald Rumsfeld as a result of his leadership and its impacts on the war in Iraq. Scholars noted that the event was unprecedented in the way that multiple officers "devoid of partisan political motive, felt the need or obligation to speak out publicly regarding policy and leadership during an ongoing conflict" [@cook_revolt_2008, 7]. While the "Revolt of the Generals" generated concern that the military's professional ethic had been strained, others applauded the generals' honesty and forthrightness and claimed that concerns about a dip in the professionalism among military officers was overblown. Frederick Kagan, for instance, claimed, "there is no danger to the republic in a handful of retired generals speaking their minds. There is great danger in making vital decisions about an on-going armed struggle without hearing the views of all available experts" [@kagan_let_2006]. 

Whether one agrees with Kagan's general assessment regarding the impact of opinion pieces authored by military officers, it is important to understand the ways in which opinion commentary is different from other types of speech made by military officers, such as that provided during formal testimony to Congress. Risa Brooks, for instance, argues that opinion or commentary is distinguished by both its "public nature" as well as the "scope and nature of the commentary" [@brooks_militaries_2009, 218]. Building off of Brooks, I would add that the action of authoring opinion commentary is completely voluntary, and thus, a deliberate action. Crafting a written op-ed is different from even giving interviews on television, because writing an op-ed requires the author to deliberately prepare and submit a viewpoint for publication. A deliberate and carefully-crafted message further underscores the degree to which the author believes the message he or she is sending. It is difficult to "misspeak" during or to be "misquoted" from an opinion-editorial. 

In short, we should conclude that military officers who engage in authoring commentary or opinion pieces genuinely believes the message he or she is sending. As Brooks points out, however, the content of the specific message matters, and thus, the act of writing an op-ed in and of itself is not problematic. An op-ed that strictly points out the pitfalls of the nation's combat strategy while avoiding blatant criticism of a civilian official may not be inappropriate at all, whereas an op-ed written by a retired military officer that argues in favor of a specific Presidential candidate or policy is entirely inappropriate, at least by the normative standards of non-partisanship and civilian control of the military. Therefore, at least two questions should be asked when examining an opinion piece authored by a retired military officer: first, does the piece violate either the norm of civilian control of the military or the military as an apolitical institution, and second, does the op-ed address an appropriate topic that the military officer, on the basis of his or her professional expertise, is qualified to address?  

## The Role of Retired Military Officers

Beyond these two questions that an individual officer (active or retired) should ask when thinking about writing an op-ed, there is, I argue, an even larger and broader question: what message or messages is the Republic receiving from the collective body of military leaders who provide opinion-laden commentary? Because military elites tend to provide this type of commentary more often than those on active duty, this question is particularly relevant for retired military elites. The collective tone, nature, and message sent by retired military elites serves as a barometer of sorts regarding the role that the retired officer corps sees for itself in interacting with the American public. Are retired military elites upholding norms of non-partisanship and civilian control of the military? When they are not, should we view these instances as a rare aberration, or are there clear trends that can be detected that should legitimately cause concern? 

One reason to believe that a shift in the "role beliefs" among retired military officers may be occurring is the increasing political behavior of retired military elites during recent national political events. During the Presidential Campaign of 2016, for instance, Retired Army General Martin Dempsey, who previously served as Chairman of the Joint Chiefs of Staff, published an opinion-editorial piece in _The Washington Post_ which criticized retired Army Major General Mike Flynn and Marine Corps General John Allen for participating in the respective Republican and Democratic conventions. Dempsey wrote, "As generals, they have an obligation to uphold our apolitical traditions. They have just made the task of their successors – who continue to serve in uniform and are accountable for our security - more complicated. It was a mistake for them to participate as they did. It was a mistake for our presidential candidates to ask them to do so" [@dempsey_military_2016]. Thus, there was both a clear expectation and a concern among some prominent military elites that their fellow retired generals and admirals were not adhering to the norms of civilian control and non-partisanship of the military institution. Additionally, after four years of the Trump administration, during which several retired military officers served in prominent positions within the administration, and now that President-Elect Biden has, at least as of the time of writing this paper, nominated another retired general for the position of Secretary of Defense, there is a renewed interest in the normative roles that retired military elites should and actually do have in America. 

By systematically examining the op-eds written by retired military officers in major US publications over time, I contend that it is possible to identify potential changes (or none) in the role beliefs held by retired US military officers. In doing so, the key area of focus must be the degree to which retired military officers have upheld vice strayed from the norms of civilian control and non-partisanship of the military, as well as the degree to which retired officers have addressed topics that are appropriately rooted in relevant military expertise. If there are patterns of violating these norms that can be detected, it may indicate that the role beliefs of certain members of the retired military community have likewise changed. On the other hand, if there has been little deviation over time from these foundational norms, it is possible to conclude that military retirees view their normative roles in politics (role beliefs) in similar ways as previous generations of retired military officers. 

## Hypotheses 

Given the theoretical importance of the norms of civilian control and non-partisanship of the military institution, the rise of domestic political polarization and prestige of the US military over the past several decades, and the unique contributions that retired military elites make in a democratic society, four distinct hypotheses are derived, which will be examined and tested in the analysis to follow. These hypotheses include:

\singlespace
**H1: Increasing political polarization is associated with an increase in commentary authored by retired military officers that criticizes civilian leadership or makes clear partisan statements (potentially indicating a drop in adherence among retired officers to the principle of civilian control).**

\doublespace

\singlespace
**H2: Increasing political polarization is associated with an increase in commentary authored by retired military officers that addresses topics that are outside of traditional national security debates.**

\doublespace

\singlespace
**H3: During periods of high prestige and credibility, the aggregate level of commentary offered by retired military officers is higher relative to lower periods of prestige and credibility.**

\doublespace

\singlespace
**H4: During periods of active ground combat, military elites will increasingly comment on topics related to the war, and comment less on other domestic issues.**

\doublespace

\singlespace


\doublespace


# Methodological Approach and the Data

```{r introduction, include = FALSE}

# Read in the Data

dataset <- read_csv(here("data", "MilEditorials.csv"))
dataset %>% print()
# Only read in the first 390 rows - is there a better way to get rid of N/A data?

# Replace NA with 0 in the data for variables letter_ed to online_only, and leg_act #to role
adjdataset <- dataset[1:390,] %>%
  mutate_at(
    .vars = vars(letter_ed:online_only, Leg_Act:Role),
    .funs = ~ case_when(is.na(.) ~0, TRUE ~ 1)
) %>% mutate(
    PubYr = lubridate::year(PubDate)
) %>% view()

summary(adjdataset)
names(adjdataset) 

# count the number of observations by type of binary variable and store as new dataframe
countbytype  <- adjdataset %>% 
  count(Topic) %>% view()


# count the number of observations by source
countbysource <- adjdataset %>% 
  count(Source) %>% view()

#count the number of times each source occurs in the data
WSJcount <- adjdataset %>% count(Source=="WSJ")
WSJcount

NYTcount <- adjdataset %>% count(Source=="NYT")
NYTcount

WaPOcount <- adjdataset %>% count(Source=="WaPO")
WaPOcount

USATodaycount <- adjdataset %>% count(Source=="USA Today")
USATodaycount

LATimescount <- adjdataset %>% count(Source=="LA Times") %>% print()
LATimescount

# Create a new variable called Yrcount, which is a count of the yearly number of published op-eds

adjdataset <- adjdataset %>% 
  group_by(PubYr) %>% 
  mutate(Yrcount=n()) %>% 
  view()

# number of authors and author teams
adjdataset %>% 
  group_by(Author) %>% 
  count() %>% 
  view()
```

The methodological approach used in this analysis consists of two broad steps. The first is to analyze the topics which retired military officers have been writing about, and the second is to determine, regardless of the topic addressed in a publication, whether an op-ed has violated the norms of non-partisanship or civilian control of the military. The data consists of 390 opinion-editorial publications and letters to the editor authored by retired military officers and published in major US newspapers from 1979-2020. I include both opinion-editorials and letters to the editor because both display a viewpoint that required the author to craft a deliberate message, and to submit the piece for consideration to be published by an editorial board at a particular newspaper. Because I am interested in messages that required the officer to craft a deliberate message, I do not include published interviews with retired military officers, nor any reprinted speeches or remarks that a retired military officer gave at another venue (while searching for the data, there were a few instances in which newspapers published remarks of officers made at events). I also do not include short forms of analysis, such as that provided by the "Monkey Cage" from _The Washington Post_. 

The five newspapers that I examine in depth include the _The Wall Street Journal_ (WSJ), _The New York Times_ (NYT), _The Washington Post_ (WaPo), _The Los Angeles Times_, and _The USA Today_. With the exception of _The USA Today_, all of the newspapers have long and established histories (the _USA Today_ began in 1982 [@noauthor_about_nodate]). Moreover, publishing an opinion piece or letter in these newspapers is often a competitive process. It is not uncommon to see opinion pieces in these prominent newspapers published by heads of state, members of congress, governmental figures, and leaders of industry and academia. Therefore, examining these particular five sources - as opposed to smaller, less known newspapers - helps select the most prominent viewpoints that retired military officers greatly wanted to convey to the public. Finally, these newspapers, particularly _The Wall Street Journal_ (WSJ), _The New York Times_ (NYT), and _The Washington Post_ (WaPo) are likely the most circulated among the defense and national security communities, as they are located in New York City and Washington D.C., where think tanks, international organizations, and the Pentagon are based.

There are both advantages and disadvantages to confining my data to select US newspapers, as I have. The primary advantage is that these sources offer longevity across a time period of nearly forty-two years; in doing so, they provide the researcher with an opportunity to detect patterns across several decades. The primary disadvantage is that by limiting myself to only the five newspapers I examine, I am not including a large number of modern social media and blogging sites, contemporary webpages, and other digital offerings that likewise showcase a large number of opinions of retired military officers. 

Although this is a limitation in terms of selection, the methodological approach adopted in this analysis allows for an examination of the same newspaper sources over time. This is important, as it affords the researcher the opportunity to detect potential within-source variation. However, it should be noted that the overall influence and impact of newspapers is likely less than it was several decades ago. Social media has taken over as the primary means in which a majority of Americans receive their news today [@martin_how_2018], and even after accounting for the advent of online editions of newspapers in the late 1990s and early 2000s, the overall circulation of newspapers (print and online editions combined) is still down today compared to earlier years. As a means of rough comparison, estimated total daily newspaper circulation in the US in 2000 was 62 million for weekday and 54 million for weekend editions, respectively, while in 2018, the same figures were 29 million and 31 million [@barthel_leading_2019]. Still, the five newspapers examined are reasonably well known, and while it can't be proven in this paper, most individuals would consider it an accomplishment to author an opinion piece published in any one of these five examined sources.

To obtain the data, I use a series of key-word searches, such as "retired military officer," "retired from the Army, Navy, Marine Corps, Air Force," "former Chairman of the Joint Chiefs of Staff," "General," "Admiral," and so forth in several different databases, including Factiva, Lexus-Uni, and ProQuest Historical Newspapers. ^[For exact replication of the database search parameters, please see the Appendix.] Each of these databases classifies historical newspaper articles differently: some databases include editorials as a separate source type to search for (making the acquisition of data easier), while others do not, requiring a search of all articles that contain the key words (which therefore results in a significantly larger pool of articles to search through and find the op-eds and letters I am looking for). I include in my search publications that were made available both in print and online format, though I did not double-count: if the same op-ed was published both in print and online and published by the same author, I count the article only one time. ^[To my knowledge, there is no data set that yet includes the data assembled for this research project. Therefore, I cannot conclusively say that I actually found all of the opinion pieces and letters authored by retired military officers and published in these sources within the time periods examined in this paper. It is possible, for instance, that I left out an important search parameter that did not trigger a certain number of op-eds that were actually authored by retired military officers. While it is difficult to put a precise degree of confidence in the method employed to obtain the data, my personal reaction is that the method employed caught greater than 95% of the actual observations. I offer this as a "ballpark" assessment based off several searches of individual newspapers in a given year in which I compared the results I obtained using the search parameters described against a method in which I literally looked through every op-ed published in a given calendar year.]

In some instances, especially where the same retired military officers authored multiple publications, the officer did not always refer to him or herself as a retired military officer. This occurred in several instances in which the officer was, ostensibly, writing from the perspective of someone other than a retired military officer, such as that of a political appointee to a high level governmental position (after retirement from the military, Army General Barry McCaffrey, for instance, served as President Clinton's Director of the Office of National Drug Control Policy). Nonetheless, because these officers had previously identified themselves as retired military officers in earlier publications, and thus, because members of the public might reasonably recognize the author as a prominent retired military figure, I include the subsequent publications in the data, and simply control for whether the author identified him or herself as a retired military officer in a specific publication. 

Some caution must be taken up front so as to not over-promise and under-deliver concerning what the data is actually capable of presenting. The first caution to consider is that opinion-editorial publications to some degree reflect not only the views of retired military officers, but also the desire of the press, or in economic terms, the "demand" of the press to publish the views of these officers. Moreover, because op-eds are to some degree designed to advocate for or against a certain course of action, op-eds will likely present somewhat provocative material. As a result, under periods of harmonious civil-military relations, there may be either less of a demand from the press to publish the viewpoints of retired military officers, or less of a supply side effect for retired military officers to write about topics in the first place. Moreover, both the supply and demand for commentary written by retired military elites will likely be somewhat driven by exogenous events, such as wars. A second cautionary note is that over the course of several decades, underlying societal changes, including political polarization, have occurred. These changes have likewise likely impacted the willingness of the press to publish certain viewpoints of military officers, and/or the desire of military officers to write about certain topics in the first place. It is important that these societal-level effects be identified and, if possible, controlled for. 

## Summary Statistics 
The data set includes a total of 390 observations, penned by 216 distinct different authors or author teams. Eight of the observations are written by anonymous authors, but it is unknown whether the same unknown author penned several of these eight observations. Included in the data are cases in which at least one retired military officer co-authored a piece with one or more civilian authors. ^[There are several co-authored pieces. For example, where General (Retired) David Petraeus wrote one publication and another one was written by General (Retired) David Petraeus and Michael O'Hanlon, I counted these as two different authors for coding purposes.] There are also several retired military officers who penned several observations. The most prolific of these officers includes Retired Army Lieutenant General William Odom (22 publications in total), Retired Air Force General Michael Hayden (20 publications in total), and Retired Marine Lieutenant Colonel and Former National Security Adviser Robert McFarlane (18 publications). 

```{r, include=FALSE, echo=FALSE}
# show summary stats for yearly count, rank, and then a breakdown according to source
# yearly count
yearstats <- adjdataset %>% 
  select(PubYr, Yrcount) %>% 
  distinct(PubYr, Yrcount) %>% 
  print()

summary(yearstats)
sd(yearstats$Yrcount)
var(yearstats$Yrcount)

#rank
rankstats <- adjdataset %>% 
  filter(AuthRank <=10) %>% 
  select(Author, AuthRank) %>% print()

summary(rankstats)
sd(rankstats$AuthRank)
var(rankstats$AuthRank)

#sex
adjdataset %>% group_by(AuthSex) %>% count()

#service
adjdataset %>% group_by(Service) %>% count()

Descriptive_Stats <- tribble(
  ~Variable, ~Min, ~Max, ~Mean, ~SD, ~N, 
  'Author Rank', 4, 10, 8.34, 1.99, 390,
  'Number of Yearly Publications', 1, 27, 9.51, 6.84, 390)

Sex_Service_Stats <- tribble(~Breakdown, ~Observations, ~Percentage,
   'Male Authors', 374, 95.9,
  'Female Authors', 3, .8,
  'Mixed  Authorship', 5, 1.3,
  'Sex Unknown', 8, 2,
  'Army', 205, 52.5,
  'Navy',  56, 14.3,
  'Marine Corps', 47, 12.1,
  'Air Force', 63, 16.2,
  'Coast Guard',3,.8,
  'Mixed Service Authorship', 16, 4.1)
  

 rankdistrib <- adjdataset %>% group_by(AuthRank) %>% count()

```

In addition to coding each publication for the source and a count of publications in a given year, several bits of biographical information about each author is collected, including the service in which the author served, the rank achieved by the author during his or her time in service, and the sex of the author. Summary statistics for the variables of author rank and the annual count of publications are displayed in Table 1, below. Of note, the variable for author rank is coded equivalently to the officer's military pay grade. For example, a rank of "10" means the officer achieved the pay grade of O-10, which is a four star general or admiral. A rank value of 7 refers to a one star general or admiral. A rank of 4 corresponds to a pay grade of O-4, which in the Army, Air Force, and Marine Corps, refers to the rank of Major. In the Navy and Coast Guard, a pay grade of O-4 refers to the rank of Lieutenant Commander. A rank of 5 in the Army, Air Force, and Marine Corps refers to a Lieutenant Colonel, whereas in the Navy and Coast Guard, a rank of 5 denotes a Commander.  Finally, a rank of 6 in the Army, Air Force, and Marine Corps refers to a Colonel, whereas in the Navy and Coast Guard, a rank of 6 corresponds to the rank of Captain. 


```{r table1, echo=FALSE, include=TRUE, warning=FALSE, error=FALSE}
T1 <- kable(Descriptive_Stats, 
            format = 'simple', 
            booktabs = TRUE, 
            caption = "Descriptive Statistics for Author Rank and Yearly Number of Publications Breakdown of Observations by Author Sex and Service Affiliation",
            align = c('l', 'c', 'c', 'c'),
            digits = 2)

T1
```

Table 2, below, shows the breakdown of observations by author sex and service affiliation. Mixed authorship refers to publications that were authored by a mixture of male and female authors, whereas mixed service authorship denotes publications that were authored by more than one individual and together, as an author team, had served in multiple service branches - for example, a retired Colonel who had served in the Air Force and a retired Admiral who had served in the Navy write an op-ed together. 

```{r table2, echo=FALSE, include=TRUE, warning=FALSE, error=FALSE}
T1 <- kable(Sex_Service_Stats, 
            format = 'simple', 
            booktabs = TRUE, 
            caption = "Breakdown of Observations by Author Sex and Service Affiliation",
            align = c('l', 'c', 'c', 'c'),
            digits = 2)

T1
```

Unsurprisingly, as the military's largest service, retired military elites who served in the Army comprise the largest segment (52.5%) of authors in the sample. Authors who served in the Navy and Air Force, whose active duty populations in 2020 were of comparable size, featured similarly in the sample (14.3% and 16.2%, respectively). Representation from authors who served in the Marine Corps (12.1%) and Coast Guard (.8%) round out the sample. In terms of the gender of the author, again unexpectedly, male authors dominate the sample (95.9%). Because the sample includes publications dating back to 1979-2020, a probable reason for this is the difference in gender at the highest ranks of service - a feature that will likely change in the coming years as the restrictions on women serving in specific positions within the military have been removed.

Figure 1, below, plots the yearly count of publications authored by retired military officers across all of the major newspapers. The smoothed trend line indicates that the average yearly count of publications rose steadily from 1979 until the early 2000s, at which time, a decline occurred until approximately 2008. From 2008 until the present day, the average yearly count of publications authored by retired military officers in the major US newspapers examined in this study rose heavily during the Obama administration years before briefly falling, but then resuming the high count during the Trump administration. 
```{r plot1, fig.align="center", warning=FALSE, error=FALSE, message=FALSE, include=TRUE, echo=FALSE, fig.cap="Number of Op-Eds and/or Letters to the Editor Authored by Retired Military Officers and Published in Major US Newspapers (WSJ, NYT, WaPo, USA Today, LA Times), 1979-2020"}
  
# construct a scatter plot with line and smoothing

ggplot(data=adjdataset, aes(x=PubYr, y=Yrcount)) + geom_point() + geom_line() + geom_smooth(method="loess") + labs(
  x="Year", 
  y="yearly count", 
  caption="Note: Data as of 1 December 2020"
  ) + scale_x_continuous(limits=c(1979, 2020)) + scale_y_continuous(limits=c(-7, 28)) + 
geom_vline(xintercept = 1981, size=.2, colour="tomato") + 
  geom_vline(xintercept=1989, size=.2, colour="tomato") + 
  geom_vline(xintercept=1993, size=.2, colour="dodgerblue") + geom_vline(xintercept=2001, size=.2, colour="tomato") + 
  geom_vline(xintercept=2009, size=.2, colour="dodgerblue") +
  geom_vline(xintercept=2017, size=.2, colour="tomato") +
  geom_text(aes(x=1981, label="Reagan",y=10), colour="tomato", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=1989, label="Bush 41",y=15), colour="tomato", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=1993, label="Clinton",y=20), colour="dodgerblue", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=2001, label="Bush 43",y=-2), colour="tomato", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=2009, label="Obama",y=-2), colour="dodgerblue", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=2017, label="Trump",y=-2), colour="tomato", angle=90, vjust=1, text=element_text(size=9)) 
  
```
It is noteworthy, from a macro perspective, to consider all of the impacts on the US military over the past four decades and more. Numerous substantive events, debates, and changes have occurred in this time period for retired military officers to weigh in on. In terms of external threats and wars, this period includes the final years of the Cold War against the Soviet Union (through 1991), the Gulf War in Iraq (1990-1991), operations in the Balkans (primarily 1993-1995), and the post-9/11 wars in Iraq (2003-present day) and Afghanistan (2001-present day), as well as a host of smaller but important military operations in Grenada (1983), Panama (1989), Somalia (1992-1993), and Haiti (1994-1995). The military has also undergone significant social and institutional changes in the time period examined. In the late 1970s, the military had ended the draft only years before (1973) following the conclusion of the Vietnam War. The debates over "Don't Ask, Don't Tell" (instituted in 1994 and repealed in 2011), as well as policies regarding women in the combat arms (2015) are also included in the data, as are numerous annual discussions regarding the size of military budgets.

Returning to a description of the statistics, Figure 2 below displays the distribution of publications by author rank (left) and source (right). As the distribution plot according to author rank shows, the publications tend to be authored by those of a higher rank. 185 of the 390 publications (47.4%) were authored by retired 4 star general or flag officers (Generals or Admirals). Generals or Admirals account for 276 of the total 390 publications (70.8%). The fact that high ranking officers tend to publish in the newspapers examine is unsurprising, and confirms the competitive nature of publishing an opinion piece in one of the premier newspapers in the nation. At the same time, there are enough authors who are not retired generals or admirals (29.2%) to show that those who do not attain the rank of general or admiral still have a respectable opportunity to publish their views in the examined sources.  

The distribution according to source, shown on the right side of Figure 2, reveals that the overwhelming number (379, or 97.2%) of opinion publications were authored by retired military officers in the _New York Times_, _Washington Post_, or _Wall Street Journal._ Only 11 of the 390 publications (2.8%) were authored in either the _USA Today_ or the _Los Angeles Times_.  This is not surprising. Though they distribute worldwide and are read across the country, the _Washington Post_, _New York Times_, and _Wall Street Journal_ are based in the nation's capital and New York City. Think tanks and international organizations abound in both cities, and thus, these three newspapers are likely more read by those who work in the defense or foreign affairs community. 

```{r plot2, include=TRUE, echo=FALSE, warning=FALSE, fig.cap="Opinion Pieces authored by Retired Military Officers Broken Down by Author Rank (L) and Newspaper Source (R), 1979-2020"}
colplot1 <- ggplot(data=rankdistrib, aes(x=AuthRank, y=n)) + labs(x="Pay Grade", y="Number of Publications") + geom_col() + geom_text(aes(label=n), position=position_dodge(width=.9), vjust=-.25)

colplot2 <- ggplot(data=countbysource, aes(x=Source, y=n)) + labs(x="Source", y="Number of Publications") + geom_col() + geom_text(aes(label=n), position=position_dodge(width=.9), vjust=-.25) 

grid.arrange(colplot1, colplot2, ncol=2)


```

## First Layer Analysis: What Are Retired Military Officers Addressing?

Having established that there has been a relative increase in the overall number of opinion pieces authored by retired military officers and published in major US newspapers since 1979, the next critical step is to determine whether there has been change, and to what degree, in what retired military officers have actually been writing about. This requires classifying each of the observations into one of several possible types of topic. The possible range of topics should be broad enough to classify every observation, while maintaining sufficient distinction between each of the topics so as to make useful comparisons between the topics. 

I generate a categorical variable, "topic", that can take one of eight possible categories. The first is **foreign policy**, if the publication generally refers to American foreign policy or overarching National Security policy. The next category of publication is **war-fighting and operations**, which I code those publications that address an ongoing conflict, or a central aspect of the military's ability to fight and win on the battlefield. Of note, I also include those publications that deal with most intelligence procedures, as in most cases, the authors of these publications tied intelligence to the military's ability to win on the battlefield. The third category is **social policy**, which include those publications that address questions such as who should serve, and why. In the data, this included publications that addressed subjects such as the "Don't Ask, Don't Tell" policy, recruitment, and the integration of women into the combat arms branches of the military. 

A fourth category is **domestic policy**, for which I include those publications that address issues that do not have a _direct_ impact on the military, or those that as the main points of emphasis, describe the general operating environment in Washington D.C. To be clear, especially in a day and age when the military has taken on greater roles, it is easy for a military author to logically connect almost any topic with the nation's defense. And though there may be convincing times when these connections need to be made, the determination used when classifying publications according to this variable was whether the military can reasonably claim that it either is or should be an expert on a particular topic addressed. In the data, for instance, publications that first and foremost addressed the national debt or the domestic economic impacts of a given policy, are classified according to this category. The fifth categorical variable for topic is **budget/weapons/troops**, which I use to classify those publications that primarily address issues such as military or defense budgets, or why the military or a particular service should or should not adopt a particular weapons platform. 

The sixth category of topic used to classify publications is **support**, which I classify those publications that are written in support of the collective military (such as op-eds that ask the nation to remember fallen Troops and their families on Memorial Day), as well as those that actually support or weigh in on specific military leaders (such as those that recall a prominent officer who passed away recently). In the data, most patriotic calls to think about the sacrifice that troops make or have made are classified in this category. The seventh category of possible topics is **civil-military relations,** which I use to classify those topics that mainly address how civilian and military leaders should or actually do share power, or how generals and admirals should offer advice in an improved manner. In the data, this also includes publications which comment on how the military and government organizations may be better organized or rearranged in order to better  implement defense policies. The eighth and final category for publication topic is **service culture**, and refers to those publications that primarily address an issue or issues that are ongoing within the military services. ^[See the Appendix for a more-detailed explanation of each variable and classification.]

In order to classify each observation by topic, I read the op-ed or publication, usually multiple times, and identify what I think is the author's main point or thesis statement. To be clear, separating these publications into one of the eight possible categories is not a perfect process. Some classifications address multiple topics, or straddle the line between two or more categories. When this occurred, the central question made when determining an observation's ultimate topic category was, "what is the central argument made in the publication?"

Table 2, below, displays the breakdown of observations by topic category. The topics of **war-fighting and operations** and **foreign policy** account for a total of 230 of the 390 observations (59% of the sample). The categories of **support**, **domestic policy**, and **budget/weapons/troops** account for a total of 119 observations, or 30.5% of the sample. Finally, the topics of **social policy**, **civil-military relations**, and **service culture** account for a combined 41 observations, or 10.5% of the sample. 

```{r table3, fig.align="center", include=TRUE, echo=FALSE, warning=FALSE}
countbytype %>%
  kable(format="latex", caption="Publication Breakdown by Subject") %>%
  kable_styling(latex_options = "striped", font_size=10) %>% 
  row_spec(0, bold=T)
```

Is this breakdown by itself surprising in any way, especially as it relates to the topic of military role beliefs? I contend that the answer is not especially. It is unsurprising that retired military officers comment heavily on matters related to war-fighting, operations, and national security policy, as these matters directly relate to the experiences of retired military leaders. If anything, it is somewhat surprising, though not necessarily cause for concern, that the number of observations addressing domestic policy matters is as high as it is, and that the number addressing civil-military relations is as low as it is. 

However, what is relevant and applicable to the issue of military role beliefs is whether there has been a change in the way retired military officers have been addressing domestic political matters over time. Figure 3, below, plots the proportion of annual publications authored by retired military leaders that address domestic policy since 1979, and reveals several interesting trends. The first is that with the exception of an outlier year (1983), no publications addressed domestic policy until 1999. The first two decades worth of data, in other words, saw retired military officers nearly shun domestic politics altogether. The second interesting trend is that since that time period, retired military officers have become increasingly active in addressing domestic political matters when authoring opinion publications. The trend line in blue, moreover, drops between roughly the years 2009 and 2012, during the first Obama Administration. The final trend is that since 2012, the proportion of annual opinion publications authored by retired military officers which address domestic political matters has increased significantly. While this trend has increased under the Trump administration, it's origin can be found under the Obama and George W. Bush Administrations. 

```{r, include=FALSE, echo=FALSE}
## Create several new variables.  

# The first is a variable called "case-count" that counts the number of times by publishing year an #article published that calls for legislative action, criticizes or is #insubordinate, or is blatantly #partisan. 

#The second variable, prop_yr, is simply a 
# proportion of annual publications where these conditions hold true.

# The third variable is called dom_pol_count which counts the number of publications
# by publishing year that are written on topics of domestic policy, and the fourth 
# variable is a proportion of this type of article by year. 

adjdataset <- adjdataset %>% group_by(PubYr) %>% 
  mutate(
    case_count=sum(Leg_Act ==1 | Crit_Insub==1 | Endorse_Partisan==1), 
    prop_yr=(case_count/Yrcount),
    dom_pol_count=sum(Topic=="DomesticPolicy"),
    dom_pol_prop=(dom_pol_count/Yrcount)
  ) %>% view()
```

```{r plot3, include=TRUE, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, fig.cap="Proportion of Opinion Publications authored by Retired Military Officers Which Primarily Address Domestic Politics, 1979-2020"}
ggplot(data=adjdataset, mapping=aes(x=PubYr, y=dom_pol_prop)) + geom_point() + geom_smooth() + labs(x="Year", y="Proportion", color=NULL)+ geom_vline(xintercept = 1981, size=.2, colour="tomato")+
  geom_vline(xintercept=1989, size=.2, colour="tomato") + 
  geom_vline(xintercept=1993, size=.2, colour="dodgerblue") + geom_vline(xintercept=2001, size=.2, colour="tomato") + 
  geom_vline(xintercept=2009, size=.2, colour="dodgerblue") +
  geom_vline(xintercept=2017, size=.2, colour="tomato") +
  geom_text(aes(x=1981, label="Reagan",y=.25), colour="tomato", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=1989, label="Bush 41",y=.25), colour="tomato", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=1993, label="Clinton",y=.25), colour="dodgerblue", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=2001, label="Bush 43",y=.3), colour="tomato", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=2009, label="Obama",y=.3), colour="dodgerblue", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=2017, label="Trump",y=.4), colour="tomato", angle=90, vjust=1, text=element_text(size=9))  
  
```

## Second Layer Analysis - A More Brazen Retired Officer Corps?

Having classified what retired military officers have been addressing in publications according to topic, the next step empirically is to determine to what degree retired military officers have been upholding the two foundational norms of civilian control and non-partisanship. To measure adherence to these norms, I develop three binary indicator variables that shed light on how well the retired military officers who publish in major US newspapers adhere to these norms. 

The first indicator measures whether a publication severely criticizes a sitting President or political figure to the point that such criticism goes beyond making recommendations or pointing out shortcomings. Inclusion of such an indicator is designed to measure how willing a retired military officer is to blatantly "take a shot" at a civilian leader. Publications that go far beyond making a recommendation, and that actually are designed to sway the public's opinion regarding a specific political leader, are coded positively. Including such an indicator, moreover, enables measurement of how well retired military officers adhere to the principle of civilian control of the military. This is not to suggest that every single instance of a retired military officer criticizing a civilian leader is a crisis of civilian control of the military. However, including such an indicator will allow how frequent blatant criticism occurs.  

A second indicator is whether a publication clearly endorses or rebukes a partisan politician or party, including candidates for office. Including this indicator allows the researcher to measure the degree to which the principle of non-partisanship of the military has been upheld in the publications authored by retired military officers over time. 

The third and final indicator measures whether a publication calls on Congress or a President to undertake specific legislative actions, such as ratifying a treaty or passing a bill or budget. This indicator does not directly measure either non-partisanship or civilian control of the military, but it does measure how comfortable retired military officers feel being involved in a part of the political process, that of calling for and/or making legislative action, which is an experience that virtually all military officers do not develop during their time in uniform.^[A small number of officers may have experience working on the hill, or serving as a Congressional liaison, but I argue that this is different from actually calling upon a politician to engage in adopting a stance regarding a specific piece of legislation.] This is not to say that military officers on active duty or during retirement either should be or in fact are oblivious to this aspect of the political process. Rather, including an indicator for assertiveness (calls to pursue legislative action) measures to what extent the retired military author believes his or her military experience serves as an appropriate basis for calling for legislative action. 

Because some retired military officers in fact do go on to serve in political administrations, or run for office and serve as politicians themselves, it is important to denote whether a retired military officer wrote a publication while serving in a role that legitimately would authorize the military officer to play a more prominent role in the legislative process. Therefore, a fourth and final indicator control for whether the author serves in such a role at the time of publication is included.

```{r, include = FALSE, warning=FALSE, echo=FALSE}

# Show all articles and authors who published articles on Domestic Policy and Social Policy 
adjdataset %>% arrange(PubYr) %>% 
  filter(Topic=="DomesticPolicy" | Topic=="SocialPolicy")  %>% 
  select (Author, Title, Source, PubYr, Topic) %>% 
  print()

# Show all articles where Leg_Act, Crit_Insub, or Endorse_Partisan was equal to 1
## (in other words, articles in which the author called for legislative activity, criticized, or made
## an explicitly partisan statement)

adjdataset %>% arrange(PubYr) %>% 
  filter(Leg_Act== 1 | Crit_Insub== 1 | Endorse_Partisan ==1)  %>% 
  select (Author, Title, Source, PubYr, Topic, Leg_Act, Crit_Insub, Endorse_Partisan) %>% 
  print()


# create a df called quest_prof that contains articles that either called for legislation 
# or criticized or were expressly partisan, as well as whether someone was 
# in a role at the time that may have made such behavior ok

quest_prof <-adjdataset %>% arrange(PubYr) %>% 
  filter(Crit_Insub== 1 | Endorse_Partisan ==1 | Leg_Act ==1)  %>% 
  rename(Crit=Crit_Insub, Endorse=Endorse_Partisan, Leg=Leg_Act
  ) %>% 
  select (Author, Title, Source, PubYr, Topic, Leg, Crit, Endorse, Role) %>% 
  view ()

# Now create two variables, leg_crit_part and crit_part that take a 1 if certain 
# conditions are met.  This will come in handy for the glm regressions that
# will be conducted later.  

adjdataset <- adjdataset %>% 
  mutate(
    leg_crit_part=case_when(
      (Leg_Act==1 | Crit_Insub ==1 | Endorse_Partisan==1 & Role ==0) ~ 1,
      (Leg_Act != 1 & Crit_Insub != 1 & Endorse_Partisan !=1 | Role==1) ~ 0
    ),
    crit_part=case_when(
      (Crit_Insub ==1 | Endorse_Partisan==1 & Role ==0) ~ 1,
      (Crit_Insub != 1 & Endorse_Partisan != 1 | Role ==1) ~ 0
    )
  ) %>% 
  view()


# create a variable decade
authorcalcs <- adjdataset %>% 
  mutate(
  decade =case_when(
    (PubYr < 1980) ~ "1970s",
    (PubYr >= 1981 & PubYr < 1991) ~ "1980s",
    (PubYr >= 1991 & PubYr < 2001) ~ "1990s",
    (PubYr >= 2001 & PubYr < 2011) ~ "2000s",
    (PubYr >= 2011 & PubYr < 2021) ~ "2010s"
  )
) %>% select(Author:Service, Title:Topic, leg_crit_part, crit_part, decade) %>% print()

#count the number of publications by GO/FO author in each decade
authorcalcs %>% group_by(decade) %>% filter(AuthRank >=7) %>% select(Author) %>% count(Author) %>% view()

#create a table for the Appendix
TA1 <- quest_prof %>%
  kable(format="latex", caption="Op-Eds authored by Retired Military Officers that Called for Legislative Action, Strongly Criticized Civilian Leaders, or were Expressly Partisan, 1979-2020", longtable=T) %>%
  kable_styling(font_size=9, latex_options="striped", repeat_header_continued=T) %>% 
  column_spec(1, width="9em") %>% 
  column_spec(2, width="14em") %>% 
  column_spec(3:4, width="2.5em") %>%
  column_spec(5, width="5.6em") %>%
  column_spec(6:7, width="1.4em") %>%
  column_spec(8:9, width="3.1em") %>% 
  row_spec(0, bold=T)
```

```{r plot4, include = TRUE, echo=FALSE, warning = FALSE, error=FALSE, message=FALSE, fig.cap="Proportion of Op-Eds Which Either Argues for Legislative Action, Expressly Criticizes Civilian Leadership, or Takes Clearly Partisan Stances, 1979-2020"}

ggplot(data=adjdataset, mapping=aes(x=PubYr, y=prop_yr)) + geom_point() + geom_smooth() + labs(x="Year", y="Annual Proportion", color=NULL) + geom_vline(xintercept = 1981, size=.2, colour="tomato") + 
  geom_vline(xintercept=1989, size=.2, colour="tomato") + 
  geom_vline(xintercept=1993, size=.2, colour="dodgerblue") + geom_vline(xintercept=2001, size=.2, colour="tomato") + 
  geom_vline(xintercept=2009, size=.2, colour="dodgerblue") +
  geom_vline(xintercept=2017, size=.2, colour="tomato") +
  geom_text(aes(x=1981, label="Reagan",y=.2), colour="tomato", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=1989, label="Bush 41",y=.2), colour="tomato", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=1993, label="Clinton",y=.4), colour="dodgerblue", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=2001, label="Bush 43",y=.3), colour="tomato", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=2009, label="Obama",y=.4), colour="dodgerblue", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=2017, label="Trump",y=.05), colour="tomato", angle=90, vjust=1, text=element_text(size=9))  
```

Figure 4, above, shows over time the proportion of observations in a given year which either heavily criticizes a civilian leader, takes a clear partisan stance, or calls for specific legislative action to be taken. The smoothed trend line in blue shows a story that is similar to the frequency at which retired military officers address domestic political topics. Not until the 1990s did retired military officers frequently criticize civilian leaders, call for express legislative action, or take clear partisan stances in the publications they authored. Additionally, there is a localized minimum in the trend line around the year 2012 before a sharp increase that has persisted until the present day. 

\singlespace
## Explaining A More Assertive Retired Military Officer Corps 
\doublespace

Collectively, Figures 3 and 4 tell an interesting story: retired military officers are addressing domestic political topics more frequently and at a greater rate relative to other topics than they did in previous decades, and they are criticizing civilian leaders, taking partisan stances, and calling for legislative action at greater rates than they previously did. Said differently, retired military officers are exuding boldness through the opinion pieces they publish in ways that differ from previous decades. Having established that these trends exist, the next task is to attempt to explain their development.Fortunately, measurements for political polarization and credibility of the military exist, offering the researcher the ability conduct a quantitative analysis of the data. 

One measurement of political polarization are DW-NOMINATE (dynamic-weight, nominal three step estimation) scores, which were originally developed by the political scientists Keith Poole and Howard Rosenthal in the early 1980s [@jeffrey_b_lewis_voteview_2020]. These scores use roll-call voting data to assign member of Congress a score on the liberal-conservative dimension, ranging from -1 (extremely liberal) to 1 (extremely conservative) [@jeffrey_b_lewis_voteview_2020].^[The data is frequently relied upon by academics, journalists, and the media to understand how a variety of political bodies, to include Congress and political parties, have changed over time. Over the time period examined in this paper, polarization in each chamber of Congress, the House of Representatives and the Senate, has increased [@jeffrey_b_lewis_voteview_2020].] For instance, the distance in party mean DW-NOMINATE scores in the late 1970s was just less than .6, while the distance today for both chambers is above .8 [@jeffrey_b_lewis_voteview_2020].

A second measurement of political polarization is a measurement of affective polarization, which was first conducted by Iyengar et al using time series data from the American National Election Study [@iyengar_affect_2012; @iyengar_origins_2019]. Using a series of "feeling thermometer" questions that gauge a person's feelings about those on the opposite side of one's political spectrum, this measurement roughly captures how much "dislike" exists among the electorate for each other [@iyengar_strengthening_2018]. This measurement does not reflect an encouraging story: while the level at which citizens like their own party has remained remarkably stable for the past 40 years, the level of affection for those of the opposite political party precipitously dropped around the year 2000, a trend that has continued [@iyengar_strengthening_2018]. I include both measures of polarization - among members of Congress and among the electorate - in the analysis to follow. 

Using Gallup public opinion data, I include a measurement of the credibility of the American military as an institution. These surveys have asked Americans to rate the level of trust that they place in a variety of American institutions, such as the Presidency, Congress, Churches, and the Police. Since the late 1970s, the credibility of the American military as an institution has gradually increased [@noauthor_confidence_2020]. In 1979, 54% of those surveyed expressed either a great deal or quite a lot of confidence in the military, while in 2020, 72% expressed the same [@noauthor_confidence_2020]. Trends in political polarization (using DW-Nominate Scores and a measure of Affective Polarization) and confidence in the military are displayed in Figure Four, below.  

```{r, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
#adjust the data to account for polarization by creating a new dataset called polaradjdataset.  
#In this dataset, create variables that capture ideological distance in each the House and Senate along each of the two dimensions according to Poole and Rosenthal's DW-NOMINATE data
# and a variable for affective polarization

# scale the data for congressional polarization by multiplying by 100

polaradjdataset <-adjdataset %>% mutate (
  House_distance_dim1=abs(House_Dem_mean_nom1-House_Rep_mean_nom1)*100,
  Sen_distance_dim1=abs(Sen_Dem_mean_nom1-Sen_Rep_mean_nom1)*100,
  House_distance_dim2=abs(House_Dem_mean_nom2-House_Rep_mean_nom2)*100,
  Sen_distance_dim2=abs(Sen_Dem_mean_nom2-Sen_Rep_mean_nom2)*100,
  aff_polar=dem_diff_aff+`rep-diff_aff`
) %>% view ()

# Create a variable called polar_and_cred that uses the melt function (from the library reshape2")
#note that the id.var will serve as the common axis for these 
polar_and_cred <- melt(subset(polaradjdataset, select=c(PubYr, House_distance_dim1, aff_polar_alt, Inst_Cred_Gallup)), id.var="PubYr")
```

```{r plot5, inculde=TRUE, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, fig.cap="Polarization Levels in the US House of Representatives, Affective Polarization Levels Among the US Electorate, and American Confidence in the Military, 1979-2020"}
# 
ggplot(polar_and_cred, aes(x=PubYr, y=value)) + geom_line(aes(color=variable)) + facet_grid(variable~., scales="free_y") + theme(legend.position = "none")
```


Because there are only three observations written exclusively by retired female officers, and because there is no compelling logic for why the service of the author would increase the likelihood of violating norms when publishing an opinion piece, I do not control for the sex or service affiliation of the author. I do include source fixed effects in case there is concern that increasing polarization may reflect the fact that individual newspaper sources or other media outlets may have taken on a more-pronounced ideological bent along the political left-right spectrum as polarization has worsened, and thus, the data merely reflect the fact that that certain newspaper sources are more inclined to publish pieces that espouse a point of view reflecting the source's political leanings.^[For example, the _New York Times_ in 2020 is deemed to lean further to the political left than the current  _Wall Street Journal_, which some see as aligned with the political center or slightly right of center [for an informative view, see the Media Bias Chart created by allsides.com [@noauthor_allsides_nodate].] 

The dependent variable is a binary indicator of whether the opinion piece either called for legislative action, blatantly criticized a political leader, or clearly took a partisan stance on an issue. Because some may argue that calling for legislative action in itself does not violate norms of civilian control of the military or the non-partisanship of the military institution, the dependent variable in alternate models is an indicator for whether a publication either criticized a political leader or took a clearly partisan stance (and does not include instances in which a publication called for legislative action). The estimated model therefore takes the form: 

$$y = \beta_0 + \beta_1 MilitaryCredibility + \beta_2Polarization + \beta3AuthorRank + \beta_4Source +\epsilon $$

Because the dependent variable is dichotomous, a logistic regression is most appropriate. I run several model specifications including a regular logit model, a logit model that clusters standard errors at the author level, and a mixed model that introduces random effects at the author level. The intent of including models which cluster standard errors or introduce random effects at the author level and introduce is to determine the degree to which the data can distinguish if the individual authors themselves are an important determinant of whether an opinion piece criticizes civilians, calls for legislation, or takes a partisan stance. In other words, the logit model without clustered errors at the author level assumes that authors have no distinguishing characteristics that would incline them towards behavior that would result in the dependent variable taking on a value of 1, whereas both the mixed model with author random effects and the logit model that clusters errors at the author level allows for the possibility that individual author characteristics may explain the dependent variable in important ways that the model does not otherwise account for. The regressions proceed on 379 of the 390 observations (97.2%) of the sample. Due to complete separation between the dependent variable involving sources with a few number of observations (_Los Angeles Times_ and _USA Today_), observations from these sources are dropped.  

The results are displayed in Table 4, below.  

```{r logit analysis, include=FALSE, echo=FALSE, warning=FALSE}
# check for separation (complete or quasi-complete)
polaradjdataset %>% 
  filter(Leg_Act == 1 | Crit_Insub ==1 | Endorse_Partisan == 1) %>% group_by(Source) %>% 
  view()

#next, conduct logit regressions where the DV is first, values of leg_crit_part and
# then, where the DV is crit_part
# model 1 

mod1 <- glm(leg_crit_part ~ Inst_Cred_Gallup + House_distance_dim1 + AuthRank,
    family=binomial(link="logit"),
    data=polaradjdataset
    )
summary(mod1)
tidy(mod1)
glance(mod1)

# model 2 - logit with different DV

mod2 <- glm(crit_part ~ Inst_Cred_Gallup + House_distance_dim1 + AuthRank,
    family=binomial(link="logit"),
    data=polaradjdataset
    )

summary(mod2)
tidy(mod2)
glance(mod2)

#model 3 - logit using aff polar 
mod3 <- glm(leg_crit_part ~ Inst_Cred_Gallup + aff_polar_alt + AuthRank,
    family=binomial(link="logit"),
    data=polaradjdataset
    )

summary(mod3)

#model 4 - logit using aff polar with different DV

mod4 <- glm(crit_part ~ Inst_Cred_Gallup + aff_polar_alt + AuthRank,
    family=binomial(link="logit"),
    data=polaradjdataset
    )

summary(mod4)

# models 5-8 include Source as an IV, but the standard error estimates are too large
# likely because there is not enough variation within Sources (LA Times is the problem)

mod5 <- glm(leg_crit_part ~ Inst_Cred_Gallup + House_distance_dim1 + AuthRank + Source,
    family=binomial(link="logit"),
    data=polaradjdataset
    )
summary (mod5)


mod6 <- glm(crit_part ~ Inst_Cred_Gallup + House_distance_dim1 + AuthRank + Source,
    family=binomial(link="logit"),
    data=polaradjdataset
    )
summary (mod6)

mod7 <- glm(leg_crit_part ~ Inst_Cred_Gallup + aff_polar_alt + AuthRank + Source,
    family=binomial(link="logit"),
    data=polaradjdataset
    )
summary(mod7)

mod8 <- glm(crit_part ~ Inst_Cred_Gallup + aff_polar_alt + AuthRank + Source,
    family=binomial(link="logit"),
    data=polaradjdataset
    )
summary(mod8)

# I am now going to drop observations from the LA Times and USA Today  which will enable me to obtain some #better, more #precise estimates of the data. 

xtabs(leg_crit_part ~ Source, polaradjdataset)
xtabs(crit_part ~ Source, polaradjdataset)
slimdata <- polaradjdataset %>% filter(Source != "LA Times" & Source !="USA Today") %>% print()
xtabs(leg_crit_part ~ Source, slimdata)
xtabs(crit_part ~ Source, slimdata)

slimdata %>% view()

slimdata %>% filter(leg_crit_part ==1) %>% group_by(Author) %>% count(Author) %>% arrange(n) %>% view() 

slimdata %>% filter(crit_part ==1) %>% group_by(Author) %>% count(Author) %>%  arrange(n) %>% view ()

# Now run a straght logit GLM including which includes Source as an IV
# Them conduct model with robust standard errors at the author level
# Then conduct a mixed model where we include author random effects

# DV is leg_Crit_part, house_distance_dim1 is IV
mod9 <- glm(leg_crit_part ~ Inst_Cred_Gallup + House_distance_dim1 + AuthRank + Source, family=binomial(link="logit"), data=slimdata)
summary(mod9)

mod9robust <- robustify(mod9, cluster=Author)
summary(mod9robust)

mod9mixed <- glmer(leg_crit_part ~ Inst_Cred_Gallup + House_distance_dim1 + AuthRank + Source + (1 | Author), family=binomial, data=slimdata)
summary(mod9mixed)

# DV is leg_Crit_part, aff_polar_alt is IV

mod11 <- glm(leg_crit_part ~ Inst_Cred_Gallup + aff_polar_alt + AuthRank + Source, family=binomial(link="logit"), data=slimdata)
summary(mod11)

mod11robust <- robustify(mod11, cluster=Author)
summary(mod11robust)

mod11mixed <- glmer(leg_crit_part ~ Inst_Cred_Gallup + aff_polar_alt + AuthRank + Source + (1 | Author), family=binomial, data=slimdata)
summary(mod11mixed)

# DV is crit_part, IV is house_distance_dim1

mod10 <- glm(crit_part ~ Inst_Cred_Gallup + House_distance_dim1 + AuthRank + Source, family=binomial(link="logit"), data=slimdata)
summary(mod10)

mod10robust <- robustify(mod10, cluster=Author)
summary(mod10robust)

mod10mixed <- glmer(crit_part ~ Inst_Cred_Gallup + House_distance_dim1 + AuthRank + Source + (1 | Author), family=binomial, data=slimdata)
summary(mod10mixed)

# DV is crit_part, IV is aff_polar_alt

mod12 <- glm(crit_part ~ Inst_Cred_Gallup + aff_polar_alt + AuthRank + Source, family=binomial(link="logit"), data=slimdata)
summary(mod12)

mod12robust <- robustify(mod12, cluster=Author)
summary(mod12robust)

mod12mixed <- glmer(crit_part ~ Inst_Cred_Gallup + aff_polar_alt + AuthRank + Source + (1 | Author), family=binomial, data=slimdata)
summary(mod12mixed)



```
\singlespace
```{r table4, include=TRUE, echo=FALSE, results="asis", warning=FALSE, message=FALSE}
stargazer(mod9, mod11robust, mod9mixed, mod12, mod10robust, mod12mixed,
          type="latex",
          title="Regression Results", 
          covariate.labels = c('Mil Cred', 'Cong. Polar', 'Aff. Polar', 'Rank', 'SourceWaPo'),
          model.names=TRUE,
          multicolumn=FALSE,
          column.sep.width = "-15pt",
          dep.var.caption = "Odds (log) of Opinion Publications Authored by Retired Military Officers",
          dep.var.labels.include=FALSE,
          column.labels = c('Legislation, Criticize Civilians, or Partisan', 'Criticize Civilians or Partisan'),
          column.separate=c(3, 3),
          align=TRUE)
         
```


\doublespace

The results shown in Table 3 indicate that across five of six model specifications, the coefficient for polarization, regardless of how it is measured, is statistically significant. In models 1-3, where the dependent variable is an indicate for whether an opinion publications calls for legislation, criticizes civilians, or takes an expressly partisan stance, all three model specifications show that the measure of polarization plays a significant role in determining the odds of the dependent variable occurring. Substantively, a coefficient of .087 in Model 1 (logistic regression) indicates that a one unit increase in the mean ideological distance between political parties as measured by roll call voting is associated with a 9.1 percent increase in the odds that a publication authored by a retired military officer calls for legislative action, blatantly criticizes a civilian official, or expresses a clearly partisan stance over an issue. In model 2 (logistic regression with clustered standard errors at the author level), a coefficient of .029 indicates that a one unit increase in the level of affective polarization among the electorate is associated with a 2.9 percent increase in the odds of a publication calling for legislative action, criticizing a civilian official, or expressing a clearly partisan stance over an issue. 

While these coefficients appear to be of different size, scaling the measures of polarization are important. For instance, within the House of Representatives in the 112th Congress, which began in January 2011, polarization levels were nearly six points higher than the level that existed in the 111th Congress. Such a drastic change in the polarization level within Congress would roughly predict a 54.6 percent increase in the odds of a publication retired by a military officer calling for legislative action, criticizing a civilian leader, or taking a clear partisan stance. Regarding changes in the levels of affective polarization, data from the 2012 ANES indicate that the aggregate level of affective polarization increased by just over 16 points from the previous Presidential Election year (2008), which would correspond to a 46.4 percent increase in the odds of a publication authored by a retired military officer calling for legislative action, criticizing a civilian official, or taking a clear partisan stance. Thus, the coefficients for the respective levels of polarization are substantively similar. Even in Model 3 (Mixed Model with author random effects), the coefficient for polarization increases from its estimate in Model 1 to a value of .091, which corresponds to a 9.5 percent increase in the odds of a publication calling for legislative action, criticizing a civilian official, or taking a clear partisan stance for every one unit increase in the mean ideological distance between political parties as measured by roll call voting in Congress. Because the coefficient in Model 3 is still statistically significant even after introducing author random effects, we would interpret this as signifying that the individual author characteristics do not substantively weaken the impact of polarization. 

When the dependent variable no longer includes whether a publication calls for distinct legislative action to be taken, and thus, indicates either criticism of a civilian official or taking a clearly partisan stance, the models (models 4-6) tell a very similar story with the exception of the mixed model (model 6). Compared to the estimates of the coefficients in Models 1 and 2, the coefficients in Models 4 and 5 are relatively larger. A coefficient for polarization in Model 5 of .100 which corresponds to a 10.5 percent increase in the odds of a publication criticizing a civilian official or taking a clear partisan position with every one unit increase as measured by Congressional roll call voting. A coefficient of .037 in Model 4 corresponds to a 3.7 percent increase in the odds of a publication criticizing a civilian official or taking a clear partisan position with every one unit increase in the level of affective polarization among the electorate. 

In Model 6, which includes author random effects, the coefficient for affective polarization is no longer statistically significant, and thus, we cannot rule out the possibility that individual author characteristics are driving changes in the dependent variable, as opposed to the level of affective polarization itself. Such an interpretation also suggests that not every individual author reacts in the same way to increasing levels of polarization - which actually confirms much of the theoretical literature about polarization. Still, the fact that coefficients measuring polarization in two ways are significant in five of six models confirms the role of polarization in presenting publications authored by retired military officers that deviate from the norms of civilian control of the military and non-partisanship. If we conceptualize the commentary proffered by retired military officers in major newspapers as a significant portion of the overall messaging from the military community, even if it is offered by a relatively small number of total retired military officers, the impact of polarization on adherence to military norms is striking.

In terms of the other coefficients besides measurements of polarization, three points are striking. The first is that coefficients for the estimated impacts of military credibility are statistically significant in only two of six models, but that in all six models, the sign of the coefficient is negative. Relative to the consistent statistical significance of the coefficients for measurements of polarization, we conclude that changes in military credibility have less explanatory power, if any, in whether a publication authored by retired military officers fails to adhere to norms of civilian control and non-partisanship. The second interesting point is that the author's rank is significant in also just two of six models, and so we cannot conclude that author rank explains significant changes in our dependent variable of interest. The final point is that across all model specifications, not a single coefficient for source is statistically significant. This suggests that even if all of the three examined sources for the regression (Washington Post, Wall Street Journal, and New York Times) have changed over time as a result of polarization in terms of the types of content each source is willing to publish, changes in an opinion or commentary publication authored by a retired military officer breaking norms of civilian control or non-partisanship have not depended on which particular source the author publishes in. Said differently, regardless of the specific newspaper source, the changes in commentary or opinion publications authored by military retirees have been driven more by changes in polarization than any other variable controlled for across the different models. 

We can also use the regression results from Table 3 to transform the results of our logistic regressions into predicted probabilities. Using the results of the logistic regression in Model 4, which included a measure for affective polarization, we set the source as the _New York Times_ and hold values of institutional credibility of the military and author rank at their respective mean values of 71.1 and 8.39. These predicted probabilities are graphed in Figure 6, below. 

```{r predicted probabilities, include=FALSE, warning=FALSE, message=FALSE, error=FALSE}
# Use Model 4 (Logit) from Above, where DV is crit_part and there are no Source Fixed Effects
# Hold values of Author Rank and Institutional Credibility at their mean
# Vary the Level of Affective Polarization from its minimum value to an arbitrary 
# level of 120, which is higher than it currently is but not theoretically impossible.  

predictiondata <- tibble(aff_polar_alt=seq(min(slimdata$aff_polar_alt), 100, by=1), Inst_Cred_Gallup = mean(slimdata$Inst_Cred_Gallup), Source="NYT", AuthRank=mean(slimdata$AuthRank)) %>% crossing(crit_part = c(0,1))

predictiondata
augment(mod12, newdata = predictiondata)


# prediction data
predictions <- augment(mod12, newdata=predictiondata) %>% 
  mutate(
    .prob=plogis(.fitted),
    conf.low = plogis(.fitted -(qnorm(.975) * .se.fit)),
    conf.high = plogis(.fitted + (qnorm(.975) * .se.fit)),
    pred_final_rec = case_when(
      .prob >= .5 ~ "Likely",
      .prob < .5 ~ " Not likely"
    )
  ) 

predictions %>% arrange(.prob, .fitted, crit_part, desc(.fitted)) %>% 
  view()

# now let's try some graphs
# notice that the predictions are not in (0, 1)
ggplot(predictions) +
  aes(x = .fitted) +
  geom_histogram()

# if we wanted predictions on the probability scale,
#   we predict on the log scale,
#   then apply inverse-logit (logistic CDF)
# use plogis(): cumulative probability of the logistic distribution

# demonstration: scatter log odds (x) and predicted probability (y)
# with 95% CIs
# (CIs don't look smooth in this example because .se.fit depends on X data)
predictions %>%
  ggplot() +
  aes(x = .fitted, y = .prob) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high)) +
  geom_smooth() 

```

```{r plot6, include=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Predicted Probability of Polarization\n on Retired Military Elite Opinion Publications"}
predictions%>%
  ggplot() +
  aes(x = aff_polar_alt, y = .prob) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high)) +
  geom_smooth() + labs(
    x="Level of Affective Polarization\n Among the US Electorate",
    y="Predicted Probability of an Op-Ed Published in NYT\n by Retired Military Elites which Criticizes Civilians\n or Takes a Clear Partisan Stance"
  ) + scale_y_continuous(breaks=seq(-.1, .8, .1)) + 
  geom_vline(xintercept = 48.81, size=.2, colour="tomato") + 
  geom_vline(xintercept=59.62, size=.2, colour="tomato") + 
  geom_vline(xintercept=68.001, size=.2, colour="dodgerblue") + 
  geom_vline(xintercept=84.4, size=.2, colour="dodgerblue") +
  geom_vline(xintercept=79.55, size=.2, colour="tomato") +
  geom_text(aes(x=48.81, label="Reagan, 1980 ANES",y=.3), colour="tomato", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=59.62, label="Bush 43, 2000 ANES",y=.4), colour="tomato", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=68.001, label="Obama, 2008 ANES",y=.4), colour="dodgerblue", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=84.4, label="Obama, 2012 ANES",y=.52), colour="dodgerblue", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=79.55, label="Trump, 2016 ANES",y=.50), colour="tomato", angle=90, vjust=1, text=element_text(size=9))  
```

As the x axis of Figure 6 reveals, the highest levels of affective polarization within the electorate were recorded during the ANES administered in 2012, in the months leading up to the start of President Obama's second term. In fact, the level of affective polarization actually dropped as measured by the ANES fielded in 2016, as President Trump won the Presidency. Figure 6 shows a clear increase in the predicted point estimate of the probability of a publication authored by a retired military officer failing to adhere to the norms of civilian control of the military and/or non-partisanship as the level of affective polarization among the electorate varies. Using this model shows, in fact, that at ANES levels that existed at the time of the election of President Reagan in 1980, the predicted probability of an opinion publication in _The New York Times_ authored by a retired military officer failing to adhere norms of professionalism or non-partisanship was roughly 8.0 percent. At ANES levels that existed when President Trump was elected, that probability increased to 21.7 percent. It remains to be seen how much worse polarization will become, but the predicted impacts of any increase in polarization do not bode well for the future of military professionalism. If we project that the nation is now back at the level of affective polarization as that which existed as measured at the time of President Obama being elected to a second term, and should the same level of change in affective polarization occur that took place between President Obama's first and second terms (roughly 16 points in the level of affective polarization), the model would predict a nearly 37% probability that an opinion publication published in the the _New York Times_ and authored by a retired military officer would violate norms of civilian control or non-partisanship. 

In terms of the four hypotheses this paper set out to explore and examine, the data seems to tell a convincing story about the relative impact of increasing political polarization on the propensity of retired military officers adhering to the norms of civilian control of the military and non-partisanship, and particularly whether these norms are followed in the opinion publications retired officers seek to publish in major US newspapers. The regression analysis indicated that in fact, the level of political polarization is a statistically significant indicator of the likelihood of retired military elites criticizing civilian officials or making clearly partisan statements, thus lending confirmation of _H1_. Additionally, the descriptive statistics, and particularly the trend in Figure 3, indicated that increasing levels of political polarization are associated with military retirees addressing domestic political topics, thus providing support for _H2_. Figure 3 also revealed in a general sense that the frequency with which retired military elites write commentary or opinion pieces in major US newspapers is associated with an increase in the level of prestige and credibility the military institution enjoys among American society. Finally, the data indicated that periods of active ground combat tend to focus retired military elites to write on topics related to combat. This finding is supported by Figures 1, 3 and 4, each of which revealed a clear local minimum in the number of publications authored by retired military officers (Figure 1), the proportion of publications that addressed domestic topics (Figure 3), and the proportion of publications calling for legislative action, criticizing civilian leaders, or making boldly partisan claims. These local minima occurred during the periods of surge operations in Iraq and/or Afghanistan during the late 2000s through early 2010s, when the US military had been engaged overseas for several years with frequent deployments, enduring a steady frequency of American casualties. This confirms the broad insight previous scholars have advanced regarding how the demands of combat and war-fighting tend to focus certain professional behaviors of the military [@snider_once_2012]. 

# Can the Military Remain Professional?

This study demonstrated that although the vast number of opinion publications authored by retired military officer uphold the military's norms of civilian control and non-partisanship, worrisome trends are evident that suggest retired military elites are increasingly failing to uphold the central norms of civilian control of the military and the non-partisanship of the military institution in a significant portion of the messaging that they conduct with the American public: deliberately written opinion and/or commentary pieces which retired military officers author in major US newspapers. The data clearly reveal that the manner in which retired military officers address the American public has changed with respect to upholding military norms and standards over the past four decades, even while holding the source of a newspaper constant. These trends are alarming for at least three reasons, and speak not only to the degree of seriousness of the problem, but the role that political polarization has played in driving certain members of the retired military officer community to engage in this behavior.  

A primary reason that these trends are alarming stems from the influence that even a small number of retired military elites can potentially have. While one can argue in theory that Americans should not be concerned if a handful or even several dozen retired military officers break certain norms of professionalism on the basis that there are more than 7,000 living retired general and flag officers, such a figure should not serve as a baseline comparison when analyzing the particular remarks of retired military officers that are published in prominent US newspapers. Only a small number of pieces are chosen to be published by the editorial boards of major newspapers whose circulations range in the tens of millions, and it is important that this small number of retired military officers showcase to the American public if not a unifying message, at least a unifying adherence to the fundamental norms of civilian control of the military and non-partisanship that all members of the military institution are expected to follow. 

A second reason why these trends are concerning, I argue, is because members of the active duty force still look, in many cases, to these retired officers who are violating norms of professionalism as role models and figures whose behavior is worthy of emulation. This is not to suggest that every single instance of a retired military officer breaching a professional norm has a major impact on the military profession, but it is to suggest that at some point, many instances of behavior constitute a pattern. The evidence presented in this paper, I argue, presents clear patterns of failing to uphold foundational norms of professionalism among the retired military community. At this point, considering that several active duty military generals have previously attempted to "reign in" or at least plead with members of the retired military community to limit behaviors which would tarnish the image of a professional military, it is doubtful that anything less than numerous and urgent pleas from fellow retired military elites and current four star generals and admirals would work to reign some of the most blatant offenders of professional norms.  Even then, however, the outcome is likely uncertain whether such actions would attenuate these sorts of behaviors.  

A third and final reason that the data presented in this paper is concerning is that amid all of the cries from pundits and scholars alike of civilian leaders acting to "politicize the military," a fair bit of politicization of the military is occurring via the written words of retired military officers. For an institution such as the military that espouses to be a profession that looks inward and polices the behavior of those who stray from professional norms, it is time that retired military officers - and especially, the community of retired flag and general officers - look inward, and ask both themselves and their peers, "are we upholding professional norms in everything that we do and say?" If the answer to this question is "no," these retired military officers should put partisan preferences aside and commit themselves to solving the problem. In an age marked by increasing polarization and divisiveness, however, this may just require retired military officers to perhaps do the hardest things of all: put the pen down and hold one's tongue.

One of the long-term challenges facing the active military in an age of political polarization is integrating the multiple identities each soldier carries with them in such a way that each can still proudly wear the uniform and bear the identity of an American service member. Though this challenge is not new for the US military, increasing political polarization has significantly raised the level of difficulty the military faces in tackling such a vital challenge. Scholars who have studied multi-ethnic militaries across the world have argued that it is military leaders, not civilians, who ultimately devise the most effective solutions to the problems posed by soldiers who carry multiple loyalties, and interestingly, only as a result of combat requirements [@peled_question_1998]. Perhaps the American challenge is even greater in that at least for the near term future, there is no major war on the immediate horizon that will bring about the coming together of disparate loyalties and identities among future soldiers. While the knowledge, skills, and experiences of retired military officers are important, and can and should be used to help solve myriad problems facing the country, the impacts of political polarization are threatening to undermine the commitment of certain members of the military, to include military retirees, to professional norms. Without addressing and correcting the trend of retired military officers violating professional norms swiftly, there could be significant repercussions on the military's ability to gain and maintain public trust, as well as impacts on how future generations of military officers conceive of their normative role in politics. 
\singlespace
\pagebreak

# Appendix 1

\doublespace
## Word and Data-Base Searches to Obtain Data

I did two different sets of searches.  This first set of search terms was applied to the Factiva and Nexus-Uni data bases. In each of these databases, I had to select the "type of article" and included options such as commentary/opinion and letters to the editor.
The specific terms of search included:

"retired Army General" or "former Army General" or "retired Navy Admiral" or "former Navy Admiral" or "retired Marine Corps General" or "former Marine Corps General" or "former Air Force General" or "retired Air Force General" or "retired Coast Guard Admiral" or "former Coast Guard Admiral" or "retired from the Marine Corps" or "retired from the Army" or "retired from the Navy" or "retired from the Air Force" or "retired from the Coast Guard" or "supreme commander of" or "commander of" or "Chairman of the Joint Chiefs" or "chairman of the joint chiefs" or "Supreme Commander of" or "Chief of Staff" or "Chief of Naval Operations" or "retired Army" or "retired Navy" or "retired Air Force" or "retired Marine Corps" or "retired Coast Guard" or "retired military" or "commandant of" or "director of" or "retired from the military"

The second set of search terms was applied to Pro Quest Historical Newspapers, which the University of Wisconsin had access to for both the Wall Street Journal and the New York Times. Factiva included article from the Wall Street Journal only back to the middle of 1994, so I relied heavily on Pro Quest to search through the Wall Street Journal from 1979-1994. Unlike the search terms used for Factiva and Nexus-Uni, however, I came to quickly realize that the opinion pieces I sought were not categorized as opinion-editorials or letters to the editor, but rather as "articles."  This required significantly more time searching for the desired data, as more articles had to be searched through and sifted. The specific terms of search included:

(retired OR former OR “chief of staff”) AND (“Army General” OR Gen. OR general OR "Navy Admiral" OR Adm. OR admiral OR "Marine Corps General" OR "Coast Guard Admiral" OR "Air Force General" OR "Chairman of the Joint Chiefs" OR "colonel" OR "captain" OR "commander" OR "chief of naval operations" OR "commandant" OR "director of")

## Observations used for Regression Analysis

The following table contains those observations whose content either called for legislative action, criticized civilian leaders, or took clearly partisan stances.

\singlespace

```{r table5, include=TRUE, echo=FALSE}
TA1
```

First, 17 of the 49 publications are, upon a closer reading, determined to not reflect possible violations of military professionalism. This is worth showcasing in order to draw attention to the fact that there are instances of retired military officers authoring opinion pieces which address the subjects of social policy related to the military, domestic policy, and figures by name while simultaneously upholding standards of military professionalism. However, the data reveals that in practice, such instances are questionable. The practical implication is simple and straightforward: if military officers decide to write about topics other than war fighting and operations, troop support and institutional culture, or strategic advice, they need to carefully craft their argument so as to not violate norms of military professionalism.

Second, and resulting from the first, is that across the entire data set, instances of potential violations of military professionalism in the op-eds published by retired military officers in major US newspapers is a somewhat rare occurrence. Indeed, the entire count of potentially questionable publications with respect to military professionalism across the data set is just 32 of 344 total observations, or 9.3% of the entire sample. Thus, ample support exists to confirm the first hypothesis, that the vast majority of opinion pieces authored by retired military officers uphold standards of military professionalism. 

A third and interesting finding, however, is that the first publishing year in which an opinion piece penned by a retired officer potentially failed to uphold professional standards is 1998, nearly halfway across the sample in terms of publishing year. Why is this the case? We will deal more with the suspected culprits of this trend, rising political polarization and rising institutional credibility of the military, later in this paper. The point to make is that without question, retired military officers before 1998 had plenty of opportunities either to air partisan preferences and/or to address topics that may have similarly pushed the limits of military professionalism. In fact, many of these retired officers would have been veterans of the Vietnam and Korean Wars, and perhaps, like Maxwell Taylor, an author who is in the sample data, veterans of World War Two. Moreover, the data shows that these retirees wrote opinion pieces ( a total of 46 observations were written before the first observation in Table 4 raises a question about military professionalism), but that in doing so, they stayed clearly within the bounds of military professionalism. 

Fourth, Table 4 reveals that several retired military officers wrote publications not from the perspective of a retired military officer, but from the perspective of other professional and political roles that these officers served in after retiring from the military. These include op-eds by: Retired Air Force Lieutenant General Claudius Watts, who wrote an op-ed regarding the advantages of all-male education while serving as the president of the Citadel; Retired Army General Barry McCaffrey, who penned an opinion piece while serving in the administration of Bill Clinton as the Director of the Office of National Drug Control Policy; and Retired Marine Corps Colonel John Murtha, who also served in the US House of Representatives. Each of these authors addressed topics that one should reasonably expect them to on the basis of the roles each was serving in. This would include, in the case of John Murtha, making obviously partisan agreements and thus, revealing political preferences. 

Table 4 reveals, fifth, that several publications are deemed as possibly having violated military professionalism in that they address topics that may be considered outside of the realm of proper military expertise. It is worth briefly commenting on several of these examples. In his _Washington Post_ editorial, "Home Should Not Be a War Zone," Retired Army General Stanley McChrystal addresses the tragic nature of mass shootings in America. He states in the article, "...as a combat veteran and proud American, I believe we need a national response to the gun violence that threatens so many of our communities" [@mcchrystal_home_2016]. He later, in 2018, pens an op-ed entitled "Good Riddance" in which he discusses the evolution of his views on Robert E. Lee and other Confederate soldiers and symbols [@mcchyrstal_stanley_good_2018]. Retired Army Brigadier General Ty Seidule similarly, in 2020, advocates for the renaming of several Army installations [@seidule_ty_what_2020]. Other examples of observations coded as potentially violating military professionalism based on the topic of the publication include articles penned by Retired Navy Admiral Mike Mullen (entitled "The Refugees We Need") and co-authors Retired Army Lieutenant General Mark Hertling and Retired Naval Admiral Robert Natter (entitled "Cutting Refugee Admissions Will Have Severe Consequences for the US Military").  

These articles are certainly not unreasonable in that they, in the opinion of the author of this article, express reasonable thoughts and ideas to many of the nation's pressing problems, including gun violence, the appropriate legacy of Confederate symbols, and national immigration policy. To be clear, many Americans agree with the stance these officers take in the articles they have published. One can even make a strong, perhaps even a compelling argument, why these officers might be uniquely qualified to address aspects of some of these topics in that they possess expertise commanding large and powerful organizations charged to carry out violence on behalf of the country (related to the issue of gun control), that they may have experience serving and leading troops on bases named after Confederate generals (related to the issue of Confederate symbols and bases named after Confederate generals), and that they have have considerable experience serving alongside soldiers of multiple ethnicities as well as considerable experience in Iraq and Afghanistan (related to the issue of improving American immigration policy). 

But I contend that it is also not wrong to insist that these and other contentious national issues are most appropriately solved by elected civilian officials and not by retired military officers. Interestingly, in July of 2020, General Mark Milley, the Chairman of the Joint Chiefs of Staff, was asked by US Representative Anthony Brown, D-MD, during a hearing held by the House Armed Services Committee, to "comment on the naming of Army installations after Confederate soldiers," and whether the continued use of the names of these bases "reflects the values" that the military is trying to instill in the nation's armed forces [@milley_mark_general_2020]? Milley's response was instructive. He offered the Congressmen the idea that even as the Department of Defense continues to take a hard look at the issue, the decision to remove those names should reflect the decision to adopt those names in the first place in that such a decision was "a political decision" [@milley_mark_general_2020]. Milley's answer was the appropriate one, I contend, because it recognized that regardless of how he feels about this issue, the issue is not for him (or the military) to solve. Many other topics addressed by publications listed in Table 4, to include gun violence and immigration, fall into the same realm.

A sixth finding is that there are roughly three clusters of observations in Table Four with respect to the year of publication. In the first cluster (1998-1999), the publications reveal multiple grievances between certain retired military officers and the manner in which President Clinton was planning or using military force abroad. Only two of the five observations in this first cluster are coded as possibly having violated standards of military professionalism. 

The second cluster refers to publications penned between 2004-2006. Most of these comprise part of or are related to the so-called "Revolt of the Generals," which occurred near the 2004 US midterm elections and referred to sharp disagreements that entered the public arena between multiple US generals and admirals and President George W. Bush's Secretary of Defense, Donald Rumsfeld. These disagreements were directly related to these military leaders' disagreement about Rumsfeld's leadership regarding the planning and execution of the early years of the war in Iraq. Retired Army General Tommy Franks penned two observations that essentially provided public support for the President, George W. Bush, while Retired Army Major General Paul Eaton, a critic of Rumsfeld, authored two of the pieces published in this cluster. Of the seven observations published in Table 4 between 2004-2006, all but one contained possible violations of the concept of military professionalism. 

The third and largest cluster occurs between 2016-2019.  The 23 observations in this cluster reveals concern over the candidate and later President Donald Trump. The titles alone of many of these opinion pieces reveal, at a minimum, clear partisan preferences by many of the authors. Retired Air Force General Michael Hayden, Navy Admirals Mike Mullen and William McRaven, and Army General Stanley McChrystal are among the retired officers who authored several publications during this time period. All but four of the observations in this cluster are coded as having potentially violated standards of military professionalism. Of note, one observation in this cluster entitled, "Fists Raised at West Point" and published in the Washington Post is actually a compendium of two anonymous letters, each one taking an opposite stance on the appropriateness of an incident in which a photograph at West Point's 2016 graduation ceremony captured a graduating female cadet raising a clenched a fist, an action which some saw as an inappropriate political gesture related to the growing Black Lives Matter movement [@anonymous_author_fists_2016].  

\singlespace
\pagebreak









