---
title: "True Colors? The Military Institution and Presidential Campaign Advertisements, 2000-2016"
author: "Peter M. Erickson"
date: |
  `r format(Sys.time(), '%B %d, %Y')`
abstract: |
  
  _Using data from the Wisconsin Advertising Project (WiscAd) and the Wesleyan Media Project (WMP) consisting of Presidential Campaign television advertisements used in the five Presidential Elections from 2000 -2016, this chapter examines variation in the use of military figures and symbols in presidential campaign television advertisements. The results of this large-N, quantitative analysis indicate that civilian political candidates are increasingly using military actors (to include retired military officers, veterans, and veteran family members) to undertake clearly partisan behaviors while appearing in television campaign advertisements. The statistical results suggest that variation in this type of political behavior undertaken primarily at the hands of civilian actors is the result of changes in the overall level of the military's prestige. These results provide support for the notion that increasing levels of military prestige provide an opportunity for civilian actors to involve the military in politics in ways that challenge the central principles of civil-military relations._  
  
geometry: margin = 1.15in              # margin widths
fontsize: 12pt                         # text size
bibliography: dissertation.bib         # bibliography file
biblio-style: authoryear               # citation style
urlcolor: blue                         # links to external URLs
citecolor: black                       # links to bibliography
linkcolor: magenta                     # internal figures, footnotes, etc.
indent: true                          # paragraph format
                                       # true for indented paragraphs
                                       
header-includes:
    - \usepackage{setspace}
    - \usepackage{multirow}
    - \usepackage{indentfirst}
    - \usepackage{caption}
    - \usepackage{rotating, graphicx}
   
    
output: 
  bookdown::pdf_document2:
    keep_tex: true
    latex_engine: pdflatex
    number_sections: false
    toc: false
    fig_caption: true
    citation_package: biblatex
    includes: 
      in_header: 
        - latex-preamble.tex
  bookdown::htmldocument2: default
  bookdown::word_document2:
      toc: false
---

\doublespacing 

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo=FALSE,
  include=FALSE,
  message=FALSE,
  error=FALSE,
  fig.height=3,
  fig.width=4,
  out.width="90%")
```

```{r, include = FALSE, echo = FALSE}
library("here")       # file paths
library("tidyverse")  # workhorse package
library("tidylog")
library("kableExtra")
library("knitr")
library("ggdag")
library("dagitty")
library("gridExtra")
library("broom")
library("dplyr")
library("reshape2")
library("pacman")
library("ggrepel")
library("cowplot")
library("stargazer")
library("bucky")
library("lme4")
library("brglm")
library("mice")
library("miceadds")
```

\newpage

# Introduction

On October 10, 2020, then Presidential candidate Joe Biden's campaign team aired an advertisement touting the former Vice President's national security credentials. This particular advertisement is narrated by a man identified as Jim, an "Iraq War Veteran," who, with a personal touch, describes the efforts of the former Vice President in delivering Mine Resistant Ambush Protected (MRAP) vehicles to Iraq and Afghanistan. These vehicles were known for their heavy armor, which likely saved the lives and limbs of servicemen and women who were deployed at the time to the Middle East. The advertisement was later pulled by the Biden Campaign after a retired Army Lieutenant General, Sean MacFarland, objected to the advertisement on the grounds that the ad used images of him without his consent [@moore_biden_2020]. The images in question showed then Lieutenant General MacFarland, in uniform, shaking hands with Biden as the former Vice President arrived to Iraq [@moore_biden_2020]. 

In objecting to the ad, MacFarland noted that he had not granted permission to the Biden Campaign to use images which featured the general in such a recognizable manner. Further, MacFarland expressed concern that voters would assume that his appearance in the ad constituted an endorsement of Biden, and noted that, "A number of people have understandably inferred that my appearance [in this Biden Campaign advertisement] constitutes an endorsement of the former Vice President. It does not. To be clear, I have not endorsed President Trump, either. I object to the use of ANY military personnel in uniform in political ads - full stop" [MacFarland quoted in @moore_biden_2020]. 

This incident illustrates one of the many ways in which political candidates use the military, especially during elections.^[To be clear, President Trump's campaign likewise used military symbols and imagery in many advertisements. For a salient example in which the Chairman of the Joint Chiefs of Staff appeared in one of Trump's advertisements, see @seligman_top_2020.] While civilians have long appealed to the public by harnessing the popularity of the military or of individual military figures in a variety of ways, the practice of doing so during a closely contested election season underscores how a non-partisan institution can find itself - willingly, or in the case of MacFarland, unwillingly - thrust into the political spotlight.   

This chapter examines variation in the use of the military during Presidential campaign television advertisements. The content of campaign advertisements is primarily the purview of civilian rather than military actors, and thus, examining this particular form of behavior in depth will facilitate a comparison between political behaviors undertaken by military actors (the focus of last chapter) vice civilian actors. Of interest is whether any discernible variation can be detected in the degree to which civilian actors have used the military in ways that violate the principles of civil-military relations, and to account for this variation. 

This chapter proceeds in four parts. First, this chapter briefly revisits the theory argued previously in this dissertation, which posits that the levels of political polarization and the prestige of the military impact the degree to which military and civilian actors adhere to the central principles of civil-military relations. In the second part of the chapter, I present the methodology and the data used in this chapter, which include roughly 3,600 presidential campaign television advertisements aired on behalf of the five elections occurring inclusively from 2000 - 2016. With the exception of some of the data that is missing, these advertisements comprise all presidential campaign television advertisements used in these elections.^[So far as I know, this dissertation chapter comprises the only scholarly effort undertaken with this particular data for the purposes of coding the use of military symbols, actors, and behaviors.]

The third part of the chapter explores and analyzes the data in detail. The analysis first explores variation in the types of military imagery that have been featured in campaign advertisements, and then explores variation in the ways that various military actors have committed expressly partisan acts during these advertisements, such as endorsing or attacking a political candidate. The results show that there is significant variation in how civilians have used the military, to include its symbols and actors, in presidential campaign television advertisements over the time period examined. Furthermore, the results provide moderately strong support for the notion that the level of military prestige drives the choice of civilians to use the military in ways that challenge one or more of the central principles of civil-military relations, and especially the principle of non-partisanship. In the fourth and final part of the chapter, I conclude by integrating the results of this chapter with the results obtained in the previous chapter. I also lay the groundwork and motivate the qualitative case studies which begin in the subsequent chapter. 

# A Brief Review of the Theory

In this section of the chapter, I briefly review the theory presented earlier in chapter two of this dissertation. I will not review in detail the three central principles of civil-military relations, as these have been covered in depth in the previous two chapters. Instead, the next several paragraphs will focus on a review of the theoretical expectation that the level of military prestige _may_ impact the willingness of _both_ civilian and military leaders to behave in ways that fail to adhere to the central principles of civil-military relations. 

The level of military prestige, I argued in chapter two, is one of two independent variables that shape the constraining influence of the three central principles of civil-military relations, which include the principle of civilian control of the military, the principle of non-partisanship of the military institution, and the principle of non-interference of the military into certain realms of state politics. I argued that high levels of military prestige present military and civilian actors with _opportunities_ to involve the military in politics, and thus, when military prestige is high, civilian and military actors are more likely to engage in behaviors that violate these central principles. 

The other independent variable that impacts the political behavior of civilian or military actors, I argued in chapter two, is the level of political polarization. I argued in chapter two that rising political polarization provides a "motive" for military actors in particular to intervene in politics, and thus, when polarization is high relative to when it is low, military actors are increasingly likely to engage in behaviors that violate the central principles of civil-military relations. 

## The Impact of Military Prestige on Military and Civilian Leaders

Chapter two argued that the level of military prestige is important because higher levels of military prestige increase the _potential_ that the military be used, either by civilian or military actors, to impact public opinion. This central claim arose mainly from the political communication literature about political elites, which argues that the public often looks to elites who are considered or believed by the public to be "knowledgeable" and "trustworthy" [@lupia_democratic_1998, 69-76] when forming decisions or opinions [@zaller_nature_1992]. 

Chapter two noted that although this and other literature on elite communication [@saunders_no_2017; @guisinger_mapping_2017; @baum_relationships_2008] is not specifically written from the perspective of military leaders in particular, applying this literature to military leaders and military elites in particular seems plausible, especially when the variable of military prestige is conceptualized as a relational indicator of the overall level of trust and confidence placed by the public in the military. Chapter two argued for a broad definition of the variable military prestige, and noted that several different pathways might ultimately impact the level of prestige of a military, including defeat or victory in war, and an overall assessment of how central or important the military is thought to be by the nation. The point of doing so was to allow room for the level of military prestige to change as a result of multiple possible factors. Regardless of how it changes, chapter two argued that when the public trusts and respects the military and its leaders at relatively high levels, it is reasonable to conclude that the public simultaneously perceives the military and its leaders as possessing the traits of knowledge and trustworthiness, and thus, as the prestige of the military rises, it is increasingly likely that the public _may_ also listen or become persuaded by the military and its leaders. 

Chapter two also emphasized that the _potential_ to impact public opinion is different from the actual _ability_ or _efficacy_ of the military to impact public opinion, as several experimental studies have shown [@golby_military_2012; @golby_listening_2013]. For the purposes of this dissertation, it is the belief of civilian and military actors in the _potential_ of the military to impact public opinion when the level of military prestige is relatively high that matters, and not whether a more prestigious military actually can and does persuade the public to a greater degree.   

The theory also explained in chapter two that military and civilian leaders both become aware of the military's increasing potential to move public opinion as the prestige of the military rises. In terms of how this impacts civilian and military actor behavior, an important question then becomes what type of outcome civilian and military actors may seek to influence. Civilian leaders, whose goal is to stay in political power or to achieve their desired policy choices, become increasingly willing to violate one or more of the principles of civil-military relations, and therefore to engage in a political behavior or activity that does so. Military leaders, too, may be more willing to behave in ways that violate one or more of the central principles of civil-military relations as the military institution becomes more prestigious. Chapter two also articulated a number of forms or types of behaviors that civilian and military actors engage in that, depending on the context in which these behaviors occur, violate one or more of the central principles of civil-military relations. These included, for example, the civilian actor behaviors of _electoral swaggering_, _partisan stacking_ and _the military as sales reps_, and the military actor behaviors of _explicit insubordination_, _running for office_, and what Risa Brooks has called "politicking" and "shoulder tapping" [@brooks_militaries_2009, 219]. The theory strongly advocated for evaluating each of these and many other types of political behaviors engaged in by civilian and military actors through the lens of their impact to the three central principles of civil-military relations. 

In summary, chapter two argued that rising military prestige presents an _opportunity_ for civilian and military actors to use the military to sway the public. Depending on what policy outcomes or issues civilian and military actors seek to influence - whether it is an election, or a use of force decision - this can result in civilian and military leaders engaging in political behaviors that challenge the principles of civilian control, non-partisanship, and non-interference of the military institution. When civilian and military leaders believe that the military is highly prestigious and therefore capable of swaying the public, both actor types, on the whole, are less likely to adhere to and thus, be constrained by, the central principles of civil-military relations. Combined with the presence (or absence) of a high level of military prestige, which affords military actors with a "motive" to engage in political behaviors that violate the principles of civil-military relations, chapter two argued that in total, four possible domestic structures emerge, each of which corresponds to distinct patterns of political activity that involve the military. 

# Methodology

In order to examine potential variation in the political behaviors undertaken by civilian actors that involve the military, we must examine a behavior that is primarily the purview of civilian actors. Such an approach parallels the approach implemented in the previous chapter, which explored retired military officer opinion commentary, a political behavior that, I argued, is undertaken primarily by military actors. This chapter examines the civilian use of the military in campaign advertisements, a form of behavior that most approximates the civilian actor behavior of _Electoral Swaggering_ that was described previously in chapter two. 

In this chapter, I examine all of the presidential campaign advertisements that aired on television in the five elections that occurred between 2000 and 2016, inclusively, which amount to roughly 3,600 campaign ads in total. During this period, the level of US military prestige, as measured by Gallup polling that tracks the percentage of the public that has a "great deal or quite a lot" of trust in the military, ranges from 64 (in 2000) to 82 (in 2003 and 2009) [@noauthor_confidence_2020].^[In the years that presidential elections occur, the lowest value is 64 (2000) and the highest is 75 (2004 and 2012), however.] 

To be clear, this does not by itself constitute quite the degree of variation that a researcher hopes for, but there are other things that can and should be leveraged. First, from a qualitative perspective, this period captures (albeit barely) the significant change in the international environment marked by the attacks on the United States that occurred on September 11, 2001. These attacks were the proximate cause of two wars that the US military fought beginning that same year. Notwithstanding any debates about its effectiveness during these conflicts, the continual deployment of the US military to Iraq and Afghanistan likely increased the prestige of the military in a sense that may not be entirely captured by the Gallup polling data.

Additionally, when we also factor in the credibility ratings of other predominant political institutions, a greater shift in what we might call "relative" military prestige becomes slightly more evident. For example, the gap between the percentage of Americans surveyed who expressed a "great deal or quite a lot" in the military compared to Congress was 40% (in favor of the military) in 2000 and 64% (also in favor of the military) in 2016. The percentage of surveyed Americans who placed a "great deal or quite a lot" of trust in the military, Congress, and the gap between these two measures (the "relative" level of military prestige) is displayed in Figure \@ref(fig:trust-ch4). 

```{r trust-ch4, warning=FALSE, echo=FALSE, include=TRUE, fig.height=4, fig.width=6, fig.cap = "Public Trust in the US Military, Congress, 2000 - 2021"}

df_trust_ch4 <- read_csv(here("data","inst-trust-ch4.csv")) %>% print()

df_trust_consolidated_ch4 <- df_trust_ch4 %>% 
  mutate(Rel_Mil_Pres=Military-Congress) %>% pivot_longer(Congress:Rel_Mil_Pres, names_to = "Value", values_to = "Score") %>% distinct () %>% print()

df_trust_consolidated_ch4 %>% ggplot(aes(x=Year, y=Score, color=Value)) + geom_smooth(se=FALSE) + labs(x="Year", y="Percentage Surveyed Expressing\n A Great Deal or Quite a Lot of\n Confidence In...", subtitle="Data Available from 'Confidence in Institutions,'\n https://news.gallup.com/poll/1597/Confidence-Institutions.aspx") + scale_color_discrete(name="Level of Trust",
                         breaks=c("Congress", "Military",  "Rel_Mil_Pres"),
                         labels=c("Congress", "The Military","Difference Between Congress and the Military")) + theme(legend.position="bottom") + 
  geom_vline(xintercept=2001, size=.2, colour="tomato") + 
  geom_vline(xintercept=2009, size=.2, colour="dodgerblue") +
  geom_vline(xintercept=2017, size=.2, colour="tomato") +
  geom_text(aes(x=2000, label="Bush 43",y=17), colour="tomato", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=2008, label="Obama",y=30), colour="dodgerblue", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=2016, label="Trump",y=30), colour="tomato", angle=90, vjust=1, text=element_text(size=9))   
```

At the very least, the period from 2000 to 2016 generally saw relative increases in the prestige of the military. These increases are small when we consider trust in the military in its raw form, but larger when viewed in relative terms compared to Congress. Both measurements may be useful when exploring the impact of military prestige on the civilian propensity to use the military in presidential campaign advertisements in ways that violate the central principles of civil-military relations. 

Polarization over the same period also generally increased, except for measurements of affective polarization that rely on measurements provided by American National Election Survey, which show a dip in affective polarization between 2016 [@jeffrey_b_lewis_voteview_2020; @noauthor_american_nodate; @iyengar_affect_2012]. In short, the time period examined in this chapter, 2000 - 2016, broadly reflects movement between a domestic environment characterized by moderate polarization and moderate military prestige to one marked by higher polarization and higher military prestige. Because both of the independent variables of polarization and military prestige are changing over this time period, the statistical analysis will enable us to better disentangle the relative impacts of these (and other) variables. 

## Civilian Use of Military Imagery and Actors in Campaign Ads

Examining how civilians have featured the military in the content of campaign advertisements provides several advantages from a research perspective that are worth briefly articulating. The first is that, similar to the behavior of retired military officer opinion commentary, there is no _a priori_ expectation that featuring a military image or a military actor in a presidential campaign advertisement must violate, or automatically violates, one or more principles of civil-military relations. In fact, there is no expectation that civilian candidates choose to feature the military in campaign advertisements over and above any other institution, or over any other topic in general. Civilian political candidates and their campaign teams have an almost limitless array of options when producing campaign ads. 

A second significant advantage afforded by examining campaign advertisements is that they comprise an integral part of the democratic process. Like other democracies, the United States holds elections at fixed intervals, and thus, we can expect campaign advertisements to regularly occur in periods leading up to an election. We can and should also expect that the content of campaign ads will likely change over time, and for a variety of factors. The issues that are important to voters are not static, and the broad political atmosphere changes over time. Of course, we would expect civilian political candidates and their campaign teams to construct advertisements around those issues and topics they think will be most effective in persuading the public so as to secure electoral victory. At times, perhaps when wars are fought, for example, we might expect advertisements to feature the military in some noticeable way. While we cannot predict what specific issues will be on the docket four or eight years from now, we can suspect, if the theory offered earlier in this dissertation is accurate, that changes in the relative level of military prestige _could_ impact the degree to which civilian candidates feature or showcase the military in the content of campaign advertisements. 

How might civilian actors fail to adhere to the central principles of civil-military relations through the content of campaign advertisements? It is unlikely that the content of a campaign advertisement would violate the principle of civilian control. This is because, as has been stated earlier in this dissertation, civilians have a responsibility to maintain and enforce the principle of civilian control, but usually do not expressly violate the principle through acts they commit. Rather, civilian behavior "erodes" the principle of civilian control when civilians excessively defer to the military such that civilians no longer adequately influence the policy making process [@beliakova_erosion_2021]. Thus, a campaign advertisement that uses the military in such a way that violates the principle of civilian control would take on a rather extreme form, perhaps if an advertisement featured a retired military actor urging others to mount a coup or stage some other act of obvious insubordination. 

It is far more likely that featuring the military in television campaign advertisements would break the other two principles of civilian control - the principle of non-partisanship and non-interference of the military. A violation of the principle of non-partisanship of the military vis-a-vis campaign advertisements would occur if an advertisement clearly intimated that the military favored or endorsed a particular candidate, or the policies or political stances of a candidate, or a particular political party over another. Such a message, deliberate or not, could and likely would communicate to the American public that the military (or a faction of the military, or particular leaders within the military) favors a particular political entity over another. Examples of such behavior abound in almost every election, as this chapter will show. 

A violation of the principle of "non-interference" of the military into certain realms of state policy might occur if a campaign advertisement featured a military actor weighing in on policy arenas that are largely disconnected from national security, defense, and other essential roles the military is expected to fill in American society [@ben-meir_civil-military_1995]. Hypothetically speaking, this could occur if a retired general appeared in a campaign advertisement and spoke predominantly about a candidate's plan to reform social security or to reduce the national debt, and why particular candidate's plans to tackle these topics are more sound than another candidate's plans on the same topics. In this example, it would be reasonable to conclude that the advertisement violated the principle of non-interference because the retired military officer is likely not an expert on the topics of social security reform and the national debt. I do not question that a large number of reasonably intelligent retired military officers, like many retired individuals at large within American society, are smart and can understand the topics of social security reform and reigning in national debt. At issue, though, in this hypothetical example, is the choice by the civilian candidate or campaign team to use a military figure to convince the American public of the candidate's merits on these particular topics. It is doubtful that a retired general, admiral, colonel, or captain are truly the most qualified experts to weigh the candidates' stances over these issues that are somewhat removed from topics that distinctly involve national security, and thus, such an act violates the principle of non-interference. 

## Hypotheses Explored in this Chapter

The five hypotheses presented in chapter two are listed below. Given that this chapter focuses on a type of political undertaken primarily by civilian actors, this chapter primarily addresses H3. It will not address H1, which was addressed in the previous chapter, nor H2, as both of these hypotheses are addressed towards military rather than civilian actors. This chapter will, however, inform H4 and H5.  

H1: Military actors increasingly violate the central principles of civil-military relations when polarization is high relative to periods of low polarization. 

H2: Military actors increasingly violate the central principles of civil-military relations in public and visible ways when military prestige is high relative to periods of low military prestige. 

H3: Civilian actors increasingly violate the central principles of civil-military relations when military prestige is high relative to periods of low military prestige. 

H4: Military and civilian actors engage in relatively more visible and public forms of political behavior that involve the military institution when military prestige is high relative to periods of low military prestige. 

H5: The most extreme levels of overall political activity occur in environments that are characterized by simultaneously high levels of political polarization and military prestige. 

## The Data

The data consist of all (more than 3,600) presidential campaign television advertisements from the 2000 - 2016 presidential elections. ^[Unfortunately, the raw data used to code the 1996 election is no longer available to review; hence, this study begins with the 2000 Presidential Election. Additionally, raw data for the 2020 Election is not available at the time of this writing, but is projected to become available sometime in late 2022 or 2023.] The data is made available from the Wisconsin Advertising Project (WiscAD) and Wesleyan Media Project (WMP). For the 2000, 2004, and 2008 elections, the data come from the WiscAd, and consist of individual PDF documents that contain a full transcript and several still images of the advertisement [@goldstein_political_2002; @goldstein_presidential_2007; @goldstein_presidential_2011]. For 2012 and 2016 elections, the data comes from the WMP, and consists of full motion video, to include sound [@fowler_presidential_2017; @fowler_presidential_2020].^[See the Appendix for further details regarding how the data was cleaned and merged to enable a comparison over time.]

In the statistical analysis portion of this chapter, I also rely on data covering polarization, military prestige, and military casualties. For polarization data, I primarily rely on DW-Nominate (dynamic-weight, three step estimation) scores, which are constructed using roll-call voting data from the US House of Representatives and the US Senate [@lewis_voteview_nodate]. I also include "affective" polarization measurements, which are generally constructed using a series of "feeling thermometer" scores from the American National Election Study (ANES) [@iyengar_affect_2012; @iyengar_strengthening_2018; @iyengar_origins_2019].  

As a proxy for military prestige, I rely on Gallup public opinion data that captures the level of trust that Americans have placed in the military and a variety of other institutions over the past several decades [@noauthor_confidence_2020]. Finally, I use a variety of open source data sources from the Department of Defense regarding US military casualties to measure levels of hostile casualties sustained by US military forces over the past several decades.^[These sources include @department_of_defense_dcas_2011; @department_of_defense_dcas_2021; @department_of_defense_dcas_2021-1; @department_of_defense_dcas_2021-2; @department_of_defense_dcas_2021-3; @department_of_defense_dcas_2021-4; @department_of_defense_dcas_2021-5; @department_of_defense_dmdc_2021.]  

After gathering the data, I implement an empirical strategy that at its root, looks to answer two major questions. The first question, does this particular advertisement have any type of military image or symbol featured in the advertisement? At a minimum, answering this first question will help paint a picture regarding how frequently military imagery generally appears in campaign advertisements. The second question I seek to answer is whether a particular advertisement features the military in some way that violates the principles of civil-military relations. Answering these two main questions and presenting the results in a meaningful way is the aim of th next section of this chapter. 

# Analysis

```{r data, include=FALSE, echo=FALSE}

# reading in the raw data
df_campaign_ads_read <- read_csv(here("data", "refined_master_imagery_ch4_my_edits.csv"))
summary(df_campaign_ads_read)

# truncate the data after the last row (observation 3666) 
# rename the variable 'missing' to 'missing_ad', and replace all N/A values of 'missing_ad' with #0

df_campaign_ads_raw <- df_campaign_ads_read [1:3666,] %>% select(year:missing) %>% rename(missing_ad=missing) %>% mutate_at(
  .vars = vars(missing_ad), 
  .funs = ~ case_when(is.na(.) ~ 0, TRUE ~1) 
) %>% view()

# create another, identical data frame for conducting MICE later
df_campaign_ads_raw_mice <- df_campaign_ads_raw


```

In this portion of the chapter, I analyze the data and present the results. First, I present and analyze the general frequency and type of military symbols and imagery that appear within and across the data. This first step is meant to serve as a more general overview of the ways in which the military appears - in some fashion - in campaign advertisements over time.  Second, I analyze the frequency with which particular campaign advertisements involve the military in ways that violate one or more of the principles of civil-military relations. This second step is the more substantive of the two in terms of examining which factors impact the degree to which civilian actors violate the principles of civil-military relations.     

In total, the data consist of 3,666 advertisements across the five presidential elections that occurred from 2000 - 2016, inclusively. 89 of these observations (2.4%) are missing.^[After contacting WiscAd and WMP, there is no reason to suspect that the data is not missing at random. Later in the statistical analysis, I conduct multiple imputation to deal with the missing data, but want to first provide the "raw" or "as is" facts given the data that is available.] After cleaning and merging the data across the several data sets, I capture several variables that WiscAd and WMP have already coded for that may be helpful for the statistical analysis conducted later in this chapter. These variables, and a short description of each, are displayed in Table \@ref(tab:existing-variables). 

```{r existing-variables, include=TRUE, echo=FALSE}

text_vars <- tribble(~"Variable Name", ~"Question", ~"Possible Values",
  "Creative",  "What is the title of the ad?", "varies by each ad",
  "Count", "How many times did this ad air on a particular media market?", "varies by each ad", 
  "Length", "How long is the ad (in seconds)", "0-120",
  "Party", "What is the Party of the favored candidate?", "Republican; Democrat; Other",
  "Sponsor", "Who sponsored the advertisement?", "1 - Candidate; 2 - Party; 3 - Coordinated Between a Candidate and Party; 4 - Interest Group/Other",
  "f_mention", "Is the favored candidate mentioned in the ad?", "1 - Yes; 0 - No",
  "f_picture", "Is the favored candidate pictured in the ad?", "1 - Yes; 0 - No",
  "f_narrate", "Does the favored candidate narrate the ad?", "1 - Yes; 0- No",
  "o_mention", "Is/are the opposition candidate/s mentioned in the ad?", "1 - Yes; 0 - No",
  "o_picture", "Is/are the opposition candidate/s pictured in the ad?", "1 - Yes; 0 - No",
  "ad_tone", "In your judgement, is the primary purpose of the ad to promote a specific candidate, attack a candidate, contrast the candidate, or other?", "1 - contrast; 2- promote; 3 - attack; 4 - other",
  "prty_mn", "Does the ad mention the party label of the favored candidate or the opponent?", "0 - No, 1 - Yes, favored candidate's party; 2 - Yes, opposition candidate's party; 3 - Yes, both candidate's parties; 4 - yes, favored candidate's party but only in the paid for by line"

  )


text_vars %>% kable(booktabs=T, caption="Existing Variables from WiscAd and WMP") %>%
  kable_styling(full_width = F, font_size=10.5, latex_options="striped" ) %>% 
  column_spec(1, width="8em") %>% 
  column_spec(2, width="12em") %>% 
  column_spec(3, width="15em") 


```

I then develop four dichotomous variables that capture various types of military images and symbols.^[I came up with these four variables myself by first viewing approximately 250 advertisements across the five data sets to determine what categories would be most meaningful in terms of distinguishing the different types of images and symbols.] These four variables denote the use of active duty US soldiers or troops, veterans, combat, or US military hardware. 

The first kind of military imagery, that of US soldiers or troops, include images of US servicemen and women who are in uniform. It could include a lone soldier saluting, or thousands of US servicemen and women training, or a host of other possible scenarios. The key thing I look for are images of US military uniforms, military medals, or insignia that clearly denote US soldiers. The second type of military imagery relates to veterans, which I consider distinct from images of troops. This indicator variable looks for any image of a veteran, which could include gatherings of veterans, such as an American Legion or Veterans of Foreign Wars group, or imagery of a wounded veteran recovering in a hospital, or a veteran, for example, who is in a wheelchair during a later stage of life.

The third indicator accounts for images of combat. This could include things such as an explosion, a soldier driving a tank, or a scene containing fighting that clearly depicts US soldiers involved. This category does not include images of fighting that clearly involve non-US soldiers, such as soldiers of a foreign military or members of a terrorist group. The fourth indicator variable for a type of military imagery accounts for images of US military hardware. This variable captures advertisements that contain images or scenes of any major military hardware item, such as a naval vessel, fighter jet, or military radar system. 

I allow for the possibility that an advertisement portrays or includes more than one type of imagery. Again, the purpose of each of these four variables is to obtain a general idea for the prevalence of military imagery across the data. These four variables do not necessarily indicate that a campaign advertisement has violated any of the central principles of civil-military relations. 

Finally, I develop four additional variables that measure several other items of interest. These include the overall purpose of the advertisement, whether the candidate is a veteran, whether the advertisement explicitly mentions or portrays that the featured candidate is a veteran, and whether the advertisement itself is entirely focused on military, defense, veteran, and/or or national security issues. These variables, the question they intend to answer, and the potential range of values these variables can take, are displayed in Table \@ref(tab:el-ads-additional).

```{r el-ads-additional, include=TRUE, echo=FALSE}

text_vars <- tribble(~"Variable Name", ~"Question", ~"Possible Values",
  "Act_Duty_Imagery", "Does the advertisement feature imagery of US troops in uniform?", "1 - Yes; 0 - No",
  "Vet_Imagery", "Does the advertisement feature imagery of Veterans (VFW, veteran gatherings, etc.)?", "1 - Yes; 0 - No",
  "
  US_Mil_Hardware_Imagery", "Does the advertisement feature imagery of US military hardware (planes, tanks, helicopters, ships)?", "1 - Yes; 0 - No",
  "Combat_Imagery", "Does the advertisement feature images of past or present combat involving US military forces (prior wars, scenes of combat)?", "1 - Yes, 0 - No",
  "Def_Vet_Ad_Theme", "Is the ad itself focused entirely on military, defense, veteran, or national security issues?", "1 - Yes, military/defense/veteran issues only; 0 - other topics",
    "Ad_Purpose",  "What is the overall purpose of the ad", "1-promote or disparage candidates for an election; 2 - promote/disparage a specific issue or proposition that is on the ballot, 3 - to discuss a general theme (the economy, immigration, etc.) that does not do 1 or 2;  4 - other",
  "Can_Vet", "Is the candidate a Veteran?", "1 - Yes; 0 - No",
  "Can_Vet_Mention", "Does the ad explicitly state or portray the candidate as a Veteran?", "1 - Yes, 0 - No",
  "Vet_Fam", "Does the ad feature a Retired Military Elite, Veteran, or Veteran's family speaking?", "3 - Veteran and Family; 2 - Veteran's Family; 1 - Veteran; 0 - No",
  "Vet_Fam_Endorse", "Does the Military Figure who appears in a prominent way explicitly endorse the candidate for office?", "1 - Yes; 0 - No; 99-N/A",
  "Vet_Fam_Attack", "Does the Military Figure who appears in a prominent way explicitly attack or disparage other candidates running for office?", "1 - Yes; 0 - No; 99 - N/A",
  "Vet_Fam_Out_Issues", "Does the Military Figure who appears in a prominent way address topics other than the candidate's personal characteristics or defense/veteran (broadly related) issues?", "1 - Yes, the vet or vet's family addresses topics outside of the candidate's characteristics and/or veterans/defense (broadly related) issues; 0 - No, the vet or vet's family only addresses the candidate's characteristics and/or veterans/defense issues; 99 - N/A"
  

  )
 
text_vars %>% kable(booktabs=T, caption="Additional Variables / Anticipated Coding Scheme") %>%
  kable_styling(full_width = F, font_size=10.5, latex_options="striped" ) %>% 
  column_spec(1, width="8em") %>% 
  column_spec(2, width="14em") %>% 
  column_spec(3, width="16em") %>% add_footnote(label = "Vet_Fam_Endorse, Vet_Fam_Attack, and Vet_Fam_Out_Issues can take a value of N/A if the variable Vet_Fam equals 0.")
```

## Summary Statistics and Descriptive Data

Table \@ref(tab:desc-stats) shows, for each election cycle, the total number of advertisements aired on the nation's top media markets, the number of advertisements missing from the data, the proportion of the data that is missing, the number of advertisements that featured some form of military imagery in each election, and the proportion of each election's ads that contained military imagery. 

```{r initial-stats, include=F, echo=F, warning=F, message=F}
desc_stats <- tribble(
  ~'Election Cycle', ~'Total Ads (on hand and missing)', ~'# Ads Missing', ~'% Ads Missing', ~'# Ads on hand with Military Imagery', ~'% of Ads on hand with Military Imagery',
  "2000", "347", "39", "11.2", "20", "6.5",
  "2004", "650", "3", ".05", "136", "20.9",
  "2008", "966", "45", "4.7", "138", "15",
  "2012", "789", "1", ".1", "85", "10.8",
  "2016", "914", "1", ".1", "171", "18.7",
  "2000-2016", "3666", "89", "2.4", "550", "15.0")
```

```{r desc-stats, include=TRUE, echo=FALSE}
 
kable(desc_stats, booktabs=T, caption="Summary Statistics, The Military in Presidential Campaign Advertisements, 2000-2016") %>%
  kable_styling(full_width = F, font_size=10.5, latex_options="striped") %>% 
  column_spec(1:2, width="5em") %>% 
  column_spec(3:4, width="6.5em") %>% 
  column_spec(5:6, width="6.5em") 


```
### The Prominence of Military Imagery in US Presidential Elections

Table \@ref(tab:desc-stats) shows that 15 percent of the presidential campaign television ads featured military imagery of some kind. The least and most saturated elections in terms of military imagery, respectively, were the 2000 and 2004 elections, when 6.5% and 20.9% of advertisements featured some kind of military imagery.^[The 2000 Election contains the highest incidence of missing data, at 11.2%. Still, it is very unlikely that the 2000 Election is _not_ the least saturated with military imagery. For the 2000 Election to eclipse the second-least saturated election (2012), at least 18 of the 39 missing ads (46%) from the 2000 Election would have to contain military imagery. As the data is missing randomly, this is certainly possible, but very highly unlikely.] Following the 2004 Election, the percentage of campaign ads containing military imagery declined in the 2008 and 2012 Elections before rising sharply again in 2016. 

There are a number of possible explanations behind these broad trends. One involves the arc of American foreign policy, and the military's role in shaping it between 2000 and 2016. The September 11, 2001 attacks ushered in a distinct role for the military as the US conducted lengthy military operations, first in Afghanistan in 2001 and later in Iraq in 2003. That military imagery contained high relative proportions of military imagery during the 2004 and 2008 Elections is mostly unsurprising. The war in Iraq was a major theme in the 2004 Election, when the incumbent, George W. Bush, ran against the Democratic challenger, John Kerry, a decorated combat veteran of the Vietnam War. At the time of the 2008 election, the surge of US military forces in Iraq had shown signs that it was likely working to at least alleviate the level of sectarian violence in Iraq. Moreover, US casualties, after peaking in 2007, had also been declining steadily throughout 2008 [@us_department_of_defense_dcas_2021]. This is likely a major reason why the proportion of campaign ads featuring military imagery declined in 2008 relative to the proportion of ads featuring military imagery in the 2004 election. 

The arc of US foreign policy, and the US military's role in it, becomes a bit more complicated from 2008 to 2016. First, from 2008 to 2012, the US' involvement in Iraq ended, but its involvement in Afghanistan increased. Indeed, the last of US forces withdrew from Iraq at the end of 2011 when Operation _New Dawn_ concluded.^[US forces would return to Iraq as part of the counter-ISIS effort several years later, and they remain there, albeit in relatively small numbers, as of writing this dissertation in early 2022.] But, President Barack Obama has also in 2009 decided to increase the number of US forces in Afghanistan as part of a surge of US forces in order to conduct a rigorous counter-insurgency campaign. Though the Afghan surge did not bring the acclaim that the earlier surge to Iraq did, there were nonetheless important victories along the way, perhaps none as important as the killing of Osama bin Laden, the architect of the 9/11 attacks, in 2011. Additionally, the number of US casualties in Afghanistan peaked in the summer of 2011, and declined afterwards, especially in the months leading up to the 2012 election [@us_department_of_defense_dcas_2021-1]. At the time of the election in November of 2012, few observers could point to Afghanistan and say that the surge of forces there had not worked, but at the same time, there was little if any appetite - from politicians and the public, alike - to do anything substantially greater there. Combined with the fact that 2012 was a reelection year for the incumbent, Barack Obama, the relatively low percentage of campaign advertisements featuring the military in 2012 is, in my view, rather unsurprising. 

However, in the 2016 election, the percentage of advertisements featuring military imagery rose sharply compared to the 2012 election. The passage of time facilitated an opportunity for Americans and political candidates to think about the trajectory of US foreign and national security policy in broad and critical terms, and in particular, to evaluate US involvement in a series of messy situations and crises that had unfolded abroad, particularly in the Middle East and Eastern Europe. These included conflict in Libya and an increasingly deteriorating civil war in Syria. In particular, President Obama's handling of events in Syria in 2013, which included making but failing to enforce a now infamous "red line" policy position that alluded to US military action in the event the Assad regime used chemical weapons (which Assad did), troubled many Americans [@rhodes_inside_2018].

In many respects, Hillary Clinton, the Democratic Candidate for President in 2016, was largely seen as an extension of Barack Obama, particularly in the realm of foreign policy. Clinton had served as Obama's Secretary of State, and she was portrayed extremely negatively after a US Ambassador and three other Americans were killed on September 12, 2012, when terrorists attacked the US consulate in Benghazi, Libya [@bixby_parents_2016]. Donald Trump, an outside candidate who ran and ultimately won as a Republican, instead ran on a platform of putting "America First," which included promises to cease fighting endless wars and significantly rebuilding the US military. The combination of perceived foreign policy missteps between 20012 and 2016 and the promise of fortifying the US' strategic position likely led to the relative increase in military imagery in the 2016 presidential election. 

```{r data imputation, include=FALSE, echo=FALSE}

## use multiple imputation to handle missing data

# create a variable for candidate name

df_campaign_ads_raw_mice$candidate <- factor(sub("/.*$", "", df_campaign_ads_raw_mice$creative)) %>% print()

df_campaign_ads_raw_mice <- df_campaign_ads_raw_mice %>% mutate(
  party_reg=case_when(
    party=="DEMOCRAT" ~ "DEMOCRAT",
    party=="REPUBLICAN" ~ "REPUBLICAN",
    (party!= "REPUBLICAN") & (party!= "DEMOCRAT") ~ "OTHER"
  ),
  party_reg_numeric=case_when(
    party_reg=="DEMOCRAT" ~ 1,
    party_reg=="REPUBLICAN" ~ 0,
    party_reg=="OTHER" ~2
  )) %>% print()


df_campaign_ads_raw_mice$prom_mil_fig <- factor(df_campaign_ads_raw_mice$prom_mil_fig, levels=setdiff(unique(df_campaign_ads_raw_mice$prom_mil_fig), c("")))

df_campaign_ads_raw_mice_ready <- (df_campaign_ads_raw_mice[,!sapply(df_campaign_ads_raw_mice, is.character)])

df_campaign_ads_raw_mice_ready <- df_campaign_ads_raw_mice_ready %>% mutate(
   house_polar = case_when(
    year==2000 ~ 77.713,
    year == 2004 ~ 79.302,
    year == 2008 ~ 80.765,
    year == 2012 ~ 86.142,
    year == 2016 ~ 87.644
    ), 
  sen_polar = case_when(
    year==2000 ~ 65.833,
    year == 2004 ~ 64.992,
    year == 2008 ~ 68.744,
    year == 2012 ~ 74.383,
    year == 2016 ~ 81.304
  ),
  aff_polar = case_when(
    year==2000 ~ 61.692,
    year == 2004 ~ 68.052,
    year == 2008 ~ 69.519,
    year == 2012 ~ 84.689,
    year == 2016 ~ 67.754,
  ),
  HostCasRate = case_when(
    year==2000 ~ 1.23874924,
    year == 2004 ~ 52.0801226,
    year == 2008 ~ 25.1029256,
    year == 2012 ~ 29.4917871,
    year == 2016 ~ 1.229532
  ),
  mil_prestige = case_when(
    year== 2000 ~ 64,
    year == 2004 ~ 75,
    year == 2008 ~ 71,
    year == 2012 ~ 75,
    year == 2016 ~ 73
  ), 
  dummy_04 = case_when(
    year == 2004 ~1, 
    year != 2004 ~0
  )
)

# goes through variables except count and l, made them factors, and changed 99 to 0
for (varname in setdiff(names(df_campaign_ads_raw_mice_ready), c("count", "l", "house_polar", "sen_polar", "aff_polar", "HostCasRate", "mil_prestige"))) {
    if (length(setdiff(unique(df_campaign_ads_raw_mice_ready[[varname]]), 99)) > 2)
        df_campaign_ads_raw_mice_ready[[varname]] <- factor(ifelse(df_campaign_ads_raw_mice_ready[[varname]]==99,0,df_campaign_ads_raw_mice_ready[[varname]]), levels=sort(setdiff(unique(df_campaign_ads_raw_mice_ready[[varname]]), 99)))
    else
        df_campaign_ads_raw_mice_ready[[varname]] <- as.integer(ifelse(df_campaign_ads_raw_mice_ready[[varname]]==99,0,df_campaign_ads_raw_mice_ready[[varname]]))
}

## conduct multiple  imputation

#make a prediction matrix
prediction_matrix <- make.predictorMatrix(as.matrix(df_campaign_ads_raw_mice_ready), blocks = make.blocks( c("act_dty_img", "vet_img","us_mil_hard_img", "us_cmbt_img", "vet_fam", "vet_fam_end", "vet_fam_atk", "vet_fam_out_issues"), partition="scatter"))

prediction_matrix[, "candidate"] <- -2

# change the variable candidate to a type of integer
df_campaign_ads_raw_mice_ready$candidate <- as.integer(df_campaign_ads_raw_mice_ready$candidate)

# put in variables you don't want to use for the imputation
#prediction_matrix[, c("")] <- 0

df_campaign_ads_raw_mice_imputed <- mice(data=df_campaign_ads_raw_mice_ready, method="2l.bin", predictorMatrix=prediction_matrix, maxit=10, m=20, blocks=as.list(rownames(prediction_matrix)))

```

### Specific Types of Military Imagery

Moving beyond the general incidence of military imagery across presidential elections from 2000 - 2016, Figure \@ref(fig:img-type) displays the incidence of military imagery by type in these elections. The figure shows that imagery of active duty troops and veterans are the most common forms of military imagery. It is also clear that the relative gap between active duty military imagery and veteran imagery has widened, as veteran imagery has emerged as the most common form of imagery beginning in the 2012 election.

```{r data selection results, include=FALSE, echo=FALSE}
#create a variable called Yrcount that counts the number of ads per election cycle
#create a variable called "mil_img" if there is any type of military imagery category in a 
# given ad
### analysis with original code on the data that we have on hand

df_campaign_ads_raw_on_hand <- df_campaign_ads_raw %>%
  filter(missing_ad != 1) %>% 
  group_by(year) %>% 
  mutate(
    Yrcount=n(),
    mil_img = case_when(
    (vet_fam != 0) | (act_dty_img ==1 | vet_img ==1 | us_mil_hard_img ==1 | us_cmbt_img==1) ~ 1, 
  (vet_fam == 0) & (act_dty_img ==0 & vet_img ==0 & us_mil_hard_img ==0 & us_cmbt_img==0) ~ 0
),
  no_mil_img = case_when(
    (vet_fam != 1) & (act_dty_img !=1 & vet_img !=1 & us_mil_hard_img !=1 & us_cmbt_img!=1) ~ 1, 
  (vet_fam != 0) | (act_dty_img ==1 | vet_img ==1 | us_mil_hard_img ==1 | us_cmbt_img==1) ~ 0
  )
    ) 


# create a new df called 'adj_df' that 
# takes the original df and adds additional variables 
#(1) sum_mil_img, the total # of ads per election cycle containing military imagery 
#(mil_img ==1); 
#(2) sum_no_mil_img, the total # of ads per cycle containing no military imagery
#(3) mil_fig, if an ad features a prominent veteran or veteran's family in the ad
# (mil_fig==1)
#(4) sum_mil_fig_cycle, the total # of ads per cycle that feature a veteran or family endorser
# sum_mil_fig_cycle
#(5) sum_def_vet_ad_theme, the total # of ads per cycle that address entirely matters
# of national security, defense, or veterans issues
adj_df <- df_campaign_ads_raw_on_hand %>% group_by(year) %>% mutate(
    sum_mil_img_cycle=sum(mil_img),
    sum_no_mil_img_cycle=sum(no_mil_img),
    prop_mil_img = sum_mil_img_cycle/Yrcount,
    act_duty_img_total=sum(act_dty_img),
    prop_act_dty_img=(act_duty_img_total/sum_mil_img_cycle),
    vet_img_total=sum(vet_img),
    prop_vet_img=(vet_img_total/sum_mil_img_cycle),
    us_cmbt_img_total=sum(us_cmbt_img),
    prop_us_cmbt_img=(us_cmbt_img_total/sum_mil_img_cycle),
    us_mil_hard_img_total=sum(us_mil_hard_img),
    prop_us_mil_hard_img=(us_mil_hard_img_total/sum_mil_img_cycle),
    mil_fig=case_when(vet_fam != 0 ~1,
                      vet_fam == 0 ~ 0),
    sum_mil_fig_cycle=sum(mil_fig),
    sum_def_vet_ad_theme=sum(def_vet_ad_theme),
    def_theme_and_mil_img=case_when((def_vet_ad_theme==1) & (mil_img==1) ~1,
                                    (def_vet_ad_theme!=1) | (mil_img==0) ~0),
    sum_def_theme_and_mil_img=sum(def_theme_and_mil_img),
    not_def_theme_but_mil_img=case_when((def_vet_ad_theme!=1) & (mil_img==1) ~ 1,
                                         (mil_img==0) | (def_theme_and_mil_img==1) ~0),
    sum_not_def_theme_but_mil_img=sum(not_def_theme_but_mil_img),
    non_def_vet_ad_theme=case_when((def_vet_ad_theme==0) ~1,
                                  (def_vet_ad_theme==1) ~ 0),
    sum_non_def_vet_ad_theme=sum(non_def_vet_ad_theme)
    ) %>% view()

```


In 2004 and in 2008, combat imagery is more common than imagery containing military hardware, but in all other elections, more images of military hardware were evident than combat imagery. The most reasonable explanation for this difference is the fact that American combat operations in Iraq and Afghanistan loomed large over the 2004 and 2008 elections, respectively. It is also quite clear from Figure \@ref(fig:img-type) that the 2000 election contained far less military imagery - across all imagery types - than subsequent elections. This likely reflects the post-9/11 era increase in the general prominence or the centrality of the military as an institution in politics relative to the immediate elections held in the pre-9/11 era. 

```{r img-type, include=TRUE, echo=FALSE, error=FALSE, warning=FALSE, fig.height=4, fig.width=6, fig.cap="Number of Presidental Campaign TV Ads Displaying Military Imagery, By Type of Military Imagery"}

## prepare a graph for raw counts and proportions of image type 

# raw counts
graphable_img_types_df <- adj_df %>% 
  select(year, act_duty_img_total, vet_img_total, us_cmbt_img_total, us_mil_hard_img_total) %>% pivot_longer(act_duty_img_total:us_mil_hard_img_total, names_to = "statistic", values_to = "Number") %>% distinct () %>% view ()

plot_img_type_count <- ggplot(data=graphable_img_types_df, aes(fill=statistic, x=year, y=Number)) + geom_bar(position="dodge", stat="identity") + labs(x="Election", y="Number of Instances") + scale_fill_discrete(name="Military Imagery Type",
                         breaks=c("act_duty_img_total", "vet_img_total", "us_cmbt_img_total", "us_mil_hard_img_total"),
                         labels=c("Active-Duty Military", "Veteran", "Combat", "Mil. Hardware")) + theme(legend.position="bottom") + scale_x_continuous(breaks=seq(2000, 2016, 4))


# proportion
graphable_img_prop_df <- adj_df %>% 
  select(year, prop_act_dty_img, prop_vet_img, prop_us_cmbt_img, prop_us_mil_hard_img) %>% pivot_longer(prop_act_dty_img:prop_us_mil_hard_img, names_to = "statistic", values_to = "Proportion") %>% distinct () %>% view ()

plot_img_type_prop <- ggplot(data=graphable_img_prop_df, aes(fill=statistic, x=year, y=Proportion)) + geom_bar(position="dodge", stat="identity") + labs(x="Election", y="Proportion of Type of Military Imagery") + scale_fill_discrete(name="Military Imagery Type",
                         breaks=c("prop_act_dty_img", "prop_vet_img", "prop_us_cmbt_img", "prop_us_mil_hard_img"),
                         labels=c("Active-Duty Military", "Veteran", "Combat", "Mil. Hardware")) + theme(legend.position="bottom") + scale_x_continuous(breaks=seq(2000, 2016, 4))

plot_img_type_count

plot_img_type_prop


```


### The Use of Military Imagery in non Defense-Themed Ads

To a large extent, we can expect that some form of military imagery would appear in advertisements that are squarely focused on issues related to defense, and that the prominence of defense issues might change, as described above, over time as particular events unfold. Yet there is another way to gauge how civilian candidates choose to leverage the military for their electoral advantages. This involves looking for instances of advertisements that are _not_ centered on defense-related issues, but which feature military imagery anyways. In such a case, the focus or theme of a particular advertisement is on an issue or a broader set of issues than those that pertain squarely to defense or national security, but by still including some type of military imagery, the candidate is, I argue, nonetheless signaling to the public that the military matters to a certain degree.

Figure \@ref(fig:non-def-theme-imagery) shows the proportion of advertisements in each election cycle that include military imagery but are _not_ squarely focused on issues pertaining to defense or veterans themes. The smoothing line in blue clearly depicts an upward trend, noting that campaign advertisements are increasingly featuring military imagery, even in those advertisements that are not directly addressing an issue related to defense or veterans issues. The data is consistent with the notion that communicating the importance of the military to the public by means of including some form of military imagery, even in ads that are not directly related to issues of defense and national security, weighs heavily in the logic of civilian political candidates and their campaign teams. 

```{r non-def-theme-imagery, include=TRUE, echo=FALSE, warning=FALSE, fig.height=4, fig.width=6, fig.cap="Military Imagery in non-Defense or non-Veteran Related Presidential Campaign Advertisements, 2000-1016"}

facet_wrap_nondef <- adj_df %>% select(year, Yrcount, sum_def_vet_ad_theme, sum_def_theme_and_mil_img, sum_not_def_theme_but_mil_img) %>% mutate(
  prop_not_def_but_mil_img = sum_not_def_theme_but_mil_img/Yrcount
  ) %>% select(year, prop_not_def_but_mil_img) %>% 
  distinct()

# now make the data into longer format for graphing and prepare to graph

facet_wrap_nondef <- facet_wrap_nondef  %>% pivot_longer(prop_not_def_but_mil_img, names_to = "question", values_to = "value")

plot_nondef_img <- ggplot(data=facet_wrap_nondef, aes(fill=question, x=year, y=value)) + geom_bar(position="dodge", stat="identity") + labs(x="Election", y="Proportion of Advertisements") + scale_fill_discrete(name="Ad Category",
                         breaks=c("prop_not_def_but_mil_img"),
                         labels=c("Ads that are focused on Non-Defense/Veteran Issues\n But Contain Military Imagery")) + theme(legend.position="bottom") + scale_x_continuous(breaks=seq(2000, 2016, by = 4)) + geom_smooth()

plot_nondef_img

```


## Are Campaign Ads Increasingly Challenging Principles of Civil-Military Relations?

Having described the data in the previous section, I now pivot to answering a second, and for the purposes of the theory argued earlier in this dissertation, more substantive question, which is whether and to what degree the military images and symbols that feature in presidential campaign advertisements violate one or more of the principles of civil-military relations. 

In order to answer this important question, I must first determine which advertisements feature a military figure in a prominent way. That is, I find advertisements that include a military figure who appears in an advertisement in a noteworthy manner, or in a way that is qualitatively distinct from, for example, a passing unnamed veteran who delivers a salute on camera. I am rather interested in finding advertisements which include figures such as retired generals or admirals, veterans, or veteran family members who behave in ways that are deliberate and purposeful in terms of delivering the message of the advertisement. These types of behaviors include a military figure narrating an advertisement, or endorsing or attacking a political candidate. 

I therefore develop several additional variables which capture various military actors who appear in advertisements in prominent ways, as well as some of the potential behaviors these actors may commit. The first is a categorical variable that captures instances of various military actors _appearing_ in a prominent way in the campaign advertisement. The values this variable can take include an elite military actor (such as retired generals or admirals), a military veteran (who is not a retired general or admiral), a family member of a veteran, or a veteran _and_ members of his or her family. 

These categorical values are important because each type of military actor is likely designed to portray a slightly different type of message. Elite military actors, such as retired general and admirals, convey the rank and authority of high-ranking military officers. Civilian candidates might choose elite military actors to deliver a message in the hopes that the public will listen to them on the basis of their accrued knowledge and expertise. Non-elite veterans, on the other hand, may not be subject matter experts on specific and often complex foreign policy issues, but they might nonetheless be chosen by civilian political candidates to convey a message regarding the broader value of service to one's country. A non-elite veteran is likely in a better position, in other words, to hypothetically state in an advertisement, "I fought for my country; now I hope the country does its part by making prescription drugs more affordable" than a retired multistar general or admiral. Veteran family members, understandably, offer an even more personal and intimate connection to the public. Civilian political candidates may use veterans and/or their family members to craft emotional, heartfelt messages with the public. 

I then create variables to capture what actions, if any, these military figures who appear in a prominent way perform. I develop a series of indicator variables that correspond to behaviors that violate either the principle of non-partisanship of the military and the principle of non-interference of the military. Violation of the military's non-partisanship principle is captured by indicator variables that catalog instances of a military figure explicitly endorsing a political candidate for office, or attacking a political candidate running for office. Violation of the principle of non-interference of the military, on the other hand, is captured by an indicator variable that captures whether the military figure addresses issues that fall outside of the military figure's military expertise.  

### Prominent Appearances by Military Figures in Campaign Ads

Figure \@ref(fig:mil-endorse) shows the number of times in each election cycle that military figures appear in prominent ways during presidential campaign advertisements (in instances where the the viewer is informed that a certain candidate was "endorsed" by several former members of the Joint Chiefs of Staff, whose photo is displayed during the ad, I have labeled this category as "retired generals/admirals pictured only"). 

Figure \@ref(fig:mil-endorse) portrays an interesting narrative. There were no instances of military figures appearing prominently in campaign advertisements in the 2000 Election. Then, in the 2004 election, there was an explosion of military figure prominent appearances. In 2008 and 2012, military figure prominent appearances gradually declined. Yet in 2016, the number of military figure prominent appearances increased - not quite to the levels experienced during the 2004 election, but to nonetheless relatively high levels. 

```{r mil-endorse, include=TRUE, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, fig.height=5, fig.width=7.5, fig.cap="Prominent Appearances of Military Figures in Presidential Campaign TV Ads, By Type of Military Actor, 2000-2016  "}

# create a df with the variables I will need
prom_figs_df <- adj_df %>% filter(vet_fam!=0) %>% select (year, count, creative, party, sponsor, ad_purpose, can_vet, can_vet_mention, vet_fam, vet_fam_end, vet_fam_atk, vet_fam_out_issues, prom_mil_fig) %>% group_by(year, party) %>% view()

# create counts of all of the pertinent variables such as 
# counts of prominent veteran appearances, prominent vet family appearances, etc.

prom_figs_df <- prom_figs_df %>% 
  group_by(year) %>% mutate(
    veteran_prominent=case_when(prom_mil_fig=="vet" ~ 1, prom_mil_fig!="vet" ~0),
    total_veteran_prominent=sum(veteran_prominent),
    elite_prominent=case_when(prom_mil_fig=="elite" ~1, prom_mil_fig !="elite" ~0),
    total_elite_prominent=sum(elite_prominent),
    family_vet_prominent=case_when(prom_mil_fig=="family" ~1, prom_mil_fig !="family" ~0),
    total_family_vet_prominent=sum(family_vet_prominent),
    elite_prominent_gen_sense=case_when(prom_mil_fig=="elitegeneralsense"~1, prom_mil_fig!="elitegeneralsense" ~0),
    total_elite_mil_fig_gen_sense=sum(elite_prominent_gen_sense),
    elite_pol_mil_fig=case_when(prom_mil_fig=="elitepolfig" ~1, prom_mil_fig!="elitepolfig"~0),
    total_elite_pol_mil_fig=sum(elite_pol_mil_fig),
     total_elite_appearances=total_elite_prominent+total_elite_mil_fig_gen_sense+total_elite_pol_mil_fig,
    total_prominent_appearances=total_veteran_prominent + total_elite_prominent + total_family_vet_prominent + total_elite_mil_fig_gen_sense + total_elite_pol_mil_fig
    ) %>% view()

# begin making the first facet / panel graph wrap 

facet_wrap1 <- prom_figs_df %>% select(year, veteran_prominent:total_prominent_appearances) %>% distinct() %>% select(-veteran_prominent, -elite_prominent, -family_vet_prominent, -elite_prominent_gen_sense, -elite_pol_mil_fig, -total_elite_appearances) %>% view()

# now make the data into longer format for graphing and prepare to graph

facet_wrap1 <- facet_wrap1  %>% pivot_longer(total_veteran_prominent:total_prominent_appearances, names_to = "question", values_to = "value") %>% view ()

plot3 <- ggplot(data=facet_wrap1, aes(fill=question, x=year, y=value)) + geom_bar(position="dodge", stat="identity") + labs(x="Election", y="Number of Ads Featuring Prominent\n Appearances by Military Figures") + scale_fill_discrete(name="Actor",
                         breaks=c("total_veteran_prominent", "total_elite_prominent", "total_family_vet_prominent", "total_elite_mil_fig_gen_sense", "total_elite_pol_mil_fig", "total_prominent_appearances"),
                         labels=c("Vets", "Retired Gen./Adm.","Family Members of Vets", "Retired Gen./Adm. Mentioned/Pictured Only", "Pol.Figures w/Mil. Backgrounds", "Total Appearances")) + theme(legend.position="bottom") + scale_x_continuous(breaks=seq(2000, 2016, 4))

plot3

```

The 2004 presidential election deserves a quick note to orient the reader to some of the underlying reasons for the high number of military figure prominent appearances (42 in total) in that election's campaign advertisements. In 2004, a number of veterans, family member of veterans, and retired officers coalesced to attack the war record of the Democratic Candidate, Senator John Kerry. The group, which was an official political action committee known as the "Swiftboat Veterans for Truth," claimed that Kerry had exaggerated his war record and was undeserving of many of the heroic medals he had received [@ullman_whats_2008]. The group even published a book that turned out to be a _New York Times_ bestseller, whose title embodied the group's main charge against Kerry: "Unfit for Command" [@buckaloo_swiftboat_nodate]. 

Although it is hard to state conclusively, the attacks levied by the Swiftboat Veterans for Truth against Kerry were almost certainly effective. One scholar argues that the combination of the veracity of the Swiftboat ads, the fact that they largely featured real veterans as opposed to paid actors, and the timing of their release (the ads were released in August, the "doldrums period between conventions," which facilitated these ads having quite a bit of exposure in the media) badly hurt Kerry's chances as a candidate [@devlin_contrasts_2005, 292].

It is not my central purpose in this chapter to render a normative judgement on the actions of the Swiftboat Veterans. One can reasonably state that such behavior was inherently partisan, for the underlying purpose of the group was to prevent Kerry from ascending to the Presidency. From this perspective, the actions of the Swiftboat Veterans must be viewed as wrong in that they violated the military's non-partisan ethic. On the other hand, one might also conclude that the attacks came primarily from Kerry's fellow Vietnam Veterans and for the central purpose of, at least from the Swiftboat Veterans perspective, correcting Kerry's untruthful and embellished war record. Considering the divisions in American society that the Vietnam War revealed, one might reasonably conclude that while the impact of the Swiftboat Veterans actions was partisan, their main intent (though certainly not that of the Republican Party) was to set the record straight on Kerry rather than to meddle in partisan politics. 

Regardless of one's normative opinion on the actions of the Swiftboat Veterans for Truth in 2004, there is a much more significant story that needs to be explained given the data presented in Figure \@ref(fig:mil-endorse), in my view. This significant story involves the fact that there are still a relatively large number of military figure prominent appearances in elections besides 2004, when the candidates running for office were not veterans (2012 and 2016). 

### Violations of the Principles of Non-Partisanship and Non-Interference

Figure \@ref(fig:actor-principle) displays the number of distinct advertisements aired in each election cycle in which retired generals and admirals, veterans, and veteran family members explicitly endorse or attack a presidential candidate running for office, or address issues that are tangential if not altogether outside of military expertise during presidential elections from 2000 - 2016. 

```{r actor-principle-prep, include=FALSE, echo=FALSE, error=FALSE, warning=FALSE}

prom_figs_df %>% view()

facet_wrap2 <- prom_figs_df %>% group_by(year) %>% mutate(
  vet_endorse=case_when(prom_mil_fig=="vet" & vet_fam_end==1 ~ 1, prom_mil_fig!="vet" | vet_fam_end != 1 ~0),
  vet_endorse_total=sum(vet_endorse),
  elite_endorse=case_when((prom_mil_fig=="elite" | prom_mil_fig=="elitegeneralsense" | prom_mil_fig=="elitepolfig") & vet_fam_end==1 ~ 1, (prom_mil_fig!="elite" | prom_mil_fig!="elitegeneralsense" | prom_mil_fig!="elitepolfig") | vet_fam_end!=1 ~ 0), 
  elite_endorse_total=sum(elite_endorse), 
  family_endorse=case_when(prom_mil_fig=="family" & vet_fam_end==1 ~1, prom_mil_fig!="family" | vet_fam_end != 1 ~0),
  family_endorse_total=sum(family_endorse),
  vet_attack=case_when(prom_mil_fig=="vet" & vet_fam_atk==1 ~ 1, prom_mil_fig!="vet" | vet_fam_atk != 1 ~0),
  vet_attack_total=sum(vet_attack),
  elite_attack=case_when((prom_mil_fig=="elite" | prom_mil_fig=="elitegeneralsense" | prom_mil_fig=="elitepolfig") & vet_fam_atk==1 ~ 1, (prom_mil_fig!="elite" | prom_mil_fig!="elitegeneralsense" | prom_mil_fig!="elitepolfig") | vet_fam_atk!=1 ~ 0), 
  elite_attack_total=sum(elite_attack), 
  family_attack=case_when(prom_mil_fig=="family" & vet_fam_atk==1 ~1, prom_mil_fig!="family" | vet_fam_atk != 1 ~0),
  family_attack_total=sum(family_attack),
  vet_issues=case_when(prom_mil_fig=="vet" & vet_fam_out_issues==1 ~ 1, prom_mil_fig!="vet" | vet_fam_out_issues != 1 ~0),
  vet_issues_total=sum(vet_issues),
  elite_issues=case_when((prom_mil_fig=="elite" | prom_mil_fig=="elitegeneralsense" | prom_mil_fig=="elitepolfig") & vet_fam_out_issues==1 ~ 1, (prom_mil_fig!="elite" | prom_mil_fig!="elitegeneralsense" | prom_mil_fig!="elitepolfig") | vet_fam_out_issues!=1 ~ 0), 
  elite_issues_total=sum(elite_issues), 
  family_issues=case_when(prom_mil_fig=="family" & vet_fam_out_issues==1 ~1, prom_mil_fig!="family" | vet_fam_out_issues != 1 ~0),
  family_issues_total=sum(family_issues),
  endorse_total=vet_endorse_total+elite_endorse_total+family_endorse_total,
  attack_total=vet_attack_total+elite_attack_total+family_attack_total,
  issues_total=vet_issues_total+elite_issues_total+family_issues_total
) %>% select(year, party, sponsor, ad_purpose, prom_mil_fig, total_veteran_prominent, total_family_vet_prominent, total_elite_appearances, total_prominent_appearances, vet_endorse_total, elite_endorse_total, family_endorse_total, vet_attack_total, elite_attack_total, family_attack_total, vet_issues_total, elite_issues_total, family_issues_total, endorse_total, attack_total, issues_total) %>% view()


# this line contains all of the information needed for the subsequent graphs!
facet_wrap2graph <- facet_wrap2 %>% select(year, total_veteran_prominent:issues_total) %>% distinct() %>% view()

#all military figures

milfigures_plot <- facet_wrap2graph %>% select(endorse_total, attack_total, issues_total, total_prominent_appearances) %>% print()

milfigures_plot_graph <- milfigures_plot %>% pivot_longer(endorse_total:total_prominent_appearances, names_to="category", values_to="value") %>% view()

milfigures_plot_graph <- ggplot(milfigures_plot_graph, aes(fill=category, x=year, y=value)) + geom_bar(position="dodge", stat="identity") + theme_bw() + theme(legend.position=c(.4, .85)) + scale_x_continuous(breaks=seq(2000, 2016, by = 4)) + scale_y_continuous(limits=c(0, 55, by=5)) + labs(x="Election", y="Number of Instances") + scale_fill_grey(start=0.2, end=0.8, name="All Military Figures",
                         breaks=c("attack_total", "endorse_total", "issues_total", "total_prominent_appearances"),
                         labels=c("Attack Candidate", "Endorse Candidate","Address Topics Outside of Expertise", "Total Appearances"))

milfigures_plot_graph

# prepare to make individual plots

#elite graphs
elite_plot <- facet_wrap2graph %>% select(elite_endorse_total, elite_attack_total, elite_issues_total, total_elite_appearances) %>% print()

elite_plot_graph <- elite_plot %>% pivot_longer(elite_endorse_total:total_elite_appearances, names_to = "category", values_to = "value") %>% view()

elite_plot_graph <- ggplot(elite_plot_graph, aes(fill=category, x=year, y=value)) + geom_bar(position="dodge", stat="identity") + theme_bw() + theme(legend.position=c(.4, .85)) + scale_x_continuous(breaks=seq(2000, 2016, by = 4)) + scale_y_continuous(limits=c(0, 14, by=2)) + labs(x="Election", y="Number of Instances") + scale_fill_grey(start=0.2, end=0.8, name="Retired Generals/Admirals",
                         breaks=c("elite_attack_total", "elite_endorse_total", "elite_issues_total", "total_elite_appearances"),
                         labels=c("Attack Candidate", "Endorse Candidate","Address Topics Outside of Expertise", "Total Appearances"))

elite_plot_graph
  


#vet graphs

vet_plot <- facet_wrap2graph %>% select(vet_endorse_total, vet_attack_total, vet_issues_total, total_veteran_prominent) %>% print()

vet_plot_graph <- vet_plot %>% pivot_longer(vet_endorse_total:total_veteran_prominent, names_to="category", values_to="value") %>% view ()

vet_plot_graph$category <- gsub("vet_","", vet_plot_graph$category)

vet_plot_graph <- ggplot(vet_plot_graph, aes(fill=category, x=year, y=value)) + geom_bar(position="dodge", stat="identity") + theme_bw() + theme(legend.position=c(.4, .85)) + scale_x_continuous(breaks=seq(2000, 2016, by = 4)) + scale_y_continuous(limits=c(0, 30, by=4)) + labs(x="Election", y="Number of Instances") + scale_fill_grey(start=0.2, end=0.8, name="Veteran",
                         breaks=c("attack_total", "endorse_total", "issues_total", "total_veteran_prominent"),
                         labels=c("Attack Candidate", "Endorse Candidate","Address Topics Outside of Expertise", "Total Appearances"))

vet_plot_graph

# family graphs
family_plot <- facet_wrap2graph %>% select(family_endorse_total, family_attack_total, family_issues_total, total_family_vet_prominent) %>% print()

family_plot_graph <- family_plot %>% pivot_longer(family_endorse_total:total_family_vet_prominent, names_to="category", values_to="value") %>% print()

family_plot_graph <- ggplot(family_plot_graph, aes(fill=category, x=year, y=value)) + geom_bar(position="dodge", stat="identity") + theme_bw() + theme(legend.position=c(.5, .85)) + scale_x_continuous(breaks=seq(2000, 2016, by = 4)) + scale_y_continuous(limits=c(0, 20, by=4)) + labs(x="Election", y="Number of Instances") + scale_fill_grey(start=0.2, end=0.8, name="Family Members of Vets",
                         breaks=c("family_attack_total", "family_endorse_total", "family_issues_total", "total_family_vet_prominent"),
                         labels=c("Attack Candidate", "Endorse Candidate","Address Topics Outside of Expertise", "Total Appearances"))

family_plot_graph

```

```{r actor-principle, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.height=10, fig.width=8,fig.cap="Instances of A) Retired Generals and Admirals, B) Veterans, and C) Family Members of Veterans Engaging in Forms of Partisan Activity (Attack and Endorse Political Candidates, and Address Topics Outside of Military Expertise) in Presidential Campaign Television Ads, 2000 - 2016"}

plot_grid(milfigures_plot_graph, elite_plot_graph, vet_plot_graph, family_plot_graph, labels = "AUTO")

```
Moreover, Figure \@ref(fig:actor-principle) reveals several interesting points. First, Panel A shows that across all types of military figures who appear prominently in a campaign ad, these figures engage in overtly partisan behaviors when they do appear, such as attacking or endorsing candidates. The 2004 election elicited the greatest number of prominent military figure appearances (42 in total), which included 21 attacks, 11 endorsements, and 1 instance of addressing a topic that was tangential to military expertise. In 2016, on the other hand, out of 32 total military figure prominent appearances, there were 25 attacks, 15 endorsements, and three instances of addressing a topic that was tangential to military expertise. To the degree that it _may_ have been possible for a military figure to prominently appear in a campaign ad and not engage in an overtly partisan behavior - a claim that is questionable to begin with - the data indicate that, rather unsurprisingly, when military figures appear, they routinely and increasingly engage in explicitly partisan acts. 

This trend seems to hold across all types of military figures. For instance, Panel B indicates that of the two retired general/admiral appearances in the 2000 election, one involved addressing a topic outside of military expertise, but none explicitly endorses or attacked a political candidate. In 2008, the number of retired general/admiral appearances equaled the number of endorsements. And in 2016, of the ten retired generals or admirals who appeared prominently in campaign ads, these appearances included seven attacks and five endorsements. 

### On which Party's Behalf are Military Figures Violating the Principles of Civil-Military Relations? 

Given that civilian actors are increasingly airing campaign ads that feature military actors committing explicitly partisan behaviors, it is important to discern whether such behavior is serving the interests of one political party over another. One might suspect, given a moderate amount of literature that the US military more closely associates with the Republican Party than the Democratic Party [@davis_attitudes_2001; @ricks_widening_1997; see also @golby_duty_2011], that military figures engage in more partisan activity on behalf of the Republican rather than the Democratic Party. However, the results indicate otherwise. 

Figure \@ref(fig:behavior-party) displays the total number of military figure prominent appearances, endorsements, attacks, and instances of addressing issues that are tangential to military expertise separated by the party supported by the advertisement. The results indicate that from 2000-2016, with the exception of the 2012 election, military figures who appear prominently during campaign ads engage in partisan behaviors more often on behalf of Democratic rather than Republican candidates. 

```{r behavior-party-prep, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE, error=FALSE}

# partisan appearances
mil_fig_appear_by_party <- prom_figs_df %>% ungroup() %>% select(year, party) %>% group_by(year, party) %>% mutate(count = n()) %>% distinct() %>% rename(value=count) %>% view ()

mil_fig_appear_by_party_graph <- ggplot(mil_fig_appear_by_party, aes(fill=party, x=year, y=value)) + geom_bar(position="dodge", stat="identity", width=.95) + theme_bw() + theme(legend.position=c(.5, .80)) + scale_x_continuous(breaks=seq(2000, 2016, by = 4)) + scale_y_continuous(limits=c(0, 40, by=4)) + labs(x="Election", y="Number") + scale_fill_grey(start=0.2, end=0.8, name="Military Figure Appearances on Behalf of", breaks=c("DEMOCRAT", "REPUBLICAN", "OTHER"), labels=c("Democratic Candidate", "Republican Candidate","Other"))

mil_fig_appear_by_party_graph
                                        
# partisan endorsements
mil_fig_endorse_by_party <- prom_figs_df %>% filter(vet_fam_end==1) %>% ungroup() %>% select(year, party, vet_fam_end) %>% group_by(year, party, vet_fam_end) %>% mutate(count = n()) %>% distinct() %>% rename(value=count) %>% print()

mil_fig_endorse_by_party_graph <- ggplot(mil_fig_endorse_by_party, aes(fill=party, x=year, y=value)) + geom_bar(position="dodge", stat="identity", width=.95) + theme_bw() + theme(legend.position=c(.5, .80)) + scale_x_continuous(breaks=seq(2000, 2016, by = 4)) + scale_y_continuous(limits=c(0, 12, by=2)) + labs(x="Election", y="Number") + scale_fill_grey(start=0.2, end=0.8, name="Military Figure Endorsements\n on Behalf of", breaks=c("DEMOCRAT", "REPUBLICAN", "OTHER"), labels=c("Democratic Candidates", "Republican Candidates","Other"))

mil_fig_endorse_by_party_graph

# partisan attacks
mil_fig_attack_by_party <- prom_figs_df %>% filter(vet_fam_atk==1) %>% ungroup() %>% select(year, party, vet_fam_atk) %>% group_by(year, party, vet_fam_atk) %>% mutate(count = n()) %>% distinct() %>% rename(value=count) %>% print()

mil_fig_attack_by_party_graph <- ggplot(mil_fig_attack_by_party, aes(fill=party, x=year, y=value)) + geom_bar(position="dodge", stat="identity", width=.95) + theme_bw() + theme(legend.position=c(.4, .80)) + scale_x_continuous(breaks=seq(2000, 2016, by = 4)) + scale_y_continuous(limits=c(0, 24, by=4)) + labs(x="Election", y="Number") + scale_fill_grey(start=0.2, end=0.8, name="Military Figure Attacks in\n Campaign Ads on Behalf of", breaks=c("DEMOCRAT", "REPUBLICAN", "OTHER"), labels=c("Democratic Candidates", "Republican Candidates","Other"))

mil_fig_attack_by_party_graph


# partisan address issues outside of expertise
mil_fig_issues_by_party <- prom_figs_df %>% filter(vet_fam_out_issues==1) %>% ungroup() %>% select(year, party, vet_fam_out_issues) %>% group_by(year, party, vet_fam_out_issues) %>% mutate(count = n()) %>% distinct() %>% rename(value=count) %>%print()

mil_fig_issues_by_party_graph <- ggplot(mil_fig_issues_by_party, aes(fill=party, x=year, y=value)) + geom_bar(position="dodge", stat="identity", width=.95) + theme_bw() + theme(legend.position=c(.5, .80)) + scale_x_continuous(breaks=seq(2000, 2016, by = 4)) + scale_y_continuous(limits=c(0, 6, by=2)) + labs(x="Election", y="Number") + scale_fill_grey(start=0.2, end=0.8, name="Military Figures Addressing Topics Outside\n of Military Expertise In Campaign Ads on Behalf of ", breaks=c("DEMOCRAT", "REPUBLICAN", "OTHER"), labels=c("Democratic Candidates", "Republican Candidates","Other"))

mil_fig_issues_by_party_graph


```

```{r behavior-party, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE, error=FALSE, fig.cap="Military Figure A) Appearances, B) Endorsements, C) Attacks, and D) Addressing Topics that are Tangential to National Security in Presidential Campaign Television Ads, 2000 - 2016, By Party Affiliation of Sponsor of Ad", fig.height=10, fig.width=9}

plot_grid(mil_fig_appear_by_party_graph, mil_fig_endorse_by_party_graph, mil_fig_attack_by_party_graph, mil_fig_issues_by_party_graph, labels = "AUTO")

```

Panels A and C of Figure \@ref(fig:behavior-party) show that the _only_ times that Democratic-affiliated military figure partisan activity during campaign ads did not outpace that of Republican-affiliated military figure appearances occurred in 2012, when the number of military figure appearances on behalf of Republican candidates were slightly higher (Panel A), and when military figures attacked political candidates more than Democratic-affiliated military figures (Panel C). 

Are these results inconsistent with literature suggesting that the military is more closely aligned with the Republican than the Democratic Party? I suggest that in answering this important question, it is important to remember the distinction between _non-visible ideological alignment_ and _visible partisan behavior_. The traditional view that the military is ideologically affiliated with the Republican Party more than the Democratic Party relies on a thorough body of survey data in which researchers compared the attitudes of military officers in the late 1990s to a diverse set of civilian groups [@feaver_soldiers_2001].^[These surveys were spearheaded by the Triangle Institute for Security Studies (TISS), a joint effort between researchers at Duke University and the University of North Carolina at Chapel-Hill.]. This survey data itself constitutes an impressive body of scholarship aimed at determining whether and the degree to which a gap existed between the military and civilians, and what form such a gap took [@feaver_soldiers_2001; see also @ricks_widening_1997; @golby_duty_2011; and @golby_thanks_2016]. 

The TISS survey data - more than 20 years old - showed that the military overwhelmingly identified with the Republican Party, to the point that among military leaders, 63.9% identified with Republicans, compared to 8.1% with Democrats, 16.7% with Independents, and 11.3% with no or another political party [@holsti_chasms_2001, 28]. When asked to identify their own ideological affiliation in liberal or conservative terms, similar results emerged among military leaders: only 4.4% identified as "very liberal" or "somewhat liberal," compared to 37.5% of civilian non-veteran leaders and 28.5% of the public who were not veterans [@holsti_chasms_2001, 33]. Some scholars of civil-military relations saw the ideological gap reflected in these results as concerning in that the ideological beliefs of the military officer corps varied too much from those of the American public. 

However, there is a significant difference between invisible ideological preferences held by military officers and visible partisan behaviors committed by them. The latter _proves_ that the military is failing to uphold to a principle of non-partisanship. The former, on the other hand, may still a cause for concern, but only insofar as an ideological imbalance may lead to _behavior_ that favors one party over another. 

A lot has changed in the more than two decades since the TISS survey data was collected. It is possible, and perhaps likely, that the military no longer shares as strong as an ideological connection to the Republican Party as it did in the late 1980s and 1990s. More recent data is necessary in order to draw such a conclusion. 

It could also be that the trends uncovered in this chapter of military figures engaging in explicitly partisan behavior more frequently on behalf of Democratic candidates than Republican candidates indicate that Democratic candidates want to convey to the public that they are not weaker than or inferior to Republican candidates on defense-related issues. Democratic candidates may desire to employ a military figure to endorse them, or to attack their political opponent, because doing so might make the Democratic candidate appear strong. Such a possibility is consistent with the "issue ownership" literature in American politics, which suggests that the political parties are known to have advantages in certain issue areas [@petrocik_issue_2003; @sides_consequences_2007]. Such a possibility poses the basis for fertile research in the future, especially as polarized politics in the US endure and as several recent events - to include the withdrawal of the military from Afghanistan, and very contentious Congressional testimony by military leaders in its aftermath - may signal that any actual or perceived ties between the military and the GOP are weakening. 

## Explaining Campaign Ads that Feature Explicitly Partisan Behavior by Military Figures

The next several paragraphs reveal the results of regression analysis for a series of different model specifications. I control for military prestige by including a measure of public trust and confidence in the military, as reported through Gallup public opinion polling regarding trust in institutions [@noauthor_confidence_2020]. This control variable is intended to capture the intuition that civilians are more likely to use the military for their own partisan gain and purposes as military prestige rises. 

Consistent with the regression analysis conducted in the previous chapter, I include several measurements of polarization. These include separate measurements of ideology in the US House and Senate [@jeffrey_b_lewis_voteview_2020], and a measure of affective polarization, which measures how partisans who identify with one political party generally feel about the other political party as reported in data from the American National Election Studies (ANES) [@iyengar_affect_2012; @iyengar_strengthening_2018; @iyengar_origins_2019]. 

In some model specifications, I include several other control variables that could theoretically drive variation in the level and frequency of explicitly partisan behavior conducted by military figures in civilian campaign advertisements. These include a measure of military casualties sustained in a given year, and whether at least one candidate mentioned in the advertisement is a veteran.^[This variable is constructed from a number of sources, and consists of the number of hostile US casualties sustained in a given calendar year per 100,000 Active Duty military forces. I include in this measurement those casualties that are the result of what the Department of Defense classifies as either "hostile action" or a "terrorist attack." Annual data is provided from the Department of Defense from 1980-2010.  I then used data from the Department of Defense regarding the wars in Iraq and Afghanistan to construct a measurement for years outside of this time range. See @department_of_defense_dcas_2011; @department_of_defense_dcas_2021; @department_of_defense_dcas_2021-1; @department_of_defense_dcas_2021-2; @department_of_defense_dcas_2021-3; @department_of_defense_dcas_2021-4; @department_of_defense_dcas_2021-5; @department_of_defense_dmdc_2021 for more details.] I control for the level of military casualties sustained through hostile action in a given year because troop casualty levels may make civilian candidates more willing to use military figures to attack and/or endorse political candidates, especially if the views of these candidates on the use of force strongly align or clash with those held by a particular type of military figure. I include a candidate's veteran status because civilian candidates who themselves are veterans may likewise spur their political opponents to employ military figures in ways that undermine the veteran candidates service record or views regarding the use of force. 

The dependent variables in each model is a dichotomous indicator of the log odds of a military figure appearing prominently in a campaign advertisement (Models 1-3) and explicitly endorsing a candidate, attacking a candidate, or addressing a topic that falls outside of the military figure's expertise (Models 4-8). Models 1-6 are logistic regressions, and models 7-8 are mixed models that includes year random effects to allow for the possibility that a particular election cycle in a given year has special characteristics that are driving variation in the dependent variable. The results of these regressions are displayed in Table \@ref(tab:regression6) on the following page.

### Regression Analysis

```{r reg data manipulation, include=FALSE, echo=FALSE, message=FALSE, error=FALSE}

# prep the data 
reg_df <- adj_df %>% select (year, count, creative, party, sponsor, ad_purpose, can_vet, can_vet_mention, vet_fam, vet_fam_end, vet_fam_atk, vet_fam_out_issues, prom_mil_fig, def_vet_ad_theme:sum_non_def_vet_ad_theme) %>% group_by(year, party) %>% view() 


# create explicit variables for each partisan action of endorse, attack, speak out on tangential issues.  
# vet_fam_end ==1 or if vet_fam_atk ==1 or if vet_fam_out_issues ==1

# also, use include several control variables to the data frame.  
reg_df <- reg_df %>% mutate(
  partisan_act=case_when(vet_fam_end ==1 | vet_fam_atk ==1 | vet_fam_out_issues ==1 ~ 1, (vet_fam_end !=1 & vet_fam_atk!=1 & vet_fam_out_issues!=1)~0),
  dummy_04=case_when(
    year == 2004 ~ 1, 
    year != 2004 ~0), 
  partisan_appearance=case_when(
    vet_fam > 0 ~ 1, 
    vet_fam ==0 ~ 0),
  endorse=case_when(
    vet_fam_end == 1 ~ 1, 
    vet_fam_end != 1 ~ 0), 
  attack=case_when(
    vet_fam_atk == 1 ~ 1, 
    vet_fam_atk != 1 ~ 0),
  speak_out=case_when(
    vet_fam_out_issues == 1 ~ 1, 
    vet_fam_out_issues != 1 ~ 0),
  party_reg=case_when(
    party=="DEMOCRAT" ~ "DEMOCRAT",
    party=="REPUBLICAN" ~ "REPUBLICAN",
    (party!= "REPUBLICAN") & (party!= "DEMOCRAT") ~ "OTHER"
  ),
  party_reg_numeric=case_when(
    party_reg=="DEMOCRAT" ~ 1,
    party_reg=="REPUBLICAN" ~ 0,
    party_reg=="OTHER" ~2
  ),
  house_polar = case_when(
    year==2000 ~ 77.713,
    year == 2004 ~ 79.302,
    year == 2008 ~ 80.765,
    year == 2012 ~ 86.142,
    year == 2016 ~ 87.644
    ), 
  sen_polar = case_when(
    year==2000 ~ 65.833,
    year == 2004 ~ 64.992,
    year == 2008 ~ 68.744,
    year == 2012 ~ 74.383,
    year == 2016 ~ 81.304
  ),
  aff_polar = case_when(
    year==2000 ~ 61.692,
    year == 2004 ~ 68.052,
    year == 2008 ~ 69.519,
    year == 2012 ~ 84.689,
    year == 2016 ~ 67.754,
  ),
  HostCasRate = case_when(
    year==2000 ~ 1.23874924,
    year == 2004 ~ 52.0801226,
    year == 2008 ~ 25.1029256,
    year == 2012 ~ 29.4917871,
    year == 2016 ~ 1.229532
  ),
  mil_prestige = case_when(
    year== 2000 ~ 64,
    year == 2004 ~ 75,
    year == 2008 ~ 71,
    year == 2012 ~ 75,
    year == 2016 ~ 73
  ), 
  prty_ofc = case_when(
    year== 2000 ~ 1,
    year == 2004 ~ 0,
    year == 2008 ~ 0,
    year == 2012 ~ 1,
    year == 2016 ~ 1
  ),
  prop_year_mil_img = (sum_mil_img_cycle / Yrcount) * 100, 
  prop_not_defense_theme_but_mil_img = (sum_not_def_theme_but_mil_img / Yrcount) * 100
  ) %>% view()


### Here I construct my regression models (mostly logit to begin with).

## DV: Prominent Appearance by any military figure, vet family member, etc. Where partisan_appearance ==1. 

mod_house <- glm(partisan_appearance ~ mil_prestige + house_polar,
    family=binomial(link="logit"),
    data=reg_df
    )

summary(mod_house)
tidy(mod_house)
glance(mod_house)

mod_senate <- glm(partisan_appearance ~ mil_prestige + sen_polar,
    family=binomial(link="logit"),
    data=reg_df
    )

summary(mod_senate)
tidy(mod_senate)
glance(mod_senate)

mod_aff <- glm(partisan_appearance ~ mil_prestige + aff_polar,
    family=binomial(link="logit"),
    data=reg_df
    )

summary(mod_aff)
tidy(mod_aff)
glance(mod_aff)


## DV: The occurrence of a Partisan Act, which is defined as either an endorsement, an attack, 
# or speaking out about an issue.  Where partisan_act == 1.

mod1_house <- glm(partisan_act ~ mil_prestige + house_polar + party_reg,
    family=binomial(link="logit"),
    data=reg_df
    )

summary(mod1_house)
tidy(mod1_house)
glance(mod1_house)

mod1_senate <- glm(partisan_act ~ mil_prestige + sen_polar + HostCasRate + can_vet + party_reg,
    family=binomial(link="logit"),
    data=reg_df
    )

summary(mod1_senate)
tidy(mod1_senate)
glance(mod1_senate)

mod1_aff <- glm(partisan_act ~ mil_prestige + aff_polar + HostCasRate + can_vet + party_reg,
    family=binomial(link="logit"),
    data=reg_df
    )

summary(mod1_aff)
tidy(mod1_aff)
glance(mod1_aff)

# use this model, below, for predicted probability
mod1_aff_pp <- glm(partisan_act ~ mil_prestige + aff_polar + HostCasRate + can_vet + party_reg_numeric,
    family=binomial(link="logit"),
    data=reg_df
    )

summary(mod1_aff_pp)
tidy(mod1_aff_pp)
glance(mod1_aff_pp)


## DV: Now I go through each partisan act specifically: the first is endorsement. 
# I also include now controls for the hostile casualty rate and whether the candidate was a veteran.   

mod2_house <- glm(endorse ~ mil_prestige + house_polar + HostCasRate + can_vet,
    family=binomial(link="logit"),
    data=reg_df
    )

summary(mod2_house)
tidy(mod2_house)
glance(mod2_house)


mod2_senate <- glm(endorse ~ mil_prestige + sen_polar + HostCasRate + can_vet,
    family=binomial(link="logit"),
    data=reg_df
    )

summary(mod2_senate)
tidy(mod2_senate)
glance(mod2_senate)

mod2_aff <- glm(endorse ~ mil_prestige + aff_polar + HostCasRate + can_vet,
    family=binomial(link="logit"),
    data=reg_df
    )

summary(mod2_aff)
tidy(mod2_aff)
glance(mod2_aff)


## DV: Attacking a candidate.  
  
mod3_house <- glm(attack ~ mil_prestige + house_polar + HostCasRate + can_vet,
    family=binomial(link="logit"),
    data=reg_df
    )

summary(mod3_house)
tidy(mod3_house)
glance(mod3_house)


mod3_senate <- glm(attack ~ mil_prestige + sen_polar + HostCasRate + can_vet,
    family=binomial(link="logit"),
    data=reg_df
    )

summary(mod3_senate)
tidy(mod3_senate)
glance(mod3_senate)

mod3_aff <- glm(attack ~ mil_prestige + aff_polar + HostCasRate + can_vet,
    family=binomial(link="logit"),
    data=reg_df
    )

summary(mod3_aff)
tidy(mod3_aff)
glance(mod3_aff)


## DV: Speaking out on Other Issues.   
  
mod4_house <- glm(speak_out ~ mil_prestige + house_polar + HostCasRate + can_vet,
    family=binomial(link="logit"),
    data=reg_df
    )

summary(mod4_house)
tidy(mod4_house)
glance(mod4_house)


mod4_senate <- glm(speak_out ~ mil_prestige + sen_polar + HostCasRate + can_vet,
    family=binomial(link="logit"),
    data=reg_df
    )

summary(mod4_senate)
tidy(mod4_senate)
glance(mod4_senate)

mod4_aff <- glm(speak_out ~ mil_prestige + aff_polar + HostCasRate + can_vet,
    family=binomial(link="logit"),
    data=reg_df
    )

summary(mod4_aff)
tidy(mod4_aff)
glance(mod4_aff)


### Now move into some OLS regressions on:
# - the proportion of ads in a given cycle with military imagery.

macro_img_reg_df <- reg_df %>% ungroup(year, party) %>% select(year, Yrcount, mil_prestige, house_polar, sen_polar, aff_polar, HostCasRate, prop_year_mil_img, prop_not_defense_theme_but_mil_img) %>% group_by(year) %>% distinct()

mod5_mil_img_cycle_house <- lm(prop_year_mil_img ~ mil_prestige + house_polar + HostCasRate, data=macro_img_reg_df)
summary(mod5_mil_img_cycle_house)

mod5_mil_img_cycle_sen <- lm(prop_year_mil_img ~ mil_prestige + sen_polar + HostCasRate, data=macro_img_reg_df)
summary(mod5_mil_img_cycle_sen)

mod5_mil_img_cycle_aff <- lm(prop_year_mil_img ~ mil_prestige + aff_polar + HostCasRate, data=macro_img_reg_df)
summary(mod5_mil_img_cycle_aff)

# - the proportion of ads with military imagery that are NOT related to an all-defense ad but still have military imagery. 

mod6_mil_img_cycle_house <- lm(prop_not_defense_theme_but_mil_img ~ mil_prestige + house_polar + HostCasRate, data=macro_img_reg_df)
summary(mod6_mil_img_cycle_house)

mod6_mil_img_cycle_sen <- lm(prop_not_defense_theme_but_mil_img ~ mil_prestige + sen_polar + HostCasRate, data=macro_img_reg_df)
summary(mod6_mil_img_cycle_sen)

mod6_mil_img_cycle_aff <- lm(prop_not_defense_theme_but_mil_img ~ mil_prestige + aff_polar + HostCasRate, data=macro_img_reg_df)
summary(mod6_mil_img_cycle_aff)


### When complete, do I want to do a mixed model or a model with robust standard errors at the 'year' level
## to allow for the possibility that some election years are just, simply, different than others in meaningful ways?

mod7_endorse_house <- robustify(mod2_house, cluster=year)
summary(mod7_endorse_house)

mod7_attack_senate <- robustify(mod3_senate, cluster=year)
summary(mod7_attack_senate)

# Then conduct a mixed model where we include author random effects
mod8_partisan_act_mixed <- glmer(partisan_act ~ mil_prestige + house_polar + party_reg + (1 | year), family=binomial, data=reg_df)
summary(mod8_partisan_act_mixed)

mod9_partisan_act_mixed <- glmer(partisan_act ~ mil_prestige + sen_polar + party_reg + (1 | year), family=binomial, data=reg_df)
summary(mod9_partisan_act_mixed)

## Endogeneity Checks
# Regress military prestige on partisan_act and polarization

mod10_ch4_endog_house <- lm(mil_prestige ~ partisan_act + house_polar + HostCasRate, data=reg_df)
summary(mod10_ch4_endog_house)

mod10_ch4_endog_house_dummy <- lm(mil_prestige ~ partisan_act + house_polar + HostCasRate + dummy_04, data=reg_df)
summary(mod10_ch4_endog_house_dummy)

mod10_ch4_endog_sen <- lm(mil_prestige ~ partisan_act + sen_polar + HostCasRate + can_vet, data=reg_df)
summary(mod10_ch4_endog_sen)

mod10_ch4_endog_sen_dummy <- lm(mil_prestige ~ partisan_act + sen_polar + HostCasRate + can_vet + dummy_04, data=reg_df)
summary(mod10_ch4_endog_sen_dummy)

## Now include an interaction term for crit_part_interfere and polarization

# with house measurements

mod10_ch4_endog_house_int_simple <- lm(mil_prestige ~ partisan_act + house_polar + partisan_act*house_polar, data=reg_df)
summary(mod10_ch4_endog_house_int_simple)

mod10_ch4_endog_house_int_simple_dummy <- lm(mil_prestige ~ partisan_act + house_polar + partisan_act*house_polar + dummy_04, data=reg_df)
summary(mod10_ch4_endog_house_int_simple_dummy)

mod10_ch4_endog_house_int <- lm(mil_prestige ~ partisan_act + house_polar + HostCasRate + can_vet +  partisan_act*house_polar, data=reg_df)
summary(mod10_ch4_endog_house_int)

mod10_ch4_endog_house_int_dummy <- lm(mil_prestige ~ partisan_act + house_polar + HostCasRate + can_vet +  partisan_act*house_polar + dummy_04, data=reg_df)
summary(mod10_ch4_endog_house_int_dummy)


# with sen measurements

mod10_ch4_endog_sen_int_simple <- lm(mil_prestige ~ partisan_act + sen_polar + partisan_act*sen_polar, data=reg_df)
summary(mod10_ch4_endog_sen_int_simple)

mod10_ch4_endog_sen_int_simple_dummy <- lm(mil_prestige ~ partisan_act + sen_polar + partisan_act*sen_polar + dummy_04, data=reg_df)
summary(mod10_ch4_endog_sen_int_simple_dummy)

mod10_ch4_endog_sen_int <- lm(mil_prestige ~ partisan_act + sen_polar + HostCasRate + can_vet + partisan_act*sen_polar, data=reg_df)
summary(mod10_ch4_endog_sen_int)

mod10_ch4_endog_sen_int_dummy <- lm(mil_prestige ~ partisan_act + sen_polar + HostCasRate + can_vet + partisan_act*sen_polar + dummy_04, data=reg_df)
summary(mod10_ch4_endog_sen_int_dummy)


# for MICE regressions using imputed data
mod_house_imp <- with(df_campaign_ads_raw_mice_imputed, glm(vet_fam ~ mil_prestige + house_polar, family=binomial(link="logit")))

summary(pool(mod_house_imp))

mod_sen_imp <- with(df_campaign_ads_raw_mice_imputed, glm(vet_fam ~ mil_prestige + sen_polar, family=binomial(link="logit")))

summary(pool(mod_sen_imp))



```

\singlespace
```{r Regression Results_ch4, include=TRUE, echo=FALSE, results="asis", warning=FALSE, message=FALSE}
stargazer(mod_house, mod_senate, mod_aff, mod1_house, mod1_senate, mod1_aff, mod8_partisan_act_mixed, mod9_partisan_act_mixed,
          type="latex",
          label="tab:regression6",
          title="Regression Results", 
          covariate.labels = c('Trust in Military', 'House Polarization', 'Senate Polarization', 'Aff. Polarization', 'Hostile Casualty Rate', 'Candidate-Veteran', 'Party-Other', 'Party-GOP'),
          dep.var.labels = c('Military Figure Appearing Prominently', 'Endorsing or Attacking Candidate, or Addressing Topic Outside of Military Expertise'),
          model.names=TRUE,
          header=FALSE,
          column.sep.width = "1pt",
          align=TRUE,
          font.size="small",
          no.space=TRUE,
          omit.stat=c("f", "ser"),
          float.env="sidewaystable"
)
         
```
\doublespace

The regression results provide general support for H3, which stated that civilian actors increasingly leverage the military for their own political purposes and advantages when military prestige is high relative to periods of low military prestige. Across all model specifications, trust in the military is positive and statistically significant, suggesting that civilians employ the military and military figures differently when prestige is relatively higher. In other words, the statistical evidence supports the notion that increasing military prestige indeed provides an _opportunity_ for civilian actors to involve the military in politics. 

In models 1-3, where the dependent variable is the log odds of a military figure appearing prominently in a campaign advertisement, and only controls are the levels of trust in the military and political polarization, the coefficient on polarization is negative, which interestingly suggests that increasing polarization decreases the likelihood that a civilian candidate will employ the military to appear in a prominent way, or to commit an explicitly partisan act, such as endorsing or attacking a candidate during the campaign advertisement. This is a potentially interesting finding, as it suggests that rising polarization could offset increasing military prestige. Statistically speaking, however, across all model specifications, the coefficient on various measures of polarization are not consistently significant. The coefficient for polarization is even positive in model 8. Therefore, I do not strongly suspect that polarization has consistent explanatory power in terms of impacting how civilian candidates employ the military in campaign ads, given the results of these regressions. 

In models 4-8, where the dependent variable is the log odds of a military figure either endorsing or attacking a candidate, or addressing a topic that is outside of military expertise, military prestige continues to remain statistically significant and positive, even as we introduce other control variables. However, as models 5-6 show, when we include controls for the hostile casualty rate and whether at least one of the candidates mentioned in an advertisement is a veteran, these other control variables also are statistically significant. The negative coefficients on hostile casualty rate in models 5-6 suggest that rising casualty levels decrease the likelihood that a military figure commits an explicitly partisan act while appearing in a campaign advertisement, and the positive coefficient on a candidate's veteran status suggests that military figures are in fact more likely to endorse or attack candidates who themselves served in the military. These findings are interesting in themselves, and further research could investigate these claims. 

Substantively, the coefficient for trust in the military of .381 (model 6) indicates that a one unit increase in percentage of the public that places a high degree or quite a lot of confidence in the military results in a 46.4 percent increase in the odds that a military figure endorses or attacks a political candidate, or addresses a topic that is outside of military expertise, while appearing in a campaign advertisement. 

Furthermore, the statistically significant and negative coefficient for Party-GOP (models 4-8) suggests that advertisements aired in support of GOP candidates are less likely to feature military figures committing explicitly partisan acts of behavior relative to the baseline group, which are ads aired in support of the Democratic party.^[The coefficient for Party-Other captures ads that are aired in support of candidates who represent parties other than the Republicans or Democrats, such as the Libertarian or Green parties.] However, I suspect that the high number of military figure endorsements and attacks that occurred in campaign ads during the 2004 election, when John Kerry ran against George W. Bush, are largely responsible for these results. The key statistical finding for the purposes of the theory offered in this dissertation, and the hypotheses offered in this chapter, is that even when these additional controls of party, hostile casualty rate, and candidate veteran status are included, the coefficient for trust in the military remains positive and significant. 

Finally, the coefficient for trust in the military remains positive in models 7 -8, which are mixed models that control only for trust in the military and polarization, but include year random effects to allow for the possibility that particular, unobserved characteristics impact a particular election differently. In both of these specifications, the coefficient for trust in the military remains positive and statistically significant. These findings reinforce the general intuition that the level of military prestige plays an important role in whether and how civilian candidates employ military figures in campaign ads, and in particular, the degree to which they wish these figures who do appear to engage in explicitly partisan behaviors.

## Predicted Probability

In order to further demonstrate the substantive impact of military prestige on the likelihood that military figures engage in acts of behavior that violate the principle of non-partisanship or non-interference, I transform the regression results from Table \@ref(tab:regression6) into predicted probabilities. Figure \@ref(fig:pred-prob-ch4-mod1-aff) plots the predicted probability of military figures either endorsing or attacking a political candidate, or addressing topics that are tangential to military expertise, while appearing in a presidential campaign advertisement. Using the results of the logistic regression in Model 6, I fix the level of affective polarization and the hostile troop casualty rate at their respective means. I hold the value of candidate-veteran at 0, denoting that the candidates portrayed in a hypothetical ad are not veterans, and hold the value of the party that the advertisement supports as the Republican Party. The vertical colored lines indicate the level of military prestige as measured by public opinion polling in certain points of time in US history [@noauthor_confidence_2020]. The color of the line indicates the party of the President who was in office at the time. 

```{r predicted-probabilities-ch4, include=FALSE, warning=FALSE, message=FALSE, error=FALSE}

# Use Model 6 (Logit) from Above. Hold values of hostile casualty rates and affective polarization at their means, and set the party as the Republican Party, and the candidate as a non-veteran. 

predictiondata_ch4 <- tibble(mil_prestige=seq(50, 85, by=1), aff_polar = mean(reg_df$aff_polar), HostCasRate=mean(reg_df$HostCasRate), party_reg_numeric=1, can_vet=0) %>% crossing(partisan_act = c(0,1)) 

predictiondata_ch4 %>% view()
augment(mod1_aff_pp, newdata = predictiondata_ch4, se_fit=TRUE)

# prediction data
predictions_ch4 <- augment(mod1_aff_pp, newdata=predictiondata_ch4, se_fit=TRUE) %>% 
  mutate(
    .prob=plogis(.fitted),
    conf.low = plogis(.fitted - (qnorm(.975) * .se.fit)),
    conf.high = plogis(.fitted + (qnorm(.975) * .se.fit)),
    pred_final_rec = case_when(
      .prob >= .5 ~ "Likely",
      .prob < .5 ~ " Not likely"
    )
  )

predictions_ch4 %>% arrange(.prob, .fitted, partisan_act, desc(.fitted)) %>% 
  view()

# now let's try some graphs
# notice that the predictions are not in (0, 1)
ggplot(predictions_ch4) +
  aes(x = .fitted) +
  geom_histogram()

# if we wanted predictions on the probability scale,
#   we predict on the log scale,
#   then apply inverse-logit (logistic CDF)
# use plogis(): cumulative probability of the logistic distribution

# demonstration: scatter log odds (x) and predicted probability (y)
# with 95% CIs
# (CIs don't look smooth in this example because .se.fit depends on X data)
predictions_ch4 %>%
  ggplot() +
  aes(x = .fitted, y = .prob) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high)) +
  geom_smooth() 

```

```{r pred-prob-ch4-mod1-aff, include=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Predicted Probability of Military Prestige/n on Civilian Use of Military Figures\n Committing Partisan Acts in Campaign Ads", fig.height=4.5, fig.width=6}
predictions_ch4%>%
  ggplot() +
  aes(x = mil_prestige, y = .prob) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high)) +
  geom_smooth() + labs(
    x="Level of Military Prestige",
    y="Predicted Probability of Military Figure Committing\n An Explicitly Partisan Act During a Campaign Advertisement"
  ) + scale_y_continuous(breaks=seq(-.1, .9, .1)) + 
  geom_vline(xintercept = 50, size=.2, colour="tomato") + 
  geom_vline(xintercept=85, size=.2, colour="tomato") + 
  geom_vline(xintercept=64, size=.2, colour="dodgerblue") + 
  geom_vline(xintercept=82, size=.2, colour="dodgerblue") +
  geom_vline(xintercept=58, size=.2, colour="tomato") +
  geom_vline(xintercept=69, size=.2, colour="dodgerblue") +
  geom_text(aes(x=50, label="1981",y=.5), colour="tomato", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=85, label="Feb 1991",y=.2), colour="tomato", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=64, label="1994",y=.5), colour="dodgerblue", angle=90, vjust=1, text=element_text(size=9)) + 
  geom_text(aes(x=82, label="2003",y=.7), colour="tomato", angle=90, vjust=-.2, text=element_text(size=9)) + 
  geom_text(aes(x=58, label="1975",y=.5), colour="dodgerblue", angle=90, vjust=-.2, text=element_text(size=9))  + 
  geom_text(aes(x=69, label="2021", y=.5), colour="dodgerblue", angle=90, vjust=1, text=element_text(size=9))
```

The vertical axis of the graph remains at or near zero for values of military prestige that are somewhat moderate, but at higher levels of military prestige, the vertical axis climbs steadily. When the level of military prestige is at 85, as was the case at the completion of the Gulf War against Iraq in February 1991, there is a greater than 50% predicted probability that a campaign advertisement ran by a civilian candidate features a military figure engaging in a partisan act during the advertisement. High levels of military prestige also existed in 2003, presumably as a result of the invasion of Iraq that began in the same year. 

The data examined in this study, which ran from 2000 - 2016, unfortunately do not allow for a richer comparison of the military's inclusion in presidential campaign ads over time. It would be extremely valuable, in my view, to be able to extend the analysis further back in time, in order to see whether at values of lower military prestige than were examined in this chapter - such as those which existed in the late 1970s and early 1980s - the trends uncovered in this chapter still hold. Similarly, this chapter did not include data from the 2020 election - an election that was contentious to say the least. Future work can, and should, extend the analysis completed in this chapter, as doing so might refine the role that military prestige plays in the calculus of civilians when determining what content to display during campaign advertisements. 

## Alternate Explanations

Before concluding the statistical analysis, it is worth thinking about potential alternate explanations that could also be responsible for driving the increase in partisan actions committed by military figures who appear prominently in campaign advertisements. What else, besides the level of military prestige, might be responsible for the patterns and trends uncovered in this chapter?

In my view, the most compelling alternate explanation is not necessarily an altogether alternate explanation, but rather a more precise definition or measurement of the components that shape military prestige. In developing the theory offered in this dissertation in chapter two, I noted that military prestige is a comprehensive and multi-faceted concept. Specifically, I argued in chapter two that the level of military prestige is shaped by several components, including institutional features of the military, which impact its relative influence within and across the whole of government, as well as societal features, to include the degree that the military is esteemed by the public. These components of military prestige are likely further comprised of other components or building blocks. For example, the degree to which the military is esteemed by the public likely rests on several factors, including the degree of felt separation between those who serve and those who do not (one of "the gaps" between the military and the public), and the perceived performance of the military in recent events, such as wars [@feaver_soldiers_2001;@schake_warriors_2016]. We can, and this dissertation has, measured the combined effects of all of these component factors of military prestige by relying on measures of public trust in the military [@noauthor_confidence_2020].

The point in revisiting the notion of military prestige being shaped by multiple subcomponents is to argue that it is _possible_ that one or more of these subcomponents - institutional influence, societal esteem, and/or performance in recent wars or conflicts - is more responsible for shaping variation in the dependent variables examined in this chapter more than the other subcomponents. However, it is difficult, if not impossible, to measure many of these subcomponents of military prestige, or at least to do so reliably. For example, did the United States lose in Iraq and in Afghanistan? The answer to this question is not crystal clear, and even if it were, the perceived answer to this question might very well have been different if we had asked the same question at the height of these conflicts, say in 2004 or in 2008.^[Future scholarship, with the benefit of hindsight and the passage of time, may render more definitively whether such wars ought to be coded dichotomously, but applying such a measurement scheme to these wars, in this chapter, would be irresponsible, in my view.] 

Although measuring the individual subcomponents of military prestige is difficult, it is nonetheless possible to at least trace the _logic_ for why one type of subcomponent might be more responsible for shaping the level of military prestige than other subcomponents, at least over the time period examined in this chapter. As Figure \@ref(fig:pred-prob-ch4-mod1-aff) shows, for example, the highest measurements of military prestige occurred right at time periods when the military had either won a war (February 1991) or was initiating a major war (2003). At the very least, Iraq had not yet become rife with sectarian violence, nor had the American public began to weary of the US' involvement there. Figure \@ref(fig:pred-prob-ch4-mod1-aff) also shows, importantly, that the lowest moments of military prestige occurred in the 1970s and early 1980s, which coincide with the era after Vietnam. The end of the Vietnam War, and the immediate years which followed, were difficult years for the military. Rife with indiscipline, the military and those who served in it returned from Vietnam not respected or revered by the American public, but ostracized. One military historian, Roger Spiller, has noted that the United States Army in particular was an "institutional wreck" at the end of the Vietnam War [@spiller_shadow_1997, 42]. It is no wonder, then, that the US military's prestige was indeed lowest at the end of the Vietnam War and in the years to follow. 

Under such logic, one might conclude that it is the military's performance in war that shapes not only the level of military prestige that arises, but importantly, also the civilian choice to leverage the military in politics, including how and the degree to which the military appears in campaign advertisements. My view, however, is that while such a view is entirely possible, it is nonetheless incomplete, for the reason that the other subcomponents of military prestige - the military's relative institutional influence and its relationship with society - are also varying at the same time, and at least partly as a result of the military's performance in war. For example, the military's relationship with society is always evolving, and the Vietnam War was fought under a compulsory (draft) system of military service. The Gulf War, on the other hand, involved an all-volunteer force, as have all conflicts since the mid-1970s. Additionally, the military's institutional influence within the government has certainly changed over time, especially as institutional reforms such as the Goldwater-Nichols Act have taken root within the military, and as the military's role in conflicts such as Iraq and Afghanistan vis-a-vis other institutions, such as the State Department, have varied. Simply put, the relationship and correlation between the military's performance in war, its relative institutional influence within government, and the military's relationship with society make a definitive judgment very difficult regarding which of these factors is most responsible for driving the level of military prestige. So we cannot rule out the possibility that one or more subcomponents of military prestige are more responsible than others in shaping military prestige and the civilian choice to feature the military in campaign ads, but neither can we definitively state that this is the case. 

It is also important to briefly comment on whether norm change is a potential alternate explanation from the perspective of civilian actors. In other words, is it possible that civilian candidates and their campaign teams have simply increasingly adopted the behavior of using the military - images, symbols, and military figures - in campaign ads? Have civilian actors simply come to see the behavior of using the military in campaign ads - and in ways that violate the central principles of civil-military relations - as less "taboo" than it used to be [@tannenwald_nuclear_1999]? 

In my view, norm change might be _somewhat_ responsible for the variation uncovered in this chapter, but nonetheless _not the primary_ reason for the variation. In other words, to the degree that norm change may play a role in shaping the decision by civilians to feature the military in campaign ads, its role, if any, should be considered small and less compelling than changes in the level of military prestige. There are two main reasons to suspect that this is a balanced and accurate view of the trends uncovered in this chapter.

First, it is true that while civilian actors are expected to adhere to the principles of civil-military relations, they do not constitute a professional body in the way that the military does, and thus, civilian actors may not be expected to adhere as strongly to the principles of non-partisanship and non-interference of the military as the military is. Thus, it is entirely possible that civilian candidates might learn from the perceived success of civilian candidates using the military in prominent ways during campaign advertisements in previous elections, and then use the military in similar ways themselves.

Second, however, the empirical record uncovered in this chapter does not reflect a story of civilian political candidates gradually learning from and implementing such behavior. For instance, Table \@ref(tab:desc-stats) in the beginning of this chapter showed that the percentage of ads in each election cycle that contain military imagery does not vary linearly over time. The same trends were also revealed in the majority of graphs featured in this chapter. If norm change is in fact the most compelling explanation of the variation in the use of the military in campaign ads by civilian actors, we would more than likely see more linear patterns of the military appearing or behaving in particular ways during campaign advertisements over time. The empirical patterns we see, however, do not seem to support such an explanation. Thus, while civilian political candidates can and likely do learn lessons from what works and does not work with respect to campaign advertising in elections, the record does not support norm change as the most compelling and underlying reason for driving patterns of civilian incorporation of the military into campaign advertising during presidential elections. 

# An Assessment of the 'Motives and Opportunities' Framework in the US Case

This section of the chapter pauses to offer an initial assessment of the motives and opportunities framework offered in chapter two of this dissertation. It is appropriate to conduct such an assessment now, having completed two major empirical analyses, each of a political behavior conducted primarily by one type of actor engaged in the conduct of civil=military relations. The previous chapter examined opinion commentary authored by military figures and published in major newspapers, and this chapter examined the incorporation of the military by civilian political candidates and their campaign teams into presidential campaign advertisements. In this section, I will briefly review what we now know, highlight what we do not yet know, and outline the way forward for the remainder of the dissertation. 

## What We Know 

In chapter two of this dissertation, I outlined four hypotheses, which are listed below for reference. 

H1: Military actors are increasingly active in a partisan manner when polarization is high relative to periods of low polarization.

H2: The overall level of political behaviors committed by military actors that challenge the fundamental principles of civil-military relations is higher in environments marked by high polarization and high military prestige relative to other types of environments. 

H3: Civilian actors increasingly leverage the military for their own political purposes and advantages when military prestige is high relative to periods of low military prestige. 

H4: The overall level of political behaviors committed by civilian actors that challenge the fundamental principles of civil-military relations is higher in environments marked by high polarization and high military prestige relative to other types of environments. 

At this point in the dissertation, I have found strong support for H1 and H3, and mixed support for H2 and H4. Strong support for H1 was found in chapter three of this dissertation, which concluded that rising political polarization is the most consistent and statistically significant explanation for the likelihood that published retired military officer commentary violates either the principle of civilian control, non-partisanship, or non-interference. Even after controlling for other variables, to include the author's rank achieved during their time on active duty, the hostile casualty rate sustained by military forces in a given year, and whether an opinion piece was authored during an election year, the coefficients for political polarization across a variety of model specifications were positive and statistically significant, lending strong support for the notion that in the US framework of civil-military relations, political polarization, especially when conceptualized as the clash of worldviews that shape and influence the fundamental values that influence, shape, and govern society, serve as a _motive_ for the military to intervene in politics.

Strong support for H3 was found in this fourth chapter of the dissertation, which, after examining the content of presidential campaign advertising, a behavior that is primarily conducted by civilian actors, concluded that rising military prestige is positively and statistically significantly associated with not only greater frequency of military imagery appearing in campaign advertisements, but greater likelihood that campaign advertisements will include military figures who commit behaviors that violate the principle of non-partisanship and/or non-interference of the military. This finding itself contributes to the scholarly understanding of how various actors are increasingly willing to overlook normative standards out of a desire to see his or her preferred candidate win office [see also @graham_democracy_2020]. 

The empirical record uncovered in this chapter also showed that military figures tend to endorse or attack political candidates, or speak out on issues that are tangential to their own military expertise, on behalf of Democratic presidential candidates moreso than on behalf of Republican candidates. This result is consistent with the "issue ownership" literature, and in particular the notion that suggests the Republican party is generally considered more skilled in matters of defense and national security and thus, more closely tied to the military, than the Democratic Party [@petrocik_issue_2003; @sides_consequences_2007]. However, this result also provides a richer contextual understanding of previous scholarly literature that found that members of the military more closely associate with the Republican party than the Democratic party [@ricks_widening_1997; @davis_attitudes_2001; @holsti_chasms_2001; see also @golby_duty_2011]. It could be that this past literature, which is 10-25 years old, is no longer valid in that service members' attitudes no longer more closely associate with the Republican party than with the Democratic party. Regardless, there is an important distinction between political attitudes and performed political behaviors, and this chapter clearly demonstrated that with respect to one form of the latter, more are occurring on behalf of the Democratic party than on behalf of the Republicans. Even after controlling for other variables, to include whether a candidate is a veteran, and the hostile casualty rate sustained by military forces in a given year, the level of military prestige proves to be a positive and statistically significant factor associated with how military images appear and the behavior of military figures in presidential campaign ads. These results provide strong support for the notion that military prestige indeed provides an _opportunity_ for civilian actors to involve the military in politics. 

Therefore, we now know that at a minimum, each actor type involved in the conduct of civil-military relations in the US seems to respond differently to changing levels of military prestige (civilians) and political polarization (military actors). When integrated, it seems, then, that there is general support for the notion of a "motives and opportunities" framework that exists in the contemporary United States for the military's involvement in politics, at least with respect to two very specific political behaviors. Such a result is in line with previous scholarly works that have used a motives and opportunities framework in general and other comparative contexts [@finer_man_1962; @taylor_politics_2003]. 

## What We Do Not Yet Know

However, there is still much that is unclear regarding the relationship between political polarization, military prestige, and the political behaviors that challenge the central principles of civil-military relations. There are at least two items of interest that remain unknown at the moment, each of which I address in turn. 

The first involves an issue of external validity with respect to the types of political behaviors each actor type can make. The trends uncovered in chapters three and four thus far each involve only one type of political behavior, and it is not clear the extent to which, if at all, the same trends apply to other types of political behaviors that each actor type can commit. While a statistical analysis of every type of political behavior that civilian and military actors can commit is not possible in this dissertation, H2 and H4 both nonetheless make predictions about the _overall level_ of political behaviors that each actor type commits when the relative levels of political polarization and military prestige are combined, and when one of the four types of domestic environments, as shown in Table \@ref(tab:poss-environment-ch4), exists. 

Each hypothesis in particular posits that each actor type will engage in more political behaviors that violate the central principles of civil-military relations when both polarization and military prestige exist at relatively high levels. As of now, the data uncovered in chapters 3 and 4 do not facilitate making such a pronouncement. Each chapter involved focusing on an important, but nonetheless only one, type of political behavior. The analysis completed thus far does not allow the researcher to claim that multiple types of political behaviors might react the same way as a result of changes in the level of political polarization and military prestige.  

A second item of interest that remains unknown at this time involves how military actors in particular respond to levels of military prestige. The main thrust of this dissertation, and this chapter in particular, examines how civilian actors respond to rising military prestige. And indeed, the notion that rising military prestige provides an _opportunity_ for civilians to involve the military in politics, rather than an _opportunity_ for the military to involve itself in politics, is the primary claim made in this dissertation. 

Still, it would be helpful if the relationship between rising military prestige and military actors could be clarified. In chapter two of this dissertation, I argued that it was not clear if military actors uniformly respond to rising military prestige the same way, as theoretically, at least two responses might be possible. The first was that military actors might respond to higher levels of military prestige by being emboldened to behave in ways that challenge or violate the principles of civil-military relations. The basis for this expectation stems from the fact that these military actors know that they are popular and respected by the public, and that they might believe that because this is true, they can, on occasion, flaunt the principles to which they otherwise must adhere to. The second possible response, however, was that military actors who know they are prestigious do not want to lose the potential public influence that they have, and therefore would not act or behave in such a way so as to jeopardize the strong rapport that they have a public. One response to higher military prestige expects a bolder military that selectively violates the principles of civil-military relations, while one predicts a military that more tightly adheres to them. Further clarifying the direction of this possible relationship, even if a uniform expectation is not possible, would solidify that the relationship between civilian actors and military prestige truly indeed is the right relationship to emphasize. 

## The Way Forward

This dissertation now pivots to explore historical case studies, which will help better inform the impact of polarization and military prestige on the overall level of political behaviors committed by civilian and military actors, and clarify the relationship between rising military prestige and the political behaviors committed by military actors in particular. The first case study, undertaken in the next chapter, explores political behaviors committed by civilian and military actors involving the military in the period from immediately after the US Civil War through the middle of the Gilded Age. This historical era runs approximately from 1865 - 1885, and involves the evolution of an era marked by high political polarization and high military prestige to one characterized by high polarization and low military prestige. 

The second case study, undertaken immediately after the first, examines the political behaviors committed by civilian and military actors from the end of the Vietnam War to the period immediately after the Gulf War. This historical era runs from approximately 1973 - 1992, and spans a period in which political polarization and military prestige evolved from levels in which both were relatively low, to one in which political polarization was still low, but military prestige was high. 

In each of these case studies, military prestige varies, while the level of polarization remains constant within each. This is a deliberate decision, largely stemming from the fact that the empirical analysis thus far completed has involved somewhat greater variation in the level of polarization than in the level of military prestige. However, each of the case studies involves a different relative level of polarization. Thus, together, the case studies will enable a meaningful comparison, both within and across cases, to better understand how polarization and military prestige impact the political behaviors committed by civilian and military actors.

\singlespace

\pagebreak

# Appendix to Chapter Four

\doublespace

This chapter Appendix roughly aligns with the chronological order of the chapter.   

## Note on the Campaign Media Analysis Group and its Role in the Data

WiscAD and WMP gather the television advertisements with the assistance of a commercial company, the Campaign Media Analysis Group (CMAG). CMAG provides several pieces of information when releasing the data, such as the media market in which a particular advertisement aired, the estimated cost of running a particular advertisement, the time of day the advertisement was run, and other variables. A group of scholars and graduate students affiliated with WiscAD and WMP then conduct additional coding on the advertisements. This coding varies to some extent by election cycle, but otherwise remains markedly similar. The additional variables that are coded for in each election cycle include items such as the specific issues that an advertisement mentions or addresses during the advertisement, whether a candidate is featured during the advertisement, and whether a given advertisement takes on a negative tone, and more. 

Cleaning the data from five separate data sets was a considerable project. I had to merge several variables and implement a common naming scheme across the data in order to enable a meaningful comparison of the data over time. The .rmd file that contains the cleaning instructions is available for review. 

## Robustness and Endogeneity Tests

In this section of the Appendix, I conduct a series of robustness checks and tests for endogeneity using regression analysis. I present the results and briefly discuss the results and implications of each in turn. 

### Robustness Tests

The dependent variable used in the chapter's main regression table was the likelihood that a military figure appeared in an campaign advertisement and violated any of the central principles of civil-military relations by conducting one of three specific types of behaviors: endorsing a candidate for office, attacking a candidate running for office, or addressing an outside issue. the first two behaviors are different manifestations of violating the principle of non-partisanship, while the third is a manifestation of violating the principle of non-interference of the military. 

In the regression table to follow, I make each of these three behaviors a separate dependent variable in order to test the finding that the level of military prestige drives how civilians implement military figures in campaign ads, and specifically, whether we can be precise to predict that changes in the level of military prestige are robust to particular types of behaviors, rather than a conglomerate of the three behaviors, as used in the main regression table. I also include several OLS models where the dependent variable is the proportion of ads in each election cycle that contains military imagery altogether, of any type. 

The results of these regressions are displayed in Table \@ref(tab:regression7).

\singlespace

```{r Regression Robustness Tests_ch4, include=TRUE, echo=FALSE, results="asis", warning=FALSE, message=FALSE}
stargazer(mod2_house, mod2_senate, mod3_senate, mod3_aff, mod4_house, mod4_aff, mod5_mil_img_cycle_house, mod5_mil_img_cycle_sen, mod5_mil_img_cycle_aff,
          type="latex",
          label="tab:regression7",
          title="Robustness Checks (Violations of Each Principle Separately and Proportion of Ads Featuring Military Imagery in Each Election Cycle)", 
          covariate.labels = c('Trust in Military', 'House Polarization', 'Senate Polarization', 'Aff. Polarization', 'Hostile Casualty Rate', 'Candidate-Veteran'),
          dep.var.labels = c('Endorse', 'Attack', 'Address Outside Issue', 'Proportion of Ads with Military Imagery'),
          model.names=TRUE,
          header=FALSE,
          column.sep.width = "1pt",
          align=TRUE,
          font.size="small",
          no.space=TRUE,
          omit.stat=c("f", "ser"),
          float.env="sidewaystable"
          
          )
```
\doublespace

There are several important takeaways from Table \@ref(tab:regression7). The first is that the finding from the main regression table regarding military prestige is robust to two of the three behavioral specifications for violating the central principles of civil-military relations. Indeed, as models 1-4 show, changes in the level of trust in the military are positive and statistically significant when the dependent variable is instances of endorsing a political candidate (models 1-2) as well as instances of attacking a political candidate (models 3-4). However, the results are not robust to a specification of the dependent variable as addressing issues that are tangential to military expertise, as shown in models 5-6 of Table \@ref(tab:regression7). Thus, the results predict that as the military becomes increasingly trusted by the public, civilian political candidates are more likely to use military figures in campaign ads to make endorsements of political candidates, and to render attacks against political opponents, but we cannot state that rising levels of military trust will also definitively result in increasing instances of military figures addressing topics that fall outside of military expertise. 

### Checking for Endogeneity

In this section of the Appendix, I again check for endogeneity, and in particular, the relationship between military prestige and instances of military figures appearing in civilian campaign advertisements who either endorse or attack political candidates, or address a topic that is tangential to or outside of military expertise. Table \@ref(tab:regression7) presents the result of several OLS regression models in which the dependent variable is military prestige, as measured by the level of public trust and confidence in the military. I again include several control variables, including different measures of polarization, as well as an interaction term between the level of polarization and instances in which military figures appear in campaign ads and violate one or more of the central standards of civil-military relations. 

\singlespace
```{r endogeneity checks_ch4, include=TRUE, echo=FALSE, results="asis", warning=FALSE, message=FALSE}

stargazer(mod10_ch4_endog_house, mod10_ch4_endog_sen, mod10_ch4_endog_house_int_simple, mod10_ch4_endog_sen_int_simple, mod10_ch4_endog_house_int, mod10_ch4_endog_sen_int,
          type="latex",
          label="tab:regression7",
          title="Regression Results: Endogeneity Checks (Regress Military Prestige on Polarization and an Instance of a Military Figure Violating a Central Principle during a Campaign Ad)", 
          covariate.labels = c('Violation', 'House Polarization', 'Senate Polarization', 'Hostile Casualty Rate', 'Candidate-Veteran', 'Violation*House Polarization', 'Violation*Senate Polarization'),
          dep.var.labels = c('Military Prestige'),
          model.names=TRUE,
          header=FALSE,
          column.sep.width = "1pt",
          align=TRUE,
          font.size="small",
          no.space=TRUE,
          omit.stat=c("f", "ser")
          )
```
\doublespace

The main concern again in checking for endogeneity deals with the directionality of the system of variables involved. In the particular setting of this chapter, then, we are attempting to determine whether there is strong evidence that instances of a military figure violating a central principle drives the level of military prestige, as opposed to what I have claimed - that military prestige drives the civilian choice or willingness of civilian candidates to use military figures who behave in ways that violate or challenge the principles of civil-military relations. 

The results of the regressions in Table \@ref(tab:regression7) suggest that there is low-moderate endogeneity in the data, but ultimately, not to a level that calls into question the results obtained in the main regression table in this chapter. From left to right, models 1-6 first involve no interaction term between polarization and an instance of a military figure violating one of the principles of civil-military relations (models 1-2), then include models in which an interaction term is included, but controls for the hostile casualty rate and whether are a candidate is a veteran are removed (models 3-4), and then finally include models in which these controls are re-instated along with appropriate interaction terms (models 5-6). 

The coefficient on violation is positive and highly statistically significant in models 1-4, which could prove problematic by itself, but when we reinstate other control variables in models 5-6, the statistical significance for the coefficient on violation falls away. In addition, in models 5-6, the coefficient for the interaction term between violation*polarization (house and senate, respectively) is statistically insignificant and slightly negative. 

I run a series of additional checks that include a dummy variable for the 2004 Election, as that was an election held at a time at which the military was highly prestigious, and when casualties were relatively high, when several of the candidates were veterans (to include George W. Bush and John Kerry, the main candidates), and an election in which the Swiftboat Veterans and other military figures appeared rather frequently and engaged in frequent partisan behaviors. I am interested in determining whether the coefficient for the dummy on 2004, in other words, picks up a lot of the magnitude and statistical significance that was present on the coefficient for violation in the previous regression table. 

The results are displayed in Table \@ref(tab:regression8). 

\singlespace
```{r endogeneity checks_ch4_part2, include=TRUE, echo=FALSE, results="asis", warning=FALSE, message=FALSE}

stargazer(mod10_ch4_endog_house_dummy, mod10_ch4_endog_sen_dummy, mod10_ch4_endog_house_int_simple_dummy, mod10_ch4_endog_sen_int_simple_dummy, mod10_ch4_endog_house_int_dummy, mod10_ch4_endog_sen_int_dummy,
          type="latex",
          label="tab:regression8",
          title="Regression Results: Endogeneity Checks with a 2004 dummy (Regress Military Prestige on Polarization and an Instance of a Military Figure Violating a Central Principle during a Campaign Ad)", 
          covariate.labels = c('Violation', 'House Polarization', 'Senate Polarization', 'Hostile Casualty Rate', 'Candidate-Veteran', '2004 Election', 'Violation*House Polarization', 'Violation*Senate Polarization'),
          dep.var.labels = c('Military Prestige'),
          model.names=TRUE,
          header=FALSE,
          column.sep.width = "1pt",
          align=TRUE,
          font.size="small",
          no.space=TRUE,
          omit.stat=c("f", "ser")
          )
```
\doublespace

The key takeaway from Table \@ref(tab:regression8) is that including the dummy for the 2004 Election, as predicted, absorbs quite a bit of the magnitude and statistical significance, again stemming from the correlation and multicollinearity that exists between the dependent variable and several of the independent variables in Table \@ref(tab:regression7). Most importantly, inclusion of the 2004 Election dummy, as shown in models 5 - 6, results in statistically insignificant coefficients on instances of a violation of one of the central principles and the interaction term between violation and polarization. 

## Additional Figures 

This section of the Appendix features several additional graphs that help descriptively present the data.  

Figure \@ref(fig:def-theme) displays for each election cycle the number of national security-themed ads, the number of ads that contain some form of military imagery, the number of non-national security themed ads, and the total number of advertisements. Although these statistics do not reflect challenges to the central principles of civil-military relations, they nonetheless provide a useful starting point regarding the general prominence of military and defense related issues during the five presidential elections from 2000 - 2016.

```{r def-theme, include=TRUE, echo=FALSE, fig.cap="Presidential Campaign Advertisements Featuring Military Imagery, Separated by Ads that are Entirely Focused on National Security Topics", fig.height=4, fig.width=7}

plots_data <- adj_df %>% select(year, Yrcount, sum_mil_img_cycle, sum_no_mil_img_cycle, sum_def_vet_ad_theme, sum_non_def_vet_ad_theme, sum_def_theme_and_mil_img, sum_not_def_theme_but_mil_img) %>% view()

# now send this data frame to the pivot_longer function to make it more easily graphable
# use the distinct () function to keep only pertinent values
 graph_plot1_data <- plots_data %>% select(year, Yrcount, sum_def_vet_ad_theme, sum_non_def_vet_ad_theme, sum_mil_img_cycle) %>% pivot_longer(Yrcount:sum_mil_img_cycle, names_to = "statistic", values_to = "Number") %>% distinct () %>% view ()

plot1 <- ggplot(data=graph_plot1_data, aes(fill=statistic, x=year, y=Number)) + geom_bar(position="dodge", stat="identity") + labs(x="Election", y="Number of Presidential Campaign Ads") + scale_fill_discrete(name="Value",
                         breaks=c("sum_def_vet_ad_theme", "sum_mil_img_cycle", "sum_non_def_vet_ad_theme", "Yrcount"),
                         labels=c("National Security-Themed Ads", "Ads that Featured Some Form of Military Imagery", "Non-National Security Themed Ads", "Total Number of Ads")) + theme(legend.position="bottom")

plot1 

```

## To What Extent Is the Military an Enduring Campaign Issue?  

It is worth examining the degree to which the military stands out in a general sense in campaign advertisements over time. In other words, while exogenous events such as wars and international crises are likely to drive changes in the agendas of political candidates, and thus, changes in national security-themed campaign advertisements that feature the military, there might also be some baseline level of attention that the military receives from political candidates that is separate from wars or prominent international crises in which the military plays a noteworthy role.

Figure \@ref(fig:mil-img) below displays the number of non-national security related advertisements in each election cycle that feature some form of military imagery. In these campaign advertisements, therefore, there were themes presented other than those that focus entirely on issues of national security, defense, or veterans issues, but military imagery nonetheless appeared in the advertisement. Of all of the various non-security related issues and themes that are raised during a campaign advertisement, the military nonetheless made the cut. 

```{r mil-img, warning=FALSE, error=FALSE, include=TRUE, echo=FALSE, fig.height=4, fig.width=6, fig.cap= "Non-National Security Themed Campaign Ads that Featured Military Imagery"}

graph_plot2_data <- plots_data %>% select(year, sum_not_def_theme_but_mil_img, sum_non_def_vet_ad_theme) %>% pivot_longer(sum_not_def_theme_but_mil_img:sum_non_def_vet_ad_theme, names_to = "statistic", values_to = "Number") %>% distinct () %>% view ()

plot2 <- ggplot(data=graph_plot2_data, aes(fill=statistic, x=year, y=Number)) + geom_bar(position="dodge", stat="identity") + labs(x="Election", y="Number of Presidential Campaign Ads") + scale_fill_discrete(name="Value",
                         breaks=c("sum_non_def_vet_ad_theme", "sum_not_def_theme_but_mil_img"),
                         labels=c("Non-National Security Themed Ads", "Non-National Security Themed Ads That Contained Military Imagery")) + theme(legend.position="bottom")

plot2

```

Note the relative consistency in the fact that from 2000 - 2016, military imagery featured in a healthy proportion of campaign advertisements that were not focused on issues of national security, defense, or veterans issues. These results suggest that at least for the past 20 years, American political candidates have conveyed to the public their desire to support and strengthen the military in a very general sense. Although the data examined do not extend far back enough in time to prove the following conclusion, a trend of political candidates appealing to the public by acknowledging and featuring the military - at least in a general way during campaign advertisements - reflects the relatively high level of military prestige that the US military has commanded since at least 2000.

## Data Acknowledgement and Disclaimers 

## 2000

The data were obtained from a joint project of the Brennan Center for Justice at New York University School of Law and Professor Kenneth Goldstein of the University of Wisconsin-Madison, and includes media tracking data from the Campaign Media Analysis Group in Washington, D.C. The Brennan Center-Wisconsin project was sponsored by a grant from The Pew Charitable Trusts. The opinions expressed in this article are those of the author(s) and do not necessarily reflect the views of the Brennan Center, Professor Goldstein, or The Pew Charitable Trusts.

## 2004

The data were obtained from a project of the University of Wisconsin Advertising Project includes media tracking data from TNSMI/Campaign Media Analysis Group in Washington, D.C. The University of Wisconsin Advertising Project was sponsored by a grant from The Pew Charitable Trusts. The opinions expressed in this article are those of the author(s) and do not necessarily reflect the views of the University of Wisconsin Advertising Project or The Pew Charitable Trusts.

## 2008

The data were obtained from a project of the University of Wisconsin Advertising Project includes media tracking data from TNSMI/Campaign Media Analysis Group in Washington, D.C. The opinions expressed in this article are those of the author(s) and do not necessarily reflect the views of the University of Wisconsin Advertising Project.

## 2012

The data were obtained from the Wesleyan Media Project, a collaboration between Wesleyan University, Bowdoin College, and Washington State University, and includes media tracking data from Kantar/Campaign Media Analysis Group in Washington, D.C. The Wesleyan Media Project was sponsored in 2012 by grants from The John S. and James L. Knight Foundation and The John D. and Catherine T. MacArthur Foundation. The opinions expressed in this article are those of the author(s) and do not necessarily reflect the views of the Wesleyan Media Project, Knight Foundation, MacArthur Foundation or any of its affiliates.

## 2016

The data were obtained from the Wesleyan Media Project, a collaboration between Wesleyan University, Bowdoin College, and Washington State University, and includes media tracking data from Kantar/Campaign Media Analysis Group in Washington, D.C. The Wesleyan Media Project was sponsored in 2016 by a grant from The John S. and James L. Knight Foundation. The opinions expressed in this article are those of the author(s) and do not necessarily reflect the views of the Wesleyan Media Project, Knight Foundation, or any of its affiliates.

## Other Acknowledgements

All regression tables used in chapter three were produced with the help of the Stargazer package, designed by Marek Hlavac [@hlavac_stargazer_2018].

\pagebreak
\singlespace
\newpage



