---
title: "05-campaign_ads"
author: "Pete Erickson"
date: "2/15/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, include = FALSE, echo = FALSE}
library("here")       # file paths
library("tidyverse")  # workhorse package
library("tidylog")
library("kableExtra")
library("knitr")
library("ggdag")
library("dagitty")
library("gridExtra")
library("broom")
library("dplyr")
library("reshape2")
library("stargazer")
library("rstatix")
library("lme4")
library("brglm")
library("bucky")
library("haven")
library("writexl")
```

# Notes for Data Cleaning

In this document, I have essentially cleaned 5 x data sets containing information about various presidential campaign television advertisements. As each data set is unique, I have taken care to standardize the data sets and prepare them in such a way so as to facilitate future coding. In the paragraphs to follow, I first outline the procedures taken before embarking on these coding procedures and tasks.  

I desired to obtain two main outputs from cleaning this data. The first was a list of observations, along with the associated relevant variables, that were "likely" to feature what I was looking for - in my case, some form of US military imagery (personnel, uniforms, veterans, scenes of combat, etc.). The second was the opposite of the first - a list of observations that were "unlikely" to feature images of the military. Together, these two lists shoud consist of all of the data, and be mutually exclusive, i.e., an observation that is on one list is not on the other.

# Procedures Used to Clean and Organize the Data

Data from the 2016 and 2012 data sets came from the Wesleyan Media Project (WMP), whereas the data from the 2008, 2004, and 2000 data sets came from the Wisconsin Advertising Project (WiscAd). Each data set is structured somewhat differently. One can detect a fairly straightforward improvement each subsequent election, i.e. the 2000 data is the "roughest," and the data improves each year.  

In the most basic sense, the main procedure used to separate the "likely" from "unlikely" observations involved identifying and using existing variables which I felt would be most correlated to the military in some sense. These existing variables varied each election cycle, but generally dealt with the issue of what topic did a particular observation address? Most of the data sets coded observations for the topic(s) each advertisement addressed, and some data sets included additional controls that, by my reasoning, provided even further justification for including / excluding certain observations. 

The obvious topical discriminators were any advertisements that were coded for discussing/addressing defense, military, or veterans-related issues, as I felt that these were the most likely to contain ads that included military imagery and symbols. Given the Iraq and Afghanistan wars that occurred in the post 9/11 era, I also included in the "likely" category any explicit variables that identified Iraq, Afghanistan, terrorism, or war on terror, or similar words (this was not possible for the 2000 election, as 9/11 had not yet occurred). 

I then included any other foreign policy variables that could be reasonably associated with the US military. Admittedly, this involved some judgement on my part. I included those observations that involved explicit mentions of issues such as China, Iran, or Israel, mainly because I felt that these would, on average, be more likely to be associated with US military imagery than other electoral issues, such as elemantary education and natural resources. 

Finally, but importantly, there were, on occasion, other variables that I could include that also helped me separate the likely from the unlikely observations. One data set (2004), for instance, had a variable for "actor credibility" (the variable was 'ECFCRED'), which sought to determine the source of credibility in an actor that appeared in an advertisement. Sure enough, one possible value that this variable could take was '13,' or "veteran/military" personnel.  Therefore, for the 2004 data set, I included this particular value and variable to help narrow the list. 

# Results as of 9 August 2021

In this paragraph, I briefly highlight some notable points about the data, particularly as it relates to the separation of likely versus unlikely ads, as well as missing data.  

* Likely - 895 total Ads (12 observations are missing, however)
* Unlikely - 3143 Ads (however, screening proceeds on 2970 ads, as the year 2000 includes 173 ads that are entirely coded as "N/A" for topics)
* Total - 4038 Ads

## Important Files and Filenames

df_comb_init_cads_edits - the file I used consisting of the "likely" ads, which I then watched and conducted further coding on

df_comb_init_cads_unlikely_edits_screened - the file I used to screen what consisted of the "unlikely" ads

ch6_df_comb_reg_data - the file I used that contained all of the ads (need to add in the "N/A" ads from 2000) to conduct regression.  Note: I had to compile this particular data set using a series of variables from the two prior files. 

(not yet used) ch6_Master_Imagery - this is a file that I have created, but not yet used, to conduct analysis on all of the ads for purposes of imagery types only. 

*** note - I will code by hand on the file "reg_data" any ads that include a prominent military person; there is at least one add where this must be done (2008 PRES/MOVEON OBAMACAN), and there might be another (2006 PRES/ELSUPERPAC CARELIZ SP) ***


## Remaining Work

* there are several ads that require a spanish speaker/translator...several on the init_cads_edits list, and at least one on the unlikely_edits_screened file.  
* need a decision from my committee about whether I should just go ahead and code ALL of the ads and be done with it. 

```{r data2016, include=FALSE, echo=FALSE, warning=FALSE}

# This section of code "cleans" the data used from the 2016 data set.  

## (1) To begin with, these are the issue numbers that help to 
# narrow down which ad themes may feature the military.
# Issue60-Military, 
# Issue 61 - Foreign Policy (generic)
# Issue 62-Veterans, 
# Issue 64 - Nuclear Proliferation
# Issue 65 - China
# Issue 66 - Middle East
# Issue 67-Afghanistan, 
# Issue 68-September 11th
# Issue69-Terrorism, 
# Issue70-Iraq
# Issue71-Israel
# Issue 72 - Iran
# Issue 73-ISIL or ISIS 
# Issue 74 - Syria
# 14 total issue areas


# first read in the data and then summarize it
df <- read_dta(here("data", "wmp-pres-2016-v1.1.dta"))
summary(df)



#Get an initial idea for how many ads there are are, which ones to watch, split. 
# Coding to find the Total # of Ads: 914
dfsplit  <- df %>% select(creative, issue60, issue61, issue62, issue64, issue65, issue66, issue67, issue68, issue69, issue70, issue71, issue72, issue73, issue74) %>% distinct() %>% view()

#Ads to watch : 252
dfsplit_watch <- dfsplit %>% filter(issue60==1 | issue61==1 |issue62==1 | issue64==1 | issue65==1 |issue66==1 | issue67==1 | issue68==1 | issue69==1 | issue70==1 | issue71==1 | issue72==1 | issue73==1 | issue74==1) %>% view()

# Ads to not watch : 659
dfsplit_dontwatch <- dfsplit %>% filter(issue60 !=1 & issue61 !=1 & issue62 !=1 & issue64 !=1 & issue65 !=1 & issue66 !=1 & issue67 !=1 & issue68 !=1 & issue69 !=1 & issue70 !=1 & issue71 != 1 & issue72 !=1 & issue73 !=1 & issue74 !=1) %>% view()

# Ads that are coded as NA for all of these issue areas: 3
dfsplit_NA <- dfsplit %>% filter(is.na(issue60) & is.na(issue61) & is.na(issue62) & is.na(issue64) & is.na(issue65) & is.na(issue66) & is.na(issue67) & is.na(issue68) & is.na(issue69) & is.na(issue70) & is.na(issue71) & is.na(issue72) & is.na(issue73) & is.na(issue74)) %>% view()



# Now create a df by filtering and then grouping by the ID of the advertisement (variable "creative")
robustdf <- df %>% filter(issue60==1 | issue61==1 |issue62==1 | issue64==1 | issue65==1 |issue66==1 | issue67==1 | issue68==1 | issue69==1 | issue70==1 | issue71==1 | issue72==1 | issue73==1 | issue74==1) %>% group_by(issue60, issue61, issue62, issue64, issue65, issue66, issue67, issue68, issue69, issue70, issue71, issue72, issue73, issue74, creative) %>% count(creative) %>% view()


# begin forming a clean data set for 2016 that can be merged with other data across the years
# step 1: filter the data using the filter command by issue area 
# step 2: select the appropriate variables to include in the data using the 'select' verb
# step 3: group the results using the 'group_by' verb

clean_df_2016 <- df %>% 
  filter(issue60==1 | issue61==1 |issue62==1 | issue64==1 | issue65==1 |issue66==1 | issue67==1 | issue68==1 | issue69==1 | issue70==1 | issue71==1 | issue72==1 | issue73==1 | issue74==1) %>% 
  select(creative, l, election, tonecmag, sponsorcmag, party, sponsor, f_mention, f_picture, f_narrate, o_mention, o_picture, ad_tone, prty_mn, issue60, issue61, issue62, issue64, issue65, issue66, issue67, issue68, issue69, issue70, issue71, issue72, issue73, issue74) %>%
  group_by(creative) %>% view()


# initial cut at a data set to join to other data sets.
# I've already selected which issue areas matter, so I can drop those here and 
# because these issue areas will change over time. 
# Next, I can group by the variables I am interested in. 
# Then, I create a variable for count, referring to the number of times an ad was run
# in a particular election cycle, and an ad for the election cycle more generally. 
# I then use the "distinct" function to keep only the unique rows that matter, which is key. 
# Then I rearrange the data and display it according to descending "count" order. 

initial_2016_join <- clean_df_2016 %>% 
  select(-issue60, -issue61,-issue62, -issue64, -issue65, -issue66, -issue67, -issue68, -issue69, -issue70, -issue71, -issue72, -issue73, -issue74) %>% 
  group_by(creative, l, party, sponsor) %>%
  mutate(
    count = n(), 
    year=2016
  ) %>% 
distinct() %>% 
  select(year, count, creative:prty_mn) %>% 
  arrange(desc(count)) %>% 
  view()

df_2016 <- initial_2016_join %>% select (-election, -tonecmag, -sponsorcmag) %>% view()
# total of 252 ads to watch


##########Unlikely Ads ###########

# the following paragraph is one line of code that enabled me to gather the vital variables for when I developed the master imagery list
# it handles the variables that I need to do a meaningul comparison for the ads that are unlikely to feature military imagery
unprob_df_2016 <- df %>% 
  filter((issue60 !=1 & issue61 !=1 & issue62 !=1 & issue64 !=1 & issue65 !=1 & issue66 !=1 & issue67 !=1 & issue68 !=1 & issue69 !=1 & issue70 !=1 & issue71 != 1 & issue72 !=1 & issue73 !=1 & issue74 !=1) | (is.na(issue60) & is.na(issue61) & is.na(issue62) & is.na(issue64) & is.na(issue65) & is.na(issue66) & is.na(issue67) & is.na(issue68) & is.na(issue69) & is.na(issue70) & is.na(issue71) & is.na(issue72) & is.na(issue73) & is.na(issue74))) %>% 
  select(creative, l, election, tonecmag, sponsorcmag, party, sponsor, f_mention, f_picture, f_narrate, o_mention, o_picture, ad_tone, prty_mn, issue60, issue61, issue62, issue64, issue65, issue66, issue67, issue68, issue69, issue70, issue71, issue72, issue73, issue74) %>% 
  group_by(creative, l) %>% 
  mutate(
    count = n(), 
    year=2016
  ) %>% 
  distinct() %>% 
  select(year, count, creative, l, party, sponsor, f_mention, f_picture, f_narrate, o_mention, o_picture, ad_tone, prty_mn) %>% 
  arrange(desc(count)) %>% distinct() %>% 
  view()

## this list above includes 662 total ads, including three that were "N/A" on all pertinent issue areas

# below is the list that is "unprobable" to contain the ads I think I need to watch
clean_df_2016_unprobably <- df %>% 
  filter(issue60 !=1 & issue61 !=1 & issue62 !=1 & issue64 !=1 & issue65 !=1 & issue66 !=1 & issue67 !=1 & issue68 !=1 & issue69 !=1 & issue70 !=1 & issue71 != 1 & issue72 !=1 & issue73 !=1 & issue74 !=1) %>% 
  select(creative, l, party, issue60, issue61, issue62, issue64, issue65, issue66, issue67, issue68, issue69, issue70, issue71, issue72, issue73, issue74) %>%
  group_by(creative)

initial_2016_join_unprobable <- clean_df_2016_unprobably %>% 
  group_by(creative, l) %>%
  mutate(
    count = n(), 
    year=2016
  ) %>% 
distinct() %>% 
  select(year, count, creative, l, party, issue60, issue61, issue62, issue64, issue65, issue66, issue67, issue68, issue69, issue70, issue71, issue72, issue73, issue74) %>% 
  arrange(desc(count)) %>% 
  view()


# 659 ads to NOT watch

df_2016_notlikely <- initial_2016_join_unprobable %>% select(-issue60, -issue61, -issue62, -issue64, -issue65, -issue66, -issue67, -issue68, -issue69, -issue70, -issue71, -issue72, -issue73, -issue74) %>% view()

#total of 659 not to watch

```


```{r data2012, include=FALSE, echo=FALSE, warning=FALSE}
# Issue60-Military, 
# Issue61-Foreign Policy (generic)
# Issue 62-Veterans
# Issue 64 - nuclear proliferation
# Issue 65-China
# ISsue 66 - Middle East
# Issue 67-Afghanistan, 
# ISsue 68-September 11
# Issue69-Terrorism 
# Issue70- Iraq
# Issue 71 - Israel
# Issue 72 - Iran
# 12 issue areas


# read in the data and then summarize it
df2012 <- read_dta(here("data", "wmp-pres-2012-v1.2_compress.dta"))
summary(df2012)



#Get an initial idea for how many ads there are are, which ones to watch, split. 
df_2012_split  <- df2012 %>% select(creative, issue60, issue61, issue62, issue64, issue65, issue66, issue67, issue68, issue69, issue70, issue71, issue72) %>% distinct() %>% view()
# Coding to find the Total # of Ads: 735

df_2012_split_watch <- df_2012_split %>% filter(issue60==1 | issue61==1 |issue62==1 | issue64==1 | issue65==1 |issue66==1 | issue67==1 | issue68==1 | issue69==1 | issue70==1 | issue71==1 | issue72==1) %>% view()
#Ads to watch : 102

df_2012_split_dontwatch <- df_2012_split %>% filter(issue60 !=1 & issue61 !=1 & issue62 !=1 & issue64 !=1 & issue65 !=1 & issue66 !=1 & issue67 !=1 & issue68 !=1 & issue69 !=1 & issue70 !=1 & issue71 != 1 & issue72 !=1) %>% view()
# Ads to not watch : 622

df_2012_split_NA <- df_2012_split %>% filter(is.na(issue60) & is.na(issue61) & is.na(issue62) & is.na(issue64) & is.na(issue65) & is.na(issue66) & is.na(issue67) & is.na(issue68) & is.na(issue69) & is.na(issue70) & is.na(issue71) & is.na(issue72)) %>% view()
#Ads that are N/A or not coded: 11



# create a robust df by filtering and then grouping by the ID of the advertisement (variable "creative")
robustdf2012 <- df2012 %>% filter(issue60==1 | issue61==1 | issue62==1 | issue64==1 | issue65==1 | issue66==1 | issue67==1 | issue68==1 | issue69==1 | issue70==1 | issue71==1 | issue72==1) %>% group_by(issue60, issue61, issue62, issue64, issue65, issue66, issue67, issue68, issue69, issue70, issue71, issue72, creative) %>% count(creative) %>% view()
# count - 102


# begin forming a clean data set for 2012 that can be merged with other data across the years
# step 1: filter the data using the filter command by appropriate issue area 
# step 2: select the appropriate variables to include in the data using the 'select' verb
# step 3: group the results using the 'group_by' verb
clean_df_2012 <- df2012 %>% 
  filter(issue60==1 | issue61==1 | issue62==1 | issue64==1 | issue65==1 | issue66==1 | issue67==1 | issue68==1 | issue69==1 | issue70==1 | issue71==1 | issue72==1) %>% 
  select(creative, l, election, tonecmag, sponsorcmag, affiliation, sponsorwmp, f_mention, f_picture, f_narrate, o_mention, o_picture, ad_tone, prty_mn, issue60, issue61, issue62, issue64, issue65, issue66, issue67, issue68, issue69, issue70, issue71, issue72) %>% 
  group_by(creative) %>% view()


# initial cut at a data set to join to other data sets.
# I've already selected which issue areas matter, so I can drop those here and 
# because these issue areas will change over time. 
# Next, I can group by the variables I am interested in. 
# Then, I create a variable for count, referring to the number of times an ad was run
# in a particular election cycle, and an ad for the election cycle more generally. 
# I then use the "distinct" function to keep only the unique rows that matter, which is key. 
# Then I rearrange the data and display it according to descending "count" order. 

initial_2012_join <- clean_df_2012 %>% 
  select(-issue60, -issue61, -issue62, -issue64, -issue65, -issue66, -issue67, -issue68, -issue69, -issue70, -issue71, -issue72) %>% 
  group_by(creative, l, affiliation, sponsorwmp, prty_mn) %>%
  mutate(
    count = n(), 
    year=2012
  ) %>% 
distinct() %>% 
  select(year, count, creative:prty_mn) %>% 
  arrange(desc(count)) %>% 
  view()

# 107 ads to watch

df_2012 <- initial_2012_join %>% 
  select(-election, -tonecmag, -sponsorcmag) %>% 
  rename(party=affiliation, sponsor=sponsorwmp) %>% view()




##########################Not Likely############

# the following paragraph is one line of code that enabled me to gather the vital variables for when I developed the master imagery list
# it handles the variables that I need to do a meaningul comparison for the ads that are unlikely to feature military imagery
unprob_df_2012 <- df2012 %>% 
  filter((issue60 !=1 & issue61 !=1 & issue62 !=1 & issue64 !=1 & issue65 !=1 & issue66 !=1 & issue67 !=1 & issue68 !=1 & issue69 !=1 & issue70 !=1 & issue71 != 1 & issue72 !=1) | (is.na(issue60) & is.na(issue61) & is.na(issue62) & is.na(issue64) & is.na(issue65) & is.na(issue66) & is.na(issue67) & is.na(issue68) & is.na(issue69) & is.na(issue70) & is.na(issue71) & is.na(issue72))) %>% 
  select(creative, l, election, tonecmag, sponsorcmag, affiliation, sponsorwmp, f_mention, f_picture, f_narrate, o_mention, o_picture, ad_tone, prty_mn, issue60, issue61, issue62, issue64, issue65, issue66, issue67, issue68, issue69, issue70, issue71, issue72) %>% 
  group_by(creative, l) %>% 
  mutate(
    count = n(), 
    year=2012
  ) %>% 
  distinct() %>% 
  select(year, count, creative, l, affiliation, sponsorwmp, f_mention, f_picture, f_narrate, o_mention, o_picture, ad_tone, prty_mn) %>% 
  arrange(desc(count)) %>% rename(party=affiliation, sponsor=sponsorwmp) %>% distinct() %>% 
  view()


# below is the list that is "unprobable" to contain the ads I think I need to watch
clean_df_2012_unprobably <- df2012 %>% 
  filter(issue60 !=1 & issue61 !=1 & issue62 !=1 & issue64 !=1 & issue65 !=1 & issue66 !=1 & issue67 !=1 & issue68 !=1 & issue69 !=1 & issue70 !=1 & issue71 != 1 & issue72 !=1) %>% 
  select(creative, l, affiliation, issue60, issue61, issue62, issue64, issue65, issue66, issue67, issue68, issue69, issue70, issue71, issue72) %>%
  group_by(creative)

initial_2012_join_unprobable <- clean_df_2012_unprobably %>% 
  group_by(creative, l) %>%
  mutate(
    count = n(), 
    year=2012
  ) %>% 
distinct() %>% 
  select(year, count, creative, l, affiliation, issue60, issue61, issue62, issue64, issue65, issue66, issue67, issue68, issue69, issue70, issue71, issue72) %>% 
  arrange(desc(count)) %>% 
  rename(party=affiliation) %>% 
  view()

# 672 ads to NOT watch

df_2012_notlikely <- initial_2012_join_unprobable %>% select(-issue60, -issue61, -issue62, -issue64, -issue65, -issue66, -issue67, -issue68, -issue69, -issue70, -issue71, -issue72) %>% view()

```

```{r data2008, include=FALSE, echo=FALSE, warning=FALSE}
# Issues in General are coded differently than WMP data from 2012 and 2016

# defpolicy
# vets
#SEPT11
#terror
#iraq
#surge
#mideast
#ISSUE_IRAQ
#FOR_NIRAQ
#china
#nukes
#iran
#afghan
# DEF_NIRQ
# 14 total issue areas


# read in the data and then summarize it
df2008 <- read_dta(here("data", "WiscAds2008_Presidential.dta"))
summary(df2008)


#Get an initial idea for how many ads there are are, which ones to watch, split. 
df_2008_split  <- df2008 %>% select(creative, defpolicy, vets, SEPT11, terror, iraq, surge, mideast, iran, afghan, DEF_NIRQ, ISSUE_IRAQ, FOR_NIRQ, nukes, china) %>% distinct() %>% view()
# Coding to find the Total # of Ads: 934



df_2008_split_watch <- df_2008_split %>% filter(defpolicy==1 | vets==1 | SEPT11==1 | terror ==1 | iraq ==1 | surge ==1 | mideast ==1 | iran ==1 | afghan ==1 | DEF_NIRQ == 1 | ISSUE_IRAQ == 1 | FOR_NIRQ == 1 | nukes == 1 | china == 1) %>% view()
#Ads to watch : 255

df_2008_split_dontwatch <- df_2008_split %>% filter(defpolicy!=1 & vets!=1 & SEPT11!=1 & terror !=1 & iraq !=1 & surge !=1 & mideast !=1 & iran !=1 & afghan !=1 & DEF_NIRQ != 1 & ISSUE_IRAQ != 1 & FOR_NIRQ != 1 & nukes != 1 & china != 1) %>% view()
# Ads to not watch : 679

df_2008_split_NA <- df_2008_split %>% filter(is.na(defpolicy) & is.na(vets) & is.na(SEPT11) & is.na(terror) & is.na(iraq) & is.na(surge) & is.na(mideast) & is.na(iran) & is.na(afghan) & is.na(DEF_NIRQ) & is.na(ISSUE_IRAQ) & is.na(FOR_NIRQ) & is.na(nukes) & is.na(china)) %>% view()
#Ads that are N/A or not coded: 0



# create a larger df by filtering and then grouping by the ID of the advertisement (variable "creative")

robustdf2008 <- df2008 %>% filter(defpolicy==1 | vets==1 | SEPT11==1 | terror ==1 | iraq ==1 | surge ==1 | mideast ==1 | iran ==1 | afghan ==1 | DEF_NIRQ == 1 | ISSUE_IRAQ == 1 | FOR_NIRQ == 1 | nukes == 1 | china == 1) %>% group_by(defpolicy, vets, SEPT11, terror, iraq, surge, mideast, iran, afghan, DEF_NIRQ, ISSUE_IRAQ, FOR_NIRQ, nukes, china, creative) %>% count(creative) %>% view()
#count - 255

# begin forming a clean data set for 2008 that can be merged with other data across the years
# step 1: filter the data using the filter command by appropriate issue area 
# step 2: select the appropriate variables to include in the data using the 'select' verb
# step 3: group the results using the 'group_by' verb
clean_df_2008 <- df2008 %>% 
  filter(defpolicy==1 | vets==1 | SEPT11==1 | terror ==1 | iraq ==1 | surge ==1 | mideast ==1 | iran ==1 | afghan ==1 | DEF_NIRQ == 1 | ISSUE_IRAQ == 1 | FOR_NIRQ == 1 | nukes == 1 | china == 1) %>% 
  select(creative, l, GROUP_NA, party, sponsor, FC_MNTN, FC_APER, OP_MNTN, AD_TONE, PRTY_MN, defpolicy, vets, SEPT11, terror, iraq, surge, mideast, iran, afghan, DEF_NIRQ, ISSUE_IRAQ, FOR_NIRQ, nukes, china) %>% 
  group_by(creative) %>% view()

# initial cut at a data set to join to other data sets.
# I've already selected which issue areas matter, so I can drop those here and 
# because these issue areas will change over time. 
# Next, I can group by the variables I am interested in. 
# Then, I create a variable for count, referring to the number of times an ad was run
# in a particular election cycle, and an ad for the election cycle more generally. 
# I then use the "distinct" function to keep only the unique rows that matter, which is key. 
# Then I rearrange the data and display it according to descending "count" order. 

initial_2008_join <- clean_df_2008 %>% 
  select(-defpolicy, -vets, -SEPT11, -terror, -iraq, -surge, -mideast, -iran, -afghan, -DEF_NIRQ, -ISSUE_IRAQ, -FOR_NIRQ, -nukes, -china) %>%
  group_by(creative, l, party) %>%
  mutate(
    count = n(), 
    year=2008
  ) %>% 
distinct() %>% 
  select(year, count, creative:PRTY_MN) %>% 
  arrange(desc(count)) %>% 
  view()
#260 total ads

# drop the "GROUP_NA" variable
df_2008_almost <- initial_2008_join %>% select(-GROUP_NA) %>% view()

# now use the mutate function to create multiple
#variables
#recode values of sponsor to match the 2012 and 2016 schemes (flip 4 and 3)

df_2008 <- df_2008_almost %>% mutate(
  f_mention=case_when(FC_MNTN==1 | FC_MNTN==3 | FC_MNTN==4 | FC_MNTN==5 ~ 1,
                      FC_MNTN==0 | FC_MNTN==2 | FC_MNTN==98 | FC_MNTN==99 ~ 0),
  f_picture=case_when(FC_MNTN==2 | FC_MNTN==3 | FC_APER==1 | FC_APER==2 ~ 1, 
                      FC_APER==0 | FC_APER==98 | FC_APER==99 | FC_MNTN != 2 | FC_MNTN != 3 ~ 0),
  f_narrate=case_when(FC_APER==1 | FC_APER==2 ~ 1, 
                      FC_APER == 0 | FC_APER==98 | FC_APER==99 ~ 0),
  o_mention=case_when(OP_MNTN == 1 | OP_MNTN ==3 ~ 1,
                      OP_MNTN ==0 | OP_MNTN==2 | OP_MNTN == 98 | OP_MNTN ==99 ~ 0),
  o_picture=case_when(OP_MNTN==2 | OP_MNTN==3 ~1,
                      OP_MNTN==0 | OP_MNTN==1 | OP_MNTN==98 | OP_MNTN==99 ~ 0),
  party=case_when(party==1 ~ "DEMOCRAT",
                  party==2 ~ "REPUBLICAN",
                  party != 1 | party!=2 ~ "OTHER"),
  sponsor=case_when(sponsor==1~1,
                    sponsor==2 ~2,
                    sponsor==3~4,
                    sponsor==4~3
                    )
  ) %>% 
  select(-FC_MNTN, -FC_APER, -OP_MNTN) %>% 
  relocate(AD_TONE, PRTY_MN, .after = o_picture) %>% 
  rename(ad_tone=AD_TONE, prty_mn=PRTY_MN) %>% 
  view()

# 260 total ads


###########UNLIKELY####################

# the following paragraph is one line of code that enabled me to gather the vital variables for when I developed the master imagery list
# it handles the variables that I need to do a meaningul comparison for the ads that are unlikely to feature military imagery
unprob_df_2008 <- df2008 %>% 
  filter(defpolicy!=1 & vets!=1 & SEPT11!=1 & terror !=1 & iraq !=1 & surge !=1 & mideast !=1 & iran !=1 & afghan !=1 & DEF_NIRQ != 1 & ISSUE_IRAQ != 1 & FOR_NIRQ != 1 & nukes != 1 & china != 1) %>% 
  select(creative, l, GROUP_NA, party, sponsor, FC_MNTN, FC_APER, OP_MNTN, AD_TONE, PRTY_MN) %>% 
  group_by(creative, l) %>% mutate(
    count = n(), 
    year=2008,
    f_mention=case_when(FC_MNTN==1 | FC_MNTN==3 | FC_MNTN==4 | FC_MNTN==5 ~ 1,
                      FC_MNTN==0 | FC_MNTN==2 | FC_MNTN==98 | FC_MNTN==99 ~ 0),
  f_picture=case_when(FC_MNTN==2 | FC_MNTN==3 | FC_APER==1 | FC_APER==2 ~ 1, 
                      FC_APER==0 | FC_APER==98 | FC_APER==99 | FC_MNTN != 2 | FC_MNTN != 3 ~ 0),
  f_narrate=case_when(FC_APER==1 | FC_APER==2 ~ 1, 
                      FC_APER == 0 | FC_APER==98 | FC_APER==99 ~ 0),
  o_mention=case_when(OP_MNTN == 1 | OP_MNTN ==3 ~ 1,
                      OP_MNTN ==0 | OP_MNTN==2 | OP_MNTN == 98 | OP_MNTN ==99 ~ 0),
  o_picture=case_when(OP_MNTN==2 | OP_MNTN==3 ~1,
                      OP_MNTN==0 | OP_MNTN==1 | OP_MNTN==98 | OP_MNTN==99 ~ 0),
  party=case_when(party==1 ~ "DEMOCRAT",
                  party==2 ~ "REPUBLICAN",
                  party != 1 | party!=2 ~ "OTHER"),
  sponsor=case_when(sponsor==1~1,
                    sponsor==2 ~2,
                    sponsor==3~4,
                    sponsor==4~3
                    )
  ) %>% 
  distinct() %>% 
  arrange(desc(count)) %>% 
  select(-FC_MNTN, -FC_APER, -OP_MNTN, -GROUP_NA) %>% 
  relocate(AD_TONE, PRTY_MN, .after = o_picture) %>% 
  rename(ad_tone=AD_TONE, prty_mn=PRTY_MN) %>% 
  select(year, count, creative, party, sponsor, f_mention, f_picture, f_narrate, o_mention, o_picture, ad_tone, prty_mn) %>% group_by(creative) %>% distinct() %>% view()


# below is the list that is "unprobable" to contain the ads I think I need to watch
clean_df_2008_unprobably <- df2008 %>% 
  filter(defpolicy!=1 & vets!=1 & SEPT11!=1 & terror !=1 & iraq !=1 & surge !=1 & mideast !=1 & iran !=1 & afghan !=1 & DEF_NIRQ != 1 & ISSUE_IRAQ != 1 & FOR_NIRQ != 1 & nukes != 1 & china != 1) %>% 
  select(creative, l, party, defpolicy, vets, SEPT11, terror, iraq, surge, mideast, iran, afghan, DEF_NIRQ, ISSUE_IRAQ, FOR_NIRQ, nukes, china) %>% mutate(
    party=case_when(party==1 ~ "DEMOCRAT",
                  party==2 ~ "REPUBLICAN",
                  party != 1 | party!=2 ~ "OTHER")
  ) %>% group_by(creative)

initial_2008_join_unprobable <- clean_df_2008_unprobably %>% 
  group_by(creative, l) %>%
  mutate(
    count = n(), 
    year=2008
  ) %>% 
distinct() %>% 
  select(year, count, creative, l, party, defpolicy, vets, SEPT11, terror, iraq, surge, mideast, iran, afghan, DEF_NIRQ, ISSUE_IRAQ, FOR_NIRQ, nukes, china) %>% 
  arrange(desc(count)) %>% 
  view()

# 706 ads to NOT watch

df_2008_notlikely <- initial_2008_join_unprobable %>% select(-defpolicy, -vets, -SEPT11, -terror, -iraq, -surge, -mideast, -iran, -afghan, -DEF_NIRQ, -ISSUE_IRAQ, -FOR_NIRQ, -nukes, -china) %>% view()



```

```{r}
# NOT RUN {
dfex <- tibble(
  x = sample(10, 100, rep = TRUE),
  y = sample(10, 100, rep = TRUE)
)
dfex %>% view()

nrow(dfex)
nrow(distinct(dfex))
nrow(distinct(dfex, x, y))

distinct(dfex, x)
distinct(dfex, y)

# Can choose to keep all other variables as well
distinct(df, x, .keep_all = TRUE)
distinct(df, y, .keep_all = TRUE)

# You can also use distinct on computed variables
distinct(df, diff = abs(x - y))

# The same behaviour applies for grouped data frames
# except that the grouping variables are always included
df <- tibble(
  g = c(1, 1, 2, 2),
  x = c(1, 1, 2, 1)
) %>% group_by(g)
df %>% distinct()
df %>% distinct(x)

# Values in list columns are compared by reference, this can lead to
# surprising results
tibble(a = as.list(c(1, 1, 2))) %>% glimpse() %>% distinct()
tibble(a = as.list(1:2)[c(1, 1, 2)]) %>% glimpse() %>% distinct()
# }
```


```{r data2004, include=FALSE, echo=FALSE, warning=FALSE}
# WiscAd Data; slight difference from 2008 data

#EISSUE: if coded 50 - then Defense/Military
#ESSUE: if coded 51 - Missile Defense/Star Wars; 
#EISSUE: if coded 52 - Veterans
#EISSUE: if coded 53 - Foreign Policy
#EISSUE: if coded 54 - Bosnia
#EISSUE: if coded 55 - China
#EISSUE: if coded 57 - terrorism
#EISSUE: if coded 58 - Middle East;
# EISSUE: if coded 59 - Afghanistan; 
#EISSUE: if coded 83 - September 11
# A total of 10 x topic areas plus 
# 2 x explicit mention variables () ESEPT11, eterror  )

# read in the data and then summarize it
df2004 <- read_dta(here("data", "WiscAds_2004 Presidential.dta"))
summary(df2004)


#Get an initial idea for how many ads there are are, which ones to watch, split. 
df_2004_split  <- df2004 %>% select(creative, EISSUE1, EISSUE2, EISSUE3, EISSUE4, ESEPT11, eterror, EACTOR1, EACTOR2, EACTOR3, EACTOR90, ecfcred, EISSUE90) %>% distinct() %>% view()
# Coding to find the Total # of Ads: 644

df_2004_split_watch <- df_2004_split %>% filter(EISSUE1 == 50 | EISSUE1 == 51 | EISSUE1 == 52 | EISSUE1 == 53 | EISSUE1 == 54 | EISSUE1 == 55 | EISSUE1 == 57 | EISSUE1 == 58 | EISSUE1 == 59 | EISSUE1 == 83 | EISSUE2 == 50 | EISSUE2 == 51 | EISSUE2 == 52 | EISSUE2 == 53 | EISSUE2 == 54 | EISSUE2 == 55 | EISSUE2 == 57 | EISSUE2 == 58 | EISSUE2 == 59 | EISSUE2 == 83 | EISSUE3 == 50 | EISSUE3 == 51 | EISSUE3 == 52 | EISSUE3 == 53 | EISSUE3 == 54 | EISSUE3 == 55 | EISSUE3 == 57 | EISSUE3 == 58 | EISSUE3 == 59 | EISSUE3 == 83 | EISSUE4 == 50 | EISSUE4 == 51 | EISSUE4 == 52 | EISSUE4 == 53 | EISSUE4 == 54 | EISSUE4 == 55 | EISSUE4 == 57 | EISSUE4 == 58 | EISSUE4 == 59 | EISSUE4 == 83 | ESEPT11==1 | eterror==1) %>% view()
#Ads to watch : 185


### several weeks after coding of ads began, I adjust the selection criteria to include 
# three variables I had initially missed.
# the first of these is the family of 'EACTOR' variables (1, 2, 3, and 90)...these variables
# measures who features prominently in the ad, one of which is a response for
# military personnel.  
# The second variable is ECFCRED, which is the source of credibility of 
# the actor in the variable, which can take the value of 'military'
# After including these associated variables, I conduct a further analysis. 
# Finally, thereis also a variable called EISSUE90, which is a "catch all," but I noticed that one 
# of these ads took the value of "iraq"

df_2004_split_watch_revised <- df_2004_split %>% filter(EISSUE1 == 50 | EISSUE1 == 51 | EISSUE1 == 52 | EISSUE1 == 53 | EISSUE1 == 54 | EISSUE1 == 55 | EISSUE1 == 57 | EISSUE1 == 58 | EISSUE1 == 59 | EISSUE1 == 83 | EISSUE2 == 50 | EISSUE2 == 51 | EISSUE2 == 52 | EISSUE2 == 53 | EISSUE2 == 54 | EISSUE2 == 55 | EISSUE2 == 57 | EISSUE2 == 58 | EISSUE2 == 59 | EISSUE2 == 83 | EISSUE3 == 50 | EISSUE3 == 51 | EISSUE3 == 52 | EISSUE3 == 53 | EISSUE3 == 54 | EISSUE3 == 55 | EISSUE3 == 57 | EISSUE3 == 58 | EISSUE3 == 59 | EISSUE3 == 83 | EISSUE4 == 50 | EISSUE4 == 51 | EISSUE4 == 52 | EISSUE4 == 53 | EISSUE4 == 54 | EISSUE4 == 55 | EISSUE4 == 57 | EISSUE4 == 58 | EISSUE4 == 59 | EISSUE4 == 83 | ESEPT11==1 | eterror==1 | EACTOR1 ==16 | EACTOR2 == 16 | EACTOR3 ==16 | EACTOR90=="family of soldiers" | ecfcred==13 | EISSUE90=="IRAQ") %>% view()
#Ads to watch : 239 revised ads

df_2004_split_dontwatch <- df_2004_split %>% filter(EISSUE1 != 50 & EISSUE1 != 51 & EISSUE1 != 52 & EISSUE1 != 53 & EISSUE1 != 54 & EISSUE1 != 55 & EISSUE1 != 57 & EISSUE1 != 58 & EISSUE1 != 59 & EISSUE1 != 83 & EISSUE2 != 50 & EISSUE2 != 51 & EISSUE2 != 52 & EISSUE2 != 53 & EISSUE2 != 54 & EISSUE2 != 55 & EISSUE2 != 57 & EISSUE2 != 58 & EISSUE2 != 59 & EISSUE2 != 83 & EISSUE3 != 50 & EISSUE3 != 51 & EISSUE3 != 52 & EISSUE3 != 53 & EISSUE3 != 54 & EISSUE3 != 55 & EISSUE3 != 57 & EISSUE3 != 58 & EISSUE3 != 59 & EISSUE3 != 83 & EISSUE4 != 50 & EISSUE4 != 51 & EISSUE4 != 52 & EISSUE4 != 53 & EISSUE4 != 54 & EISSUE4 != 55 & EISSUE4 != 57 & EISSUE4 != 58 & EISSUE4 != 59 & EISSUE4 != 83 & ESEPT11!=1 & eterror!=1) %>% view()
# Ads to not watch : 455

df_2004_split_NA <- df_2004_split %>% filter(is.na(EISSUE1) & is.na(EISSUE2) & is.na(EISSUE3) & is.na(EISSUE4) & is.na(ESEPT11) & is.na(eterror)) %>% view()
#Ads that are N/A or not coded: 2


# a more robust version of ads, if we include:
robustdf2004 <- df2004 %>% filter(EISSUE1 == 50 | EISSUE1 == 51 | EISSUE1 == 52 | EISSUE1 == 53 | EISSUE1 == 54 | EISSUE1 == 55 | EISSUE1 == 57 | EISSUE1 == 58 | EISSUE1 == 59 | EISSUE1 == 83 | EISSUE2 == 50 | EISSUE2 == 51 | EISSUE2 == 52 | EISSUE2 == 53 | EISSUE2 == 54 | EISSUE2 == 55 | EISSUE2 == 57 | EISSUE2 == 58 | EISSUE2 == 59 | EISSUE2 == 83 | EISSUE3 == 50 | EISSUE3 == 51 | EISSUE3 == 52 | EISSUE3 == 53 | EISSUE3 == 54 | EISSUE3 == 55 | EISSUE3 == 57 | EISSUE3 == 58 | EISSUE3 == 59 | EISSUE3 == 83 | EISSUE4 == 50 | EISSUE4 == 51 | EISSUE4 == 52 | EISSUE4 == 53 | EISSUE4 == 54 | EISSUE4 == 55 | EISSUE4 == 57 | EISSUE4 == 58 | EISSUE4 == 59 | EISSUE4 == 83 | ESEPT11==1 | eterror==1) %>% group_by(EISSUE1, EISSUE2, EISSUE3, EISSUE4, ESEPT11, eterror, creative) %>% count(creative) %>% view()
#count 185

# revised df, with additional selection criteria
robustdf2004_revised <- df2004 %>% filter(EISSUE1 == 50 | EISSUE1 == 51 | EISSUE1 == 52 | EISSUE1 == 53 | EISSUE1 == 54 | EISSUE1 == 55 | EISSUE1 == 57 | EISSUE1 == 58 | EISSUE1 == 59 | EISSUE1 == 83 | EISSUE2 == 50 | EISSUE2 == 51 | EISSUE2 == 52 | EISSUE2 == 53 | EISSUE2 == 54 | EISSUE2 == 55 | EISSUE2 == 57 | EISSUE2 == 58 | EISSUE2 == 59 | EISSUE2 == 83 | EISSUE3 == 50 | EISSUE3 == 51 | EISSUE3 == 52 | EISSUE3 == 53 | EISSUE3 == 54 | EISSUE3 == 55 | EISSUE3 == 57 | EISSUE3 == 58 | EISSUE3 == 59 | EISSUE3 == 83 | EISSUE4 == 50 | EISSUE4 == 51 | EISSUE4 == 52 | EISSUE4 == 53 | EISSUE4 == 54 | EISSUE4 == 55 | EISSUE4 == 57 | EISSUE4 == 58 | EISSUE4 == 59 | EISSUE4 == 83 | ESEPT11==1 | eterror==1 | EACTOR1 ==16 | EACTOR2 == 16 | EACTOR3 ==16 | EACTOR90=="family of soldiers" | ecfcred==13 | EISSUE90=="IRAQ") %>% group_by(EISSUE1, EISSUE2, EISSUE3, EISSUE4, ESEPT11, eterror, EACTOR1, EACTOR2, EACTOR3, EACTOR90, ecfcred, EISSUE90, creative) %>% count(creative) %>% view()
#count 239

# revised dontwatch list
df_2004_split_dontwatch_revised <- df_2004_split %>% filter(EISSUE1 != 50 & EISSUE1 != 51 & EISSUE1 != 52 & EISSUE1 != 53 & EISSUE1 != 54 & EISSUE1 != 55 & EISSUE1 != 57 & EISSUE1 != 58 & EISSUE1 != 59 & EISSUE1 != 83 & EISSUE2 != 50 & EISSUE2 != 51 & EISSUE2 != 52 & EISSUE2 != 53 & EISSUE2 != 54 & EISSUE2 != 55 & EISSUE2 != 57 & EISSUE2 != 58 & EISSUE2 != 59 & EISSUE2 != 83 & EISSUE3 != 50 & EISSUE3 != 51 & EISSUE3 != 52 & EISSUE3 != 53 & EISSUE3 != 54 & EISSUE3 != 55 & EISSUE3 != 57 & EISSUE3 != 58 & EISSUE3 != 59 & EISSUE3 != 83 & EISSUE4 != 50 & EISSUE4 != 51 & EISSUE4 != 52 & EISSUE4 != 53 & EISSUE4 != 54 & EISSUE4 != 55 & EISSUE4 != 57 & EISSUE4 != 58 & EISSUE4 != 59 & EISSUE4 != 83 & ESEPT11!=1 & eterror!=1 & EACTOR1 !=16 & EACTOR2 != 16 & EACTOR3 !=16 & EACTOR90!="family of soldiers" & ecfcred!=13 & EISSUE90!="IRAQ") %>% view()
# REvised Ads to not watch : 399




# begin forming a clean data set for 2004 that can be merged with other data across the years
# step 1: filter the data using the filter command by appropriate issue area 
# step 2: select the appropriate variables to include in the data using the 'select' verb
# step 3: group the results using the 'group_by' verb
clean_df_2004 <- df2004 %>% 
  filter(EISSUE1 == 50 | EISSUE1 == 51 | EISSUE1 == 52 | EISSUE1 == 53 | EISSUE1 == 54 | EISSUE1 == 55 | EISSUE1 == 57 | EISSUE1 == 58 | EISSUE1 == 59 | EISSUE1 == 83 | EISSUE2 == 50 | EISSUE2 == 51 | EISSUE2 == 52 | EISSUE2 == 53 | EISSUE2 == 54 | EISSUE2 == 55 | EISSUE2 == 57 | EISSUE2 == 58 | EISSUE2 == 59 | EISSUE2 == 83 | EISSUE3 == 50 | EISSUE3 == 51 | EISSUE3 == 52 | EISSUE3 == 53 | EISSUE3 == 54 | EISSUE3 == 55 | EISSUE3 == 57 | EISSUE3 == 58 | EISSUE3 == 59 | EISSUE3 == 83 | EISSUE4 == 50 | EISSUE4 == 51 | EISSUE4 == 52 | EISSUE4 == 53 | EISSUE4 == 54 | EISSUE4 == 55 | EISSUE4 == 57 | EISSUE4 == 58 | EISSUE4 == 59 | EISSUE4 == 83 | ESEPT11==1 | eterror==1) %>% 
  select(creative, spotleng, Party, Sponsor, EFC_MNTN, EFC_APER, EOP_MNTN, EAD_TONE, EPRTY_MN, EISSUE1, EISSUE2, EISSUE3, EISSUE4, ESEPT11, eterror) %>% 
  group_by(creative) %>% view()

## revised

clean_df_2004_revised <- df2004 %>% 
  filter(EISSUE1 == 50 | EISSUE1 == 51 | EISSUE1 == 52 | EISSUE1 == 53 | EISSUE1 == 54 | EISSUE1 == 55 | EISSUE1 == 57 | EISSUE1 == 58 | EISSUE1 == 59 | EISSUE1 == 83 | EISSUE2 == 50 | EISSUE2 == 51 | EISSUE2 == 52 | EISSUE2 == 53 | EISSUE2 == 54 | EISSUE2 == 55 | EISSUE2 == 57 | EISSUE2 == 58 | EISSUE2 == 59 | EISSUE2 == 83 | EISSUE3 == 50 | EISSUE3 == 51 | EISSUE3 == 52 | EISSUE3 == 53 | EISSUE3 == 54 | EISSUE3 == 55 | EISSUE3 == 57 | EISSUE3 == 58 | EISSUE3 == 59 | EISSUE3 == 83 | EISSUE4 == 50 | EISSUE4 == 51 | EISSUE4 == 52 | EISSUE4 == 53 | EISSUE4 == 54 | EISSUE4 == 55 | EISSUE4 == 57 | EISSUE4 == 58 | EISSUE4 == 59 | EISSUE4 == 83 | ESEPT11==1 | eterror==1 | EACTOR1 ==16 | EACTOR2 == 16 | EACTOR3 ==16 | EACTOR90=="family of soldiers" | ecfcred==13 | EISSUE90=="IRAQ") %>% 
  select(creative, spotleng, Party, Sponsor, EFC_MNTN, EFC_APER, EOP_MNTN, EAD_TONE, EPRTY_MN, EISSUE1, EISSUE2, EISSUE3, EISSUE4, ESEPT11, eterror, EACTOR1, EACTOR2, EACTOR3, EACTOR90, ecfcred, EISSUE90) %>% 
  group_by(creative) %>% view()


# initial cut at a data set to join to other data sets.
# I've already selected which issue areas matter, so I can drop those here and 
# because these issue areas will change over time. 
# Next, I can group by the variables I am interested in. 
# Then, I create a variable for count, referring to the number of times an ad was run
# in a particular election cycle, and an ad for the election cycle more generally. 
# I then use the "distinct" function to keep only the unique rows that matter, which is key. 
# Then I rearrange the data and display it according to descending "count" order. 

initial_2004_join <- clean_df_2004 %>% 
  select(-EISSUE1, -EISSUE2, -EISSUE3, -EISSUE4, -ESEPT11, -eterror) %>% 
  group_by(creative, spotleng, Party) %>%
  mutate(
    count = n(), 
    year=2004
  ) %>% 
distinct() %>% 
  select(year, count, creative:EPRTY_MN) %>% 
  arrange(desc(count)) %>% 
  view()
# 191 Ads

#revised

initial_2004_join_revised <- clean_df_2004_revised %>% 
  select(-EISSUE1, -EISSUE2, -EISSUE3, -EISSUE4, -ESEPT11, -eterror, -EACTOR1, -EACTOR2, -EACTOR3, -EACTOR90, -ecfcred, -EISSUE90) %>% 
  group_by(creative, spotleng, Party) %>%
  mutate(
    count = n(), 
    year=2004
  ) %>% 
distinct() %>% 
  select(year, count, creative:EPRTY_MN) %>% 
  arrange(desc(count)) %>% 
  view()
# 246 Ads


# use mutate function to create multiple variables to match 2012/2016 schemes
#recode values of sponsor and adtone to match the 2012 and 2016 schemes (flip 4 and 3)
df_2004 <- initial_2004_join %>% mutate(
  f_mention=case_when(EFC_MNTN==1 | EFC_MNTN==3 | EFC_MNTN==4 | EFC_MNTN==5 ~ 1,
                      EFC_MNTN==0 | EFC_MNTN==2 | EFC_MNTN==98 | EFC_MNTN==99 ~ 0),
  f_picture=case_when(EFC_MNTN==2 | EFC_MNTN==3 | EFC_APER==1 | EFC_APER==2 ~ 1, 
                      EFC_APER==0 | EFC_APER==98 | EFC_APER==99 | EFC_MNTN != 2 | EFC_MNTN != 3 ~ 0),
  f_narrate=case_when(EFC_APER==1 | EFC_APER==2 ~ 1, 
                      EFC_APER == 0 | EFC_APER==98 | EFC_APER==99 ~ 0),
  o_mention=case_when(EOP_MNTN == 1 | EOP_MNTN ==3 ~ 1,
                      EOP_MNTN ==0 | EOP_MNTN==2 | EOP_MNTN == 98 | EOP_MNTN ==99 ~ 0),
  o_picture=case_when(EOP_MNTN==2 | EOP_MNTN==3 ~1,
                      EOP_MNTN==0 | EOP_MNTN==1 | EOP_MNTN==98 | EOP_MNTN==99 ~ 0),
  ad_tone=case_when(EAD_TONE == 2 ~1,
                    EAD_TONE == 3 ~ 2,
                    EAD_TONE == 1 ~ 3,
                    EAD_TONE <= 98 ~4),
  Party=case_when(Party==1 ~ "DEMOCRAT",
                  Party==2 ~ "REPUBLICAN",
                  Party != 1 | Party!=2 ~ "OTHER"),
  sponsor=case_when(Sponsor==1 ~ 1,
                    Sponsor ==2 ~2,
                    Sponsor ==3 ~4,
                    Sponsor ==4 ~3 )
  )%>% 
  select(-EFC_MNTN, -EFC_APER,-EOP_MNTN, -EAD_TONE, -Sponsor) %>% 
  relocate(EPRTY_MN, .after = ad_tone) %>% 
  rename(
    l=spotleng,
    party=Party,
    prty_mn=EPRTY_MN
  ) %>% 
  relocate(sponsor, .after = party) %>% 
  view()

# 191 ads to watch

df_2004_revised <- initial_2004_join_revised %>% mutate(
  f_mention=case_when(EFC_MNTN==1 | EFC_MNTN==3 | EFC_MNTN==4 | EFC_MNTN==5 ~ 1,
                      EFC_MNTN==0 | EFC_MNTN==2 | EFC_MNTN==98 | EFC_MNTN==99 ~ 0),
  f_picture=case_when(EFC_MNTN==2 | EFC_MNTN==3 | EFC_APER==1 | EFC_APER==2 ~ 1, 
                      EFC_APER==0 | EFC_APER==98 | EFC_APER==99 | EFC_MNTN != 2 | EFC_MNTN != 3 ~ 0),
  f_narrate=case_when(EFC_APER==1 | EFC_APER==2 ~ 1, 
                      EFC_APER == 0 | EFC_APER==98 | EFC_APER==99 ~ 0),
  o_mention=case_when(EOP_MNTN == 1 | EOP_MNTN ==3 ~ 1,
                      EOP_MNTN ==0 | EOP_MNTN==2 | EOP_MNTN == 98 | EOP_MNTN ==99 ~ 0),
  o_picture=case_when(EOP_MNTN==2 | EOP_MNTN==3 ~1,
                      EOP_MNTN==0 | EOP_MNTN==1 | EOP_MNTN==98 | EOP_MNTN==99 ~ 0),
  ad_tone=case_when(EAD_TONE == 2 ~1,
                    EAD_TONE == 3 ~ 2,
                    EAD_TONE == 1 ~ 3,
                    EAD_TONE <= 98 ~4),
  Party=case_when(Party==1 ~ "DEMOCRAT",
                  Party==2 ~ "REPUBLICAN",
                  Party != 1 | Party!=2 ~ "OTHER"),
  sponsor=case_when(Sponsor==1 ~ 1,
                    Sponsor ==2 ~2,
                    Sponsor ==3 ~4,
                    Sponsor ==4 ~3 )
  )%>% 
  select(-EFC_MNTN, -EFC_APER,-EOP_MNTN, -EAD_TONE, -Sponsor) %>% 
  relocate(EPRTY_MN, .after = ad_tone) %>% 
  rename(
    l=spotleng,
    party=Party,
    prty_mn=EPRTY_MN
  ) %>% 
  relocate(sponsor, .after = party) %>% 
  view()

# 246 ads to watch



###########UNLIKELY##############

# the following paragraph is one line of code that enabled me to gather the vital variables for when I developed the master imagery list
# it handles the variables that I need to do a meaningul comparison for the ads that are unlikely to feature military imagery
unprob_df_2004 <- df2004 %>% 
  filter(EISSUE1 != 50 & EISSUE1 != 51 & EISSUE1 != 52 & EISSUE1 != 53 & EISSUE1 != 54 & EISSUE1 != 55 & EISSUE1 != 57 & EISSUE1 != 58 & EISSUE1 != 59 & EISSUE1 != 83 & EISSUE2 != 50 & EISSUE2 != 51 & EISSUE2 != 52 & EISSUE2 != 53 & EISSUE2 != 54 & EISSUE2 != 55 & EISSUE2 != 57 & EISSUE2 != 58 & EISSUE2 != 59 & EISSUE2 != 83 & EISSUE3 != 50 & EISSUE3 != 51 & EISSUE3 != 52 & EISSUE3 != 53 & EISSUE3 != 54 & EISSUE3 != 55 & EISSUE3 != 57 & EISSUE3 != 58 & EISSUE3 != 59 & EISSUE3 != 83 & EISSUE4 != 50 & EISSUE4 != 51 & EISSUE4 != 52 & EISSUE4 != 53 & EISSUE4 != 54 & EISSUE4 != 55 & EISSUE4 != 57 & EISSUE4 != 58 & EISSUE4 != 59 & EISSUE4 != 83 & ESEPT11!=1 & eterror!=1 & EACTOR1 !=16 & EACTOR2 != 16 & EACTOR3 !=16 & EACTOR90!="family of soldiers" & ecfcred!=13 & EISSUE90!="IRAQ") %>% 
  select(creative, spotleng, Party, Sponsor, EFC_MNTN, EFC_APER, EOP_MNTN, EAD_TONE, EPRTY_MN, EISSUE1, EISSUE2, EISSUE3, EISSUE4, ESEPT11, eterror, EACTOR1, EACTOR2, EACTOR3, EACTOR90, ecfcred, EISSUE90) %>% 
  group_by(creative, spotleng, Party) %>% mutate(
    count = n(), 
    year=2004,
    f_mention=case_when(EFC_MNTN==1 | EFC_MNTN==3 | EFC_MNTN==4 | EFC_MNTN==5 ~ 1,
                      EFC_MNTN==0 | EFC_MNTN==2 | EFC_MNTN==98 | EFC_MNTN==99 ~ 0),
  f_picture=case_when(EFC_MNTN==2 | EFC_MNTN==3 | EFC_APER==1 | EFC_APER==2 ~ 1, 
                      EFC_APER==0 | EFC_APER==98 | EFC_APER==99 | EFC_MNTN != 2 | EFC_MNTN != 3 ~ 0),
  f_narrate=case_when(EFC_APER==1 | EFC_APER==2 ~ 1, 
                      EFC_APER == 0 | EFC_APER==98 | EFC_APER==99 ~ 0),
  o_mention=case_when(EOP_MNTN == 1 | EOP_MNTN ==3 ~ 1,
                      EOP_MNTN ==0 | EOP_MNTN==2 | EOP_MNTN == 98 | EOP_MNTN ==99 ~ 0),
  o_picture=case_when(EOP_MNTN==2 | EOP_MNTN==3 ~1,
                      EOP_MNTN==0 | EOP_MNTN==1 | EOP_MNTN==98 | EOP_MNTN==99 ~ 0),
  ad_tone=case_when(EAD_TONE == 2 ~1,
                    EAD_TONE == 3 ~ 2,
                    EAD_TONE == 1 ~ 3,
                    EAD_TONE <= 98 ~4),
  Party=case_when(Party==1 ~ "DEMOCRAT",
                  Party==2 ~ "REPUBLICAN",
                  Party != 1 | Party!=2 ~ "OTHER"),
  sponsor=case_when(Sponsor==1 ~ 1,
                    Sponsor ==2 ~2,
                    Sponsor ==3 ~4,
                    Sponsor ==4 ~3 )
  ) %>% 
  distinct() %>% 
  arrange(desc(count)) %>% 
  rename(
    l=spotleng,
    party=Party,
    prty_mn=EPRTY_MN
  )%>% 
  select(year, count, creative, l, party, sponsor, f_mention, f_picture, f_narrate, o_mention, o_picture, ad_tone, prty_mn) %>%  
  view()

unprob_df_2004 %>% select(creative) %>% distinct() %>% view() 


# below is the list that is "unprobable" to contain the ads I think I need to watch


clean_df_2004_unprobably <- df2004 %>% 
  filter(EISSUE1 != 50 & EISSUE1 != 51 & EISSUE1 != 52 & EISSUE1 != 53 & EISSUE1 != 54 & EISSUE1 != 55 & EISSUE1 != 57 & EISSUE1 != 58 & EISSUE1 != 59 & EISSUE1 != 83 & EISSUE2 != 50 & EISSUE2 != 51 & EISSUE2 != 52 & EISSUE2 != 53 & EISSUE2 != 54 & EISSUE2 != 55 & EISSUE2 != 57 & EISSUE2 != 58 & EISSUE2 != 59 & EISSUE2 != 83 & EISSUE3 != 50 & EISSUE3 != 51 & EISSUE3 != 52 & EISSUE3 != 53 & EISSUE3 != 54 & EISSUE3 != 55 & EISSUE3 != 57 & EISSUE3 != 58 & EISSUE3 != 59 & EISSUE3 != 83 & EISSUE4 != 50 & EISSUE4 != 51 & EISSUE4 != 52 & EISSUE4 != 53 & EISSUE4 != 54 & EISSUE4 != 55 & EISSUE4 != 57 & EISSUE4 != 58 & EISSUE4 != 59 & EISSUE4 != 83 & ESEPT11!=1 & eterror!=1) %>% 
  select(creative, spotleng, Party, EISSUE1, EISSUE2, EISSUE3, EISSUE4, ESEPT11, eterror) %>%
  group_by(creative) %>% view()

initial_2004_join_unprobable <- clean_df_2004_unprobably %>% 
  group_by(creative, spotleng) %>%
  mutate(
    count = n(), 
    year=2004,
    Party=case_when(Party==1 ~ "DEMOCRAT",
                  Party==2 ~ "REPUBLICAN",
                  Party != 1 | Party!=2 ~ "OTHER")
  ) %>% 
distinct() %>% 
  select(year, count, creative, spotleng, Party, EISSUE1, EISSUE2, EISSUE3, EISSUE4, ESEPT11, eterror) %>% rename(party=Party) %>% arrange(desc(count)) %>% 
  view()

# 462 ads to NOT watch

df_2004_notlikely <- initial_2004_join_unprobable %>% 
  select(-EISSUE1, -EISSUE2, -EISSUE3, -EISSUE4, -ESEPT11, -eterror) %>% 
  rename(
    l=spotleng
    ) %>% view()


```


```{r data2000, include=FALSE, echo=FALSE, warning=FALSE}
# WiscAd Data; slight difference from 2004 data
# instead of creative, title of ad called "custitle"
#q32-35 deal with campaign themes: 

# 50 - Defense
# 51 - missile defense/star wars; 
# 52 - Veterans
# 53 - Foreign Policy
# 54 - Bosnia; 
# 55 - China;
# 59 - Other defense/foreign policy issues
# 7 total issues


# read in the data and then summarize it
df2000 <- read_dta(here("data", "Political Advertising in 2000.dta"))
summary(df2000)

# this data is different from the other data sets in that it
# includes advertisements from multiple types of races (house, senate, President, etc.)
# Must therefore first filter for Presidential 
# as of 1 September 21, I am trimming down races to only presidential issues (where q3==95). Previously, q3 could also == 80 or 99
# where 80 == issue advocacy or 99 == OTHER

#Get an initial idea for how many ads there are are, which ones to watch, split. 
df_2000_split  <- df2000 %>% select(custitle, q3, q32, q33, q34, q35) %>% 
  filter(q3 == 95) %>% 
  distinct() %>% view()
# Coding to find the Total # of Presidential Ads: 344


df_2000_split_watch <- df_2000_split %>% filter(q32 == 50 | q33 == 50 | q34 == 50 | q35 == 50 | q32 == 51 | q33 == 51 | q34 == 51 | q35 == 51 | q32 == 52 | q33 == 52 | q34 == 52 | q35 == 52 | q32 == 53 | q33 == 53 | q34 == 53 | q35 == 53 | q32 == 54 | q33 == 54 | q34 == 54 | q35 == 54 | q32 == 55 | q33 == 55 | q34 == 55 | q35 == 55 | q32 == 59 | q33 == 59 | q34 == 59 | q35 == 59) %>% view()
#Ads to watch : 17

df_2000_split_dontwatch <- df_2000_split %>% filter(q32 != 50 & q33 != 50 & q34 != 50 & q35 != 50 & q32 != 51 & q33 != 51 & q34 != 51 & q35 != 51 & q32 != 52 & q33 != 52 & q34 != 52 & q35 != 52 & q32 != 53 & q33 != 53 & q34 != 53 & q35 != 53 & q32 != 54 & q33 != 54 & q34 != 54 & q35 != 54 & q32 != 55 & q33 != 55 & q34 != 55 & q35 != 55 & q32 != 59 & q33 != 59 & q34 != 59 & q35 != 59) %>% view()
# Ads to not watch : 327

# Therefore, Iam only interested in viewing/coding 327+17 ads = 344 total. 

# begin forming a clean data set for 2004 that can be merged with other data across the years
# step 1: filter the data using the filter command by appropriate issue area 
# step 2: select the appropriate variables to include in the data using the 'select' verb
# step 3: group the results using the 'group_by' verb
clean_df_2000 <- df2000 %>% 
  filter(q32 == 50 | q33 == 50 | q34 == 50 | q35 == 50 | q32 == 51 | q33 == 51 | q34 == 51 | q35 == 51 | q32 == 52 | q33 == 52 | q34 == 52 | q35 == 52 | q32 == 53 | q33 == 53 | q34 == 53 | q35 == 53 | q32 == 54 | q33 == 54 | q34 == 54 | q35 == 54 | q32 == 55 | q33 == 55 | q34 == 55 | q35 == 55 | q32 == 59 | q33 == 59 | q34 == 59 | q35 == 59) %>% 
  select(custitle, len, q5, sponsor, q3, q12, q13, q14, q15, q20, q21, q22, q32, q33, q34, q35) %>% 
  filter(q3 == 95) %>% group_by(custitle) %>% view()


# initial cut at a data set to join to other data sets.
# I've already selected which issue areas matter, so I can drop those here and 
# because these issue areas will change over time. 
# Next, I can group by the variables I am interested in. 
# Then, I create a variable for count, referring to the number of times an ad was run
# in a particular election cycle, and an ad for the election cycle more generally. 
# I then use the "distinct" function to keep only the unique rows that matter, which is key. 
# Then I rearrange the data and display it according to descending "count" order. 

initial_2000_join <- clean_df_2000 %>% 
  select(-q3, -q32, -q33, -q34, -q35) %>% 
  group_by(custitle, len, sponsor) %>%
  mutate(
    count = n(), 
    year=2000
  ) %>% 
distinct() %>% 
  select(year, count, custitle:q22) %>% 
  arrange(desc(count)) %>% 
  view()
# 18 ads


# below, I create several variables (using other, similar existing variables) 
# to match existing variables in the 2012 and 2016 data sets.  
# these variables include f_mention, f_picture, f_narrate, o_mention, o_picture
# ad_tone, prty_mn, Party, and sponsor.

df_2000 <- initial_2000_join %>% mutate(
  f_mention=case_when(q12== 1 | q12== 4 ~ 1,
                      q12==0 | q12 ==3 ~ 0),
  f_picture=case_when(q12==2 | q12==4 ~ 1, 
                      q12 ==1 | q12 ==3 | q12==0 ~ 0),
  f_narrate=case_when(q15==1 ~ 1, 
                      q15 == 0 | q15 == 9 ~ 0),
  o_mention=case_when(q13 == 4 | q13 ==2 ~ 1,
                      q13 ==0 | q13==1 | q13 == 3 ~ 0),
  o_picture=case_when(q13==3 | q13==2 ~1,
                      q13==0 | q13==1 | q13==4 ~ 0),
  ad_tone=case_when(q14 == 2 ~ 1,
                    q14 == 3 ~ 2,
                    q14 == 1 ~ 3,
                    q14 ==0 | q14==4 ~ 4),
  prty_mn=case_when(q22==0 | q22==1 ~ 0,
                    q22==4 ~ 1,
                    q22 == 2 ~2,
                    q22 == 3 ~3),
  Party=case_when(q5==1 ~ "DEMOCRAT",
                  q5==2 ~ "REPUBLICAN",
                  q5 != 1 | q5 != 2 ~ "OTHER"),
  sponsor=case_when(sponsor ==1 | sponsor ==2 | sponsor ==9 | sponsor ==10 | sponsor == 11 | sponsor ==12 ~ 1,
                    sponsor ==3 | sponsor ==4 | sponsor ==13 | sponsor == 14 | sponsor ==15 | sponsor ==16 ~ 2,
                    sponsor == 7 | sponsor ==8 ~ 3,
                    sponsor ==5 | sponsor ==6 | sponsor ==17 | sponsor == 18 | sponsor ==80 ~ 4)
  ) %>% 
  select(-q5, -q12, -q13, -q14, -q15, -q20, -q21, -q22) %>% 
  relocate(Party, .after = len) %>% 
  rename(
    creative=custitle, 
    l=len,
    party=Party
  ) %>% 
  view()



##########Unlikely############

unprob_df_2000 <- df2000 %>% 
  filter(q32 != 50 & q33 != 50 & q34 != 50 & q35 != 50 & q32 != 51 & q33 != 51 & q34 != 51 & q35 != 51 & q32 != 52 & q33 != 52 & q34 != 52 & q35 != 52 & q32 != 53 & q33 != 53 & q34 != 53 & q35 != 53 & q32 != 54 & q33 != 54 & q34 != 54 & q35 != 54 & q32 != 55 & q33 != 55 & q34 != 55 & q35 != 55 & q32 != 59 & q33 != 59 & q34 != 59 & q35 != 59) %>% 
  select(custitle, len, q5, sponsor, q3, q12, q13, q14, q15, q20, q21, q22, q32, q33, q34, q35) %>% 
  filter(q3 == 95) %>% group_by(custitle, len, sponsor) %>%
  mutate(
    count = n(), 
    year=2000,
    f_mention=case_when(q12== 1 | q12== 4 ~ 1,
                      q12==0 | q12 ==3 ~ 0),
  f_picture=case_when(q12==2 | q12==4 ~ 1, 
                      q12 ==1 | q12 ==3 | q12==0 ~ 0),
  f_narrate=case_when(q15==1 ~ 1, 
                      q15 == 0 | q15 == 9 ~ 0),
  o_mention=case_when(q13 == 4 | q13 ==2 ~ 1,
                      q13 ==0 | q13==1 | q13 == 3 ~ 0),
  o_picture=case_when(q13==3 | q13==2 ~1,
                      q13==0 | q13==1 | q13==4 ~ 0),
  ad_tone=case_when(q14 == 2 ~ 1,
                    q14 == 3 ~ 2,
                    q14 == 1 ~ 3,
                    q14 ==0 | q14==4 ~ 4),
  prty_mn=case_when(q22==0 | q22==1 ~ 0,
                    q22==4 ~ 1,
                    q22 == 2 ~2,
                    q22 == 3 ~3),
  Party=case_when(q5==1 ~ "DEMOCRAT",
                  q5==2 ~ "REPUBLICAN",
                  q5 != 1 | q5 != 2 ~ "OTHER"),
  sponsor=case_when(sponsor ==1 | sponsor ==2 | sponsor ==9 | sponsor ==10 | sponsor == 11 | sponsor ==12 ~ 1,
                    sponsor ==3 | sponsor ==4 | sponsor ==13 | sponsor == 14 | sponsor ==15 | sponsor ==16 ~ 2,
                    sponsor == 7 | sponsor ==8 ~ 3,
                    sponsor ==5 | sponsor ==6 | sponsor ==17 | sponsor == 18 | sponsor ==80 ~ 4)
  ) %>% distinct() %>% select(-q5, -q12, -q13, -q14, -q15, -q20, -q21, -q22) %>% 
  relocate(Party, .after = len) %>% 
  rename(
    creative=custitle, 
    l=len,
    party=Party
  ) %>% select(year, count, creative, l, party, sponsor, f_mention, f_picture, f_narrate, o_mention, o_picture, ad_tone, prty_mn) %>% arrange(desc(count)) %>% 
  view()
  
  

# below is the list that is "unprobable" to contain the ads I think I need to watch
clean_df_2000_unprobably <- df2000 %>% 
  filter(q32 != 50 & q33 != 50 & q34 != 50 & q35 != 50 & q32 != 51 & q33 != 51 & q34 != 51 & q35 != 51 & q32 != 52 & q33 != 52 & q34 != 52 & q35 != 52 & q32 != 53 & q33 != 53 & q34 != 53 & q35 != 53 & q32 != 54 & q33 != 54 & q34 != 54 & q35 != 54 & q32 != 55 & q33 != 55 & q34 != 55 & q35 != 55 & q32 != 59 & q33 != 59 & q34 != 59 & q35 != 59) %>% 
  select(custitle, len, q3, q5, q32, q33, q34, q35) %>% 
  filter(q3== 95) %>% 
  group_by(custitle) %>% view()

initial_2000_join_unprobable <- clean_df_2000_unprobably %>% 
  group_by(custitle, len) %>%
  mutate(
    count = n(), 
    year=2000
  ) %>% 
distinct() %>% 
  select(year, count, custitle, len, q3, q5, q32, q33, q34, q35) %>% 
  arrange(desc(count)) %>% 
  view()

# 328 ads to NOT watch. 

df_2000_notlikely <- initial_2000_join_unprobable %>% mutate(
  party=case_when(q5==1 ~ "DEMOCRAT",
                  q5==2 ~ "REPUBLICAN",
                  q5 != 1 | q5 != 2 ~ "OTHER")
  ) %>% select(-q3, -q5, -q32, -q33, -q34, -q35) %>% 
   rename(
    creative=custitle, 
    l=len) %>% drop_na() %>% view()

```

```{r join_data_frames, include=FALSE, echo=FALSE, warning=FALSE}

# Here, bind the rows of all pertinent data frames and 
# then write to an excel file for future coding. 

# File for Likely
df_comb_init_cads <- bind_rows(df_2016, df_2012, df_2008, df_2004, df_2000) %>% view()
df_comb_init_cads %>% write_xlsx(here("data", "df_comb_init_cads.xlsx"))

# Revised File for Likely
revised_df_comb_init_cads <- bind_rows(df_2016, df_2012, df_2008, df_2004_revised, df_2000) %>% view()
revised_df_comb_init_cads %>% write_xlsx(here("data", "revised_df_comb_init_cads.xlsx"))

# File for Initial Unlikely File
df_comb_init_cads_unlikely <- bind_rows(df_2016_notlikely, df_2012_notlikely, df_2008_notlikely, df_2004_notlikely, df_2000_notlikely) %>% view() 
df_comb_init_cads_unlikely %>% write_xlsx(here("data", "df_comb_init_cads_unlikely.xlsx"))

## File for the Master Imagery Check; Have to Carefully Comb Through Years 2000, 2004, and 2008
ch6_master_imagery_check_roughdraft <- bind_rows(unprob_df_2016, unprob_df_2012, unprob_df_2008, unprob_df_2004, unprob_df_2000) %>% view()
ch6_master_imagery_check_roughdraft %>% write_xlsx(here("data", "ch6_master_imagery_roughdraft.xlsx"))

# Now, I use the Master Imagery File to Do Some More Data Sorting
# I produce a final excel file from which to bounce off the regression file

master_imageryDF <- read_csv(here("data", "ch6_Master_Imagery.csv")) %>% ungroup() %>% view()

refined_master_imagery_ch4 <- master_imageryDF %>% group_by(year) %>% arrange(year, desc(count)) %>% distinct(.keep_all=TRUE) %>% view()

refined_master_imagery_ch4 %>% write_xlsx(here("data", "refined_master_imagery_ch4.xlsx"))

```
