---
title: "True Colors? The Military Institution and Campaign Advertisements"
author: "Peter M. Erickson"
date: |
  `r format(Sys.time(), '%B %d, %Y')`
abstract: |
  
  _Using data from the Wisconsin Advertising Project and the Wesleyan Media Project consisting of Presidential Campaign advertisements used in five Presidential Elections from 2000 -2016, I conduct a large-n study to explore variation in the ways that campaign advertisements have featured the military over time. From a theoretical standpoint, the chapter argues that the military's relatively high prestige, combined with polarization that has dramatically increased over the time interval examined, has led to civilian actors using the military in ways that expressly challenge fundamental standards of American civil-military relations, to include the non-partisanship of the military institution. This chapter ultimately provides strong evidence that increasing polarization has led to both military and civilian actors becoming increasingly willing to challenge principles of civil-military relations._  
  
geometry: margin = 1.15in              # margin widths
fontsize: 12pt                         # text size
bibliography: dissertation.bib         # bibliography file
biblio-style: authoryear               # citation style
urlcolor: blue                         # links to external URLs
citecolor: black                       # links to bibliography
linkcolor: magenta                     # internal figures, footnotes, etc.
indent: true                          # paragraph format
                                       # true for indented paragraphs
                                       
header-includes:
    - \usepackage{setspace}
    - \usepackage{multirow}
    - \usepackage{indentfirst}
   
    
output: 
  bookdown::pdf_document2:
    keep_tex: true
    latex_engine: pdflatex
    number_sections: false
    toc: false
    fig_caption: true
    citation_package: biblatex
    includes: 
      in_header: 
        - latex-preamble.tex
  bookdown::htmldocument2: default
  bookdown::word_document2:
      toc: false
---

\doublespacing 

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo=FALSE,
  include=FALSE,
  message=FALSE,
  error=FALSE,
  fig.height=3,
  fig.width=4,
  out.width="90%")
```

```{r, include = FALSE, echo = FALSE}
library("here")       # file paths
library("tidyverse")  # workhorse package
library("tidylog")
library("kableExtra")
library("knitr")
library("ggdag")
library("dagitty")
library("gridExtra")
library("broom")
library("dplyr")
library("reshape2")
library("pacman")
library("ggrepel")
```

\newpage

# Introduction

On October 10, 2020, then candidate Joe Biden's Campaign Team aired an advertisement touting the former Vice President's national security credentials, including his experience working with the military as both Vice President and before that, US Senator from Delaware. The advertisement is narrated by a man identified as Jim, an "Iraq War Veteran," who describes the role of the former Vice President in delivering Mine Resistant Ambush Protected (MRAP) vehicles, whose heavy armor helped protect servicemen and women while deployed to Iraq and Afghanistan. Within days, however, the advertisement was pulled by the Biden Campaign after a retired Army Lieutenant General, Sean McFarland, objected to the advertisement on the grounds that the ad used images of him without his consent [@moore_biden_2020]. These images showed the general, then in uniform and on active duty, shaking hands with Biden as the former Vice President arrived to Iraq [@moore_biden_2020]. 

In objecting to the advertisement, MacFarland noted that he had not granted permission to the Biden Campaign to use these images in the advertisement, but also added that in his opinion, it is normatively wrong for the military - especially active duty forces - to appear in campaign advertisements altogether, as doing so jeopardizes the military's non-partisan ethic. Indeed, MacFarland expressed these sentiments in a public statement after the advertisement aired: "A number of people have understandably inferred that my appearance [in this Biden Campaign advertisement] constitutes an endorsement of the former Vice President. It does not. To be clear, I have not endorsed President Trump, either. I object to the use of ANY military personnel in uniform in political ads - full stop" [MacFarland quoted in @moore_biden_2020]. 

This incident serves as one of several during the 2020 Presidential Election in which political candidates used the military in ways that push against, challenge, and in some cases, violate standards which govern the non-partisan ideals of the military.^[President Trump's Campaign likewise used military symbols and imagery inappropriately, and perhaps on more than one occasion. For a salient example in which the Chairman of the Joint Chiefs of Staff appeared in one of Trump's advertisements, see @seligman_top_2020.] And while civilians have long appealed to the public by harnessing the popularity of the military or of individual military figures in a variety of ways, the practice of doing so during election seasons highlights the ways in which polarized politics challenge the military and its members in seeking to remain politically neutral. 

This chapter systematically explores variation in the use of the military in Presidential Campaign television advertisements. In so doing, clear variation is evident in how civilian candidates seeking to win office have featured the military in campaign advertisements in an attempt to sway voters. Furthermore, this chapter provides clear evidence that the military institution is becoming increasingly associated with partisan campaigns during election periods, and suggests that civilian candidates - and in some cases, former members of the military - have become increasingly willing to set aside the military's non-partisan ethic in order to maximize the likelihood that these candidates win office. This chapter argues that increasing political polarization over the past several years has rendered civilian actors less willing to abide by the principle of non-partisanship of the military institution, resulting in these actors engaging in political behaviors (campaign advertisements) that further challenge the principle of non-partisanship and other standards of healthy civil-military relationships.

This chapter also makes several contributions to the scholarly literature on the topics of civil-military relations, American politics, and international relations. First, this chapter provides yet another data point to advance the scholarly understanding of how political behavior involving the military institution has deviated from the normative principles of civilian control, non-partisanship, and the "non-interference" of the military. The chapter reveals that not only have military actors been impacted by political polarization, a development borne out in the previous chapter which examined retired officer opinion commentary, but civilian actors too have been impacted by polarization, resulting in civilians committing political behaviors which deviate from normative principles out of a desire to see their preferred candidate elected [see also @graham_democracy_2020]. A second contribution this chapter makes is that, engaging with the American politics literature on the topic of "issue ownership" [@petrocik_issue_1996; @petrocik_issue_2003; @sides_consequences_2007], it shows how both political parties have attempted to portray the military - to varying degrees - as supportive of its platform, initiatives, and ideas. Ultimately, the chapter suggests that civilian actors, who have the final say in the content of campaign advertisements, must be more willing than they currently are to avoid using the military in ways that place the military's non-partisan ethic at further risk. The failure to do so, even if it does not result in physical harm to the United States in the short term, can only lead to increasingly problematic civil-military relations in the long term.  

This chapter proceeds in three parts. First, I review the theory advanced previously in this dissertation regarding the impact of polarization and prestige on the ability of central principles of civil-military relations to constrain the behavior of both military and civilian actors. In doing so, I apply the theory directly to campaign advertisements, tying into the "issue ownership" [@petrocik_issue_1996; @petrocik_issue_2003; @sides_consequences_2007] literature within American politics. In the second part of the chapter, I present the data and outline the methodology used to test the hypotheses. Third, and before concluding, I raise and address potential confounding variables and alternate explanations that may also play a role. 

# Governing Principles of Civil-Military Relations

In this chapter, the focus is on civilian actors, whereas the previous chapter focused on a particular type of behavior committed by military actors - that of public opinion commentary authored by retired military officers. Deliberately focusing on civilian actors, and the political behaviors that they can commit which involve the military institution, is an important undertaking for several reasons. In the most basic sense, because the study of civil-military relationships involve both civilian and military actors, examining the behaviors of both sets of actors yields a richer understanding of the dynamics which shape these relationships. Furthermore, because the fundamental claim made in this dissertation is that polarization and military prestige comprise the variables which alter the constraining influence of principles that shape civil-military relations, examining the behaviors of both sets of actors in a deliberate fashion allows the researcher to subject these claims, and the predictive power that arises from these claims, to greater scrutiny.   

A key theme in this dissertation has been the importance of three fundamental, governing principles of American civil-military relations. While there is always some irreducible tension in civil-military relationships due to the very nature of these relationships [@feaver_civil-military_1996], a state's civil-military relationship tends to be relatively healthy and vibrant when the three principles of civilian control, non-partisanship, "non-interference" of the military into certain affairs of the state are adhered to by both sets of actors. One of the reasons this tends to be the case, I argued, is because these principles tend to constrain the behavior of both sets of actors in unique ways, in spite of ample historical evidence that shows there are nonetheless periodic challenges that surface with respect to civil-military relations.  

While these standards or principles have different application and precise meaning in certain country contexts, scholars generally conceptualize each of these standards not in dichotomous black and white terms, but rather along a gradual scale that varies. For this reason, describing civil-military relationships - in any context - usually involves a fair amount of nuance and description in relative terms. For example, scholars have noted that it is more helpful to assess whether the principle of civilian control of the military is healthy and vibrant in a given country or context less in concrete terms of "yes" or "no," but rather along lines that examine the _relative_ power between military and non-military actors, and by examining which actor has ultimate decision-making responsibility within a given political system [@bruneau_civil-military_2019; @brooks_shaping_2008]. 

In this dissertation, the focus is on US civil-military relationships, and in particular, determining which variables, if any, alter the ability of the three governing principles to constrain the behavior of civilian and military actors. The central argument of this dissertation is that two variables, that of political polarization and military prestige, impact the ability of the fundamental principles of civil-military relations to constrain the behavior of civilian and military actors. In making this claim, furthermore, I argued that increasing political polarization serves as the "motive" and increasing military prestige as the "opportunity" for civilian and military actors to buck the constraining influence of these principles, building off previous scholarship that likewise incorporates a "motives" and "opportunities" framework to explain, though not in the US context, military intervention into politics [@finer_man_1962; @taylor_politics_2003].

Polarization and military prestige work in unique but important ways to change the willingness of civilian and military actors to abide by the principles of civilian control, non-partisanship, and "non-interference." As the level of political polarization rises, the theory claims, the _ideological unity_ among civilian and military actors decreases. This stems from scholars who have argued that contemporary political polarization in the US involves not only disagreement about policy, which is an inherent feature of politics since the beginning of the world, but rather disagreement about the fundamental values, worldviews, and ideals that do and should govern a society [@deneen_why_2018; @mohler_jr_gathering_2020]. Furthermore, as the military's level of overall prestige rises, its ability to pose a "latent political influence" [@brooks_militaries_2009] or level of "political capital" [@albright_william_2006] also rises, referring to the _potential_ of the military to sway public opinion.

This dissertation also claims that _both_ polarization and military prestige need to be considered when determining the degree to which the governing principles of civil-military relations effectually constrain the behaviors of civilian and military actors. A total of four potential domestic environments are possible, each of which reflects a high versus low relative value of each independent variable. These environments are displayed in Table \@ref (tab:poss-environment) below.

```{r poss-environment, echo=FALSE, include=TRUE, error=FALSE, message=FALSE}

text_env <- tribble(~"Environment", ~"Low Polarization", ~"High Polarization",
  "Low Prestige", "Actors Ideologically United; Private Behaviors Involving the Military Are Likely", 
  "Actors Ideologically Divided; Private Behaviors Involving the Military Likely",  
  "High Prestige", "Actors Ideologically United; Public Behaviors Involving the Military Likely", 
  "Actors Ideologically Divided; Public Behaviors Involving the Military Likely" 
  )
 

  kable(text_env, booktabs=T, caption="Types of Domestic Environments for Civil-Military Relations and their Impact on Political Behaviors") %>%
  kable_styling(full_width = F, font_size=8, latex_options="striped") %>% 
  column_spec(1, width="10em") %>% 
  column_spec(2:3, width="20em") 
```

## Civilian Actors, Campaign Advertisements, and the Military

Examining campaign advertisements used in recent Presidential Elections provides a unique way to detect potential variation in the degree to which civilians are adhering to the principles of civil-military relations. Campaign advertisements, like military officer commentary, need not be problematic with respect to civil-military relations in every instance. After all, campaign advertisements, at least in democratic communities, comprise an integral part of the democratic process as candidates seek to win office. Moreover, like military officer commentary, campaign advertisements are inherently designed to shape public opinion. Whereas opinion commentary ostensibly seeks to sway public opinion over an important contemporary issue or event, campaign advertisements seek to sway the public to vote a certain way.

By combining the predicted impacts of polarization and prestige with respect to campaign advertisements, it is possible to fashion unique domestic political environments in which the military would feature, to varying degrees, throughout the course of political campaigns. In environments where military prestige is low, civilian actors are less inclined to appeal to the public using the military institution. When military prestige is low, there is less "demand" from the public to hear and see the nation's military, and thus, civilian actors, who wish to sway as many citizens as possible to vote a certain way, will not feature the military as often and as prominently than when military prestige is at higher levels. 

If and when military prestige is at relatively high levels, the corresponding degree of political polarization that prevails in society will determine if and how the military features in campaign advertisements. When political polarization is low, potential voters are less likely to look past behaviors that challenge or violate standards of civil-military relations. Thus, environments of high military prestige and low political polarization are likely to feature the military portrayed in campaign ads in relatively neutral ways. Alternately, when political polarization is high, potential voters are more likely to desire the outcome that corresponds with their preferred ideological slant over and above a process that respects and adheres to civil-military principles. Therefore, when military prestige and political polarization is high, we can expect campaign advertisements to prominently feature the military to the public, but in ways that push the boundaries or limits of civil-military principles. These predictions lead more generally to the formation of the following hypotheses:

## Hypotheses

The preceding theoretical predictions lead to three distinct hypotheses, which are listed below:

H4: When military prestige is high, civilian leaders will increasingly attempt to leverage or harness the military (either as an institution, or individual military figures who possess outsized levels of prestige/popularity) for their own political purposes and advantages relative to when military prestige is low. 

H5: When polarization is high, civilian leaders will employ the military (either as an institution, or individual military figures who possess outsized levels of prestige/popularity) to support (or denounce) policy initiatives that align with (or oppose) underlying ideologies associated with a polarizing issue or cause relative to when polarization is low. 

H6: When polarization and military prestige are high, civilian political parties will incresingly compete for the military institution and in increasingly public ways relative to when polarization and military prestige are low. 

# Methodology and the Data

In this chapter, I seek to conduct a quantitative analysis combined with content analysis of Presidential campaign advertisements. The Wisconsin Advertising Project (WiscAD) and Wesleyan Media Project (WMP) are two entities that have systematically tracked campaign advertising beginning in the 1996 Presidential Election.^[Unfortunately, the raw data used to code the 1996 election is no longer available to review; hence, this study begins with the 2000 Presidential Election. Additionally, raw data for the 2020 Election is not available at the time of this writing, but is projected to become available sometime in 2022.] Both WiscAD and WMP gather the television advertisements with the assistance of a commercial company, the Campaign Media Analysis Group (CMAG), which provides several pieces of information when releasing the data, such as the media market in which a particular advertisement aired, the estimated cost of running a particular advertisement, the time of day the advertisement was run, etc. 

WiscAD and the WMP then gather additional information on the advertisements, which varies to some extent by election cycle, but otherwise remains markedly similar. These information includes items such as the specific issues that an advertisement mentions or addresses during the advertisement, whether a candidate is featured during the advertisement, and whether a given advertisement takes on a negative tone, and more (see a future chapter Appendix for a comparison, by year, of media market breakdowns). 

The overall methodological goal of this chapter is to determine if and how there has been changes in the way that the military has featured during these campaign advertisements over the given time period. Specifically, I am interested in capturing any changes to how the military has been featured in ways that challenge or expressly violate two of the three governing principles of civil-military relations: non-partisanship of the military institution, and the "non-interference" of the military into certain civilian affairs. I am also interested, however, in capturing any changes to the civilian actors themselves that are using the military in ways that challenge these norms. The world of campaign advertising has indeed grown increasingly complex: political parties, candidates, and other interest groups all generate political advertisements of different varieties. It is therefore important to capture variation not only in how the military is being used in campaign ads over time, but also potential variation in which civilian actors, if any, are driving these changes. Such changes, moreover, would be of particular use in tying into the "issue ownership" scholarship of American politics [@petrocik_issue_1996; @petrocik_issue_2003; @sides_consequences_2007].

## Classification, Coding, and Empirical Strategy

I propose conducting a three step empirical strategy. First, I have to determine which advertisements to view and code, capturing at the same time any important existing variables to apply to later analysis. Second, I have to watch the applicable advertisements and code them for variables that I think capture the potential variation that I am interested in capturing. Third, I then have to conduct analysis consisting of descriptive statistics and several regression models to determine the impact, if any, of polarization and prestige in influencing any relevant dependent variables. I describe this strategy in greater detail in the paragraphs to follow. 

First, using data coded by the WiscAD and WMP, I determine which ads in each election cycle address or cover topics that are the most likely to feature military symbols, imagery, or personnel. In briefly reviewing the data on this, there are two possible ways to conceptualize this (maybe more?). One way, which is a relatively "narrow" option, is to examine in each cycle the ads that are coded for addressing matters of "military/defense" or "veterans" issues. Another way, the relatively "wider" option, is to include these issues but to also include any exogenous topics that might also feature the military in some way. These exogenous issue areas vary by year, but include issues such as "September 11th" or "Afghanistan" or "Iraq." The number of advertisements per election cycle which fall into relatively "narrow" versus "wide" categories are listed below in Table \@ref(tab:el-ads-numeric).

```{r el-ads-numeric, include=TRUE, echo=FALSE}

text_tbl_ads <- tribble(~"Campaign Year", ~"Narrow Value", ~"Wide Value",
  "2000", "83", "89",
  "2004", "103", "164",
  "2008", "213", "254",
  "2012", "30", "57",
  "2016", "117", "200",
  "Total Values", "546","764"
  )
 

text_tbl_ads %>% kable(booktabs=T, caption="Coding Scheme for WiscAd and WMP Campaign Advertisements") %>%
  kable_styling(full_width = F, font_size=8, latex_options="striped" ) %>% 
  column_spec(1, width="10em") %>% 
  column_spec(2, width="15em") %>% 
  column_spec(3, width="10em") 


```

The second step is to watch these advertisement and to code them for several additional variables which capture the changes predicted by the hypotheses previously outlined. These variables are mostly dichotomous or categorical and are displayed in Table \@ref(tab:el-ads-additional).

```{r el-ads-additional, include=TRUE, echo=FALSE}

text_vars <- tribble(~"Variable Name", ~"Question", ~"Possible Values",
  "Can_Vet", "Is the candidate a Veteran?", "1 - Yes; 0 - No",
  "Can_Vet_Mention", "Does the ad explicitly state or portray the candidate as a Veteran?", "1 - Yes, 0 - No",
  "Vet_Fam", "Does the ad feature a Veteran or Veteran's family speaking?", "3 - Veteran and Family; 2 - Veteran's Family; 1 - Veteran; 0 - No",
  "Vet_Fam_Endorse", "Does the speaking Veteran or Veteran's Family explicitly endorse the President for office?", "1 - Yes; 0 - No",
  "Vet_Fam_Attack", "Does the speaking Veteran or Veteran's Family attack or disparage other candidates running for office?", "1 - Yes; 0 - No",
  "Vet_Fam_Issues", "Does the speaking Veteran or Veteran's Family address other topics outside of supporting the candidate?", "3 - Other Issues (List); 2 - National Security or Defense Issues; 1 - Veterans; 0 - No",
  "Act_Duty_Imagery", "Does the advertisement feature imagery of active duty troops?", "1 - Yes; 0 - No",
  "Vet_Imagery", "Does the advertisement feature imagery of Veterans?", "1 - Yes; 0 - No",
  "Service_Imagery", "Does the advertisement feature military service imagery (planes, tanks, past images of combat, etc.?", "1 - Yes; 0 - No",
  "General_Ad_Theme", "Is the ad itself an ad focused on the military, defense, or national security, or are the military/defense themes in the ad one part of an advertisement that addresses other topics?", "1 - military or defense only; 0 - other topics"

  )
 

text_vars %>% kable(booktabs=T, caption="Coding Scheme for WiscAd and WMP Campaign Advertisements") %>%
  kable_styling(full_width = F, font_size=8, latex_options="striped" ) %>% 
  column_spec(1, width="10em") %>% 
  column_spec(2, width="15em") %>% 
  column_spec(3, width="10em") 
```

The third and final step is to present descriptive statistics, and to run a series of regressions using the relevant variables and data. The descriptive statistics will show whether there has been variation along several of the important variables. For instance, the coding scheme will detect if there has been an increase in the number or proportion of advertisements which feature military veterans speaking and endorsing candidates during advertisements, or whether veteran families have been featured more often in recent years as polarization has worsened. The descriptive statistics will also capture whether advertisements have increasingly incorporated certain types of military imagery, such as imagery of active duty troops.

Then, I will run a series of regressions in which the dependent variables are several of the variables listed in Table \@ref(tab:el-ads-additional). I will include, as control variables, various measurements of polarization and military prestige, and determine the degree to which there is association between these variables.  

# Alternate Explanations and Potential Issues

As I dive into this research agenda, it is important to maintain awareness of what else - besides polarization and military prestige - could drive changes in the way that civilian actors feature the military in campaign advertisements. I think there are at least two likely culprits of alternate explanations that critics of the approach laid out in this proposal might point to. These culprits are external events, and overall changes in the media environment. 

Some critics might say that exogenous events, such as relatively unpopular war in Iraq at the time of the 2004 Presidential Election, might drive potential ways in which civilian candidates portrayed the military during campaign advertisements. To a certain extent, I think that these critics would be correct. However, I think that while external events might shape the _prominence_ or _salience_ of certain topics featured in certain campaigns over other topics - a completely reasonable expectation, in my opinion - these external events would not drastically alter how close campaign advertisements challenge or violate civil-military norms themselves. Therefore, a key consideration in conducting this analysis is really arguing for or against which variables listed in Table \@ref(tab:el-ads-additional) measure certain things. For example, we might expect the variable "Service_Imagery" to increase in 2004 at the height of the Iraq War, as advertisements might show images of the combat taking place, tanks, planes, etc. But other variables, such as "Vet_Fam_Issues", would not likely by itself increase as a result of the War. I suspect that there is no clean way around this confounder, except to strongly state my reasoning as appropriate and acknowledging that there will always be, to some degree, the issue of exogenous events that drive the salience of certain topics during elections. 

A second alternate explanation, that of changes in the media environment in general driving variation in how the military is featured in campaign advertisements, might also be something that a critique would point out as being important. On this point, too, I share this concern. However, at some point, I also feel as if polarization - at least how I have described it in the theory section of this dissertation - speaks to these changes in the media environment. In other words, we know that polarization has impacted a lot of things, to include the media. Thus, while my argument in this dissertation focused on how polarization impacts the willingness of actors to abide by civil-military principles, it does not discount the fact that polarization has also impacted much more.  Still, I think it is important that I describe that this is an important consideration, but one that I likely will be unable to parse out in a more meaningful matter, at least in this chapter. 

# Conclusion

A key argument made previously in this dissertation is that polarization and prestige significantly influence the way in which civilian and military actors abide by fundamental principles of civil-military relations, and that as a result, these variables also change how these actors commit certain behaviors that involve the military institution. Campaign advertisements are one type of political behavior committed by civilian actors that are a regular feature of the political landscape and the democratic process. Presidential campaign television advertisements are displayed every four years in an attempt to persuade voters to vote for a certain candidate over another. While these advertisements do not necessarily have to feature the military at all, nor necessarily in ways that challenge the military's non-partisan ethic and a principle of "non-interference" by the military into certain state affairs, recent political campaigns have demonstrated that certain advertisements can, in fact, be problematic along these lines. 

Conducting a systematic investigation of campaign advertisements for the five Presidential Elections from 2000-2016 will reveal potential variation in how the military has featured in these advertisements. In this document, I have proposed a research agenda and an empirical strategy for collecting the data which will ultimately display any potential variation in this arena to scholars, and important contribution in itself, and determine the degree to which political polarization and military prestige have driven these changes. 

\singlespace

\newpage


