---
title: "True Colors? The Military Institution and Campaign Advertisements"
author: "Peter M. Erickson"
date: |
  `r format(Sys.time(), '%B %d, %Y')`
abstract: |
  
  _Using data from the Wisconsin Advertising Project and the Wesleyan Media Project consisting of Presidential Campaign advertisements used in five Presidential Elections from 2000 -2016, this research proposal seeks to conduct a large-n study to determine whether there has been variation in the ways that campaign advertisements have featured the military over time. This proposal will form the basis of a chapter of my upcoming dissertation, and will argue that the military's relatively high prestige, combined with political polarization that has dramatically increased over the time interval examined, has led to civilian actors using the military institution in ways that expressly challenge fundamental standards of American civil-military relations, to include the non-partisanship of the military institution._  
  
geometry: margin = 1.15in              # margin widths
fontsize: 12pt                         # text size
bibliography: dissertation.bib         # bibliography file
biblio-style: authoryear               # citation style
urlcolor: blue                         # links to external URLs
citecolor: black                       # links to bibliography
linkcolor: magenta                     # internal figures, footnotes, etc.
indent: true                          # paragraph format
                                       # true for indented paragraphs
                                       
header-includes:
    - \usepackage{setspace}
    - \usepackage{multirow}
    - \usepackage{indentfirst}
   
    
output: 
  bookdown::pdf_document2:
    keep_tex: true
    latex_engine: pdflatex
    number_sections: false
    toc: false
    fig_caption: true
    citation_package: biblatex
    includes: 
      in_header: 
        - latex-preamble.tex
  bookdown::htmldocument2: default
  bookdown::word_document2:
      toc: false
---

\doublespacing 

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo=FALSE,
  include=FALSE,
  message=FALSE,
  error=FALSE,
  fig.height=3,
  fig.width=4,
  out.width="90%")
```

```{r, include = FALSE, echo = FALSE}
library("here")       # file paths
library("tidyverse")  # workhorse package
library("tidylog")
library("kableExtra")
library("knitr")
library("ggdag")
library("dagitty")
library("gridExtra")
library("broom")
library("dplyr")
library("reshape2")
library("pacman")
library("ggrepel")
```

\newpage

# Introduction

On October 10, 2020, then candidate Joe Biden's campaign team aired an advertisement touting the former Vice President's national security credentials, including his experience working with the military as both Vice President and before that, US Senator from Delaware. The advertisement is narrated by a man identified as Jim, an "Iraq War Veteran," who describes the role of the former Vice President in delivering Mine Resistant Ambush Protected (MRAP) vehicles, whose heavy armor helped protect servicemen and women while deployed to Iraq and Afghanistan. Within days, however, the advertisement was pulled by the Biden Campaign after a retired Army Lieutenant General, Sean McFarland, objected to the advertisement on the grounds that the ad used images of him without his consent [@moore_biden_2020]. These images showed the general, then in uniform and on active duty, shaking hands with Biden as the former Vice President arrived to Iraq [@moore_biden_2020]. 

In objecting to the advertisement, MacFarland noted that he had not granted permission to the Biden Campaign to use these images in the advertisement, but also added that in his opinion, it is normatively wrong for the military - especially active duty forces - to appear in campaign advertisements altogether, as doing so jeopardizes the military's non-partisan ethic. Indeed, MacFarland expressed these sentiments in a public statement after the advertisement aired: "A number of people have understandably inferred that my appearance [in this Biden Campaign advertisement] constitutes an endorsement of the former Vice President. It does not. To be clear, I have not endorsed President Trump, either. I object to the use of ANY military personnel in uniform in political ads - full stop" [MacFarland quoted in @moore_biden_2020]. 
This incident serves as one of several during the 2020 Presidential Election in which political candidates used the military in ways that push against, challenge, and in some cases, violate standards that govern the non-partisan ideals of the military.^[President Trump's Campaign likewise used military symbols and imagery inappropriately, and perhaps on more than one occasion. For a salient example in which the Chairman of the Joint Chiefs of Staff appeared in one of Trump's advertisements, see @seligman_top_2020.] And while civilians have long appealed to the public by harnessing the popularity of the military or of individual military figures in a variety of ways, the practice of doing so during election seasons highlights the ways in which polarized politics challenge the military and its members in seeking to remain politically neutral. 
In this research proposal, I lay out a research agenda to determine whether, and to what degree, there has been variation in the way in which civilian presidential candidates have featured the military during campaign advertisements over time. It is an interesting question to consider, because campaign advertisements, unlike commentary authored by retired military officers (the focus of the previous chapter in the dissertation), fall under the ultimate purview of civilian actors. If there is, in fact, notable variation in the ways that the military as an institution has been featured in the course of Presidential campaign ads, it is important to then understand why these changes have occurred. 

Ultimately, the final version of the chapter for which this research proposal is based will seek to argue that the military institution is becoming increasingly associated with partisan campaigns during election periods, and that civilian candidates - and in some cases, former members of the military - have become increasingly willing to set aside the military institution's non-partisan ethic in order to maximize the likelihood that specific candidates win office (or other specific candidates lose office). The final version of the chapter will also argue that the twin factors of increasing political polarization and the prestige of the military have combined such that civilian actors are less willing to abide by the principle of non-partisanship of the military institution while harnessing a popular institution (the military) in an attempt to win voters.

This chapter also makes several contributions to the scholarly literature on the topics of civil-military relations, American politics, and international relations. First, this chapter provides yet another data point to advance the scholarly understanding of how political behavior involving the military institution has deviated from the normative principles of civilian control, non-partisanship, and the "non-interference" of the military - principles that this dissertation has argued are critical to maintaining healthy civil-military relationships. The probable findings of this chapter will likely demonstrate that not only have military actors been impacted by political polarization, a finding confirmed in the previous chapter examining retired officer opinion commentary, but also that civilian elites (Presidential candidates running for office) have likewise been impacted by polarization such that civilian actors are committing behaviors that deviate from normative principles out of a desire to win office, a finding that will broaden our understanding of how actors are increasingly willing to overlook normative standards out of a desire to see his or her preferred candidate win office  [see also @graham_democracy_2020]. A second likely contribution this chapter will make involves the American Politics literature on the topic of "issue ownership" [@petrocik_issue_2003; @sides_consequences_2007]. Specifically, I suspect that this chapter will provide clear evidence that _both_ mainstream American political parties have attempted to portray the military - to varying degrees - as supportive of its platform, initiatives, and ideas. Ultimately, this chapter will likely suggest that civilian actors, who have the final say in the content of campaign advertisements, must be more willing than they currently are to avoid using the military in ways that place the military's non-partisan ethic at further risk. The failure to do so, even if it does not result in physical harm to the United States in the short term, will lead to increasingly problematic civil-military relations in the long term. 

This research proposal proceeds in three parts. First, the proposal briefly reviews the theory advanced previously in the dissertation, which argued that the respective levels of political polarization and the prestige of the military impact the ability of three central principles of civil-military relations to constrain the behavior of both military and civilian actors. In doing so, this proposal applies this theory through the lens of the topic of "issue ownership" [@petrocik_issue_2003; @sides_consequences_2007]. In the second and main part of this research proposal, I present initial hypotheses, data, and otherwise outline the methodology that I plan to use in crafting the full chapter. Emphasis is placed on describing what variables I plan to code for, what regressions I would run, and otherwise explaining what evidence I need to present in order to test my hypotheses. In the third and final section of this proposal, I raise potential confounding variables and address alternate explanations that could challenge my anticipated results.   

# Governing Principles of Civil-Military Relations

One of the key themes in this dissertation has been the importance of three governing principles of American civil-military relations. While there is always some irreducible tension in civil-military relationships due to the very nature of these relationships [@feaver_civil-military_1996], a state's civil-military relationship tends to be relatively healthy and vibrant when the three principles of civilian control of the military, non-partisanship of the military, and the "non-interference" of the military into certain affairs of the state are adhered to by both sets of actors.^[For a full definition of these principles, please see Chapter 1 of the dissertation.] These three principles also tend to constrain the behavior of both sets of actors, military and civilian, in unique ways, in spite of ample historical evidence that shows the periodic and sometimes intense challenges that have surfaced with respect to civil-military relations. These principles tend to constrain the behavior of both military and civilian actors for a variety of reasons that involve both the American public's expectations, as well as the realization of both sets of actors that more harmonious outcomes with respect to the nation's security are achieved when these principles are generally adhered to, followed, and respected.  

This dissertation has also emphasized that each of the three standards or governing principles have a different application and precise meaning in certain country contexts. For this reason, the degree to which these principles are respected, and thus, the constraining influence these principles have on the actors involved in civil-military relations, changes significantly cross-nationally. Nonetheless, scholars have generally conceptualized each of the governing principles or standards not in dichotomous black and white terms, but rather along gradual scales that allow for a rich and nuanced classification of a country's civil-military relations. For example, scholars have noted that it is more helpful to assess whether the principle of civilian control of the military is healthy and vibrant in a given country or context less in concrete terms of "yes" or "no," but rather along lines that examine the _relative_ power between military and non-military actors, and by examining which actor has ultimate decision-making responsibility within a given political system [@bruneau_civil-military_2019; @brooks_shaping_2008]. 

This dissertation then argued that in the American context, the variables of political polarization and military prestige uniquely impact the ability of the principles of civil-military relations to constrain the behavior of civilian and military actors. Furthermore, in making this claim, this dissertation argued that increasing political polarization serves as the "motive" and increasing military prestige as the "opportunity" for civilian and military actors to buck the constraining influence of the principles of civilian control, non-partisanship of the military, and the "non-interference" of the military institution into realms of state politics. Moreover, in assigning the variable of political polarization the role of "motive" and the variable of military prestige the role of "opportunity," this dissertation follows previous scholarship that likewise incorporates a "motives" and "opportunities" framework to explain, though not in the US context, military intervention into politics [@finer_man_1962; @taylor_politics_2003].

The theory chapter of this dissertation argued that polarization and military prestige work in unique but important ways to change the willingness of civilian and military actors to abide by the principles of civilian control, non-partisanship, and "non-interference." The theory specifically argued that as the level of political polarization rises, the _ideological unity_ among civilian and military actors decreases. This stems from scholars in the American Politics sub-field and from political theorists who have argued that contemporary political polarization in the US involves not only disagreement about policy, but rather disagreement about the fundamental values, worldviews, and ideals that do and should govern a society [@deneen_why_2018; @mohler_jr_gathering_2020]. 

Furthermore, the theory posited earlier in the dissertation argued that as the military's level of overall prestige rises, its ability to pose a "latent political influence" [@brooks_militaries_2009], or the military's ultimate level of "political capital" [@albright_william_2006], also rises. In other words, as the level of the military's prestige rises, the military's ability to potentially impact public opinion and to engage in public versus private behaviors, increases as well. Thus, a highly prestigious military holds the potential to enact political costs on civilian leaders, appeal to the public, and otherwise shape the world of politics, if it so chooses. 

Finally, the theory posited earlier in this dissertation claimed that _both_ political polarization and military prestige need to be considered when determining the degree to which the governing principles of civil-military relations effectually constrain the behaviors of civilian and military actors. The theory argued that just because one variable exists at a relatively high level, the other variable might be relatively low, and vice versa. In all, a total of four potential domestic environments are possible, each of which reflects a high versus low relative value of each independent variable. These environments are displayed in Table \@ref(tab:poss-environment) above.

```{r poss-environment, echo=FALSE, include=TRUE, error=FALSE, message=FALSE}

text_env <- tribble(~"Environment", ~"Low Polarization", ~"High Polarization",
  "Low Prestige", "Actors Ideologically United; Private Behaviors Involving the Military Are Likely", 
  "Actors Ideologically Divided; Private Behaviors Involving the Military Likely",  
  "High Prestige", "Actors Ideologically United; Public Behaviors Involving the Military Likely", 
  "Actors Ideologically Divided; Public Behaviors Involving the Military Likely" 
  )
 

  kable(text_env, booktabs=T, caption="Types of Domestic Environments for Civil-Military Relations and their Impact on Political Behaviors") %>%
  kable_styling(full_width = F, font_size=8, latex_options="striped") %>% 
  column_spec(1, width="10em") %>% 
  column_spec(2:3, width="20em") 
```

## Civilian Actors, Campaign Advertisements, and the Military

In this chapter, the focus is on a type of behavior committed by civilian actors, whereas the previous chapter focused on a particular type of behavior committed by military actors - that of public opinion commentary authored by retired military officers. Deliberately focusing on civilian actors, and the political behaviors that they can commit which involve the military institution, is an important undertaking. 

In the most basic sense, because the study of civil-military relationships involve both civilian and military actors, examining the behaviors of both sets of actors yields a richer understanding of the dynamics that shape these relationships. Specific to this dissertation, because the fundamental claim is that polarization and military prestige comprise the variables which alter the constraining influence of principles that shape civil-military relations, it is necessary to examine the behaviors of _both_ sets of actors in a deliberate fashion. Doing so allows the researcher to subject the theory made in this dissertation, and the predictive power that arises from the theory's claims, to greater scrutiny. Having subjected military actors and the behavior of public opinion commentary to intense scrutiny in the previous chapter, I now turn to civilian actors and the behavior of campaign advertisements. 

While the behavior of crafting and airing campaign advertisements are not the same as retired officer opinion commentary, the two behaviors are similar in important ways that are worth highlighting, particularly for the purposes of showcasing why campaign advertisements make for a good empirical study of a type of behavior conducted by civilian actors. First, like military officer commentary, campaign advertisements need not be problematic with respect to civil-military relations. In other words, there is no _a priori_ expectation that campaign advertisements would, or must, violate or challenge the principles of civilian control, non-partisanship, and "non-interference" of the military. After all, campaign advertisements, especially in democratic communities, comprise an integral part of the democratic process as candidates seek to win office at regularly scheduled intervals. Second, like military officer commentary, campaign advertisements are inherently designed to shape public opinion. Whereas opinion commentary ostensibly seeks to sway public opinion over an important contemporary issue or event, campaign advertisements seek to sway the public to vote a certain way, either by supporting or opposing a particular candidate, or supporting or opposing a particular legislative action, or both. And third, like retired officer opinion commentary, campaign advertisements are a regular and reoccurring feature of political life. Just as most contemporary newspapers run opinion pieces or op-eds every day or week, every Presidential Election in American history has involved efforts by candidates to advertise, albeit through different forms, mediums and technological avenues. 

For all of these reasons, a rigorous empirical analysis of campaign advertisements over time may show variation in the degree to which the military has played a part, or featured, in these advertisements. If there is variation, then such variation can and should be explained. In short, examining campaign advertisements used in recent presidential elections provides a unique way to detect potential variation in the degree to which civilians are adhering to the fundamental principles of civil-military relations.

How would one know if civilians are _not_ adhering to, and instead, are challenging if not violating, the fundamental principles of civil-military relations through the content of campaign advertisements? Earlier in this dissertation, I argued that civilian actors cannot, by definition, violate the principle of civilian control, but they can violate or challenge the other two principles: that of non-partisanship of the military, and the "non-interference" of the military into certain realms of state policy. 

A violation of the principle of non-partisanship of the military vis-a-vis campaign advertisements would occur if an advertisement clearly intimated that the military favored or endorsed a particular candidate over another, or the policies or political stances of a candidate over another, or a political party over another. Such a message, deliberate or not, could and likely would communicate to the American public that the military (or a faction of the military, or particular leaders within the military) favors a particular political entity over another, in direct conflict with a legion of scholars, practitioners, and former military officers who all have emphasized the importance of maintaining the military's non-partisan ethic [@huntington_soldier_1957]. 

A violation of the principle of "non-interference" of the military into certain realms of state policy, again expanding on the definition of such a principle from Chapter 1 of this dissertation, would occur vis-a-vis campaign advertisements if the military was portrayed in such a way that it weighed in on policy arenas that are largely disconnected from national security, defense, and otherwise other essential roles the military is expected to fill in American society [@ben-meir_civil-military_1995]. 

There are myriad ways in which campaign advertisements _could_ challenge or violate these two principles, but a simple example will likely make the violations of each respective principle more clear. Suppose that an active duty military officer, in uniform, appeared in a campaign advertisement and stated, "I support Candidate X for President, because Candidate X will be better than Candidate Y as President." The military officer appearing in uniform and stating those words would clearly communicate the officer's favoring of one partisan political candidate over that of another, and thus, constitute a violation to the non-partisan aspirational nature of the military as an institution. However, if the officer also said, in addition to the above endorsement of the political candidate, the words, "In addition, Candidate X has a better tax plan that will stimulate the economy, and a better plan to improve our public education system than Candidate Y," the advertisement would also violate the principle of "non-interference." By publicly weighing in on the issues of national wealth (taxes) and public education, the officer has weighed in on two issues that, at least in the US context, are clearly outside of the quiver of issues for which the US public expects the military to fill an influential role by providing relevant expertise. 

To be clear, my argument is that if the officer had instead uttered the words, "In addition, Candidate X's plan for withdrawing US Troops from the Middle East is better than Candidate Y's" after endorsing Candidate X, such behavior still constitutes an inappropriate violation of the principle of non-partisanship of the military, but not a violation of the principle of "non-interference." However, for the purposes of this study, the fact that an advertisement only violated one vice two central principles of civil-military relations should not be interpreted as being any less severe of a violation. The important thing to keep in mind is that both instances go against the normative, fundamental principles of civil-military relations, albeit different principles and in different ways. 

Before proceeding, one word of caution is warranted regarding civilian actors, the potential principles of civil-military relations, and campaign advertisements. The world of political advertisements has become increasingly complex in the sense that today, campaign advertisements and political advertisements more broadly are created and aired by a diverse set of candidates, interest groups, and other coalitions. As the data in the following section will likely show, advertisements sponsored by veteran-affiliated interest groups have also proliferated over the past several elections. Although these veteran-affiliated interest groups are not run or chaired by sitting active duty military members, I argue that because of their direct association with the military, we should expand the aperture of potential principles of civil-military relations to be violated through campaign advertisements to include that of civilian control. In other words, it is my contention that although civilian candidates cannot violate the principle of civilian control by definition, interest groups that are directly associated with the military and who claim or portray to the public to represent the military, can. Thus an advertisement aired by a veteran-affiliated interest group that sharply criticizes a particular civilian candidate or that candidate's policy stances has, I argue, violated the principle of civilian control.

## Predictions and Hypotheses

Having briefly reviewed the relevant literature and summarized the theory posited earlier in the dissertation, this research proposal now examines the predicted impacts of political polarization and military prestige on adherence to the standards of civil-military relations with respect to political campaign advertisements. Using Table \@ref(tab:poss-environment) as a guide, it is possible to fashion four unique domestic political environments in which the military would feature, to varying degrees, throughout the course of political campaigns. 

In environments where military prestige is low, civilian actors are less inclined to appeal to the public using the military institution. Furthermore, when military prestige is relatively low, there is less "demand" from the public to hear and see the nation's military, and thus, civilian actors, who wish to sway as many citizens as possible to vote a certain way, will not feature the military as often and as prominently than when the military's prestige is relatively high. 

The level of political polarization present in society is also important. However, if military prestige is low, it is unlikely that civilians will deliberately choose to feature the military in campaign advertisements, regardless of the level of political polarization. 

However, when military prestige is relatively high, civilian actors are more likely to seek to feature the military during campaign advertisements. When the military is relatively respected and trusted as an institution by the public, the military holds a greater _potential_ to impact public opinion [@albright_william_2006; @brooks_militaries_2009]. Thus, it is reasonable to predict that the military will be more prevalent, broadly speaking, in campaign advertisements when the relative level of military prestige is high rather than low. 

And yet, the level of political polarization is important with respect to _how_ the military is featured during campaign advertisements (when the level of military prestige is high). When political polarization is low, potential voters are less likely to look past behaviors that challenge or violate the central principles of civil-military relations. Additionally, the political environment itself is relatively less charged when polarization is low. Thus, environments of high military prestige and low political polarization are likely to feature the military portrayed in campaign ads in relatively neutral ways. Therefore, when political polarization is low and military prestige is high, it is unlikely to see campaign advertisements that feature the military in ways that violate principles of non-partisanship or "non-interference" of the military.  

Alternately, when political polarization is high, potential voters are more likely to overlook instances of behavior that challenge principles of civil-military relations as voters instead desire the outcome (their preferred candidate winning office) that corresponds with their ideological slant over and above a process that respects and adheres to civil-military principles [@graham_democracy_2020]. Moreover, the political environment itself is highly charged when polarization is relatively high, increasing the likelihood of hard fought and bitter political campaigns between candidates seeking office. Thus, environments of high military prestige and high political polarization are likely to feature the military portrayed in campaign ads in non-neutral ways. In this type of environment, we would expect to see the military in campaign advertisements in ways that violate principles of non-partisanship or "non-interference" of the military. 

## Hypotheses

These predictions lead more generally to the formation of the following hypotheses:

H4: When military prestige is high, civilian leaders will attempt to leverage or harness the military (either as an institution, or individual military figures who possess outsized levels of prestige/popularity) for their own political purposes and advantages relative to when military prestige is low.

H5: When polarization is high, the level of military prestige moderates the propensity of civilian leaders to employ the military (as an institution, or individual military actors) in public ways to support (or denounce) policy initiatives or political candidates. When polarization is high and military prestige is high, civilian actors will use the military in public ways to advocate publicly for policies and/or partisan political candidates. When polarization is high and military prestige is low, however, civilians actors are less inclined to use the military to publicly advocate for policies and/or partisan political candidates.  

H6:  When polarization and prestige are high, there is increased public competition between the civilian political parties for the military institution relative to when polarization and military prestige are low. 

# Methodology and the Data

The data used in this chapter consist of presidential campaign advertisements from the 2000 - 2016 elections. ^[Unfortunately, the raw data used to code the 1996 election is no longer available to review; hence, this study begins with the 2000 Presidential Election. Additionally, raw data for the 2020 Election is not available at the time of this writing, but is projected to become available sometime in 2022.] The data is made available from the Wisconsin Advertising Project (WiscAD) and Wesleyan Media Project (WMP), which are two entities that have systematically tracked campaign advertising beginning with the 1996 Presidential Election. For 2012 and 2016, the data comes from WMP, and consists of full motion video, to include sound [@fowler_presidential_2017; @fowler_presidential_2020]. For 2008, 2004, and 2000, the data comes from WiscAd, and consists of a PDF document containing a full transcript of the ad as well as several images of the advertisement [@goldstein_political_2002; @goldstein_presidential_2007; @goldstein_presidential_2011]. 

WiscAD and WMP gather the television advertisements with the assistance of a commercial company, the Campaign Media Analysis Group (CMAG). CMAG provides several pieces of information when releasing the data, such as the media market in which a particular advertisement aired, the estimated cost of running a particular advertisement, the time of day the advertisement was run, and other variables. 

A group of scholars and graduate students affiliated with WiscAD and WMP then conduct additional coding on the advertisements. This coding varies to some extent by election cycle, but otherwise remains markedly similar. The additional variables that are coded for in each election cycle include items such as the specific issues that an advertisement mentions or addresses during the advertisement, whether a candidate is featured during the advertisement, and whether a given advertisement takes on a negative tone, and more (see a future chapter Appendix for a comparison, by year, of media market breakdowns). 

The overall methodological goal of this chapter is to determine if and how there has been changes in the way that the military has featured during these campaign advertisements over the given time period. Specifically, I am interested in capturing any changes to how the military has been featured in ways that challenge or expressly violate the three governing principles of civil-military relations: civilian control of the military, the non-partisanship of the military institution, and the "non-interference" of the military into certain civilian affairs. 

In order to proceed in answering such questions, a three part empirical strategy is necessary. First, I have to select the right data from which to conduct further coding. That is, it is necessary to select data that features or is likely to feature the military. Second, the empirical strategy must then code for key variables that measure the variation that I wish to capture. That is, I require a coding classification scheme that can capture potential challenges or violations to the three principles of civilian control, non-partisanship of the military institution, and "non-interference" of the military. The third and final portion of the empirical strategy consists of the analytic steps to determine whether there has been, in a statistical sense, variation over time in the way that the military has featured in campaign advertisements. This portion includes the descriptive statistics and regression analyses used to convey the results of the coding and classification scheme. In the section to follow, I detail these procedures. 

## Data Selection

The first step of the empirical strategy consists of data selection. In order to accomplish this step, I need to determine which ads that I need to watch in order to conduct further coding. For this to occur, I first determine which advertisements by election cycle are the most likely to feature the military in any way, shape, or form, such as military symbols, images, scenes, or personnel. 

In order to do this, I harness the issue classification coding already done by coders at WiscAd and WMP. Coders from WiscAd and WMP coded each advertisement for a series of dozens of dichotomous indicators, such as, "are any of these issues mentioned in the ad?" [@goldstein_presidential_2011]. These issues vary every election cycle, but many of the same issues are found across multiple election cycles. For each election cycle, I seek to cast a wide net, and identify the issues that are most strongly correlated with the military and national defense more broadly, and the individual advertisements that mention at least one of these issues. I look for issues include items such as the military and defense, veterans affairs, China, Iran, terrorism, September 11, the war in Iraq, Afghanistan, the surge, and a host of other issues. Table \@ref(tab:election-issues-cycle) shows the issues by election cycle that I deem as being highly correlated with a probable mention or feature of the military in some way.

```{r election-issues-cycle, include=TRUE, echo=FALSE}


text_tbl_issues <- tribble(~"Campaign Year", ~"Issues Highly Correlated with the Military", ~"Number of Issue Areas or Explicit Mentions",
  "2000", "Defense; Missile Defense/Star Wars; Veterans; Foreign Policy; Bosnia; China; Other Defense/Foreign Policy Issues", "7 Issue Areas",
  "2004", "Defense; Missile Defense/Star Wars; Veterans; Foreign Policy; Bosnia; China; Terrorism; Middle East; Afghanistan; September 11", "10 Issue Areas plus two explicit mentions (September 11 and terrorism)",
  "2008", "Defense Policy; Veterans; September 11; Terrorism; Iraq; Surge; Middle East; Defense Issue NOT related to Iraq; Iraq as an Issue; China; Nuclear Weapons; Iran; Afghanistan; Other Defense Issue NOT related to Iraq", " 10 Issue Area plus four explicit mentions (September 11, Terrorism, Surge, Iraq)",
  "2012", "Military; Foreign Policy (generic); Veterans; Nuclear Proliferation; China; Middle East; Afghanistan; September 11; Terrorism; Iraq; Israel; Iran", "12 Issue Areas",
  "2016", "Military; Foreign Policy (generic); Veterans; Nuclear Proliferation; China; Middle East; Afghanistan; September 11; Terrorism; Iraq; Israel; Iran; ISIL or ISIS; Syria", "14 Issue Areas"
  )

text_tbl_issues %>% kable(booktabs=T, caption="Selecting Advertisements For Further Coding") %>%
  kable_styling(full_width = F, font_size=8, latex_options="striped" ) %>% 
  column_spec(1, width="10em") %>% 
  column_spec(2, width="25em") %>% 
  column_spec(3, width="10em")

 


```

After cleaning the data and separating all advertisements by election cycle that mention at least one of the issues mentioned in Table \@ref(tab:election-issues-cycle), I obtain an estimate of the number of advertisements that require viewing and further coding. These results are displayed in Table \@ref(tab:el-ads-numeric) above. As Table \@ref(tab:el-ads-numeric) indicates, there are approximately 840 advertisements in total that comprise my anticipated complete dataset.  

Importantly, there are also a lot of advertisements in the data that I have deemed "unlikely" to feature the military. In fact, 4805 of the total 5645 advertisements in the data, or nearly 85% of the total number of advertisements, are part of this "unlikely" group. It is important to also rule out, at least with reasonable confidence, that the data selection strategy has not overlooked potential advertisements that actually do feature the military in some way, which could lead to the result in measurement error, bias, and a host of other problems. 

I am in the process of engaging statisticians at how to do this precisely, but an initial reasonable expectation (after talking with them!) is that I also watch a random sample of the advertisements that are part of the "unlikely" group, and determine an error rate. A significant question, however, and one for which I do not yet have the answer, is how large this random sample needs to be (the statisticians with whom I am in contact are thinking about this, and I will be in touch with them over the course of this project).  Early indications suggest that it need consist of approximately 10% of the group, or roughly 480 ads. However, I need to confirm this with the statisticians with whom I am in contact, as well as confirm an acceptable error rate, that would reasonably suggest the data selection strategy is sound.   

```{r el-ads-numeric, include=TRUE, echo=FALSE}

text_tbl_ads <- tribble(~"Campaign Year", ~"# Ads Likely", ~" # Ads Unlikely", ~"Total Presidential Campaign Cycle Ads",
  "2000", "30", "700","730",
  "2004", "191", "464","655",
  "2008", "260", "706","966",
  "2012", "107", "672","779",
  "2016", "252", "659", "911",
  "Total Values", "840","4805", "5645"
  )
 

text_tbl_ads %>% kable(booktabs=T, caption="Advertisements by Election Cycle that are Likely, Unlikely (and the total) to Feature the Military") %>%
  kable_styling(full_width = F, font_size=8, latex_options="striped" ) %>% 
  column_spec(1, width="10em") %>% 
  column_spec(2, width="15em") %>% 
  column_spec(3, width="10em") 


```

 

## Data Coding

The second stage of the empirical strategy involves coding the advertisements in order to capture the particular variation I am interested, namely, the variation in how the military has featured in political advertisements, and more specifically, whether and how campaign advertisements have challenged or violated one or more of the three principles of civil-military relations. 

First, it is important to capture any variables that WiscAd and WMP have already coded for and that may be instructive when analyzing the data. Although each codebook for each election cycle is distinct, there are variables, upon close inspection, that are similar across the years. Therefore, I first clean the data, rename several variables, and change the possible values each variable can take in order to ensure that coding classifications are uniform. Ultimately, I determine that 11 variables will aid in the time-series analysis I wish to conduct. 

These variables, and a short description of each, are displayed in Table \@ref(tab:existing-variables) on the following page. These variables contain helpful information that speak to the individual advertisement (creative and length) as well as the political and partisan nature (Party, sponsor, and other variables) of the advertisement. 

```{r existing-variables, include=TRUE, echo=FALSE}

text_vars <- tribble(~"Variable Name", ~"Question", ~"Possible Values",
  "Creative",  "What is the title of the ad?", "varies by each ad",
  "Length", "How long is the ad (in seconds)", "0-120",
  "Party", "What is the Party of the favored candidate?", "Republican; Democrat; Other",
  "Sponsor", "Who sponsored the advertisement?", "1 - Candidate; 2 - Party; 3 - Coordinated Between a Candidate and Party; 4 - Interest Group/Other",
  "f_mention", "Is the favored candidate mentioned in the ad?", "1 - Yes; 0 - No",
  "f_picture", "Is the favored candidate pictured in the ad?", "1 - Yes; 0 - No",
  "f_narrate", "Does the favored candidate narrate the ad?", "1 - Yes; 0- No",
  "o_mention", "Is/are the opposition candidate/s mentioned in the ad?", "1 - Yes; 0 - No",
  "o_picture", "Is/are the opposition candidate/s pictured in the ad?", "1 - Yes; 0 - No",
  "ad_tone", "In your judgement, is the primary purpose of the ad to promote a specific candidate, attack a candidate, contrast the candidate, or other?", "1 - contrast; 2- promote; 3 - attack; 4 - other",
  "prty_mn", "Does the ad mention the party label of the favored candidate or the opponent?", "0 - No, 1 - Yes, favored candidate's party; 2 - Yes, opposition candidate's party; 3 - Yes, both candidate's parties; 4 - yes, favored candidate's party but only in the paid for by line"

  )


text_vars %>% kable(booktabs=T, caption="Existing Variables from WiscAd and WMP") %>%
  kable_styling(full_width = F, font_size=8, latex_options="striped" ) %>% 
  column_spec(1, width="10em") %>% 
  column_spec(2, width="15em") %>% 
  column_spec(3, width="20em") 


```

Next, I develop several variables specifically related to the military. In order to develop and refine these variables, I conducted a pilot coding of 10% of the sample data, ensuring that I watched 10% of each election cycle's advertisements that were "likely" to feature the military. 
First, I determine that it is important to capture advertisements that feature military imagery or symbols in some way. Therefore, I develop four dichotomous variables related to imagery, including whether the advertisement features imagery of US soldiers or troops, imagery of veterans, images of combat, or images of US military hardware. Imagery of US soldiers or troops include images of US servicemen and women in uniform of any kind, and could include a single soldier saluting, or thousands of US servicemen and women training, or a host of other possible scenarios. The key thing I look for are images of US military uniforms, medals, or insignia that are clearly related to US soldiers and troops. 

The second indicator accounts for imagery associated with veterans, which I consider being distinct from active soldiers who are in uniform as well as medals and insignia that often accompany US troops and units. This indicator variable for veteran imagery specifically looks for any image of a veteran, which could include gathering of veterans, such as a Veterans of Foreign Wars or American Legion group, or imagery of a wounded veteran recovering in a hospital, or a veteran in wheelchair later in life. The third indicator accounts for images of combat. These could include things such as an explosion, or a scene containing fighting involving US soldiers. The fourth indicator accounts for US military hardware, such as imagery of a naval vessel, a tank, an airplane, radar, etc. 

The purpose of these four indicators is to obtain a general level of what types of imagery associated with the military the advertisement contains. By themselves, the four indicators do not necessarily indicate that any of the central principles of civil-military relations have been challenged or violated, but rather give the researcher a general and broad indication regarding the ways in which the advertisement features the military.

The second class of variables involves a more direct measurement of an advertisement's challenge or potential violation of one or more of the central principles of civil-military relations. I use four additional indicator/categorical variables here. The first accounts for whether the advertisement includes imagery of a veteran or veteran's family who plays a central, personal role in the advertisement. Here, I conceive of a veteran as a former service member, regardless of rank or military service affiliation. In many cases, this denotes high ranking retired or former military officers, and in other cases, it denotes servicemen and women who were wounded in action. 

This variable can take on a value of no; yes, a veteran; yes, a veteran's family; or yes, a veteran and his/her family. In conducting the pilot coding for this project, the data reveal several advertisements in which former servicemen and women - or their family members - speak or appear in political advertisements in a prominent role that is greater than a single and more general image of veterans appearing in an advertisement, which is accounted for in the earlier variables described.

The remaining three indicator variables describe what actions, if any, the veteran and/or veteran's family members do when appearing in the advertisement in a prominent role. In particular, these indicators account for whether the veteran/veteran's family member(s) explicitly endorse a political candidate for office, attack or disparage other candidates who are running for office, and finally, whether the veteran and/or the veteran's family members address issues other than or outside of the candidate's personal characteristics or defense/veteran issues, broadly speaking. 

Furthermore, these three indicator variables that examine whether a veteran or veteran's family member endorses a candidate, disparages a candidate, and/or speaks on issues outside of a candidate's personal characteristics and defense or veteran issues help to measure respective violations of the three central principles of civil-military relations: civilian control of the military, the non-partisanship of the military, and the "non-interference" of the military into certain realms of state policy. While some of these indicator variables are clear cut, others are a bit more fuzzy and constitute a type of gray space that likely present at least a challenge to, if not an outright violation of, one or more of the central civil-military relationship principles.

```{r el-ads-additional, include=TRUE, echo=FALSE}

text_vars <- tribble(~"Variable Name", ~"Question", ~"Possible Values",
  "Ad_Purpose",  "What is the overall purpose of the ad", "1-promote or disparage candidates for an election; 2 - promote/disparage a specific issue or proposition that is on the ballot, 3 - to discuss a general theme (the economy, immigration, etc.) that does not do 1 or 2;  4 - other",
  "Can_Vet", "Is the candidate a Veteran?", "1 - Yes; 0 - No",
  "Can_Vet_Mention", "Does the ad explicitly state or portray the candidate as a Veteran?", "1 - Yes, 0 - No",
  "Vet_Fam", "Does the ad feature a Veteran or Veteran's family speaking?", "3 - Veteran and Family; 2 - Veteran's Family; 1 - Veteran; 0 - No",
  "Vet_Fam_Endorse", "Does the speaking Veteran or Veteran's Family explicitly endorse the candidate for office?", "1 - Yes; 0 - No; 99-N/A",
  "Vet_Fam_Attack", "Does the speaking Veteran or Veteran's Family explicitly attack or disparage other candidates running for office?", "1 - Yes; 0 - No; 99 - N/A",
  "Vet_Fam_Out_Issues", "Does the speaking Veteran or Veteran's Family address topics other than the candidate's personal characteristics or defense/veteran (broadly related) issues?", "1 - Yes, the vet or vet's family addresses topics outside of the candidate's characteristics and/or veterans/defense (broadly related) issues; 0 - No, the vet or vet's family only addresses the candidate's characteristics and/or veterans/defense issues; 99 - N/A",
  "Act_Duty_Imagery", "Does the advertisement feature imagery of US troops in uniform?", "1 - Yes; 0 - No",
  "Vet_Imagery", "Does the advertisement feature imagery of Veterans (VFW, veteran gatherings, etc.)?", "1 - Yes; 0 - No",
  "us_mil_hard_Imagery", "Does the advertisement feature imagery of US military hardware (planes, tanks, helicopters, ships)?", "1 - Yes; 0 - No",
  "Combat_Imagery", "Does the advertisement feature images of past or present combat involving US military forces (prior wars, scenes of combat)?", "1 - Yes, 0 - No",
  "Def_Vet_Ad_Theme", "Is the ad itself focused entirely on military, defense, veteran, or national security issues?", "1 - Yes, military/defense/veteran issues only; 0 - other topics"

  )
 

text_vars %>% kable(booktabs=T, caption="Additional Variables / Anticipated Coding Scheme") %>%
  kable_styling(full_width = F, font_size=8, latex_options="striped" ) %>% 
  column_spec(1, width="10em") %>% 
  column_spec(2, width="15em") %>% 
  column_spec(3, width="20em") %>% add_footnote(label = "The variables Vet_Fam_Endorse, Vet_Fam_Attack, and Vet_Fam_Out_Issues can all take the value of N/A, if the variable Vet_Fam takes the value of 0.")
```

For instance, the act of former servicemen or women publicly endorsing presidential candidates challenges the principle of non-partisanship of the military in perhaps the clearest of ways, because such an action conveys to the public that the military institution favors a particular partisan political candidate. The act of a former servicemen or woman's family endorsing a political candidate is likely less clear of a violation to the principle of non-partisanship, but nonetheless problematic because there is still some level of association conveyed in the advertisement between the military and a particular partisan political candidate. 

It is also possible that former servicemen or women, and/or their families, do not explicitly endorse a partisan candidate for office during the advertisement, but instead explicitly disparage or attack other particular candidates running for office, or their policies and political views. The pilot coding of the data revealed several instances of this occurring. The act of disparaging or attacking a political candidate without explicitly endorsing another candidate ostensibly allows the former serviceman/woman and/or family member to claim that he or she did not endorse a candidate for office, and importantly, for the civilian candidate, interest group, or organization response for the advertisement to claim the same, and thus, that such action does not violate the principle of non-partisanship by their actions. Such reasoning is questionable at best, in this author's view, as the disparaging of one political candidate amounts to the reverse side of the same coin of endorsement of a candidate for office, and thus, is tantamount to a challenge or violation of the military's non-partisan ethic. 

The final indicator included in this class of variables is whether the veteran and/or veteran's family speaks only to the candidate's personal characteristics and/or veteran/defense issues in a broad sense, or instead addresses topics that are outside of these categories in endorsing and/or disparaging political candidates. This variable is designed to capture potential challenges to the central principle of "non-interference" of the military into certain realms of state policy. Here, the variable examines if the veteran or veteran's family addresses policy issues that are not directly related to the candidate's traits or to defense or veterans issues, such as the candidate's plans for domestic spending, social security, education, or a host of other potential areas. Instances where this occurs, if any, would reveal a person associated with the military addressing or weighing in on topics for which the person does not reasonably possess relevant expertise, and thus, would challenge or violate the principle of "non-interference" by the military. 

In wrapping this section of variables up, note that there is no variable that captures violation of the principle of civilian control, because, as was mentioned earlier, care must be taken to determine if and when this principle has been challenged by a particular campaign advertisement. In particular, one should account for what kind of a group or sponsor created the advertisement (using the sponsor variable from Table \@ref(tab:existing-variables)), and whether the advertisement disparaged a particular candidate, to determine whether the principle of civilian control has been challenged or violated by a particular advertisement.  

Finally, in the course of conducting the pilot coding for the data, I also develop four additional variables that measure a variety of helpful items. These include the overall purpose of the advertisement, whether the candidate is a veteran, whether the advertisement explicitly mentions that the candidate is a veteran, and whether the advertisement itself is focused on military, defense, veteran, or national security issues. 

In total, after completing the pilot coding of 10% of the available data, I determine that 12 variables should be added to the data, and that each advertisement should be coded along these 12 variables. These variables, the question they intend to answer, and the potential range of values these variables can take, are displayed in Table \@ref(tab:el-ads-additional). 

It is too early to present a meaningful preview of the data, but I did want to point out that of the roughly 85 advertisements that I have coded, approximately 13 feature an ad in which a veteran or veteran's family member appears in a prominent role, and roughly 47 of the 85 advertisements feature an advertisement in which one or more of the variables for various sorts of imagery (troops/soldiers in uniform or medals/insignia; veterans, scenes of combat, or US military hardware) is applicable. Just under half of the pilot coding advertisements in the "likely" category are completely meaningless in terms of featuring military imagery of any sort, or veterans/veteran family members in a prominent role. My reaction to these early statistics is positive, in the sense that I think the selection and classification strategy, at least at this early point in the process, is reasonably sound.  

## Data Analysis

The third and final step of the empirical strategy is to analyze the data in a meaningful and compelling way. I plan to conduct this in two sequential steps. The first is to present the descriptive statistics of the data, and the second is to run a series of regressions involving several variables within the data as well as several control variables. I describe each of these steps in further detail in the paragraphs to follow. 

In presenting the descriptive statistics of the data, it is important to show, over time, to what degree the military has featured in campaign advertisements throughout the course of presidential campaigns. In order to show this, I will present a series of statistics that show, by election cycle, the number and proportion of campaign advertisements that featured one or more of the military imagery variables (troops/soldiers, veterans, combat, and military hardware). These statistics will inform the reader of the general prominence of the military in each election cycle. 

Of note, we might anticipate the general prominence of the military to vary according to exogenous events over time. For example, in the 2000 election, the military might not be as prominent as the military in the 2000 and 2004 elections. The 2000 election occurred before the events of 9/11, which thrust the US into wars in Afghanistan and Iraq in the years to follow. 

The descriptive statistics will then show, over time, the instances in which a veteran or veteran's family member appeared in campaign advertisements, endorsed a candidate, disparaged a candidate, and/or addressed issues outside of the candidate's personal traits or defense issues broadly speaking. This portion of the descriptive statistics will help show if and when there is a clear "start date" in which veterans and/or their family appeared in campaign advertisements in a prominent manner.

The second step of data analysis involves running a series of regressions that involve pertinent variables and controls. In this section, I anticipate running a series of logistic (logit) regressions, where the dependent variables are the four variables designed to capture challenges or violations of the three principles of civil-military relations (whether a veteran or a veteran's family member appears in the ad in a prominent role, endorses a candidate, disparages a candidate, addresses issues outside of the candidate's personal characteristics or defense/veteran issues). The main control variables in each of these regressions will consist of measures of political polarization and military prestige. Specifically, I will use two different measurements of political polarization that I previously used in an earlier chapter of the dissertation on retired officer commentary, to include DW-NOMINATE scores and measurements of affective polarization, both of which are standard and conventional measurements of polarization in the American Politics literature [@iyengar_affect_2012; @hare_polarization_2014; @iyengar_strengthening_2018; @jeffrey_b_lewis_voteview_2020]. 

In order to measure military prestige, I plan to rely on public opinion data stemming from Gallup polling that asks Americans to rate the level of trust they place in various institutions [@noauthor_confidence_2020]. This measurement of military prestige is a broad and general measurement of the level of trust that Americans have placed in the military over time, but it does have drawbacks, one of which is that it is generally unresponsive to micro-level factors that could alter how prestigious the military is across American society at any given point in time. 

In terms of other controls, I may include a measure of public opinion for the war in Iraq/Afghanistan. I am curious what the reader of this research project thinks on this point in particular. I do not think that it is necessary to control for many other variables, though it may be necessary to include election-year fixed effects to control for the overall characteristics of each election cycle. Finally, it may be necessary to form an interaction term between polarization and military prestige, the essence of which speaks to the heart of my overall research project and the specific hypotheses developed for exploration in this chapter. 

# Anticipated Results, Alternate Explanations and the Way Forward

While it is too soon to conclusively predict the results of this research agenda, there are a couple of general predictions revealed by the coding of pilot ads that are likely to stand as I continue to code the data. The first is that campaign advertisements in every election cycle do in fact feature the military in some way. From 2000 to 2016, the military is featured during campaign advertisements, at least in a broad and general sense. This is more or less good news for the research agenda, as such a finding suggests that there is a strong likelihood of finding clear variation in how the military was featured across time in these advertisements. 

The second finding is also somewhat unsurprising, and that is that the 2016 election in particular looks to be an election that really brought out a lot of changes in the way the military featured in campaign advertisements. Already in the pilot coding, several advertisements reveal instances in which several former servicemen and women, and/or their families, endorsed or disparaged one or more of the candidates seeking office. While this is not surprising given that the 2016 election pitted two deeply unpopular candidates in the general election in Hillary Clinton and Donald Trump, it does likely raise an empirical question that I and other scholars will have to contend with at some point in this research project: is it polarization, or is it Donald Trump, that is driving the results here? There are no easy answers to this question, but I will have to confront it at some point in this chapter. 

In all likelihood, there will be at least a strong correlation between the variables of polarization and military prestige and the dependent variables used in this study, instances of former servicemen and women, and/or their families, who either endorse a candidate, disparage a candidate, or address issues that are unrelated to the personal characteristics of a candidate or defense/veteran issues more broadly during political advertisements. And while the correlation is a good result to obtain, I also have to acknowledge and address potential confounding variables in order to prove that the general theory posited earlier in the dissertation has merit.  

With respect to potential confounding variables and alternate explanations besides polarization and military prestige that are driving changes in the way that civilian actors feature the military in campaign advertisements, I think there are at least two likely culprits of alternate explanations that critics of the approach laid out in this proposal might point to and be concerned about. These culprits are exogenous events, and overall changes in the media environment. 

Some critics might say that exogenous events, such as a relatively unpopular war in Iraq at the time of the 2004 Presidential Election, might drive potential ways in which civilian candidates portrayed the military during campaign advertisements. To a certain extent, I think that these critics would be correct. However, I think that while external events might shape the _prominence_ or _salience_ of certain topics featured in certain campaigns over other topics - a completely reasonable expectation, in my opinion - these external events would not drastically the degree that campaign advertisements challenge or violate civil-military principles themselves.

In other words, it is reasonable to expect that the military is featured generally more in 2004 campaign advertisements than in 2000 campaign advertisements, but it does not necessarily follow that this would automatically mean that 2004 would see greater frequency of advertisements that challenge or violate central principles of civil-military relations. 

A second alternate explanation, that of changes in the media environment in general driving variation in how the military is featured in campaign advertisements, might also be something that a critique would point out as being important. I share this concern. I plan to speak to mass media scholars to get help in addressing this concern. I also am a little concerned that changes in the media environment are possibly highly correlated with polarization, which could further problematize the interpretation of the results. 

Moving forward, I see three major tasks, which I intend to conduct simultaneously as I move through this research agenda. The first is to continue coding and, in the process, confirm that my data selection strategy, which separates the data into "likely" and "unlikely" categories of featuring the military, is sound. This specifically involves confirming with statisticians that watching a sample of the "unlikely" category of advertisements, followed by determining an error rate from the random sample that I view, is sound from a statistical standpoint. 

The second task is to dig further into the "issue ownership" and "mass media" literatures. The first will help me better situate this chapter in terms of its contribution to the overall scholarly literature, while the second will help me to address potential confounding explanations in the overall media environment that could impact the results I obtain. To be fair, I feel as if much of this research agenda, as it currently stands, lies more in the American Politics literature, but I am actively looking for ways to bring it more into the domestic politics of international relations literature. There is undoubtedly some overlap here, but I need to determine the best way to make domestic political campaign advertisements attractive and relevant to IR scholars. I think there is a way to do this, but I admit that I have more work to do in this arena.  
The third and final task I see is to be prepared to describe the anticipated results from this project as interesting _even if_ there is a null finding, or if one or both of my hypotheses are unable to be fully proven. I anticipate that some scholars will rightly question whether polarization and military prestige are truly doing the work that I think they are, and point out some of the objections previously addressed here. I need to anticipate these objections more and really think about to what degree these objections hold water, and to what degree these objections do not. 

# Conclusion

A key argument made previously in this dissertation is that polarization and prestige significantly influence the way in which civilian and military actors abide by fundamental principles of civil-military relations, and that as a result, these variables also change how these actors commit certain behaviors that involve the military institution. Campaign advertisements are one type of political behavior committed by civilian actors that are a regular feature of the political landscape and the democratic process. Presidential campaign television advertisements are displayed every four years in an attempt to persuade voters to vote for a certain candidate over another. While these advertisements do not necessarily have to feature the military at all, nor necessarily in ways that challenge the military's non-partisan ethic and a principle of "non-interference" by the military into certain state affairs, recent political campaigns have demonstrated that certain advertisements can, in fact, be problematic along these lines. 

Conducting a systematic investigation of campaign advertisements for the five Presidential Elections from 2000-2016 will reveal potential variation in how the military has featured in these advertisements. In this document, I have proposed a research agenda and an empirical strategy for collecting the data which will ultimately display any potential variation in this arena to scholars, and important contribution in itself, and determine the degree to which political polarization and military prestige have driven these changes. 

# Descriptive Statistics
 
```{r data, include=FALSE, echo=FALSE}
# read in the raw data
raw_df <- read_csv(here("data", "df_comb_init_cads_edits.csv"))
summary(raw_df)

# truncate the data at the first 840 observations, and drop 3 x variables (notes, endorser, and #end_who)

# also, replace NA with 0 for the variable "finished" - will come in handy later. 
df <- raw_df[1:840,] %>% select(-notes, -endorser, -end_who) %>% 
  mutate_at(
    .vars = vars(finished),
    .funs = ~ case_when(is.na(.) ~ 0, TRUE ~1)
) %>% view()
```
 
```{r data selection results, include=FALSE, echo=FALSE}
#create a variable called Yrcount that counts the number of ads per election cycle
#create a variable called "mil_img" if there is any type of military imagery category in a 
# given ad
df <- df %>% 
  filter(finished ==1) %>% 
  group_by(year) %>% 
  mutate(
    Yrcount=n(),
    mil_img = case_when(
    (vet_fam != 0) | (act_dty_img ==1 | vet_img ==1 | us_mil_hard_img ==1 | us_cmbt_img==1) ~ 1, 
  (vet_fam == 0) & (act_dty_img ==0 & vet_img ==0 & us_mil_hard_img ==0 & us_cmbt_img==0) ~ 0
),
  no_mil_img = case_when(
    (vet_fam != 1) & (act_dty_img !=1 & vet_img !=1 & us_mil_hard_img !=1 & us_cmbt_img!=1) ~ 1, 
  (vet_fam != 0) | (act_dty_img ==1 | vet_img ==1 | us_mil_hard_img ==1 | us_cmbt_img==1) ~ 0
  )
    ) %>% 
  view()

# create a new df called 'adj_df' that 
# takes the original df and adds additional variables 
#(1) sum_mil_img, the total # of ads per election cycle containing military imagery 
#(mil_img ==1); 
#(2) sum_no_mil_img, the total # of ads per cycle containing no military imagery
#(3) mil_fig, if an ad features a prominent veteran or veteran's family in the ad
# (mil_fig==1)
#(4) sum_mil_fig_cycle, the total # of ads per cycle that feature a veteran or family endorser
# sum_mil_fig_cycle
#(5) sum_def_vet_ad_theme, the total # of ads per cycle that address entirely matters
# of national security, defense, or veterans issues
adj_df <- df %>% group_by(year) %>% mutate(
    sum_mil_img_cycle=sum(mil_img),
    sum_no_mil_img_cycle=sum(no_mil_img),
    mil_fig=case_when(vet_fam != 0 ~1,
                      vet_fam == 0 ~ 0),
    sum_mil_fig_cycle=sum(mil_fig),
    sum_def_vet_ad_theme=sum(def_vet_ad_theme),
    def_theme_and_mil_img=case_when((def_vet_ad_theme==1) & (mil_img==1) ~1,
                                    (def_vet_ad_theme!=1) | (mil_img==0) ~0),
    sum_def_theme_and_mil_img=sum(def_theme_and_mil_img),
    not_def_theme_but_mil_img=case_when((def_vet_ad_theme!=1) & (mil_img==1) ~ 1,
                                         (mil_img==0) | (def_theme_and_mil_img==1) ~0),
    sum_not_def_theme_but_mil_img=sum(not_def_theme_but_mil_img),
    non_def_vet_ad_theme=case_when((def_vet_ad_theme==0) ~1,
                                  (def_vet_ad_theme==1) ~ 0),
    sum_non_def_vet_ad_theme=sum(non_def_vet_ad_theme)
    ) %>% view()

desc_stats <- tribble(
  ~'Election Cycle', ~'Total Ads', ~'Ads Selected for Examination', ~'Ads Selected but Missing from the Data', ~'Ads Examined that Contained Some Form of Military Imagery', ~'Ads Examined that Prominently Feature a Veteran or Veteran Family', 
  2000, 730, 28, 2, 15, 2,
  2004, 652, 190, 1, 85, 26,
  2008, 957, 251, 9, 100, 13,
  2012, 779, 107, 0, 44, 14,
  2016, 910, 251, 0, 120, 31)

```

In this section of the chapter, I present the descriptive statistics for the data examined. Table \@ref(tab:desc-stats) shows, for each election cycle, the total number of advertisements aired on the nation's top media markets, the number of advertisements that met the selection criteria and were subsequently examined, the number of selected advertisements that were missing, the number of advertisements that featured some form of military imagery, and the number of advertisements that featured either a veteran and/or a veteran's family in a prominent way. In total, 840 total advertisements were selected for examination (of these, 12 were missing from the data). Of the 828 advertisements coded, 364 (44%) contained some form of military imagery, while 464 (56%) did not. 

```{r desc-stats, include=TRUE, echo=FALSE}
 
kable(desc_stats, booktabs=T, caption="Summary Statistics, The Military in Presidential Campaign Advertisements, 2000-2016") %>%
  kable_styling(full_width = F, font_size=11, latex_options="striped") %>% 
  column_spec(1:2, width="8em") %>% 
  column_spec(3:6, width="12em") 


```

Although the descriptive statistics alone do not provide complete answers the questions sought after in this chapter, they do however provide some information regarding the general prominence of military and defense related issues during a given election cycle. Table \@ref(tab:def-prominence) displays the percentage of presidential campaign advertisements that featured at least one of the forms of military imagery coded for (active duty troops or insignia, veterans, military hardware, or combat), as well as a possible reason to generally describe why the military featured relatively prominently or non-prominently in a given election cycle. 

Note, for example, that during the 2000 presidential election, roughly two percent of presidential campaign advertisements featured military imagery of some sort, whereas during the 2004 election, approximately thirteen percent of presidential campaign ads featured some form of military imagery. likely reason for this sharp increase between these two election cycles is the September 11, 2001 attacks on the US, which was subsequentlyfollowed by respective military operations against Afghanistan later in 2001, and Iraq in 2003. Therefore, in 2004, when George W. Bush ran as the incumbent against the Democratic challenger, John Kerry (who was a decorated combat veteran of the Vietnam War), the direction of the war effort in Iraq featured as a major campaign theme, and thus, a greater portion of advertisements featured some flavor of military imagery.   

Relative to the figure of 13 percent in 2004, the next three presidential elections - 2008, 2012, and 2016 - saw two declines followed by a sharp rise in the proportion of campaign advertisements that featured the military. A full unpacking of these trends requires a much lengthier description than can be given here, but a likely if perhaps oversimplistic explanation would identify several trends and important events that occurred in the twelve years between the 2004 and 2016 elections. In 2008, the surge of US military forces in Iraq had shown signs that it was, albeit imperfectly, working in terms of reducing sectarian violence in Iraq, and thus, there was less division among the US public about how to proceed in that country. US casualties had also been declining steadily throughout the calendar year of 2008 after peaking in 2007, further strengthening the argument that the Surge of US forces over 2007 - 2008 had tamed much of the violence in Iraq [@us_department_of_defense_dcas_2021]. 

```{r def-prominence, include=TRUE, echo=FALSE}
def_prominence <- tribble(
  ~'Election Cycle', ~'Percentage of Ads Containing Fome Form of Military Imagery', ~'Likely Reason', 
  '2000 Election', 2, 'Late 1990s Pre-9/11 Era; the USA was not yet steadily at war',
  '2004 Election', 13, 'The War in Iraq was a year and a half old, but was already controversial; whether to stay or change course in Iraq was a major campaign theme',
  '2008 Election', 9.5, 'The Surge in Iraq had began to show signs that it was working; thus, there was less controversy about what to do in Iraq compared to the 2004 Election',
  '2012 Election', 5.6, 'The US had by this time drawn down heavily in Iraq, and had implemented a surge in Afghanistan roughly two years prior to the election. Additionally, Osama Bin Laden had been killed in 2011 by US military forces',
  '2016 Election', 13.1, 'The ongoing wars in Iraq and Afghanistan had receded somewhat from the center of US public attention; however, the rise of ISIS, Russian aggression, and other general issues regarding Syria as well as Benghazi featured prominently as a Campaign Theme')

kable(def_prominence, booktabs=T, caption="Percentage of Presidential Campaign Ads that Feature Military Imagery, 2000 - 2016") %>%
  kable_styling(full_width = F, font_size=11, latex_options="striped") %>% 
  column_spec(1:2, width="8em") %>% 
  column_spec(3, width="14em") 



```
From 2008 to 2012, the US' involvement in Iraq lessened while its involvement in Afghanistan increased. Indeed, President Barack Obama in 2009 decided to increase the number of US forces in Afghanistan as part of a surge of US forces. Though the Afghan surge did not bring the acclaim that the earlier surge to Iraq did, there were nonetheless important victories along the way, perhaps none as important as the killing of Osama bin Laden, the architecht of the 9/11 attacks, in early 2011. Additionally, US casualties in Afghanistan had peaked in the summer of 2011, and declined afterwards, especially in the months leading up to the 2012 election [@us_department_of_defense_dcas_2021-1]. At the time of the election in November of 2012, few observers could point to Afghanistan and say that the surge of forces there had not worked. At the same time, there was no appetite - from politicians and the public, alike - to do anything substantially greater there. Therefore, the relatively low percentage of campaign advertisements featuring the military in 2012 is unsurprising. 

However, in the 2016 election, the percentage of advertisements featuring military imagery rose sharply compared to the 2012 election. Despite the surges in Iraq and Afghanistan being long finished at this point, the passage of time also facilitated the opportunity for Americans and political candidates to think about the trajectory of US foreign and national security policy in broader and more critical terms. Hillary Clinton, the Democratic Candidate for President, was largely seen as an extension of Barack Obama, particularly in the realm of foreign policy, as Clinton had served as Obama's Secretary of State. Moreover, since the 2012 election, several US foreign policy choices were viewed rather negatively, including Obama's handling of events in Syria in 2013 (including Obama's infamous "red line" statement) and concern over rising Russian aggression. Hillary Clinton in particular was portrayed extremely negatively after a US Ambassador and three other Americans were killed on September 12, 2012, when terrorists attacked the US consulate in Banghazi, Libya [@bixby_parents_2016]. Donald Trump, an outside candidate who ran and won as a Republican, instead ran on a platform of putting "America First," which included promises to cease fighting endless wars and rebuilding the US military. For these reasons, one can understand how and why military imagery, at least in a general sense, would have been more prominent in 2016 than in 2012. 

# Can Candidates Afford to Bypass or Ignore the Military?

During advertisements in which foreign or national security policy is the prominent theme or issue discussed, one can expect a reasonable chance to see pictures of troops in uniform, tanks or airplanes, or other forms of military imagery. But it is not clear that in other advertisements, many of which are 30 seconds in length and discuss a variety of themes or issues, the military would be one of these issues that "make the cut" and are chosen by candidates to be conveyed to the public.

Figure \@ref(fig:def-theme) displays the number and portion of campaign advertisements by election cycle separated by those that entirely address a national security/defense/military theme versus that do not. 

```{r def-theme, include=TRUE, echo=FALSE, fig.cap="Presidential Campaign Advertisements Featuring Military Imagery, Separated by Ads that are Entirely Focused on National Security Topics"}

plot1 <- ggplot(data=adj_df, aes(x=year, y=n)) + labs(x="Source", y="Number of Campaign Advertisements") + geom_col() + geom_text(aes(label=n), position=position_dodge(width=.9), vjust=-.25) 

colplot2

```

\singlespace

\pagebreak

# Appendix

## 2000

The data were obtained from a joint project of the Brennan Center for Justice at New York University School of Law and Professor Kenneth Goldstein of the University of Wisconsin-Madison, and includes media tracking data from the Campaign Media Analysis Group in Washington, D.C. The Brennan Center-Wisconsin project was sponsored by a grant from The Pew Charitable Trusts. The opinions expressed in this article are those of the author(s) and do not necessarily reflect the views of the Brennan Center, Professor Goldstein, or The Pew Charitable Trusts.

## 2004

The data were obtained from a project of the University of Wisconsin Advertising Project includes media tracking data from TNSMI/Campaign Media Analysis Group in Washington, D.C. The University of Wisconsin Advertising Project was sponsored by a grant from The Pew Charitable Trusts. The opinions expressed in this article are those of the author(s) and do not necessarily reflect the views of the University of Wisconsin Advertising Project or The Pew Charitable Trusts.

## 2008

The data were obtained from a project of the University of Wisconsin Advertising Project includes media tracking data from TNSMI/Campaign Media Analysis Group in Washington, D.C. The opinions expressed in this article are those of the author(s) and do not necessarily reflect the views of the University of Wisconsin Advertising Project.

## 2012

The data were obtained from the Wesleyan Media Project, a collaboration between Wesleyan University, Bowdoin College, and Washington State University, and includes media tracking data from Kantar/Campaign Media Analysis Group in Washington, D.C. The Wesleyan Media Project was sponsored in 2012 by grants from The John S. and James L. Knight Foundation and The John D. and Catherine T. MacArthur Foundation. The opinions expressed in this article are those of the author(s) and do not necessarily reflect the views of the Wesleyan Media Project, Knight Foundation, MacArthur Foundation or any of its affiliates.

## 2016

The data were obtained from the Wesleyan Media Project, a collaboration between Wesleyan University, Bowdoin College, and Washington State University, and includes media tracking data from Kantar/Campaign Media Analysis Group in Washington, D.C. The Wesleyan Media Project was sponsored in 2016 by a grant from The John S. and James L. Knight Foundation. The opinions expressed in this article are those of the author(s) and do not necessarily reflect the views of the Wesleyan Media Project, Knight Foundation, or any of its affiliates.

\newpage



