---
title: "Chapter 5 (Case Study, 1865-1892): Senior US Military Leaders in an Age of Declining Military Relevance "
author: "Peter M. Erickson"
date: |
  `r format(Sys.time(), '%B %d, %Y')`
abstract: |
  
  _This historical case study compares the patterns of political behavior engaged in by civilian and military actors involving the military institution from roughly 1865 - 1885. During this time period, the prestige of the military declined as the post-Civil War military took on constabulary duties and shrank in size. Using process tracing, I describe how the high levels of polarization and military prestige that were evident at the beginning of this historical era combined to create a toxic political environment in which both senior military and civilian leaders engaged in behaviors that violated the central principles of civil-military relations. I then argue that as the prestige of the military declined, the potential political influence of the military, leading to military and civilian actors engaging in less severe instances of political behavior that violated the central principles of civil-military relations._  
  
geometry: margin = 1.15in              # margin widths
fontsize: 12pt                         # text size
bibliography: dissertation.bib         # bibliography file
biblio-style: authoryear               # citation style
urlcolor: blue                         # links to external URLs
citecolor: black                       # links to bibliography
linkcolor: magenta                     # internal figures, footnotes, etc.
indent: true                          # paragraph format
                                       # true for indented paragraphs
                                       
header-includes:
    - \usepackage{setspace}
    - \usepackage{multirow}
    - \usepackage{indentfirst}
   
    
output: 
  bookdown::pdf_document2:
    keep_tex: true
    latex_engine: pdflatex
    number_sections: false
    toc: false
    fig_caption: true
    citation_package: biblatex
    includes: 
      in_header: 
        - latex-preamble.tex
  bookdown::htmldocument2: default
  bookdown::word_document2:
      toc: false
---

\doublespacing 

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo=FALSE,
  include=FALSE,
  message=FALSE,
  error=FALSE,
  fig.height=3,
  fig.width=4,
  out.width="90%")
```

```{r, include = FALSE, echo = FALSE}
library("here")       # file paths
library("tidyverse")  # workhorse package
library("tidylog")
library("kableExtra")
library("knitr")
library("ggdag")
library("dagitty")
library("gridExtra")
library("broom")
library("dplyr")
library("reshape2")
library("pacman")
library("ggrepel")
```

\newpage
# Introduction
	
This chapter compares and contrasts the patterns of political behavior engaged in by military and civilian actors from the end of the American Civil War in 1865 to roughly 20 years later, in 1885. Using the case study method and through the use of process tracing, this chapter explores the "motives and opportunities" framework developed in chapter two of this dissertation.

The goals of this chapter are four fold: first, to argue that a decline in the military's prestige actually occurred over the period from 1865-1885, and to trace this decline; second, to argue that the decline in the level of military prestige over the period examined resulted in a reduction in the "opportunity" for both civilian and military leaders to involve the military in politics; three, to show that throughout the time period examined, the relatively constant yet high level of polarization provided military leaders with a "motive" to engage in political behaviors that involved the military institution; and four, to argue that the latter portion of the case study was altogether a less corrosive environment for the conduct of civil-military relations than the immediate years after the Civil War.  

This chapter proceeds in five parts. First, I explain what case studies are, and why they augment the research agenda already undertaken in this dissertation. In this section, I also defend the logic of the two broad case studies which comprise both this chapter as well as the next. Second, I describe the general predictions regarding the patterns of political behavior that should be evident based on the "motives and opportunity" framework previously set forth in this dissertation. 

The third and fourth parts of this chapter comprise the heart of the 1865 - 1885 case study. Each of these parts explores the earlier and latter parts of case itself. The fifth and final part of this chapter then evaluates and compares the differences in the patterns of political behaviors that occurred during these two eras. Using process tracing, I argue that the reduction in the military's prestige that occurred from 1865 - 1885 resulted in both civilian and military actors engaging in less public and less severe forms of political behavior involving the military. I conclude that the case study examined in this chapter fulfills most of the predictions made by the theory, but that it also differs from the theory in a couple of important ways. 

# Summary of Results Obtained, and How Case Studies Help
	
Before diving into the specific case study explored in this chapter, it is important to lay the logical groundwork for what to look for when conducting case studies, and to state why case studies matter in the first place. In order to do this well, I wish to briefly summarize the empirical results already obtained in the previous two chapters, as doing so will help motivate the specific case studies explored in the remainder of this dissertation. 
## A Summary of Results Already Obtained

Each of the two previous large-N chapters examined one form or type of a political behavior that is primarily (though not exclusively) engaged in by one type of civil-military actor. Chapter three examined opinion commentary authored by retired military officers and published in major newspapers, and chapter four examined the use of military figures in civilian presidential campaign advertisements. 

The statistical analysis undertaken in each of the previous chapters indicated that increasing levels of polarization drives retired military officers to author opinion commentary in ways that violate the principles of civil-military relations, and that increasing levels of military prestige drives civilian actors to use military figures in ways that violate the principles of civil-military relations. These initial results are important because they established a clear connection between each type of actor, military and civilian, and a particular form of political behavior that each engages in involving the military institution. 

Still, there is much of the theory that is not yet addressed by these previous results. In particular, the results have not clearly indicated whether and to what degree military actors are also sensitive to changes in the level of military prestige, as the theory predicts they would be. Additionally, because the previous chapters explored only one type of political behavior in depth, conclusions could not be reached concerning other forms of political behavior that each actor type can engage in. Historical case studies will instead attempt to examine an era, rather than one particular form of behavior, in order to establish overall patterns about the multiple political behaviors that occur in a given period of time. This is important because, as the US experience proves, the levels of polarization and military prestige vary over time. In terms of the five hypotheses developed in chapter two of this dissertation (and listed below), the two case studies undertaken in this and the next chapter will help shed light on all five, but in particular, H2, H4, and H5. 

H1: Military actors increasingly violate the central principles of civil-military relations when polarization is high relative to periods of low polarization. 

H2: Military actors increasingly violate the central principles of civil-military relations in public and visible ways when military prestige is high relative to periods of low military prestige. 

H3: Civilian actors increasingly violate the central principles of civil-military relations when military prestige is high relative to periods of low military prestige. 

H4: Military and civilian actors engage in relatively more visible and public forms of political behavior that involve the military institution when military prestige is high relative to periods of low military prestige. 

H5: The most extreme levels of overall political activity occur in environments that are characterized by simultaneously high levels of political polarization and military prestige.  

## Methodological Advantages Afforded by Case Studies

A case study, in the words of Brady and Collier (2010), "may be understood as a temporally and spatially bounded instance of a specified phenomenon" [@brady_rethinking_2010, 208]. These same authors define the concept of process-tracing as involving "the examination of 'diagnostic' pieces of evidence within a case that contribute to supporting or overturning alternative explanatory hypotheses" [@brady_rethinking_2010, 208]. Brady and Collier also argue that the use of process tracing within a case study is valuable for at least two main reasons.

The first advantage of process tracing through case studies is that it can inform whether the independent variables influence the dependent variable through the pathways, mechanisms, and sequences claimed to be at work in a given theory [@brady_rethinking_2010, 208-209]. In the context of this particular dissertation, this means that the use of case studies can help clarify the nature and the extent to which, in a given historical era, the level of political polarization serves as a "motive" for military actors to violate the central principles of civil-military relations, and the level of military prestige provides an "opportunity" for both civilian and military actors to do the same.  

This also means that case studies can help diagnose the degree to which endogeneity is a problem within a given theory. In this dissertation, this means the case studies should clearly illustrate that military and civilian actors engage in political behaviors that involve the military as a result of the level of the military's prestige and the level of political polarization. The case studies should not, in other words, demonstrate a relationship that works in the other direction.^[In the United States context in particular, the more likely problem with endogeneity, if there is one, exists between the independent variable of military prestige and the dependent variable of political behaviors involving the military, rather than with the relationship between polarization and the dependent variable. This is the case because as I have defined polarization in chapter two as the contestation of worldviews and moral judgments, and how they battle for primacy in the political sphere, any worldview differences are likely formed prior to the military's behavior. I recognize this may not be the case in other contexts outside of mature democracies.] 

A second advantage of process tracing within a case study is that its use enables the researcher to evaluate potential alternate explanations that may also be thought to influence the dependent variable [@brady_rethinking_2010, 208]. In this dissertation, this means that the use of process tracing within a case study can help inform the degree to which other explanations other than the levels of polarization and military prestige might drive the degree to which military and civilian actors are constrained by the central principles of civil-military relations. In the theory chapter of this dissertation, several potential alternate explanations were identified, including norm change and the prevailing conflict or threat environment. A good case study, therefore, should enable a meaningful comparison of these alternate explanations against the explanation offered by the theory.

For both of these reasons, the case study is an important analytical tool, both in general but also in this dissertation. While large-N statistical analyses (of the type undertaken in chapters 3 and 4 of this dissertation) and qualitative case studies may in fact, as Gerring (2011) describes, "explore the world in different ways," the two methods should not be viewed as "rival genres," but rather as complimentary methods that ultimately help "synthesize" each other [@gerring_case_2011]. 

## Case Selection Considerations

In this section, I describe the cases used in this and the next chapter, and defend the logic behind these choices. An ideal case study involves ample variation in one or both of the independent variables for the purposes of showing how the variation that occurs in one or more of the independent variables results in changes in the dependent variable. The ideal case study then allows for the researcher to argue that the reasons offered by the theory are responsible for the changes that occur in the dependent variable.

In this dissertation, the two independent variables of political polarization and military prestige always exist at some level. Additionally, these two independent variables change, but not necessarily in isolation from the other. That is, there is every reason to expect, and indeed, the empirical record demonstrates, that the levels of polarization and military prestige often both vary over the same time period, and sometimes in the same direction. When both variables change, especially in the same direction, it can be difficult to disentangle the relative influence of each variable. To the degree that it is possible, however, I want to find cases in which one variable clearly changes and the other variable either remains relatively constant, or at least changes relatively less than the other. In cases where such a pattern emerges, I can more easily discern the relative influences of the independent variables on the dependent variable.   

It is important to mention one other related but important consideration. Because the independent variables examined in this dissertation both change, I also want to try and limit the length of time comprising each case study, or at least define it as precisely as possible so as to limit the number of other changes occurring in the same time period that might also impact the dependent variable. In short, this dissertation needs to examine a relatively short and focused time period in which primarily one independent variable changes. 

Although United States history affords a number of potentially useful cases for this dissertation, a great place to begin is historical eras in which the independent variables hold extreme values. For instance, given polarization data that relies on Congressional roll call voting data, polarization levels are the highest in US history in two distinct eras, one ranging from approximately 1879-1912, and the other from 2004 - 2022. The same data reveals that polarization was lowest from approximately 1940 - 1970. 

The independent variable of military prestige is a little more difficult to gauge over time. Data on public trust in the US military dates back only to 1975 [@noauthor_confidence_2020]. However, there are a couple of historical signposts available that, in conjunction with other indicators that exist, such as the size and budget of the military, can help triangulate the relative levels of military prestige prevalent in a particular historical era. 

As a quick example, it is reasonable to conclude that the post-World War Two time period is one of high military prestige, given the scope of the victory achieved by US military forces and the size of the impact the victory wrought for the United States in terms of the role it went on to assume in the world. After World War Two, the size of the military sharply increased, and it assumed truly global responsibilities - something it has not cast off in the years since.  

## A Description of the Selected Cases

Given that the previous large-N statistical chapters focused on generally the contemporary modern United States (1979-2020 in chapter 3, and 2000 - 2016 in chapter 4), I am deliberately choosing cases that are mostly outside of this time period. I do so on the basis that exploring different time periods than those already examined will better inform the broad predictions made by the theory in this dissertation. 

The first case study I will conduct, and the focus of this particular chapter, is an examination of the roughly 16 years after the US Civil-War, from approximately 1865 - 1881. I will argue that over this period, polarization remained relatively high, while the relative prestige of the US military declined. Thus, this case study involves an examination of the political behaviors involving the military in a domestic environment that changes from one marked by high polarization and high military prestige, to one characterized by high polarization and lower military prestige. The main independent variable changing in this particular case study, then, is that of military prestige. The goal of this case study is to determine the degree to which the military's declining prestige resulted in less of an "opportunity" for both civilian and military actors to involve the military in politics. At the same time, the case study should show evidence that the enduring high polarization of this period nonetheless provided military actors with a "motive" to intervene in politics.  

The second case study, undertaken in the next chapter, is an examination of the period between the end of the Vietnam War and the Gulf War, from approximately 1973 - 1990. During this time period, military prestige remained relatively low, as the US military rebuilt in the wake of the Vietnam War. Moreover, the military had not yet achieved its 1991 battlefield victory in the Gulf War, an event which many in the military and civil-military relations scholars have come to view as a critical moment in terms of the military's rebirth after Vietnam. Over the same period, however, polarization rises, beginning in the mid-late 1970s. Thus, this case study involves an examination of a domestic environment that changes from relatively low military prestige and low polarization, to one marked by low military prestige and higher polarization. The goal of this case study is to determine the degree to which rising polarization strengthened the "motive" for the military to intervene in politics. Simultaneously, however, the case study should show evidence that because of the military's enduring low level of prestige, neither civilian nor military actors possessed a significant "opportunity" to involve the military in politics prior to the victory US forces achieved in the Gulf War.    

# Theoretical Predictions for the Post US Civil War Era, 1865-1881

## High Polarization and High Military Prestige

The levels of polarization and military prestige were both very high upon the end of the Civil War. According to the theory laid out in this dissertation, there should be evidence that due to the high level of polarization, there were significant ideological divisions within the military officer corps, reflecting the high level of polarization that existed in the nation at the time. Furthermore, there should be evidence that the high level of polarization prevalent at the time served as a "motive" for military officers to intervene in politics, and that military actors engaged in political behaviors based on these motives in ways that violated one or more of the central principles of civil-military relations. 

The level of military prestige was also relatively high at the beginning of this period, however. Thus, there should also be evidence that civilian and military actors engaged in behaviors that sought to harness the "opportunity" to involve the military in politics. These behaviors, according to the theory offered earlier in chapter two, should be relatively visible to the public as both sets of actors seek to leverage the prestige of the military. 

Finally, the theory predicts that the simultaneously high levels of polarization and prestige will result in civilian and military actors engaging in a comprehensive form of military politicization. This means that civilian and military political behavior should, together, reflect a rather extreme level of political activity, and that both actors will do so in ways that undermine one or more of the principles of civilian control, non-partisanship, and non-interference of the military.   

```{r poss-environment2, echo=FALSE, include=TRUE, error=FALSE, message=FALSE}

text_env_revised <- tribble(~"Environment", ~"Low Polarization", ~"High Polarization",
  "Low Prestige", "No politicization; low levels of overall political activity.", 
  "Military (induced) politicization; moderate levels of overall political activity.",  
  "High Prestige", "Military and Civilian (induced) politicization; moderate levels of overall political activity.", 
  "Comprehensive politicization; extreme levels of overall political activity.  " 
  )
 

  kable(text_env_revised, booktabs=T, caption="Types of Domestic Environments for Civil-Military Relations and their Impact on Political Behaviors") %>%
  kable_styling(full_width = F, font_size=10, latex_options="striped") %>% 
  column_spec(1, width="8em") %>% 
  column_spec(2:3, width="18em") 

```
## High Polarization and Low(er) Military Prestige

As time passes, I argue that the relative level of military prestige declines. According to the theory offered in this dissertation, the most significant prediction stemming from declining military prestige is that the "opportunity" for civilian and military actors to involve the military in politics diminishes. The theory predicts that declining military prestige is associated with a reduction in the perceived potential of the military to influence public opinion, which in turn predicts that civilian and military actors will become less likely to undertake political behaviors that involve the military in politics in highly public ways. Because polarization remains high in the latter portion of this case study, however, the theory predicts that military actors nonetheless maintain a "motive" to intervene in politics. Thus, in a domestic environment categorized by "motive" without "opportunity," the theory predicts that military actors will be the primary type of actor inducing political activity, and that they will engage in less overt forms of political behavior than when military prestige is high. 

In the pages to follow, I argue that during the post-US Civil War era, the character and severity of political behaviors involving the military undertaken by civilian and military actors changes. Early in this period, military and civilian actors engaged in highly visible political behaviors that involved the military. Towards the end of the period, however, civilian and military leaders reacted to the military's diminishing political influence by engaging in fewer and less overt forms of political behavior involving the military. By the end of the period examined, military actors were the primary actors engaging in political behaviors, but even these were relatively less overt and less severe than those undertaken in the years immediately after the end of the Civil War. 

However, this case study also reveals some noteworthy exceptions to the predictions made by the theory, particularly in the fact that even in an era of overall diminishing military prestige, several individual military figures retained significant personal prestige throughout the era. Highly popular and well known military actors undertook public political behaviors themselves, or were the object of civilian efforts to leverage the military for explicit partisan and/or political advantage. I interpret this finding - that individual military figures have been highly prestigious even when the military as an institution is not - not so much as a major theoretical problem for the theory of this dissertation, but rather as an important cautionary note in that military prestige often operates at two different levels, that of the institution and that of the individual military leader. This finding adds richness to the ways in which scholars should both conceptualize of and measure the variable of military prestige and all that it entails.  
				
# Part 1: Johnson, Grant, Congress, and the Army, 1865 - 1868

In early 1868, the United States endured a significant political crisis when Congress impeached President Andrew Johnson, who had succeeded Abraham Lincoln. By a single vote cast in May 1868, the US Senate voted not to remove Johnson as President. Two days before Johnson's acquittal, the nation's highest ranking military figure, General Ulysses S. Grant, was nominated to run as the Republican Party's candidate for office. Less than a year later, Grant would occupy the presidency, an office which he would hold for two terms. 

Despite occurring 154 years ago, the US Army and its senior leaders played a much larger role in the events leading up to Johnson's impeachment that is often appreciated or understood. As this case study will show, moreover, the Army and its leaders played a key role in the events surrounding Johnson's impeachment and Grant's subsequent presidency because of the tremendous impact of high levels of political polarization and military prestige on the behaviors of both civilian and military leaders at the time. 

On numerous occasions from 1865-1868, civilian actors sought to leverage the military's prestige for their own partisan gain. President Andrew Johnson as well as his opponents both attempted to pull Grant and other highly prestigious military actors into their respective political camps, and to portray to the American public that certain military leaders were indeed political allies. 

At the same time, military leaders, including Grant himself, engaged in several instances of political behavior that, even by historical standards, violated the principles of civil-military relations. These military leaders did so, I argue, mainly because of their deep personal convictions related to multiple political issues at stake during the period of Reconstruction. These political issues at stake during the period of US Reconstruction, moreover, had obvious and significant moral undertones, as I show in the paragraphs to follow. Military leaders were also very aware of their enormous political influence, and while most of them did not seek to harness their prestige, some, at key points during these years, did.  

I argue that the character of the polarization that prevailed at the time clearly served as a "motive" for senior military leaders to engage in political behaviors that violated the central principles of civil-military relations, and that the enormous level of military prestige likewise provided an "opportunity" that civilian and military actors sought and did leverage. In this front section of the case study, I first set the scene, establishing that polarization and military prestige operated at relatively high levels upon the end of the US Civil War in 1865. Then, using process tracing, I describe how high levels of polarization and military prestige shaped the political behaviors undertaken by civilian and military actors in the events leading to Johnson's impeachment and Grant's nomination to the Presidency.  

## Extremely High Level of Polarization, 1865-1868

It is difficult to state the precise level of polarization that prevailed at the end of the Civil War, but there is no doubt that the level was significantly high. The fact that a Civil War - the only one in US history - had just occurred, primarily over the issue of slavery, provides an immediate basis for this assertion.  

Standard measurements of polarization in Congress, in fact, usually begin with the 45th Congress, which met from 1877-1879 [@jeffrey_b_lewis_voteview_2020]. This is again because of the Civil War and its aftermath. When the first state, South Carolina, seceded from the Union in December 1860, until the Compromise of 1877 effectively ended reconstruction, several southern states either lacked Congressional representation altogether or had been prevented from electing members of Congress on their own, absent of federal interference.^[The Compromise of 1877 was, in essence, a compromise about the 1876 Presidential Election of 1876. Democrats more or less agreed to a compromise such that the Republican candidate, Rutherford B. Hayes, would become the President, while Republicans in Congress agreed that the last of federal troops would depart Lousiana and South Carolina, effectively ending Reconstruction. See @hoogenboom_rutherford_1995, 274-294 for an account of this compromise by one of Hayes' biographers.] Still, the lack of a precise measure of polarization until the late 1870s need not leave the reader wondering whether polarization was high at the end of the Civil War in 1865.  

Polarization was high for several reasons upon the end of the Civil War as the country sought to "bind up the nation's wounds" [@lincoln_lincolns_1865]. These reasons involved the confluence of several factors, each of which is significant in its own right: slavery and the Civil War had just ended, and the nation now had to to chart a new course.^[For a sampling of the rich literature that describes these issues in depth, see @mckitrick_andrew_1988; @green_reconstructing_2010; @sheehan-dean_northern_2014; @smith_stormy_2017; @sheehan-dean_northern_2019.] In charting a new course as a country, however, the US faced enormous questions about the institution of new political freedoms - which ones, and how quickly they should be granted - to more than four million former slaves [@green_reconstructing_2010, 186-193]. These political freedoms in question included the entirely new rights of voting and owning land [@green_reconstructing_2010, 201].  

The US government at the end of the Civil War also faced significant other legal and moral questions pertaining to how to handle and relate to the states that had seceded. Indeed, McKitrick (1988) argues that Reconstruction, at its heart, boiled down to two central questions: "what had the war done to the rebellious states? And what was the new relation of those states - or districts - to the federal government" [@mckitrick_andrew_1988, 96]? It is not possible to unpack all of the differing views regarding how the federal government should proceed with reconstructing the nation, nor the actors enjoined to each of these views. It is sufficient, rather, to state that views on the topic of Reconstruction ranged widely at the time. 

Some politicians, such as Thaddeus Stevens, a Pennsylvania Congressman who bitterly opposed Johnson and who was known as a Republican 'Radical,' believed that the southern states should be administered as "conquered provinces," much in the same way a victor in war might exercise rule over a formerly hostile foreign power and its inhabitants [@mckitrick_andrew_1988, 99-101]. Stevens, and others who held to this view, believed that the states which had seceded, as well as those who had fought for the Confederacy, had forfeited any of the legal rights that they had once possessed under the Constitution, that the entire way of life in the south would have to be forcefully remade, and perhaps most controversially, that private property belonging to southerners should be seized, confiscated, and redistributed [@mckitrick_andrew_1988, 99-101]. 

On the other end of the spectrum of views regarding reconstruction is what McKitrick (1988) calls the "southern theory" of Reconstruction [@mckitrick_andrew_1988, 97-98]. This set of views essentially advocated for a relatively quick restoration of full rights to the former Confederate states on the basis that these states had tried to secede, but had failed, and should be welcomed back into the Union with little or no hindrance [@mckitrick_andrew_1988, 97-98]. True, this set of views argues, several southern states had seceded, but they had also fought, lost, and suffered in the process. Now that the war was over, there was little need for punitive action. McKitrick (1988) also points out that among the most ardent supporters of the southern theory of reconstruction were those who had fought for the union, including General William Sherman, who, after Lee's surrender at Appomattox, infamously offered generous peace terms to Confederate General Albert Sidney Johnston [@mckitrick_andrew_1988, 97-98].^[These terms of peace included a general pronouncement of amnesty for former Confederate officers.] But Sherman quickly encountered political turmoil. Indeed, when Sherman's superior, General Grant, learned what had happened, he intervened and revoked the terms, cognizant that Sherman, by brokering terms of political restoration, had not only crossed into a realm that only statesmen should occupy, but had offered terms that were politically unpalatable [@mckitrick_andrew_1988, 98; see also @simpson_ulysses_1988, 95-100]. 

In the middle of the spectrum of views concerning reconstruction were many others, which for space constraints, cannot be unpacked here [see @mckitrick_andrew_1988, 93-119]. The important thing to underscore is that each of the different views regarding how the south should be reconstructed involved a host of significant moral but also practical issues and beliefs about what actions the federal government should take with respect to the southern states and its inhabitants, including both former confederates and former slaves. Should former confederate officers and soldiers be granted amnesty for their past actions in taking up arms against the union? Should these men be allowed to hold political office, and if so, would they have to denounce their past behavior or declare loyalty to the federal government? How much help should the federal government provide, and how quickly should they provide it, to ensure the implementation of new freedoms to former slaves?^[See @mckitrick_andrew_1988, 120-152 and @green_reconstructing_2010, 186-194 for a concise treatment of these issues]. 

Adding to the weight of all of these major questions was the issue of _who_ was ultimately responsible for overseeing and implementing the nation's reunification. Was the President ultimately responsible, or should Congress set the terms for welcoming the recalcitrant southern states back into the Union after four years of Civil War [@mckitrick_andrew_1988, 103-109]? Although President Lincoln had claimed significant executive powers, some of which rested on perhaps questionable invocations of wartime necessity, Republicans in Congress were not willing to grant Lincoln's successor the same benefit of the doubt, especially in the wake of Lincoln's assassination and because the war, at least in terms of battles fought in the open, had ended [@mckitrick_andrew_1988, 106-107; @hearn_impeachment_2000, 29-32].  

The Army and its senior leaders came to adopt strong opinions about these and related issues, especially as the Army became a direct participant in the process of Reconstruction. Indeed, the Army came to embark on what should be considered nothing short of an occupation of the American south that would last more than a decade. For much of this unprecedented period, Army forces in the south and their commanders took on incredible responsibilities for overseeing civilian government, administering martial law, and exercising authorities similar to those that exist in wartime [@downs_after_2015; @lang_wake_2017]. 

Like the public, however, Army leaders were divided about many of the same issues and questions that divided Congress and the nation regarding the implementation of reconstruction. Some military leaders, including Ulysses S. Grant and Phil Sheridan, came to adopt and practice views that were consistent with those of the Republicans in Congress, and favored strong and consistent involvement by both the Army and the federal government, while others, such as William Sherman and Winfield Scott Hancock, held views that generally favored states' rights and thus, a more limited role for the Army during reconstruction [@milton_age_1930; @hesseltine_ulysses_1935; @jordan_winfield_1988;@simpson_let_1991; @marszalek_sherman_2007; @white_american_2016; @chernow_grant_2017]. At times, the viewpoints that military officers held regarding issues involving Reconstruction served as "motives" for these officers to engage in political behaviors that challenged, and sometimes violated, the central principles of civil-military relations.^[It is clear that in the historiography of this period, biases are present. Some historians, such as Milton (1930) are obviously sympathetic towards Andrew Johnson, and thus very critical of Grant. Others, such as  White (2016) and Chernow (2017) are extremely critical towards Johnson, and thus ebullient in their praise of Grant. Still others, in my view, including Hesseltine (1935) and Simpson (1988) seem to fall somewhere in the middle. In this chapter, I am seeking to remain as neutral as I possibly can, and to present the events that unfolded in this period as objectively as possible. The impeachment of Andrew Johnson, and the Army's closeness to the events leading up to it, was undoubtedly an atypical event. Yet these events also clearly demonstrate, perhaps like none other in American history to date, how high levels of polarization and military prestige combined to serve as critical factors in arguably the country's greatest civil-military relations crisis.] 
  
## Extremely High Level of Military Prestige, 1865-1868

The level of military prestige was also high at the end of the Civil War, and for one central reason: the Army had achieved victory in preventing the dissolution of the nation. Although this author could not locate any quantitative public opinion data indicating the level of public trust or confidence in the military, similar to measures that exist for the latter half of the 20th and the 21st Centuries, there are a number of qualitative indicators that suggest the military was highly esteemed by the public in the northern states which had comprised the Union. This was especially true in the case of individual senior military leaders, rather than for all members who had fought in support of the Union cause, likely because to have served in the American Civil War on either side was not necessarily all that unique. Indeed, historical estimates suggest that upwards of ten percent of the entire country's population at the time served on either side of the war [@noauthor_civil_2012; @noauthor_facts_2021]. 

While it is not possible to list all of the individuals who achieved heroic stature as a result of their martial exploits during the Civil War, it is worth pointing out several, who, in the years after the war, remained in the Army. Among these included General Philip Sheridan, the famed cavalry officer; General William Sherman, known for his ruthless march through Georgia near the close of the war; General Winfield Scott Hancock, who repelled the Confederate charge at Cemetery Ridge at the battle of Gettysburg; and many, many others.^[These include figures such as John M. Schofield, who would play a significant role in the Army's reform efforts late in the 19th Century.] However, while these officers were well known and popular figures among the public, the prestige that accompanied General Ulysses S. Grant, who commanded all Union forces at the close of the war, was truly second to none. 

Grant was the last of Lincoln's battlefield commanders and the one who ultimately brought about the defeat of Lee's Army. Grant had accumulated significant attention in the national press, like many other Union generals, for his martial exploits in the years before Appomattox, as a result of victories at Fort Donelson, Vicksburg, Chattanooga, and Petersburg. A West Point graduate and veteran of the Mexican War, Grant had resigned his officer commission in 1854 as a Captain. However, after several years in failed private business ventures, Grant was recommissioned as a Colonel in the Illinois Militia when the Civil War broke out in 1861, and he quickly established himself as an able fighter and commander [@white_american_2016].

So popular and so recognized by the American public was Grant at the end of the Civil War that Hesseltine (1935) argues Grant "was rapidly on his way to becoming a saint" [@hesseltine_ulysses_1935, 47]. Furthermore, multiple biographers of Grant recount that the general spent several weeks during the late summer of 1865 on a traveling tour of the northern states, where he was fervently met at every stop, and often regaled with large and lavish receptions, parties, and other gifts [@hesseltine_ulysses_1935, 57-62; @chernow_grant_2017, 558-560]. A significant portion of the northern press was also infatuated with Grant. For example, a _New York Daily Tribune_ article written on April 20, 1865, said the following of the general: "Three captains created the art of war as it exists today in Europe - Marlborough, Frederick, and Napoleon. Grant's last campaign entitles him to rank with either of the two former, and in several important elements of military greatness places him on a level even with the great French Captain" [@noauthor_grants_1865]. 

Qualitative historical evidence also demonstrates that other prominent members of the public were cognizant of the high prestige Grant carried with the American public at the end of the Civil War. For example, Horace White, the editor for the _Chicago Tribune_, wrote in a letter to Congressman Elihu Washburn at the end of 1865 that, "I am persuaded that Grant is to be the next President if he gives us half a chance to nominate him" [White quoted in @hesseltine_ulysses_1935, 63]. Chernow (2017) goes so far as saying that in the years after the Civil War, Grant "became the supreme prize in American politics, sought by all parties" [@chernow_grant_2017, 597].

Though the centrality of the military to the nation would fade in the years after the Civil War, it remained quite strong for a time, especially as the complexity and difficulty of the post-Civil War years became evident. Now serving as the commanding general of the Army and with headquarters in Washington D.C., Ulysses S. Grant found himself in the middle of the high stakes political debates about how Reconstruction should proceed, and what policies Reconstruction should entail. The same would be true of many of Grant's subordinate officers, many of whom were also highly prestigious (at least in the north) and well known figures, as they took on important roles and duties related to Reconstruction. 

For the first several weeks, and perhaps months, the relationship between Grant and President Johnson was relatively free of friction, even, as Hesseltine describes, a "political cauldron" was boiling all around [@hesseltine_ulysses_1935, 54]. Over time, however, Grant and several of his subordinate commanders would no longer be able, nor desire, to remain outside of politics. 

In the paragraphs to follow, I recount and describe the political behaviors engaged in by civilian and military leaders from roughly May 1865 - May 1868 that threatened the principles of civil-military relations. I contend that civilian actors engaged in overtly political behaviors that reflected a desire to harness the prestige commanded by several of the military's leaders at the time. I also argue that military actors engaged in political behaviors as a result of _both_ the high level of polarization that existed at the time as well as the high degree of prestige these officers commanded with the public. I begin by describing how the high level of military prestige impacted the political behaviors of civilian and military actors, and then describe the role of polarization - as reflected in the significant moral questions that military actors faced at the time - in shaping the political behaviors engaged in by military actors. 
				
## How Military Prestige Shaped Political Behaviors Involving the Military, 1865-1868

As divisions over reconstruction became evident in 1865 between President Johnson and his opponents in Congress, both sought to bring the military into each's respective camp with respect to how best to restore the southern states to the nation. Johnson and members of his administration, as well as members of Congress, sought General Grant's approval on the specific measures that Reconstruction should entail. Indeed, in the months that followed, as the divisions between Johnson and Congress became increasingly irreconcilable, both Johnson and his opponents in Congress would increasingly wonder, as White (2016) notes, "where did Grant stand" [@white_american_2016, 429]?

### 1865 - Fall of 1866

One of the first political behaviors the President engaged in was sending Grant on a tour of the South in late 1865. The purpose of sending Grant to the South was to refute a previous report made by Carl Schurz, a journalist and former general whom Johnson had also sent to assess the general political atmosphere of the southern states. Schurz's report, however, was politically problematic for Johnson because it portrayed southern attitudes as "angry and defiant," and indicated that former slaves who had been freed were "languishing in wretched conditions of poverty, reinforced by Black Codes that trapped them in a new subservience" [@chernow_grant_2017, 563]. In short, Schurz's report did not describe a south that was remorseful and prepared to rapidly reunify with the nation, but rather one that required a lengthier and more robust federal presence [@chernow_grant_2017, 563]. 

Johnson hoped that Grant, after visiting the south, would instead write a report that argued the political conditions in the south favored embarking on a relatively quick process of restoration, in line with Johnson's views on Reconstruction [@hesseltine_ulysses_1935, 60; see also @chernow_grant_2017, 563-564]. If the President could obtain such a report by Grant, a highly influential and prestigious military figure, then Johnson could employ the report as a political weapon against his opponents in Congress. This is indeed the course of action Johnson took. After Grant returned from the visit, Johnson summoned Grant to a cabinet meeting where the general "expressed his conviction that every consideration called for the early reestablishment of the Union [@hesseltine_ulysses_1935, 61]. Grant's report was then sent off to the Senate, and used by Johnson's administration to defend against attacks from Radical Republicans [@hesseltine_ulysses_1935, 61; @chernow_grant_2017, 565-566].^[Chernow (2017) argues that Grant's visit was superficial in that it was rushed and largely featured sham interactions with southern whites who praised Grant. Chernow also notes that Grant regretted his report soon after furnishing it. See @chernow_grant_2017, 565-566 for more details. Furthermore, both Hesseltine (1935) and Milton (1930) argue that in the wake of Grant furnishing his report, Radical Republicans fervently tried to get Grant to denounce it. See @hesseltine_ulysses_1935, 65; @milton_age_1930, 290 for more details.] 

In sending Grant to the southern states, the President violated at least two of the central principles of civil-military relations. Johnson's behavior in dispatching Grant broke the principle of non-partisanship because there was, effectively, no possible way that Grant could generate a report that did not anger either Johnson's supporters or his opponents in Congress. Indeed, Johnson's chief purpose in sending Grant was provide both the American public and Congress the impression that Grant's views on Reconstruction favored his own. Johnson knew that if Grant were to validate and approve of his own political views, his opponents in Congress would be disadvantaged, if even for a short period of time.  

But Johnson's dispatching of Grant to the south also broke the principle of non-interference of the military because it had the effect of making a sitting military commander investigate, evaluate, and publicize prevailing political conditions in ways that were comparable to what a reporter or journalist does. To see this, we have to consider that Grant was sent for the primary purpose of refuting a report that already been made by Carl Schurz, a career journalist who in 1866 would become the editor of the _Detroit Post._^[Schurz was also a former Union Army general, but it is important to note that he was a volunteer who had attained the rank of general, rather than a career Army officer who became a general. This matters because as Schurz first made his trip to the south, he was corresponding with the press as a reporter, not in the capacity of a former Army general. See @simpson_let_1991, 116-188 for more details.] Grant and his Army entourage that went to refute Schurz's report did not possess any notably superior skills, neither as investigators nor as reporters. Moreover, if anything, it is almost certain that Grant's popularity prevented him during his trip from accessing areas and people who would have evinced far less rosy of a situation than Grant made [@hesseltine_ulysses_1935, 60]. In asking Grant to fill a role that was initially and better suited for a press correspondent, President Johnson compelled the nation's top Army officer to assume a role that was tangential, if not altogether unrelated, to his role as the general-in-chief of the Army.^[If one counters on the grounds that Grant had a network of military commanders in the south who knew more about what was transpiring in the south than perhaps press correspondents, and thus, that Johnson did not violate the principle of non-interference through his actions, it should be emphasized that Johnson did not send Grant with the motivation to genuinely learn what was happening in the south. This would, I agree, not have been problematic from the perspective of the principle of non-interference of the military, as military commanders are obligated to advise the president, which in part, includes keeping the president informed of developments that occur. But the historical record indicates that this is not what Johnson intended when he sent Grant to refute Schurz's report. Instead, the record shows that Johnson already knew of the troublesome events beginning to transpire in the south, and that he sought Grant's prestige to gain an advantage over his opponents in Congress.] Indeed, Johnson could have dispatched others to refute Schurz's report, including perhaps other press correspondents who likened to his views. Instead, in sending Grant, Johnson sought to invoke Grant's enormous prestige, knowing that both the public and Congress would be interested in Grant's subsequent evaluation.^[A reasonable observer might also conclude that Johnson's behavior also violated the principle of civilian control in that dispatching Grant, the President delegated too much power to the military. I concur with this sentiment in general, but I am wary of wading into the specific debate about whether the military held (or was delegated by civilians) too much power during Reconstruction. This is an important question, but one that is difficult to answer, particularly in retrospect. I do take this on in the conclusion of this section of the chapter, but desire to point out here that it is better, in my view, to emphasize that in many ways, what Johnson and his opponents were wrestling over Reconstruction was precisely the question of how much power the military should hold during a particular era. This means, essentially, that there was little if any difference between political policy and military policy during Reconstruction. Several scholars point this out using slightly different terms. Downs (2015) makes the compelling point that in many ways, American reconstruction was a fight between Johnson and Congress over who - the President or Congress - has the power to end a special type of (Civil) war, and the Army's role in such a questions. See @downs_after_2015, 7-8 for an overview. Hyman (1960) is perhaps even more stark. He argues that in essence, two armies existed when the Civil War ended; one was controlled by the President, and had duties on the American frontier and along the US border with Mexico. The other army is the one that implemented reconstruction in the south. This army, Hyman argues, is the one that Congress would never allow President Johnson to control. The stakes in reconstruction were simply too high. See @hyman_santon_1960 for more details, especially pp. 86-87.]

After Grant's report became public knowledge, the final months of 1865 and early months of 1866 evinced even greater divisions between Johnson and his Republican opponents in Congress. As time passed, it became increasingly clear that Johnson's conception of Reconstruction clearly favored southern whites with ties to the former confederacy, and at the same time, disadvantaged both former slaves and whites who had supported the union. For instance, newly freed blacks frequently faced increasing danger as southern state and local governments enacted a series of "black codes" that authorized a variety of discriminatory measures against blacks and former slaves [see for example, @mckitrick_andrew_1988, 204-205; @chernow_grant_2017, 571]. 

Though the 13th Amendment (which formally abolished slavery in the United States) went into effect in December of 1865, Johnson continually vetoed legislation that sought to expand the rights and political freedoms of the former slaves.^[Again, the reasons for Johnson's vetoing these bills involve significant differences in legal and constitutional views, which I do not take up here. There is no doubt that by today's standards, Johnson held racist views, but it would be an oversimplification, at least in my opinion, to entirely account for Johnson's vetoing these pieces of legislation as stemming from his views on race. His racist views played perhaps the dominant part, and should not be ignored, but the role of other factors, to include Johnson's legal convictions about the Constitution and its powers, are worth considering, if for no other reason because they illustrate the complexity of both the task and the debates concerning American reconstruction. See @mckitrick_andrew_1988 for more details.] These included bills which sought to extend the life of the Freedmen's Bureau, an organization created in 1865 to "provide food, shelter, clothing, medical services, and land to displaced southerners, including newly freed African Americans" [@noauthor_us_nodate], and later, a civil rights bill which, while "silent on voting rights," nonetheless sought to make the federal government the "guarantor of basic liberties" [@chernow_grant_2017, 570].^[Both of these pieces of legislation would eventually end up passing as Congress voted to override vetoes by Johnson.]

With these legislative battles between Congress and the President unfolding, Johnson engaged in another obvious political behavior involving the military ahead of the upcoming midterm elections of 1866. Despite the already noteworthy shortcomings of Johnson's Reconstruction policies up until that time, Johnson sought to show the nation that his approach towards the south was the right one, and that Grant, his prestigious military commander, thought so too. Though Johnson was not up for reelection in 1866, conservative Republicans and moderate Democrats - who favored Johnson - were. In a very real way, then, the midterm elections were a referendum on Johnson's policies in the south.  

Johnson and other civilian leaders attempted to harness General Grant's prestige and that of other military leaders for partisan electoral gain in the months leading up to the elections, in ways both large and small. As an example of a relatively small political behavior, on several occasions, civilian actors deliberately orchestrated their public appearances in such a way so as to physically appear alongside of popular military figures so as to give the impression that military leaders approved of the policies of these civilian actors. For instance, on August 18, 1866, President Johnson hosted "a delegation from the National Union Convention in Philadelphia" at the White House [@chernow_grant_2017, 576]. In spite of the clearly partisan nature of the gathering, Grant reluctantly accepted Johnson's invitation to attend, as he "felt duty-bound" to do so [@chernow_grant_2017, 576]. Ahead of speaking to his supporters in the East Room of the White House, the President summoned Grant to stand next to him during the entirety of his remarks [@chernow_grant_2017, 577]. After the event, Grant supposedly returned to his headquarters, where he was "full of indignation" at having been manipulated by the President [Badeau, 1877, as cited in @chernow_grant_2017, 577]. Ultimately, Grant's physical proximity to the President on that particular August evening insinuated, at least to one reporter present at the event, that Grant fervently endorsed Johnson's policies [McFeely, 2002, as cited in @chernow_grant_2017, 577]. 

A far greater instance of Johnson and other civilian leaders harnessing Grant's prestige and that of other military leaders occurred at the end of the summer of 1866, when Johnson insisted that Grant and several other military figures accompany him on his "Swing Around the Circle," a two week long public relations tour of several Northern states and populous cities. Traveling by train and designed to drum up support ahead of the elections, the President brought with him several of his cabinet members, but also several military figures, including General Grant, Admiral David Farragut, and General George Meade. Much the same way the President had done at the White House earlier in August, appearing alongside popular military figures in overtly partisan settings intended to convey to the public that these military figures approved of Johnson's political policies, which included strong denunciation of the 14th Amendment, which the Senate had passed previously in June [@hesseltine_ulysses_1935, 74-76; @chernow_grant_2017, 577-581].^[The 14th Amendment granted citizenship to all persons born in the United States, including former slaves. Except for Tennessee, the southern states rejected the 14th Amendment initially, but the Military Restoration Acts of 1867 later required ratification of both the 13th and 14th Amendments for states to be restored to the Union.]

By almost all accounts, the President's trip turned out to be a public relations disaster. Johnson often argued with the crowds at almost every stop, and many of those who turned out asked instead to hear from Grant and Farragut rather than from the President [@hesseltine_ulysses_1935, 75; @chernow_grant_2017, 577]. Interestingly, Milton (1930, pp. 362-369), Hesseltine (1935, pp. 74-75), and Simpson (1991, 148) argue that the crowds who cheered for Grant and Farragut and those who heckled Johnson did so as a result of deliberate orchestrations devised by Radical Republicans in Congress and their allies in the press, who, upon learning that several well known military figures would accompany Johnson on his speaking tour, devised a plan to prevent Johnson from capitalizing on the prestige these military figures brought. If these explanations are valid, even in part, then it is clear that not only Johnson, but also Johnson's opponents, acted in ways that sought to appropriate the prestige of military figures for their own partisan advantages. In essence, Johnson and his opponents competed, at times in ways that were not subtle at all, for the military's prestige. Actions to harness the prestige of military figures by one civilian actor were met by counteractions to prevent this from occurring by another civilian actor. 

This political maneuvering at the hands of civilian actors - Johnson but also his opponents in Congress - for the prestige of particular military actors violated the principle of military non-partisanship. Johnson's behavior at the White House in August, as well as his insisting that several military figures accompany him on his "Swing Around the Circle," and the behavior by Johnson's opponents when they learned of these developments, were all undertaken for the purpose of influencing public opinion. These actors believed that portraying particular military leaders as supportive of their own policies related to Reconstruction would prove politically beneficial, particularly ahead of the 1866 midterm elections. ^[Some historians remark that in part, the trip likely marked somewhat of an internal turning point for Grant, personally, in terms of his internal convictions regarding his conduct vis-a-vis the central principles of civil-military relations. For instance, Hesseltine (1935, p. 75) argues that accompanying the President on this particular trip enabled Grant to draw conclusions regarding how unpopular the President and his policies were.]

### Late 1866 - Johnson's Impeachment in early 1868

As the nation's political climate continued to worsen from the fall of 1866 until Johnson's impeachment early in 1868, Congress and the President were at odds to the point that many feared open violence might erupt [@hesseltine_ulysses_1935, 76-77]. Significant race riots in 1866, first in Memphis and then in New Orleans, had killed dozens of freed blacks and sympathetic whites [@mckitrick_andrew_1988, 422-427; @hyman_santon_1960, 90]. Combined with continued issues regarding what authorities and jurisdictions military commanders in the south possessed, to include whether commanders could declare martial law and oversee matters pertaining to civilian governance, partisan battle lines had been more less drawn by this point, even among prominent military officers: along with the Secretary of War, Edwin Stanton, Generals Grant, Sheridan, Pope, and Meigs clearly favored views of Reconstruction held by the Republicans in Congress, while Generals Sherman and Hancock favored Johnson's [@hyman_santon_1960, 93]. 

During this period, Johnson again engaged in several blatant political behaviors that violated the principles of civil-military relations. The most obvious of these involved efforts to minimize the influence and in some cases, replace, key military commanders on the basis of partisan convictions rather than military excellence. Johnson was largely unsuccessful in these efforts, and his eventual impeachment occurred for attempting to do the same with Stanton, his civilian Secretary of War. Nonetheless, his attempts to fill key positions with military actors loyal to his political views shows how the combination of a highly polarized environment and a highly prestigious military contributed to the undertaking of actions that ran counter to several of the central principles of civil-military relations.   
The first of Johnson's schemes involved sidestepping the powerful influence of the combined team of Grant and Stanton, who by this point (early 1867), historians are quick to point out that while not being friends, had both come to believe that Congress, not the President, had the Army's best interests at heart with respect to Reconstruction [@hyman_santon_1960]. Too prestigious to dismiss outright, Johnson sought to send Grant to Mexico as part of a diplomatic mission, and instead, to replace Stanton with General Sherman, who Hyman argues was the "nation's third most popular man" [@hyman_santon_1960, 94]. Johnson apparently considered making the move official in conjunction with leaking to the press a private letter he had received from Sherman earlier in the year which contained support for his policies [@hesseltine_ulysses_1935, 78]. 

Johnson's scheme floundered, and neither Grant nor Stanton (this particular time) were replaced. Though Sherman was sympathetic to Johnson's views on Reconstruction, Sherman could not bring himself to be part of Johnson's political schemes. Grant simply refused Johnson's directive to go to Mexico, and in a rather dramatic manner. In front of an assembled cabinet, Johnson asked the Attorney General whether his orders to send Grant to Mexico were illegal, ostensibly in order to publicly shame Grant into complying with the order. Hesseltine (1935) depicts Grant as standing up before the Attorney General could respond, and stating, "I can answer that question Mr. President, without referring to the Attorney General. I am an officer of the army, and bound to obey your military orders. But this is a civil office, a purely diplomatic duty that you offer me, and I cannot be compelled to undertake itno power on earth can compel me to it" [Badeau, 1877 as cited in @hesseltine_ulysses_1935, 79]. 

Though Johnson decided not to press the issue of sending Grant to Mexico any further, the entire incident, including Grant's refusal to obey Johnson, reveals how a high level of military prestige shaped both civilian and military actor behavior. To begin with, if replacing Stanton had even a chance to work in the first place, he would have to be replaced by someone who was highly regarded by the public, which explains Johnson's selection of Sherman to replace Stanton. Appointing an officer to replace Stanton with considerably less renown than Sherman would likely not have been as successful, politically. Though Sherman was known to support Johnson's view of Reconstruction over that of the Republicans in Congress, Johnson's opponents in Congress would have likely accepted Sherman as the Secretary of War on the grounds that he was nonetheless a well known and highly regarded general. 

Grant's refusal of Johnson's directive to go to Mexico is also instructive, however. Grant stated in effect that the order was illegal, and that he therefore did not have to follow it. Although Johnson's directive to send Grant to Mexico was clearly done in order for the President to obtain a political advantage in that it would remove Grant from interfering with Johnson's political wishes, I am not convinced that the order was itself explicitly illegal. It is more likely the case that in refusing Johnson's order, Grant knew that he could in that both Republicans in Congress, as well as the public, would back him. There is no doubt that Grant wished this incident not to have transpired in the first place, but nonetheless, Grant knew that the prestige he carried rendered him politically unassailable, even as he refused the President's order.  

On March 2nd, 1867, Congress passed (again, over a Presidential veto) the first of several Military Reconstruction Acts. This legislation split the south into five military districts, each of which would be overseen by an army general. The legislation intended for these military commanders to hold enormous sway over the daily conduct of state and local governments, which were to be regarded by the commanders as "provisional" in nature, pending their "full" restoration to the union [Grant, 1867 as cited in @chernow_grant_2017, 587]. The legislation also dictated that all orders to the Army from the President had to be routed through Grant [@chernow_grant_2017, 587].^[The Military Reconstruction Acts also empowered Grant to appoint these five generals himself, but even Grant had reservations about this, believing that the President, as the Commander in Chief, should have a strong say in who these generals are. See @hesseltine_ulysses_1935, 83 and @chernow_grant_2017, 585 for more details.] At the same time, Congress also enacted the Tenure of Office Act, which sought to preserve Stanton's place in Johnson's cabinet, and thus his link to Republicans in Congress [@hearn_impeachment_2000, 110-119; @chernow_grant_2017, 586-587]. This piece of legislation specifically declared that the President could not remove cabinet officials that he had not appointed himself without the Senate's approval [@chernow_grant_2017, 586].^[Chernow (2017, p. 586) notes that Grant personally felt the Tenure of Office Act was "ridiculous" in that it essentially rendered the President powerless, but that he nonetheless sought to obey the law that had been passed.] Indeed, Johnson had inherited Stanton, who had also served as Lincoln's Secretary of War. Together, the Military Reconstruction and the Tenure of Office Acts effectively cut off the President from exercising meaningful control of the Army in the south. 

As Congress passed a second and third Reconstruction Act in March and July of that year, the tension between the President and Congress continued to significantly escalate, and the Army found itself right in the middle of it all. In August 1867, the situation reached a boiling point. The President first informed Stanton that he was suspended (not "removed," so as to avoid being instantly in violation of the Tenure of Office Act), and proceeded to appoint Grant as the Secretary of War _ad interim_ until a permanent replacement would be named [@hyman_santon_1960, 98]. Weeks later, the President then removed General Philip Sheridan, who commanded the 5th Military District (Lousiana and Texas), after Sheridan removed the Governors of both Lousiana and Texas [@white_american_2016, 448-449; @hyman_santon_1960; 98-99]. Sheridan, a favored subordinate of Grant's who favored an aggressive Republican policy towards Reconstruction, had helped register thousands of blacks to vote in the district, earning widespread condemnation from Johnson and southern whites. Johnson also later removed the commanders of the 2nd and 3rd Military Districts, Generals Daniel Sickles and John Pope, for similar reasons. Johnson's replacement commanders were all known to be sympathetic to Johnson's views on Reconstruction, especially General Winfield Scott Hancock, who replaced Sheridan in the 5th Military District.  

In early January of 1868, the situation involving Johnson, Grant, and the Army moved towards a climax. The Senate first released a report refusing to concur with Stanton's suspension, which in turn raised questions as to whether Grant's serving as the _ad interim_ War Secretary was legal or not [@mckitrick_andrew_1988, 501-505]. Grant did not wish to find out the answer, and thus he attempted to vacate the office, which Johnson would not let him easily do [@mckitrick_andrew_1988, 501-502].

Historians, it seems to this author, have for years attempted to ascertain exactly what sort of understanding Grant and Johnson came to regarding Grant's continued service as the _ad interim_ War Secretary, and when such an understanding was reached. Those who are sympathetic towards Johnson argue that the two men agreed that Grant would either keep the office, even if the Senate or the Supreme Court would hold proceedings to decide the legality of Stanton's removal or that of the Tenure of Office Act, or that Grant would vacate the office sufficiently ahead of time so as to allow Johnson to name a replacement, and thus, to not be found in violation of the law [@milton_age_1930, 481-485]. Other historians, mainly those who are more charitable towards Grant, take a different view [@chernow_grant_2017, 602-607]. Regardless of which point of view is historically accurate, it is clear that following another heated cabinet meeting in January of 1868, Johnson's administration engaged with the press in multiple efforts to discredit Grant and to portray him as willfully acting in a "duplicitous" manner [@chernow_grant_2017, 605; @milton_age_1930, 483-485; @mckitrick_andrew_1988, 503; @hesseltine_ulysses_1935, 106-107]. For several weeks, the nation hung on these and other reports that detailed what was happening within Johnson's administration. These final but significant political behaviors of leaking information and giving interviews to the press by the Johnson Administration were intended to discredit Grant and, in the process, to boost Johnson's political capital. 

Leaking this information to the press and following up with interviews about what had transpired within the cabinet primarily violated the principles of civilian control and non-partisanship. These political behaviors violated the principle of civilian control because they reflected that by this point, the Administration was no longer attempting to exercise oversight of the military through any established mechanisms. Indeed, to be fair to Johnson and also historically accurate, the regular mechanisms of civilian oversight available to the Johnson Administration (or any Presidential Administration beforehand) had been removed by Congress, effectually making the Commander in Chief an observer of, rather than a participant within, civil-military relations. Leaking and giving interviews to the press by civilian actors to discredit Grant also violated the principle of non-partisanship, however. While it was well known that Grant differed with the Commander in Chief over Reconstruction policies by this point, the public nature of the leaks and interviews given by the Johnson Administration only served to publicize these differences. Ultimately, the House of Representatives voted to impeach Johnson in March of 1868, and Johnson went on to avoid conviction in the Senate by a single vote following his trial in the Senate in May. Grant went on to easily capture the Republican nomination for President the same month. 

From 1865-1868, Johnson and Congress sought to leverage the prestige of key military leaders over policies related to Reconstruction. Both Johnson, members of Congress, and their allies engaged in behaviors either to court approval with highly influential military actors or to give the impression of such approval to the public. These political behaviors ranged from the use of prestigious military figures for specific instances of political gain, such as when Johnson sent Grant to the American south to refute Carl Schurz's initial observations, to orchestrating partisan appearances in public alongside military figures, such as when Johnson brought several military actors with him on his "Swing Around the Circle Tour" ahead of the midterm elections in 1866, to attempting to discredit military figures in the press, such as when Johnson's Administration leaked the content of discussions between himself and Grant during a heated cabinet meeting in January of 1868. In other instances, military figures undermined the central principles of civil-military relations by asserting the prestige that they carried. This occurred perhaps most recognizably when Grant, arguably the nation's most prestigious figure, refused Johnson's directive to be sent to Mexico. The consistent and public nature of political behaviors undertaken by civilian and military actors during the period from 1865-1868 strongly suggest that high military prestige played an important role compelling both sets of actors to engage in these behaviors. 

## The Role of Polarization on Political Behaviors Involving the Military, 1865-1868
			
In this section of the chapter, I briefly show how the high level of polarization impacted the political behaviors engaged in by military actors in particular from 1865-1868. As the preceding paragraphs made clear, military actors were confronted with choices and decisions that they could not avoid about the nature of Reconstruction, especially as they formed conclusions about events unfolding in the south. 

As I have described above, the political questions Army officers faced during Reconstruction involved significant moral components. The political problem of Reconstruction was thus inseparable with questions regarding the restoration of one's fellow citizens to the nation, the treatment and awarding of political freedoms to former slaves, and the legacy of a painful and costly Civil War. Over time, these moral concerns, all of which were or would become embedded in policies involving Reconstruction, weakened the constraining influence of the central principles of civil-military relations on military actors. Military actors thus engaged in political behaviors that violated the central principles of civil-military relations, but nonetheless advanced the moral views to which they subscribed. 

It would be difficult to capture all of the instances in which military commanders violated the principles of civil-military relations as a result of moral convictions reflected in polarized debates during this period, simply because there would be too many to list. Nonetheless, even a few noteworthy examples clearly show that military actors could not, and ultimately did not, ignore the moral undertones of Reconstruction. Thus, in a very real and discernible way, the high polarization present in this historical period provided a _motive_ for military actors to act in ways that violated the principles of civil-military relations. 

In June of 1867, for instance, following General Sheridan's removal of the governors of both Texas and Lousiana from office, President Johnson, before removing Sheridan, had his Attorney General, Henry Stanbery, produce an opinion strongly denouncing Sheridan's actions and the Reconstruction Acts passed by Congress up to that point altogether [@hesseltine_ulysses_1935, 85; @chernow_grant_2017, 589]. Johnson ordered Grant to distribute Stanbery's opinion to his commanders, and indeed, Grant complied; however, Grant also let his commanders know "they could freely interpret it as they chose" [@chernow_grant_2017, 589]. Grant's behavior in this instance constitutes a form of selective compliance with a directive, and constituted a violation of the principle of civilian control. 

After Stanbery's opinion was published, Sheridan then wrote Grant a letter in which Sheridan upbraided the President [@hesseltine_ulysses_1935, 85-86]. The letter was leaked and published, though it is not clear who was responsible for the leak. The President ultimately decided not to punish Sheridan for the letter on the grounds that it had been written, ostensibly at least from Sheridan's perspective, as a private rather than public letter [@hesseltine_ulysses_1935, 85-86]. However, Republicans in Congress were greatly happy with Sheridan's criticism of the President, and used it as grounds to make further overtures towards Grant [@hesseltine_ulysses_1935, 85-86]. Sheridan's letter to Grant was not intended as a public appeal, but his letter to Grant nonetheless ended up having precisely such an effect. 

Then, after Johnson finally removed Sheridan, an action that Grant greatly protested, Grant told his commanders not to reinstate any of the southern politicians Sheridan had previously removed [@chernow_grant_2017, 596]. The President had first asked for Grant's opinion on replacing Sheridan, to which Grant responded in a private letter that it was "unmistakably the expressed wish of the country that General Sheridan not be removedthis is a Republic where the will of the people is the law of the land. I beg that their voice may be heard" [Grant, 1867 as cited in @milton_age_1930, 458]. After Johnson removed Sheridan anyways, Grant then furnished the contents of the previously private letter to several press outlets, including the _Army and Navy Journal_, which was popular with military audiences [@chernow_grant_2017, 596]. By directing his subordinates not to reinstate any politicians Sheridan had removed, Grant obstructed the President's wishes, thus violating the principle of civilian control. By then leaking a letter to the press, his behavior further undermined the principle of civilian control but also, due to its likely applaud among Republicans in Congress, further challenged the principle of non-partisanship. 

By engaging in these political behaviors, Grant and Sheridan could no longer ignore the trajectory of Johnson's policies on Reconstruction. The institution of Black Codes in many southern states, the occurrence of race riots in Memphis and New Orleans, and the seemingly quick restoration of southern politicians who had advocated rebellion with the union just a few years before, were all met with indifference, if not support, by the President. These developments that occurred within 18 months of the formal end of the Civil War signaled to many senior military leaders that Johnson's Reconstruction policies would ultimately squander that which the nation and its Army had so significantly sacrificed during the Civil War [@hyman_santon_1960, 87; @benedict_new_1973, 351]. These moral convictions led to the formation of what can be considered an alliance of sorts between Grant, Stanton, many of the 'Radical' Republicans in Congress, and several of Grant's subordinate military commanders, all of whom shared the fundamental belief that the Army needed to be more heavily involved in, rather than standing more aloof from, Reconstruction policy in the South. 

Some historians note that Grant in particular became more openly opposed to the President and his policies following the "Swing Around the Circle Tour" in the late Summer of 1866, when Grant clearly perceived how unpopular both Johnson's policies and the President himself were with the American public [for instance, see @hesseltine_ulysses_1935, 75]. This is an important consideration in some respects, especially insofar as the political philosophy that Grant would exercise as President was one that viewed the Executive's office as the executor of the "will of the people" [@hesseltine_ulysses_1935, 93]. 

I do not doubt that Grant's behavior, and perhaps that of other military commanders, took on a tone that was more obviously against the President upon Grant learning firsthand how unpopular the President was. But I believe that Johnson's lack of popularity with the American public was not the primary factor compelling Grant and other military leaders to selectively comply with orders, or to challenge presidential directives, or form alliances with Republicans in Congress. The historical record as a whole seems to portray Grant and other military leaders becoming increasingly convinced over time that significant and irreconcilable moral differences existed between the President's views of Reconstruction and their own, and that these differences would result in the nation following altogether different trajectories. 

In other words, it seems much more likely, based on the historical evidence, that Grant and many of his subordinates did not especially wish to confront the President through their actions by selectively complying with directives, nor did they do especially wish to form political alliances with the President's opponents in Congress, but that they did so only after concluding that doing so was a moral imperative. The purposes for which the Civil War had been fought, Grant and many of his subordinate commanders believed, stood in opposition to President Johnson's policies involving Reconstruction.

## Summary of the 1865-1868 Period

A high level of military prestige provided civilian and military leaders an opportunity to engage in political behaviors involving the military from 1865 - 1868. The high level of polarization prevalent at the same time likewise provided military actors a motive for engaging in behaviors that challenged the principles of civil-military relations. A summary of these political behaviors engaged in by both sets of actors from 1865-1868 is captured in Table \@ref(tab:pol-behav-cs1).

```{r pol-behav-cs1, echo=FALSE, include=TRUE, error=FALSE, message=FALSE}

text_tbl1 <- tribble(~"Actor", ~"Tactic Political Behavior", ~"Instance", ~"Central Principles Challenged",
  "President Andrew Johnson (Civilian)", "Partisan Stacking", "Andrew Johnson replaces Generals Sheridan, Sickles, and Pope as Military District Commanders", "Non-partisanship",
  "President Andrew Johnson (Civilian)", "The Military as Sales Reps", "Andrew Johnson directs Generals Grant and Meade, along with Admiral Farragut, to accompany him on a tour of the nation ahead of the 1866 midterm elections", "Non-partisanship",
  "President Andrew Johnson (Civilian)", "Brass in the Cabinet", "Johnson seeks to replace Stanton as Secretary of War with General Sherman", "Non-partisanship",
  "President Andrew Johnson (Civilian)", "Public Appeal", "Johnson leaks the content of a cabinet meeting to the Press, and follows up with several interviews about it in early 1868", "Civilian control; non-partisanship",
  "General Grant (military)", "Selective Obedience/Compliance", "Grant publishes Attorney General Stanbery's legal opinion of the Military Reconstruction Acts in conjunction with President Johnson's wishes, but also instructs his commanders to make their own interpretations", "Civilian control",
  "General Sheridan (military)", "Public Appeal (perhaps inadvertent)", "Sheridan writes a letter to Grant in which he upbraids the President's views on Reconstruction", "Civilian control; non-partisanship",
  "General Grant (military)", "Explicit Insubordination", "Grant refuses to be sent to Mexico by Johnson", "Civilian control",
  "General Grant (military)", "Public Appeal", "After Johnson removes Sheridan, Grant leaks a private letter he had sent to Johnson in which he protested the move", "Civilian control; non-partisanship"
  )
 

text_tbl1 %>% kable(booktabs=T, caption="Political Behaviors Involving the Military, 1865-1868") %>%
  kable_styling(full_width = F, font_size=8, latex_options="striped" ) %>% 
  column_spec(1, width="10em") %>% 
  column_spec(2, width="12em") %>% 
  column_spec(3, width="16em") %>% 
  column_spec(4, width="14em") 

```

# Part 2 - The Army, Domestic Disturbance, and Failed Reform Efforts, 1877 - 1881

Johnson's impeachment and Grant's subsequent ascendance to the Presidency ensured that the Army would play a pivotal role in the political life of the American South during the Reconstruction period. Eventually, this role came to an end. when following the contested Presidential Election of 1876, President Grant pulled the last remaining federal troops from the south. This was the result of a deal struck between Republicans and Democrats that ultimately bestowed the White House to the Republican candidate, Rutherford B. Hayes, in exchange for the return of southern local rule free of federal military interference [@woodward_reunion_1991, 6-8; @marszalek_sherman_2007, 428; @wooster_united_2021, 242].

From 1877 - 1881, the regular Army encountered a number of difficult circumstances amid a tense and uncertain political environment. To begin with, the Army encountered significant friction as it looked for ways to institute a number of reforms to shape the institution for the next conflict [@clark_preparing_2017].^[Clark (2017) tells the full story of the Army's efforts to reform after the Civil War in his book, _Preparing for War: The Emergence of the Modern U.S. Army, 1815-1917_. See especially his fourth chapter on this point, "The Civil War's Legacy."] The 1870s and 1880s would proved to be difficult times for Army leaders to convince Congress and the nation that significant reforms were required, particularly in the absence of a compelling international threat.^[Grandstaff (1998) and Clark (2017) conclude argue that the late 1890s and early 1900s, after the Spanish-American War, is the period in which the Army became more successful at gaining support for their intended reforms. See @grandstaff_preserving_1998, 537 and @clark_preparing_2017 for more details.] A second difficult issue the Army faced after the Civil War involved apprehensions about its domestic role. The political tasks carried out by the Army during Reconstruction proved deeply unpopular with many Americans, and with southern Democrats in particular [@byler_civil-military_2006, 6-11]. Thus the Army had to navigate entrenched partisan forces that, in the wake of Reconstruction, sought to define the terms under which the Army could be and should be used domestically, especially during a number of domestic labor disputes in the late 1870s and early 1880s that involved federal military forces. Inherent in each of these difficult circumstances was the polarized nature of late 19th Century US politics altogether. Indeed, as the United States industrialized and its population expanded, it continued to wrestle with a number of significant political questions over issues such as immigration, the treatment of native Americans, and how to best continue to ensure civil freedoms to millions of former slaves in the southern United States [@azari_back_2016]. 

In this section of the chapter, I argue that in the midst of these still very difficult political circumstances, the military and its leaders did not pose as significant of a political influence as they had in the immediate years after the Civil War, from 1865-1868. As a result, civilian and military actors engaged in generally fewer and less public instances of political behaviors that challenged the principles of civilian control, non-partisanship, and non-interference of the military. 

I first set the scene by briefly showing that the level of polarization remained high during these years, from 1877-1881. I then trace the decline in the relative prestige of the military as an institution from 1868-1877. I make the point that this historical period in particular accorded with a decline in the prestige of the military as an institution, even while a number of individual military figures retained an outsized level of personal prestige on the basis of their martial exploits during the Civil War.   

I then narrate several significant events that involved the Army from 1877 - 1881, including the use of the Army in several domestic labor disputes and bureaucratic fights. I make the case that the general qualities of the political behaviors engaged in by civilian and military actors from 1877 - 1881 were noticeably different than those which both types of actors had engaged in from 1865 - 1868. In particular, civilians did not engage in the same intensity of efforts to harness or otherwise leverage the prestige of the military or its leaders for their own partisan advantage because the military offered a relatively smaller political advantage to civilian actors. This pattern in political behaviors prevailed, even as military leaders cared deeply about many of the issues at stake at the time, because they too were cognizant that they represented an institution which, at the time, carried diminished political clout. 

## High Level of Polarization, 1877-1885

The level of polarization remained quite high as Reconstruction ended. When we view Figure \@ref(fig:polar-chamber), we can see that after a couple of very small drops in the level of polarization in the US House, polarization only climbed beginning in the mid 1880s all the way until after the start of the 20th Century. Polarization in the Senate reflects a very similar trajectory. 

Scholars have noted that polarization in the 21st century shares more than just a couple of passing similarities with the polarization that prevailed in the late 19th century. These scholars make the point that in the late 19th issues related to "race, morality, immigration, and themes related to populism" contributed to American polarization at the time [@azari_back_2016, 93]. These themes remained salient as the United States emerged from Reconstruction and embarked on a process of significant industrialization throughout the decades of the late 19th Century [@skowronek_building_1982]. In short, the level of polarization remains relatively static across the entire period examined in this case study, from 1865 - 1881 (and beyond).

## Declining Military Prestige, 1868-1877 

However, the relative level of military prestige declined from 1868 - 1877. This decline was not the result of a single event, but rather the confluence of several trends that converged on the Army during the Reconstruction period and immediately afterward. These interrelated trends include the downsizing of the Army after the Civil War; the perception of many Americans of the Army as a result of its involvement in Reconstruction and the assumption of constabulary duties by military forces on the American frontier; and finally, the lack of an emergence of a clear international threat for military forces to prepare against. The combined effect of all these factors on the relationship between the Army and American society was that as an institution, it was not as central to the life of the nation, nor as respected, as it had been at the close of the Civil War.

When the Civil War ended, the Army downsized significantly, and it would continue to downsize for roughly a decade. Within half a year of the end of the Civil War, "800,000 of the 1 million men in the Union's Grand Army of the Republic were mustered out of federal service" [@laurie_role_1997, 21]. Four years later, by 1869, the Army's authorized end strength stood at 37,313 [Utley, 1974 and Cooper, 1986 as cited in @clark_preparing_2017, 99]. By the end of 1875, the Army was capped at 25,000 men, including officers and enlisted men [@laurie_role_1997, 21].^[Laurie and Cole (1997, p. 22) also note that only 2,105 officers remained in the Army after 1871 - a figure that did not change by more than 100 men until the close of the century.]

As it dwindled in numbers, the regular Army also had to contend with those who opposed reforming the Army. Many opponents of reform of the regular Army believed that the volunteer concept and individual state militias, on the whole, had served the nation well, and thus, that efforts to institute reforms for the regular Army were not required. Many of these opponents had at least a moderate degree of political traction because they had themselves served as volunteers in the Civil War, many achieving high rank and acclaim. This was perhaps most true of Senator John Logan, who in 1870 served as the House Military Affairs Chairman, and who had been passed over for command of the Army of the Tennessee by General William Sherman in 1864 [@clark_preparing_2017, 100]. 

The reforms sought by senior Army leaders would vary throughout the late 19th century, and would meet with varying degrees of success and failure. Early on, the intellectual question at stake between Sherman and opponents such as Logan was the degree to which military officers required special training and preparation in order to hone battlefield skills. Logan and others believed, based on their own experiences, that "the overwhelming determent of success as a commander was natural 'genius'," and thus they opposed the creation of the special professional military education programs Sherman and others sought [@clark_preparing_2017, 100].^[Key figures who advocated for reforms on behalf of the regular Army included, in addition to General Sherman, Colonel Emory Upton and Major General John M. Schofield. For excellent biographies of these three individuals, respectively, see @marszalek_sherman_2007; @connelly_john_2006; and @fitzpatrick_emory_2017.] Other significant questions about the Army during peacetime, including how the Army should man itself and against whom it should prepare to defend [@grandstaff_preserving_1998, 536-537]. The broader point is that in the early years after the Civil War, not everyone agreed on what purpose the Army served in the first place, and how to achieve that purpose in the absence of a clear international threat.

The Army's relationship with society in general also weakened as a result of its involvement in Reconstruction tasks and activities. Though it is difficult to precisely gauge the impact of Reconstruction on the military's relationship with society, we can reasonably ascertain that the Army was most disliked and distrusted by Southern Whites. To be clear, and as the front part of this case study made clear, there was significant variation in the way that individual military commanders in charge of federal forces exercised authorities over southern civilian governments. However, Byler (2006) notes that of particular concern was the way that federal forces had been used at or near voting locations during elections, particularly after 1870, when the military districts created by the Military Reconstruction Acts of 1867 ceased to formally function [@byler_civil-military_2006, 8]. The Army had also become increasingly scattered in a geographical sense. Laurie and Cole (1997) note, for instance, that upon the end of Reconstruction, "all of the cavalry forces and three fourths of the infantry" had moved to geographical outposts "west of the Mississippi Rive to battle hostile Indians and otherwise police the frontier" [@laurie_role_1997, 22]. For many Americans, the lengthened political involvement of the Army during Reconstruction, as well as its central role in instituting contentious policies against Native Americans, rendered the Army increasingly unpopular [@byler_civil-military_2006, 10]. Suspecting the impact of these developments on the Army's relationship with American society, Major General John Pope, while speaking to a gathering of Civil War Veterans in 1873, expressed concern that the Army might "properly and naturally become an object of suspicion and dislike" in the years to come [Cozzens, 2000 as cited in @byler_civil-military_2006, 1].

Like Pope, other Army leaders concluded by the mid-late 1870s that the Army occupied a less central position for the nation than it had previously occupied at the close of the Civil War. They did not like this development, nor accept it, but nonetheless, most of them, at the very minimum, were cognizant of it. For instance, General Sherman, in a private letter to his son in 1877, said, "our enemy is not now the Indians, but Congress...it is simply shameful that our pay and our Army organization should depend on the very men we whipped 12 years ago - Such however is the manner of a Representative Government" [Sherman, 1877 as cited in @marszalek_sherman_2007, 432]. Furthermore, Cooper (1980) notes that Major General O.O. Howard decried the fact that the public no longer praised and supported the Army, and lamented that "our people, little by little, have fallen back into the old ways of doing and thinking" [Howard, 1897 as cited in @cooper_army_1980, 31]. These sentiments are important because they reflect an acknowledgement by senior Army leaders that the institution was no longer operating from the same position it had occupied at the end of the Civil War. 

My argument is that this relative drop in the prestige of the military between 1868 and 1877 also impacted the form and quality of the political behaviors that civilian and military actors engaged in that involved the military between 1877 and 1881. Whereas President Johnson and his opponents in Congress regularly jockeyed for the favor of Grant and other senior military leaders and to portray the military's acquiescence to the public, civilian leaders from 1877 - 1881 rarely sought to harness the Army for their own partisan political advantages. Military leaders, on the other hand, when they did engage in political activities from 1877 - 1881, engaged in political behaviors that were far less visible and obvious than those that military leaders had engaged in from 1865 - 1868. 

## How Military Prestige Shaped Political Behaviors Involving the Military, 1877-1881

In this section, I narrate two related events or developments that occurred in the period from 1877 - 1881. I then describe the political behaviors that each actor type, military and civilian, undertook as these developments unfolded. Even as the level of polarization remained very high during this period, civilian actors did not attempt to harness the military and its leaders to the same degree that it had a decade earlier, and military leaders, cognizant that the role of the Army had changed significantly, shifted to engaging in political behaviors that were relatively covert and less visible to the public.

The first development was the use of the Army to quell what would end up becoming several domestic disturbances.^[Cooper (1980) and Laurie and Cole (1997) chronicle the use of the military in responding to these domestic incidents, which occurred from the late 1870s - 1890s. I do not address all of these particular instances in this chapter. See @cooper_army_1980 and @laurie_role_1997 for excellent and in-depth accounts of these events.] The first of these disturbances is generally referred to as _The Great Railway Strike of 1877_, which began in Maryland in July of that year before spreading to West Virginia, Pennsylvania, and ultimately, to Chicago. This and other subsequent strikes occurred as workers, particularly in the rail and coal industries, protested against what they perceived to be unfair and inequitable working conditions, to include the enaction of several wage cuts after the widespread economic downturn of 1873 [@cooper_army_1980, 43-45]. At first, state and local officials tried to respond to these strikes with local law enforcement and state National Guard units, but these proved woefully insufficient to handle the strikers, who had not only ceased all train traffic, but in some cases, had turned violent [@cooper_army_1980, 43-45; @laurie_role_1997, 30-32]. Despite having some initial hesitancy about involving the Army in what might be perceived as a "class war", President Hayes dispatched the Regular Army to maintain law and order [@laurie_role_1997, 32]. 

The timing of this strike intersected with an Army whose political position had been weakened in important ways. To begin with, Congress had failed to pass an appropriations bill funding the Army past July 1, meaning that the Army was essentially serving without pay. The failed appropriations bill, moreover, was a deliberate orchestration by southern Democrats against both the Army and Republicans for how the Army had been used throughout the Reconstruction era, and in particular, its use to secure polling sites during the contested Presidential Election of 1876 that had occurred just 8 months prior [@cooper_army_1980, 47-48; @laurie_role_1997, 31-32]. When Congressional Republicans failed to commit to any meaningful restrictions on the Army's use domestically after the contested elections, democrats "blocked the 1877 appropriations bill" [@clark_preparing_2017, 115].

The Army was generally successful in quelling the _Great Strike of 1877_, and many citizens and press outlets perceived the Army as such [@cooper_army_1980, 60-61; @laurie_role_1997, 52-55]. General Winfield Scott Hancock, who had commanded the Army's response in the East to this disturbance, likewise thought that the Army had performed well, stating in a private letter to his fellow General, John Schofield, that, "the troops have lost the government no prestige...it was the moral force of the United States government that was displayed - not its physical force" [Hancock, 1877 as cited in @cooper_army_1980, 61]. Yet even after this event in which the Army was perceived to have performed admirably, neither military nor civilian leaders engaged in visible political behaviors touting its success. The general political environment remained apathetic towards the Army. Not until 1894, when the next "great nationwide strike" occurred, did Congress allocate any additional funds for the purpose of dealing with domestic labor disputes [@laurie_role_1997].^[In 1894, the _Pullman Strike_ originated in the Pacific Northwest, in Coeur d'Alene, Idaho. See @laurie_role_1997, 163-178; @cooper_army_1980, 165-196 for more details.] 

The second development involves legislative provisions involving the Army undertaken by Congress the following year in 1878. Indeed, the army appropriations bill passed by Congress in 1878 undertook two important provisions. First, at the behest of Congressional Democrats, the bill instituted the _posse comitatus_ law. In response to what they perceived to be the heavy handed and inappropriate use of the Army throughout the Reconstruction period, Democrats "forbade the use of the Army as a posse to aid local officials in dealing with domestic turmoil without the express order of the President, and only then for very specific purposes" [@cooper_army_1980, 83]. Second, the 1878 army appropriations act established a bipartisan committee to examine "military organization and reform" [@clark_preparing_2017, 116]. The committee came to be known as the _Burnside Commission,_ chaired by Senator Ambrose Burnside, a former Union General, and would consider a number of important Army reforms, including its size, the number of officers, organization, promotion, retirement, and the commanding general's relationship to the Secretary of War.^[Fitzpatrick (2017) and Clark (2017) are among the contemporary historians who write about these reforms extensively. By the late 1870s, the combined team of General Sherman and Lieutenant Colonel (and later, Colonel) Emory Upton was well underway in laying the intellectual framework for a number of important ideas about the Army in peacetime, including how the Army might rapidly expand if a large war broke out. See also Connelly (2006).]

The Burnside proposal addressed Army reform from a number of perspectives, but it ultimately failed to pass. The proposal itself was somewhat of a compromise in that it recommended to maintain the overall size of the Army, while reducing the overall number of officers, including the number of generals [@fitzpatrick_emory_2017; @clark_preparing_2017, 118-121]. The Army did not receive the proposed legislation uniformly, either. Some leaders, such as Generals Sherman and Schofield, cautiously and perhaps reluctantly approved of the legislation, believing that it would provide the best chance for the Army to obtain any meaningful reforms at any point in the foreseeable future, while others, such as Sheridan, did not ultimately support the legislation [@clark_preparing_2017, 118-121]. Several historians explain the legislation's ultimate failure as reflective of the overall national sentiment at the time, which as Francis V. Greene would note nearly four decades later, was that, "this country [the US] would never be engaged in war; and therefore, in the midst of such pressing problems as the building of railroads, the resumption of specie payments, the silver question, and the tariff, there was no time to think about the needs of the Army. There was no 'military situation in the United States' worth thinking about [Greene, 1919 as cited in @fitzpatrick_emory_2017, 223]. 

What is especially important to note, however, is _how_ military and civilian actors engaged in political behaviors to support and/or defeat the proposed legislation, and what forms of political behaviors these actors undertook. Some military actors indeed did lobby on behalf of the proposed legislation, both in support of and in opposition to the bill. When they did, however, these military actors addressed primarily military audiences, and in some cases, politicians directly, and rarely addressed the general public (although there were some exceptions, which I note below). Other senior Army leaders also remained notably quiet altogether about the legislation, and did not engage in significant efforts either to ensure the bill's passage or its defeat. Furthermore, these political activities undertaken by military actors as a response to the Burnside reforms rarely, if ever, violated the principles of civil-military relations.

To begin with, a number of officers wrote letters to the Burnside Committee expressing their opinions on the proposed changes. For the most part, the letters these officers wrote constituted totally appropriate behavior in that the officers were responding to legitimate Congressional inquiry over topics that involved the military. Some officers went above and beyond the writing of letters in response to Congressmen, however.  Fitzpatrick (2006) notes, for instance, that some staff officers in Washington D.C., who in particular were concerned that the legislation would reduce their authority and influence by aligning the army staff bureaus under the commanding general, produced and distributed informational pamphlets to members of congress [@fitzpatrick_emory_2017, 221]. Senator Burnside himself, upon realizing that his legislation was stalling, noted, "I must say that some of these staff officers have gone beyond the line of duty, particularly...in Washington" [Burnside, 1878 as cited in @fitzpatrick_emory_2017, 221]. These behaviors can be viewed as acts of _shoulder tapping_ and _alliance building_ [@brooks_militaries_2009]. They also threatened the principle of non-partisanship, and perhaps to a lesser degree, the principle of civilian control. 

Military actors also engaged in a series of public appeals regarding the legislation, but these appeals were qualitatively distinct from those made during the period from 1865 - 1868. For instance, General Sherman passed on letters written by Generals Schofield, Sheridan, and Pope, as well as Lieutenant Colonel Upton, to the _Army and Navy Journal_ [@connelly_john_2006; 243]. Two things are worth noting about this. The first is that Sherman's behavior here should not be considered a leak per se, but rather an action undertaken with the author's concurrence and for the purposes of informing rather than influencing. The second thing to note is that the _Army and Navy Journal_ was a periodical intended primarily for consumption by military audiences. Sherman also made it clear, however, that though he thought the legislation contained more good than it did bad, he would not actively lobby on behalf of the bill [@connelly_john_2006; 243; @clark_preparing_2017, 118]. 

## Summary of the 1877-1881 Period

Give a summary. 

# Implications of this Case Study

\newpage

\singlespace